---
title: "Reading WordNet synsets"
subtitle: "Treat word meaning as a set, not a spelling"
author: "Alex Farach"
description: "Learn how WordNet groups words into synsets, why one word can cover many senses, and where a curated lexical resource helps or fails."
categories: [similarity, WordNet, word senses]
---

A museum cataloguer labels a drawer `bank` and then hesitates. The drawer might
hold river photographs, money records, or a toy building with a coin slot.

The label is a single string. The meaning depends on the sense attached to it.

A **synset** is a set of words that share one sense. WordNet is a hand-built
lexical database that records those sets and links them to broader meanings such
as **hypernyms**, the "is-a" parents in a hierarchy.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- define a synset in plain language;
- count senses for a word string;
- compare a sense count with a simple chance baseline;
- list synonyms inside one synset;
- walk hypernym links up an is-a hierarchy; and
- name what WordNet can and cannot support.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the WordNet extract

The lesson reads four local CSV files made from Open English WordNet 2024. No
code downloads WordNet while the page renders. The metadata file records the
source, licence, and SHA-256 fingerprints.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)
library(knitr)

source("R/permutation-null.R")

senses <- read_csv(
  "data/wordnet/wordnet-senses.csv",
  na = character(),
  col_types = cols(
    lemma = col_character(),
    part_of_speech = col_character(),
    sense_id = col_character(),
    synset_id = col_character(),
    sense_order = col_integer()
  )
)

synsets <- read_csv(
  "data/wordnet/wordnet-synsets.csv",
  na = character(),
  col_types = cols(
    synset_id = col_character(),
    part_of_speech = col_character(),
    definition = col_character(),
    example = col_character()
  )
)

members <- read_csv(
  "data/wordnet/wordnet-members.csv",
  na = character(),
  col_types = cols(
    synset_id = col_character(),
    lemma = col_character()
  )
)

relations <- read_csv(
  "data/wordnet/wordnet-relations.csv",
  na = character(),
  col_types = cols(
    synset_id = col_character(),
    relation = col_character(),
    target_synset_id = col_character(),
    target_definition = col_character(),
    target_members = col_character()
  )
)

metadata <- read_csv(
  "data/wordnet/wordnet-metadata.csv",
  na = character(),
  col_types = cols(
    field = col_character(),
    value = col_character()
  )
)

metadata_summary <- tibble(
  item = c("lemmas", "senses", "synsets", "synset membership rows", "relations"),
  value = c(
    n_distinct(senses$lemma),
    nrow(senses),
    nrow(synsets),
    nrow(members),
    nrow(relations)
  )
)

kable(
  metadata_summary,
  col.names = c("Item", "Count"),
  caption = "Size of the local WordNet extract used in this lesson",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(senses) > 400L,
  n_distinct(senses$lemma) == 38L,
  nrow(synsets) > 400L,
  nrow(members) > nrow(synsets),
  sum(relations$relation == "hypernym") > 300L,
  sum(relations$relation == "hyponym") > sum(relations$relation == "hypernym"),
  metadata$value[metadata$field == "source_version"] == "Open English Wordnet 2024"
)
```

This slice covers 38 lemmas, not all of English. It is enough to show how a
lexical database separates spelling from meaning.

## Read the senses of one word

A **sense** is one listed meaning of a word. The string `bank` has 18 senses in
this extract: 10 noun senses and 8 verb senses. The first noun senses split
river land, finance, piles, rows, reserves, and gambling funds.

```{r}
bank_senses <- senses |>
  filter(lemma == "bank") |>
  count(part_of_speech, name = "senses") |>
  arrange(part_of_speech)

bank_noun_senses <- senses |>
  filter(lemma == "bank", part_of_speech == "noun") |>
  arrange(sense_order) |>
  left_join(synsets, by = c("synset_id", "part_of_speech")) |>
  slice_head(n = 6) |>
  transmute(
    sense = sense_order,
    definition
  )

lemma_sense_counts <- senses |>
  count(lemma, name = "sense_count")

bank_sense_null <- permutation_null(
  observed = sum(senses$lemma == "bank"),
  replicate_fn = function() {
    lemma_sense_counts$sense_count[
      match(sample(lemma_sense_counts$lemma, 1L), lemma_sense_counts$lemma)
    ]
  },
  replicates = 1000L,
  seed = 5801L,
  alternative = "greater"
)

kable(
  bank_noun_senses,
  col.names = c("Noun sense", "Definition"),
  caption = "The first six noun senses of bank in the local WordNet extract",
  row.names = FALSE
)

kable(
  bank_sense_null |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  col.names = c("Observed", "Null mean", "Null 5th pct", "Null 95th pct", "p-value", "Replicates", "Alternative"),
  caption = "Sense count for bank compared with random lemmas in the extract",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  sum(bank_senses$senses) == 18L,
  identical(bank_senses$part_of_speech, c("noun", "verb")),
  identical(bank_senses$senses, c(10L, 8L)),
  nrow(bank_noun_senses) == 6L,
  any(str_detect(bank_noun_senses$definition, "financial institution")),
  any(str_detect(bank_noun_senses$definition, "body of water")),
  bank_sense_null$observed == 18,
  bank_sense_null$null_high > bank_sense_null$observed,
  bank_sense_null$p_value > 0.05
)
```

A word string is not a unit of meaning. If a search, model, or count treats every
`bank` as the same thing, it has made a simplifying choice. The null table adds a
second caution: this extract was built from ambiguous teaching examples, so 18
senses for `bank` is not unusual inside this small file.

Any lexical method will return an answer for a known word. The safer habit is to
ask what answer appears after the target word is compared with other words drawn
from the same source.

## Look inside a synset

For the financial-institution sense, the synset contains four member lemmas.
These are four names WordNet links to the same sense.

```{r}
financial_bank_synset <- senses |>
  filter(lemma == "bank", part_of_speech == "noun", sense_order == 2L) |>
  pull(synset_id)

financial_bank_members <- members |>
  filter(synset_id == financial_bank_synset) |>
  arrange(lemma) |>
  mutate(role = if_else(lemma == "bank", "queried word", "other member"))

kable(
  financial_bank_members,
  col.names = c("Synset ID", "Lemma", "Role"),
  caption = "Members of the financial-institution synset for bank",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  length(financial_bank_synset) == 1L,
  nrow(financial_bank_members) == 4L,
  "bank" %in% financial_bank_members$lemma,
  "depository financial institution" %in% financial_bank_members$lemma,
  sum(financial_bank_members$role == "other member") == 3L
)
```

Synonymy here is local to one sense. `bank` and `banking company` share the
finance sense; `river bank` belongs elsewhere.

## Walk an is-a chain

A hypernym says that one meaning is a kind of another meaning. In this extract,
the professor-position sense of `chair` walks upward through a position synset,
then through a work synset, then to `activity`.

```{r}
first_hypernym <- function(synset) {
  next_row <- relations |>
    filter(synset_id == synset, relation == "hypernym") |>
    arrange(target_synset_id) |>
    slice_head(n = 1)

  if (nrow(next_row) != 1L) {
    stop("No hypernym recorded for synset: ", synset, call. = FALSE)
  }

  next_row
}

select_chair_synset <- function() {
  candidates <- senses |>
    filter(lemma == "chair", part_of_speech == "noun") |>
    left_join(synsets, by = c("synset_id", "part_of_speech")) |>
    filter(str_detect(definition, regex("professorship|professor", ignore_case = TRUE))) |>
    arrange(sense_order)

  if (nrow(candidates) != 1L) {
    stop("Expected one professor-position chair sense in this extract.", call. = FALSE)
  }

  candidates$synset_id[[1]]
}

walk_hypernyms <- function(start_synset, steps) {
  path <- start_synset

  for (step in seq_len(steps)) {
    path <- c(path, first_hypernym(path[length(path)])$target_synset_id)
  }

  path
}

lookup_definition <- function(id) {
  local_definition <- synsets$definition[synsets$synset_id == id]
  if (length(local_definition) > 0L) {
    return(local_definition[[1]])
  }

  external_definition <- relations$target_definition[match(id, relations$target_synset_id)]
  if (is.na(external_definition)) {
    stop("No definition available for synset: ", id, call. = FALSE)
  }

  external_definition
}

lookup_members <- function(id) {
  local_members <- members |>
    filter(synset_id == id) |>
    arrange(lemma) |>
    pull(lemma)

  if (length(local_members) > 0L) {
    return(paste(local_members, collapse = ", "))
  }

  external_members <- relations$target_members[match(id, relations$target_synset_id)]
  if (is.na(external_members)) {
    stop("No members available for synset: ", id, call. = FALSE)
  }

  external_members
}

chair_synset <- select_chair_synset()
path_ids <- walk_hypernyms(chair_synset, steps = 3L)

hypernym_path <- tibble(
  level = seq_along(path_ids) - 1L,
  synset_id = path_ids,
  members = vapply(path_ids, lookup_members, character(1)),
  definition = vapply(path_ids, lookup_definition, character(1))
)

kable(
  hypernym_path,
  col.names = c("Level", "Synset ID", "Members", "Definition"),
  caption = "A four-step hypernym path from chair to activity",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  chair_synset == "oewn-00599171-n",
  identical(
    hypernym_path$synset_id,
    c("oewn-00599171-n", "oewn-00587299-n", "oewn-00583425-n", "oewn-00408356-n")
  ),
  all(c("chair, professorship", "activity") %in% unname(hypernym_path$members)),
  nrow(hypernym_path) == 4L
)
```

The walk assumes this extract contains the chosen `chair` sense and three
hypernym links above it. The functions stop when that assumption is false, so a
missing lemma cannot produce a half-empty path that looks valid.

The hierarchy is useful for broadening a search. A reader looking for occupations
might want `chair`, `office`, and `job` to meet at a broader node. The same
hierarchy can be too coarse for a task that needs the academic rank, the
furniture sense, or a named committee role.

## Keep the limits attached

WordNet is a curated snapshot with a version and a licence. It is not a direct
measurement of English. It covers much common vocabulary, while domain terms,
slang, new coinages, and local uses can be missing.

Sense inventories also draw boundaries. A fine split that helps one annotation
task can be too fine for another, so a WordNet sense should be treated as a
chosen inventory entry rather than a final answer about meaning.

WordNet is still good for teaching the difference between words and meanings,
expanding a controlled search, checking whether two labels share a listed sense,
and building small rule-based features where coverage can be inspected.

```{r}
limit_check <- tibble(
  check = c("version recorded", "licence recorded", "target lemmas recorded"),
  value = c(
    metadata$value[metadata$field == "source_version"],
    metadata$value[metadata$field == "source_licence"],
    metadata$value[metadata$field == "n_lemmas_found"]
  )
)

kable(
  limit_check,
  col.names = c("Check", "Value"),
  caption = "Provenance fields carried with the WordNet extract",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  limit_check$value[limit_check$check == "version recorded"] == "Open English Wordnet 2024",
  limit_check$value[limit_check$check == "licence recorded"] == "https://creativecommons.org/licenses/by/4.0",
  as.integer(limit_check$value[limit_check$check == "target lemmas recorded"]) == n_distinct(senses$lemma)
)
```

## What to remember

- A synset is a set of words sharing one sense.
- `bank` has 18 senses in this extract, so one spelling is not one meaning.
- A simple null check keeps the sense count from sounding stronger than it is.
- Hypernym links make an is-a hierarchy that can broaden a search.
- WordNet is useful when a curated inventory is acceptable and limited coverage
  can be inspected.

The cataloguer keeps the `bank` drawer only after adding sense labels to the
folders inside it.

## Sources

- [Open English WordNet](https://en-word.net/)
- [Open English WordNet GitHub repository](https://github.com/globalwordnet/english-wordnet)
- [Creative Commons Attribution 4.0 International licence](https://creativecommons.org/licenses/by/4.0/)
- [Princeton WordNet project](https://wordnet.princeton.edu/)
