---
title: "Measuring distance between texts"
subtitle: "Compare string distance with vector distance"
description: "Learn how string and vector distances answer different similarity questions, and why raw-count Euclidean distance tracks document length."
categories: [similarity, distance measures, inaugural addresses]
---

A library volunteer receives two boxes of index cards. One box has typed names
with small spelling errors, and the other has paragraphs from speeches.

The word "similar" is doing two jobs. Matching a mistyped name is not the same
task as comparing the vocabulary of two paragraphs.

A **distance measure** turns a comparison into a number. Small distance means
close. The hard part is choosing a distance that matches the question.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- compare edit distance and Jaro-Winkler distance for names;
- build paragraph vectors from word counts;
- compare cosine, Euclidean, and Jaccard distance;
- test whether a distance mostly tracks text length; and
- explain why the same data can give different rankings.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Compare strings directly

**String distance** compares characters. Levenshtein distance counts edits:
insertions, deletions, and substitutions. Jaro-Winkler distance gives extra
credit to strings that share an early prefix, which often helps with names.

```{r}
library(dplyr)
library(tibble)
library(stringr)
library(stringdist)
library(quanteda)
library(Matrix)
library(ggplot2)
library(knitr)

source("R/permutation-null.R")

within_tolerance <- function(observed, expected, tolerance = 1e-3) {
  abs(observed - expected) < tolerance
}

name_matches <- tibble(
  recorded = c("Jonson", "Jonson", "Martha", "Martha", "Stevenson", "Stevenson"),
  candidate = c("Johnson", "Jansen", "Marta", "Marsha", "Stephenson", "Steven")
) |>
  mutate(
    levenshtein = stringdist(recorded, candidate, method = "lv"),
    jaro_winkler = stringdist(recorded, candidate, method = "jw", p = 0.1),
    levenshtein_rank = rank(levenshtein, ties.method = "first"),
    jaro_winkler_rank = rank(jaro_winkler, ties.method = "first")
  )

stevenson_rows <- name_matches |>
  filter(recorded == "Stevenson")

kable(
  name_matches |>
    mutate(jaro_winkler = round(jaro_winkler, 4)),
  col.names = c("Recorded", "Candidate", "Levenshtein", "Jaro-Winkler", "Levenshtein rank", "Jaro-Winkler rank"),
  caption = "Two string distances rank the same candidate names differently",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(name_matches) == 6L,
  identical(stevenson_rows$candidate, c("Stephenson", "Steven")),
  identical(stevenson_rows$levenshtein, c(2, 3)),
  all(within_tolerance(stevenson_rows$jaro_winkler, c(0.0726, 0.0667), tolerance = 5e-4)),
  stevenson_rows$candidate[which.min(stevenson_rows$levenshtein)] == "Stephenson",
  stevenson_rows$candidate[which.min(stevenson_rows$jaro_winkler)] == "Steven"
)
```

For `Stevenson`, Levenshtein picks `Stephenson` because it needs fewer edits.
Jaro-Winkler picks `Steven` because the shared opening is strong. A name-matching
project has to choose the error pattern it cares about.

## Build paragraph vectors

A **vector** is a row of numbers. In a document-feature matrix, each row is a
paragraph and each column is a word feature. The entry says how many times that
word appears in that paragraph.

```{r}
source("R/inaugural-corpus.R")

paragraphs <- inaugural_paragraphs()

paragraph_tokens <- tokens(
  paragraphs$paragraph,
  remove_punct = TRUE,
  remove_numbers = TRUE
) |>
  tokens_tolower()

paragraph_dfm <- dfm(paragraph_tokens)
docnames(paragraph_dfm) <- paragraphs$paragraph_id

count_matrix <- as(paragraph_dfm, "dgCMatrix")
token_lengths <- Matrix::rowSums(count_matrix)
squared_lengths <- Matrix::rowSums(count_matrix^2)
dot_products <- as.matrix(Matrix::tcrossprod(count_matrix))

euclidean_squared <- outer(squared_lengths, squared_lengths, "+") - 2 * dot_products
euclidean_squared[euclidean_squared < 0 & euclidean_squared > -1e-8] <- 0
euclidean_distance <- sqrt(euclidean_squared)

vector_norms <- sqrt(squared_lengths)
cosine_similarity <- dot_products / outer(vector_norms, vector_norms)
cosine_distance <- 1 - cosine_similarity

binary_matrix <- Matrix::drop0(count_matrix)
binary_matrix@x <- rep(1, length(binary_matrix@x))
feature_counts <- Matrix::rowSums(binary_matrix)
intersections <- as.matrix(Matrix::tcrossprod(binary_matrix))
unions <- outer(feature_counts, feature_counts, "+") - intersections
jaccard_distance <- 1 - intersections / unions

pair_index <- which(upper.tri(euclidean_distance), arr.ind = TRUE)
pair_distances <- tibble(
  first = pair_index[, 1],
  second = pair_index[, 2],
  length_difference = abs(token_lengths[first] - token_lengths[second]),
  euclidean = euclidean_distance[pair_index],
  cosine = cosine_distance[pair_index],
  jaccard = jaccard_distance[pair_index],
  first_speech = paragraphs$speech_id[first],
  second_speech = paragraphs$speech_id[second]
)

kable(
  tibble(
    item = c("Paragraphs", "Speeches", "People", "Features", "Paragraph pairs"),
    value = c(nrow(paragraphs), n_distinct(paragraphs$speech_id), n_distinct(paragraphs$president), nfeat(paragraph_dfm), nrow(pair_distances))
  ),
  col.names = c("Item", "Value"),
  caption = "Paragraph vectors built from the inaugural-address corpus",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(paragraphs) == 1377L,
  n_distinct(paragraphs$speech_id) == 60L,
  n_distinct(paragraphs$president) == 40L,
  nfeat(paragraph_dfm) == 9327L,
  nrow(pair_distances) == 947376L,
  as.integer(min(token_lengths)) == 22L,
  as.integer(max(token_lengths)) == 984L,
  all(is.finite(pair_distances$euclidean)),
  all(pair_distances$cosine >= 0 & pair_distances$cosine <= 1 + 1e-8)
)
```

The matrix has 1,377 paragraph rows and 9,327 word columns. Each paragraph
becomes a point in a high-dimensional space.

## Compare three vector distances

Euclidean distance measures straight-line distance between count vectors. Cosine
distance compares direction, so multiplying a paragraph vector by the same
constant leaves its cosine direction unchanged. Jaccard distance uses word
presence and ignores repeated counts.

```{r}
length_correlations <- pair_distances |>
  summarise(
    `Euclidean distance` = cor(euclidean, length_difference),
    `Cosine distance` = cor(cosine, length_difference),
    `Jaccard distance` = cor(jaccard, length_difference)
  ) |>
  tidyr::pivot_longer(
    everything(),
    names_to = "distance",
    values_to = "correlation_with_length_difference"
  )

same_speech_examples <- pair_distances |>
  mutate(
    first_paragraph = paragraphs$paragraph_id[first],
    second_paragraph = paragraphs$paragraph_id[second],
    first_length = as.integer(token_lengths[first]),
    second_length = as.integer(token_lengths[second])
  ) |>
  filter(first_speech == second_speech) |>
  arrange(desc(length_difference), euclidean) |>
  slice_head(n = 3) |>
  transmute(
    first_paragraph,
    second_paragraph,
    speech = first_speech,
    first_length,
    second_length,
    length_difference = as.integer(length_difference),
    euclidean = round(euclidean, 2),
    cosine = round(cosine, 3),
    jaccard = round(jaccard, 3)
  )

kable(
  length_correlations |>
    mutate(correlation_with_length_difference = round(correlation_with_length_difference, 3)),
  col.names = c("Distance", "Correlation with length difference"),
  caption = "Raw-count distance compared with paragraph length difference",
  row.names = FALSE
)

kable(
  same_speech_examples,
  col.names = c("First paragraph", "Second paragraph", "Speech", "First length", "Second length", "Length difference", "Euclidean", "Cosine", "Jaccard"),
  caption = "Long and short paragraphs from the same speech can be far apart under Euclidean distance",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(within_tolerance(length_correlations$correlation_with_length_difference, c(0.951, -0.268, 0.275), tolerance = 1e-3)),
  same_speech_examples$first_paragraph[1] == "1841-Harrison-p10",
  same_speech_examples$second_paragraph[1] == "1841-Harrison-p22",
  same_speech_examples$length_difference[1] == 915L,
  within_tolerance(same_speech_examples$euclidean[1], 147.46, tolerance = 0.01),
  within_tolerance(same_speech_examples$cosine[1], 0.330, tolerance = 0.001)
)
```

The Euclidean correlation with length difference is 0.951, a strong warning that
raw-count distance is mostly measuring paragraph size. The cosine correlation is
negative, -0.268, because longer paragraphs share many ordinary words with other
long paragraphs. That matters for this corpus: nineteenth-century paragraphs are
longer than later paragraphs, so a length-shaped method can echo the era pattern
seen in the corpus lessons.

```{r}
same_speech_observed <- pair_distances |>
  filter(first_speech == second_speech) |>
  summarise(mean_cosine_distance = mean(cosine)) |>
  pull(mean_cosine_distance)

speech_labels <- paragraphs$speech_id
same_speech_null <- permutation_null(
  observed = same_speech_observed,
  replicate_fn = function() {
    shuffled_labels <- sample(speech_labels)
    pair_distances |>
      filter(shuffled_labels[first] == shuffled_labels[second]) |>
      summarise(mean_cosine_distance = mean(cosine)) |>
      pull(mean_cosine_distance)
  },
  replicates = 1000L,
  seed = 5902L,
  alternative = "less"
)

kable(
  same_speech_null |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  col.names = c("Observed", "Null mean", "Null 5th pct", "Null 95th pct", "p-value", "Replicates", "Alternative"),
  caption = "Same-speech paragraph cosine distance compared with shuffled speech labels",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  same_speech_null$observed < same_speech_null$null_low,
  same_speech_null$p_value < 0.01,
  same_speech_null$replicates == 1000L
)
```

A distance formula will fill a matrix for any paragraphs it receives. The
permutation check asks whether paragraphs from the same speech are closer than a
label shuffle would produce. Here they are, but the raw-count Euclidean result is
still dominated by length.

```{r}
#| label: fig-length-distance
#| fig-cap: "Length difference and distance for a deterministic sample of paragraph pairs."
#| fig-alt: "A faceted scatter plot. Euclidean distance rises sharply with length difference, while cosine and Jaccard show much weaker patterns."

set.seed(5901)
plot_pairs <- pair_distances |>
  slice_sample(n = 3000) |>
  tidyr::pivot_longer(
    c(euclidean, cosine, jaccard),
    names_to = "distance",
    values_to = "value"
  )

ggplot(plot_pairs, aes(x = length_difference, y = value)) +
  geom_point(alpha = 0.18, size = 0.7) +
  facet_wrap(~distance, scales = "free_y") +
  labs(
    x = "Difference in paragraph length, in tokens",
    y = "Distance",
    title = "Raw Euclidean distance mostly follows length"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  nrow(plot_pairs) == 9000L,
  identical(sort(unique(plot_pairs$distance)), c("cosine", "euclidean", "jaccard")),
  max(plot_pairs$length_difference) > 900L
)
```

The plot turns the correlation into a visible warning. A long paragraph and a
short paragraph can look far apart under raw-count Euclidean distance even when
they come from the same speech.

## What to remember

- String distance compares character sequences.
- Vector distance compares numeric representations of text.
- Levenshtein and Jaro-Winkler can rank the same name candidates differently.
- On these paragraph counts, Euclidean distance correlates 0.951 with length
  difference.
- Cosine distance compares direction, which reduces the length problem for raw
  counts.

The volunteer keeps two labels on the boxes: spelling repair for names,
representation choice for paragraphs.

## Sources

- [stringdist package reference](https://cran.r-project.org/package=stringdist)
- [quanteda tokens documentation](https://quanteda.io/reference/tokens.html)
- [United States copyright guidance for government works](https://www.usa.gov/government-copyright)
