---
title: "Finding similar documents"
subtitle: "Nearest neighbours depend on the representation"
description: "Learn how speech-level document similarity changes under raw counts, tf-idf, and stopword removal."
categories: [similarity, document similarity, inaugural addresses]
---

A reading group wants one companion speech for a long weekend assignment. They
ask for the nearest neighbour to a selected address.

The request sounds simple until the group asks what "nearest" means. A method can
compare word counts, rare-word weights, or a version with common words removed.

**Document similarity** is a score that compares two documents after they have
been turned into features. The score is only as meaningful as that representation
for the question at hand.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- build speech-level document-feature matrices;
- compute cosine similarity across 60 speeches;
- list the most and least similar pairs under tf-idf;
- show that nearest neighbours change when features change;
- test an authorship pattern against shuffled labels; and
- notice when document length is part of the signal.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build speech vectors

**Tf-idf** stands for term frequency times inverse document frequency. It keeps
words from each speech, then gives more weight to words that are less common
across the full set of speeches.

```{r}
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(quanteda)
library(quanteda.textstats)
library(knitr)

source("R/inaugural-corpus.R")
source("R/permutation-null.R")

within_tolerance <- function(observed, expected, tolerance = 1e-3) {
  abs(observed - expected) < tolerance
}

paragraphs <- inaugural_paragraphs()

original_speeches <- tibble(
  speech_id = names(quanteda::data_corpus_inaugural),
  year = as.integer(quanteda::data_corpus_inaugural$Year),
  surname = as.character(quanteda::data_corpus_inaugural$President),
  first_name = as.character(quanteda::data_corpus_inaugural$FirstName),
  party = as.character(quanteda::data_corpus_inaugural$Party),
  text = as.character(quanteda::data_corpus_inaugural)
) |>
  mutate(
    president = str_squish(paste(first_name, surname)),
    era = factor(
      if_else(year < 1900, "before 1900", "1900 or later"),
      levels = c("before 1900", "1900 or later")
    ),
    original_words = str_count(text, "\\S+")
  )

retained_by_speech <- paragraphs |>
  group_by(speech_id) |>
  summarise(retained_words = sum(paragraph_words), .groups = "drop")

paragraph_era_summary <- paragraphs |>
  group_by(era) |>
  summarise(median_kept_paragraph_words = median(paragraph_words), .groups = "drop")

retention_by_era <- original_speeches |>
  select(speech_id, era, original_words) |>
  left_join(retained_by_speech, by = "speech_id") |>
  mutate(retained_words = coalesce(retained_words, 0L)) |>
  group_by(era) |>
  summarise(
    speeches = n(),
    original_words = sum(original_words),
    retained_words = sum(retained_words),
    share_retained = retained_words / original_words,
    .groups = "drop"
  ) |>
  left_join(paragraph_era_summary, by = "era")

speeches <- paragraphs |>
  group_by(speech_id, year, president, surname, party, era) |>
  summarise(
    text = paste(paragraph, collapse = "\n"),
    speech_tokens = sum(paragraph_words),
    .groups = "drop"
  )

make_speech_dfm <- function(remove_stops = FALSE, tfidf = FALSE) {
  speech_tokens <- tokens(
    speeches$text,
    remove_punct = TRUE,
    remove_numbers = TRUE
  ) |>
    tokens_tolower()

  if (remove_stops) {
    speech_tokens <- tokens_remove(speech_tokens, stopwords("en"))
  }

  speech_dfm <- dfm(speech_tokens)
  docnames(speech_dfm) <- speeches$speech_id

  if (tfidf) {
    speech_dfm <- dfm_tfidf(speech_dfm)
  }

  speech_dfm
}

raw_dfm <- make_speech_dfm(remove_stops = FALSE, tfidf = FALSE)
tfidf_dfm <- make_speech_dfm(remove_stops = FALSE, tfidf = TRUE)
tfidf_no_stop_dfm <- make_speech_dfm(remove_stops = TRUE, tfidf = TRUE)

representation_table <- tibble(
  representation = c("raw counts", "tf-idf", "tf-idf with stopwords removed"),
  documents = c(ndoc(raw_dfm), ndoc(tfidf_dfm), ndoc(tfidf_no_stop_dfm)),
  features = c(nfeat(raw_dfm), nfeat(tfidf_dfm), nfeat(tfidf_no_stop_dfm))
)

surname_collisions <- speeches |>
  distinct(president, surname) |>
  count(surname, name = "people") |>
  filter(people > 1L) |>
  arrange(surname)

identity_table <- tibble(
  item = c("speeches", "people", "surnames", "surnames used by two people"),
  value = c(nrow(speeches), n_distinct(speeches$president), n_distinct(speeches$surname), nrow(surname_collisions))
)

kable(
  representation_table,
  col.names = c("Representation", "Documents", "Features"),
  caption = "Three speech representations used for similarity comparisons",
  row.names = FALSE
)

kable(
  retention_by_era |>
    mutate(share_retained = round(share_retained, 3)),
  col.names = c("Era", "Speeches", "Original words", "Retained words", "Share retained", "Median kept paragraph words"),
  caption = "Words retained after rebuilding speeches from paragraphs of at least 25 words",
  row.names = FALSE
)

kable(
  identity_table,
  col.names = c("Item", "Count"),
  caption = "The corpus has 60 speeches by 40 people but only 36 surnames",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(paragraphs) == 1377L,
  nrow(speeches) == 60L,
  n_distinct(speeches$president) == 40L,
  n_distinct(speeches$surname) == 36L,
  identical(surname_collisions$surname, c("Adams", "Bush", "Harrison", "Roosevelt")),
  identical(representation_table$documents, c(60L, 60L, 60L)),
  identical(representation_table$features, c(9327L, 9327L, 9193L)),
  all(within_tolerance(retention_by_era$share_retained, c(0.998, 0.914), tolerance = 0.001)),
  identical(retention_by_era$median_kept_paragraph_words, c(112, 54))
)
```

The helper rebuilds each speech by joining only paragraphs with at least 25
words, so the "speeches" below are retained-paragraph versions of the addresses.
The filter keeps nearly all words before 1900 and a smaller share from 1900 or
later. The `president` field is the full name: 60 speeches come from 40 people,
while four surnames each refer to two people.

Comparing every speech with every other speech gives 1,770 pairs, because
`60 * 59 / 2 = 1,770`.

## Rank all speech pairs

Cosine similarity ranges from 0 to 1 for these nonnegative vectors. Larger values
mean the two speeches point in more similar vocabulary directions.

```{r}
cosine_matrix <- function(speech_dfm) {
  similarity_matrix <- as.matrix(
    textstat_simil(speech_dfm, method = "cosine", margin = "documents")
  )
  diag(similarity_matrix) <- NA_real_
  similarity_matrix
}

cosine_pairs <- function(speech_dfm) {
  similarity_matrix <- cosine_matrix(speech_dfm)
  pair_index <- which(upper.tri(similarity_matrix), arr.ind = TRUE)

  tibble(
    first = rownames(similarity_matrix)[pair_index[, 1]],
    second = colnames(similarity_matrix)[pair_index[, 2]],
    similarity = similarity_matrix[pair_index]
  ) |>
    left_join(
      speeches |> select(first = speech_id, first_year = year, first_president = president, first_tokens = speech_tokens),
      by = "first"
    ) |>
    left_join(
      speeches |> select(second = speech_id, second_year = year, second_president = president, second_tokens = speech_tokens),
      by = "second"
    )
}

tfidf_pairs <- cosine_pairs(tfidf_dfm) |>
  arrange(desc(similarity), first, second)

most_similar_pairs <- tfidf_pairs |>
  slice_head(n = 5) |>
  mutate(group = "Most similar")

least_similar_pairs <- tfidf_pairs |>
  arrange(similarity, first, second) |>
  slice_head(n = 5) |>
  mutate(group = "Least similar")

pair_table <- bind_rows(most_similar_pairs, least_similar_pairs) |>
  transmute(
    group,
    first,
    second,
    first_president,
    second_president,
    first_tokens,
    second_tokens,
    similarity = round(similarity, 4)
  )

least_with_short_washington <- least_similar_pairs |>
  filter(first == "1793-Washington" | second == "1793-Washington") |>
  nrow()

kable(
  pair_table,
  col.names = c("Group", "First speech", "Second speech", "First president", "Second president", "First tokens", "Second tokens", "Cosine similarity"),
  caption = "Most and least similar speech pairs under speech-level tf-idf",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(tfidf_pairs) == 1770L,
  tfidf_pairs$first[1] == "1817-Monroe",
  tfidf_pairs$second[1] == "1821-Monroe",
  within_tolerance(pair_table$similarity[1], 0.2903, tolerance = 1e-4),
  pair_table$first[6] == "1793-Washington",
  pair_table$second[6] == "1905-Roosevelt",
  within_tolerance(pair_table$similarity[6], 0.0066, tolerance = 1e-4),
  least_with_short_washington >= 4L
)
```

The strongest pair under this representation is the two Monroe speeches, with
cosine similarity about 0.2903. Four of the five weakest rows include the short
1793 Washington speech, which has 135 retained tokens. Shortness, not a reading
of subject matter, puts that speech at the bottom of the ranking.

```{r}
tfidf_similarity_matrix <- cosine_matrix(tfidf_dfm)

mean_tfidf_similarity <- tibble(
  speech_id = rownames(tfidf_similarity_matrix),
  mean_similarity = rowMeans(tfidf_similarity_matrix, na.rm = TRUE)
) |>
  left_join(speeches |> select(speech_id, speech_tokens), by = "speech_id")

length_similarity_table <- mean_tfidf_similarity |>
  summarise(
    measure = "Mean tf-idf similarity and speech token count",
    correlation = cor(mean_similarity, speech_tokens)
  )

kable(
  length_similarity_table |>
    mutate(correlation = round(correlation, 3)),
  col.names = c("Measure", "Correlation"),
  caption = "Mean tf-idf similarity compared with speech token count",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  within_tolerance(length_similarity_table$correlation, 0.861, tolerance = 0.001),
  length_similarity_table$correlation > 0.8
)
```

Mean tf-idf similarity still correlates strongly with speech length. Tf-idf
changes the representation, but it does not remove the length gradient from this
corpus.

## Change the representation

A nearest neighbour is the highest-scoring other document. The code below asks
for the nearest neighbour of the 1793 Washington speech three ways: raw counts,
tf-idf, and tf-idf after removing stopwords.

```{r}
nearest_for <- function(speech_dfm, target) {
  similarity_matrix <- cosine_matrix(speech_dfm)
  target_scores <- similarity_matrix[target, ]
  target_scores[target] <- NA_real_

  tibble(
    neighbour = names(target_scores),
    similarity = as.numeric(target_scores)
  ) |>
    arrange(desc(similarity), neighbour) |>
    slice_head(n = 1)
}

target_speech <- "1793-Washington"
nearest_table <- bind_rows(
  nearest_for(raw_dfm, target_speech) |>
    mutate(representation = "raw counts"),
  nearest_for(tfidf_dfm, target_speech) |>
    mutate(representation = "tf-idf"),
  nearest_for(tfidf_no_stop_dfm, target_speech) |>
    mutate(representation = "tf-idf, stopwords removed")
) |>
  left_join(
    speeches |> select(neighbour = speech_id, year, president, speech_tokens),
    by = "neighbour"
  ) |>
  transmute(
    target = target_speech,
    representation,
    neighbour,
    year,
    president,
    speech_tokens,
    similarity = round(similarity, 4)
  )

kable(
  nearest_table,
  col.names = c("Target", "Representation", "Nearest neighbour", "Year", "President", "Neighbour tokens", "Cosine similarity"),
  caption = "One target speech gets three nearest neighbours under three representations",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nearest_table$target, rep("1793-Washington", 3L)),
  identical(nearest_table$neighbour, c("1841-Harrison", "1861-Lincoln", "1885-Cleveland")),
  all(within_tolerance(nearest_table$similarity, c(0.8474, 0.0552, 0.0512), tolerance = 1e-4)),
  n_distinct(nearest_table$neighbour) == 3L
)
```

The three answers differ. The raw-count neighbour for the shortest speech is the
longest retained speech, and the score is about 0.8474. That column is dominated
by shared high-frequency words and length, so it should not be read as a topical
match.
```{r}
nearest_all <- function(speech_dfm) {
  similarity_matrix <- cosine_matrix(speech_dfm)

  map_dfr(
    rownames(similarity_matrix),
    function(target) {
      scores <- similarity_matrix[target, ]
      tibble(
        target,
        neighbour = names(scores)[which.max(scores)],
        similarity = max(scores, na.rm = TRUE)
      )
    }
  ) |>
    left_join(
      speeches |> select(target = speech_id, target_president = president, target_surname = surname),
      by = "target"
    ) |>
    left_join(
      speeches |> select(neighbour = speech_id, neighbour_president = president, neighbour_surname = surname),
      by = "neighbour"
    ) |>
    mutate(
      same_president = target_president == neighbour_president,
      same_surname = target_surname == neighbour_surname
    )
}

raw_neighbours <- nearest_all(raw_dfm)
raw_hub_table <- raw_neighbours |>
  count(neighbour, name = "nearest_neighbour_count", sort = TRUE) |>
  slice_head(n = 10)

harrison_hub_count <- raw_neighbours |>
  filter(neighbour == "1841-Harrison") |>
  nrow()

kable(
  raw_hub_table,
  col.names = c("Raw-count nearest neighbour", "Number of target speeches"),
  caption = "How often each speech is the nearest neighbour under raw counts",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(raw_neighbours) == 60L,
  max(raw_hub_table$nearest_neighbour_count) == 7L,
  harrison_hub_count == 2L,
  raw_hub_table$neighbour[1] == "1985-Reagan"
)
```

The raw-count hub check does not make `1841-Harrison` a corpus-wide hub; it is
the nearest neighbour for two speeches. The broader warning is still clear.
Several speeches become nearest neighbours for many targets under raw counts,
which is another sign that the column measures common-word mass more than a
careful vocabulary match.

## Separate vocabulary signal from authorship signal

When one person has more than one address in the corpus, repeated phrasing can
make those speeches neighbours. That can be useful for authorship or style
questions and distracting for topic questions.

```{r}
tfidf_neighbours <- nearest_all(tfidf_dfm)
same_president_count <- sum(tfidf_neighbours$same_president)

authorship_examples <- tfidf_neighbours |>
  filter(same_president) |>
  arrange(desc(similarity), target) |>
  slice_head(n = 6) |>
  transmute(
    target,
    neighbour,
    president = target_president,
    similarity = round(similarity, 4)
  )

same_president_observed <- tfidf_pairs |>
  filter(first_president == second_president) |>
  summarise(mean_similarity = mean(similarity)) |>
  pull(mean_similarity)

author_labels <- speeches$president
names(author_labels) <- speeches$speech_id

same_president_null <- permutation_null(
  observed = same_president_observed,
  replicate_fn = function() {
    shuffled_labels <- sample(author_labels)
    names(shuffled_labels) <- names(author_labels)
    tfidf_pairs |>
      filter(shuffled_labels[first] == shuffled_labels[second]) |>
      summarise(mean_similarity = mean(similarity)) |>
      pull(mean_similarity)
  },
  replicates = 1000L,
  seed = 6002L,
  alternative = "greater"
)

kable(
  authorship_examples,
  col.names = c("Target", "Nearest neighbour", "President", "Cosine similarity"),
  caption = "Examples where a speech's nearest tf-idf neighbour has the same president",
  row.names = FALSE
)

kable(
  same_president_null |>
    mutate(across(where(is.numeric), ~ round(.x, 3))),
  col.names = c("Observed", "Null mean", "Null 5th pct", "Null 95th pct", "p-value", "Replicates", "Alternative"),
  caption = "Same-president tf-idf similarity compared with shuffled author labels",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(tfidf_neighbours) == 60L,
  same_president_count == 19L,
  authorship_examples$target[1] == "1817-Monroe",
  authorship_examples$neighbour[1] == "1821-Monroe",
  within_tolerance(authorship_examples$similarity[1], 0.2903, tolerance = 1e-4),
  same_president_null$observed > same_president_null$null_high,
  same_president_null$p_value < 0.01
)
```

Under tf-idf, 19 of the 60 speeches have a nearest neighbour by the same person.
The label-shuffle null says same-person pairs are more similar than chance in
this corpus. That is evidence of repeated wording or style, not proof that the
speeches discuss the same subjects.

## What to remember

- Document similarity compares representations, not documents in the abstract.
- Speech-level tf-idf gives 1,770 pairwise comparisons for 60 speeches.
- The nearest neighbour of one speech changed across raw counts, tf-idf, and
  stopword removal.
- Same-person neighbours can show authorship or style rather than subject matter.
- Similarity functions return a ranking even when length or labels explain it.

The reading group chooses a companion speech only after naming what kind of
similarity it wants.

## Sources

- [quanteda document-feature matrix documentation](https://quanteda.io/reference/dfm.html)
- [quanteda.textstats similarity documentation](https://quanteda.io/reference/textstat_simil.html)
- [quanteda inaugural corpus documentation](https://quanteda.io/reference/data_corpus_inaugural.html)
- [United States copyright guidance for government works](https://www.usa.gov/government-copyright)
