---
title: "Training word vectors"
subtitle: "Check whether neighbours carry information"
description: "Learn how distributed word representations are trained from local context and why a small corpus can make neighbour lists look better than they are."
categories: [similarity, word embeddings, inaugural addresses]
---

A curator wants to tag a speech archive without reading every paragraph. She
asks which words sit near `freedom`, `union`, and `peace` in the language of the
collection.

A table of neighbours appears quickly. It looks meaningful, but every
word-vector model can return a nearest neighbour for any word it kept.

A **distributed word representation** is one numeric vector per word, learned
from the words that appear near it. This lesson loads one saved local model, then
checks whether an easy related pair separates from arbitrary pairs.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- read a pinned word2vec skip-gram model;
- inspect the preprocessing and training settings stored with it;
- explain why the fitted model is saved rather than trained during render;
- compare an expected related pair with random pairs; and
- summarize the full distribution of cosine similarities.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load one saved word2vec model

Word2vec comes from work by Mikolov and colleagues on learning word vectors from
nearby words. The **skip-gram** version uses the current word to predict words in
its window. The R package splits input on whitespace, so the training strings
should already be lowercased with punctuation and numbers removed.

```{r}
library(dplyr)
library(tibble)
library(stringr)
library(readr)
library(quanteda)
library(word2vec)
library(ggplot2)
library(knitr)

source("R/inaugural-corpus.R")

within_tolerance <- function(observed, expected, tolerance = 1e-4) {
  abs(observed - expected) < tolerance
}

metadata <- read_csv(
  "data/word2vec/word2vec-metadata.csv",
  na = character(),
  col_types = cols(
    field = col_character(),
    value = col_character()
  )
)
metadata_values <- setNames(metadata$value, metadata$field)

paragraphs <- inaugural_paragraphs()
training_tokens <- tokens(
  paragraphs$paragraph,
  remove_punct = TRUE,
  remove_numbers = TRUE
) |>
  tokens_tolower()

model <- word2vec::read.word2vec("data/word2vec/inaugural-word2vec.bin")
embedding_matrix <- as.matrix(model)
punctuation_rows <- rownames(embedding_matrix)[grepl("[[:punct:]]", rownames(embedding_matrix))]

training_summary <- tibble(
  item = c(
    "training documents",
    "tokens after cleanup",
    "vocabulary rows",
    "dimensions",
    "punctuation rows",
    "unique word pairs"
  ),
  value = c(
    metadata_values[["training_documents"]],
    metadata_values[["training_tokens"]],
    metadata_values[["vocabulary_rows"]],
    metadata_values[["dimensions"]],
    metadata_values[["punctuation_rows"]],
    metadata_values[["unique_pairs"]]
  )
)

hyperparameter_table <- tibble(
  setting = c("algorithm", "dimensions", "window", "iterations", "minimum count", "threads"),
  value = metadata_values[c("algorithm", "dimensions", "window", "iterations", "minimum_count", "threads")]
)

model_fingerprint <- tibble(
  field = c("generator", "model SHA-256", "reproducibility note"),
  value = metadata_values[c("generator", "sha256_model", "reproducibility")]
)

kable(
  training_summary,
  col.names = c("Item", "Value"),
  caption = "Training text and vocabulary checks stored with the saved word2vec model",
  row.names = FALSE
)

kable(
  hyperparameter_table,
  col.names = c("Setting", "Value"),
  caption = "Hyperparameters for the pinned local word2vec skip-gram model",
  row.names = FALSE
)

kable(
  model_fingerprint,
  col.names = c("Field", "Value"),
  caption = "Provenance for the committed word2vec model",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(paragraphs) == as.integer(metadata_values[["training_documents"]]),
  as.integer(sum(ntoken(training_tokens))) == as.integer(metadata_values[["training_tokens"]]),
  nrow(embedding_matrix) == as.integer(metadata_values[["vocabulary_rows"]]),
  ncol(embedding_matrix) == as.integer(metadata_values[["dimensions"]]),
  identical(punctuation_rows, "</s>"),
  length(punctuation_rows) == as.integer(metadata_values[["punctuation_rows"]]),
  as.integer(metadata_values[["unique_pairs"]]) == choose(nrow(embedding_matrix), 2),
  hyperparameter_table$value[hyperparameter_table$setting == "threads"] == "1",
  hyperparameter_table$value[hyperparameter_table$setting == "window"] == "5",
  nchar(metadata_values[["sha256_model"]]) == 64L,
  file.exists("data/word2vec/inaugural-word2vec.bin"),
  all(c("freedom", "liberty", "duties", "union", "government", "peace", "war") %in% rownames(embedding_matrix))
)
```

The metadata records 1,377 training documents, 134,419 cleaned tokens, 2,709
vocabulary rows, and 3,667,986 unique word pairs. It also records one punctuation
row, the trainer's `</s>` marker. Punctuation was removed before training, so
forms such as `peace.` and `peace,` do not compete with `peace`.

The same training code with the same seed and `threads = 1L` produced neighbour
lists in a different order on each fit. The similarity values moved only slightly,
about 0.03 here, but the ranking changed. This page loads one saved fit, and the
metadata records its SHA-256 fingerprint. A result you cannot reproduce is a
result you cannot check; pinning the artefact is the normal fix when exact words
will be printed.

## Print neighbours, then distrust the glow

A nearest-neighbour list returns the words with highest cosine similarity to a
query word. Cosine similarity compares vector direction; larger values mean
closer under the trained representation.

```{r}
nearest_words <- function(word, n = 6L) {
  predict(model, newdata = word, type = "nearest", top_n = n)[[1]] |>
    as_tibble() |>
    mutate(
      similarity_band = case_when(
        similarity >= 0.8 ~ "0.80 or higher",
        similarity >= 0.7 ~ "0.70 to 0.79",
        TRUE ~ "below 0.70"
      )
    )
}

neighbour_table <- bind_rows(
  nearest_words("union") |>
    mutate(query = "union"),
  nearest_words("freedom") |>
    mutate(query = "freedom"),
  nearest_words("peace") |>
    mutate(query = "peace")
) |>
  select(query, neighbour = term2, similarity, similarity_band, rank)

expected_neighbours <- c(
  "texas", "attachment", "system", "jurisdiction", "confederation", "communities",
  "human", "evil", "mankind", "dignity", "civilization", "ancient",
  "friendship", "lasting", "world", "cultivate", "amity", "prosperity"
)

kable(
  neighbour_table |>
    mutate(similarity = round(similarity, 3)),
  col.names = c("Query", "Nearest word", "Cosine similarity", "Cosine similarity band", "Rank"),
  caption = "Nearest neighbours from the pinned local word2vec model",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(neighbour_table) == 18L,
  identical(sort(unique(neighbour_table$query)), c("freedom", "peace", "union")),
  identical(sort(unique(neighbour_table$rank)), 1:6),
  identical(neighbour_table$neighbour, expected_neighbours),
  all(neighbour_table$similarity > 0.85),
  all(neighbour_table$similarity_band == "0.80 or higher")
)
```

Some neighbours look plausible. That is not enough. A neighbour list always
returns something, and a fluent label can make a weak signal feel stronger than
it is. Pinning the file makes the printed words checkable; it does not make the
small model trustworthy by itself.

## Compare a related pair with random pairs

The useful question is not whether the top neighbour list looks nice. The useful
question is whether a pair chosen to be easy, `freedom / liberty`, clearly
separates from arbitrary pairs. Using an easy pair is conservative: if this pair
barely separates, harder related pairs should not be trusted without more data.

```{r}
cosine_between_words <- function(first, second) {
  first_vector <- embedding_matrix[first, ]
  second_vector <- embedding_matrix[second, ]
  sum(first_vector * second_vector) /
    sqrt(sum(first_vector^2) * sum(second_vector^2))
}

freedom_liberty_cosine <- cosine_between_words("freedom", "liberty")

set.seed(6101)
vocabulary <- rownames(embedding_matrix)
random_pair_indices <- replicate(
  1000L,
  sample.int(length(vocabulary), 2L),
  simplify = FALSE
)

random_pair_cosines <- vapply(
  random_pair_indices,
  function(pair) cosine_between_words(vocabulary[pair[1]], vocabulary[pair[2]]),
  numeric(1)
)

random_pair_summary <- tibble(
  statistic = c("mean", "median", "5th percentile", "95th percentile"),
  cosine_similarity = c(
    mean(random_pair_cosines),
    median(random_pair_cosines),
    as.numeric(quantile(random_pair_cosines, 0.05)),
    as.numeric(quantile(random_pair_cosines, 0.95))
  )
)

kable(
  bind_rows(
    tibble(statistic = "freedom / liberty", cosine_similarity = freedom_liberty_cosine),
    random_pair_summary
  ) |>
    mutate(cosine_similarity = round(cosine_similarity, 3)),
  col.names = c("Statistic", "Cosine similarity"),
  caption = "An expected related pair compared with 1,000 random word pairs",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  length(random_pair_cosines) == 1000L,
  within_tolerance(freedom_liberty_cosine, 0.7722, tolerance = 1e-4),
  within_tolerance(mean(random_pair_cosines), 0.7522, tolerance = 1e-4),
  median(random_pair_cosines) > freedom_liberty_cosine,
  as.numeric(quantile(random_pair_cosines, 0.95)) > freedom_liberty_cosine
)
```

The random-pair distribution reaches into the same range as `freedom / liberty`.
In this saved fit the pair scores about 0.772, while random pairs average about
0.752. The model can still be useful for prompting inspection, but a high cosine
score is not enough by itself.

## Show the similarity distribution

The distribution below uses every unique word pair in the cleaned vocabulary. It
also marks the `freedom / liberty` pair with a vertical line.

```{r}
normalized_embeddings <- embedding_matrix / sqrt(rowSums(embedding_matrix^2))
word_similarity_matrix <- tcrossprod(normalized_embeddings)
diag(word_similarity_matrix) <- NA_real_
word_similarities <- word_similarity_matrix[upper.tri(word_similarity_matrix)]
freedom_liberty_percentile <- mean(word_similarities <= freedom_liberty_cosine)

similarity_summary <- tibble(
  statistic = c("word pairs", "median", "95th percentile", "freedom / liberty percentile"),
  value = c(
    length(word_similarities),
    median(word_similarities),
    as.numeric(quantile(word_similarities, 0.95)),
    freedom_liberty_percentile
  )
)

kable(
  similarity_summary |>
    mutate(value = if_else(statistic == "word pairs", as.character(as.integer(value)), sprintf("%.3f", value))),
  col.names = c("Statistic", "Value"),
  caption = "Distribution of all pairwise word-vector cosine similarities",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  length(word_similarities) == as.integer(metadata_values[["unique_pairs"]]),
  within_tolerance(freedom_liberty_percentile, 0.4442, tolerance = 1e-4),
  within_tolerance(median(word_similarities), 0.7938, tolerance = 1e-4),
  within_tolerance(as.numeric(quantile(word_similarities, 0.95)), 0.9354, tolerance = 1e-4),
  freedom_liberty_percentile < 0.5
)
```

`freedom / liberty` sits at the `r round(freedom_liberty_percentile * 100)`th
percentile of the full pairwise distribution in this saved model. The pair was
chosen to be easy, yet it lands below the median pair.

```{r}
#| label: fig-word2vec-cosines
#| fig-cap: "Pairwise cosine similarities among words kept by the pinned local word2vec model."
#| fig-alt: "A histogram concentrated toward high positive cosine values. A vertical line for freedom slash liberty falls left of the median rather than near the far right edge."

plot_similarities <- tibble(cosine_similarity = word_similarities)

ggplot(plot_similarities, aes(x = cosine_similarity)) +
  geom_histogram(bins = 40, boundary = 0, color = "white") +
  geom_vline(xintercept = freedom_liberty_cosine, linewidth = 1, linetype = "dashed") +
  labs(
    x = "Cosine similarity",
    y = "Word pairs",
    title = "Many arbitrary word pairs sit above an easy related pair"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  nrow(plot_similarities) == as.integer(metadata_values[["unique_pairs"]]),
  within_tolerance(median(plot_similarities$cosine_similarity), 0.7938, tolerance = 1e-4),
  freedom_liberty_cosine < median(plot_similarities$cosine_similarity)
)
```

Published embeddings are trained on billions of tokens. Large training text gives
the model many more contexts for separating near neighbours from accidental
neighbours.

## What to remember

- A word2vec model gives one vector to each kept word.
- The pinned model used skip-gram, 50 dimensions, window 5, 5 iterations,
  minimum count 5, and one thread.
- The word2vec package expects text that is already cleaned before whitespace
  splitting.
- Neighbour lists need a pinned artefact when exact words are printed.
- In this saved fit, even `freedom / liberty` lands below the median of all
  pairwise similarities.

The curator keeps the neighbour table as a prompt for reading, not as evidence
that the archive has revealed its themes.

## Sources

- [word2vec R package reference](https://cran.r-project.org/package=word2vec)
- [Original word2vec paper](https://arxiv.org/abs/1301.3781)
- [quanteda tokens documentation](https://quanteda.io/reference/tokens.html)
- [quanteda inaugural corpus documentation](https://quanteda.io/reference/data_corpus_inaugural.html)
- [United States copyright guidance for government works](https://www.usa.gov/government-copyright)
