---
title: "Reading words in context"
subtitle: "Compare one word vector with one token occurrence vector"
description: "Learn how spaCy returns a context-sensitive token representation and why two uses of the same word can have different vectors."
categories: [similarity, contextual embeddings, spaCy]
---

A student underlines `bank` twice in one sentence. The first `bank` raises
interest rates; the second sits beside a river.

The spelling is identical. The surrounding words make the two occurrences
different.

A **contextualized word representation** gives a vector to a token occurrence
rather than to a word type. The same word can therefore receive two different
vectors in the same document.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- start the pinned local spaCy pipeline through the project helper;
- confirm that the small pipeline has no static word-vector table;
- read the token tensor produced for one sentence;
- compare token occurrence vectors with an all-pair baseline;
- check same-sense controls; and
- separate contextual encoding from word-sense identification.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the local spaCy pipeline

spaCy is a Python library. This project reaches it through `R/use-spacy.R`, which
pins the Python environment and prevents downloads during rendering.

```{r}
library(reticulate)
library(spacyr)
library(dplyr)
library(tibble)
library(knitr)

source("R/use-spacy.R")

within_tolerance <- function(observed, expected, tolerance = 1e-3) {
  abs(observed - expected) < tolerance
}

pipeline <- use_project_spacy()
spacy_py <- import("spacy", delay_load = FALSE)
nlp <- spacy_py$load("en_core_web_sm")

pipeline_table <- tibble(
  item = c("pipeline", "static vector rows", "static vector columns"),
  value = c(
    pipeline$model,
    as.character(nlp$vocab$vectors$shape[[1]]),
    as.character(nlp$vocab$vectors$shape[[2]])
  )
)

kable(
  pipeline_table,
  col.names = c("Item", "Value"),
  caption = "The pinned small spaCy pipeline has no static word-vector table",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  pipeline$model == "en_core_web_sm",
  as.integer(nlp$vocab$vectors$shape[[1]]) == 0L,
  as.integer(nlp$vocab$vectors$shape[[2]]) == 0L
)
```

The zero-row vector table matters. This small pipeline does not store one
reusable vector for every word. It still produces an internal tensor for the
tokens it reads.

## Extract token occurrence vectors

A **tensor** is an array of numbers. For this sentence, spaCy returns one row per
token and 96 columns per token.

```{r}
text <- "The bank raised interest rates. He sat on the river bank and watched."
doc <- nlp(text)

tokens <- vapply(
  seq_len(length(doc)) - 1L,
  function(i) doc[i]$text,
  character(1)
)

tensor <- py_to_r(doc$tensor)

bank_positions <- which(tokens == "bank")
the_positions <- which(tokens %in% c("The", "the"))

token_table <- tibble(
  position = seq_along(tokens),
  token = tokens,
  target = case_when(
    position %in% bank_positions ~ "bank occurrence",
    position %in% the_positions ~ "the occurrence",
    TRUE ~ "other token"
  )
)

kable(
  token_table,
  col.names = c("Position", "Token", "Role in this lesson"),
  caption = "Tokens in the sentence and the positions compared below",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(tokens, c("The", "bank", "raised", "interest", "rates", ".", "He", "sat", "on", "the", "river", "bank", "and", "watched", ".")),
  identical(dim(tensor), c(15L, 96L)),
  identical(bank_positions, c(2L, 12L)),
  identical(the_positions, c(1L, 10L)),
  nrow(token_table) == 15L
)
```

The two `bank` tokens are at positions 2 and 12. They share spelling, but they do
not share the same tensor row.

## Compare occurrence vectors with a baseline

Cosine similarity compares vector direction. A score of 1 would mean the two rows
point in exactly the same direction. The two `bank` occurrence vectors score
about 0.3466.

```{r}
cosine <- function(first, second) {
  sum(first * second) / sqrt(sum(first^2) * sum(second^2))
}

all_pair_index <- which(
  upper.tri(matrix(NA_real_, nrow = length(tokens), ncol = length(tokens))),
  arr.ind = TRUE
)

all_pair_cosines <- apply(
  all_pair_index,
  1,
  function(pair) cosine(tensor[pair[1], ], tensor[pair[2], ])
)

bank_cosine <- cosine(tensor[bank_positions[1], ], tensor[bank_positions[2], ])
the_cosine <- cosine(tensor[the_positions[1], ], tensor[the_positions[2], ])
bank_percentile <- mean(all_pair_cosines <= bank_cosine)
the_percentile <- mean(all_pair_cosines <= the_cosine)

all_pair_summary <- tibble(
  statistic = c("mean", "median", "5th percentile", "95th percentile"),
  cosine = c(
    mean(all_pair_cosines),
    median(all_pair_cosines),
    as.numeric(quantile(all_pair_cosines, 0.05)),
    as.numeric(quantile(all_pair_cosines, 0.95))
  )
)

comparison_table <- tibble(
  comparison = c("bank / bank", "the / the"),
  first_position = c(bank_positions[1], the_positions[1]),
  second_position = c(bank_positions[2], the_positions[2]),
  cosine_similarity = c(bank_cosine, the_cosine),
  percentile_of_all_pairs = c(bank_percentile, the_percentile)
)

kable(
  all_pair_summary |>
    mutate(cosine = round(cosine, 4)),
  col.names = c("All-pair baseline", "Cosine"),
  caption = "Cosine similarities across all 105 token pairs in the example sentence",
  row.names = FALSE
)

kable(
  comparison_table |>
    mutate(
      cosine_similarity = round(cosine_similarity, 4),
      percentile_of_all_pairs = round(percentile_of_all_pairs, 3)
    ),
  col.names = c("Comparison", "First position", "Second position", "Cosine similarity", "Percentile of all token pairs"),
  caption = "Repeated word strings compared with the all-pair baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  length(all_pair_cosines) == 105L,
  within_tolerance(bank_cosine, 0.3466, tolerance = 1e-3),
  within_tolerance(the_cosine, 0.5322, tolerance = 1e-3),
  within_tolerance(mean(all_pair_cosines), 0.0476, tolerance = 1e-3),
  within_tolerance(median(all_pair_cosines), -0.0074, tolerance = 1e-3),
  within_tolerance(as.numeric(quantile(all_pair_cosines, 0.95)), 0.4475, tolerance = 1e-3),
  bank_percentile > 0.9,
  the_percentile > bank_percentile
)
```

The `bank` pair is higher than an arbitrary token pair from the same sentence:
it sits around the 91st percentile of the all-pair baseline. That supports a
modest claim. The encoder gives different vectors to occurrences, and the two
`bank` rows are still closer than most random token pairs in this short text.

## Check same-sense controls

The stronger claim would be that the distance isolates word sense. The controls
below test that claim by changing sense and grammatical role separately.

```{r}
tensor_for_text <- function(text) {
  doc <- nlp(text)
  tokens <- vapply(
    seq_len(length(doc)) - 1L,
    function(i) doc[i]$text,
    character(1)
  )

  list(tokens = tokens, tensor = py_to_r(doc$tensor))
}

bank_pair_cosine <- function(text) {
  parsed <- tensor_for_text(text)
  positions <- which(parsed$tokens == "bank")

  if (length(positions) != 2L) {
    stop("Expected exactly two bank tokens.", call. = FALSE)
  }

  cosine(parsed$tensor[positions[1], ], parsed$tensor[positions[2], ])
}

bank_control_sentences <- tibble(
  comparison = c(
    "different senses, different roles",
    "same financial sense, same role",
    "same river sense",
    "same financial sense, different roles"
  ),
  text = c(
    "The bank raised interest rates. He sat on the river bank and watched.",
    "The bank raised interest rates. The bank approved the loan.",
    "He sat on the river bank and watched. She walked along the muddy bank.",
    "The bank approved the loan. She walked into the bank."
  ),
  cosine_similarity = c(
    bank_pair_cosine(text[1]),
    bank_pair_cosine(text[2]),
    bank_pair_cosine(text[3]),
    bank_pair_cosine(text[4])
  )
)

role_gap <- abs(
  bank_control_sentences$cosine_similarity[bank_control_sentences$comparison == "different senses, different roles"] -
    bank_control_sentences$cosine_similarity[bank_control_sentences$comparison == "same financial sense, different roles"]
)

kable(
  bank_control_sentences |>
    mutate(cosine_similarity = round(cosine_similarity, 4)),
  col.names = c("Comparison", "Sentences", "Cosine similarity"),
  caption = "Same-sense controls for two occurrences of bank",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(bank_control_sentences) == 4L,
  all(within_tolerance(
    bank_control_sentences$cosine_similarity,
    c(0.3466, 0.9422, 0.5900, 0.3447),
    tolerance = 1e-3
  )),
  role_gap < 0.01,
  bank_control_sentences$cosine_similarity[2] > 0.9,
  bank_control_sentences$cosine_similarity[2] > bank_control_sentences$cosine_similarity[1]
)
```

The reversal is the lesson. Same financial sense in different grammatical roles
scores about 0.3447, almost the same as different senses in different roles at
about 0.3466. Same financial sense in the same role scores about 0.9422. In these
four probes, the grammatical-role contrast is at least as large as the
dictionary-sense contrast. This is an illustration, not a word-sense evaluation:
there is no benchmark or independent sense labeling here.

## Contrast with one vector per word

The word2vec lesson trains one vector for each kept word type. A
one-vector-per-word model cannot give river `bank` and finance `bank` two rows in
the same sentence. A contextual pipeline can.

That does not mean the vector knows which dictionary sense is meant. The
`en_core_web_sm` tok2vec layer, short for token-to-vector, is the part of the
pipeline that turns each token into numbers, and it is a shared encoder trained
for tagging, parsing, and named-entity recognition. It has no word-sense
objective, so it carries the information useful for those tasks. A representation
that changes with context is not the same thing as a model that knows which sense
you meant. This is a small pipeline's encoder, not a large language model.

```{r}
final_checks <- tibble(
  check = c("token rows", "tensor dimensions", "bank cosine rounded", "same-role financial cosine rounded", "static vector rows"),
  value = c(
    as.character(nrow(tensor)),
    paste(ncol(tensor), "columns"),
    sprintf("%.4f", bank_cosine),
    sprintf("%.4f", bank_control_sentences$cosine_similarity[2]),
    as.character(nlp$vocab$vectors$shape[[1]])
  )
)

kable(
  final_checks,
  col.names = c("Check", "Value"),
  caption = "Final checks for the contextual representation example",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(tensor) == 15L,
  ncol(tensor) == 96L,
  within_tolerance(bank_cosine, 0.3466, tolerance = 1e-3),
  within_tolerance(bank_control_sentences$cosine_similarity[2], 0.9422, tolerance = 1e-3),
  as.integer(nlp$vocab$vectors$shape[[1]]) == 0L,
  identical(tokens[bank_positions], c("bank", "bank")),
  identical(tokens[the_positions], c("The", "the"))
)
```

## What to remember

- A contextualized representation gives one vector to each token occurrence.
- The pinned small spaCy pipeline has 0 rows in its static vector table.
- The sentence produces a 15 by 96 tensor.
- The two `bank` occurrence vectors score about 0.3466, above most token pairs in
  that sentence.
- In these four probes, grammatical role matters at least as much as the
  dictionary-sense contrast; that is not a word-sense benchmark.

The student keeps both underlines and adds a note: same spelling, different
occurrence vectors, not a sense label.

## Sources

- [spaCy model documentation](https://spacy.io/models/en)
- [spaCy Doc tensor attribute](https://spacy.io/api/doc#attributes)
- [spaCy tok2vec architecture documentation](https://spacy.io/api/tok2vec)
- [reticulate package documentation](https://rstudio.github.io/reticulate/)
- [spacyr package reference](https://spacyr.quanteda.io/)
