---
title: "Finding speeches unlike the rest"
subtitle: "Rank cosine distance and test length"
author: "Alex Farach"
description: "Learn how outlier detection depends on the text representation and why length belongs beside distance."
categories: [signals and discovery, outlier detection, inaugural addresses]
---

Owen has one shelf for speeches that deserve closer reading. He asks for the addresses most unlike the rest, expecting a neutral ranking.

The ranking arrives with a catch. An **outlier** is a case that looks far from other cases under a chosen measurement. For text, changing the features can change what looks far away, and very short documents can look strange because they have little vocabulary to average.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- build speech-level document-feature matrices;
- compute cosine distance from the average speech;
- report word length next to distance;
- test how much distance tracks length; and
- compare rank agreement with a shuffled-feature null.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build the representations

A **representation** is the version of the text that a method sees. Here one representation keeps content words after removing stopwords, another gives those content words tf-idf weights, and a third keeps all words as raw counts. **Cosine distance** compares the direction of two word-count vectors. A value closer to 1 means farther from the average vector.

```{r}
library(dplyr)
library(tibble)
library(quanteda)
library(proxy)
library(knitr)

source("R/inaugural-corpus.R")
source("R/permutation-null.R")

paragraphs <- inaugural_paragraphs()

speeches <- paragraphs |>
  summarise(text = paste(paragraph, collapse = "\n"), .by = c(speech_id, year, president, party))

make_speech_dfm <- function(remove_stops = TRUE) {
  speech_tokens <- tokens(
    speeches$text,
    remove_punct = TRUE,
    remove_numbers = TRUE
  ) |>
    tokens_tolower()

  if (remove_stops) {
    speech_tokens <- tokens_remove(speech_tokens, stopwords("en"))
  }

  speech_dfm <- dfm(speech_tokens)
  docnames(speech_dfm) <- speeches$speech_id
  speech_dfm
}

cosine_from_matrix <- function(matrix_counts) {
  average_speech <- colMeans(matrix_counts)
  numerator <- as.numeric(matrix_counts %*% average_speech)
  denominator <- sqrt(rowSums(matrix_counts * matrix_counts)) * sqrt(sum(average_speech * average_speech))
  1 - numerator / denominator
}

cosine_from_average <- function(speech_dfm) {
  cosine_from_matrix(as.matrix(speech_dfm))
}

content_dfm <- make_speech_dfm(remove_stops = TRUE)
content_tfidf_dfm <- dfm_tfidf(content_dfm)
all_words_dfm <- make_speech_dfm(remove_stops = FALSE)

lengths <- tibble(
  speech_id = speeches$speech_id,
  year = speeches$year,
  president = speeches$president,
  total_words = ntoken(all_words_dfm),
  content_words = ntoken(content_dfm)
)

shortest_speech <- lengths |>
  slice_min(total_words, n = 1, with_ties = FALSE)

representation_summary <- tibble(
  representation = c("Content words, raw counts", "Content words, tf-idf", "All words, raw counts"),
  documents = c(ndoc(content_dfm), ndoc(content_tfidf_dfm), ndoc(all_words_dfm)),
  features = c(nfeat(content_dfm), nfeat(content_tfidf_dfm), nfeat(all_words_dfm))
)

kable(
  representation_summary,
  col.names = c("Representation", "Documents", "Features"),
  caption = "Three speech representations for outlier detection",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(paragraphs), 1377L),
  identical(ndoc(content_dfm), 60L),
  identical(nfeat(content_dfm), 9193L),
  identical(nfeat(all_words_dfm), 9327L),
  identical(shortest_speech$speech_id, "1793-Washington"),
  identical(as.integer(shortest_speech$total_words), 135L),
  identical(as.integer(lengths$content_words[lengths$speech_id == "1793-Washington"]), 62L)
)
```

Cosine distance compares direction rather than raw size. Length still matters because very short speeches can have unusual word mixes.

## Rank distance from the average

The next table uses content words with raw counts. It shows the most distant speeches and the closest speeches under that one representation.

```{r}
content_rank <- lengths |>
  mutate(
    distance = cosine_from_average(content_dfm),
    representation = "content words, raw counts"
  ) |>
  arrange(desc(distance))

most_typical <- content_rank |>
  arrange(distance) |>
  slice_head(n = 5) |>
  mutate(group = "Least distant")

most_distant <- content_rank |>
  slice_head(n = 5) |>
  mutate(group = "Most distant")

rank_table <- bind_rows(most_distant, most_typical) |>
  transmute(
    group,
    speech_id,
    year,
    president,
    total_words,
    content_words,
    shortest_in_corpus = if_else(total_words == min(lengths$total_words), "yes", "no"),
    distance = round(distance, 3)
  )

kable(
  rank_table,
  col.names = c("Group", "Speech", "Year", "President", "Total words", "Content words", "Shortest in corpus", "Cosine distance"),
  caption = "Most and least distant speeches using content-word raw counts",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(content_rank$speech_id[1], "1793-Washington"),
  identical(content_rank$speech_id[2], "2021-Biden"),
  identical(content_rank$speech_id[3], "1865-Lincoln"),
  identical(content_rank$speech_id[nrow(content_rank)], "1897-McKinley"),
  identical(as.numeric(round(content_rank$distance[1], 3)), 0.677),
  identical(as.numeric(round(min(content_rank$distance), 3)), 0.235)
)
```

The 1793-Washington speech is the most distant under this representation, and it is the shortest speech at 135 total words. A 135-word document has little vocabulary to average, so a handful of words can drive its position.

## Check the length confound

A **confound** is a second factor that can explain a result the analyst might otherwise attribute to the method's target. Here the candidate confound is document length.

```{r}
tfidf_rank <- lengths |>
  mutate(
    distance = cosine_from_average(content_tfidf_dfm),
    representation = "content words, tf-idf"
  ) |>
  arrange(desc(distance))

all_words_rank <- lengths |>
  mutate(
    distance = cosine_from_average(all_words_dfm),
    representation = "all words, raw counts"
  ) |>
  arrange(desc(distance))

all_rankings <- bind_rows(content_rank, tfidf_rank, all_words_rank)

length_correlations <- all_rankings |>
  summarise(
    correlation = round(cor(total_words, distance), 3),
    .by = representation
  )

length_threshold <- 1000L

restricted_top <- all_rankings |>
  filter(total_words >= length_threshold) |>
  arrange(desc(distance), .by_group = TRUE) |>
  slice_head(n = 5, by = representation) |>
  transmute(
    representation,
    speech_id,
    year,
    total_words,
    distance = round(distance, 3)
  )

kable(
  length_correlations,
  col.names = c("Representation", "Correlation between total words and distance"),
  caption = "Distance is strongly related to speech length",
  row.names = FALSE
)

kable(
  restricted_top,
  col.names = c("Representation", "Speech", "Year", "Total words", "Cosine distance"),
  caption = "Top outliers after restricting to speeches with at least 1,000 words",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length_correlations$correlation, c(-0.738, -0.908, -0.534)),
  identical(restricted_top$speech_id[restricted_top$representation == "content words, raw counts"][1], "1813-Madison"),
  identical(restricted_top$speech_id[restricted_top$representation == "content words, tf-idf"][1], "1813-Madison"),
  identical(restricted_top$speech_id[restricted_top$representation == "all words, raw counts"][1], "2001-Bush"),
  !"1793-Washington" %in% restricted_top$speech_id,
  !"2021-Biden" %in% restricted_top$speech_id
)
```

The correlations are negative under all three representations: -0.738 for content-word counts, -0.908 for content-word tf-idf, and -0.534 for all-word counts. Shorter speeches tend to sit farther from the average. Once speeches under 1,000 words are removed, 1793-Washington and 2021-Biden leave the top lists. The outlier ranking is substantially a length ranking.

## Compare the feature choices with a null

Now add tf-idf weights and then keep stopwords in a separate raw-count representation. This changes both the weighting rule and the feature set. A **rank agreement** statistic checks whether the three distance rankings put speeches in similar order.

A **permutation null** reruns the rank-agreement calculation after feature counts are randomized. The **p-value** is the fraction of shuffled agreements that are at least as large as the observed agreement.

```{r}
feature_comparison <- all_rankings |>
  slice_head(n = 5, by = representation) |>
  transmute(
    representation,
    speech_id,
    year,
    president,
    total_words,
    content_words,
    distance = round(distance, 3)
  )

shared_content_tfidf <- length(intersect(
  content_rank$speech_id[1:5],
  tfidf_rank$speech_id[1:5]
))

shared_content_all <- length(intersect(
  content_rank$speech_id[1:5],
  all_words_rank$speech_id[1:5]
))

rank_agreement <- function(distance_data) {
  mean(c(
    cor(distance_data$content_raw, distance_data$content_tfidf, method = "spearman"),
    cor(distance_data$content_raw, distance_data$all_words, method = "spearman"),
    cor(distance_data$content_tfidf, distance_data$all_words, method = "spearman")
  ))
}

observed_distances <- lengths |>
  transmute(
    speech_id,
    content_raw = cosine_from_average(content_dfm),
    content_tfidf = cosine_from_average(content_tfidf_dfm),
    all_words = cosine_from_average(all_words_dfm)
  )

observed_rank_agreement <- rank_agreement(observed_distances)

tfidf_matrix <- function(matrix_counts) {
  document_frequency <- colSums(matrix_counts > 0)
  idf <- log(nrow(matrix_counts) / pmax(document_frequency, 1))
  sweep(matrix_counts, 2, idf, `*`)
}

random_matrix <- function(length_vector, probabilities) {
  t(vapply(length_vector, function(one_length) {
    as.numeric(rmultinom(n = 1L, size = one_length, prob = probabilities))
  }, numeric(length(probabilities))))
}

content_matrix <- as.matrix(content_dfm)
all_words_matrix <- as.matrix(all_words_dfm)
content_lengths <- rowSums(content_matrix)
all_words_lengths <- rowSums(all_words_matrix)
content_probabilities <- colSums(content_matrix)
all_words_probabilities <- colSums(all_words_matrix)

rank_agreement_null <- permutation_null(
  observed = observed_rank_agreement,
  replicate_fn = function() {
    random_content <- random_matrix(content_lengths, content_probabilities)
    random_all_words <- random_matrix(all_words_lengths, all_words_probabilities)

    random_distances <- tibble(
      content_raw = cosine_from_matrix(random_content),
      content_tfidf = cosine_from_matrix(tfidf_matrix(random_content)),
      all_words = cosine_from_matrix(random_all_words)
    )

    rank_agreement(random_distances)
  },
  replicates = 500L,
  seed = 5702L,
  alternative = "greater"
)

rank_agreement_display <- rank_agreement_null |>
  mutate(across(c(observed, null_mean, null_low, null_high, p_value), \(x) round(x, 3)))

kable(
  feature_comparison,
  col.names = c("Representation", "Speech", "Year", "President", "Total words", "Content words", "Cosine distance"),
  caption = "Top five outliers across three representations",
  row.names = FALSE
)

kable(
  rank_agreement_display,
  col.names = c("Observed", "Null mean", "Null low", "Null high", "p-value", "Replicates", "Alternative"),
  caption = "Permutation null for rank agreement across representations",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(feature_comparison), 15L),
  identical(all_words_rank$speech_id[1], "1793-Washington"),
  identical(all_words_rank$speech_id[3], "2001-Bush"),
  identical(tfidf_rank$speech_id[1:5], content_rank$speech_id[1:5]),
  identical(shared_content_tfidf, 5L),
  identical(shared_content_all, 4L),
  identical(as.numeric(round(observed_rank_agreement, 3)), 0.770),
  identical(as.numeric(rank_agreement_display$null_mean), 0.953),
  identical(as.numeric(rank_agreement_display$p_value), 1)
)
```

Content-word raw counts and content-word tf-idf share all five top speeches. Keeping stopwords shares four of five with content-word raw counts and changes the third slot from 1865-Lincoln to 2001-Bush. The average Spearman rank agreement across the three full rankings is 0.770. When document lengths are held fixed but features are randomly reassigned, the null mean is 0.953 with a 90% interval from 0.939 to 0.966. The observed agreement does not exceed that shuffled-feature null (p = 1). Cross-representation agreement is therefore not reassuring by itself.

Outlier methods always rank something first. The shuffle removes document-specific word patterns while preserving lengths, so it asks whether the three representations agree more than feature noise would. They do not. The shortest-speech result remains a length-confounded reading lead.

## What to remember

- Cosine distance ranks each speech against the average vector.
- The content-word representations have 9,193 features; the all-word representation has 9,327.
- Under content-word raw counts, 1793-Washington is farthest and 1897-McKinley is closest to the average.
- Distance is strongly tied to length here; the shortest speeches drive the top of the list.
- The three representation rankings do not agree more than shuffled features predict after lengths are fixed.

Use outlier scores to choose what to read next. Do not treat the ranking as a property of the speech apart from the features you chose.

## Sources

- [proxy package documentation for distance measures](https://cran.r-project.org/package=proxy)
- [Aggarwal, Outlier Analysis](https://doi.org/10.1007/978-3-319-47578-3)
- [quanteda document-feature matrix documentation](https://quanteda.io/reference/dfm.html)
- [quanteda tokenization documentation](https://quanteda.io/reference/tokens.html)
