---
title: "Deciding what counts as a word"
subtitle: "Compare word, sentence, and subword tokenization"
description: "Learn why tokenization is a rule-based decision before any word count can begin."
categories: [word parsing, tokenization, workforce research]
---

A planning meeting begins with the Lab's labeled sentences on the table. The
team has 28 reviewed text units from a job board page and a training flyer.
Six came from the flyer, and 22 came from job-detail lines.

The labels answer one question: what does each text unit say? The next question
is smaller and more mechanical. Before the Lab can count words, it has to decide
what a word is.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain why a token is created by a rule;
- compare whitespace and package word tokenizers;
- check whether sentence tokenization changes row counts;
- train a subword model without writing files into the project folder; and
- explain what actually decides how many pieces a text costs.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Start with the sentence on the page

The first code the reader sees brings in readr for files, dplyr and tibble for
tables, purrr for repeated checks, stringr for strings, tokenizers for word and
sentence splitting, and tokenizers.bpe for subword splitting. A **tibble** is
the tidyverse table shape that prints its dimensions and each column's type.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(tokenizers.bpe)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

document_counts <- sentences |>
  mutate(
    source_group = if_else(
      document_id == "F001",
      "training flyer",
      "job-detail rows"
    )
  ) |>
  count(source_group, name = "rows")

first_sentence <- sentences$text[1]

knitr::kable(
  document_counts,
  col.names = c("Source group", "Rows"),
  caption = "The 28 labeled text units by source",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(
    document_counts$rows[document_counts$source_group == "training flyer"],
    6L
  ),
  identical(
    document_counts$rows[document_counts$source_group == "job-detail rows"],
    22L
  ),
  identical(
    first_sentence,
    "Paid 12-week training is provided."
  )
)
```

The first sentence fits on one line: `Paid 12-week training is provided.` A
**token** is one piece of text produced by a splitting rule. The same visible
sentence can produce different tokens because the rule makes different choices.

## Split on spaces

One simple rule is to split wherever there is whitespace. Here stringr uses the
ICU text library to find spaces, tabs, and similar separators. That rule gives
five pieces. The period stays attached to `provided.`, and `12-week` stays in
one piece.

```{r}
whitespace_tokens <- str_split(
  first_sentence,
  "\\s+",
  simplify = FALSE
)[[1]]

knitr::kable(
  tibble(
    token_number = seq_along(whitespace_tokens),
    token = whitespace_tokens
  ),
  col.names = c("Token number", "Whitespace token"),
  caption = "Tokens made by splitting on whitespace",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length(whitespace_tokens), 5L),
  identical(
    whitespace_tokens,
    c("Paid", "12-week", "training", "is", "provided.")
  )
)
```

This rule is easy to explain, but it keeps punctuation and hyphenated material
attached to nearby letters. That may or may not fit the count the Lab wants.

## Use a word tokenizer

`tokenizers::tokenize_words()` applies a different rule. It lowercases text,
drops punctuation, and splits the hyphenated phrase. The same sentence becomes
six tokens: `paid`, `12`, `week`, `training`, `is`, and `provided`.

```{r}
default_word_tokens <- tokenize_words(first_sentence)[[1]]

knitr::kable(
  tibble(
    token_number = seq_along(default_word_tokens),
    token = default_word_tokens
  ),
  col.names = c("Token number", "Default word token"),
  caption = "Tokens from tokenizers::tokenize_words()",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length(default_word_tokens), 6L),
  identical(
    default_word_tokens,
    c("paid", "12", "week", "training", "is", "provided")
  )
)
```

That rule is useful when `Paid` and `paid` should count together. It is a poor
fit if the Lab needs to keep punctuation as its own evidence.

## Keep case and punctuation visible

The same function can be asked not to lowercase and not to remove punctuation.
With those settings, `Paid` stays capitalized, the hyphen becomes its own
token, and the full stop becomes its own token. The sentence becomes eight
tokens.

```{r}
case_punctuation_tokens <- tokenize_words(
  first_sentence,
  lowercase = FALSE,
  strip_punct = FALSE
)[[1]]

token_count_summary <- tibble(
  rule = c(
    "split on whitespace",
    "default word tokenizer",
    "keep case and punctuation"
  ),
  tokens = c(
    length(whitespace_tokens),
    length(default_word_tokens),
    length(case_punctuation_tokens)
  )
)

knitr::kable(
  token_count_summary,
  col.names = c("Rule", "Tokens from the first sentence"),
  caption = "One sentence under three tokenization rules",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length(case_punctuation_tokens), 8L),
  identical(
    case_punctuation_tokens,
    c("Paid", "12", "-", "week", "training", "is", "provided", ".")
  )
)
```

Five, six, and eight are all correct counts under the rule that produced them.
The count alone does not tell the reader which rule was used.

## Check sentence splitting before word counts

The file already stores one sentence per row. The Lab still checks that a
sentence tokenizer sees the same 28 pieces before moving on. This guard matters
because some files store several sentences in one row, while others break a
sentence across rows.

```{r}
sentence_pieces <- tokenize_sentences(sentences$text)
sentence_piece_counts <- map_int(sentence_pieces, length)
source_sentence_counts <- sentences |>
  mutate(sentence_pieces = sentence_piece_counts) |>
  summarise(
    rows = n(),
    pieces = sum(sentence_pieces),
    .by = document_id
  )

knitr::kable(
  source_sentence_counts,
  col.names = c("Document ID", "Rows", "Sentence pieces"),
  caption = "Sentence tokenizer results by document",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(sentence_piece_counts, rep(1L, 28L)),
  identical(sum(source_sentence_counts$pieces), 28L)
)
```

For these rows, `tokenizers::tokenize_sentences()` returns one piece per row.
That means the file's sentence boundary choice and the tokenizer's boundary
choice agree on this teaching dataset.

## Train a subword tokenizer

Modern language models often split text into **subwords**, pieces smaller than
some words and larger than single letters when the model has learned a common
pattern. **Byte pair encoding (BPE)** starts from single characters and
repeatedly glues together the pair that occurs most often. The model below
trains only on the 28 Riverton sentences.

The `vocab_size = 250L` argument is a modeling choice that controls how many
pieces the trainer can keep. The `coverage = 1` argument tells the trainer to
keep every character it sees in the training text; characters outside that set
have no piece. The special marker `<UNK>` is the model's way of saying it has no
piece for this, short for "unknown". The code also sets `threads = 1L` so the
result does not depend on how work is split across processors.

::: {.callout-note}
The explicit `model_path` sends the model to a temporary file during rendering.
:::

```{r}
corpus_file <- tempfile(fileext = ".txt")
write_lines(sentences$text, corpus_file)

bpe_model <- bpe(
  corpus_file,
  model_path = tempfile(fileext = ".bpe"),
  vocab_size = 250L,
  coverage = 1,
  threads = 1L
)

first_vocabulary_rows <- bpe_model$vocabulary |>
  slice_head(n = 10)

knitr::kable(
  first_vocabulary_rows,
  col.names = c("ID", "Subword"),
  caption = "The first rows of the BPE vocabulary",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(bpe_model$vocabulary), 250L),
  identical(
    bpe_model$vocabulary$subword[1:6],
    c("<PAD>", "<UNK>", "<BOS>", "<EOS>", "▁", "e")
  )
)
```

The vocabulary has 250 entries. The first four rows are special markers, then
the boundary marker `▁` and single characters. The marker `▁` means that a
subword begins at a word boundary.

## Compare familiar and unfamiliar text

When the BPE model sees the first sentence, common pieces such as `training`,
`is`, and `provided.` stay together. The rarer `12-week` material fragments.

```{r}
first_sentence_subwords <- bpe_encode(
  bpe_model,
  x = "Paid 12-week training is provided.",
  type = "subwords"
)[[1]]

unseen_phrase_subwords <- bpe_encode(
  bpe_model,
  x = "unpredictability biotechnology",
  type = "subwords"
)[[1]]

subword_summary <- tibble(
  text = c(
    "Paid 12-week training is provided.",
    "unpredictability biotechnology"
  ),
  pieces = c(
    length(first_sentence_subwords),
    length(unseen_phrase_subwords)
  ),
  subwords = c(
    str_c(first_sentence_subwords, collapse = " | "),
    str_c(unseen_phrase_subwords, collapse = " | ")
  )
)

knitr::kable(
  subword_summary,
  col.names = c("Text", "Subword pieces", "Subwords"),
  caption = "BPE pieces for familiar and unfamiliar text",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    first_sentence_subwords,
    c(
      "▁Paid",
      "▁1",
      "2",
      "-",
      "w",
      "eek",
      "▁training",
      "▁is",
      "▁provided."
    )
  ),
  identical(length(first_sentence_subwords), 9L),
  identical(length(unseen_phrase_subwords), 21L)
)
```

The first sentence takes nine subword pieces. The phrase `unpredictability
biotechnology`, which does not appear in the Riverton rows, takes 21 pieces in
this toy model. Larger released tokenizers use different training data,
vocabularies, normalizers, and unknown policies, so their piece counts must be
measured rather than inferred from this result.

It is tempting to end there with a rule: unfamiliar material costs more pieces.
The next section shows that the rule is false as stated.

## What actually decides the number of pieces

This model was trained on 28 English lines written in the Latin script, and
`coverage = 1` told the trainer to keep every character it saw in that text. A
character outside that set has no piece of its own, and the model falls back to
`<UNK>`.

Watch what that does to the piece count. The comparison below adds two texts the
model has never seen: a Chinese phrase meaning "training course" and a Greek
word. It also switches from `type = "subwords"` to `type = "ids"`, which reports
the vocabulary entry each piece actually resolved to.

```{r}
unknown_id <- bpe_model$vocabulary$id[
  bpe_model$vocabulary$subword == "<UNK>"
]

texts_to_encode <- c(
  familiar_sentence = "Paid 12-week training is provided.",
  unseen_english = "unpredictability biotechnology",
  unseen_chinese = "\u57F9\u8BAD\u8BFE\u7A0B",
  unseen_greek = "\u0393\u03BB\u03C9\u03C3\u03C3\u03B1"
)

piece_report <- tibble(
  example = names(texts_to_encode),
  text = unname(texts_to_encode),
  ids = map(
    unname(texts_to_encode),
    \(one_text) bpe_encode(bpe_model, x = one_text, type = "ids")[[1]]
  ),
  subwords = map(
    unname(texts_to_encode),
    \(one_text) bpe_encode(bpe_model, x = one_text, type = "subwords")[[1]]
  )
) |>
  mutate(
    characters = str_length(text),
    pieces = map_int(ids, length),
    unknown_pieces = map_int(
      ids,
      \(id_vector) sum(id_vector == unknown_id)
    ),
    shown_as_subwords = map_chr(
      subwords,
      \(piece_vector) str_c(piece_vector, collapse = " | ")
    ),
    display_text = case_when(
      example == "unseen_chinese" ~
        "<span lang=\"zh\">培训课程</span>",
      example == "unseen_greek" ~
        "<span lang=\"el\">Γλωσσα</span>",
      TRUE ~ text
    ),
    display_subwords = case_when(
      example == "unseen_chinese" ~ str_c(
        "<span lang=\"zh\">",
        shown_as_subwords,
        "</span>"
      ),
      example == "unseen_greek" ~ str_c(
        "<span lang=\"el\">",
        shown_as_subwords,
        "</span>"
      ),
      TRUE ~ shown_as_subwords
    )
  )

knitr::kable(
  piece_report |>
    select(display_text, characters, pieces, unknown_pieces, display_subwords),
  col.names = c(
    "Text",
    "Characters",
    "Pieces",
    "Unknown pieces",
    "What the subword view shows"
  ),
  caption = "Piece counts and unknown pieces for four texts",
  row.names = FALSE,
  escape = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(unknown_id, 1L),
  identical(piece_report$pieces, c(9L, 21L, 2L, 2L)),
  identical(piece_report$unknown_pieces, c(0L, 0L, 1L, 1L)),
  piece_report$pieces[3] < piece_report$pieces[1],
  identical(
    piece_report$shown_as_subwords[3],
    str_c("\u2581 | ", texts_to_encode[["unseen_chinese"]])
  ),
  str_detect(piece_report$display_text[3], fixed("lang=\"zh\"")),
  str_detect(piece_report$display_text[4], fixed("lang=\"el\"")),
  str_detect(piece_report$display_subwords[3], fixed("lang=\"zh\"")),
  str_detect(piece_report$display_subwords[4], fixed("lang=\"el\""))
)
```

The Chinese phrase is the least familiar text on the page and costs two pieces,
where the fully familiar English sentence costs nine. Nothing was learned about
Chinese. One of those two pieces is `<UNK>`, the model's way of recording that
it has no representation at all.

That is the whole correction. The number of pieces depends on the tokenizer and
the vocabulary it learned, on the script and the language, on the normalization
the model applies before splitting, and above all on what the model does with a
character it does not know. Under this model's unknown policy, unfamiliar
material can cost fewer pieces than familiar material, because collapsing to
`<UNK>` is cheap and lossy.

The `shown_as_subwords` column is a second warning. It displays the Chinese
characters as though the model had a piece for them, because the subword view
reconstructs surface text rather than reporting vocabulary entries. Count
unknowns from the ids, never from the printed pieces.

### Check that the pieces can be turned back into text

A tokenizer that cannot rebuild its input has lost something. Decoding the ids
answers that question directly.

```{r}
round_trip <- piece_report |>
  mutate(
    decoded = map_chr(
      ids,
      \(id_vector) bpe_decode(bpe_model, id_vector)[[1]]
    ),
    recovers_input = decoded == text
  )

knitr::kable(
  round_trip |>
    select(display_text, pieces, unknown_pieces, decoded, recovers_input),
  col.names = c(
    "Text",
    "Pieces",
    "Unknown pieces",
    "Decoded from ids",
    "Recovers the input"
  ),
  caption = "Round-trip results for the four texts",
  row.names = FALSE,
  escape = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    round_trip$recovers_input,
    c(TRUE, TRUE, FALSE, FALSE)
  ),
  all(str_detect(
    round_trip$decoded[!round_trip$recovers_input],
    fixed("<UNK>")
  )),
  identical(
    round_trip$recovers_input,
    round_trip$unknown_pieces == 0L
  )
)
```

Both Latin-script texts come back character for character, including the
punctuation and the digits. Both unseen-script texts come back as `<UNK>`, and
the check confirms the pattern exactly: the texts that fail to round-trip are
precisely the texts that contained an unknown piece.

Round-trip recovery tests reversible text identity. It does not test source
alignment: some tokenizers can return offsets into the original input even when
an unknown id cannot decode to its text. A task that marks spans or highlights
matches must test those offset mappings separately.

The result gives the team a rule for the meeting: a word count must name the
rule that made the words, and a piece count must name the vocabulary, the
script it covers, and what it does with everything else.

## What to remember

- Tokenization turns text into pieces according to a chosen rule.
- The first Riverton sentence becomes 5, 6, or 8 word tokens under three rules.
- Sentence tokenization preserves the 28 rows in this file.
- This toy BPE used larger pieces for some material present in its training
  text; that pattern is not universal.
- Piece counts depend on the vocabulary, the script, the normalization the
  model applies, and its policy for unknown characters.
- Under an unknown-piece policy, unfamiliar text can cost fewer pieces and lose
  everything it contained.
- Count unknown pieces from the ids, and check that the ids decode back to the
  input.

The Lab can start counting only after it writes down the token rule. For these
sentences, that rule changes the unit being counted before any summary appears.

## Sources

- [tokenizers documentation for word and sentence tokenizers](https://docs.ropensci.org/tokenizers/reference/basic-tokenizers.html)
- [tokenizers.bpe package documentation](https://cran.r-project.org/package=tokenizers.bpe)
- [Neural Machine Translation of Rare Words with Subword Units](https://aclanthology.org/P16-1162/)
- [SentencePiece: A simple and language independent subword tokenizer](https://aclanthology.org/D18-2012/)
