---
title: "Deciding what counts as a word"
subtitle: "Compare word, sentence, and subword tokenization"
author: "Alex Farach"
description: "Learn why tokenization is a rule-based decision before any word count can begin."
categories: [word parsing, tokenization, workforce research]
---

A planning meeting begins with the Lab's labeled sentences on the table. The
team has 28 reviewed text units from a job board page and a training flyer.
Six came from the flyer, and 22 came from job-detail lines.

The labels answer one question: what does each sentence say? The next question
is smaller and more mechanical. Before the Lab can count words, it has to decide
what a word is.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain why a token is created by a rule;
- compare whitespace and package word tokenizers;
- check whether sentence tokenization changes row counts;
- train a subword model without writing files into the project folder; and
- describe why unfamiliar text can require more subword pieces.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Start with the sentence on the page

The first code the reader sees brings in readr for files, dplyr and tibble for
tables, purrr for repeated checks, stringr for strings, tokenizers for word and
sentence splitting, and tokenizers.bpe for subword splitting. A **tibble** is
the tidyverse table shape that prints its dimensions and each column's type.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(tokenizers.bpe)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

document_counts <- sentences |>
  mutate(
    source_group = if_else(
      document_id == "F001",
      "training flyer",
      "job-detail rows"
    )
  ) |>
  count(source_group, name = "rows")

first_sentence <- sentences$text[1]

stopifnot(
  identical(nrow(sentences), 28L),
  identical(
    document_counts$rows[document_counts$source_group == "training flyer"],
    6L
  ),
  identical(
    document_counts$rows[document_counts$source_group == "job-detail rows"],
    22L
  ),
  identical(
    first_sentence,
    "Paid 12-week training is provided."
  )
)

knitr::kable(
  document_counts,
  col.names = c("Source group", "Rows"),
  caption = "The 28 labeled text units by source",
  row.names = FALSE
)
```

The first sentence fits on one line: `Paid 12-week training is provided.` A
**token** is one piece of text produced by a splitting rule. The same visible
sentence can produce different tokens because the rule makes different choices.

## Split on spaces

One simple rule is to split wherever there is whitespace. Here stringr uses the
ICU text library to find spaces, tabs, and similar separators. That rule gives
five pieces. The period stays attached to `provided.`, and `12-week` stays in
one piece.

```{r}
whitespace_tokens <- str_split(
  first_sentence,
  "\\s+",
  simplify = FALSE
)[[1]]

stopifnot(
  identical(length(whitespace_tokens), 5L),
  identical(
    whitespace_tokens,
    c("Paid", "12-week", "training", "is", "provided.")
  )
)

knitr::kable(
  tibble(
    token_number = seq_along(whitespace_tokens),
    token = whitespace_tokens
  ),
  col.names = c("Token number", "Whitespace token"),
  caption = "Tokens made by splitting on whitespace",
  row.names = FALSE
)
```

This rule is easy to explain, but it keeps punctuation and hyphenated material
attached to nearby letters. That may or may not fit the count the Lab wants.

## Use a word tokenizer

`tokenizers::tokenize_words()` applies a different rule. It lowercases text,
drops punctuation, and splits the hyphenated phrase. The same sentence becomes
six tokens: `paid`, `12`, `week`, `training`, `is`, and `provided`.

```{r}
default_word_tokens <- tokenize_words(first_sentence)[[1]]

stopifnot(
  identical(length(default_word_tokens), 6L),
  identical(
    default_word_tokens,
    c("paid", "12", "week", "training", "is", "provided")
  )
)

knitr::kable(
  tibble(
    token_number = seq_along(default_word_tokens),
    token = default_word_tokens
  ),
  col.names = c("Token number", "Default word token"),
  caption = "Tokens from tokenizers::tokenize_words()",
  row.names = FALSE
)
```

That rule is useful when `Paid` and `paid` should count together. It is a poor
fit if the Lab needs to keep punctuation as its own evidence.

## Keep case and punctuation visible

The same function can be asked not to lowercase and not to remove punctuation.
With those settings, `Paid` stays capitalized, the hyphen becomes its own
token, and the full stop becomes its own token. The sentence becomes eight
tokens.

```{r}
case_punctuation_tokens <- tokenize_words(
  first_sentence,
  lowercase = FALSE,
  strip_punct = FALSE
)[[1]]

token_count_summary <- tibble(
  rule = c(
    "split on whitespace",
    "default word tokenizer",
    "keep case and punctuation"
  ),
  tokens = c(
    length(whitespace_tokens),
    length(default_word_tokens),
    length(case_punctuation_tokens)
  )
)

stopifnot(
  identical(length(case_punctuation_tokens), 8L),
  identical(
    case_punctuation_tokens,
    c("Paid", "12", "-", "week", "training", "is", "provided", ".")
  )
)

knitr::kable(
  token_count_summary,
  col.names = c("Rule", "Tokens from the first sentence"),
  caption = "One sentence under three tokenization rules",
  row.names = FALSE
)
```

Five, six, and eight are all correct counts under the rule that produced them.
The count alone does not tell the reader which rule was used.

## Check sentence splitting before word counts

The file already stores one sentence per row. The Lab still checks that a
sentence tokenizer sees the same 28 pieces before moving on. This guard matters
because some files store several sentences in one row, while others break a
sentence across rows.

```{r}
sentence_pieces <- tokenize_sentences(sentences$text)
sentence_piece_counts <- map_int(sentence_pieces, length)
source_sentence_counts <- sentences |>
  mutate(sentence_pieces = sentence_piece_counts) |>
  summarise(
    rows = n(),
    pieces = sum(sentence_pieces),
    .by = document_id
  )

stopifnot(
  identical(sentence_piece_counts, rep(1L, 28L)),
  identical(sum(source_sentence_counts$pieces), 28L)
)

knitr::kable(
  source_sentence_counts,
  col.names = c("Document ID", "Rows", "Sentence pieces"),
  caption = "Sentence tokenizer results by document",
  row.names = FALSE
)
```

For these rows, `tokenizers::tokenize_sentences()` returns one piece per row.
That means the file's sentence boundary choice and the tokenizer's boundary
choice agree on this teaching dataset.

## Train a subword tokenizer

Modern language models often split text into **subwords**, pieces smaller than
some words and larger than single letters when the model has learned a common
pattern. **Byte pair encoding (BPE)** starts from single characters and
repeatedly glues together the pair that occurs most often. The model below
trains only on the 28 Riverton sentences.

The `vocab_size = 250L` argument is a modeling choice that controls how many
pieces the trainer can keep. The `coverage = 1` argument tells the trainer to
keep every character it sees in the training text; characters outside that set
have no piece. The special marker `<UNK>` is the model's way of saying it has no
piece for this, short for "unknown". The code also sets `threads = 1L` so the
result does not depend on how work is split across processors.

::: {.callout-note}
The explicit `model_path` sends the model to a temporary file during rendering.
:::

```{r}
corpus_file <- tempfile(fileext = ".txt")
write_lines(sentences$text, corpus_file)

bpe_model <- bpe(
  corpus_file,
  model_path = tempfile(fileext = ".bpe"),
  vocab_size = 250L,
  coverage = 1,
  threads = 1L
)

first_vocabulary_rows <- bpe_model$vocabulary |>
  slice_head(n = 10)

stopifnot(
  identical(nrow(bpe_model$vocabulary), 250L),
  identical(
    bpe_model$vocabulary$subword[1:6],
    c("<PAD>", "<UNK>", "<BOS>", "<EOS>", "▁", "e")
  )
)

knitr::kable(
  first_vocabulary_rows,
  col.names = c("ID", "Subword"),
  caption = "The first rows of the BPE vocabulary",
  row.names = FALSE
)
```

The vocabulary has 250 entries. The first four rows are special markers, then
the boundary marker `▁` and single characters. The marker `▁` means that a
subword begins at a word boundary.

## Compare familiar and unfamiliar text

When the BPE model sees the first sentence, common pieces such as `training`,
`is`, and `provided.` stay together. The rarer `12-week` material fragments.

```{r}
first_sentence_subwords <- bpe_encode(
  bpe_model,
  x = "Paid 12-week training is provided.",
  type = "subwords"
)[[1]]

unseen_phrase_subwords <- bpe_encode(
  bpe_model,
  x = "unpredictability biotechnology",
  type = "subwords"
)[[1]]

subword_summary <- tibble(
  text = c(
    "Paid 12-week training is provided.",
    "unpredictability biotechnology"
  ),
  pieces = c(
    length(first_sentence_subwords),
    length(unseen_phrase_subwords)
  ),
  subwords = c(
    str_c(first_sentence_subwords, collapse = " | "),
    str_c(unseen_phrase_subwords, collapse = " | ")
  )
)

stopifnot(
  identical(
    first_sentence_subwords,
    c(
      "▁Paid",
      "▁1",
      "2",
      "-",
      "w",
      "eek",
      "▁training",
      "▁is",
      "▁provided."
    )
  ),
  identical(length(first_sentence_subwords), 9L),
  identical(length(unseen_phrase_subwords), 21L)
)

knitr::kable(
  subword_summary,
  col.names = c("Text", "Subword pieces", "Subwords"),
  caption = "BPE pieces for familiar and unfamiliar text",
  row.names = FALSE
)
```

The first sentence takes nine subword pieces. The phrase `unpredictability
biotechnology`, which does not appear in the Riverton rows, takes 21 pieces in
this toy model. A tokenizer used with a real language model learns tens of
thousands of pieces from far more text and would likely split that phrase into a
handful of pieces. What carries over is the direction: unfamiliar material costs
more pieces than familiar material, not the size of the effect.

The result gives the team a rule for the meeting: a word count must name the
rule that made the words.

## What to remember

- Tokenization turns text into pieces according to a chosen rule.
- The first Riverton sentence becomes 5, 6, or 8 word tokens under three rules.
- Sentence tokenization preserves the 28 rows in this file.
- BPE represents familiar material with larger pieces and unfamiliar material
  with smaller pieces.
- Text that looks unfamiliar to the model can take more pieces to represent.

The Lab can start counting only after it writes down the token rule. For these
sentences, that rule changes the unit being counted before any summary appears.

## Sources

- [tokenizers documentation for word and sentence tokenizers](https://docs.ropensci.org/tokenizers/reference/basic-tokenizers.html)
- [tokenizers.bpe package documentation](https://cran.r-project.org/package=tokenizers.bpe)
- [Neural Machine Translation of Rare Words with Subword Units](https://aclanthology.org/P16-1162/)
