---
title: "Building a vocabulary from tokens"
subtitle: "Choose which words to keep and measure what falls outside"
description: "Learn how frequency cutoffs and out-of-vocabulary checks shape a small workforce text vocabulary."
categories: [word parsing, vocabulary, workforce research]
---

After the tokenization check, the coordinator prints the Riverton Workforce
Lab's word tokens on a single sheet. The list looks manageable until the
repeated words and one-time words sit side by side. A vocabulary cannot keep
everything in a large project without a rule.

The team decides to test the rule on its small teaching data. It will build a
word list from the job-detail rows, then ask how much of the training flyer
falls outside that list.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- distinguish tokens from types;
- count a word vocabulary and its one-time terms;
- apply minimum-frequency cutoffs;
- measure out-of-vocabulary words on held-out text; and
- compare a word vocabulary with a subword vocabulary.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Count tokens and types

The first visible chunk loads readr for files, dplyr, tidyr, and tibble for
tables, purrr for repeated work, tokenizers for word tokens, and tokenizers.bpe
for subword tokens. A **tibble** is a table that prints its size and column
types.

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(purrr)
library(tokenizers)
library(tokenizers.bpe)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

word_token_rows <- sentences |>
  transmute(
    sentence_id,
    token = tokenize_words(text)
  ) |>
  unnest_longer(token)

word_counts <- word_token_rows |>
  count(token, name = "frequency", sort = TRUE)
top_word_counts <- word_counts |>
  slice_head(n = 8)

knitr::kable(
  top_word_counts,
  col.names = c("Token type", "Frequency"),
  caption = "The eight most frequent word types in all 28 rows",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(word_token_rows), 153L),
  all(!is.na(word_token_rows$sentence_id)),
  identical(nrow(word_counts), 96L),
  identical(sum(word_counts$frequency == 1L), 72L),
  identical(
    top_word_counts$token,
    c(
      "is",
      "a",
      "required",
      "the",
      "training",
      "are",
      "experience",
      "of"
    )
  ),
  identical(
    top_word_counts$frequency,
    c(13L, 7L, 7L, 5L, 5L, 4L, 4L, 3L)
  )
)
```

A **token** is one occurrence. A **type** is one distinct token value. These
counts use the default `tokenizers::tokenize_words()` rule: lowercase text, drop
punctuation, and split hyphenated material. Across the 28 rows, that rule
creates 153 tokens and 96 types. Of those types, 72 appear exactly once.

The most frequent word types are ordinary function words and repeated terms
from the notices: `is` appears 13 times, `a` and `required` appear 7 times each,
`the` and `training` appear 5 times each, `are` and `experience` appear 4 times
each, and `of` appears 3 times.

## Split the data before testing a vocabulary

An honest check cannot build a vocabulary and test it on the same rows. The Lab
uses the 22 job-detail rows to build the word list and the 6 flyer rows to test
what the list has not seen.

```{r}
job_rows <- sentences |>
  filter(document_id != "F001")
flyer_rows <- sentences |>
  filter(document_id == "F001")

job_token_rows <- job_rows |>
  transmute(
    sentence_id,
    token = tokenize_words(text)
  ) |>
  unnest_longer(token)
flyer_token_rows <- flyer_rows |>
  transmute(
    sentence_id,
    token = tokenize_words(text)
  ) |>
  unnest_longer(token)

job_word_counts <- job_token_rows |>
  count(token, name = "frequency", sort = TRUE)

split_summary <- tibble(
  split = c("job-detail rows", "training flyer"),
  rows = c(nrow(job_rows), nrow(flyer_rows)),
  tokens = c(nrow(job_token_rows), nrow(flyer_token_rows)),
  types = c(nrow(job_word_counts), n_distinct(flyer_token_rows$token))
)

knitr::kable(
  split_summary,
  col.names = c("Split", "Rows", "Tokens", "Types"),
  caption = "Word-token counts after splitting by document source",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(job_rows), 22L),
  identical(nrow(flyer_rows), 6L),
  identical(nrow(job_token_rows), 133L),
  identical(nrow(job_word_counts), 86L),
  identical(nrow(flyer_token_rows), 20L),
  all(!is.na(c(job_token_rows$sentence_id, flyer_token_rows$sentence_id)))
)
```

The job-detail rows contain 133 tokens and 86 types. The flyer rows contain 20
tokens. The split is small, but it keeps the test from grading a vocabulary on
the same text that created it.

## Keep words by minimum frequency

A **minimum-frequency cutoff** keeps only types that appear at least a stated
number of times. The code below asks how many job-detail types remain and how
many job-detail tokens those remaining types cover.

The shorthand `\(cutoff)` means "make a small function whose input is
`cutoff`."

```{r}
frequency_cutoffs <- tibble(minimum_frequency = 1:3) |>
  mutate(
    kept_types = map_int(
      minimum_frequency,
      \(cutoff) sum(job_word_counts$frequency >= cutoff)
    ),
    covered_tokens = map_int(
      minimum_frequency,
      \(cutoff) {
        kept <- job_word_counts |>
          filter(frequency >= cutoff) |>
          pull(token)
        sum(job_token_rows$token %in% kept)
      }
    ),
    coverage = covered_tokens / nrow(job_token_rows),
    coverage_label = sprintf(
      "%s/%s (%s%%)",
      covered_tokens,
      nrow(job_token_rows),
      round(100 * coverage)
    )
  )

knitr::kable(
  frequency_cutoffs |>
    select(
      minimum_frequency,
      kept_types,
      coverage_label
    ),
  col.names = c(
    "Minimum frequency",
    "Types kept",
    "Share of job tokens covered"
  ),
  caption = "Vocabulary size and coverage under three cutoffs",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(frequency_cutoffs$kept_types, c(86L, 18L, 8L)),
  identical(frequency_cutoffs$covered_tokens, c(133L, 65L, 45L)),
  identical(
    frequency_cutoffs$coverage_label,
    c("133/133 (100%)", "65/133 (49%)", "45/133 (34%)")
  )
)
```

The minimum-frequency-1 row is a reference point: it keeps every job-detail
type, so full coverage is true by construction. The stricter rows show how many
job-detail tokens lose their own word type under each cutoff.

## Measure out-of-vocabulary words

An **out-of-vocabulary** token is a token that the vocabulary does not contain.
Here the vocabulary comes from the 22 job-detail rows. The test asks which
flyer tokens fall outside that job-built list.

A retrieval system can keep a different kind of vocabulary. A search index
stores whole terms and counts for retrieval, while a tokenizer or subword model
stores the pieces it will use to encode new text.

```{r}
job_vocabulary <- job_word_counts$token
unknown_flyer_rows <- flyer_token_rows |>
  filter(!token %in% job_vocabulary)
unknown_flyer_types <- sort(unique(unknown_flyer_rows$token))
unknown_summary <- tibble(
  measure = c(
    "flyer tokens",
    "unknown flyer tokens",
    "unknown share"
  ),
  value = c(
    as.character(nrow(flyer_token_rows)),
    as.character(nrow(unknown_flyer_rows)),
    sprintf(
      "%s/%s (about half)",
      nrow(unknown_flyer_rows),
      nrow(flyer_token_rows)
    )
  )
)

knitr::kable(
  unknown_summary,
  col.names = c("Measure", "Value"),
  caption = "Out-of-vocabulary rate on the held-out flyer rows",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(unknown_flyer_rows), 10L),
  identical(nrow(flyer_token_rows), 20L),
  all(!is.na(unknown_flyer_rows$sentence_id)),
  identical(
    unknown_summary$value[unknown_summary$measure == "unknown share"],
    "10/20 (about half)"
  ),
  identical(
    unknown_flyer_types,
    c(
      "15",
      "apply",
      "by",
      "classes",
      "house",
      "october",
      "open",
      "riverton",
      "stipend",
      "support"
    )
  )
)
```

The unknown flyer types are `15`, `apply`, `by`, `classes`, `house`, `october`,
`open`, `riverton`, `stipend`, and `support`.

For those 10 token occurrences, the job-built word list has no entry.

## Try the same split with subwords

**Byte pair encoding (BPE)** starts from single characters and repeatedly glues
together the pair that occurs most often. The Lab trains a BPE model on the
job-detail rows only, then encodes the flyer rows. This gives the subword
vocabulary the same held-out test as the word vocabulary.

The word tokenizer lowercases both sources, so the BPE comparison does the same.
Holding case constant matters: otherwise the comparison would change both the
tokenization method and the case policy. The `vocab_size = 200L` argument is a
modeling choice that controls how many pieces the trainer can keep. The
`coverage = 1` argument tells the trainer to keep every character it sees in the
lowercased job-detail text. The special marker `<UNK>` is the model's way of
saying it has no piece for this, short for "unknown".

::: {.callout-note}
The explicit `model_path` sends the model to a temporary file during rendering.
:::

```{r}
job_corpus_file <- tempfile(fileext = ".txt")
write_lines(tolower(job_rows$text), job_corpus_file)

job_bpe_model <- bpe(
  job_corpus_file,
  model_path = tempfile(fileext = ".bpe"),
  vocab_size = 200L,
  coverage = 1,
  threads = 1L
)

flyer_subwords <- bpe_encode(
  job_bpe_model,
  x = tolower(flyer_rows$text),
  type = "subwords"
)
flyer_ids <- bpe_encode(
  job_bpe_model,
  x = tolower(flyer_rows$text),
  type = "ids"
)
flyer_word_ids <- bpe_encode(
  job_bpe_model,
  x = flyer_token_rows$token,
  type = "ids"
)
flat_flyer_subwords <- unlist(flyer_subwords, use.names = FALSE)
flat_flyer_ids <- unlist(flyer_ids, use.names = FALSE)
unknown_id <- job_bpe_model$vocabulary$id[
  job_bpe_model$vocabulary$subword == "<UNK>"
]
unknown_flyer_piece_count <- as.integer(sum(flat_flyer_ids == unknown_id))
flyer_word_bpe <- flyer_token_rows |>
  mutate(
    bpe_ids = flyer_word_ids,
    has_unknown_id = map_lgl(
      bpe_ids,
      \(ids) any(ids == unknown_id)
    )
  )
unknown_flyer_word_count <- sum(flyer_word_bpe$has_unknown_id)
flyer_unknown_by_line <- tibble(
  sentence_id = flyer_rows$sentence_id,
  text = flyer_rows$text,
  pieces = map_int(flyer_ids, length),
  unknown_pieces = map_int(
    flyer_ids,
    \(ids) as.integer(sum(ids == unknown_id))
  )
)
subword_test_summary <- tibble(
  measure = c(
    "BPE vocabulary rows",
    "flyer word tokens",
    "word tokens with an unknown id",
    "flyer subword pieces",
    "unknown piece ids"
  ),
  value = c(
    as.character(nrow(job_bpe_model$vocabulary)),
    as.character(nrow(flyer_word_bpe)),
    as.character(unknown_flyer_word_count),
    as.character(length(flat_flyer_ids)),
    as.character(unknown_flyer_piece_count)
  )
)

knitr::kable(
  subword_test_summary,
  col.names = c("Measure", "Value"),
  caption = "BPE results on the held-out flyer rows",
  row.names = FALSE
)

knitr::kable(
  flyer_unknown_by_line,
  col.names = c("Sentence ID", "Flyer line", "Subword pieces", "Unknown ids"),
  caption = "Unknown subword ids by flyer line",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(job_bpe_model$vocabulary), 200L),
  identical(length(flat_flyer_subwords), 68L),
  identical(length(flat_flyer_ids), 68L),
  identical(as.integer(unknown_id), 1L),
  identical(unknown_flyer_word_count, 0L),
  identical(unknown_flyer_piece_count, 0L),
  identical(flyer_unknown_by_line$sentence_id, flyer_rows$sentence_id),
  identical(flyer_unknown_by_line$unknown_pieces, rep(0L, 6L))
)
```

The like-for-like comparison is now at the word-token level. The whole-word
vocabulary has no entry for 10 of 20 flyer tokens. With both sources lowercased,
none of the 20 word tokens contains a BPE unknown id, and none of the 68 BPE
pieces is unknown. The subword model can assemble unseen words from characters
and pieces learned in the job rows.

That result belongs to this 200-piece model and this case policy. It is not a
general promise that BPE removes out-of-vocabulary material. Case-sensitive
input, a different writing system, a smaller character coverage, or another
tokenizer can produce unknown ids. Comparing 10 word tokens with 68 pieces would
also mix denominators, so the conclusion uses the 20 word-token outcomes.

## What to remember

- Tokens are occurrences; types are distinct token values.
- The 28 Riverton rows produce 153 word tokens and 96 word types.
- Most word types in this small file appear exactly once.
- Frequency cutoffs shrink a vocabulary and reduce token coverage.
- Held-out text exposes out-of-vocabulary words.
- With the same lowercase policy, this BPE model encodes all 20 held-out word
  tokens; that result is setting-specific, not a general guarantee.

These percentages belong only to this 28-sentence teaching corpus. They show
how vocabulary choices behave on the Riverton example, not how often these
patterns occur in workforce text more generally.

## Sources

- [tokenizers documentation for word tokenizers](https://docs.ropensci.org/tokenizers/reference/basic-tokenizers.html)
- [tokenizers.bpe package documentation](https://cran.r-project.org/package=tokenizers.bpe)
- [Transformers tokenization algorithms](https://huggingface.co/docs/transformers/tokenizer_summary)
- [Neural Machine Translation of Rare Words with Subword Units](https://aclanthology.org/P16-1162/)
