---
title: "Building a vocabulary from tokens"
subtitle: "Choose which words to keep and measure what falls outside"
author: "Alex Farach"
description: "Learn how frequency cutoffs and out-of-vocabulary checks shape a small workforce text vocabulary."
categories: [word parsing, vocabulary, workforce research]
---

After the tokenization check, the coordinator prints the Riverton Workforce
Lab's word tokens on a single sheet. The list looks manageable until the
repeated words and one-time words sit side by side. A vocabulary cannot keep
everything in a large project without a rule.

The team decides to test the rule on its small teaching data. It will build a
word list from the job-detail rows, then ask how much of the training flyer
falls outside that list.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- distinguish tokens from types;
- count a word vocabulary and its one-time terms;
- apply minimum-frequency cutoffs;
- measure out-of-vocabulary words on held-out text; and
- compare a word vocabulary with a subword vocabulary.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Count tokens and types

The first visible chunk loads readr for files, dplyr and tibble for tables,
purrr for repeated work, tokenizers for word tokens, and tokenizers.bpe for
subword tokens. A **tibble** is a table that prints its size and column types.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(tokenizers)
library(tokenizers.bpe)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

word_tokens <- unlist(
  tokenize_words(sentences$text),
  use.names = FALSE
)
word_counts <- tibble(token = word_tokens) |>
  count(token, name = "frequency", sort = TRUE)
top_word_counts <- word_counts |>
  slice_head(n = 8)

knitr::kable(
  top_word_counts,
  col.names = c("Token type", "Frequency"),
  caption = "The eight most frequent word types in all 28 rows",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(length(word_tokens), 153L),
  identical(nrow(word_counts), 96L),
  identical(sum(word_counts$frequency == 1L), 72L),
  identical(
    top_word_counts$token,
    c(
      "is",
      "a",
      "required",
      "the",
      "training",
      "are",
      "experience",
      "of"
    )
  ),
  identical(
    top_word_counts$frequency,
    c(13L, 7L, 7L, 5L, 5L, 4L, 4L, 3L)
  )
)
```

A **token** is one occurrence. A **type** is one distinct token value. These
counts use the default `tokenizers::tokenize_words()` rule: lowercase text, drop
punctuation, and split hyphenated material. Across the 28 rows, that rule
creates 153 tokens and 96 types. Of those types, 72 appear exactly once.

The most frequent word types are ordinary function words and repeated terms
from the notices: `is` appears 13 times, `a` and `required` appear 7 times each,
`the` and `training` appear 5 times each, `are` and `experience` appear 4 times
each, and `of` appears 3 times.

## Split the data before testing a vocabulary

An honest check cannot build a vocabulary and test it on the same rows. The Lab
uses the 22 job-detail rows to build the word list and the 6 flyer rows to test
what the list has not seen.

```{r}
job_rows <- sentences |>
  filter(document_id != "F001")
flyer_rows <- sentences |>
  filter(document_id == "F001")

job_tokens <- unlist(
  tokenize_words(job_rows$text),
  use.names = FALSE
)
flyer_tokens <- unlist(
  tokenize_words(flyer_rows$text),
  use.names = FALSE
)
job_word_counts <- tibble(token = job_tokens) |>
  count(token, name = "frequency", sort = TRUE)

split_summary <- tibble(
  split = c("job-detail rows", "training flyer"),
  rows = c(nrow(job_rows), nrow(flyer_rows)),
  tokens = c(length(job_tokens), length(flyer_tokens)),
  types = c(nrow(job_word_counts), n_distinct(flyer_tokens))
)

knitr::kable(
  split_summary,
  col.names = c("Split", "Rows", "Tokens", "Types"),
  caption = "Word-token counts after splitting by document source",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(job_rows), 22L),
  identical(nrow(flyer_rows), 6L),
  identical(length(job_tokens), 133L),
  identical(nrow(job_word_counts), 86L),
  identical(length(flyer_tokens), 20L)
)
```

The job-detail rows contain 133 tokens and 86 types. The flyer rows contain 20
tokens. The split is small, but it keeps the test from grading a vocabulary on
the same text that created it.

## Keep words by minimum frequency

A **minimum-frequency cutoff** keeps only types that appear at least a stated
number of times. The code below asks how many job-detail types remain and how
many job-detail tokens those remaining types cover.

The shorthand `\(cutoff)` means "make a small function whose input is
`cutoff`."

```{r}
frequency_cutoffs <- tibble(minimum_frequency = 1:3) |>
  mutate(
    kept_types = map_int(
      minimum_frequency,
      \(cutoff) sum(job_word_counts$frequency >= cutoff)
    ),
    covered_tokens = map_int(
      minimum_frequency,
      \(cutoff) {
        kept <- job_word_counts |>
          filter(frequency >= cutoff) |>
          pull(token)
        sum(job_tokens %in% kept)
      }
    ),
    coverage = covered_tokens / length(job_tokens),
    coverage_label = sprintf(
      "%s/%s (%s%%)",
      covered_tokens,
      length(job_tokens),
      round(100 * coverage)
    )
  )

knitr::kable(
  frequency_cutoffs |>
    select(
      minimum_frequency,
      kept_types,
      coverage_label
    ),
  col.names = c(
    "Minimum frequency",
    "Types kept",
    "Share of job tokens covered"
  ),
  caption = "Vocabulary size and coverage under three cutoffs",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(frequency_cutoffs$kept_types, c(86L, 18L, 8L)),
  identical(frequency_cutoffs$covered_tokens, c(133L, 65L, 45L)),
  identical(
    frequency_cutoffs$coverage_label,
    c("133/133 (100%)", "65/133 (49%)", "45/133 (34%)")
  )
)
```

The minimum-frequency-1 row is a reference point: it keeps every job-detail
type, so full coverage is true by construction. The stricter rows show how many
job-detail tokens lose their own word type under each cutoff.

## Measure out-of-vocabulary words

An **out-of-vocabulary** token is a token that the vocabulary does not contain.
Here the vocabulary comes from the 22 job-detail rows. The test asks which
flyer tokens fall outside that job-built list.

A retrieval system can keep a different kind of vocabulary. A search index
stores whole terms and counts for retrieval, while a tokenizer or subword model
stores the pieces it will use to encode new text.

```{r}
job_vocabulary <- job_word_counts$token
unknown_flyer_tokens <- flyer_tokens[
  !flyer_tokens %in% job_vocabulary
]
unknown_flyer_types <- sort(unique(unknown_flyer_tokens))
unknown_summary <- tibble(
  measure = c(
    "flyer tokens",
    "unknown flyer tokens",
    "unknown share"
  ),
  value = c(
    as.character(length(flyer_tokens)),
    as.character(length(unknown_flyer_tokens)),
    sprintf(
      "%s/%s (about half)",
      length(unknown_flyer_tokens),
      length(flyer_tokens)
    )
  )
)

knitr::kable(
  unknown_summary,
  col.names = c("Measure", "Value"),
  caption = "Out-of-vocabulary rate on the held-out flyer rows",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length(unknown_flyer_tokens), 10L),
  identical(length(flyer_tokens), 20L),
  identical(
    unknown_summary$value[unknown_summary$measure == "unknown share"],
    "10/20 (about half)"
  ),
  identical(
    unknown_flyer_types,
    c(
      "15",
      "apply",
      "by",
      "classes",
      "house",
      "october",
      "open",
      "riverton",
      "stipend",
      "support"
    )
  )
)
```

The unknown flyer types are `15`, `apply`, `by`, `classes`, `house`, `october`,
`open`, `riverton`, `stipend`, and `support`.

For those 10 token occurrences, the job-built word list has no entry.

## Try the same split with subwords

**Byte pair encoding (BPE)** starts from single characters and repeatedly glues
together the pair that occurs most often. The Lab trains a BPE model on the
job-detail rows only, then encodes the flyer rows. This gives the subword
vocabulary the same held-out test as the word vocabulary.

The `vocab_size = 200L` argument is a modeling choice that controls how many
pieces the trainer can keep. The `coverage = 1` argument tells the trainer to
keep every character it sees in the job-detail text. Characters outside that set
have no piece. The special marker `<UNK>` is the model's way of saying it has no
piece for this, short for "unknown".

::: {.callout-note}
The explicit `model_path` sends the model to a temporary file during rendering.
:::

```{r}
job_corpus_file <- tempfile(fileext = ".txt")
write_lines(job_rows$text, job_corpus_file)

job_bpe_model <- bpe(
  job_corpus_file,
  model_path = tempfile(fileext = ".bpe"),
  vocab_size = 200L,
  coverage = 1,
  threads = 1L
)

flyer_subwords <- bpe_encode(
  job_bpe_model,
  x = flyer_rows$text,
  type = "subwords"
)
flyer_ids <- bpe_encode(
  job_bpe_model,
  x = flyer_rows$text,
  type = "ids"
)
flat_flyer_subwords <- unlist(flyer_subwords, use.names = FALSE)
flat_flyer_ids <- unlist(flyer_ids, use.names = FALSE)
first_flyer_subwords <- flyer_subwords[[1]]
first_flyer_ids <- flyer_ids[[1]]
unknown_id <- job_bpe_model$vocabulary$id[
  job_bpe_model$vocabulary$subword == "<UNK>"
]
unknown_flyer_piece_count <- as.integer(sum(flat_flyer_ids == unknown_id))
unknown_flyer_piece_label <- sprintf(
  "%s/%s (%.1f%%)",
  unknown_flyer_piece_count,
  length(flat_flyer_ids),
  100 * unknown_flyer_piece_count / length(flat_flyer_ids)
)
flyer_unknown_by_line <- tibble(
  text = flyer_rows$text,
  pieces = map_int(flyer_ids, length),
  unknown_pieces = map_int(
    flyer_ids,
    \(ids) as.integer(sum(ids == unknown_id))
  )
)
first_flyer_piece_table <- tibble(
  piece_number = seq_along(first_flyer_subwords),
  displayed_subword = first_flyer_subwords,
  model_id = first_flyer_ids,
  is_unknown_id = first_flyer_ids == unknown_id
)
capital_letters <- function(text) {
  characters <- strsplit(paste(text, collapse = ""), "")[[1]]
  sort(unique(characters[grepl("[A-Z]", characters)]))
}
job_capital_letters <- capital_letters(job_rows$text)
flyer_capital_letters <- capital_letters(flyer_rows$text)
missing_flyer_capitals <- setdiff(flyer_capital_letters, job_capital_letters)
subword_test_summary <- tibble(
  measure = c(
    "BPE vocabulary rows",
    "flyer subword pieces",
    "unknown ids",
    "first flyer line pieces",
    "missing flyer capital letters"
  ),
  value = c(
    as.character(nrow(job_bpe_model$vocabulary)),
    as.character(length(flat_flyer_ids)),
    unknown_flyer_piece_label,
    as.character(length(first_flyer_subwords)),
    paste(missing_flyer_capitals, collapse = " ")
  )
)

knitr::kable(
  subword_test_summary,
  col.names = c("Measure", "Value"),
  caption = "BPE results on the held-out flyer rows",
  row.names = FALSE
)

knitr::kable(
  flyer_unknown_by_line,
  col.names = c("Flyer line", "Subword pieces", "Unknown ids"),
  caption = "Unknown subword ids by flyer line",
  row.names = FALSE
)

knitr::kable(
  first_flyer_piece_table,
  col.names = c("Piece", "Displayed subword", "Model ID", "Unknown ID"),
  caption = "Displayed pieces and model ids for the first flyer line",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(job_bpe_model$vocabulary), 200L),
  identical(length(flat_flyer_subwords), 91L),
  identical(length(flat_flyer_ids), 91L),
  identical(as.integer(unknown_id), 1L),
  identical(unknown_flyer_piece_count, 7L),
  identical(
    unknown_flyer_piece_label,
    "7/91 (7.7%)"
  ),
  identical(
    flyer_unknown_by_line$unknown_pieces,
    c(4L, 3L, 0L, 0L, 0L, 0L)
  ),
  identical(
    job_capital_letters,
    c("A", "C", "E", "N", "O", "P", "R", "S", "T", "W")
  ),
  identical(
    missing_flyer_capitals,
    c("D", "F", "H", "I", "K", "L", "U", "V")
  ),
  identical(length(first_flyer_subwords), 22L),
  identical(
    first_flyer_subwords,
    c(
      "▁R",
      "IV",
      "E",
      "R",
      "T",
      "O",
      "N",
      "▁",
      "S",
      "KILL",
      "S",
      "▁",
      "O",
      "P",
      "E",
      "N",
      "▁",
      "H",
      "O",
      "U",
      "S",
      "E"
    )
  )
)
```

The honest comparison is 10 of 20 flyer word tokens with no entry, versus 7 of
91 subword pieces with the unknown id. Subwords reduce the problem sharply; they
do not remove it.

The display strings from `bpe_encode(type = "subwords")` show text from the
input, not the model's own ids, so they hide unknown pieces. Checking
`bpe_encode(type = "ids")` shows the model result. The unknown pieces appear
because the flyer is printed in capitals and the job-detail rows contain only
the capital letters A, C, E, N, O, P, R, S, T, and W. The flyer also uses D, F,
H, I, K, L, U, and V. Here, the `7/91` count belongs to this 200-piece
character-based BPE model with `coverage = 1` and an `<UNK>` id for unseen
characters. Other tokenizers use different out-of-vocabulary policies. A
byte-level tokenizer can avoid this particular unknown-piece problem, while
another subword vocabulary may still leave unknown material in a different
place.

## What to remember

- Tokens are occurrences; types are distinct token values.
- The 28 Riverton rows produce 153 word tokens and 96 word types.
- Most word types in this small file appear exactly once.
- Frequency cutoffs shrink a vocabulary and reduce token coverage.
- Held-out text exposes out-of-vocabulary words.
- Subwords reduce unknown material here, but they do not remove it.

These percentages belong only to this 28-sentence teaching corpus. They show
how vocabulary choices behave on the Riverton example, not how often these
patterns occur in workforce text more generally.

## Sources

- [tokenizers documentation for word tokenizers](https://docs.ropensci.org/tokenizers/reference/basic-tokenizers.html)
- [tokenizers.bpe package documentation](https://cran.r-project.org/package=tokenizers.bpe)
- [Transformers tokenization algorithms](https://huggingface.co/docs/transformers/tokenizer_summary)
- [Neural Machine Translation of Rare Words with Subword Units](https://aclanthology.org/P16-1162/)
