---
title: "Looking inside words with a dictionary"
subtitle: "Use hunspell analyses without mistaking them for contextual morphology"
author: "Alex Farach"
description: "Learn what an English spelling dictionary can and cannot say about stems, affixes, ambiguity, and context in workforce text."
categories: [word parsing, morphology, workforce research]
---

Twenty-eight reviewed sentences come from a job board page and a training
flyer. The team wants to know whether `shifts` and `shift` should
count as the same word when it summarizes the vocabulary.

That question matters because a worker may search for training, while a job
posting may say `shifts`, `certification`, or `certificate`. The team needs a
way to look inside a word before it decides what to group.

**Morphology** is the internal structure of words. A **stem** carries the core
meaning, and **affixes** are added pieces such as plural `s` or past-tense `ed`.
English marks a small amount of grammar this way, so English speakers often
notice word parts only when a form looks odd.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The dictionary output is real and comes from the
local hunspell package.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- describe a stem and an affix in plain language;
- read a dictionary-based word analysis;
- distinguish a useful decomposition from a misleading one;
- find unanalyzed and ambiguous word types in a small vocabulary; and
- explain why dictionary analysis differs from contextual morphology in
  sentences.
:::

This lesson takes a narrow first step. It asks what an English spelling
dictionary can say about isolated word forms. That is different from
**contextual morphology**, where a tagger or morphologizer labels a token in a
sentence with feature bundles such as `Number=Plur` or `Tense=Past`.

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Read the workforce sentences

To read the sentences, the code brings in readr for files, dplyr and tibble for
tables, purrr for repeated list checks, stringr for text checks, tokenizers for
word splitting, and hunspell for dictionary analysis. It then reads the 28
reviewed workforce sentences.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(hunspell)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

document_counts <- sentences |>
  count(document_id, name = "sentence_count") |>
  arrange(document_id)

stopifnot(
  identical(nrow(sentences), 28L),
  identical(
    document_counts$document_id,
    c("F001", sprintf("J%03d", 1:6))
  ),
  identical(
    document_counts$sentence_count,
    c(6L, 4L, 4L, 4L, 3L, 3L, 4L)
  ),
  str_detect(sentences$text[1], fixed("12-week training"))
)

knitr::kable(
  document_counts,
  col.names = c("Document ID", "Sentences"),
  caption = "The 28 Riverton text units by source document",
  row.names = FALSE
)
```

The flyer has six rows, and the job-detail documents have 22 rows. The sample
fits in a small table, but the vocabulary is large enough for several kinds of
word behavior.

## Ask the dictionary for analyses

`hunspell_analyze()` checks each word against an English spelling dictionary and
its affix rules. The output is compact. `st:` gives a stem, and `fl:` gives a
flag used by the dictionary's rule system. A flag is a code in the dictionary,
not a reader-friendly grammar label.

In the code, `\(value)` is R shorthand for a small function. Here it means that
each returned vector is named `value` before it is pasted into one string.

```{r}
example_words <- c(
  "shifts",
  "requires",
  "certification",
  "running",
  "forklift"
)

word_analyses <- hunspell_analyze(example_words)
word_stems <- hunspell_stem(example_words)

analysis_table <- tibble(
  word = example_words,
  analyses = map_chr(word_analyses, \(value) paste(value, collapse = "; ")),
  stems = map_chr(word_stems, \(value) paste(value, collapse = ", "))
)

stopifnot(
  identical(word_analyses[[1]], " st:shift fl:S"),
  identical(word_analyses[[2]], " fl:A st:quire fl:S"),
  identical(
    word_analyses[[3]],
    c(" st:certification", " st:certificate fl:N")
  ),
  identical(word_analyses[[4]], " st:running"),
  identical(word_analyses[[5]], " st:forklift"),
  identical(word_stems[[1]], "shift"),
  identical(word_stems[[2]], "quire"),
  identical(word_stems[[3]], c("certification", "certificate")),
  identical(word_stems[[4]], "running"),
  identical(word_stems[[5]], "forklift")
)

knitr::kable(
  analysis_table,
  col.names = c("Word", "Allowed analyses", "Returned stems"),
  caption = "Five words analyzed by hunspell",
  row.names = FALSE
)
```

For `shifts`, the dictionary finds the stem `shift` and marks the plural form
with flag `S`.

The entry for `requires` gives a warning. Hunspell can produce an analysis that
is internally legal and completely false. It allows `re-` in front of dictionary
words, and `quire` happens to be a dictionary word. The two have nothing to do
with each other: `require` came into English whole, from Old French `requerre`,
from Latin `requirere`. The rule found a path, not a history.

`certification` has two analyses. It can stand as `certification`, or it can be
connected to `certificate` through another rule. The letters alone do not choose
between those analyses.

With `running`, the word is listed as its own dictionary entry, so the analyzer
returns `running` and stops there. A common form can be stored directly.

## Count what the dictionary can see

The team next applies the same dictionary check to the vocabulary of the 28
sentences. A **word type** is one distinct word form after lowercasing. If
`training` appears several times, it counts once in this vocabulary.

```{r}
tokens_by_sentence <- tokenize_words(
  sentences$text,
  lowercase = TRUE,
  strip_punct = TRUE
)
vocabulary <- sort(
  unique(unlist(tokens_by_sentence, use.names = FALSE))
)
vocabulary_analyses <- hunspell_analyze(vocabulary)

vocabulary_summary <- tibble(
  word = vocabulary,
  analyses = vocabulary_analyses,
  analysis_count = map_int(analyses, length),
  recognized = analysis_count > 0L
)

summary_counts <- vocabulary_summary |>
  summarise(
    word_types = n(),
    analyzed_by_dictionary = sum(recognized),
    not_analyzed = sum(!recognized),
    more_than_one_analysis = sum(analysis_count > 1L),
    .groups = "drop"
  )

unanalyzed_words <- vocabulary_summary |>
  filter(!recognized) |>
  pull(word)

multi_analysis_words <- vocabulary_summary |>
  filter(analysis_count > 1L) |>
  pull(word)

stopifnot(
  identical(summary_counts$word_types, 96L),
  identical(summary_counts$analyzed_by_dictionary, 91L),
  identical(summary_counts$not_analyzed, 5L),
  identical(summary_counts$more_than_one_analysis, 8L),
  identical(unanalyzed_words, c("12", "15", "50", "october", "riverton")),
  identical(
    multi_analysis_words,
    c(
      "available",
      "certification",
      "communication",
      "customer",
      "evening",
      "outdoors",
      "provided",
      "training"
    )
  )
)

knitr::kable(
  summary_counts,
  col.names = c(
    "Word types",
    "Analyzed by dictionary",
    "Not analyzed",
    "More than one dictionary analysis"
  ),
  caption = "Dictionary coverage for the Riverton vocabulary",
  row.names = FALSE
)
```

The dictionary analyzes 91 of the 96 word types. Five receive no analysis: three
numbers, the month `october`, and the place name `riverton`. Eight word types
receive more than one analysis, which means the word form alone leaves more than
one route through the dictionary.

## Inspect the ambiguous cases

Ambiguity is not a bug by itself. A single spelling can have several plausible
paths through a dictionary. The next table shows the Riverton word types where
hunspell returned more than one analysis.

```{r}
ambiguous_words <- vocabulary_summary |>
  filter(analysis_count > 1L) |>
  transmute(
    word,
    analyses = map_chr(analyses, \(value) paste(value, collapse = "; "))
  )

stopifnot(
  identical(nrow(ambiguous_words), 8L),
  identical(ambiguous_words$word[2], "certification"),
  str_detect(
    ambiguous_words$analyses[2],
    fixed("st:certificate fl:N")
  ),
  identical(ambiguous_words$word[8], "training")
)

knitr::kable(
  ambiguous_words,
  col.names = c("Word", "Allowed analyses"),
  caption = "Riverton word types with more than one dictionary analysis",
  row.names = FALSE
)
```

For the Lab, this table is useful because it slows the decision down. The
dictionary has supplied possible word structures, not final labels. A person
still has to decide whether the structure helps the question at hand.

## Keep dictionary analysis separate from contextual morphology

English puts some information inside word endings: `shifts` marks plural, and
`provided` marks a form related to tense or aspect. It leaves much else to word
order and nearby helper words. Languages that mark case, person, tense, and
number on the word itself place more information inside a single token. A tool
built around English spelling assumptions will miss much of that structure.

Hunspell is a spelling dictionary with affix rules. It is not a universal
morphological analyzer. It can produce an analysis that is internally legal and
completely false. Hunspell allows `re-` in front of dictionary words, and
`quire` happens to be a dictionary word. The rule found a path, not a history.

This lesson stays in a narrower lane. In NLP, contextual morphology usually
means assigning feature bundles such as `Number=Plur` or `Tense=Past` to a
token in a sentence. Those features live in the `feats` column of a treebank
and can be produced by a tagger or morphologizer, including the tagger in
Lesson 17. This page does not do that. It runs a spelling dictionary over an
out-of-context list of word types, which is a different task with a different
input and output.

## What to remember

- Morphology studies the parts inside words.
- A stem carries the core meaning; affixes modify the form.
- Dictionary rules can produce useful analyses and misleading ones.
- One word form can have more than one allowed analysis.
- Hunspell checks spelling-style rules on isolated words, not contextual
  morphology in sentences.

The team can group `shifts` with `shift` for a plain vocabulary count, but every
returned stem still needs human judgment about the text. Feature bundles in
running sentences are a different task.

## Sources

- [hunspell package reference](https://docs.ropensci.org/hunspell/reference/hunspell.html)
- [Hunspell project documentation](https://hunspell.github.io/)
- [hunspell package repository](https://github.com/ropensci/hunspell)
