---
title: "Tagging words by their job"
subtitle: "Read part-of-speech tags from a small trained model"
author: "Alex Farach"
description: "Learn how a part-of-speech tagger labels workforce text and why held-out accuracy has limits."
categories: [word parsing, part-of-speech tagging, workforce research]
---

With sentence labels finished, attention turns to single words. The team points
to a job posting and asks a smaller question: which words are the actions?

That question changes the level of analysis. The team is no longer asking
whether a whole sentence is about training or schedules. It wants to know what
each word is doing inside the sentence.

A **part-of-speech tagger** assigns a grammatical tag to each token. The tag can
say that a word is a thing, an action, a description, a helper verb, a number,
or punctuation. This lesson uses a UDPipe tagger trained on 500 hand-labeled
sentences from the Universal Dependencies English Web Treebank. A **treebank**
is a collection of sentences that people have labelled by hand, word by word.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The UDPipe model and treebank artifacts are local
files bundled with this project, so the lesson does not download anything while
it runs.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- describe common Universal POS tags in plain language;
- run a local UDPipe part-of-speech tagger on fixed tokens;
- inspect a tagging mistake in a job-posting sentence;
- read a held-out accuracy score; and
- explain why accuracy on weblog posts does not promise the same result on job
  postings.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the local files

For this read-only pass, readr opens the files, dplyr and tibble handle tables,
purrr repeats token work, tokenizers split words, udpipe tags them, and digest
checks file fingerprints. The model has already been trained, so the code only
reads artifacts from disk and records the settings needed to describe the
bundled run.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(tokenizers)
library(stringr)
library(udpipe)
library(digest)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

learning_curve <- read_csv(
  "data/treebank/tagger-learning-curve.csv",
  na = character(),
  col_types = cols(
    training_sentences = col_integer(),
    training_tokens = col_integer(),
    held_out_tokens = col_integer(),
    correct_tokens = col_integer(),
    accuracy = col_double()
  )
)

metadata <- read_csv(
  "data/treebank/treebank-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    source_url = col_character(),
    retrieved_on = col_character(),
    license = col_character(),
    fingerprint = col_character()
  )
)

builder_script_path <- "data-raw/build-treebank-tagger.R"
builder_seed_match <- read_lines(builder_script_path) |>
  str_subset("^set.seed\\(") |>
  str_match("^set.seed\\((\\d+)\\)")
builder_seed <- as.integer(builder_seed_match[1, 2])
rng_settings <- RNGkind()

learning_curve_path <- "data/treebank/tagger-learning-curve.csv"
learning_curve_fingerprint <- digest(
  paste(read_lines(learning_curve_path), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)
expected_curve_fingerprint <- metadata |>
  filter(artifact == "tagger-learning-curve.csv") |>
  pull(fingerprint)

metadata_display <- metadata |>
  filter(
    artifact %in% c(
      "en_ewt-train-excerpt.conllu",
      "en_ewt-held-out.conllu",
      "en_ewt-500-tagger.udpipe",
      "tagger-learning-curve.csv"
    )
  ) |>
  mutate(fingerprint_start = str_sub(fingerprint, 1, 12)) |>
  select(artifact, source, retrieved_on, license, fingerprint_start)

reproducibility_table <- tibble(
  item = c(
    "Builder seed from script",
    "Current R RNG kind",
    "Current normal generator",
    "Current sample generator",
    "Current R version"
  ),
  value = c(
    as.character(builder_seed),
    rng_settings[[1]],
    rng_settings[[2]],
    rng_settings[[3]],
    R.version.string
  )
)

knitr::kable(
  metadata_display,
  col.names = c("Artifact", "Source", "Retrieved", "License", "SHA-256 start"),
  caption = "Local treebank artifacts and provenance used by the tagger lesson",
  row.names = FALSE
)

knitr::kable(
  reproducibility_table,
  col.names = c("Reproducibility item", "Value"),
  caption = "Recorded builder seed and the current rerun environment",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(learning_curve), 5L),
  identical(nrow(metadata_display), 4L),
  identical(builder_seed, 20260828L),
  identical(length(rng_settings), 3L),
  all(str_length(metadata$fingerprint) == 64L),
  all(str_length(metadata_display$fingerprint_start) == 12L),
  identical(
    learning_curve$training_sentences,
    c(100L, 250L, 500L, 1000L, 2000L)
  ),
  identical(
    learning_curve$correct_tokens,
    c(3301L, 3496L, 3641L, 3720L, 3778L)
  ),
  identical(
    learning_curve_fingerprint,
    "b1c614e3cdd17ae13b1959992cb440e43c7bf4533e936cd6cb58dae42f255911"
  ),
  identical(learning_curve_fingerprint, expected_curve_fingerprint)
)
```

The table prints the first 12 characters of each SHA-256 fingerprint so it stays
readable at high zoom. The full values are in
`data/treebank/treebank-metadata.csv`. A hash shows that a file has not changed
since it was made. It says nothing about whether the numbers in the file are
right. The second table mixes one build setting that is recorded in the builder
script with details from the current rerun environment. The committed model does
not carry build-time RNG kind, R version, or package-version metadata, so those
details would need to be recorded on the next rebuild. Only the 500-sentence
tagger is included in this repository, so the other four rows of the learning
curve cannot be rebuilt from the committed files.

## Tag fixed tokens

The model was trained to tag tokens it receives. It was not trained to split raw
sentences. The code therefore tokenizes each Riverton sentence first, writes one
token per line, and tells UDPipe that the input is already vertical text.

The output includes a **lemma**, the dictionary form of a word, the form you
would look up. `is` has the lemma `be`. In the code, `\(tokens)` is R shorthand
for a small function that handles one sentence's tokens at a time.

```{r}
tagger <- udpipe_load_model(
  "data/treebank/en_ewt-500-tagger.udpipe"
)

vertical <- sentences$text |>
  tokenize_words(lowercase = FALSE, strip_punct = FALSE) |>
  map_chr(\(tokens) paste(tokens, collapse = "\n"))

tagged <- udpipe_annotate(
  tagger,
  x = vertical,
  doc_id = sentences$sentence_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "none"
) |>
  as.data.frame() |>
  as_tibble()

paid_sentence_tags <- tagged |>
  filter(doc_id == "s001") |>
  select(token, upos, lemma)

knitr::kable(
  paid_sentence_tags,
  col.names = c("Token", "POS tag", "Lemma"),
  caption = "Tags for 'Paid 12-week training is provided.'",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(tagged), 178L),
  identical(
    paid_sentence_tags$token,
    c("Paid", "12", "-", "week", "training", "is", "provided", ".")
  ),
  identical(
    paid_sentence_tags$upos,
    c("PROPN", "NUM", "PUNCT", "NOUN", "NOUN", "AUX", "VERB", "PUNCT")
  ),
  identical(
    paid_sentence_tags$lemma,
    c("Paid", "12", "-", "week", "training", "be", "provid", ".")
  )
)
```

The sentence has eight tokens after keeping punctuation. Two outputs are wrong
in a way a reader can judge. `Paid` is not a proper noun. It describes the
training, and the tagger likely followed the capital letter at the start of the
sentence. The lemma `provid` for `provided` is not a word. The model guessed a
stem from a suffix rule.

## Count the tags

The tagger uses **Universal POS tags**. The **tagset** is the fixed list of
labels a tagger is allowed to choose from. `NOUN` marks things, `VERB` marks
actions, `ADJ` marks descriptions, `ADP` marks words such as `of` and `in`,
`AUX` marks helper verbs such as `is`, `DET` marks words such as `the` and `a`,
`PART` marks particles such as `to` in `to apply`, `PROPN` marks names, `PRON`
marks pronouns, `ADV` marks adverbs, `CCONJ` marks coordinating conjunctions
such as `and`, and `INTJ` marks interjections.

```{r}
tag_counts <- tagged |>
  count(upos, sort = TRUE)

punctuation_tokens <- sum(tagged$upos == "PUNCT")
tokens_without_punctuation <- nrow(tagged) - punctuation_tokens

intj_tokens <- tagged |>
  filter(upos == "INTJ") |>
  select(doc_id, token)

propn_tokens <- tagged |>
  filter(upos == "PROPN") |>
  select(doc_id, token)

expected_tag_counts <- tibble(
  upos = c(
    "NOUN",
    "PUNCT",
    "VERB",
    "AUX",
    "ADJ",
    "DET",
    "PROPN",
    "ADP",
    "NUM",
    "PART",
    "PRON",
    "ADV",
    "CCONJ",
    "INTJ"
  ),
  n = c(45L, 25L, 24L, 20L, 17L, 16L, 11L, 7L, 6L, 2L, 2L, 1L, 1L, 1L)
)

knitr::kable(
  tag_counts,
  col.names = c("POS tag", "Tagged tokens"),
  caption = "Universal POS tag counts across the 28 Riverton sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(tag_counts, expected_tag_counts),
  identical(punctuation_tokens, 25L),
  identical(tokens_without_punctuation, 153L),
  identical(intj_tokens$doc_id, "s002"),
  identical(intj_tokens$token, "No"),
  identical(
    propn_tokens$token,
    c(
      "Paid",
      "Clear",
      "RIVERTON",
      "SKILLS",
      "OPEN",
      "HOUSE",
      "DATA",
      "SUPPORT",
      "CERTIFICATE",
      "Paid",
      "October"
    )
  )
)
```

Nouns are the largest group, followed by punctuation, verbs, and helper verbs.
That matches the practical question well enough to start a review, but the
mistakes in the first sentence warn against treating the table as a perfect
grammar of the job postings.

This lesson keeps punctuation as tokens, so it tags 178 tokens. Twenty-five are
`PUNCT`; dropping them leaves the 153 word tokens counted in Lesson 15.

One row deserves a closer look. The only `INTJ` is `No` in sentence `s002`,
which is wrong in context. Of the 11 `PROPN` tokens, seven are all-capital flyer
words (`RIVERTON`, `SKILLS`, `OPEN`, `HOUSE`, `DATA`, `SUPPORT`, `CERTIFICATE`),
two are `Paid` at the start of sentences `s001` and `s025`, one is `Clear` in
`s022`, and one is `October` in `s028`, which is right. The tagger reads capital
letters as a signal for names, a systematic failure on this kind of text that is
invisible in an accuracy score.

## Read the accuracy number

This score is for 13 held-out political weblog posts from the same treebank, not
for job postings. **Held-out** means text kept back from training so it can be
reserved for scoring. The posts in both treebank excerpts are weblog
documents from 2003 to 2006, which strengthens the domain-transfer warning.

```{r}
train_path <- "data/treebank/en_ewt-train-excerpt.conllu"
held_out_path <- "data/treebank/en_ewt-held-out.conllu"

read_gold_tokens <- function(path) {
  udpipe_read_conllu(path) |>
    as_tibble() |>
    filter(!str_detect(token_id, "[^0-9]"), !is.na(upos))
}

train_newdoc_ids <- read_lines(train_path) |>
  str_subset(fixed("# newdoc id = ")) |>
  str_remove(fixed("# newdoc id = "))

held_newdoc_ids <- read_lines(held_out_path) |>
  str_subset(fixed("# newdoc id = ")) |>
  str_remove(fixed("# newdoc id = "))

treebank_years <- str_match(
  c(train_newdoc_ids, held_newdoc_ids),
  "_ENG_(\\d{4})"
)[, 2] |>
  as.integer()

train_gold <- read_gold_tokens(train_path) |>
  transmute(token, gold_upos = upos)

held <- read_gold_tokens(held_out_path) |>
  transmute(sentence_id, token, gold_upos = upos)

lookup <- train_gold |>
  count(token, gold_upos, name = "n") |>
  arrange(token, desc(n), gold_upos) |>
  group_by(token) |>
  slice(1) |>
  ungroup() |>
  select(token, baseline_upos = gold_upos)

held_vertical <- held |>
  group_by(sentence_id) |>
  summarise(text = paste(token, collapse = "\n"), .groups = "drop")

held_predictions <- udpipe_annotate(
  tagger,
  x = held_vertical$text,
  doc_id = held_vertical$sentence_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "none"
) |>
  as.data.frame() |>
  as_tibble() |>
  group_by(doc_id) |>
  mutate(token_row = row_number()) |>
  ungroup() |>
  transmute(
    sentence_id = doc_id,
    token_row,
    tagger_upos = upos
  )

scored <- held |>
  left_join(lookup, by = "token") |>
  mutate(
    baseline_upos = coalesce(baseline_upos, "NOUN"),
    seen_in_training = token %in% train_gold$token
  ) |>
  group_by(sentence_id) |>
  mutate(token_row = row_number()) |>
  ungroup() |>
  left_join(held_predictions, by = c("sentence_id", "token_row")) |>
  mutate(
    baseline_correct = baseline_upos == gold_upos,
    tagger_correct = tagger_upos == gold_upos,
    easy_gold_tag = gold_upos %in% c("PUNCT", "DET")
  )

scored_500 <- learning_curve |>
  filter(training_sentences == 500L) |>
  mutate(
    calculated_accuracy = correct_tokens / held_out_tokens,
    accuracy_percent = round(calculated_accuracy * 100, 2)
  )

learning_curve_display <- learning_curve |>
  mutate(
    accuracy_percent = round(accuracy * 100, 2)
  )

gain_100_to_500 <- learning_curve |>
  filter(training_sentences %in% c(100L, 500L)) |>
  summarise(
    points = diff(accuracy) * 100,
    .groups = "drop"
  ) |>
  pull(points)

gain_500_to_2000 <- learning_curve |>
  filter(training_sentences %in% c(500L, 2000L)) |>
  summarise(
    points = diff(accuracy) * 100,
    .groups = "drop"
  ) |>
  pull(points)

training_ratio_100_to_500 <- 500L / 100L
training_ratio_500_to_2000 <- 2000L / 500L

accuracy_summary <- scored |>
  summarise(
    held_out_tokens = n(),
    lookup_correct = sum(baseline_correct),
    lookup_accuracy = lookup_correct / held_out_tokens,
    tagger_correct = sum(tagger_correct),
    tagger_accuracy = tagger_correct / held_out_tokens,
    gain_points = (tagger_accuracy - lookup_accuracy) * 100,
    seen_tokens = sum(seen_in_training),
    seen_percent = seen_tokens / held_out_tokens * 100,
    easy_tokens = sum(easy_gold_tag),
    easy_percent = easy_tokens / held_out_tokens * 100,
    .groups = "drop"
  )

content_summary <- scored |>
  filter(!easy_gold_tag) |>
  summarise(
    held_out_tokens = n(),
    lookup_correct = sum(baseline_correct),
    lookup_accuracy = lookup_correct / held_out_tokens,
    tagger_correct = sum(tagger_correct),
    tagger_accuracy = tagger_correct / held_out_tokens,
    .groups = "drop"
  )

accuracy_comparison <- tibble(
  scope = c(
    "All held-out tokens",
    "All held-out tokens",
    "Excluding PUNCT and DET",
    "Excluding PUNCT and DET"
  ),
  system = c(
    "Lookup baseline",
    "500-sentence tagger",
    "Lookup baseline",
    "500-sentence tagger"
  ),
  correct = c(
    accuracy_summary$lookup_correct,
    accuracy_summary$tagger_correct,
    content_summary$lookup_correct,
    content_summary$tagger_correct
  ),
  tokens = c(
    accuracy_summary$held_out_tokens,
    accuracy_summary$held_out_tokens,
    content_summary$held_out_tokens,
    content_summary$held_out_tokens
  ),
  accuracy = round(correct / tokens, 4)
)

knitr::kable(
  learning_curve_display,
  col.names = c(
    "Training sentences",
    "Training tokens",
    "Held-out tokens",
    "Correct tokens",
    "Accuracy",
    "Accuracy percent"
  ),
  caption = "Held-out tagger accuracy as training data increases",
  row.names = FALSE
)

knitr::kable(
  accuracy_comparison,
  col.names = c("Scope", "System", "Correct", "Tokens", "Accuracy"),
  caption = "The 500-sentence tagger compared with a lookup baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length(train_newdoc_ids), 11L),
  identical(length(held_newdoc_ids), 13L),
  all(str_starts(train_newdoc_ids, "weblog-")),
  all(str_starts(held_newdoc_ids, "weblog-")),
  identical(range(treebank_years), c(2003L, 2006L)),
  identical(scored_500$correct_tokens, 3641L),
  identical(scored_500$held_out_tokens, 4007L),
  identical(round(scored_500$calculated_accuracy, 4), 0.9087),
  identical(round(accuracy_summary$lookup_accuracy, 4), 0.7931),
  identical(accuracy_summary$lookup_correct, 3178L),
  identical(accuracy_summary$tagger_correct, 3641L),
  identical(accuracy_summary$seen_tokens, 3140L),
  identical(round(accuracy_summary$seen_percent, 1), 78.4),
  identical(accuracy_summary$easy_tokens, 884L),
  identical(round(accuracy_summary$easy_percent, 1), 22.1),
  identical(content_summary$held_out_tokens, 3123L),
  identical(content_summary$tagger_correct, 2766L),
  identical(round(content_summary$tagger_accuracy, 4), 0.8857),
  identical(content_summary$lookup_correct, 2307L),
  identical(round(content_summary$lookup_accuracy, 4), 0.7387),
  identical(round(accuracy_summary$gain_points, 1), 11.6),
  identical(round(gain_100_to_500, 1), 8.5),
  identical(round(gain_500_to_2000, 1), 3.4),
  identical(training_ratio_100_to_500, 5),
  identical(training_ratio_500_to_2000, 4)
)
```

The lookup baseline gives each word its most frequent tag in the 500 training
sentences and guesses `NOUN` for words it has not seen. It gets 3,178 of 4,007
tokens right, or 0.7931. The trained tagger gets 3,641 of 4,007 right, or about
91 percent, which buys about 11.6 points over the lookup table. In the held-out
posts, 3,140 of 4,007 tokens, or 78.4 percent, are words that also appear in the
training sentences.

The easy tags matter. `PUNCT` and `DET` are 884 of the 4,007 held-out tokens, or
22.1 percent. Excluding them, the trained tagger scores 2,766 of 3,123, or
0.8857, and the lookup baseline scores 2,307 of 3,123, or 0.7387. The easy tokens
flatter both systems.

On this one learning curve, accuracy rises quickly at first and then more slowly.
Moving from 100 to 500 training sentences buys about 8.5 points. Moving from 500
to 2,000 buys about 3.4 more. The curve cannot separate more labeled sentences
from broader topic coverage, because each step adds new blog posts as well as
new sentences. The intervals are not the same scale: 100 to 500 is a five-fold
increase, while 500 to 2,000 is a four-fold increase.

## Keep the limits attached

The score is one number from one run over thirteen blog posts. Tokens inside a
post are not independent, because vocabulary, topic, and author repeat within
the post. The real spread is wider than the token count suggests. The builder
script sets `set.seed(20260828)`. The RNG and R-version rows above describe the
current rerun environment, not the build-time environment of the committed
model. This page still reports the bundled run only and does not add a
confidence interval, because it does not retrain or rescore on new samples or
domains.

The Riverton sentences are job postings and flyer lines, a different kind of
writing. This lesson does not claim the same rate there. On the Riverton text,
we can point to individual mistakes a reader can judge.

Universal POS tags are a useful set of bins for this task. Other projects can
choose a different tagset.

## What to remember

- A part-of-speech tagger assigns a grammatical tag to each token.
- Universal POS tags are compact labels such as `NOUN`, `VERB`, and `ADJ`.
- The local tagger labeled 178 Riverton tokens.
- Its held-out weblog-post accuracy is 3,641 of 4,007 tokens, or about 91
  percent.
- Job-posting errors must be checked directly; the weblog-post score does not
  transfer by promise.

The team can use the tagger to make a first action-word list, then inspect the
mistakes before treating the labels as evidence.

## Sources

- [UDPipe R package reference](https://bnosac.github.io/udpipe/docs/doc0.html)
- [Universal POS tags](https://universaldependencies.org/u/pos/)
- [Universal Dependencies English Web Treebank](https://universaldependencies.org/treebanks/en_ewt/index.html)
