---
title: "Drawing grammar links between words"
subtitle: "Read dependency parses from a small trained model"
author: "Alex Farach"
description: "Learn how a dependency parser links words in workforce text and why attachment scores need limits."
categories: [word parsing, dependency parsing, workforce research]
---

Word labels are ready for the 28 Riverton sentences. The team then faces a
harder decision: when a sentence says training is provided, which word
is the thing provided, and does the sentence name a provider?

That question matters because labels alone do not say how words fit together. A
sentence can contain `experience` and `required` while still saying the opposite
of a simple keyword match. A passive sentence can also leave the provider
unnamed, so the team needs links between words and a habit of checking them by
eye.

A **dependency parse** says which word each token attaches to and what the link
is called. The link points from one token to its **head**, the word it depends
on. This lesson uses a local UDPipe model trained on the same 500-sentence
English Web Treebank excerpt as the tagger in Lesson 17, with parsing turned on.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The parser and scoring files are local artifacts,
so the lesson does not download a model while it runs.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- describe a dependency parse in plain language;
- read root, compound, numeric modifier, passive subject, and punctuation links;
- compare two parses of the same sentence; and
- explain why a parser score from weblog posts needs a baseline and local checks.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the local parser inputs

The parser inputs come from local CSV files. dplyr, tibble, and purrr handle
the table work; tokenizers and udpipe keep parsing tied to fixed tokens; digest
records artifact fingerprints; and spacyr starts a comparison parse. The
project helper `use_project_spacy()` starts the pinned spaCy pipeline.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(digest)
library(tokenizers)
library(udpipe)
library(spacyr)
source("R/use-spacy.R")

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

parser_score_path <- "data/treebank/parser-held-out-score.csv"
parser_model_path <- "data/treebank/en_ewt-500-parser.udpipe"
held_out_path <- "data/treebank/en_ewt-held-out.conllu"

parser_score <- read_csv(
  parser_score_path,
  na = character(),
  col_types = cols(
    measure = col_character(),
    value = col_double()
  )
)

metadata <- read_csv(
  "data/treebank/treebank-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    source_url = col_character(),
    retrieved_on = col_character(),
    license = col_character(),
    fingerprint = col_character()
  )
)

parser_score_fingerprint <- digest(
  paste(read_lines(parser_score_path), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)
parser_model_fingerprint <- digest(
  parser_model_path,
  algo = "sha256",
  file = TRUE
)
expected_score_fingerprint <- metadata |>
  filter(artifact == "parser-held-out-score.csv") |>
  pull(fingerprint)
expected_model_fingerprint <- metadata |>
  filter(artifact == "en_ewt-500-parser.udpipe") |>
  pull(fingerprint)

held_out_lines <- read_lines(held_out_path)
held_out_tokens <- tibble(
  line = held_out_lines,
  sentence_index = cumsum(grepl("^# sent_id", held_out_lines))
) |>
  filter(!grepl("^#", line), nzchar(line)) |>
  mutate(fields = strsplit(line, "\t", fixed = TRUE)) |>
  filter(map_int(fields, length) >= 8L) |>
  transmute(
    sentence_index,
    token_id = map_chr(fields, 1),
    head_token_id = map_chr(fields, 7),
    dep_rel = map_chr(fields, 8)
  ) |>
  filter(
    !grepl("-", token_id, fixed = TRUE),
    !grepl(".", token_id, fixed = TRUE)
  )

stored_held_out_tokens <- as.integer(parser_score$value[parser_score$measure == "held_out_tokens"])
stored_correct_head <- as.integer(parser_score$value[parser_score$measure == "correct_head"])
stored_correct_head_and_relation <- as.integer(
  parser_score$value[parser_score$measure == "correct_head_and_relation"]
)
computed_unlabelled <- round(stored_correct_head / stored_held_out_tokens, 4)
computed_labelled <- round(stored_correct_head_and_relation / stored_held_out_tokens, 4)

pipeline <- use_project_spacy()
spacy_info <- spacy_pipeline_version()
spacy_info_table <- tibble(
  item = names(spacy_info),
  value = unname(unlist(spacy_info))
)

knitr::kable(
  parser_score,
  col.names = c("Measure", "Value"),
  caption = "Held-out parser scores bundled with the lesson",
  row.names = FALSE
)

knitr::kable(
  spacy_info_table,
  col.names = c("Pipeline field", "Value"),
  caption = "Local spaCy pipeline used for the comparison table",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(sentences$sentence_id[1], "s001"),
  identical(sentences$text[1], "Paid 12-week training is provided."),
  identical(pipeline$model, "en_core_web_sm"),
  identical(spacy_info$version, "3.8.0"),
  identical(spacy_info$license, "MIT"),
  identical(spacy_info$spacy, "3.8.7"),
  identical(
    parser_score$measure,
    c(
      "held_out_tokens",
      "correct_head",
      "unlabelled_attachment",
      "correct_head_and_relation",
      "labelled_attachment"
    )
  ),
  identical(
    parser_score$value,
    c(4007, 2899, 0.7235, 2629, 0.6561)
  ),
  identical(nrow(held_out_tokens), stored_held_out_tokens),
  identical(
    computed_unlabelled,
    parser_score$value[parser_score$measure == "unlabelled_attachment"]
  ),
  identical(
    computed_labelled,
    parser_score$value[parser_score$measure == "labelled_attachment"]
  ),
  all(nchar(metadata$fingerprint) == 64L),
  identical(parser_score_fingerprint, expected_score_fingerprint),
  identical(parser_model_fingerprint, expected_model_fingerprint)
)
```

The chunk checks the score file against the held-out token count, recomputes both
ratios, and matches the parser and score fingerprints in the metadata. The
parser attaches 2,899 of 4,007 held-out tokens to the right head word, about
0.72, and gets both head and relation right for 2,629 tokens, about 0.66.

## Parse fixed tokens

The parser receives one token per line because the model has no tokenizer. The
vertical text block hands udpipe each sentence as a stack of fixed tokens.

```{r}
parser <- udpipe_load_model(parser_model_path)

vertical <- sentences$text |>
  tokenize_words(lowercase = FALSE, strip_punct = FALSE) |>
  map_chr(\(tokens) paste(tokens, collapse = "\n"))

parsed <- udpipe_annotate(
  parser,
  x = vertical,
  doc_id = sentences$sentence_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "default"
) |>
  as.data.frame() |>
  as_tibble()

paid_parse <- parsed |>
  filter(doc_id == "s001") |>
  mutate(
    head_word = if_else(
      head_token_id == "0",
      "root",
      token[match(head_token_id, token_id)]
    )
  ) |>
  select(token, head_word, dep_rel)

knitr::kable(
  paid_parse,
  col.names = c("Token", "Head word", "Relation"),
  caption = "Dependency parse for 'Paid 12-week training is provided.'",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(parsed), 178L),
  identical(
    paid_parse$token,
    c("Paid", "12", "-", "week", "training", "is", "provided", ".")
  ),
  identical(
    paid_parse$head_word,
    c("week", "Paid", "Paid", "training", "provided", "provided", "root", "provided")
  ),
  identical(
    paid_parse$dep_rel,
    c("compound", "flat", "punct", "compound", "nsubj:pass", "aux:pass", "root", "punct")
  )
)
```

The parse says `provided` is the **root**, the central word of this sentence.
`training` attaches to `provided` as a passive subject, and `is` attaches as a
passive helper verb. Inside `Paid 12-week`, the parser is wrong. It makes `Paid`
part of `week` and hangs `12` off `Paid`. A reader would attach `Paid` to
`training` and `12` to `week`.

The parse answers half the team's question. `training` is the thing provided.
Nothing in the sentence says who provides it, because English lets a passive
drop the doer. There is no word for the parser to point at.

These tables use a few relation labels. `root` marks the central word.
`compound` links words inside a larger noun phrase. `flat` is for flat name-like
material, though it is a bad fit for `12` here. `nummod` marks a numeric
modifier, and `nmod` marks a nominal modifier. `nsubj:pass` marks a passive
subject, `aux:pass` marks a passive helper, `ccomp` marks a clausal complement,
and `punct` marks punctuation.

## Compare one sentence with spaCy

The same idea can come from a different tool. spaCy's English parser does not
use the Universal Dependencies label set. It uses another scheme, which is why
`nsubjpass` will not be found in the UD documentation linked below.

```{r}
spacy_s001 <- spacy_parse(
  c(s001 = "Paid 12-week training is provided."),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
) |>
  as_tibble()

spacy_paid_parse <- spacy_s001 |>
  mutate(
    head_word = if_else(
      head_token_id == token_id,
      "ROOT",
      token[match(as.character(head_token_id), as.character(token_id))]
    )
  ) |>
  select(token, head_word, dep_rel, lemma, pos)

knitr::kable(
  spacy_paid_parse,
  col.names = c("Token", "Head word", "spaCy relation", "Lemma", "POS tag"),
  caption = "spaCy parse for the same sentence",
  row.names = FALSE
)

invisible(suppressMessages(spacy_finalize()))
```

```{r}
#| include: false

stopifnot(
  identical(nrow(spacy_s001), 8L),
  identical(
    spacy_paid_parse$token,
    c("Paid", "12", "-", "week", "training", "is", "provided", ".")
  ),
  identical(
    spacy_paid_parse$head_word,
    c("training", "week", "week", "training", "provided", "provided", "ROOT", "provided")
  ),
  identical(
    spacy_paid_parse$dep_rel,
    c("nmod", "nummod", "punct", "compound", "nsubjpass", "auxpass", "ROOT", "punct")
  ),
  identical(spacy_paid_parse$lemma[7], "provide")
)
```

The two tools produce different structures, not a spelling difference in the
labels. spaCy attaches `Paid` to `training` and `12` to `week`; UDPipe attaches
`Paid` to `week` and `12` to `Paid`. Both outputs still have tokens, heads, and
labels, but they disagree about this sentence.

## Read morphology in context

The parser output also includes the `feats` column. These are **morphological
features**, compact labels for grammar carried by a token in this sentence.

```{r}
paid_features <- parsed |>
  filter(doc_id == "s001", token %in% c("is", "provided")) |>
  select(token, feats)

knitr::kable(
  paid_features,
  col.names = c("Token", "Morphological features"),
  caption = "In-context features for two tokens in sentence s001",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(paid_features$token, c("is", "provided")),
  identical(
    paid_features$feats,
    c(
      "Mood=Ind|Number=Sing|Person=3|Tense=Pres|VerbForm=Fin",
      "Tense=Past|VerbForm=Part|Voice=Pass"
    )
  )
)
```

The feature string for `is` says it is a finite present-tense helper form. The
feature string for `provided` says it is a past participle in passive voice. That
is why the relation labels include passive forms.

## Compare parsing with tagging

A parser score needs its own baseline. The chunk below uses a deliberately weak
one: attach each token to the token before it, with the first token in a sentence
attached to the root.

```{r}
tagger_curve <- read_csv(
  "data/treebank/tagger-learning-curve.csv",
  na = character(),
  col_types = cols(
    training_sentences = col_integer(),
    training_tokens = col_integer(),
    held_out_tokens = col_integer(),
    correct_tokens = col_integer(),
    accuracy = col_double()
  )
)

tagging_score <- tagger_curve |>
  filter(training_sentences == 500L) |>
  transmute(
    system = "500-sentence tagger",
    correct = correct_tokens,
    tokens = held_out_tokens,
    accuracy
  )

attachment_baseline <- held_out_tokens |>
  group_by(sentence_index) |>
  mutate(previous_head = if_else(row_number() == 1L, "0", lag(token_id))) |>
  ungroup() |>
  summarise(
    system = "Previous-token baseline, head only",
    correct = sum(head_token_id == previous_head),
    tokens = n(),
    attachment = round(correct / tokens, 4),
    .groups = "drop"
  )

attachment_scores <- bind_rows(
  attachment_baseline,
  tibble(
    system = c(
      "500-sentence parser, head only",
      "500-sentence parser, head and relation"
    ),
    correct = c(stored_correct_head, stored_correct_head_and_relation),
    tokens = c(stored_held_out_tokens, stored_held_out_tokens),
    attachment = c(computed_unlabelled, computed_labelled)
  )
)

knitr::kable(
  tagging_score,
  col.names = c("System", "Correct tokens", "Tokens scored", "Accuracy"),
  caption = "Tagging accuracy on the held-out weblog tokens",
  row.names = FALSE
)

knitr::kable(
  attachment_scores,
  col.names = c("System", "Correct tokens", "Tokens scored", "Attachment score"),
  caption = "Parser attachment scores and a previous-token baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    tagger_curve$accuracy[tagger_curve$training_sentences == 500L],
    0.9087
  ),
  identical(tagging_score$correct, 3641L),
  identical(attachment_scores$correct, c(304L, 2899L, 2629L)),
  identical(attachment_scores$tokens, c(4007L, 4007L, 4007L)),
  identical(attachment_scores$attachment, c(0.0759, 0.7235, 0.6561))
)
```

Attaching every token to the word before it gets 304 of 4,007 heads right. The
trained parser gets 2,899 heads right, about 0.72. The tagger's 0.91 belongs to
a different task with a different set of choices, so the table keeps it separate.
The head-and-relation row is stricter by construction: it can be right only when
the head is right. These numbers come from one train-and-test split of thirteen
weblog posts, and tokens inside a post are related because vocabulary, topic, and
author repeat.

## Check Riverton roots

The held-out score is a warning, not a local audit. Because the Riverton set has
only 28 sentences, the lesson can compare each parser root with a hand-chosen
root for the sentence.

```{r}
parser_roots <- parsed |>
  filter(head_token_id == "0") |>
  transmute(sentence_id = doc_id, parser_root = token)

hand_roots <- tribble(
  ~sentence_id, ~hand_root,
  "s001", "provided",
  "s002", "required",
  "s003", "available",
  "s004", "need",
  "s005", "required",
  "s006", "pays",
  "s007", "part",
  "s008", "able",
  "s009", "required",
  "s010", "preferred",
  "s011", "available",
  "s012", "receives",
  "s013", "preferred",
  "s014", "uses",
  "s015", "provided",
  "s016", "apprenticeship",
  "s017", "required",
  "s018", "outdoors",
  "s019", "required",
  "s020", "required",
  "s021", "included",
  "s022", "skill",
  "s023", "HOUSE",
  "s024", "CERTIFICATE",
  "s025", "stipend",
  "s026", "classes",
  "s027", "required",
  "s028", "Apply"
)

root_audit <- hand_roots |>
  left_join(parser_roots, by = "sentence_id") |>
  mutate(root_matches = parser_root == hand_root)

root_summary <- root_audit |>
  summarise(
    right_root = sum(root_matches),
    wrong_root = sum(!root_matches),
    total = n(),
    .groups = "drop"
  )

root_mismatches <- root_audit |>
  filter(!root_matches)

knitr::kable(
  root_mismatches,
  col.names = c("Sentence ID", "Hand root", "Parser root", "Root matches"),
  caption = "Riverton sentences where the parser root needs review",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(root_audit), nrow(sentences)),
  all(root_audit$sentence_id %in% sentences$sentence_id),
  identical(root_summary$right_root, 21L),
  identical(root_summary$wrong_root, 7L),
  identical(
    root_mismatches$sentence_id,
    c("s004", "s017", "s021", "s023", "s024", "s026", "s028")
  )
)
```

Of the 28 Riverton sentences, the parser picks the hand-marked root in 21 and
misses 7. The misses cluster in noun-heavy lines and flyer headings, where there
is little sentence structure for the parser to hold on to.

```{r}
included_parse <- parsed |>
  filter(doc_id == "s021") |>
  mutate(
    head_word = if_else(
      head_token_id == "0",
      "root",
      token[match(head_token_id, token_id)]
    )
  ) |>
  select(token, head_word, dep_rel)

knitr::kable(
  included_parse,
  col.names = c("Token", "Head word", "Relation"),
  caption = "A Riverton parse that needs human review",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(included_parse$token, c("Product", "training", "is", "included", ".")),
  identical(included_parse$head_word, c("root", "included", "included", "Product", "Product")),
  identical(included_parse$dep_rel, c("root", "nsubj:pass", "aux:pass", "ccomp", "punct"))
)
```

In `Product training is included`, the main claim is that training is included.
The parser has made `Product` the head of the sentence and hung `included`
beneath it as a subordinate clause, so the parser has the sentence inside out.

## What to remember

- A dependency parse links each token to a head word and names the relation.
- UDPipe and spaCy disagree on the heads inside `Paid 12-week`.
- A previous-token baseline gets 304 of 4,007 weblog heads right.
- The trained parser gets 2,899 of 4,007 heads right, about 0.72.
- The parser misses 7 of 28 hand-marked Riverton roots.
- Parser links on job-posting text need direct inspection.

Use dependency links as leads for reading, especially around passives. On this
corpus the useful warning is concrete: seven root choices need review before any
downstream count trusts them.

## Sources

- [UDPipe R package reference](https://bnosac.github.io/udpipe/docs/doc0.html)
- [Universal Dependencies relations](https://universaldependencies.org/u/dep/)
- [Universal Dependencies English Web Treebank](https://universaldependencies.org/treebanks/en_ewt/index.html)
