---
title: "Using a language model in a reproducible analysis"
subtitle: "Treat the model as an instrument: record it, check it against people, and correct for its errors"
description: "Learn how to use a local language model to label text inside a research workflow, keep the run reproducible, check it on a random sample reviewed by people, and correct a downstream estimate for the model's mistakes."
categories: [guide, language models, reproducibility, evaluation]
---

Ruth has the 240 records from the Calder Yard review in lesson 74 and a report
to write. A small language model could read every record in an afternoon. Her
report does not need the model to be impressive. It needs one number she can
defend: what share of the collection is responsive, and how sure she can be.

This page treats the model the way a careful researcher treats any measuring
instrument. The instrument is written down before it is used, its readings are
saved, a random sample is checked by people, and the final estimate is
corrected for the errors that sample reveals. Nothing here is specific to one
model. The same steps apply to a hosted model, a local one, or a trained
classifier.

::: {.callout-note}
The records are the constructed review collection from lesson 74. They are fictional teaching records, and their reference labels were written by the page author when the collection was built. Here those labels play the part of a careful human review.
:::

::: {.callout-tip}
## What you will learn

This page shows how to:

- decide when a language model is a reasonable instrument for a text task;
- record a model run so another person can repeat it;
- treat saved outputs as the record and reruns as a check;
- compare model labels with a random sample reviewed by people;
- correct an estimate for the model's errors with that sample; and
- find R tools that support this way of working.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Decide whether a language model is the right instrument

Language models can now annotate, classify, extract, and summarize text well
enough to be useful in research. On some annotation tasks they match or beat
crowd workers at a small fraction of the cost. Careful replications also find
them inconsistent: the same model can do well on one dataset and poorly on the
next, and small changes to a prompt's wording or format can move accuracy by
many points.

Three findings shape the advice on this page. First, a model follows a
written codebook, with definitions and edge cases, more reliably than a bare
category name, and the codebook matters more than clever phrasing. Second, with
a few hundred labelled examples for a stable task, a fine-tuned small model
often matches or beats prompting a large one, at lower cost and with more
stable output. Third, labels are inputs to an analysis rather than findings.
Even labels that agree with people most of the time can bias a downstream
estimate, because their errors are rarely random.

The earlier lessons still apply. A model does not replace corpus design
(lesson 4), a defined unit of analysis, or held-out evaluation (lesson 44). It
replaces one coder, and it needs checking like one.

## Record the instrument before using it

An **instrument record** lists everything that decides what the model says.
Here that means the model and its exact revision, the prompt, and the decoding
settings. The prompt lives in a file, so it has a fingerprint like any other
data file. It was written from the review request alone, before the model saw
any record, and was not revised after reading the model's answers.

```{r}
library(dplyr)
library(ggplot2)
library(jsonlite)
library(purrr)
library(readr)
library(stringr)
library(tibble)
library(tidyr)

source("R/use-nlg.R")
source("R/lesson-figures.R")

hash_lines <- function(path) {
  digest::digest(
    paste(read_lines(path), collapse = "\n"),
    algo = "sha256",
    serialize = FALSE
  )
}

collection_path <- "data/riverton/riverton-review-collection.csv"
prompt_path <- "data/riverton/riverton-llm-review-prompt.json"

records <- read_csv(
  collection_path,
  na = c("", "NA"),
  col_types = cols(
    .default = col_character(),
    reference_responsive = col_logical()
  )
) |>
  select(doc_id, review_request, text, reference_responsive)

prompt <- read_json(prompt_path)
model_record <- nlg_model_manifest() |>
  filter(model_key == "qwen_1_5b_instruct")

cat(prompt$system, "", prompt$user_template, sep = "\n")
```

```{r}
#| echo: false

tibble(
  item = c(
    "model", "revision", "license", "prompt ID", "prompt SHA-256",
    "decoding", "tokens allowed per answer"
  ),
  value = c(
    model_record$model_id,
    model_record$revision,
    model_record$license,
    prompt$prompt_id,
    hash_lines(prompt_path),
    "greedy: the most likely next token at every step",
    "3"
  )
) |>
  knitr::kable(
    col.names = c("Instrument", "Value"),
    caption = "Instrument record for the review labels",
    row.names = FALSE
  )
```

```{r}
#| include: false

stopifnot(
  identical(nrow(records), 240L),
  identical(sum(records$reference_responsive), 36L),
  identical(n_distinct(records$review_request), 1L),
  identical(nrow(model_record), 1L),
  identical(model_record$model_id, "Qwen/Qwen2.5-1.5B-Instruct"),
  str_detect(prompt$user_template, fixed("{request}")),
  str_detect(prompt$user_template, fixed("{record}"))
)
```

The model is an open-weight model pinned to one revision, and every runtime file
listed in the project's model manifest is checked against a recorded SHA-256
before it loads. A hosted model can change behind the same name or be retired on
the provider's schedule, so a hosted run needs the dated model version and a
saved copy of every answer.

## Label every record once and keep the outputs

The builder script `data-raw/build-llm-review-labels.R` sent this prompt, once
per record, to the pinned model and saved every raw answer with the run's
metadata. Rendering this page reads that saved run; it does not relabel the
collection. The saved file is the record of what the model said. A rerun later
is a check on that record, not a replacement for it.

Before the saved labels are used, the page checks that they still belong to
this instrument. If someone edits the prompt, swaps the model, changes the
decoding settings, or changes the collection, the check fails and the page stops
instead of reporting labels from a different run.

```{r}
labels_path <- "data/riverton/riverton-llm-review-labels.csv"

cached_labels <- read_csv(
  labels_path,
  na = character(),
  col_types = cols(
    doc_id = col_character(),
    raw_output = col_character(),
    llm_label = col_character(),
    output_tokens = col_integer(),
    ended_by_eos = col_logical()
  )
)

run_record <- read_csv(
  "data/riverton/riverton-llm-review-labels-metadata.csv",
  na = character(),
  col_types = cols(.default = col_character())
)

decoding_settings <- paste(
  "greedy; do_sample = FALSE; num_beams = 1; max_new_tokens = 3;",
  "repetition_penalty = 1"
)

parse_answer <- function(raw_output) {
  first_word <- str_extract(str_to_lower(raw_output), "[a-z]+")
  case_when(
    first_word == "yes" ~ "responsive",
    first_word == "no" ~ "not responsive",
    TRUE ~ "unparsed"
  )
}

cache_checks <- tibble(
  check = c(
    "labels file matches its recorded fingerprint",
    "prompt file and prompt ID match the run",
    "review collection matches the one the run labelled",
    "model ID and revision match the pinned manifest",
    "decoding settings match this page",
    "every saved label matches its raw answer"
  ),
  passed = c(
    hash_lines(labels_path) == run_record$fingerprint,
    hash_lines(prompt_path) == run_record$prompt_sha256 &&
      prompt$prompt_id == run_record$prompt_id,
    hash_lines(collection_path) == run_record$collection_sha256,
    run_record$model_id == model_record$model_id &&
      run_record$revision == model_record$revision,
    run_record$decoding == decoding_settings,
    all(parse_answer(cached_labels$raw_output) == cached_labels$llm_label)
  )
)
```

```{r}
#| echo: false

knitr::kable(
  cache_checks,
  col.names = c("Check before using the saved labels", "Passed"),
  caption = "The saved run is checked against the instrument record",
  row.names = FALSE
)

run_record |>
  transmute(created_on, decoding, platform, seconds_per_record) |>
  pivot_longer(everything(), names_to = "field", values_to = "value") |>
  knitr::kable(
    col.names = c("Run detail", "Value"),
    caption = "Details recorded by the builder when it labelled the collection",
    row.names = FALSE
  )
```

```{r}
#| include: false

stopifnot(
  all(cache_checks$passed),
  identical(nrow(run_record), 1L),
  identical(cached_labels$doc_id, records$doc_id),
  all(cached_labels$llm_label %in% c("responsive", "not responsive", "unparsed"))
)

label_counts <- cached_labels |>
  count(llm_label, name = "records")
count_for <- function(label) {
  value <- label_counts$records[label_counts$llm_label == label]
  if (length(value) == 0L) 0L else value
}

unparsed_sentence <- if (count_for("unparsed") == 0L) {
  "Every answer began with one of the two allowed words."
} else {
  paste(count_for("unparsed"), "answers began with neither word.")
}
```

The model answered `yes` for `r count_for("responsive")` records and `no` for
`r count_for("not responsive")`. `r unparsed_sentence` The page decided in
advance how to treat any other answer: it counts as not responsive, because the
model did not say the record was responsive.

## Rerun a few records to test the record

Greedy decoding picks the single most likely token at every step, so the same
software on the same machine usually repeats itself. Across machines, library
versions, or server batch sizes, small numerical differences can change a
choice, and "temperature 0" on a hosted service is not a guarantee. The check
below reruns eight records chosen with seed 925, the date the prompt was
written, and compares the new answers with the saved ones.

```{r}
fill_prompt <- function(request, record) {
  prompt$user_template |>
    str_replace(fixed("{request}"), request) |>
    str_replace(fixed("{record}"), record)
}

model <- load_nlg_pipeline("qwen_1_5b_instruct", "text-generation")

ask_model <- function(request, record) {
  chat_prompt <- nlg_chat_prompt(
    model$tokenizer,
    prompt$system,
    fill_prompt(request, record)
  )
  nlg_generate(model, chat_prompt, max_new_tokens = 3L)$text
}

set.seed(925)
rerun <- records |>
  slice_sample(n = 8) |>
  mutate(rerun_output = map2_chr(review_request, text, ask_model)) |>
  left_join(cached_labels |> select(doc_id, raw_output), by = "doc_id") |>
  mutate(identical_answer = rerun_output == raw_output)
```

```{r}
#| echo: false

knitr::kable(
  rerun |>
    select(doc_id, text, raw_output, rerun_output, identical_answer),
  format = "html",
  escape = TRUE,
  col.names = c(
    "Record", "Text", "Saved answer", "Answer in this render", "Identical"
  ),
  caption = "Eight saved answers rerun during this render",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(rerun), 8L),
  !anyDuplicated(rerun$doc_id),
  all(nzchar(rerun$rerun_output)),
  !anyNA(rerun$identical_answer)
)
```

In this render, `r sum(rerun$identical_answer)` of 8 rerun answers match the
saved answers exactly. The page does not require all eight to match, because a
render on a different computer may differ. It reports what happened, and the
estimates below use the saved answers either way.

## Check the labels against people on a random sample

A person now reviews a **simple random sample** of 40 records, drawn so that
every record has the same chance of selection. The seed, 2509, was fixed before
the first draw, and the sample is drawn once. Random selection is what lets the
sample speak for the whole collection; a sample of hand-picked or easy records
would not.

```{r}
sample_size <- 40L

label_levels <- c("not responsive", "responsive")

set.seed(2509)
review_sample <- records |>
  slice_sample(n = sample_size) |>
  left_join(cached_labels |> select(doc_id, llm_label), by = "doc_id") |>
  mutate(
    person_says = factor(
      if_else(reference_responsive, "responsive", "not responsive"),
      levels = label_levels
    ),
    model_says = factor(
      if_else(llm_label == "responsive", "responsive", "not responsive"),
      levels = label_levels
    )
  )

agreement <- review_sample |>
  count(person_says, model_says, name = "records", .drop = FALSE)
```

```{r}
#| echo: false

agreement |>
  pivot_wider(names_from = model_says, values_from = records) |>
  knitr::kable(
    col.names = c(
      "Person's reading", "Model says not responsive", "Model says responsive"
    ),
    caption = "Model labels compared with the person's review on the random sample",
    row.names = FALSE
  )
```

```{r}
#| include: false

sample_agreements <- sum(review_sample$person_says == review_sample$model_says)
count_cell <- function(person, model) {
  agreement$records[agreement$person_says == person & agreement$model_says == model]
}
missed <- count_cell("responsive", "not responsive")
false_flags <- count_cell("not responsive", "responsive")

error_sentence <- if (false_flags == 0L && missed > 0L) {
  paste0(
    "Here every disagreement runs the same way: the model missed ", missed,
    " responsive records and flagged no record the person called not responsive, ",
    "so a count of its labels will run low."
  )
} else {
  paste0(
    "Here the model missed ", missed, " responsive records and flagged ",
    false_flags, " records the person called not responsive."
  )
}

stopifnot(
  identical(nrow(review_sample), sample_size),
  !anyDuplicated(review_sample$doc_id),
  !anyNA(review_sample$model_says),
  identical(nrow(agreement), 4L),
  identical(sum(agreement$records), sample_size),
  identical(sample_agreements + missed + false_flags, sample_size)
)
```

The model and the person agree on `r sample_agreements` of the
`r sample_size` sampled records. Agreement alone hides which way the errors
run. `r error_sentence`

## Estimate the responsive share without trusting the model blindly

Three estimates of the responsive share are possible:

- **Model only:** the share of all 240 records the model called responsive.
  It uses every record but inherits every error the model makes.
- **Sample only:** the share of the 40 sampled records the person called
  responsive. It is unbiased but uncertain, because 40 records is a small
  sample.
- **Model plus correction:** the model-only share, plus the average
  difference between the person's reading and the model's label on the
  sample. The sample estimates the model's average error, and the correction
  subtracts that estimate.

The third estimate is the **difference estimator** from survey sampling. In
machine learning the same idea is called **prediction-powered inference**. Over
repeated random samples it is unbiased, and its precision depends on how much
the difference between the person's reading and the model's label varies from
record to record. When the model agrees with people on most records, the
interval is narrower than the sample alone would give. Each interval below is the estimate plus or minus 1.96
standard errors, with the finite-population correction for sampling 40 of 240
records.

```{r}
all_records <- nrow(records)

model_flags <- cached_labels |>
  transmute(doc_id, model_flag = as.integer(llm_label == "responsive"))

sample_values <- review_sample |>
  transmute(doc_id, person_flag = as.integer(reference_responsive)) |>
  left_join(model_flags, by = "doc_id") |>
  mutate(difference = person_flag - model_flag)

finite_population_se <- function(values, population_size) {
  sqrt((1 - length(values) / population_size) * var(values) / length(values))
}

model_only <- mean(model_flags$model_flag)
sample_only <- mean(sample_values$person_flag)
corrected <- model_only + mean(sample_values$difference)

estimates <- tibble(
  method = c("model only", "sample only", "model plus correction"),
  estimate = c(model_only, sample_only, corrected),
  standard_error = c(
    0,
    finite_population_se(sample_values$person_flag, all_records),
    finite_population_se(sample_values$difference, all_records)
  )
) |>
  mutate(
    low = pmax(0, estimate - 1.96 * standard_error),
    high = pmin(1, estimate + 1.96 * standard_error)
  )

true_share <- mean(records$reference_responsive)
```

```{r}
#| echo: false

estimates |>
  mutate(
    across(c(estimate, low, high), \(value) scales::percent(value, accuracy = 0.1)),
    interval = if_else(
      method == "model only",
      "none: no sampling, but the model's bias remains",
      str_c(low, " to ", high)
    ),
    uses = c(
      "model labels for all 240 records",
      "the person's review of 40 records",
      "model labels for all 240 records and the review of 40"
    )
  ) |>
  select(method, estimate, interval, uses) |>
  knitr::kable(
    col.names = c("Method", "Estimated responsive share", "95% interval", "Uses"),
    caption = "Three estimates of the responsive share of the collection",
    row.names = FALSE
  )
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sample_values), sample_size),
  !anyNA(sample_values$difference),
  all(sample_values$difference %in% c(-1L, 0L, 1L)),
  identical(true_share, 36 / 240),
  all(estimates$low <= estimates$estimate),
  all(estimates$high >= estimates$estimate)
)

percent_text <- function(value) scales::percent(value, accuracy = 0.1)
estimate_for <- function(name, column = "estimate") {
  estimates[[column]][estimates$method == name]
}
covers_truth <- function(name) {
  estimate_for(name, "low") <= true_share && true_share <= estimate_for(name, "high")
}

estimate_alt_text <- paste0(
  "Dot and interval chart of three estimates of the responsive share. ",
  "Model only: ", percent_text(model_only), ", with no interval. ",
  "Sample only: ", percent_text(sample_only), ", 95 percent interval ",
  percent_text(estimate_for("sample only", "low")), " to ",
  percent_text(estimate_for("sample only", "high")), ". ",
  "Model plus correction: ", percent_text(corrected), ", interval ",
  percent_text(estimate_for("model plus correction", "low")), " to ",
  percent_text(estimate_for("model plus correction", "high")), ". ",
  "A dashed line marks the true share of ", percent_text(true_share), "."
)

estimate_title <- "The model alone undercounts; a checked sample of 40 corrects it"

stopifnot(
  model_only < true_share - 0.05,
  covers_truth("model plus correction"),
  abs(corrected - true_share) < abs(model_only - true_share)
)
```

Because this collection is constructed, the true share is known:
`r scales::percent(true_share, accuracy = 0.1)`. A real project would not have
this number. It is shown only so the three methods can be judged.

```{r}
#| label: fig-llm-estimates
#| fig-cap: "Three estimates of the responsive share, with 95% intervals, against the true share known only because the collection is constructed."
#| fig-alt: !expr estimate_alt_text
#| fig-height: 3.6
#| code-fold: true
#| code-summary: "Show the plotting code"

estimate_plot <- estimates |>
  mutate(method = factor(method, levels = rev(estimates$method)))

ggplot(estimate_plot, aes(x = estimate, y = method)) +
  geom_vline(
    xintercept = true_share,
    linetype = "22",
    colour = lesson_colours[["muted"]]
  ) +
  geom_linerange(
    aes(xmin = low, xmax = high),
    linewidth = 1.1,
    colour = lesson_colours[["accent"]]
  ) +
  geom_point(size = 3.2, colour = lesson_colours[["accent"]]) +
  annotate(
    "text",
    x = true_share,
    y = 3.45,
    label = str_c("true share ", scales::percent(true_share, accuracy = 0.1)),
    hjust = -0.05,
    family = lesson_font,
    size = 3.4,
    colour = lesson_colours[["muted"]]
  ) +
  scale_x_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    title = estimate_title,
    subtitle = "Responsive share of 240 records; lines are 95% intervals",
    x = NULL,
    y = NULL
  ) +
  theme_lesson(grid = "x")
```

```{r}
#| include: false

stopifnot(
  identical(levels(estimate_plot$method), rev(estimates$method)),
  all(estimate_plot$high <= 1)
)
```

## Repeat the sample to see why the correction works

One sample could be lucky. Drawing the 40-record sample 2,000 more times, with
no new model calls, shows how each method behaves in the long run. For each
draw, the code repeats the two sample-based estimates and asks whether each
interval contains the true share.

```{r}
set.seed(2510)
repeated <- map(seq_len(2000L), \(draw) {
  sampled <- sample(all_records, sample_size)
  person <- as.integer(records$reference_responsive[sampled])
  difference <- person - model_flags$model_flag[sampled]

  tibble(
    draw = draw,
    method = c("sample only", "model plus correction"),
    estimate = c(mean(person), model_only + mean(difference)),
    standard_error = c(
      finite_population_se(person, all_records),
      finite_population_se(difference, all_records)
    )
  )
}) |>
  list_rbind() |>
  mutate(
    covers_truth = abs(estimate - true_share) <= 1.96 * standard_error
  )

long_run <- repeated |>
  summarise(
    average_estimate = mean(estimate),
    average_interval_width = mean(2 * 1.96 * standard_error),
    coverage = mean(covers_truth),
    .by = method
  )
```

```{r}
#| echo: false

long_run |>
  mutate(
    across(c(average_estimate, coverage), \(value) scales::percent(value, accuracy = 0.1)),
    average_interval_width = sprintf("%.1f points", 100 * average_interval_width)
  ) |>
  knitr::kable(
    col.names = c(
      "Method", "Average estimate", "Average interval width",
      "Intervals containing the true share"
    ),
    caption = "Two sample-based estimators over 2,000 repeated samples of 40 records",
    row.names = FALSE
  )
```

```{r}
#| include: false

run_value <- function(name, column) {
  long_run[[column]][long_run$method == name]
}
points_text <- function(value) sprintf("%.1f percentage points", 100 * value)

corrected_width <- run_value("model plus correction", "average_interval_width")
sample_width <- run_value("sample only", "average_interval_width")
corrected_coverage <- run_value("model plus correction", "coverage")
sample_coverage <- run_value("sample only", "coverage")
zero_width_draws <- sum(
  repeated$standard_error[repeated$method == "model plus correction"] == 0
)

long_run_sentence <- paste0(
  "Both methods average close to the true share. The corrected intervals are ",
  "narrower, ", points_text(corrected_width), " wide on average against ",
  points_text(sample_width), " for the sample alone, because the model's labels ",
  "settle most records and the person only has to measure its misses. They ",
  "contain the true share in ", percent_text(corrected_coverage), " of samples, ",
  "a little short of the nominal 95%, against ",
  percent_text(sample_coverage), " for the sample alone. The shortfall comes ",
  "from the simple normal-approximation interval, which is rough when the ",
  "model's misses are rare in a sample of 40. In ", zero_width_draws,
  " of the 2,000 samples none of the misses were drawn, so the corrected ",
  "interval had no width at all. A larger sample, or an interval built for ",
  "rare events, would close the gap."
)

stopifnot(
  identical(nrow(repeated), 4000L),
  identical(long_run$method, c("sample only", "model plus correction")),
  all(abs(long_run$average_estimate - true_share) < 0.02),
  corrected_width < sample_width,
  corrected_coverage < 0.95,
  corrected_coverage > 0.85,
  sample_coverage > corrected_coverage,
  zero_width_draws > 0L
)
```

`r long_run_sentence`

## A checklist for model-assisted analysis

- Choose the instrument on purpose. Prefer a rule, a dictionary, or a small
  trained model when it does the job, and use a language model when the task
  needs reading that those cannot do.
- Write the codebook and prompt before the first run, store them as a file,
  and record the file's fingerprint.
- Pin the model. Record the model name, exact revision or dated version,
  decoding settings, and software versions. Open weights that you can run
  yourself make this easier, though they do not remove every source of
  variation.
- Save every raw output with the ID of the input that produced it.
- Rerun a sample and report how many answers repeat.
- Check the labels against people on a random sample, and report both kinds
  of error.
- Correct any estimate that uses model labels, with the difference estimator
  or a related method such as prediction-powered inference or design-based
  supervised learning.
- Test a model used as a judge the same way before trusting its scores.
- Check structured output for content as well as format. A reply that matches
  a schema can still hold wrong values.
- Read and rerun code that an assistant wrote, keep a reference calculation
  for key numbers, and pin the environment.
- Disclose the model's role using a reporting checklist such as GUIDE-LLM or
  TRIPOD-LLM.

## R tools for this way of working

The packages below were checked on 25 September 2026. Versions change often,
so check again before relying on one.

| Package | What it does | Runs with local models |
|---|---|---|
| ellmer 0.5.0 | Chat with model providers, call tools, and extract structured data | Yes, through Ollama |
| mall 0.2.0 | Classify, extract, or score a data frame column row by row | Yes, designed for Ollama |
| ragnar 0.3.1 | Chunk documents, store embeddings, and retrieve passages | Partly; document conversion uses Python |
| vitals 0.4.0 | Evaluate model-based tools, including model-graded scoring | Yes, with any ellmer chat |
| tidyllm 0.6.0 | One tidy interface to several providers | Yes, through Ollama |
| ipd 0.4.1 | Inference on predicted data, including prediction-powered inference | Not needed; it uses saved labels |
| huggingfaceR 2.1.0 | Hugging Face models through Python, used on this site | Yes, fully offline once models are saved |

: R packages for model-assisted text analysis, checked 25 September 2026 {.striped}

The reference implementation of design-based supervised learning is the `dsl`
package on GitHub, which is not on CRAN.

## What to remember

- A language model is an instrument; write down the model, revision, prompt,
  and settings before using it.
- The saved outputs are the reproducible record; a rerun tests that record.
- A random sample reviewed by people shows how the model errs.
- Correct estimates with that sample; model-only numbers carry the model's bias.
- The earlier lessons' rules still apply: provenance, stable IDs, baselines,
  and held-out evaluation.

Ruth's report gives the corrected share with its interval, the sample it rests
on, and the instrument record, so a reader can check each step.

## Sources

- [Gilardi, Alizadeh, and Kubli (2023), ChatGPT outperforms crowd workers for text-annotation tasks](https://doi.org/10.1073/pnas.2305016120)
- [Kristensen-McLachlan et al. (2025), Are chatbots reliable text annotators? Sometimes](https://doi.org/10.1093/pnasnexus/pgaf069)
- [Halterman and Keith, Codebook LLMs: evaluating LLMs as measurement tools for political science concepts](https://arxiv.org/abs/2407.10747)
- [Törnberg (2024), Best practices for text annotation with large language models](https://doi.org/10.6092/issn.1971-8853/19461)
- [Edwards and Camacho-Collados (2024), Language models for text classification: is in-context learning enough?](https://aclanthology.org/2024.lrec-main.879/)
- [Sclar et al. (2024), Quantifying language models' sensitivity to spurious features in prompt design](https://arxiv.org/abs/2310.11324)
- [Angelopoulos et al. (2023), Prediction-powered inference](https://doi.org/10.1126/science.adi6000)
- [Egami et al. (2023), Using imperfect surrogates for downstream inference](https://arxiv.org/abs/2306.04746)
- [Salerno et al. (2025), ipd: an R package for conducting inference on predicted data](https://doi.org/10.1093/bioinformatics/btaf055)
- [Atil et al. (2025), Non-determinism of "deterministic" LLM system settings in hosted environments](https://aclanthology.org/2025.eval4nlp-1.12/)
- [Thinking Machines Lab (2025), Defeating nondeterminism in LLM inference](https://thinkingmachines.ai/blog/defeating-nondeterminism-in-llm-inference/)
- [Chen, Zaharia, and Zou (2024), How is ChatGPT's behavior changing over time?](https://doi.org/10.1162/99608f92.5317da47)
- [Palmer, Smith, and Spirling (2024), Using proprietary language models in academic research requires explicit justification](https://doi.org/10.1038/s43588-023-00585-1)
- [Bavaresco et al. (2025), LLMs instead of human judges? A large-scale empirical study across 20 NLP evaluation tasks](https://aclanthology.org/2025.acl-short.20/)
- [Tam et al. (2024), Let me speak freely? A study on the impact of format restrictions on LLM performance](https://aclanthology.org/2024.emnlp-industry.91/)
- [Bridgeford et al. (2026), Twelve quick tips for AI-assisted coding in science](https://doi.org/10.1371/journal.pcbi.1014428)
- [Feuerriegel et al. (2026), A reporting checklist for large language models in behavioural science](https://doi.org/10.1038/s41562-026-02492-7)
- [Gallifant et al. (2025), The TRIPOD-LLM reporting guideline](https://doi.org/10.1038/s41591-024-03425-5)
- [ellmer documentation](https://ellmer.tidyverse.org/)
- [mall documentation](https://mlverse.github.io/mall/)
- [ragnar documentation](https://ragnar.tidyverse.org/)
- [vitals documentation](https://vitals.tidyverse.org/)
