---
title: "Deduplicating documents"
subtitle: "Find repeated records before they distort a collection"
author: "Alex Farach"
description: "Learn how exact, normalized, and near-duplicate checks behave on Riverton workforce documents."
categories: [documents, deduplication, workforce research]
---

Before training and evaluation, one repeated record can make a collection look
larger than it is. The coordinator sees the same training sentence more than
once and has to decide whether it represents a repeated record or a separate
notice.

That choice affects every count that follows. A duplicate document can inflate a
common phrase, make a rare label look common, and leak between a training split
and an evaluation split.

This lesson works at the document level. A **duplicate** is a record that should
not be counted as a separate document for the question being asked.

Lesson 12 builds augmented examples with `source_family_id`; those rows need the
same split-protection idea. This page uses a smaller constructed set so the
distance mechanism is easy to inspect.

::: {.callout-note}
Riverton is a teaching case; the documents and duplicate families below are
fictional.
:::

::: {.callout-tip}
## What you will learn

After this lesson, you should be able to:

- explain why duplicate documents change downstream counts;
- find exact duplicates with a SHA-256 hash;
- find formatting-only duplicates with a normalized hash;
- compare near duplicates with character trigram distance; and
- choose a threshold by naming the error it makes cheaper.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build a small review set

The review table starts with the sentence file, then adds table shaping,
repeated calculations, text normalization, fingerprints, distance measures,
pairs, and a threshold plot.

A **hash** is a fixed-length fingerprint made from text. The same input gives the
same hash. A changed input should give a different hash, so a hash checks exact
identity rather than similarity.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(digest)
library(stringdist)
library(tidyr)
library(ggplot2)

workforce_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

source_text <- workforce_sentences |>
  filter(sentence_id == "s006") |>
  pull(text)

unrelated_text <- workforce_sentences |>
  filter(sentence_id == "s004") |>
  pull(text)

documents <- tibble(
  doc_id = c("D01", "D02", "D03", "D04", "D05", "D06"),
  source = c(
    "Riverton sentence s006",
    "Exact author copy",
    "Formatting author copy",
    "One-word author variant",
    "Near distractor",
    "Unrelated Riverton sentence s004"
  ),
  text = c(
    source_text,
    source_text,
    "  THE employer   pays for certification   TRAINING. ",
    "The employer covers certification training.",
    "The employer pays for certification exam.",
    unrelated_text
  ),
  duplicate_family = c(
    "certification-training",
    "certification-training",
    "certification-training",
    "certification-training",
    "certification-exam",
    "spreadsheet-skills"
  )
)

stopifnot(
  identical(nrow(workforce_sentences), 28L),
  identical(source_text, "The employer pays for certification training."),
  identical(unrelated_text, "Applicants need basic spreadsheet skills."),
  identical(nrow(documents), 6L),
  identical(sum(documents$duplicate_family == "certification-training"), 4L)
)

knitr::kable(
  documents |>
    select(doc_id, source, text),
  col.names = c("Document", "Source", "Text"),
  caption = "Six documents prepared for duplicate review",
  row.names = FALSE
)
```

This six-row set was written to expose a known failure mode. Four rows are
hand-marked as the same duplicate family: the source sentence, an exact copy, a
spacing-and-case copy, and a one-word paraphrase. The exam sentence changes the
claim while leaving much of the surface text intact. The spreadsheet sentence is
unrelated.

## Find exact duplicates

An **exact duplicate** has the same text byte for byte. The code hashes each text
value with SHA-256. In the code, `\(value)` is R shorthand for "for each value,
run the small calculation that follows."

```{r}
documents <- documents |>
  mutate(
    exact_hash = map_chr(
      text,
      \(value) digest::digest(
        value,
        algo = "sha256",
        serialize = FALSE
      )
    )
  )

exact_groups <- documents |>
  count(exact_hash, name = "documents") |>
  filter(documents > 1L)

exact_duplicate_pairs <- sum(choose(exact_groups$documents, 2))

stopifnot(
  identical(n_distinct(documents$exact_hash), 5L),
  identical(nrow(exact_groups), 1L),
  identical(exact_groups$documents, 2L),
  identical(exact_duplicate_pairs, 1)
)

knitr::kable(
  documents |>
    transmute(
      doc_id,
      text,
      hash_start = str_sub(exact_hash, 1, 12)
    ),
  col.names = c("Document", "Text", "SHA-256 start"),
  caption = "Exact hashes before normalization",
  row.names = FALSE
)
```

Only the source and the exact copy match, giving one exact duplicate pair. The
formatting copy receives a different fingerprint because capital letters and
extra spaces are real characters.

## Normalize before hashing

**Normalization** means applying a recorded rule before comparing text. This
rule is narrow: lowercase the text and replace repeated whitespace with one
space. It catches formatting-only differences and destroys information about
capitalization and layout.

```{r}
documents <- documents |>
  mutate(
    normalized_text = str_squish(str_to_lower(text)),
    normalized_hash = map_chr(
      normalized_text,
      \(value) digest::digest(
        value,
        algo = "sha256",
        serialize = FALSE
      )
    )
  )

normalized_groups <- documents |>
  count(normalized_hash, name = "documents") |>
  filter(documents > 1L)

normalized_duplicate_pairs <- sum(choose(normalized_groups$documents, 2))

stopifnot(
  identical(n_distinct(documents$normalized_hash), 4L),
  identical(nrow(normalized_groups), 1L),
  identical(normalized_groups$documents, 3L),
  identical(normalized_duplicate_pairs, 3),
  identical(
    documents$normalized_text[1:3],
    rep("the employer pays for certification training.", 3L)
  )
)

knitr::kable(
  documents |>
    transmute(
      doc_id,
      normalized_text,
      hash_start = str_sub(normalized_hash, 1, 12)
    ),
  col.names = c("Document", "Normalized text", "Normalized SHA-256 start"),
  caption = "Normalized hashes after lowercasing and whitespace squishing",
  row.names = FALSE
)
```

The normalized hash joins three rows, which make three duplicate pairs. That is
useful here, but risky for a flyer where line breaks or all-capital headings
carry meaning.

## Measure near duplicates

A **near duplicate** is close enough to deserve review, even when the text is not
identical. Cut each normalized string into every overlapping three-character
piece, discard order and repeats, and the result is a set of character trigrams.
Jaccard distance is the share of pieces the two sets do not have in common: 0
when the sets match, 1 when they share nothing. Because order is discarded, the
measure counts how much surface text differs, not which change matters.

```{r}
comparison_pairs <- expand_grid(
  left_doc = documents$doc_id,
  right_doc = documents$doc_id
) |>
  filter(left_doc < right_doc) |>
  left_join(
    documents |>
      select(
        left_doc = doc_id,
        left_text = text,
        left_norm = normalized_text,
        left_family = duplicate_family
      ),
    by = join_by(left_doc)
  ) |>
  left_join(
    documents |>
      select(
        right_doc = doc_id,
        right_text = text,
        right_norm = normalized_text,
        right_family = duplicate_family
      ),
    by = join_by(right_doc)
  ) |>
  mutate(
    manual_duplicate =
      left_family == right_family &
        left_family == "certification-training",
    jaccard_distance = map2_dbl(
      left_norm,
      right_norm,
      \(left, right) stringdist::stringdist(
        left,
        right,
        method = "jaccard",
        q = 3
      )
    )
  )

calibration_pair <- tibble(
  left_text = "Evening shifts require a valid forklift certification.",
  right_text = "Evening shifts require a valid forklift certificate."
) |>
  mutate(
    jaccard_distance = stringdist::stringdist(
      left_text,
      right_text,
      method = "jaccard",
      q = 3
    )
  )

calibration_distance <- calibration_pair$jaccard_distance

stopifnot(
  identical(nrow(comparison_pairs), 15L),
  identical(nrow(calibration_pair), 1L),
  identical(sum(comparison_pairs$manual_duplicate), 6L),
  identical(round(calibration_distance, 4), 0.1132),
  identical(
    comparison_pairs |>
      filter(left_doc == "D01", right_doc == "D04") |>
      pull(jaccard_distance) |>
      round(2),
    0.32
  ),
  identical(
    comparison_pairs |>
      filter(left_doc == "D01", right_doc == "D05") |>
      pull(jaccard_distance) |>
      round(4),
    0.2917
  )
)

knitr::kable(
  comparison_pairs |>
    transmute(
      pair = str_c(left_doc, right_doc, sep = " / "),
      manual_duplicate,
      jaccard_distance = round(jaccard_distance, 4)
    ),
  col.names = c("Pair", "Hand-marked duplicate", "Jaccard distance"),
  caption = "Pairwise distances across the six review documents",
  row.names = FALSE
)

knitr::kable(
  calibration_pair |>
    mutate(jaccard_distance = round(jaccard_distance, 4)),
  col.names = c("Left text", "Right text", "Jaccard distance"),
  caption = "Calibration pair for reading the Jaccard scale",
  row.names = FALSE
)
```

The calibration strings are not part of the review set; they show how to read the
scale. Two sentences that differ only in `certification` versus `certificate`
score 0.1132. In this constructed review set, the exam distractor scores 0.2917
against the source, while the meaning-preserving paraphrase scores 0.32. The
distractor changes one short word and the duplicate changes one longer phrase, so
surface overlap and meaning point in opposite directions. Real collections can
contain pairs like this; the small set is designed so the mechanism is visible.

## Sweep the threshold

A **threshold** is the distance at or below which the pipeline joins two
documents. The threshold is not learned here. The team tries several values and
counts both kinds of outcome.

```{r}
#| label: fig-dedup-threshold
#| fig-cap: "Duplicate pairs caught and distinct pairs wrongly joined at each tested threshold."
#| fig-alt: "Step chart showing true duplicate pairs caught at 3 until threshold 0.35 and 6 afterward, while wrongly joined distinct pairs move from 0 to 3, then 4, then 9."

threshold_sweep <- tibble(
  threshold = c(0, 0.2, 0.3, 0.35, 0.7, 0.99)
) |>
  mutate(
    caught_true = map_int(
      threshold,
      \(cutoff) sum(
        comparison_pairs$manual_duplicate &
          comparison_pairs$jaccard_distance <= cutoff
      )
    ),
    wrongly_joined = map_int(
      threshold,
      \(cutoff) sum(
        !comparison_pairs$manual_duplicate &
          comparison_pairs$jaccard_distance <= cutoff
      )
    )
  )

perfect_threshold <- threshold_sweep |>
  filter(caught_true == 6L, wrongly_joined == 0L)

stopifnot(
  identical(threshold_sweep$caught_true, c(3L, 3L, 3L, 6L, 6L, 6L)),
  identical(threshold_sweep$wrongly_joined, c(0L, 0L, 3L, 3L, 4L, 9L)),
  identical(nrow(perfect_threshold), 0L)
)

knitr::kable(
  threshold_sweep,
  col.names = c(
    "Threshold",
    "True duplicate pairs caught",
    "Distinct pairs wrongly joined"
  ),
  caption = "Near-duplicate results at six threshold values",
  row.names = FALSE
)

threshold_sweep |>
  tidyr::pivot_longer(
    cols = c(caught_true, wrongly_joined),
    names_to = "result",
    values_to = "pairs"
  ) |>
  mutate(
    result = case_when(
      result == "caught_true" ~ "True duplicate pairs caught",
      result == "wrongly_joined" ~ "Distinct pairs wrongly joined",
      TRUE ~ result
    )
  ) |>
  ggplot(aes(x = threshold, y = pairs, color = result)) +
  geom_step(linewidth = 0.8, direction = "hv") +
  geom_point(size = 2) +
  scale_x_continuous(breaks = threshold_sweep$threshold) +
  labs(
    x = "Jaccard distance threshold",
    y = "Pairs",
    color = NULL
  ) +
  theme_minimal()
```

For this constructed set, 0.2917 is smaller than 0.32. The distractor is closer
to the source than the paraphrase is. Because those two distances are ordered
that way, no threshold on these pairs can catch the paraphrase without also
joining the distractor. The sweep confirms that threshold choice trades one error
against another here; it is not a general law of deduplication.

## Deduplication at scale

The `expand_grid()` step used here builds an all-pairs combination, which mathematically scales at $O(N^2)$. Pairwise comparisons work for a tiny teaching set, but mathematically collapse in production. For real-world datasets, an all-pairs Jaccard distance calculation is too slow and requires too much memory.

Production systems bypass this $O(N^2)$ scaling boundary by using Locality Sensitive Hashing (LSH) and MinHash over shingled text. R packages like `textreuse` (or Python's `datasketch`) implement these algorithms to group near-duplicates across millions of documents without comparing every pair.

## What to remember

- Duplicate documents can inflate counts and leak between training and evaluation.
- Exact hashes catch only byte-for-byte repeats.
- Normalized hashes catch some formatting differences and erase formatting facts.
- Character-trigram Jaccard distance measures surface overlap, not preserved meaning.
- Near-duplicate thresholds trade missed duplicates against wrongly joined records in this set.
- The right threshold depends on which error costs more for the project.

Exact repeats can be removed automatically; near matches belong in review before
any training or evaluation split. The threshold sweep is a warning label on this
constructed set, not a universal setting for Riverton documents.

## Sources

- [digest package documentation](https://eddelbuettel.github.io/digest/)
- [stringdist package documentation](https://cran.r-project.org/package=stringdist)
- [The stringdist package for approximate string matching](https://journal.r-project.org/archive/2014/RJ-2014-011/index.html)
