---
title: "Extracting document metadata"
subtitle: "Describe a source before analyzing its text"
author: "Alex Farach"
description: "Learn how to extract HTML metadata and compute provenance fields for a Riverton source file."
categories: [documents, metadata, workforce research]
---

A saved job-board page can answer questions the sentence table cannot. The
coordinator can read the sentence text, but a later reviewer will ask a different
question: where did this document come from?

The answer depends on **metadata**, the description of a document rather than the
main content. Metadata lets a collection be filtered, cited, checked, and
audited.

Lesson 6 pulled sentences out of this page and recorded a fingerprint for the
file. This lesson looks at what the page says about itself: its title, declared
character set, headings, and absent links. Then it separates those claims from
facts computed from the copy on disk.

::: {.callout-note}
The job board page is a fictional fixture written for metadata practice.
:::

::: {.callout-tip}
## What you will learn

Work through the examples so you can:

- explain what document metadata records;
- extract a title, meta elements, headings, and link counts from HTML;
- assemble one row per document; and
- separate author claims from computed provenance facts.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Read the document description

The metadata pass reads source lines and recorded fields, builds document
records, repeats attribute work, checks strings, computes fingerprints, and
extracts HTML fields. A **provenance** field records where a document came from
and how this copy can be checked.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(digest)
library(rvest)

job_page_path <- "data/workforce/job-board.html"
job_page_lines <- read_lines(job_page_path)
job_page <- rvest::read_html(job_page_path)

page_title <- job_page |>
  rvest::html_element("title") |>
  rvest::html_text2()

meta_nodes <- job_page |>
  rvest::html_elements("meta")

meta_table <- tibble(
  element = "meta",
  attributes = map_chr(
    rvest::html_attrs(meta_nodes),
    \(attributes) str_c(names(attributes), attributes, sep = "=", collapse = "; ")
  )
)

stopifnot(
  identical(length(job_page_lines), 82L),
  identical(page_title, "Riverton workforce opportunities"),
  identical(nrow(meta_table), 1L),
  identical(meta_table$attributes, "charset=utf-8")
)

knitr::kable(
  bind_rows(
    tibble(element = "title", attributes = page_title),
    meta_table
  ),
  col.names = c("HTML element", "Recorded value"),
  caption = "Title and meta elements claimed inside the HTML document",
  row.names = FALSE
)
```

The title and `charset` value are claims written inside the file. They are useful
metadata, but they come from the document itself.

## Extract headings and links

Headings summarize the structure around the job text. Link targets show where a
page points. This fixture points nowhere: it was written without `<a>` elements,
so the link count below is zero. A real job board would usually carry apply
links; a zero count should prompt a check of the file.

```{r}
h1_text <- job_page |>
  rvest::html_elements("h1") |>
  rvest::html_text2()

h2_text <- job_page |>
  rvest::html_elements("h2") |>
  rvest::html_text2()

heading_table <- bind_rows(
  tibble(level = "h1", text = h1_text),
  tibble(level = "h2", text = h2_text)
)

heading_summary <- heading_table |>
  group_by(level) |>
  summarise(
    headings = n(),
    example = first(text),
    .groups = "drop"
  )

link_targets <- job_page |>
  rvest::html_elements("a") |>
  rvest::html_attr("href")

link_markup_count <- sum(str_detect(job_page_lines, "<a(\\s|>)"))

stopifnot(
  identical(nrow(heading_table), 7L),
  identical(nrow(heading_summary), 2L),
  identical(sum(heading_table$level == "h1"), 1L),
  identical(sum(heading_table$level == "h2"), 6L),
  identical(heading_table$text[2], "Data analyst trainee"),
  identical(length(link_targets), 0L),
  identical(link_markup_count, 0L)
)

knitr::kable(
  heading_summary,
  col.names = c("Level", "Headings", "Example"),
  caption = "Heading counts extracted from the job board page",
  row.names = FALSE
)
```

The seven headings describe the page shape without reprinting the scraped
sentence table. The zero link count is also metadata: this fixture has no
outbound targets to preserve.

## Make one row per document

A collection table needs one record per document. Long lists can be collapsed for
review while the detailed extraction stays available in code.

```{r}
document_metadata <- tibble(
  document_id = "job-board-html",
  title = page_title,
  meta_count = nrow(meta_table),
  h1_count = sum(heading_table$level == "h1"),
  h2_count = sum(heading_table$level == "h2"),
  link_count = length(link_targets),
  link_status = if_else(
    length(link_targets) == 0L,
    "no anchor elements in fixture",
    str_c(link_targets, collapse = " | ")
  )
)

stopifnot(
  identical(nrow(document_metadata), 1L),
  identical(document_metadata$meta_count, 1L),
  identical(document_metadata$h1_count, 1L),
  identical(document_metadata$h2_count, 6L),
  identical(document_metadata$link_count, 0L),
  identical(document_metadata$link_status, "no anchor elements in fixture")
)

knitr::kable(
  document_metadata,
  col.names = c(
    "Document ID",
    "Title",
    "Meta elements",
    "H1 headings",
    "H2 headings",
    "Links",
    "Link status"
  ),
  caption = "One-row metadata record for the saved job board page",
  row.names = FALSE
)
```

The row gives the coordinator a document-level handle for filtering and audit. It
keeps the zero-link result visible rather than hiding an empty vector in code.

## Add provenance fields

HTML self-description is not enough. A collection also needs fields supplied by
the collector and fields computed from the stored copy. The fingerprint here is
SHA-256 over the file text joined with `\n` line endings.

```{r}
recorded_page_metadata <- read_csv(
  "data/workforce/job-board-metadata.csv",
  na = character(),
  col_types = cols(
    source_file = col_character(),
    created_for_lesson = col_character(),
    retrieval_status = col_character(),
    records = col_character(),
    md5 = col_character(),
    rights_note = col_character()
  )
)

# Byte size depends on how the file's line endings were stored, so a size
# read straight from disk changes between systems. Measuring the same
# newline-joined text the fingerprint uses gives a number that describes the
# content rather than the copy.
source_lines <- read_lines(job_page_path)
source_text <- paste(source_lines, collapse = "\n")
source_size <- nchar(source_text, type = "bytes")
content_fingerprint <- digest::digest(
  source_text,
  algo = "sha256",
  serialize = FALSE
)

provenance <- document_metadata |>
  mutate(
    source_file = job_page_path,
    recorded_creation_date = recorded_page_metadata$created_for_lesson,
    retrieval_status = recorded_page_metadata$retrieval_status,
    size_bytes = source_size,
    sha256 = content_fingerprint,
    .after = document_id
  )

stopifnot(
  identical(nrow(recorded_page_metadata), 1L),
  identical(recorded_page_metadata$created_for_lesson, "2026-08-28"),
  identical(recorded_page_metadata$retrieval_status, "synthetic archived page"),
  identical(source_size, 3441L),
  identical(
    content_fingerprint,
    "ce1c515cfda7658ff586b4f42dd1d71c805041a6c32bc88c97b54a20f0fd5eba"
  ),
  identical(str_length(content_fingerprint), 64L),
  identical(str_length(recorded_page_metadata$md5), 32L)
)

knitr::kable(
  provenance |>
    select(
      document_id,
      source_file,
      recorded_creation_date,
      retrieval_status,
      size_bytes,
      sha256,
      title
    ),
  col.names = c(
    "Document ID",
    "Source file",
    "Recorded creation date",
    "Retrieval status",
    "Bytes",
    "SHA-256 fingerprint",
    "Author-claimed title"
  ),
  caption = "Two recorded claims and two computed facts joined to the document record",
  row.names = FALSE
)
```

The title, charset, creation date, and retrieval status are assertions someone
wrote down. Only the size and SHA-256 fingerprint were computed from the copy on
disk. The SHA-256 hash follows the course rule: read the text lines, join them
with `\n`, and hash that string. `tools::md5sum()` hashes raw file bytes, so line
endings would be part of the answer.

## What to remember

- Metadata describes a document rather than its main content.
- HTML can claim its own title, character set, headings, and links.
- A collection needs recorded provenance fields and computed file checks.
- Hash text files with a recorded line-ending rule and SHA-256.
- Author claims and computed fingerprints answer different audit questions.

For this page, the title and creation date are claims. The byte count, zero
links, and SHA-256 are checks on the saved fixture. Those checks help audit the
file; they do not make the fixture complete or externally sourced.

## Sources

- [rvest overview](https://rvest.tidyverse.org/)
- [HTML Living Standard metadata content](https://html.spec.whatwg.org/multipage/dom.html#metadata-content)
- [digest package documentation](https://eddelbuettel.github.io/digest/)
