---
title: "Identifying document language"
subtitle: "Treat language guesses as fallible classifications"
author: "Alex Farach"
description: "Learn how two language detectors behave on short Riverton text samples and why abstention can be safer than a wrong route."
categories: [documents, language identification, workforce research]
---

Short notes in several languages create a routing problem before content labels
begin. The coordinator wants to route each note to the right review queue before
anyone labels its content.

`Open house.` exposes the problem. A person reads the two-word notice as English,
while a language detector may have too little evidence to say.

**Language identification** is a classification problem. A classifier assigns a
label from a fixed set, such as `en` for English or `es` for Spanish. The risk
has two shapes: a detector can return the wrong code, or it can decline to answer
and leave the item unrouted. This lesson shows which shape these two detectors
take on a tiny teaching file.

::: {.callout-note}
The Riverton inbox and the language samples are fictional teaching materials.
:::

::: {.callout-tip}
## What you will learn

Use this lesson to practice how to:

- run two local language detectors on fixed samples;
- compare detector output with recorded labels;
- separate agreement from correctness;
- explain why short texts are hard to classify; and
- describe why returning `NA` can be safer than a wrong language code.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the labelled samples

The sample table is small, but the setup still records detector versions,
checks text length, runs the local CLD2 and CLD3 detectors, and fingerprints the
file. The samples are short strings written by the author for teaching.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)
library(cld2)
library(cld3)
library(digest)

language_path <- "data/riverton/language-samples.csv"
language_samples <- read_csv(
  language_path,
  na = character(),
  col_types = cols(
    sample_id = col_character(),
    written_in = col_character(),
    text = col_character()
  )
)

reference_metadata <- read_csv(
  "data/riverton/riverton-reference-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    fingerprint = col_character()
  )
)

language_reference <- reference_metadata |>
  filter(artifact == "language-samples.csv")

sample_fingerprint <- digest::digest(
  paste(read_lines(language_path), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)

recorded_fingerprint <- reference_metadata |>
  filter(artifact == "language-samples.csv") |>
  pull(fingerprint)

detector_versions <- tibble(
  detector = c("cld2", "cld3"),
  package_version = c(
    as.character(packageVersion("cld2")),
    as.character(packageVersion("cld3"))
  )
)

display_samples <- language_samples |>
  mutate(
    display_text = str_c(
      "<span lang=\"",
      written_in,
      "\">",
      text,
      "</span>"
    )
  )

knitr::kable(
  detector_versions,
  col.names = c("Detector package", "Version"),
  caption = "Detector package versions for this run",
  row.names = FALSE
)

knitr::kable(
  tibble(
    file = language_path,
    sha256 = sample_fingerprint
  ),
  col.names = c("File", "SHA-256 fingerprint"),
  caption = "Fingerprint for the author-written language sample file",
  row.names = FALSE
)

knitr::kable(
  display_samples |>
    select(sample_id, written_in, display_text),
  col.names = c("Sample", "Recorded language", "Text"),
  caption = "Eight labelled language samples used for detection",
  row.names = FALSE,
  escape = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(language_samples), 8L),
  identical(
    names(language_samples),
    c("sample_id", "written_in", "text")
  ),
  identical(language_samples$text[7], "Open house."),
  identical(nrow(language_reference), 1L),
  identical(sample_fingerprint, recorded_fingerprint),
  identical(
    sample_fingerprint,
    "fa84ffaf608578b8dd0e4f0dbc7425049cd9d362e9d41a1773078973afeccf67"
  ),
  all(str_detect(detector_versions$package_version, "^\\d+\\.\\d+"))
)

stopifnot(
  !any(str_detect(language_samples$text, "[<>&]")),
  all(str_detect(display_samples$display_text, fixed("<span lang=")))
)
```

The `written_in` column records the language the author meant to write, not a
verified annotation. The Spanish [Casa abierta.]{lang="es"} is grammatical, but
as an event notice it is a calque rather than an idiomatic phrase. These results come from
the detector versions printed above; a package change can return different codes
for the same strings.

## Run two detectors

CLD2 and CLD3 are trained language detectors. `NA` means the detector returned no
language label. They do not understand the whole situation around a note; they
classify the characters they receive.

```{r}
detected_samples <- language_samples |>
  mutate(
    characters = str_length(text),
    words = str_count(text, boundary("word")),
    length_group = if_else(characters <= 13L, "short", "long"),
    cld2_language = cld2::detect_language(text),
    cld3_language = cld3::detect_language(text),
    cld2_match = !is.na(cld2_language) & cld2_language == written_in,
    cld3_match = !is.na(cld3_language) & cld3_language == written_in,
    display_text = str_c(
      "<span lang=\"",
      written_in,
      "\">",
      text,
      "</span>"
    )
  )

open_house <- detected_samples |>
  filter(text == "Open house.")

knitr::kable(
  detected_samples |>
    select(
      sample_id,
      written_in,
      display_text,
      characters,
      words,
      cld2_language,
      cld3_language
    ),
  col.names = c(
    "Sample",
    "Recorded language",
    "Text",
    "Characters",
    "Words",
    "CLD2",
    "CLD3"
  ),
  caption = "Language detector output for each sample",
  row.names = FALSE,
  escape = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    detected_samples$cld2_language,
    c("en", "en", "es", "fr", "de", NA, NA, NA)
  ),
  identical(
    detected_samples$cld3_language,
    c("en", "en", "es", "fr", "de", NA, NA, NA)
  ),
  identical(open_house$written_in, "en"),
  is.na(open_house$cld2_language),
  is.na(open_house$cld3_language),
  identical(open_house$characters, 11L),
  identical(open_house$words, 2L)
)
```

For this installed version, both detectors abstain on `Open house.` by returning
`NA`. That is a useful failure for routing, because the item can be sent to human
review rather than silently routed by a wrong code.

## Count matches, abstentions, and label differences

Agreement between two detectors does not prove correctness. They can share
training data patterns and make the same mistake. The recorded labels let the
team count what happened on these eight examples.

```{r}
detector_counts <- tibble(
  detector = c("CLD2", "CLD3"),
  matched_recorded_label = c(
    sum(detected_samples$cld2_match),
    sum(detected_samples$cld3_match)
  ),
  no_label = c(
    sum(is.na(detected_samples$cld2_language)),
    sum(is.na(detected_samples$cld3_language))
  )
) |>
  mutate(
    differed_from_recorded_label =
      nrow(detected_samples) - matched_recorded_label - no_label
  )

knitr::kable(
  detector_counts,
  col.names = c(
    "Detector",
    "Matched recorded label",
    "No label",
    "Different from recorded label"
  ),
  caption = "Detector outcomes across the eight labelled samples",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(detector_counts$matched_recorded_label, c(5L, 5L)),
  identical(detector_counts$no_label, c(3L, 3L)),
  identical(detector_counts$differed_from_recorded_label, c(0L, 0L)),
  identical(sum(detected_samples$cld2_match), 5L),
  identical(sum(detected_samples$cld3_match), 5L)
)
```

Each detector matches 5 of the 8 recorded labels, returns no label for 3 of the
8, and differs from the recorded label on 0 of the 8. Those fractions describe
this tiny teaching file. Eight samples measure nothing about detector accuracy on
a workforce archive.

## Compare short and longer text

Length changes the amount of evidence a detector sees, but this file is too
small to estimate a length effect. There is no principled cutoff here. The
13-character line separates the two shortest strings from the other 6 examples.

```{r}
length_summary <- detected_samples |>
  group_by(length_group) |>
  summarise(
    samples = n(),
    cld2_matches = sum(cld2_match),
    cld2_no_label = sum(is.na(cld2_language)),
    cld3_matches = sum(cld3_match),
    cld3_no_label = sum(is.na(cld3_language)),
    .groups = "drop"
  ) |>
  arrange(length_group)

knitr::kable(
  length_summary,
  col.names = c(
    "Length group",
    "Samples",
    "CLD2 matches",
    "CLD2 no label",
    "CLD3 matches",
    "CLD3 no label"
  ),
  caption = "Detector outcomes by sample length in this file",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length_summary$length_group, c("long", "short")),
  identical(length_summary$samples, c(6L, 2L)),
  identical(length_summary$cld2_matches, c(5L, 0L)),
  identical(length_summary$cld2_no_label, c(1L, 2L)),
  identical(length_summary$cld3_matches, c(5L, 0L)),
  identical(length_summary$cld3_no_label, c(1L, 2L))
)
```

The two shortest samples produce no matches because both detectors abstain.
Among the six other samples, each detector matches 5 recorded labels and abstains
on 1. The Dutch sentence is long enough to read by eye, but both detectors return
`NA`.

## A wrong code is worse than no code

The eight labelled samples produced no wrong codes. That does not mean wrong
codes are impossible. The next string was chosen because it produces one: it is a
single hand-picked case, not a rate, and it shows what a wrong code looks like.
It also sits just above the arbitrary 13-character cutoff.

```{r}
short_probe <- tibble(
  text = c("Open house.", "Evening class."),
  recorded_language = "en"
) |>
  mutate(
    characters = str_length(text),
    cld2_language = cld2::detect_language(text),
    cld3_language = cld3::detect_language(text),
    cld2_match = !is.na(cld2_language) & cld2_language == recorded_language,
    cld3_match = !is.na(cld3_language) & cld3_language == recorded_language
  )

knitr::kable(
  short_probe |>
    select(text, recorded_language, cld2_language, cld3_language),
  col.names = c("Text", "Recorded language", "CLD2", "CLD3"),
  caption = "Two short English probes and their detected languages",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(short_probe$characters, c(11L, 14L)),
  all(is.na(short_probe$cld2_language)),
  identical(short_probe$cld3_language, c(NA, "lb")),
  identical(short_probe$cld2_match, c(FALSE, FALSE)),
  identical(short_probe$cld3_match, c(FALSE, FALSE))
)
```

`Evening class.` is English in this teaching probe. CLD2 returns `NA`, while CLD3
returns `lb`, the ISO 639-1 code for Luxembourgish. For routing, the abstention
is safer: no queue is selected until someone checks the item.

These eight samples are all written in the Latin alphabet, so this page shows
nothing about other writing systems. That is a limit of the file, not a
reassurance.

## Language identification in production

While `cld2` and `cld3` illustrate the routing problem, models such as Meta's `fasttext` are common alternatives for fast language identification. When deploying any model, keep these production realities in mind:

- **Language scope and BCP 47**: Projects must explicitly state their supported language scope upfront. When recording a language, use a standard like BCP 47 tags (e.g., `en-US`) rather than informal names.
- **Per-language recall**: A model's overall accuracy hides its failures. Always measure recall *per language*, because models routinely fail on low-resource languages.
- **Code-switching**: The single-label `cld2::detect_language()` and
  `cld3::detect_language()` calls used here return one document label. Mixed
  language text needs a detector designed for spans or tokens, or a review
  route; single-label behavior should not be generalized to every model.
- **Probability vs. certainty**: A model might return a confidence score like `0.95`, but do not treat this as a calibrated probability. It is only the model's internal score, not a statistical guarantee.

## What to remember

- Language identification may return a code or no label.
- Agreement between detectors does not prove that a language code is correct.
- The eight-row file is an inspection exercise: both detectors declined to label
  the two shortest strings rather than guessing at them.
- Returning `NA` can be safer than sending text to the wrong language queue.

Items with `NA` results belong in review, and short-text language codes should
prompt checking rather than settle a queue. The hand-picked probe demonstrates a
possible wrong code; the eight-row file demonstrates safe abstention.

## Sources

- [cld2 R package documentation](https://docs.ropensci.org/cld2/)
- [cld3 R package documentation](https://docs.ropensci.org/cld3/)
- [Google CLD3 project](https://github.com/google/cld3)
- [ISO 639 language codes](https://www.loc.gov/standards/iso639-2/php/code_list.php)
