---
title: "Parsing prices from text"
subtitle: "Turn pay mentions into amounts with units"
author: "Alex Farach"
description: "Learn how simple R rules pull wages and stipends from text while keeping ranges, missing units, and no-number cases visible."
categories: [entity enrichment, price parsing, workforce research]
---

After labelling sentences from the job board and training flyer, the team faces
a new question from the coordinator: which notices say anything concrete about
pay?

A phrase such as `$18.50 per hour` looks easy. A phrase such as `competitive pay`
looks useful to a reader but gives a computer no amount to compare. The team
needs a rule that can extract a value and also say when the text did not contain
a usable price.

::: {.callout-note}
The Riverton job board and flyer in this lesson are fictional teaching materials.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain why a number alone is not a price;
- write `stringr` patterns for money-like text;
- separate amount, currency, and pay period;
- keep ranges and missing values visible; and
- compare a small rule set with this page's expected answers.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build examples with expected answers

The setup reads the Riverton CSV with readr, shapes tables with dplyr and
tibble, and asks stringr to match the money-like patterns. The examples are
author-created pay strings in the Riverton style. A **price** here means an
amount, a currency, and a unit such as hour, year, or stipend.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)

stopifnot(all(c("readr", "dplyr", "tibble", "stringr") %in% loadedNamespaces()))

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

pay_examples <- tibble(
  example_id = sprintf("pay-%02d", 1:13),
  text = c(
    "$18.50 per hour",
    "18.50/hr",
    "$38,000 a year",
    "$17-$19 an hour",
    "competitive pay",
    "$1,250 stipend",
    "$20 hourly",
    "Pay starts at $16 per hour",
    "$42,500 annually",
    "Stipend of 750 dollars",
    "Up to $21/hour",
    "18 dollars an hour",
    "$38000 a year"
  ),
  expected_currency = c(
    "USD", NA, "USD", "USD", NA, "USD",
    "USD", "USD", "USD", "USD", "USD", "USD",
    "USD"
  ),
  expected_unit = c(
    "hour", "hour", "year", "hour", NA, "stipend",
    "hour", "hour", "year", "stipend", "hour", "hour",
    "year"
  ),
  expected_lower = c(
    18.50, 18.50, 38000, 17, NA, 1250,
    20, 16, 42500, 750, 21, 18,
    38000
  ),
  expected_upper = c(
    NA, NA, NA, 19, NA, NA,
    NA, NA, NA, NA, NA, NA,
    NA
  )
)

stopifnot(
  identical(nrow(sentences), 28L),
  identical(sentences$text[sentences$sentence_id == "s025"], "Paid training stipend"),
  identical(nrow(pay_examples), 13L)
)

knitr::kable(
  pay_examples,
  col.names = c(
    "Example ID", "Text", "Expected currency", "Expected unit",
    "Expected lower amount", "Expected upper amount"
  ),
  caption = "Author-created price examples with expected answers",
  row.names = FALSE
)
```

The table contains 13 examples. One writes a five-digit amount without a comma.
That case matters because a partial match would look like a real answer.

## Pull out the visible pieces

The pattern below uses the ICU regular expression engine through `stringr`. A
**regular expression** is a compact search pattern. This one looks for an
optional dollar sign, a number that may contain commas or cents, and an optional
second number after a hyphen or the word `to`.

```{r}
parse_prices <- function(data) {
  number_pattern <- "(?:\\d{1,3}(?:,\\d{3})+|\\d+)(?:\\.\\d+)?"
  amount_pattern <- paste0(
    "(?i)(?:\\$\\s*)?", number_pattern,
    "(?:\\s*(?:-|to)\\s*(?:\\$\\s*)?", number_pattern, ")?"
  )

  data |>
    mutate(
      amount_text = str_extract(text, amount_pattern),
      currency = case_when(
        str_detect(text, fixed("$")) ~ "USD",
        str_detect(str_to_lower(text), "\\bdollars?\\b") ~ "USD",
        TRUE ~ NA_character_
      ),
      unit = case_when(
        str_detect(str_to_lower(text), "per hour|/hr|/hour|an hour|hourly") ~ "hour",
        str_detect(str_to_lower(text), "a year|annually|annual|year") ~ "year",
        str_detect(str_to_lower(text), "stipend") ~ "stipend",
        TRUE ~ NA_character_
      ),
      amount_clean = str_remove_all(amount_text, "[$,]"),
      amount_lower = as.numeric(str_extract(amount_clean, "\\d+(?:\\.\\d+)?")),
      amount_upper = as.numeric(
        str_match(amount_clean, "(?:-|to)\\s*(\\d+(?:\\.\\d+)?)")[, 2]
      ),
      parser_result = case_when(
        is.na(amount_lower) ~ "not a price",
        is.na(unit) ~ "number without unit",
        !is.na(amount_upper) ~ "range",
        TRUE ~ "single price"
      )
    )
}

price_results <- parse_prices(pay_examples)

stopifnot(
  identical(price_results$amount_lower[3], 38000),
  identical(price_results$amount_lower[13], 38000),
  identical(price_results$amount_upper[4], 19),
  identical(price_results$parser_result[5], "not a price")
)

knitr::kable(
  price_results |>
    select(example_id, text, amount_text, currency, unit, amount_lower, amount_upper, parser_result),
  col.names = c(
    "Example ID", "Text", "Matched amount", "Currency", "Unit",
    "Lower amount", "Upper amount", "Parser result"
  ),
  caption = "Parsed amount, currency, and unit",
  row.names = FALSE
)
```

The number and the unit come from different parts of the phrase. `18.50/hr`
contains an amount and a unit, but the rule marks the currency as missing. The
rule treats every dollar sign as US dollars, which would be wrong on a board that
mixed US, Canadian, and Australian postings. Units outside this small set, such
as monthly pay, would need another branch.

## Count the special cases

A parser is easier to trust when it reports what it could not settle. The code
below counts the rows that ask for special handling.

```{r}
case_counts <- tibble(
  case = c(
    "thousands separator",
    "range with two numbers",
    "missing currency",
    "no number",
    "plain four-plus digit number"
  ),
  examples = c(
    sum(str_detect(price_results$amount_text, fixed(",")), na.rm = TRUE),
    sum(!is.na(price_results$amount_upper)),
    sum(is.na(price_results$currency)),
    sum(is.na(price_results$amount_lower)),
    sum(
      str_detect(str_remove_all(price_results$amount_text, "[$,]"), "\\d{4,}") &
        !str_detect(price_results$amount_text, fixed(",")),
      na.rm = TRUE
    )
  )
)

stopifnot(
  identical(case_counts$examples, c(3L, 1L, 2L, 1L, 1L))
)

knitr::kable(
  case_counts,
  col.names = c("Case", "Examples"),
  caption = "Cases the parser keeps visible",
  row.names = FALSE
)
```

Three examples contain thousands separators. One is a range with two amounts.
Two lack a dollar sign or dollar word. One has no number. The categories
overlap, so this table is a checklist rather than a count of distinct rows.

## Score the parser against the marking

Because the parser and expected answers come from the same author, a high score
here does not estimate accuracy. Read it only as a check that the code behaves
as this page says it should.

```{r}
values_match <- function(observed, expected) {
  (is.na(observed) & is.na(expected)) |
    (!is.na(observed) & !is.na(expected) & observed == expected)
}

price_score <- price_results |>
  summarise(
    examples = n(),
    lower_amount_matches = sum(values_match(amount_lower, expected_lower)),
    upper_amount_matches = sum(values_match(amount_upper, expected_upper)),
    unit_matches = sum(values_match(unit, expected_unit)),
    currency_matches = sum(values_match(currency, expected_currency))
  )

stopifnot(
  identical(
    unname(unlist(price_score)),
    c(13L, 13L, 13L, 13L, 13L)
  )
)

knitr::kable(
  price_score,
  col.names = c(
    "Examples", "Lower amount matches", "Upper amount matches",
    "Unit matches", "Currency matches"
  ),
  caption = "Agreement with this page's expected answers",
  row.names = FALSE
)
```

The parser matches the expected lower amount, upper amount, unit, and currency on
all 13 examples. The comma-free case would have been dangerous without the wider
number pattern: `$38000` must not be cut to `$380`.

## Check the Riverton stipend line

The real Riverton teaching file contains a stipend phrase, but not a stipend
amount. The rule should return a miss rather than inventing one.

```{r}
riverton_pay_line <- sentences |>
  filter(sentence_id == "s025") |>
  transmute(
    example_id = sentence_id,
    text,
    expected_currency = NA_character_,
    expected_unit = "stipend",
    expected_lower = NA_real_,
    expected_upper = NA_real_
  )

riverton_price_result <- parse_prices(riverton_pay_line)

stopifnot(
  identical(riverton_pay_line$text, "Paid training stipend"),
  identical(riverton_price_result$unit, "stipend"),
  is.na(riverton_price_result$amount_lower),
  identical(riverton_price_result$parser_result, "not a price")
)

knitr::kable(
  riverton_price_result |>
    select(example_id, text, amount_lower, currency, unit, parser_result),
  col.names = c("Sentence ID", "Text", "Amount", "Currency", "Unit", "Parser result"),
  caption = "The Riverton stipend line has a unit but no amount",
  row.names = FALSE
)
```

The line tells a reader that some stipend exists. It does not say how much. A
number without a unit is not a price, and a unit without a number is not an
amount the coordinator can compare.

## The limits of custom rules

A custom regular expression is a quick way to find highly constrained formats, but it breaks quickly on varied real-world data.

```{r}
failure_examples <- tibble(
  text = c("thirty-eight thousand a year", "1.250,50 / hr", "$40/hr (CAD)")
)

parse_prices(failure_examples) |>
  select(text, amount_lower, currency) |>
  knitr::kable(caption = "Representative failures of the custom regex parser")
```

- **Value normalization**: The pattern misses `"thirty-eight thousand"` entirely because it only looks for digits.
- **Locale formats**: It reads the European format `1.250,50` incorrectly (amount_lower is NA) because it expects commas as thousands separators.
- **Ambiguous currencies**: It assumes `$` means USD, which is wrong for CAD.

In standard NLP workflows, relying entirely on custom regex for entity parsing is an anti-pattern. Production systems use specialized entity parsing libraries (such as Duckling) or Named Entity Recognition (like `MONEY` entities in spaCy) to handle these variations and extract structured, normalized values across locales.

## What to remember

- A usable price needs an amount, a currency, and a unit.
- Ranges should keep both numbers.
- A parser must not return a partial number when the text contains a longer one.
- These counts describe 13 teaching examples and one Riverton line.

The stipend line can be flagged as pay-related, but it cannot support a wage
comparison because it gives no amount.

## Sources

- [stringr regular expression reference](https://stringr.tidyverse.org/articles/regular-expressions.html)
- [ICU regular expression documentation](https://unicode-org.github.io/icu/userguide/strings/regexp.html)
- [R `as.numeric()` documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/numeric.html)
