---
title: "Geocoding place names"
subtitle: "Join text mentions to an offline gazetteer"
description: "Learn how a gazetteer turns place names into coordinates while preserving ambiguous and not-found results."
categories: [entity enrichment, geocoding, workforce research]
---

On the flyer transcript, `RIVERTON SKILLS OPEN HOUSE` looks ready for a map
until `Riverton` matches two places in the reference table. The coordinator
needs the tie to stay visible.

A map needs coordinates, not only a place name. The team can join a mention to a
small reference file, then keep the unresolved cases visible.

::: {.callout-note}
The places, lab, and flyer here belong to an invented Riverton setting.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define geocoding and gazetteers in plain language;
- join place mentions to latitude and longitude;
- detect duplicate place-name matches;
- return a miss when a place is absent; and
- explain why context matters in geocoding.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the offline reference files

Geocoding means turning a place name in text into coordinates: a latitude and
longitude a map can plot. CSV files supply the examples, dplyr and tibble handle
the tables, and digest checks the gazetteer fingerprint. A **gazetteer** is a
list of place names with location information attached. Gazetteers may be public,
open-licence, or commercial reference datasets. This one is invented and local so
the lesson can run offline.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(digest)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

gazetteer <- read_csv(
  "data/riverton/riverton-gazetteer.csv",
  na = character(),
  col_types = cols(
    place_name = col_character(),
    place_type = col_character(),
    latitude = col_double(),
    longitude = col_double(),
    parent = col_character()
  )
)

metadata <- read_csv(
  "data/riverton/riverton-reference-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    fingerprint = col_character()
  )
)

gazetteer_hash <- digest(
  paste(read_lines("data/riverton/riverton-gazetteer.csv"), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)
expected_hash <- metadata |>
  filter(artifact == "riverton-gazetteer.csv") |>
  pull(fingerprint)

knitr::kable(
  gazetteer,
  col.names = c("Place name", "Type", "Latitude", "Longitude", "Parent"),
  caption = "Invented Riverton gazetteer used offline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(c("readr", "dplyr", "tibble", "digest") %in% loadedNamespaces()))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(gazetteer), 9L),
  identical(sum(gazetteer$place_name == "Riverton"), 2L),
  identical(gazetteer_hash, expected_hash)
)
```

The reference table has nine rows. The duplicate `Riverton` rows are deliberate:
one sits in Marrow County, Calder, and one sits in Tidewater. The `parent` column
names the larger place that contains a row. The coordinates are arbitrary teaching
points, not the locations of real sites.

`Riverton Skills Centre` is typed as a building here. In the entity-linking
lesson, the same words name an organisation. That is not a data-cleaning lesson
to hide; the same string can denote a place and an organisation, which is the
ORG-versus-place ambiguity that named-entity lessons have to handle.

## Join mentions to the gazetteer

The team starts with five hand-identified place mentions. One comes from the
Riverton flyer text. Four are constructed checks written in the same setting so
we can see resolved, ambiguous, and not-found cases together.

```{r}
place_mentions <- bind_rows(
  sentences |>
    filter(sentence_id == "s023") |>
    transmute(
      mention_id = sentence_id,
      source = "Riverton text",
      text,
      mention = "Riverton"
    ),
  tibble(
    mention_id = c("place-02", "place-03", "place-04", "place-05"),
    source = "constructed check",
    text = c(
      "Evening classes at Riverton Skills Centre",
      "Applicants in Riverton can apply by October 15",
      "Weekend shifts serve Bellhaven",
      "The route reaches West Calder"
    ),
    mention = c("Riverton Skills Centre", "Riverton", "Bellhaven", "West Calder")
  )
)

geocoded_rows <- place_mentions |>
  left_join(
    gazetteer,
    by = join_by(mention == place_name),
    relationship = "many-to-many"
  )

knitr::kable(
  geocoded_rows,
  col.names = c(
    "Mention ID", "Source", "Text", "Mention", "Type",
    "Latitude", "Longitude", "Parent"
  ),
  caption = "Joining place mentions to the gazetteer",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(place_mentions), 5L),
  identical(nrow(geocoded_rows), 7L),
  identical(sum(is.na(geocoded_rows$latitude)), 1L)
)
```

The `relationship` argument tells dplyr that one mention may match several rows.
That fan-out is the ambiguity this lesson is about, so the code declares it. The
join returns seven rows from five mentions because `Riverton` matches twice.
`West Calder` receives no coordinates because it is not in the gazetteer.

## Name the unresolved states

A geocoder should not pretend that every mention is solved. The next table gives
each mention a status.

```{r}
mention_status <- geocoded_rows |>
  summarise(
    matches = sum(!is.na(latitude)),
    candidate_parents = paste(na.omit(parent), collapse = ", "),
    .by = c(mention_id, source, text, mention)
  ) |>
  mutate(
    candidate_parents = if_else(candidate_parents == "", "none", candidate_parents),
    status = case_when(
      matches == 0L ~ "not found",
      matches > 1L ~ "ambiguous",
      TRUE ~ "resolved"
    )
  )

status_counts <- tibble(status = c("resolved", "ambiguous", "not found")) |>
  left_join(
    mention_status |>
      count(status, name = "mentions"),
    by = join_by(status)
  ) |>
  mutate(mentions = coalesce(mentions, 0L))

knitr::kable(
  mention_status,
  col.names = c(
    "Mention ID", "Source", "Text", "Mention", "Matches",
    "Candidate parents", "Status"
  ),
  caption = "Each mention receives a geocoding status",
  row.names = FALSE
)

knitr::kable(
  status_counts,
  col.names = c("Status", "Mentions"),
  caption = "Resolved, ambiguous, and not-found mentions",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(status_counts$mentions, c(2L, 2L, 1L)),
  identical(sum(status_counts$mentions), 5L)
)
```

Two mentions resolve to one row. Two are ambiguous. One is not found. These
counts describe only the five mentions in this lesson.

## Add context when the name is not enough

The name `Riverton` alone cannot choose between the two rows. No county appears
in the flyer, so the context below is supplied by the analyst to show the
mechanism. In a pipeline that value would need to come from a document header, a
posting address, or a stated scope for the collection.

```{r}
riverton_candidates <- geocoded_rows |>
  filter(mention_id == "s023") |>
  select(mention, latitude, longitude, parent)

supplied_context <- tibble(
  mention = "Riverton",
  parent_context = "Marrow County"
)

with_parent_context <- riverton_candidates |>
  inner_join(supplied_context, by = join_by(mention)) |>
  filter(parent == parent_context)

knitr::kable(
  riverton_candidates,
  col.names = c("Mention", "Latitude", "Longitude", "Parent"),
  caption = "The name Riverton has two candidate locations",
  row.names = FALSE
)

knitr::kable(
  with_parent_context,
  col.names = c("Mention", "Latitude", "Longitude", "Parent", "Context supplied"),
  caption = "Parent context narrows the candidate list",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(riverton_candidates), 2L),
  identical(nrow(with_parent_context), 1L),
  identical(with_parent_context$latitude, 41.821)
)
```

The context row shows a decision rule, not evidence from the flyer. Without the
parent region, a prior belief, or surrounding text, the two `Riverton` rows remain
tied.

## Geocoding in production

The local exact-match join in this lesson shows the mechanism of geocoding, but production systems rely on specialized geocoding APIs (such as OpenStreetMap/Nominatim) or packages (like `tidygeocoder` in R or `geopy` in Python) rather than raw string joins.

When using a live service, several boundaries and limits apply:
- **Match precision**: APIs can perform fuzzy matching, returning a confidence score. A low score requires uncertain-result inspection.
- **Service boundaries**: Rate limits and Terms of Service often dictate how many requests you can make.
- **Caching**: Storing results locally prevents redundant network calls and speeds up repeated processing.
- **Privacy and re-identification**: Sending text containing personal locations to an external API can leak private data. Self-hosted geocoders (like a local Nominatim instance) keep sensitive text entirely on your own infrastructure.

## What to remember

- A gazetteer is a reference list of place names with locations attached.
- Geocoding turns a place mention into coordinates.
- Joining by name can return more than one coordinate pair.
- A miss should stay a miss rather than becoming a guess.
- Context such as a parent region can reduce the candidate list.
- `ambiguous` and `not found` are first-class geocoding answers.

This teaching gazetteer can map `Riverton Skills Centre` and `Bellhaven`. The
word `Riverton` still has two candidate locations, and `West Calder` has none.

## Sources

- [GeoNames geographical database](https://www.geonames.org/)
- [Natural Earth data](https://www.naturalearthdata.com/)
- [Getty Thesaurus of Geographic Names](https://www.getty.edu/research/tools/vocabularies/tgn/)
