---
title: "Parsing time expressions"
subtitle: "Resolve dates only when the reference date is explicit"
author: "Alex Farach"
description: "Learn how simple R rules handle dates, weekdays, times of day, and durations in workforce text."
categories: [entity enrichment, temporal parsing, workforce research]
---

Calendar work stalls at `Apply by October 15`: the coordinator can see the
month and day, but not the year. A deadline needs that missing context before it
can land on a calendar.

A date parser turns a time expression into a date a computer can sort. The hard
part is knowing when the text gives a full date, when it needs a reference date,
and when it is not a calendar date at all.

::: {.callout-note}
Riverton dates and notices in this lesson are made-up teaching examples.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- separate full dates from relative dates;
- use an explicit reference date rather than `Sys.Date()`;
- recognize durations and times of day as different kinds of time language;
- compare a small temporal parser with this page's expected answers; and
- compare the rule output with spaCy entity labels.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## State the reference date

A fixed reference date anchors the examples: 2026-08-28. That date is a Friday.
A page whose output changes each day cannot be checked, so the lesson passes the
date as a variable and never calls `Sys.Date()`.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(spacyr)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

reference_date <- as.Date("2026-08-28")

temporal_examples <- tibble(
  expr_id = sprintf("time-%02d", 1:11),
  text = c(
    sentences |>
      filter(sentence_id == "s028") |>
      pull(text),
    "Evening",
    "12-week",
    "Monday",
    "next Friday",
    "next Monday",
    "2026-10-15",
    "starts in the fall",
    "Classes begin on September 14",
    sentences |>
      filter(sentence_id == "s020") |>
      pull(text),
    sentences |>
      filter(sentence_id == "s011") |>
      pull(text)
  ),
  expected_category = c(
    "reference-date needed",
    "not resolvable",
    "not resolvable",
    "reference-date needed",
    "reference-date needed",
    "reference-date needed",
    "fully resolved",
    "not resolvable",
    "reference-date needed",
    "not resolvable",
    "not resolvable"
  ),
  expected_date = as.Date(c(
    "2026-10-15",
    NA,
    NA,
    "2026-08-31",
    "2026-09-04",
    "2026-09-07",
    "2026-10-15",
    NA,
    "2026-09-14",
    NA,
    NA
  ))
)

knitr::kable(
  temporal_examples,
  col.names = c("Expression ID", "Text", "Expected category", "Expected date"),
  caption = "Temporal examples with expected answers",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(c("readr", "dplyr", "tibble", "purrr", "stringr", "spacyr") %in% loadedNamespaces()))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(temporal_examples), 11L),
  identical(format(reference_date, "%A"), "Friday")
)
```

`October 15` needs the reference year. `Evening` is a time of day without a date.
`12-week` is a duration, not a point on the calendar.

## Write a small resolver

The resolver looks for a full ISO date such as `2026-10-15`, then for a month and
day, then for weekdays. The shorthand `\(text)` means "for each text value, run
this small function."

```{r}
resolve_temporal <- function(text, reference_date) {
  lower_text <- str_to_lower(text)

  if (str_detect(text, "\\b\\d{4}-\\d{2}-\\d{2}\\b")) {
    return(list(
      category = "fully resolved",
      date = as.Date(str_extract(text, "\\b\\d{4}-\\d{2}-\\d{2}\\b"))
    ))
  }

  month_match <- str_match(
    text,
    "\\b(January|February|March|April|May|June|July|August|September|October|November|December)\\s+(\\d{1,2})\\b"
  )

  if (!is.na(month_match[1, 1])) {
    resolved <- as.Date(paste(
      format(reference_date, "%Y"),
      match(month_match[1, 2], month.name),
      month_match[1, 3],
      sep = "-"
    ))
    return(list(category = "reference-date needed", date = resolved))
  }

  weekday_names <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")

  if (str_detect(lower_text, "next friday")) {
    target_day <- 6L
    current_day <- as.POSIXlt(reference_date)$wday + 1L
    days_forward <- ((target_day - current_day) %% 7L) + 7L
    return(list(category = "reference-date needed", date = reference_date + days_forward))
  }

  found_weekday <- weekday_names[str_detect(lower_text, str_to_lower(weekday_names))]

  if (length(found_weekday) > 0L) {
    target_day <- match(found_weekday[1], weekday_names)
    current_day <- as.POSIXlt(reference_date)$wday + 1L
    days_forward <- (target_day - current_day) %% 7L
    if (days_forward == 0L) {
      days_forward <- 7L
    }
    return(list(category = "reference-date needed", date = reference_date + days_forward))
  }

  list(category = "not resolvable", date = as.Date(NA))
}

parsed_temporal <- map(
  temporal_examples$text,
  \(text) resolve_temporal(text, reference_date)
)

temporal_results <- temporal_examples |>
  mutate(
    category = map_chr(parsed_temporal, "category"),
    resolved_date = as.Date(map_chr(parsed_temporal, \(item) as.character(item$date)))
  )

knitr::kable(
  temporal_results |>
    select(expr_id, text, category, resolved_date),
  col.names = c("Expression ID", "Text", "Parser category", "Resolved date"),
  caption = "Rule-based temporal parsing results",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(temporal_results$resolved_date[1], as.Date("2026-10-15")),
  identical(temporal_results$resolved_date[4], as.Date("2026-08-31")),
  is.na(temporal_results$resolved_date[2])
)
```

The resolver turns `Monday` into the next Monday after the reference date. It
uses the same-year convention for month-day phrases, gives `next Friday` its own
branch, and treats any bare weekday name as the next occurrence. Those are policy
choices. A different rollover rule would be needed after October 15, and `next
Monday` shows that the weekday branch is too blunt.

## Score the examples

The resolver and expected answers share an author, so the score below does not
estimate parser accuracy. It checks whether this page still behaves as
described.

```{r}
date_matches <- (is.na(temporal_results$resolved_date) & is.na(temporal_results$expected_date)) |
  (!is.na(temporal_results$resolved_date) &
    !is.na(temporal_results$expected_date) &
    temporal_results$resolved_date == temporal_results$expected_date)

category_counts <- tibble(
  category = c("fully resolved", "reference-date needed", "not resolvable")
) |>
  left_join(
    temporal_results |>
      count(category, name = "examples"),
    by = join_by(category)
  ) |>
  mutate(examples = coalesce(examples, 0L))

score_summary <- tibble(
  measure = c("category matches", "date matches"),
  examples = c(
    sum(temporal_results$category == temporal_results$expected_category),
    sum(date_matches)
  )
)

knitr::kable(
  category_counts,
  col.names = c("Parser category", "Examples"),
  caption = "Temporal categories in the examples",
  row.names = FALSE
)

knitr::kable(
  score_summary,
  col.names = c("Measure", "Matching examples"),
  caption = "Agreement with this page's expected answers",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(category_counts$examples, c(1L, 5L, 5L)),
  identical(score_summary$examples, c(11L, 10L))
)
```

One example is fully resolved from the text alone. Five need the reference date.
Five are not resolvable as a single date. The `next Monday` row is the date
mismatch: the rule treats it like `Monday`, while this lesson's expected answer
uses the following week.

## Ask spaCy for a second opinion

spaCy labels spans of text as entity types such as `DATE` and `TIME`. The labels
are useful clues, not answers to the resolution problem.

```{r}
source("R/use-spacy.R")

pipeline <- use_project_spacy()
pipeline_version <- spacy_pipeline_version()

spacy_parsed <- spacy_parse(
  setNames(temporal_examples$text, temporal_examples$expr_id),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
)
spacy_entities <- entity_extract(spacy_parsed, type = "all")

spacy_checks <- spacy_entities |>
  filter(doc_id %in% c("time-01", "time-02", "time-03")) |>
  select(doc_id, entity, entity_type)

spacy_finalize()

knitr::kable(
  tibble(
    name = pipeline_version$name,
    version = pipeline_version$version,
    language = pipeline_version$lang,
    license = pipeline_version$license,
    spacy = pipeline_version$spacy
  ),
  col.names = c("Pipeline", "Version", "Language", "License", "spaCy"),
  caption = "spaCy pipeline used for the second opinion",
  row.names = FALSE
)

knitr::kable(
  spacy_checks,
  col.names = c("Expression ID", "Entity text", "spaCy label"),
  caption = "spaCy DATE and TIME labels on selected examples",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(pipeline_version$version, "3.8.0"),
  identical(pipeline_version$license, "MIT"),
  identical(pipeline_version$spacy, "3.8.7"),
  identical(
    spacy_checks$entity_type,
    c("DATE", "TIME", "DATE")
  )
)
```

spaCy labels `October 15` as `DATE`, `Evening` as `TIME`, and `12-week` as
`DATE`. The first label helps find the span, but it still needs a year. `12-week`
belongs in `DATE` under the scheme spaCy was trained on because that class
includes periods. The label is right, and it is still not a calendar date.

## Temporal parsing in production

Finding a temporal span (like spaCy tagging `next Friday` as a `DATE`) is only the first step. Extracting a fully normalized value requires a dedicated tool like Meta's Duckling or Python's `dateparser` rather than hand-coded R branches.

These tools organize key concepts that custom rules often mix together:
- **Document Creation Time (DCT) / Reference Time**: A relative phrase like `next Friday` or `tomorrow` means nothing without a strict reference date. Parsers do not enforce this automatically; callers must provide the anchor (e.g., passing `RELATIVE_BASE` to `dateparser`).
- **Granularity and representative ambiguity**: A phrase like `last month` is ambiguous. Depending on the context, does it mean the previous calendar month (a granularity of one month) or the last 30 days? A robust parser separates the span detection from the normalized value, keeping these assumptions visible.

## What to remember

- A temporal parser needs a stated reference date for relative expressions.
- Month-day dates need a same-year or rollover policy.
- Weekday phrases need more than one `next Friday` special case.
- Entity labels can find time language without settling the date.

`Apply by October 15` can be put on a 2026 calendar here because the lesson
supplies 2026 as the reference year. `Evening` and `starts in the fall` still do
not give this parser a day to place.

## Sources

- [R Date documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Dates.html)
- [stringr regular expression reference](https://stringr.tidyverse.org/articles/regular-expressions.html)
- [spaCy linguistic features and named entities](https://spacy.io/usage/linguistic-features)
