---
title: "Linking named entities"
subtitle: "Turn mentions into stable identifiers"
author: "Alex Farach"
description: "Learn how alias tables link text mentions to entity identifiers while exposing clean, ambiguous, and missed links."
categories: [entity enrichment, entity linking, workforce research]
---

Credential counts get messy when one notice says `forklift certification` and a
reference list says `Forklift Operator Licence`. The coordinator still wants to
know whether the notices point to the same credential.

**Named entity linking** connects a mention in text to a stable identifier in a
reference list. The identifier matters because spellings can change while the
record stays the same.

::: {.callout-note}
The Riverton organisations, places, and credentials here are invented for the lesson.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain why an entity identifier matters;
- join mention strings to an alias table;
- count clean links, ambiguous links, and misses;
- show how one credential can have multiple aliases; and
- state the limit of alias matching.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load entities and aliases

Three CSV files provide the reference list, aliases, and sentence examples. The
chunk checks the alias-file fingerprint, then uses dplyr and tibble for the
joins and tables. An **entity** is a thing the reference list knows about, such
as an organisation, place, or credential. The **canonical name** is the reference
list's main name for that entity. An **alias** is a surface form that may appear
in text.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(digest)

stopifnot(all(c("readr", "dplyr", "tibble", "digest") %in% loadedNamespaces()))

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

entities <- read_csv(
  "data/riverton/riverton-entities.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    canonical_name = col_character(),
    entity_type = col_character(),
    description = col_character()
  )
)

aliases <- read_csv(
  "data/riverton/riverton-aliases.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    alias = col_character()
  )
)

metadata <- read_csv(
  "data/riverton/riverton-reference-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    fingerprint = col_character()
  )
)

alias_hash <- digest(
  paste(read_lines("data/riverton/riverton-aliases.csv"), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)
expected_alias_hash <- metadata |>
  filter(artifact == "riverton-aliases.csv") |>
  pull(fingerprint)

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(entities), 8L),
  identical(nrow(aliases), 17L),
  identical(sum(aliases$alias == "Riverton"), 2L),
  identical(alias_hash, expected_alias_hash)
)

knitr::kable(
  entities,
  col.names = c("Entity ID", "Canonical name", "Type", "Description"),
  caption = "Invented Riverton entity reference list",
  row.names = FALSE
)
```

The alias table contains two rows for `Riverton`. That deliberate duplicate makes
`Riverton` an ambiguity case for this lesson. It also contains the exact strings
used below, including `forklift certification` and `Forklift Operator Licence`,
so the clean links are a property of this small teaching table.

`Riverton Skills Centre` is an organisation in this entity list. In the geocoding
lesson, the same name is a building. The difference is a normal entity-linking
problem: one name can point to an institution, a site, or both.

## Prepare mention candidates

Two mention candidates come from the Riverton sentences. Three are constructed
checks that exercise the credential and miss cases.

```{r}
real_mentions <- bind_rows(
  sentences |>
    filter(sentence_id == "s023") |>
    transmute(mention_id = sentence_id, source = "Riverton text", text, mention = "Riverton"),
  sentences |>
    filter(sentence_id == "s024") |>
    transmute(mention_id = sentence_id, source = "Riverton text", text, mention = "DATA SUPPORT CERTIFICATE")
)

mention_candidates <- bind_rows(
  real_mentions,
  tibble(
    mention_id = c("link-03", "link-04", "link-05"),
    source = "constructed check",
    text = c(
      "Applicants ask about forklift certification.",
      "The posting names a Forklift Operator Licence.",
      "Customer service badge"
    ),
    mention = c(
      "forklift certification",
      "Forklift Operator Licence",
      "Customer service badge"
    )
  )
)

stopifnot(
  identical(nrow(real_mentions), 2L),
  identical(nrow(mention_candidates), 5L),
  any(mention_candidates$mention == "forklift certification")
)

knitr::kable(
  mention_candidates,
  col.names = c("Mention ID", "Source", "Text", "Mention"),
  caption = "Mention candidates for alias matching",
  row.names = FALSE
)
```

A mention candidate is a span of text the team wants to link. This lesson starts
with hand-selected mentions so the linking step remains visible.

## Join aliases to identifiers

The alias join returns candidate identifiers. A clean link has one candidate, an
ambiguous link has more than one, and a miss has none.

```{r}
linked_rows <- mention_candidates |>
  left_join(
    aliases,
    by = join_by(mention == alias),
    relationship = "many-to-many"
  ) |>
  left_join(entities, by = join_by(entity_id))

link_status <- linked_rows |>
  summarise(
    candidate_ids = paste(na.omit(entity_id), collapse = ", "),
    candidates = sum(!is.na(entity_id)),
    .by = c(mention_id, source, text, mention)
  ) |>
  mutate(
    candidate_ids = if_else(candidate_ids == "", "none", candidate_ids),
    outcome = case_when(
      candidates == 0L ~ "miss",
      candidates == 1L ~ "clean link",
      TRUE ~ "ambiguous"
    )
  )

outcome_counts <- tibble(outcome = c("clean link", "ambiguous", "miss")) |>
  left_join(
    link_status |>
      count(outcome, name = "mentions"),
    by = join_by(outcome)
  ) |>
  mutate(mentions = coalesce(mentions, 0L))

stopifnot(
  identical(nrow(linked_rows), 6L),
  identical(outcome_counts$mentions, c(3L, 1L, 1L)),
  identical(sum(outcome_counts$mentions), 5L)
)

knitr::kable(
  link_status,
  col.names = c(
    "Mention ID", "Source", "Text", "Mention", "Candidate IDs",
    "Candidates", "Outcome"
  ),
  caption = "Alias matching outcomes",
  row.names = FALSE
)

knitr::kable(
  outcome_counts,
  col.names = c("Outcome", "Mentions"),
  caption = "Clean, ambiguous, and missed links",
  row.names = FALSE
)
```

The `relationship` argument tells dplyr that one mention may match several alias
rows. `DATA SUPPORT CERTIFICATE` links cleanly. `Riverton` maps to two place
records, so the alias alone cannot choose one. `Customer service badge` is a
miss. The 3/1/1 counts describe only these five mention candidates.

## Count by identifier

Two different aliases can point to the same credential identifier. That is the
reason the identifier matters.

```{r}
credential_mentions <- linked_rows |>
  filter(entity_type == "credential") |>
  count(entity_id, canonical_name, name = "mentions") |>
  arrange(entity_id)

stopifnot(
  identical(credential_mentions$entity_id, c("CRD-0001", "CRD-0002")),
  identical(credential_mentions$mentions, c(2L, 1L)),
  identical(sum(credential_mentions$mentions), 3L)
)

knitr::kable(
  credential_mentions,
  col.names = c("Entity ID", "Canonical name", "Mentions"),
  caption = "Credential mentions counted by stable identifier",
  row.names = FALSE
)
```

The two forklift aliases count under `CRD-0001`. Without that identifier, the
team would have to decide later whether the two spellings named the same
credential.

## What to remember

- Entity linking connects a mention to a stable identifier.
- An alias table can turn different spellings into one record.
- Duplicate aliases create ambiguity that should be counted.
- Alias matching only finds names someone has already written down.

The two forklift spellings count as one credential in these examples. The table
does not resolve which `Riverton` is meant, and a badge name absent from the alias
file stays unlinked.

## Sources

- [Wikidata data model: items and identifiers](https://www.wikidata.org/wiki/Wikidata:Introduction)
- [DBpedia Spotlight documentation](https://www.dbpedia-spotlight.org/)
- [spaCy entity linking documentation](https://spacy.io/api/entitylinker)
