---
title: "Linking named entities"
subtitle: "Turn mentions into stable identifiers"
author: "Alex Farach"
description: "Learn how alias tables generate entity candidates and why resolution and NIL handling remain separate decisions."
categories: [entity enrichment, entity linking, workforce research]
---

Credential counts get messy when one notice says `forklift certification` and a
reference list says `Forklift Operator Licence`. The coordinator still wants to
know whether the notices point to the same credential.

**Named entity linking** connects a mention in text to a stable identifier in a
reference list. The identifier matters because spellings can change while the
record stays the same.

::: {.callout-note}
The Riverton organisations, places, and credentials here are invented for the lesson.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain why an entity identifier matters;
- join mention strings to an alias table;
- count mentions with zero, one, or several candidates;
- show how one credential can have multiple aliases; and
- state the limit of alias matching.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load entities and aliases

Three CSV files provide the reference list, aliases, and sentence examples. The
chunk checks the alias-file fingerprint, then uses dplyr and tibble for the
joins and tables. An **entity** is a thing the reference list knows about, such
as an organisation, place, or credential. The **canonical name** is the reference
list's main name for that entity. An **alias** is a surface form that may appear
in text.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(digest)

stopifnot(all(c("readr", "dplyr", "tibble", "digest") %in% loadedNamespaces()))

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

entities <- read_csv(
  "data/riverton/riverton-entities.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    canonical_name = col_character(),
    entity_type = col_character(),
    description = col_character()
  )
)

aliases <- read_csv(
  "data/riverton/riverton-aliases.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    alias = col_character()
  )
)

metadata <- read_csv(
  "data/riverton/riverton-reference-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    fingerprint = col_character()
  )
)

alias_hash <- digest(
  paste(read_lines("data/riverton/riverton-aliases.csv"), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)
expected_alias_hash <- metadata |>
  filter(artifact == "riverton-aliases.csv") |>
  pull(fingerprint)

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(entities), 8L),
  identical(nrow(aliases), 17L),
  identical(sum(aliases$alias == "Riverton"), 2L),
  identical(alias_hash, expected_alias_hash)
)

knitr::kable(
  entities,
  col.names = c("Entity ID", "Canonical name", "Type", "Description"),
  caption = "Invented Riverton entity reference list",
  row.names = FALSE
)
```

The alias table contains two rows for `Riverton`. That deliberate duplicate makes
`Riverton` an ambiguity case for this lesson. It also contains the exact strings
used below, including `forklift certification` and `Forklift Operator Licence`,
so the singleton candidates are a property of this small teaching table.

`Riverton Skills Centre` is an organisation in this entity list. In the geocoding
lesson, the same name is a building. The difference is a normal entity-linking
problem: one name can point to an institution, a site, or both.

## Prepare mention candidates

Two mention candidates come from the Riverton sentences. Three are constructed
checks that exercise the credential and miss cases.

```{r}
real_mentions <- bind_rows(
  sentences |>
    filter(sentence_id == "s023") |>
    transmute(mention_id = sentence_id, source = "Riverton text", text, mention = "Riverton"),
  sentences |>
    filter(sentence_id == "s024") |>
    transmute(mention_id = sentence_id, source = "Riverton text", text, mention = "DATA SUPPORT CERTIFICATE")
)

mention_candidates <- bind_rows(
  real_mentions,
  tibble(
    mention_id = c("link-03", "link-04", "link-05"),
    source = "constructed check",
    text = c(
      "Applicants ask about forklift certification.",
      "The posting names a Forklift Operator Licence.",
      "Customer service badge"
    ),
    mention = c(
      "forklift certification",
      "Forklift Operator Licence",
      "Customer service badge"
    )
  )
)

stopifnot(
  identical(nrow(real_mentions), 2L),
  identical(nrow(mention_candidates), 5L),
  any(mention_candidates$mention == "forklift certification")
)

knitr::kable(
  mention_candidates,
  col.names = c("Mention ID", "Source", "Text", "Mention"),
  caption = "Mention candidates for alias matching",
  row.names = FALSE
)
```

A mention candidate is a span of text the team wants to link. This lesson starts
with hand-selected mentions so the linking step remains visible.

## Generate candidates from normalized aliases

In real datasets, exact casing rarely matches perfectly. Before joining, both
the mentions and aliases are lowercased so differently capitalized names can
match. This join generates candidates; it does not prove a link. A mention can
have zero, one, or several candidates, and a separate resolution policy must
decide whether to accept one or return NIL.

```{r}
library(stringr)
normalized_mentions <- mention_candidates |>
  mutate(normalized_mention = str_to_lower(mention))

normalized_aliases <- aliases |>
  mutate(normalized_alias = str_to_lower(alias)) |>
  distinct(entity_id, normalized_alias)

linked_rows <- normalized_mentions |>
  left_join(
    normalized_aliases,
    by = join_by(normalized_mention == normalized_alias),
    relationship = "many-to-many"
  ) |>
  left_join(entities, by = join_by(entity_id))

link_status <- linked_rows |>
  summarise(
    candidate_ids = paste(unique(na.omit(entity_id)), collapse = ", "),
    candidates = n_distinct(na.omit(entity_id)),
    .by = c(mention_id, source, text, mention)
  ) |>
  mutate(
    candidate_ids = if_else(candidate_ids == "", "none", candidate_ids),
    outcome = case_when(
      candidates == 0L ~ "no candidate",
      candidates == 1L ~ "one candidate",
      TRUE ~ "multiple candidates"
    )
  )

resolved_links <- link_status |>
  mutate(
    resolved_entity_id = if_else(
      candidates == 1L,
      candidate_ids,
      NA_character_
    ),
    resolution = case_when(
      candidates == 0L ~ "NIL",
      candidates == 1L ~ "accepted singleton candidate",
      TRUE ~ "needs disambiguation"
    )
  )

outcome_counts <- tibble(
  outcome = c("one candidate", "multiple candidates", "no candidate")
) |>
  left_join(
    link_status |>
      count(outcome, name = "mentions"),
    by = join_by(outcome)
  ) |>
  mutate(mentions = coalesce(mentions, 0L))

stopifnot(
  identical(nrow(linked_rows), 6L),
  identical(outcome_counts$mentions, c(3L, 1L, 1L)),
  identical(sum(outcome_counts$mentions), 5L),
  identical(
    sort(count(resolved_links, resolution, name = "mentions")$mentions),
    c(1L, 1L, 3L)
  )
)

knitr::kable(
  link_status,
  col.names = c(
    "Mention ID", "Source", "Text", "Mention", "Candidate IDs",
    "Candidates", "Candidate outcome"
  ),
  caption = "Alias matching outcomes",
  row.names = FALSE
)

knitr::kable(
  outcome_counts,
  col.names = c("Outcome", "Mentions"),
  caption = "Candidate-generation outcomes",
  row.names = FALSE
)
```

The `relationship` argument tells dplyr that one mention may match several
alias rows. `DATA SUPPORT CERTIFICATE` produces one candidate. This toy policy
accepts singleton candidates, but a production linker would still apply a
score or NIL threshold. `Riverton` maps to two place records, so the alias alone
cannot choose one. `Customer service badge` has no candidate. The 3/1/1 counts
describe only candidate generation for these five mentions.

## Count by identifier

Two different aliases can point to the same credential identifier. That is the
reason the identifier matters.

```{r}
credential_mentions <- resolved_links |>
  filter(resolution == "accepted singleton candidate") |>
  rename(entity_id = resolved_entity_id) |>
  left_join(entities, by = join_by(entity_id)) |>
  filter(entity_type == "credential") |>
  count(entity_id, canonical_name, name = "mentions") |>
  arrange(entity_id)

stopifnot(
  identical(credential_mentions$entity_id, c("CRD-0001", "CRD-0002")),
  identical(credential_mentions$mentions, c(2L, 1L)),
  identical(sum(credential_mentions$mentions), 3L)
)

knitr::kable(
  credential_mentions,
  col.names = c("Entity ID", "Canonical name", "Mentions"),
  caption = "Credential mentions counted by stable identifier",
  row.names = FALSE
)
```

The two forklift aliases count under `CRD-0001`. Without that identifier, the
team would have to decide later whether the two spellings named the same
credential.

## Real-world linking pipelines

The normalized join here shows candidate generation from an alias table.
Production entity linking systems add a separate resolution step:

- **Unicode and case normalization policy**: A pipeline must define how it handles accents (e.g., NFC/NFD normalization) and casing before joining, ensuring mentions and aliases are normalized consistently. Never report linking rates over an unnormalized exact string join.
- **Candidate generation and ranking**: When an alias is ambiguous (like `Riverton`), a modern pipeline generates candidate entities and ranks them using the surrounding sentence context.
- **NIL handling**: A robust system detects when a mention matches an alias string but refers to a *different* entity not in the knowledge base (a NIL entity), rather than blindly linking it.
- **KB versioning**: An entity's canonical name or status may change. Linking results must record the specific version of the knowledge base (KB) used.
- **Alias and candidate-recall checks**: Pipelines need ongoing evaluation to ensure the alias table contains common misspellings (alias recall) and the ranking system surfaces the correct entity (candidate recall).

## What to remember

- Entity linking connects a mention to a stable identifier.
- An alias table can turn different spellings into one record.
- Duplicate aliases create ambiguity that should be counted.
- Alias matching only finds names someone has already written down.

The two forklift spellings count as one credential in these examples. The table
does not resolve which `Riverton` is meant, and a badge name absent from the alias
file stays unlinked.

## Sources

- [Wikidata data model: items and identifiers](https://www.wikidata.org/wiki/Wikidata:Introduction)
- [DBpedia Spotlight documentation](https://www.dbpedia-spotlight.org/)
- [spaCy entity linking documentation](https://spacy.io/api/entitylinker)
