---
title: "Resolving coreference"
subtitle: "Link pronouns to the noun phrases they may refer to"
author: "Alex Farach"
description: "Learn how a nearest-noun baseline handles pronouns in a constructed Riverton passage and where it fails."
categories: [entity enrichment, coreference resolution, workforce research]
---

Pronouns make the short applicant note harder than it first appears: `she` is
clear enough to a reader, while `he` could point to more than one person.

**Coreference resolution** links different expressions that refer to the same
thing. A pronoun such as `they` may refer back to `the workers`. This lesson uses
a small baseline rule so its mistakes remain visible.

::: {.callout-note}
The Riverton passage and reference setting below are fictional.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define coreference in plain language;
- inspect noun phrases and pronouns from a spaCy parse;
- apply a nearest-agreeing-noun baseline;
- compare the baseline with this page's expected answers; and
- explain why document-level context matters.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Use a constructed passage

The lesson first checks the Riverton CSV, then uses a constructed passage because
the stored sentences are too thin for coreference teaching.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(spacyr)
source("R/use-spacy.R")

pipeline <- use_project_spacy()
pipeline_version <- spacy_pipeline_version()

stopifnot(all(c("readr", "dplyr", "tibble", "purrr", "stringr", "spacyr") %in% loadedNamespaces()))

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

riverton_parsed <- spacy_parse(
  setNames(sentences$text, sentences$sentence_id),
  pos = TRUE,
  lemma = TRUE
) |>
  as_tibble()

riverton_pronouns <- riverton_parsed |>
  filter(pos == "PRON")

teaching_passage <- paste(
  "The coordinator posted the training notice on Monday.",
  "The applicant read it before the bus arrived.",
  "The supervisor called the applicant after she finished the form.",
  "The supervisor told the coordinator that he could open another class.",
  "The workers thanked the supervisor because they needed evening seats.",
  "The centre moved the flyer near the entrance, and it stayed there all week.",
  sep = " "
)

parsed_raw <- spacy_parse(
  c(teaching_passage = teaching_passage),
  pos = TRUE,
  tag = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE,
  additional_attributes = c("morph")
)

parsed <- parsed_raw |>
  as_tibble() |>
  mutate(
    global_token = row_number(),
    morph = map_chr(morph, as.character)
  )

phrases <- nounphrase_extract(parsed_raw)

stopifnot(
  identical(pipeline_version$version, "3.8.0"),
  identical(pipeline_version$license, "MIT"),
  identical(pipeline_version$spacy, "3.8.7"),
  identical(nrow(sentences), 28L),
  identical(nrow(riverton_pronouns), 1L),
  identical(nrow(phrases), 22L),
  identical(sum(parsed$pos == "PRON"), 5L)
)

knitr::kable(
  riverton_pronouns |>
    select(doc_id, token, pos),
  col.names = c("Sentence ID", "Token", "spaCy part of speech"),
  caption = "Pronouns found in the 28 Riverton sentences",
  row.names = FALSE
)

knitr::kable(
  tibble(
    name = pipeline_version$name,
    version = pipeline_version$version,
    language = pipeline_version$lang,
    license = pipeline_version$license,
    spacy = pipeline_version$spacy
  ),
  col.names = c("Pipeline", "Version", "Language", "License", "spaCy"),
  caption = "spaCy pipeline used for the parse",
  row.names = FALSE
)

knitr::kable(
  phrases,
  col.names = c("Document", "Sentence", "Noun phrase"),
  caption = "Noun phrases extracted from the constructed passage",
  row.names = FALSE
)
```

spaCy marks one pronoun in the 28 Riverton sentences. One example is not enough
to teach pronoun linking, so the constructed passage gives the rule more chances
to fail.

The installed `en_core_web_sm` pipeline has named-entity, tagger, parser, and
other components, but no coreference component. The rule below is a baseline, not
a full coreference system.

## Apply the baseline rule

The rule is explicit: link each personal pronoun to the nearest preceding noun
phrase that has the same number feature. Here, number means singular or plural as
reported by spaCy morphology on the noun-phrase head. The thing a pronoun points
back to is called its antecedent.

```{r}
number_from_morph <- function(morph_value) {
  case_when(
    str_detect(morph_value, "Number=Plur") ~ "Plur",
    str_detect(morph_value, "Number=Sing") ~ "Sing",
    TRUE ~ NA_character_
  )
}

phrase_tokens <- parsed |>
  mutate(
    np_inside = nounphrase != "",
    np_id = if_else(np_inside, cumsum(str_detect(nounphrase, "beg")), NA_integer_)
  ) |>
  filter(np_inside)

noun_phrases <- phrase_tokens |>
  summarise(
    phrase = paste(token, collapse = " "),
    root_token = token[str_detect(nounphrase, "root")][1],
    root_pos = pos[str_detect(nounphrase, "root")][1],
    root_morph = morph[str_detect(nounphrase, "root")][1],
    start_token = min(global_token),
    end_token = max(global_token),
    .by = np_id
  ) |>
  mutate(number = number_from_morph(root_morph))

antecedents <- noun_phrases |>
  filter(root_pos %in% c("NOUN", "PROPN"), !is.na(number)) |>
  rename(antecedent = phrase, antecedent_number = number) |>
  mutate(join_key = 1L)

pronouns <- parsed |>
  filter(pos == "PRON", str_to_lower(token) %in% c("it", "she", "he", "they")) |>
  transmute(
    pronoun_id = paste0("p", row_number()),
    pronoun = token,
    sentence_id,
    pronoun_token = global_token,
    pronoun_number = number_from_morph(morph),
    join_key = 1L
  )

rule_links <- pronouns |>
  left_join(antecedents, by = join_by(join_key), relationship = "many-to-many") |>
  filter(end_token < pronoun_token, antecedent_number == pronoun_number) |>
  arrange(pronoun_id, desc(end_token)) |>
  summarise(
    rule_antecedent = first(antecedent),
    .by = c(pronoun_id, pronoun, sentence_id, pronoun_number)
  )

stopifnot(
  identical(nrow(pronouns), 5L),
  identical(nrow(rule_links), 5L),
  identical(rule_links$rule_antecedent[1], "The applicant"),
  identical(rule_links$rule_antecedent[5], "the entrance")
)

knitr::kable(
  rule_links,
  col.names = c("Pronoun ID", "Pronoun", "Sentence", "Number", "Rule antecedent"),
  caption = "Nearest preceding noun phrase with matching number",
  row.names = FALSE
)
```

The rule chooses `The applicant` for the first `it` because that is the nearest
singular noun phrase. A reader can see that `it` means the training notice.

## Score the rule

The expected answers treat `he` as genuinely ambiguous because the supervisor or
the coordinator could open another class. That row is counted separately.

```{r}
marked_coreference <- tibble(
  pronoun_id = c("p1", "p2", "p3", "p4", "p5"),
  expected_antecedent = c(
    "the training notice",
    "the applicant",
    "ambiguous",
    "The workers",
    "the flyer"
  ),
  scoreable = c(TRUE, TRUE, FALSE, TRUE, TRUE)
)

coreference_score <- rule_links |>
  left_join(marked_coreference, by = join_by(pronoun_id)) |>
  mutate(
    result = case_when(
      !scoreable ~ "ambiguous reference",
      str_to_lower(rule_antecedent) == str_to_lower(expected_antecedent) ~ "correct",
      TRUE ~ "wrong"
    )
  )

score_counts <- tibble(result = c("correct", "wrong", "ambiguous reference")) |>
  left_join(
    coreference_score |>
      count(result, name = "pronouns"),
    by = join_by(result)
  ) |>
  mutate(pronouns = coalesce(pronouns, 0L))

spacy_finalize()

stopifnot(
  identical(score_counts$pronouns, c(2L, 2L, 1L)),
  identical(sum(score_counts$pronouns), 5L)
)

knitr::kable(
  coreference_score |>
    select(pronoun_id, pronoun, rule_antecedent, expected_antecedent, result),
  col.names = c("Pronoun ID", "Pronoun", "Rule antecedent", "Expected answer", "Result"),
  caption = "Coreference baseline compared with expected answers",
  row.names = FALSE
)

knitr::kable(
  score_counts,
  col.names = c("Result", "Pronouns"),
  caption = "Correct, wrong, and ambiguous pronoun links",
  row.names = FALSE
)
```

Because the baseline and expected answers share an author, the score is only a
check on this page. The two correct answers are easy cases: `The workers` is the
only plural phrase before `they`, and the rule links `she` without using gender.
Excluding the ambiguous `he` row changes the scored denominator from five
pronouns to four. Real coreference systems need document context, syntax, and
meaning.

## Modern coreference resolution

The deterministic nearest-agreeing-noun syntax rule shown here is an outdated baseline intended only to expose the mechanism of coreference. Modern production coreference resolution is handled almost entirely by neural models (such as AllenNLP or large language models).

When evaluating modern systems, keep these limits in mind:
- **Cluster metrics**: Coreference models are evaluated using specialized clustering metrics (like MUC, B³, and CEAF) rather than simple pairwise accuracy, because errors compound when linking long chains of mentions across a document.
- **Ambiguity**: Language is frequently ambiguous (like the `he` example). Deterministic models often force a link where a human would recognize true ambiguity.
- **Stereotype and coverage asymmetry**: Coreference models often inherit training biases, successfully resolving pronouns that match occupational gender stereotypes (e.g., linking `he` to `doctor` and `she` to `nurse`) while failing on anti-stereotypical or gender-neutral (`they`) examples.

## What to remember

- Coreference links expressions that refer to the same thing.
- `en_core_web_sm` does not include a coreference resolver.
- A nearest-agreeing-noun rule is a baseline for inspection.
- Ambiguous pronouns should be marked as ambiguous, not forced into certainty.

After checking the five pronouns, the coordinator has two easy matches, two
mistakes, and one ambiguity. That is enough to explain the baseline, not enough
to use it as a resolver.

## Sources

- [spaCy linguistic features](https://spacy.io/usage/linguistic-features)
- [spaCy English pipelines](https://spacy.io/models/en)
- [Universal Dependencies features](https://universaldependencies.org/u/feat/index.html)
