---
title: "Turning reading decisions into labels"
subtitle: "Create a manual annotation codebook for workforce text"
author: "Alex Farach"
description: "Learn how people label workforce sentences without pretending that judgment is automatic or neutral."
categories: [training data, annotation, workforce research]
---

The archived job page and checked flyer transcript leave the fictional
Riverton Workforce Lab with 28 text units. To make them usable, the team must
separate training offers, application requirements, schedules, named skills,
and text that fits none of those categories. These distinctions affect which
information a worker would find while comparing notices, but the notices do
not label themselves.

A computer cannot learn these categories until people define them. **Manual
annotation** is the work of reading an item and assigning a label according to
written instructions. The hard part is not typing the label. It is deciding
what the label means and applying that meaning consistently. The task records
what a sentence explicitly says; it does not decide whether a worker qualifies
or whether a program is a good option.

::: {.callout-note}
The workforce text, labels, and research team are fictional. The examples were
created to expose ordinary annotation choices without using worker or applicant
records.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define an annotation unit and label set;
- write rules for what each label includes and excludes;
- create a versioned codebook with foundryR;
- inspect ambiguous cases rather than hiding them; and
- explain why a human label is a documented judgment, not automatic truth.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Decide what one label describes

The annotation unit is one sentence. That keeps each decision small enough to
check, while the `document_id` preserves the job posting or flyer that supplied
the text. The next chunk loads readr for files, dplyr and tibble for tables,
and stringr for text checks, then reads the CSV into a tibble.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)

workforce_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

allowed_labels <- c(
  "training",
  "requirement",
  "schedule",
  "skill",
  "other"
)

first_sentences <- workforce_sentences |>
  select(
    sentence_id,
    document_id,
    text,
    reference_label,
    uncertainty
  ) |>
  slice_head(n = 8)

stopifnot(
  identical(nrow(workforce_sentences), 28L),
  identical(
    workforce_sentences$sentence_id,
    sprintf("s%03d", 1:28)
  ),
  all(workforce_sentences$reference_label %in% allowed_labels),
  !anyDuplicated(workforce_sentences$sentence_id),
  all(workforce_sentences$derived == "false"),
  all(workforce_sentences$transformation == "verbatim")
)

knitr::kable(
  first_sentences,
  col.names = c(
    "Sentence ID",
    "Document ID",
    "Text",
    "Reference label",
    "Uncertainty"
  ),
  caption = "The first eight manually labeled sentences",
  row.names = FALSE
)
```

The file calls the reviewed result a **reference label**, not a ground-truth
label. The label records how this codebook was applied. Another defensible
codebook could divide the concepts differently.

The six flyer rows match its six-line checked transcript. No wording has been
expanded or interpreted in the annotation file.

```{r}
flyer_rows <- workforce_sentences |>
  filter(document_id == "F001")
flyer_transcript <- read_lines(
  "data/workforce/training-flyer-ground-truth.txt"
)

stopifnot(
  identical(nrow(flyer_rows), 6L),
  identical(length(flyer_transcript), 6L),
  identical(flyer_rows$source_line, as.character(1:6)),
  identical(flyer_rows$text, flyer_transcript)
)

knitr::kable(
  flyer_rows |>
    select(
      sentence_id,
      source_line,
      text,
      reference_label
    ),
  col.names = c(
    "Sentence ID",
    "Flyer line",
    "Verbatim text",
    "Reference label"
  ),
  caption = "Row-level lineage from the flyer transcript",
  row.names = FALSE
)
```

## Write the codebook before scaling up

A **codebook** records the label definitions, instructions, examples, and
version. The complete instrument lives in `R/workforce-codebook.R`. Its rules
are:

- `training`: the sentence offers or describes learning, mentoring,
  certification training, or an apprenticeship;
- `requirement`: the sentence states or prefers experience, education,
  credentials, a license, a portfolio, or a physical condition;
- `schedule`: the sentence describes when work or classes occur;
- `skill`: the sentence names an ability used in the work; and
- `other`: none of the four definitions applies; headings, deadlines,
  location, work setting, and travel belong here unless the sentence
  explicitly states another category.

The sentence's main explicit claim receives one label. Reviewers do not infer a
training offer from a job title or assume that a preferred credential is
mandatory. A required shift receives `schedule` because timing is the main
claim. Reviewers mark `needs_review` when more than one label remains
defensible.

foundryR can store the codebook as a versioned object with a strict label
schema and a content hash.

```{r}
source("R/workforce-codebook.R")
annotation_codebook <- workforce_codebook()

stopifnot(
  inherits(
    annotation_codebook,
    "foundry_codebook"
  ),
  identical(
    annotation_codebook$version,
    "1.0.0"
  ),
  identical(
    str_length(annotation_codebook$hash),
    64L
  ),
  identical(
    unclass(
      annotation_codebook$schema$
        properties$label$enum
    ),
    allowed_labels
  ),
  identical(
    unclass(
      annotation_codebook$schema$
        properties$uncertainty$enum
    ),
    c("certain", "needs_review")
  ),
  identical(
    length(annotation_codebook$examples),
    7L
  ),
  all(workforce_sentences$codebook_version == annotation_codebook$version),
  all(workforce_sentences$codebook_hash == annotation_codebook$hash),
  all(str_length(workforce_sentences$rationale) > 0),
  all(workforce_sentences$uncertainty %in% c("certain", "needs_review"))
)

annotation_codebook
```

The printed hash changes when the instructions, schema, examples, or version
changes. This makes two annotation runs easier to compare. The code above does
not call an AI service; foundryR defines and fingerprints the measurement
instrument.

## Count labels without interpreting the count

A label distribution can expose a missing category or a data-entry problem.
It cannot establish how common these ideas are across the real labor market.

```{r}
label_counts <- workforce_sentences |>
  mutate(
    reference_label = factor(
      reference_label,
      levels = allowed_labels
    )
  ) |>
  count(reference_label, name = "sentences", .drop = FALSE) |>
  transmute(
    label = as.character(reference_label),
    sentences
  )

stopifnot(
  identical(sum(label_counts$sentences), 28L),
  identical(
    as.integer(label_counts$sentences),
    c(7L, 9L, 5L, 2L, 5L)
  )
)

knitr::kable(
  label_counts,
  col.names = c("Reference label", "Sentences"),
  caption = "Label counts in the teaching dataset",
  row.names = FALSE
)
```

The counts describe this designed example only. They were not sampled from all
jobs in Riverton or from any real population.

## Keep ambiguity visible

Sentence `s013` mentions a certificate, a word that also appears in training
offers. Its main claim is that a certificate is preferred, so version 1.0.0 of
the codebook labels it as a requirement.

```{r}
ambiguous_case <- workforce_sentences |>
  filter(sentence_id == "s013")

stopifnot(
  identical(nrow(ambiguous_case), 1L),
  identical(ambiguous_case$reference_label, "requirement"),
  str_detect(ambiguous_case$text, fixed("certificate"))
)

knitr::kable(
  ambiguous_case |>
    select(
      sentence_id,
      document_id,
      text,
      reference_label,
      uncertainty,
      rationale
    ),
  col.names = c(
    "Sentence ID",
    "Document ID",
    "Text",
    "Reference label",
    "Uncertainty",
    "Rationale"
  ),
  caption = "A sentence that needs the codebook rule",
  row.names = FALSE
)
```

A reviewer who chose `training` may have followed a defensible reading rather
than worked carelessly. The team should record the disagreement, discuss the
rule, and issue a new codebook version if the definition changes. Overwriting
the label would erase the evidence needed to improve the instrument.

## Protect the people doing the work

Manual annotation can involve sensitive, repetitive, or distressing material.
Before assigning work:

1. explain the purpose and allowed use of the data;
2. remove personal information that reviewers do not need;
3. pay fairly for reading, training, and disagreement review;
4. give reviewers a way to skip harmful material;
5. separate quality feedback from surveillance; and
6. record which languages and lived experience the task requires.

Expertise depends on the question. A worker, career counselor, employer, and
labor researcher may interpret the same sentence differently. Those
perspectives should be designed into the study rather than treated as noise.

## What to remember

- Annotation labels are measurements created from written rules.
- One sentence per row keeps the decision small and traceable.
- A codebook needs a version, examples, and a change record.
- Ambiguity should lead to review, not silent correction.
- Human labels require ethical working conditions and relevant expertise.

The Lab has defined the categories and preserved its reading decisions. To test
selection without exposing known answers, the active-learning exercise starts
with five complete job postings and hides two other documents from a small
model.

## Sources

- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
- [Datasheets for Datasets](https://doi.org/10.1145/3458723)
- [Data Collection and Quality Challenges in Deep Learning](https://arxiv.org/abs/2002.11821)
- [foundryR codebook documentation](https://farach.github.io/foundryR/reference/foundry_codebook.html)
