---
title: "Turning reading decisions into labels"
subtitle: "Create a manual annotation codebook for workforce text"
description: "Learn how people label workforce text units without pretending that judgment is automatic or neutral."
categories: [training data, annotation, workforce research]
---

The archived job page and checked flyer transcript leave the fictional
Riverton Workforce Lab with 28 text units. To make them usable, the team must
separate training offers, application requirements, schedules, named skills,
and text that fits none of those categories. These distinctions affect which
information a worker would find while comparing notices, but the notices do
not label themselves.

A computer cannot learn these categories until people define them. **Manual
annotation** is the work of reading an item and assigning a label according to
written instructions. The hard part is not typing the label. It is deciding
what the label means and applying that meaning consistently. The task records
what a text unit explicitly says; it does not decide whether a worker qualifies
or whether a program is a good option.

::: {.callout-note}
The workforce text, labels, and research team are fictional. The examples were
created to expose ordinary annotation choices without using worker or applicant
records.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define an annotation unit and label set;
- write rules for what each label includes and excludes;
- create a versioned codebook with foundryR;
- run a pilot and let disagreement change the rules;
- inspect ambiguous cases rather than hiding them; and
- explain why a human label is a documented judgment, not automatic truth.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Decide what one label describes

The annotation unit is one line of source text: a single job-detail list item
from a posting, or a single line from the flyer. That keeps each decision small
enough to check, while the `document_id` preserves the job posting or flyer that
supplied the text.

Calling those lines sentences would be convenient and wrong. Two of them are
headings in capital letters, three are noun phrases with no verb, and one is an
imperative clause without terminal punctuation. The
distinction matters because a rule written for sentences, such as "label the
main claim of the text unit," has to say what it does with a heading that makes
no claim at all.

The next chunk loads readr for files, dplyr and tibble for tables,
purrr for repeated checks, and stringr for text checks, then reads the CSV into
a tibble.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)

workforce_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

allowed_labels <- c(
  "training",
  "requirement",
  "schedule",
  "skill",
  "other"
)

first_units <- workforce_sentences |>
  select(
    sentence_id,
    document_id,
    text,
    reference_label,
    uncertainty
  ) |>
  slice_head(n = 8)

knitr::kable(
  first_units,
  col.names = c(
    "Unit ID",
    "Document ID",
    "Text",
    "Reference label",
    "Uncertainty"
  ),
  caption = "The first eight manually labeled text units",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(workforce_sentences), 28L),
  identical(
    workforce_sentences$sentence_id,
    sprintf("s%03d", 1:28)
  ),
  all(workforce_sentences$reference_label %in% allowed_labels),
  !anyDuplicated(workforce_sentences$sentence_id),
  all(workforce_sentences$derived == "false"),
  all(workforce_sentences$transformation == "verbatim")
)
```

The file calls the reviewed result a **reference label**, not a ground-truth
label. The label records how this codebook was applied. Another defensible
codebook could divide the concepts differently.

### Count what the units actually are

The claim that these are not all sentences is checkable, but punctuation alone
cannot classify grammar. `Apply by October 15` is an imperative clause even
without a full stop. The six flyer lines are therefore classified explicitly
rather than inferred from their last character.

```{r}
unit_shapes <- workforce_sentences |>
  mutate(
    shape = case_when(
      sentence_id %in% c("s023", "s024") ~ "heading in capitals",
      sentence_id %in% c("s025", "s026", "s027") ~ "phrase without a verb",
      sentence_id == "s028" ~ "imperative clause",
      TRUE ~ "complete sentence"
    )
  )

shape_counts <- unit_shapes |>
  count(shape, name = "units")

knitr::kable(
  shape_counts,
  col.names = c("Shape of the unit", "Units"),
  caption = "Twenty-two sentences, two headings, three phrases, one imperative clause",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(sum(shape_counts$units), 28L),
  identical(
    unit_shapes$shape[unit_shapes$sentence_id == "s028"],
    "imperative clause"
  ),
  identical(
    sum(unit_shapes$shape == "heading in capitals"),
    2L
  ),
  identical(
    sum(unit_shapes$shape == "phrase without a verb"),
    3L
  ),
  all(
    unit_shapes$document_id[
      unit_shapes$shape != "complete sentence"
    ] == "F001"
  )
)
```

Five of the 28 units are not sentences. A sixth, `Apply by October 15`, is an
imperative clause whose flyer typography omits terminal punctuation. All six
come from a notice laid out for a reader walking past a noticeboard. `Paid
training stipend` has no verb; `RIVERTON SKILLS OPEN HOUSE` announces an event.

Both still receive a label, and the codebook has to say how. That is the reason
the rules below talk about the unit's main explicit claim rather than the
sentence's subject.

The six flyer rows match its six-line checked transcript. No wording has been
expanded or interpreted in the annotation file.

```{r}
flyer_rows <- workforce_sentences |>
  filter(document_id == "F001")
flyer_transcript <- read_lines(
  "data/workforce/training-flyer-ground-truth.txt"
)

knitr::kable(
  flyer_rows |>
    select(
      sentence_id,
      source_line,
      text,
      reference_label
    ),
  col.names = c(
    "Unit ID",
    "Flyer line",
    "Verbatim text",
    "Reference label"
  ),
  caption = "Row-level lineage from the flyer transcript",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(flyer_rows), 6L),
  identical(length(flyer_transcript), 6L),
  identical(flyer_rows$source_line, as.character(1:6)),
  identical(flyer_rows$text, flyer_transcript)
)
```

## Write the codebook before scaling up

A **codebook** records the label definitions, instructions, examples, and
version. The complete instrument lives in `R/workforce-codebook.R`. Its rules
are:

- `training`: the text unit offers or describes learning, mentoring,
  certification training, or an apprenticeship;
- `requirement`: the text unit states or prefers experience, education,
  credentials, a license, a portfolio, or a physical condition;
- `schedule`: the text unit describes when work or classes occur;
- `skill`: the text unit names an ability used in the work; and
- `other`: none of the four definitions applies; headings, deadlines,
  location, work setting, and travel belong here unless the text unit
  explicitly states another category.

The text unit's main explicit claim receives one label. Reviewers do not infer a
training offer from a job title or assume that a preferred credential is
mandatory. A required shift receives `schedule` because timing is the main
claim. Reviewers mark `needs_review` when more than one label remains
defensible.

foundryR can store the codebook as a versioned object with a strict label
schema and a content hash.

```{r}
source("R/workforce-codebook.R")
annotation_codebook <- workforce_codebook()
recorded_codebook <- workforce_codebook("1.0.0")

annotation_codebook
```

```{r}
#| include: false

stopifnot(
  inherits(
    annotation_codebook,
    "foundry_codebook"
  ),
  identical(
    annotation_codebook$version,
    "1.1.0"
  ),
  identical(recorded_codebook$version, "1.0.0"),
  !identical(annotation_codebook$hash, recorded_codebook$hash),
  identical(
    str_length(annotation_codebook$hash),
    64L
  ),
  identical(
    unclass(
      annotation_codebook$schema$
        properties$label$enum
    ),
    allowed_labels
  ),
  identical(
    unclass(
      annotation_codebook$schema$
        properties$uncertainty$enum
    ),
    c("certain", "needs_review")
  ),
  identical(
    length(annotation_codebook$examples),
    7L
  ),
  all(workforce_sentences$codebook_version == recorded_codebook$version),
  all(workforce_sentences$codebook_hash == recorded_codebook$hash),
  all(str_length(workforce_sentences$rationale) > 0),
  all(workforce_sentences$uncertainty %in% c("certain", "needs_review"))
)
```

The printed hash changes when the instructions, schema, examples, or version
changes. Version 1.1.0 corrects the unit name. The archived labels keep their
original 1.0.0 fingerprint because changing an instrument does not retroactively
change which instrument produced earlier records. A new annotation run should
use 1.1.0. The code above does not call an AI service; foundryR defines and
fingerprints the measurement instrument.

## Count labels without interpreting the count

A label distribution can expose a missing category or a data-entry problem.
It cannot establish how common these ideas are across the real labor market.

```{r}
label_counts <- workforce_sentences |>
  mutate(
    reference_label = factor(
      reference_label,
      levels = allowed_labels
    )
  ) |>
  count(reference_label, name = "units", .drop = FALSE) |>
  transmute(
    label = as.character(reference_label),
    units
  )

knitr::kable(
  label_counts,
  col.names = c("Reference label", "Text units"),
  caption = "Label counts in the teaching dataset",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(sum(label_counts$units), 28L),
  identical(
    as.integer(label_counts$units),
    c(7L, 9L, 5L, 2L, 5L)
  )
)
```

The counts describe this designed example only. They were not sampled from all
jobs in Riverton or from any real population.

## Keep ambiguity visible

Unit `s013` mentions a certificate, a word that also appears in training
offers. Its main claim is that a certificate is preferred, so version 1.1.0 of
the codebook labels it as a requirement.

```{r}
ambiguous_case <- workforce_sentences |>
  filter(sentence_id == "s013")

knitr::kable(
  ambiguous_case |>
    select(
      sentence_id,
      document_id,
      text,
      reference_label,
      uncertainty,
      rationale
    ),
  col.names = c(
    "Unit ID",
    "Document ID",
    "Text",
    "Reference label",
    "Uncertainty",
    "Rationale"
  ),
  caption = "A text unit that needs the codebook rule",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(ambiguous_case), 1L),
  identical(ambiguous_case$reference_label, "requirement"),
  str_detect(ambiguous_case$text, fixed("certificate"))
)
```

A reviewer who chose `training` may have followed a defensible reading rather
than worked carelessly. The team should record the disagreement, discuss the
rule, and issue a new codebook version if the definition changes. Overwriting
the label would erase the evidence needed to improve the instrument.

## Run a pilot before labeling everything

The rules above did not arrive fully formed. They came from a pilot: two
reviewers labeled six units independently, without discussion, and the pairs
that disagreed decided what the codebook still had to say.

```{r}
pilot_round <- tibble(
  sentence_id = c("s001", "s004", "s010", "s013", "s020", "s024"),
  reviewer_p = c(
    "training",
    "skill",
    "requirement",
    "training",
    "requirement",
    "training"
  ),
  reviewer_q = c(
    "training",
    "skill",
    "other",
    "requirement",
    "schedule",
    "other"
  )
) |>
  left_join(
    workforce_sentences |>
      select(sentence_id, text, reference_label),
    by = join_by(sentence_id)
  ) |>
  mutate(agreed = reviewer_p == reviewer_q)

pilot_agreement <- mean(pilot_round$agreed)

knitr::kable(
  pilot_round |>
    select(sentence_id, text, reviewer_p, reviewer_q, agreed),
  col.names = c(
    "Unit",
    "Text",
    "Reviewer P",
    "Reviewer Q",
    "Agreed"
  ),
  caption = "The six-unit pilot, before any rule was written",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(pilot_round), 6L),
  identical(sum(pilot_round$agreed), 2L),
  isTRUE(all.equal(pilot_agreement, 2 / 6)),
  all(!is.na(pilot_round$text)),
  identical(
    pilot_round$sentence_id[!pilot_round$agreed],
    c("s010", "s013", "s020", "s024")
  )
)
```

Two of six matched in this designed pilot. Six items cannot establish a typical
first-pass agreement rate. Here, the four disagreements identify wording that
the initial instructions did not settle.

### Turn each disagreement into a rule

The Lab wrote one rule per disagreement, then adjudicated the four units under
those rules. The adjudicator was a third reader who had not labeled the pilot.

```{r}
adjudication_log <- tibble(
  sentence_id = c("s010", "s013", "s020", "s024"),
  question = c(
    "Is a preferred condition still an entry condition?",
    "Does naming a credential make a unit a training offer?",
    "Does a required shift describe timing or an entry condition?",
    "Does a heading claim what its words name?"
  ),
  rule_added = c(
    "A preference is still an entry condition.",
    "A credential that is required or preferred is not training unless learning is offered.",
    "Timing takes precedence when a shift is required.",
    "Headings are other unless the unit explicitly states another category."
  ),
  adjudicated_label = c(
    "requirement",
    "requirement",
    "schedule",
    "other"
  )
)

instructions <- annotation_codebook$instructions
rule_clauses <- c(
  "main claim states or prefers",
  "is not training",
  "schedule takes precedence",
  "Headings, deadlines, location, work setting, and travel are other"
)

adjudication_check <- adjudication_log |>
  left_join(
    workforce_sentences |>
      select(sentence_id, reference_label, uncertainty),
    by = join_by(sentence_id)
  ) |>
  mutate(
    matches_file = adjudicated_label == reference_label,
    clause_in_codebook = map_lgl(
      rule_clauses,
      \(clause) str_detect(instructions, fixed(clause))
    )
  )

knitr::kable(
  adjudication_check |>
    select(sentence_id, question, rule_added, adjudicated_label, uncertainty),
  col.names = c(
    "Unit",
    "Question raised",
    "Rule written in response",
    "Decision",
    "Uncertainty kept"
  ),
  caption = "Four pilot disagreements, four rules, four recorded decisions",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(adjudication_check), 4L),
  all(adjudication_check$matches_file),
  all(adjudication_check$clause_in_codebook),
  identical(annotation_codebook$version, "1.1.0")
)
```

Every rule in that table is present in the current instructions, and the check
above reads them back out of the instrument rather than trusting the
description. Every adjudicated decision matches the archived label. Version
1.0.0 remains on those rows; version 1.1.0 records the corrected unit wording
for future work.

Two details stay visible. Three of the four disputed units retain
`needs_review`; `s010` is marked `certain`, so the file does not pretend every
disagreement remained unresolved. Version 1.1.0 of the codebook now names one
source-text line as the annotation unit. The historical `sentence_id` column
remains a stable identifier, not a claim that every row is a grammatical
sentence.

This pilot shows how disagreement can produce explicit rules. Its designed six
items do not estimate how much a pilot would change a larger annotation job or
how long that work would take.

## Protect the people doing the work

Manual annotation can involve sensitive, repetitive, or distressing material.
Before assigning work:

1. explain the purpose and allowed use of the data;
2. remove personal information that reviewers do not need;
3. pay fairly for reading, training, and disagreement review;
4. give reviewers a way to skip harmful material;
5. separate quality feedback from surveillance; and
6. record which languages and lived experience the task requires.

Expertise depends on the question. A worker, career counselor, employer, and
labor researcher may interpret the same text unit differently. Those
perspectives should be designed into the study rather than treated as noise.

## What to remember

- Annotation labels are measurements created from written rules.
- Name the unit precisely; a heading and a phrase are not sentences.
- One unit per row keeps the decision small and traceable.
- A codebook needs a version, examples, and a change record.
- Run a pilot and let its disagreements write the rules.
- Ambiguity should lead to review, not an unrecorded correction.
- Human labels require ethical working conditions and relevant expertise.

The Lab has defined the categories and preserved its reading decisions. To test
selection without exposing known answers, the active-learning exercise starts
with five complete job postings and hides two other documents from a small
model.

## Sources

- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
- [Datasheets for Datasets](https://doi.org/10.1145/3458723)
- [Data Collection and Quality Challenges in Deep Learning](https://arxiv.org/abs/2002.11821)
- [foundryR codebook documentation](https://farach.github.io/foundryR/reference/foundry_codebook.html)
