---
title: "Choosing which examples to label next"
subtitle: "Use active learning without confusing uncertainty with truth"
description: "Learn how a small model can rank unlabeled workforce sentences for human review."
categories: [training data, active learning, workforce research]
---

A finished codebook covers training, requirements, schedules, skills, and other
workforce text. For this exercise, five complete job
postings form the labeled starting set. A sixth posting and the flyer form an
unlabeled pool. The team could
read the pool in file order. It tests whether a small model can point
to cases that its current evidence does not separate cleanly.

**Active learning** is a loop in which a model helps choose the next items for
people to label. The model does not supply the answer. It points to examples
that meet a selection rule, and a person applies the codebook. Here, that rule
measures uncertainty about `training`, not which sentence matters most to a
worker.

::: {.callout-note}
The sentences and research team are fictional. The model is deliberately small
and exists to make the selection logic visible; it is not suitable for
workforce decisions.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- separate an initial labeled set from an annotation pool;
- explain uncertainty sampling;
- rank items by distance from a 50% model score;
- reveal labels only after selection;
- show that a selected subset moves an estimate either way; and
- name the biases that active learning can introduce.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Hide the answers in the annotation pool

The teaching file contains reviewed labels so the lesson can check its work.
During selection, the model receives labels from documents `J001` through
`J005`. The complete `J006` posting and `F001` flyer form the annotation pool.
Keeping documents intact avoids training on one part of a posting while
selecting another part from the same source. The code below opens files with
readr, handles tables with dplyr and tibble, repeats checks with purrr, and
matches text with stringr.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)

all_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

annotation_view <- all_sentences |>
  select(
    sentence_id,
    document_id,
    text,
    reference_label
  )
initial_labels <- annotation_view |>
  filter(document_id %in% sprintf("J%03d", 1:5))
annotation_pool <- annotation_view |>
  filter(document_id %in% c("J006", "F001")) |>
  select(sentence_id, document_id, text)
held_back_labels <- annotation_view |>
  filter(document_id %in% c("J006", "F001")) |>
  select(sentence_id, reference_label)

knitr::kable(
  annotation_pool,
  col.names = c(
    "Sentence ID",
    "Document ID",
    "Text"
  ),
  caption = "Ten text units waiting for annotation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(initial_labels), 18L),
  identical(nrow(annotation_pool), 10L),
  !("reference_label" %in% names(annotation_pool)),
  identical(
    annotation_pool$sentence_id,
    sprintf("s%03d", 19:28)
  ),
  length(
    intersect(
      initial_labels$document_id,
      annotation_pool$document_id
    )
  ) == 0L
)
```

The labels are stored separately and do not enter the selection calculation.
This prevents the lesson from using an answer that would be unavailable in a
real annotation round.

## Fit a small keyword model

The example uses a **Bernoulli naive Bayes** model. It records whether selected
terms appear, then compares how often those terms occur in training and
non-training sentences. Half-count smoothing prevents an observed term from
receiving a conditional probability of exactly zero or one.

This model is transparent enough to inspect, but its assumptions are strong.
It treats the selected terms as conditionally independent and ignores word
order, context, and most of the sentence.

The candidate terms come from the codebook and labeled examples, not from
reading the held-back pool. The code removes any candidate never observed in
the labeled set. The short form `\(keyword)` means a small function with one
input named `keyword`. Otherwise add-one smoothing could make an unseen term
appear more common in the smaller class merely because its denominator is
smaller.

```{r}
candidate_keywords <- c(
  "training",
  "certificate",
  "apprenticeship",
  "mentor",
  "stipend",
  "classes",
  "required",
  "preferred",
  "license",
  "diploma",
  "portfolio",
  "shift",
  "schedule",
  "skill",
  "spreadsheet",
  "communication",
  "experience"
)

make_features <- function(text, terms) {
  normalized_text <- str_to_lower(text)
  feature_columns <- terms |>
    map(\(keyword) {
      str_detect(
        normalized_text,
        sprintf("\\b%s\\w*\\b", keyword)
      )
    })

  matrix(
    unlist(feature_columns, use.names = FALSE),
    nrow = length(text),
    ncol = length(terms),
    dimnames = list(NULL, terms)
  )
}

candidate_features <- make_features(
  initial_labels$text,
  candidate_keywords
)
keywords <- candidate_keywords[
  colSums(candidate_features) > 0L
]
training_features <- candidate_features[
  ,
  keywords,
  drop = FALSE
]
is_training <- initial_labels$reference_label == "training"

smoothing <- 0.5
training_prior <- (
  sum(is_training) + smoothing
) / (
  length(is_training) + 2 * smoothing
)
training_term_count <- colSums(
  training_features[is_training, , drop = FALSE]
)
other_term_count <- colSums(
  training_features[!is_training, , drop = FALSE]
)
empirical_rate_training <- training_term_count /
  sum(is_training)
empirical_rate_other <- other_term_count /
  sum(!is_training)
term_rate_training <- (
  training_term_count + smoothing
) / (
  sum(is_training) + 2 * smoothing
)
term_rate_other <- (
  other_term_count + smoothing
) / (
  sum(!is_training) + 2 * smoothing
)
direction_is_preserved <- sign(
  term_rate_training - term_rate_other
) == sign(
  empirical_rate_training -
    empirical_rate_other
)

score_training <- function(text) {
  features <- make_features(text, keywords)
  log_odds <- log(
    training_prior / (1 - training_prior)
  ) +
    rowSums(
      sweep(
        features,
        2,
        log(term_rate_training / term_rate_other),
        "*"
      ) +
        sweep(
          !features,
          2,
          log(
            (1 - term_rate_training) /
              (1 - term_rate_other)
          ),
          "*"
        )
    )

  stats::plogis(log_odds)
}

training_scores <- score_training(annotation_pool$text)

knitr::kable(
  tibble(
    term = keywords,
    training_count = as.integer(training_term_count),
    other_count = as.integer(other_term_count),
    smoothed_training_rate = round(
      term_rate_training,
      3
    ),
    smoothed_other_rate = round(
      term_rate_other,
      3
    )
  ),
  col.names = c(
    "Term",
    "Training rows",
    "Other rows",
    "Smoothed training rate",
    "Smoothed other rate"
  ),
  caption = "Term evidence from the 18 labeled sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    keywords,
    c(
      "training",
      "certificate",
      "apprenticeship",
      "mentor",
      "required",
      "preferred",
      "license",
      "diploma",
      "portfolio",
      "shift",
      "schedule",
      "skill",
      "spreadsheet",
      "experience"
    )
  ),
  identical(length(training_scores), 10L),
  all(direction_is_preserved),
  all(training_scores > 0),
  all(training_scores < 1)
)
```

The half-count smoothing keeps probabilities away from zero and one. The
assertion also requires each smoothed difference to point in the same direction
as the observed difference. A term never seen in a training row cannot become
positive evidence for training merely because that class is smaller.

The score used next is the model's estimate under this small dataset and model.
It is not a validated probability that the label is correct.

## Select the closest scores

**Uncertainty sampling** often selects scores closest to 0.5 in a binary task.
The model is least decisive there. We calculate each score's distance from 0.5
and choose the two smallest distances. The model vocabulary contains only
terms observed in the labeled documents. Words found only in the pool cannot
acquire artificial evidence through smoothing.

```{r}
selection_table <- tibble(
  sentence_id = annotation_pool$sentence_id,
  text = annotation_pool$text,
  training_score = training_scores,
  distance_from_half = abs(
    training_scores - 0.5
  )
) |>
  arrange(distance_from_half, sentence_id)
selected_for_review <- selection_table |>
  slice_head(n = 2)

knitr::kable(
  selection_table |>
    mutate(
      training_score = round(training_score, 3),
      distance_from_half = round(
        distance_from_half,
        3
      )
    ),
  col.names = c(
    "Sentence ID",
    "Text",
    "Training score",
    "Distance from 0.5"
  ),
  caption = "Annotation candidates ordered by model uncertainty",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    selected_for_review$sentence_id,
    c("s026", "s028")
  ),
  all(diff(selection_table$distance_from_half) >= 0)
)
```

The first two rows are the next annotation candidates. Neither contains a
feature seen in the labeled set, so both receive the same score. This is a
useful failure of uncertainty-only selection: a score can reflect missing
vocabulary rather than a subtle boundary between labels.

## People still assign the labels

Only after selection do we join the held-back reference labels. This simulates
the annotator returning completed work.

```{r}
reviewed_selection <- selected_for_review |>
  select(sentence_id, text, training_score) |>
  left_join(
    held_back_labels,
    by = join_by(sentence_id),
    multiple = "error"
  )

newly_labeled <- selected_for_review |>
  select(sentence_id) |>
  left_join(
    annotation_pool,
    by = join_by(sentence_id),
    multiple = "error"
  ) |>
  left_join(
    held_back_labels,
    by = join_by(sentence_id),
    multiple = "error"
  )
updated_labels <- bind_rows(
  initial_labels,
  newly_labeled
)

knitr::kable(
  reviewed_selection,
  col.names = c(
    "Sentence ID",
    "Text",
    "Training score",
    "Human-applied label"
  ),
  caption = "The two selected sentences after annotation",
  digits = 3,
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    reviewed_selection$reference_label,
    c("schedule", "other")
  ),
  identical(nrow(updated_labels), 20L)
)
```

The two sentences receive `schedule` and `other`. Human review adds distinctions
that the binary training detector cannot express. The tie shows why a practical
selection policy may combine uncertainty with diversity and explicit coverage
goals.

## Uncertainty is only one selection rule

A model can be confidently wrong. Selecting only uncertain examples may also
overlook:

- communities or languages absent from the initial labels;
- rare cases the model scores with false confidence;
- duplicate examples that consume the annotation budget;
- important harms that occur infrequently; and
- examples far from the model's training distribution.

A stronger plan combines uncertainty with random audit samples, diversity,
coverage targets, and deliberate tests of known failure modes. Keep a
document-disjoint evaluation set that active learning never uses for feature
design, selection, or training.

## Selection changes the estimate, and not in a fixed direction

One consequence is easy to state backwards. A set of examples chosen because
the model found them hard is not a picture of the whole pool, so a number
measured on it is not a pool-level number. The tempting next sentence, that the
measured error therefore looks worse than the truth, is not a rule. It depends
on what the rule selected.

The comparison below measures one quantity, the gap between the model's score
and the label a person applied, on three sets of items: the two the uncertainty
rule chose, the two the model was most decisive about, and the whole pool.

```{r}
pool_outcomes <- annotation_pool |>
  left_join(
    held_back_labels,
    by = join_by(sentence_id),
    multiple = "error"
  ) |>
  mutate(
    training_score = training_scores,
    is_training = as.integer(reference_label == "training"),
    score_gap = abs(training_score - is_training),
    predicted_training = training_score > 0.5
  )

most_decisive <- selection_table |>
  arrange(desc(distance_from_half), sentence_id) |>
  slice_head(n = 2)

gap_for <- function(ids) {
  mean(
    pool_outcomes$score_gap[pool_outcomes$sentence_id %in% ids]
  )
}

selection_comparison <- tibble(
  evaluated_on = c(
    "two items the uncertainty rule chose",
    "two items the model was most decisive about",
    "all ten pool items"
  ),
  items = c(2L, 2L, nrow(pool_outcomes)),
  mean_score_gap = c(
    gap_for(selected_for_review$sentence_id),
    gap_for(most_decisive$sentence_id),
    gap_for(pool_outcomes$sentence_id)
  )
)

pool_accuracy <- mean(
  pool_outcomes$predicted_training ==
    (pool_outcomes$is_training == 1L)
)

knitr::kable(
  selection_comparison |>
    mutate(mean_score_gap = round(mean_score_gap, 3)),
  col.names = c(
    "Measured on",
    "Items",
    "Mean gap between score and label"
  ),
  caption = "One model, one pool, three ways of choosing what to measure",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(pool_outcomes), 10L),
  identical(
    most_decisive$sentence_id,
    c("s019", "s027")
  ),
  identical(
    round(selection_comparison$mean_score_gap, 3),
    c(0.194, 0.019, 0.112)
  ),
  selection_comparison$mean_score_gap[1] >
    selection_comparison$mean_score_gap[3],
  selection_comparison$mean_score_gap[2] <
    selection_comparison$mean_score_gap[3],
  isTRUE(all.equal(pool_accuracy, 1))
)
```

The pool value is 0.112. Selecting by uncertainty raised the measured gap to
0.194. Selecting by confidence lowered it to 0.019. Same model, same ten items,
same quantity, and the estimate moved in opposite directions because the
selection rule differed.

Neither direction is a law. A rule that finds hard cases usually flatters the
data and punishes the score; a rule that finds easy cases does the reverse; a
rule that happens to select a class the model handles well can make a weak model
look strong. The more defensible statement is also more useful: any estimate
computed on a selected subset describes that subset.

The remedy has not changed. Keep a fixed evaluation set drawn at random,
disjoint by document from anything active learning touches, and report pool-level
numbers only from it. In this exercise the classifier also happens to place all
ten pool items on the correct side of 0.5, which is worth stating plainly and
worth trusting very little: ten items, a vocabulary built from the labeled
documents, and no independent test set.

## What to remember

- Active learning chooses what people label next; it does not replace them.
- Hide pool labels until after selection.
- A score near 0.5 is uncertain only under the current model.
- Confidence is not the same as correctness or social importance.
- A number measured on selected items can land above or below the pool value.
- Audit random, rare, and underrepresented cases alongside uncertainty.

The selection trail accounts for two new reviews, but it does not broaden what
the fictional notices cover. Selection cannot repair source coverage. The Lab
therefore inspects three outside resources and asks whether any can answer the
same questions before their records are combined.

## Sources

- [Active Learning Literature Survey](https://minds.wisconsin.edu/handle/1793/60660)
- [A Sequential Algorithm for Training Text Classifiers](https://doi.org/10.1145/188490.188495)
- [A Survey of Active Learning for Text Classification](https://arxiv.org/abs/2007.14003)
- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
