---
title: "Working with several annotators"
subtitle: "Measure disagreement before combining labels"
author: "Alex Farach"
description: "Learn how to compare workforce labels from several reviewers without treating majority vote as truth."
categories: [training data, crowdsourcing, agreement]
---

The outside provider review kept unlike records separate. The Riverton
Workforce Lab now checks whether the people assigning labels make comparable
decisions. Three fictional paid reviewers apply the same codebook independently
to 11 local sentences. Nine rows match; two do not.

**Crowdsourcing** distributes tasks to a group of people, often through an
online marketplace. Access to more reviewers can increase coverage, but a
marketplace does not create expertise, fair working conditions, or valid
labels. The team must design those conditions.

::: {.callout-note}
The reviewers, annotations, and research team are fictional. The disagreements
were designed for this lesson and do not evaluate real workers.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- keep reviewer labels separate;
- calculate pairwise agreement with foundryR;
- distinguish observed agreement from chance-corrected measures;
- test how much one item moves an agreement statistic;
- find items that need adjudication and record the decision; and
- explain why majority vote is not automatic truth.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Preserve each person's answer

The annotation file stores one column per reviewer. The next chunk loads file,
table, reshaping, repeated-work, agreement, and reliability helpers, then reads
the CSV files with declared column types. No vote has been combined or
overwritten.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(foundryR)
library(irr)
library(stringr)

crowd_labels <- read_csv(
  "data/workforce/crowd_annotations.csv",
  col_types = cols(
    sentence_id = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    reviewer_a = col_character(),
    reviewer_a_uncertainty = col_character(),
    reviewer_a_rationale = col_character(),
    reviewer_b = col_character(),
    reviewer_b_uncertainty = col_character(),
    reviewer_b_rationale = col_character(),
    reviewer_c = col_character(),
    reviewer_c_uncertainty = col_character(),
    reviewer_c_rationale = col_character()
  )
)
sentence_text <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

reviewer_columns <- c(
  "reviewer_a",
  "reviewer_b",
  "reviewer_c"
)
uncertainty_columns <- paste0(
  reviewer_columns,
  "_uncertainty"
)
rationale_columns <- paste0(
  reviewer_columns,
  "_rationale"
)
allowed_labels <- c(
  "training",
  "requirement",
  "schedule",
  "skill",
  "other"
)
source("R/workforce-codebook.R")
crowd_codebook <- workforce_codebook()

stopifnot(
  identical(nrow(crowd_labels), 11L),
  inherits(crowd_labels, "tbl_df"),
  inherits(sentence_text, "tbl_df"),
  all(reviewer_columns %in% names(crowd_labels)),
  all(
    crowd_labels |>
      select(all_of(reviewer_columns)) |>
      unlist(use.names = FALSE) %in% allowed_labels
  ),
  all(
    crowd_labels |>
      select(all_of(uncertainty_columns)) |>
      unlist(use.names = FALSE) %in% c("certain", "needs_review")
  ),
  all(
    str_length(
      crowd_labels |>
        select(all_of(rationale_columns)) |>
        unlist(use.names = FALSE)
    ) > 0
  ),
  !anyDuplicated(crowd_labels$sentence_id),
  all(crowd_labels$codebook_version == crowd_codebook$version),
  all(crowd_labels$codebook_hash == crowd_codebook$hash)
)

knitr::kable(
  crowd_labels |>
    select(sentence_id, all_of(reviewer_columns)),
  col.names = c(
    "Sentence ID",
    "Reviewer A",
    "Reviewer B",
    "Reviewer C"
  ),
  caption = "Independent labels from three fictional reviewers",
  row.names = FALSE
)
```

Nine rows have complete agreement. Sentence `s010` has one different answer.
Sentence `s013` has three different answers. Keeping the columns separate
makes those disagreements visible, and the sample includes all five labels.

## Measure all reviewer comparisons

Simple agreement is the share of answers that match. **Cohen's kappa** adjusts
for the agreement expected from the reviewers' label frequencies.
**Krippendorff's alpha** is another chance-corrected measure that can support
more coders and missing labels under suitable settings.

foundryR calculates several measures together. Here we retain observed
agreement and Cohen's kappa for every reviewer pair. Directional classification
measures such as precision and recall are not shown because no reviewer is
treated as the truth.

```{r}
reviewer_pairs <- tibble(
  pair = c("A and B", "A and C", "B and C"),
  first = c(
    "reviewer_a",
    "reviewer_a",
    "reviewer_b"
  ),
  second = c(
    "reviewer_b",
    "reviewer_c",
    "reviewer_c"
  )
)

pairwise_agreement <- pmap(
  reviewer_pairs,
  \(pair, first, second) {
    measures <- foundryR::foundry_agreement(
      crowd_labels,
      estimate = second,
      truth = first
    )

    tibble(
      pair = pair,
      observed_agreement = filter(measures, metric == "accuracy")$value,
      cohen_kappa = filter(measures, metric == "cohen_kappa")$value
    )
  }
) |>
  list_rbind()

stopifnot(
  identical(
    round(pairwise_agreement$observed_agreement, 3),
    c(0.818, 0.909, 0.818)
  ),
  identical(
    round(pairwise_agreement$cohen_kappa, 3),
    c(0.756, 0.875, 0.758)
  )
)

knitr::kable(
  pairwise_agreement |>
    mutate(
      observed_agreement = round(
        observed_agreement,
        2
      ),
      cohen_kappa = round(cohen_kappa, 2)
    ),
  col.names = c(
    "Reviewer pair",
    "Observed agreement",
    "Cohen's kappa"
  ),
  caption = "Pairwise agreement across all three reviewers",
  row.names = FALSE
)
```

The B-and-C comparison is weaker than A and C. Reporting only one pair would
hide that difference.

The table shows two decimals on purpose. With 11 items, one reviewer changing
one answer moves observed agreement by about nine points, so a third decimal
describes arithmetic rather than measurement. Publishing 0.818 invites a reader
to compare it with 0.812 from another study as though the difference meant
something.

Krippendorff's alpha summarizes all three reviewers at once. `pivot_longer()`
puts the reviewer columns into one label column, then `pivot_wider()` builds the
matrix that `irr` needs. The calculation converts label names to category
numbers; the numbers carry no order.

```{r}
numeric_ratings <- crowd_labels |>
  select(sentence_id, all_of(reviewer_columns)) |>
  pivot_longer(
    cols = all_of(reviewer_columns),
    names_to = "reviewer",
    values_to = "label"
  ) |>
  mutate(category_number = match(label, allowed_labels)) |>
  select(sentence_id, reviewer, category_number) |>
  pivot_wider(
    names_from = reviewer,
    values_from = category_number
  ) |>
  arrange(sentence_id)

rating_matrix <- numeric_ratings |>
  select(all_of(reviewer_columns)) |>
  as.matrix()
all_rater_alpha <- irr::kripp.alpha(
  t(rating_matrix),
  method = "nominal"
)$value
unanimity_by_sentence <- crowd_labels |>
  select(sentence_id, all_of(reviewer_columns)) |>
  pivot_longer(
    cols = all_of(reviewer_columns),
    names_to = "reviewer",
    values_to = "label"
  ) |>
  summarise(
    unanimous = n_distinct(label) == 1L,
    .by = sentence_id
  ) |>
  arrange(sentence_id)
unanimity_share <- unanimity_by_sentence |>
  summarise(value = mean(unanimous))
unanimity_share <- unanimity_share$value

stopifnot(
  isTRUE(
    all.equal(
      round(all_rater_alpha, 3),
      0.797
    )
  ),
  isTRUE(
    all.equal(
      unanimity_share,
      9 / 11
    )
  )
)

tibble(
  measure = c(
    "Krippendorff's alpha",
    "Unanimous share"
  ),
  value = round(
    c(all_rater_alpha, unanimity_share),
    2
  )
)
```

The 11-item example is too small for a precise population estimate. These
statistics describe this annotation exercise and help locate problems; they do
not certify the codebook or any reviewer.

## Ask how much one item is worth

"Too small for a precise estimate" is a phrase readers skim past. The way to
make it concrete is to remove one item at a time and recompute, which for 11
items means 11 recalculations and no new data.

```{r}
alpha_without_item <- map_dbl(
  seq_len(nrow(rating_matrix)),
  \(dropped) {
    irr::kripp.alpha(
      t(rating_matrix[-dropped, , drop = FALSE]),
      method = "nominal"
    )$value
  }
)

sensitivity <- tibble(
  sentence_id = numeric_ratings$sentence_id,
  alpha_without_this_item = alpha_without_item
) |>
  arrange(desc(alpha_without_this_item))

alpha_range <- range(alpha_without_item)

stopifnot(
  identical(length(alpha_without_item), 11L),
  identical(
    round(alpha_range, 2),
    c(0.76, 0.91)
  ),
  identical(
    sensitivity$sentence_id[1],
    "s013"
  ),
  max(alpha_without_item) > all_rater_alpha,
  min(alpha_without_item) < all_rater_alpha
)

knitr::kable(
  sensitivity |>
    mutate(
      alpha_without_this_item = round(
        alpha_without_this_item,
        2
      )
    ),
  col.names = c("Item removed", "Alpha for the other ten"),
  caption = "Krippendorff's alpha after dropping each item in turn",
  row.names = FALSE
)
```

Alpha is 0.80 on all 11 items. Drop one item and it lands anywhere between 0.76
and 0.91. The high end is not a coincidence: removing `s013`, the item all three
reviewers read differently, is what produces it.

This is not a confidence interval, and it should not be reported as one. It is a
cruder statement and a sufficient one here: a summary this sensitive to a single
item cannot support a claim that one codebook, task, or reviewer pool is more
reliable than another. Reporting the range beside the estimate is honest;
reporting 0.797 alone is not.

Larger samples do not remove the problem, they shrink it. The same jackknife on
500 items would move alpha in the third decimal, at which point an interval from
a bootstrap is the right tool.

## Find the disagreements

An item needs review when the three labels are not identical. `left_join()`
adds the sentence text by matching the shared sentence identifier.

```{r}
disagreements <- crowd_labels |>
  left_join(
    unanimity_by_sentence,
    by = join_by(sentence_id)
  ) |>
  filter(!unanimous) |>
  left_join(
    sentence_text |>
      select(sentence_id, text),
    by = join_by(sentence_id)
  ) |>
  arrange(sentence_id)

stopifnot(
  identical(
    disagreements$sentence_id,
    c("s010", "s013")
  ),
  identical(nrow(disagreements), 2L)
)

knitr::kable(
  disagreements |>
    select(
      "sentence_id",
      "text",
      "reviewer_a",
      "reviewer_a_uncertainty",
      "reviewer_b",
      "reviewer_b_uncertainty",
      "reviewer_c",
      "reviewer_c_uncertainty"
    ),
  col.names = c(
    "Sentence ID",
    "Text",
    "Reviewer A",
    "A uncertainty",
    "Reviewer B",
    "B uncertainty",
    "Reviewer C",
    "C uncertainty"
  ),
  caption = "Sentences sent to adjudication",
  row.names = FALSE
)
```

The first sentence says experience is preferred; one reviewer chose `other`.
The second mentions a certificate and receives three different labels. Those
choices place the same notices under different questions for a person comparing
options. The codebook needs to explain the boundary rather than hide it in a
vote.

## Calculate a majority label carefully

A majority label can organize the next review step. It should not erase the
individual answers or replace adjudication.

```{r}
label_counts <- crowd_labels |>
  select(sentence_id, all_of(reviewer_columns)) |>
  pivot_longer(
    cols = all_of(reviewer_columns),
    names_to = "reviewer",
    values_to = "label"
  ) |>
  count(sentence_id, label, name = "votes")

top_vote_count <- label_counts |>
  summarise(top_votes = max(votes), .by = sentence_id)
top_vote_table <- label_counts |>
  left_join(top_vote_count, by = join_by(sentence_id)) |>
  filter(votes == top_votes) |>
  summarise(
    tied_top_labels = n(),
    top_label = paste(sort(label), collapse = ", "),
    .by = c(sentence_id, top_votes)
  )
majority_label <- top_vote_table |>
  mutate(
    majority_label = case_when(
      top_votes > length(reviewer_columns) / 2 &
        tied_top_labels == 1L ~ top_label,
      TRUE ~ "adjudicate"
    )
  ) |>
  select(sentence_id, majority_label)

combined_labels <- crowd_labels |>
  select(sentence_id) |>
  left_join(
    majority_label,
    by = join_by(sentence_id)
  ) |>
  left_join(
    unanimity_by_sentence,
    by = join_by(sentence_id)
  )

stopifnot(
  identical(
    combined_labels$majority_label,
    c(
      "training",
      "requirement",
      "schedule",
      "skill",
      "requirement",
      "training",
      "requirement",
      "other",
      "adjudicate",
      "training",
      "requirement"
    )
  ),
  identical(sum(combined_labels$unanimous), 9L)
)

knitr::kable(
  top_vote_table,
  col.names = c(
    "Sentence ID",
    "Top votes",
    "Labels tied for top",
    "Top label or labels"
  ),
  caption = "Top vote count before majority assignment",
  row.names = FALSE
)

knitr::kable(
  combined_labels,
  col.names = c(
    "Sentence ID",
    "Majority label",
    "Unanimous"
  ),
  caption = "Majority labels with unanimity retained",
  row.names = FALSE
)
```

The `unanimous` column prevents an easy majority from looking like complete
agreement. The tie check stays because teams may change the reviewer count; any
tie among top labels goes to `adjudicate`. Sentence `s013` returns `adjudicate`
because three different answers do not form a majority. The team can review the
sentence, codebook, and source labels before choosing a reference label.

## Finish the adjudication

An adjudication queue that stays a queue is not a quality process. Both flagged
items were decided by a fourth reader who had not labeled them, working from the
codebook text rather than from the vote counts.

```{r}
adjudication <- tibble(
  sentence_id = c("s010", "s013"),
  question = c(
    "Is a preferred condition an entry condition or something else?",
    "Does naming a credential make the unit a training offer?"
  ),
  rule_applied = c(
    "requirement: the main claim states or prefers an entry condition",
    "training: a credential required or preferred is not training"
  ),
  decision = c("requirement", "requirement"),
  decided_by = "fourth reader, not one of the three reviewers"
)

adjudication_outcome <- adjudication |>
  left_join(
    sentence_text |>
      select(sentence_id, text, reference_label),
    by = join_by(sentence_id)
  ) |>
  left_join(
    crowd_labels |>
      select(sentence_id, all_of(reviewer_columns)),
    by = join_by(sentence_id)
  ) |>
  mutate(
    matches_reference_file = decision == reference_label,
    reviewers_still_disagree = reviewer_a != reviewer_b |
      reviewer_b != reviewer_c
  )

stopifnot(
  identical(nrow(adjudication_outcome), 2L),
  all(adjudication_outcome$matches_reference_file),
  all(adjudication_outcome$reviewers_still_disagree),
  str_detect(
    crowd_codebook$instructions,
    fixed("still describes an entry condition")
  ),
  str_detect(
    crowd_codebook$instructions,
    fixed("is not training")
  ),
  identical(nrow(crowd_labels), 11L)
)

knitr::kable(
  adjudication_outcome |>
    select(
      sentence_id,
      text,
      reviewer_a,
      reviewer_b,
      reviewer_c,
      decision
    ),
  col.names = c(
    "Sentence ID",
    "Text",
    "Reviewer A",
    "Reviewer B",
    "Reviewer C",
    "Adjudicated label"
  ),
  caption = "Both flagged items resolved, with the reviewer answers kept",
  row.names = FALSE
)
```

Both items resolve to `requirement`, and both decisions rest on a clause the
adjudicator can point to in the codebook. The reviewer columns are still in the
table with their disagreements intact, which is why the check above can confirm
that the reviewers still differ after the item was settled.

Notice what the adjudication of `s013` did not do. It did not make the item
easy. Three trained readers reached three answers, and the sensitivity analysis
showed that this one item carries most of the distance between an alpha of 0.80
and one of 0.91. A decision closes the item; it does not retire the question,
which belongs in the next codebook revision as a worked example.

## Agreement does not prove validity

Reviewers can agree because:

- the codebook is clear;
- the item is easy;
- they share the same misunderstanding;
- the task excludes a relevant perspective; or
- the platform trained everyone toward the same shortcut.

Validity requires evidence that the labels measure the intended idea. Compare
annotations with the research question, consult people with relevant knowledge,
and test how conclusions change under reasonable label choices.

## Treat crowd work as work

A responsible plan states:

1. payment for training, annotation, and adjudication time;
2. expected task length and rejection rules;
3. privacy and data-handling requirements;
4. access to support and the ability to skip harmful material;
5. the expertise and language knowledge required; and
6. an appeal path for disputed quality decisions.

Hidden attention checks and automatic rejection can shift measurement errors
onto workers. Quality control should diagnose the task and instructions as well
as individual answers.

## What to remember

- Keep individual reviewer labels before combining them.
- Agreement measures summarize consistency, not truth.
- Report agreement to a precision the sample size can support.
- Show how far a statistic moves when one item is removed.
- Inspect disagreement items and revise the codebook when needed.
- Finish the adjudication and record which rule decided it.
- Preserve whether a majority label was unanimous.
- Fair pay, privacy, expertise, and appeal processes are part of data quality.

The comparison leaves nothing in the queue: two boundary cases were adjudicated
against named rules, and the reviewer answers that produced them are still on
the page. The following augmentation exercise keeps that annotation record
intact and creates fictional variants. Each is marked as synthetic and checked
against the same codebook before it can join training data.

## Sources

- [Inter-Coder Agreement for Computational Linguistics](https://aclanthology.org/J08-4004/)
- [Krippendorff's alpha](https://repository.upenn.edu/asc_papers/43/)
- [Cheap and Fast, but Is It Good?](https://aclanthology.org/D08-1027/)
- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
- [foundryR agreement documentation](https://farach.github.io/foundryR/reference/foundry_agreement.html)
