---
title: "Working with several annotators"
subtitle: "Measure disagreement before combining labels"
author: "Alex Farach"
description: "Learn how to compare workforce labels from several reviewers without treating majority vote as truth."
categories: [training data, crowdsourcing, agreement]
---

The outside provider review kept unlike records separate. The Riverton
Workforce Lab now checks whether the people assigning labels make comparable
decisions. Three fictional paid reviewers apply the same codebook independently
to 11 local sentences. Nine rows match; two do not.

**Crowdsourcing** distributes tasks to a group of people, often through an
online marketplace. Access to more reviewers can increase coverage, but a
marketplace does not create expertise, fair working conditions, or valid
labels. The team must design those conditions.

::: {.callout-note}
The reviewers, annotations, and research team are fictional. The disagreements
were designed for this lesson and do not evaluate real workers.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- keep reviewer labels separate;
- calculate pairwise agreement with foundryR;
- distinguish observed agreement from chance-corrected measures;
- find items that need adjudication; and
- explain why majority vote is not automatic truth.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Preserve each person's answer

The annotation file stores one column per reviewer. No vote has been combined
or overwritten.

```{r}
crowd_labels <- read.csv(
  "data/workforce/crowd_annotations.csv",
  colClasses = "character",
  check.names = FALSE
)
sentence_text <- read.csv(
  "data/workforce/workforce_sentences.csv",
  colClasses = "character",
  check.names = FALSE
)

reviewer_columns <- c(
  "reviewer_a",
  "reviewer_b",
  "reviewer_c"
)
uncertainty_columns <- paste0(
  reviewer_columns,
  "_uncertainty"
)
rationale_columns <- paste0(
  reviewer_columns,
  "_rationale"
)
allowed_labels <- c(
  "training",
  "requirement",
  "schedule",
  "skill",
  "other"
)
source("R/workforce-codebook.R")
crowd_codebook <- workforce_codebook()

stopifnot(
  identical(nrow(crowd_labels), 11L),
  all(reviewer_columns %in% names(crowd_labels)),
  all(
    unlist(
      crowd_labels[reviewer_columns],
      use.names = FALSE
    ) %in% allowed_labels
  ),
  all(
    unlist(
      crowd_labels[uncertainty_columns],
      use.names = FALSE
    ) %in% c("certain", "needs_review")
  ),
  all(
    nzchar(
      unlist(
        crowd_labels[rationale_columns],
        use.names = FALSE
      )
    )
  ),
  !anyDuplicated(crowd_labels$sentence_id),
  all(
    crowd_labels$codebook_version ==
      crowd_codebook$version
  ),
  all(
    crowd_labels$codebook_hash ==
      crowd_codebook$hash
  )
)

knitr::kable(
  crowd_labels[
    c("sentence_id", reviewer_columns)
  ],
  col.names = c(
    "Sentence ID",
    "Reviewer A",
    "Reviewer B",
    "Reviewer C"
  ),
  caption = "Independent labels from three fictional reviewers",
  row.names = FALSE
)
```

Nine rows have complete agreement. Sentence `s010` has one different answer.
Sentence `s013` has three different answers. Keeping the columns separate
makes those disagreements visible, and the sample includes all five labels.

## Measure all reviewer comparisons

Simple agreement is the share of answers that match. **Cohen's kappa** adjusts
for the agreement expected from the reviewers' label frequencies.
**Krippendorff's alpha** is another chance-corrected measure that can support
more coders and missing labels under suitable settings.

foundryR calculates several measures together. Here we retain observed
agreement and Cohen's kappa for every reviewer pair. Directional classification
measures such as precision and recall are not shown because no reviewer is
treated as the truth.

```{r}
reviewer_pairs <- data.frame(
  pair = c("A and B", "A and C", "B and C"),
  first = c(
    "reviewer_a",
    "reviewer_a",
    "reviewer_b"
  ),
  second = c(
    "reviewer_b",
    "reviewer_c",
    "reviewer_c"
  ),
  stringsAsFactors = FALSE
)

pairwise_agreement <- do.call(
  rbind,
  lapply(
    seq_len(nrow(reviewer_pairs)),
    function(pair_number) {
      measures <- foundryR::foundry_agreement(
        crowd_labels,
        estimate = reviewer_pairs$second[
          pair_number
        ],
        truth = reviewer_pairs$first[
          pair_number
        ]
      )

      data.frame(
        pair = reviewer_pairs$pair[
          pair_number
        ],
        observed_agreement = measures$value[
          measures$metric == "accuracy"
        ],
        cohen_kappa = measures$value[
          measures$metric == "cohen_kappa"
        ]
      )
    }
  )
)

stopifnot(
  identical(
    round(pairwise_agreement$observed_agreement, 3),
    c(0.818, 0.909, 0.818)
  ),
  identical(
    round(pairwise_agreement$cohen_kappa, 3),
    c(0.756, 0.875, 0.758)
  )
)

knitr::kable(
  transform(
    pairwise_agreement,
    observed_agreement = round(
      observed_agreement,
      3
    ),
    cohen_kappa = round(cohen_kappa, 3)
  ),
  col.names = c(
    "Reviewer pair",
    "Observed agreement",
    "Cohen's kappa"
  ),
  caption = "Pairwise agreement across all three reviewers",
  row.names = FALSE
)
```

The B-and-C comparison is weaker than A and C. Reporting only one pair would
hide that difference.

Krippendorff's alpha summarizes all three reviewers at once. The calculation
converts label names to category numbers; the numbers carry no order.

```{r}
numeric_ratings <- vapply(
  crowd_labels[reviewer_columns],
  function(labels) {
    match(labels, allowed_labels)
  },
  integer(nrow(crowd_labels))
)
all_rater_alpha <- irr::kripp.alpha(
  t(numeric_ratings),
  method = "nominal"
)$value
unanimity_share <- mean(
  apply(
    crowd_labels[reviewer_columns],
    1,
    function(labels) {
      length(unique(labels)) == 1L
    }
  )
)

stopifnot(
  isTRUE(
    all.equal(
      round(all_rater_alpha, 3),
      0.797
    )
  ),
  isTRUE(
    all.equal(
      unanimity_share,
      9 / 11
    )
  )
)

data.frame(
  measure = c(
    "Krippendorff's alpha",
    "Unanimous share"
  ),
  value = round(
    c(all_rater_alpha, unanimity_share),
    3
  )
)
```

The 11-item example is too small for a precise population estimate. These
statistics describe this annotation exercise and help locate problems; they do
not certify the codebook or any reviewer.

## Find the disagreements

An item needs review when the three labels are not identical.

```{r}
all_agree <- apply(
  crowd_labels[reviewer_columns],
  1,
  function(labels) {
    length(unique(labels)) == 1L
  }
)

disagreements <- crowd_labels[!all_agree, ]
disagreements <- merge(
  disagreements,
  sentence_text[c("sentence_id", "text")],
  by = "sentence_id",
  sort = FALSE
)

stopifnot(
  identical(
    disagreements$sentence_id,
    c("s010", "s013")
  ),
  identical(nrow(disagreements), 2L)
)

knitr::kable(
  disagreements[
    c(
      "sentence_id",
      "text",
      "reviewer_a",
      "reviewer_a_uncertainty",
      "reviewer_b",
      "reviewer_b_uncertainty",
      "reviewer_c",
      "reviewer_c_uncertainty"
    )
  ],
  col.names = c(
    "Sentence ID",
    "Text",
    "Reviewer A",
    "A uncertainty",
    "Reviewer B",
    "B uncertainty",
    "Reviewer C",
    "C uncertainty"
  ),
  caption = "Sentences sent to adjudication",
  row.names = FALSE
)
```

The first sentence says experience is preferred; one reviewer chose `other`.
The second mentions a certificate and receives three different labels. Those
choices place the same notices under different questions for a person comparing
options. The codebook needs to explain the boundary rather than hide it in a
vote.

## Calculate a majority label carefully

A majority label can organize the next review step. It should not erase the
individual answers or replace adjudication.

```{r}
majority_label <- apply(
  crowd_labels[reviewer_columns],
  1,
  function(labels) {
    counts <- table(
      factor(
        labels,
        levels = allowed_labels
      )
    )
    if (max(counts) > length(labels) / 2) {
      names(counts)[which.max(counts)]
    } else {
      "adjudicate"
    }
  }
)

combined_labels <- data.frame(
  sentence_id = crowd_labels$sentence_id,
  majority_label = majority_label,
  unanimous = all_agree,
  stringsAsFactors = FALSE
)

stopifnot(
  identical(
    combined_labels$majority_label,
    c(
      "training",
      "requirement",
      "schedule",
      "skill",
      "requirement",
      "training",
      "requirement",
      "other",
      "adjudicate",
      "training",
      "requirement"
    )
  ),
  identical(sum(combined_labels$unanimous), 9L)
)

knitr::kable(
  combined_labels,
  col.names = c(
    "Sentence ID",
    "Majority label",
    "Unanimous"
  ),
  caption = "Majority labels with unanimity retained",
  row.names = FALSE
)
```

The `unanimous` column prevents an easy majority from looking like complete
agreement. Sentence `s013` returns `adjudicate` because three different answers
do not form a majority. The team can review the sentence, codebook, and original
labels before choosing a reference label.

## Agreement does not prove validity

Reviewers can agree because:

- the codebook is clear;
- the item is easy;
- they share the same misunderstanding;
- the task excludes a relevant perspective; or
- the platform trained everyone toward the same shortcut.

Validity requires evidence that the labels measure the intended idea. Compare
annotations with the research question, consult people with relevant knowledge,
and test how conclusions change under reasonable label choices.

## Treat crowd work as work

A responsible plan states:

1. payment for training, annotation, and adjudication time;
2. expected task length and rejection rules;
3. privacy and data-handling requirements;
4. access to support and the ability to skip harmful material;
5. the expertise and language knowledge required; and
6. an appeal path for disputed quality decisions.

Hidden attention checks and automatic rejection can shift measurement errors
onto workers. Quality control should diagnose the task and instructions as well
as individual answers.

## What to remember

- Keep individual reviewer labels before combining them.
- Agreement measures summarize consistency, not truth.
- Inspect disagreement items and revise the codebook when needed.
- Preserve whether a majority label was unanimous.
- Fair pay, privacy, expertise, and appeal processes are part of data quality.

The comparison leaves two boundary cases for adjudication instead of folding
them into a majority vote. The following augmentation exercise keeps that
annotation record intact and creates five separate fictional variants. Each is
marked as synthetic and checked against the same codebook before it can join
training data.

## Sources

- [Inter-Coder Agreement for Computational Linguistics](https://aclanthology.org/J08-4004/)
- [Krippendorff's alpha](https://repository.upenn.edu/asc_papers/43/)
- [Cheap and Fast, but Is It Good?](https://aclanthology.org/D08-1027/)
- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
- [foundryR agreement documentation](https://farach.github.io/foundryR/reference/foundry_agreement.html)
