---
title: "Creating additional training examples"
subtitle: "Augment text without assuming the label survived"
author: "Alex Farach"
description: "Learn how to create workforce text variants and test whether they still support the intended label."
categories: [training data, augmentation, workforce research]
---

The reviewer comparison left two boundary cases visible rather than treating a
majority as truth. A different risk appears when the Riverton Workforce Lab
creates synthetic examples: the row count can grow without adding observed
evidence. The team can write additional sentences that follow known patterns.
More rows may help a model learn, but a changed sentence can lose the meaning
that justified its label. For this exercise, the team writes five fictional
variants about training and schedules. They cannot tell a worker that a real
course meets at a given time or provides paid training.

**Textual data augmentation** creates new training examples by transforming or
generating text from existing examples. Augmentation changes a dataset; it does
not collect new evidence about the labor market.

::: {.callout-note}
All sentences and organizations in this lesson are fictional. Augmented rows
are synthetic and remain marked as such.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- distinguish a new text variant from a new observation;
- create controlled template variants;
- use cmapr job titles without importing unsupported claims;
- prepare a classification-check payload with huggingfaceR; and
- test duplication, label balance, and meaning-changing edits.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Start with a narrow purpose

The team wants more examples of explicit training offers. It will not claim
that the examples describe real employers or estimate how often training is
offered. The next chunk loads package helpers for files, tables, strings, CMap
examples, and model-check payloads, then reads the file with declared column
types into a tibble.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)
library(cmapr)
library(huggingfaceR)

workforce_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)
training_examples <- workforce_sentences |>
  filter(reference_label == "training")

stopifnot(
  identical(nrow(training_examples), 7L),
  inherits(workforce_sentences, "tbl_df"),
  all(training_examples$reference_label == "training")
)

knitr::kable(
  training_examples |>
    select(sentence_id, text),
  col.names = c("Sentence ID", "Reviewed training example"),
  caption = "The seven training examples before augmentation",
  row.names = FALSE
)
```

The source labels were assigned under the manual annotation codebook. The
augmentation step must preserve that label definition.

## Use job titles as template values

cmapr includes a ten-row teaching fixture with job-title transitions. We use
three titles as vocabulary only. We do not carry over transition weights or
interpret a title change as a promotion.

```{r}
cmap_titles <- read_csv(
  cmap_example_data(),
  col_types = cols(
    sector = col_character(),
    region = col_character(),
    job_title_from = col_character(),
    job_title_to = col_character(),
    transition_weighted_count = col_double(),
    si_from = col_double(),
    si_to = col_double()
  )
)
selected_titles <- cmap_titles$job_title_from[c(3L, 5L, 8L)]

stopifnot(
  identical(
    selected_titles,
    c(
      "Data Analyst",
      "Medical Assistant",
      "Junior Developer"
    )
  )
)

selected_titles
```

The titles come from the cmapr package's teaching fixture. That file does not
contain upstream record identifiers, so this lesson does not claim that the
three strings are authenticated extracts from the full CMap dataset. The
sentences created next are fictional.

## Create controlled variants

The template states the same kind of claim for each title.

```{r}
source("R/workforce-codebook.R")
augmentation_codebook <- workforce_codebook()

template_variants <- tibble(
  source_sentence_id = "s001",
  source_family_id = "training-offer-s001",
  template_id = "training-pathway-v1",
  vocabulary_source = "cmapr example_transitions.csv",
  vocabulary_row = c(3L, 5L, 8L),
  vocabulary_commit =
    "eb2d0b3810e07782aab4a81fba1f539133f2ba6a",
  text = sprintf(
    "The %s pathway includes paid training.",
    str_to_lower(selected_titles)
  ),
  label = "training",
  synthetic = TRUE,
  method = "controlled template",
  codebook_version = augmentation_codebook$version,
  codebook_hash = augmentation_codebook$hash,
  review_status = "requires_human_review",
  split = "training_candidate"
)

schedule_variants <- tibble(
  source_sentence_id = "s026",
  source_family_id = "schedule-s026",
  template_id = "class-schedule-v1",
  vocabulary_source = "author-created template",
  vocabulary_row = NA_integer_,
  vocabulary_commit = NA_character_,
  text = sprintf(
    "%s classes meet twice a week.",
    c("Morning", "Weekend")
  ),
  label = "schedule",
  synthetic = TRUE,
  method = "controlled template",
  codebook_version = augmentation_codebook$version,
  codebook_hash = augmentation_codebook$hash,
  review_status = "requires_human_review",
  split = "training_candidate"
)

augmented_rows <- bind_rows(
  template_variants,
  schedule_variants
)
augmentation_label_order <- c(
  "training",
  "requirement",
  "schedule",
  "skill",
  "other"
)
augmentation_counts <- tibble(label = augmentation_label_order) |>
  left_join(
    augmented_rows |>
      count(label, name = "n"),
    by = join_by(label)
  ) |>
  mutate(n = coalesce(n, 0L))
source_families <- augmented_rows |>
  distinct(source_family_id) |>
  arrange(source_family_id)

stopifnot(
  identical(nrow(augmented_rows), 5L),
  !anyDuplicated(augmented_rows$text),
  !any(augmented_rows$text %in% workforce_sentences$text),
  all(augmented_rows$synthetic),
  all(augmented_rows$review_status == "requires_human_review"),
  all(augmented_rows$split == "training_candidate"),
  all(augmented_rows$codebook_hash == augmentation_codebook$hash),
  identical(template_variants$vocabulary_row, c(3L, 5L, 8L)),
  all(
    template_variants$vocabulary_commit ==
      "eb2d0b3810e07782aab4a81fba1f539133f2ba6a"
  ),
  identical(
    source_families$source_family_id,
    c("schedule-s026", "training-offer-s001")
  ),
  identical(augmentation_counts$n, c(3L, 0L, 2L, 0L, 0L))
)

knitr::kable(
  augmented_rows,
  col.names = c(
    "Source sentence",
    "Source family",
    "Template",
    "Vocabulary source",
    "Vocabulary row",
    "Vocabulary commit",
    "Synthetic text",
    "Proposed label",
    "Synthetic",
    "Method",
    "Codebook version",
    "Codebook hash",
    "Review status",
    "Intended split"
  ),
  caption = "Five controlled augmentation candidates",
  row.names = FALSE
)
```

The rows carry their source family, template, second vocabulary source,
package commit when applicable, codebook, synthetic status, intended split,
and pending review status. Any evaluation example derived from `s001` or `s026`
must stay outside the evaluation set with its whole family. The five rows
remain candidates until a person checks that each is natural, useful, and
consistent with the codebook.

## Prepare a model check without claiming a result

huggingfaceR can prepare the payload for a zero-shot classification request.
This code records the candidate labels and texts, but it does not call a model
or return predictions.

```{r}
label_check_payload <-
  hf_zero_shot_classification_payload(
    augmented_rows$text,
    candidate_labels = c(
      "training",
      "requirement",
      "schedule",
      "skill",
      "other"
    ),
    multi_label = FALSE
  )

stopifnot(
  identical(
    label_check_payload$inputs,
    augmented_rows$text
  ),
  identical(
    label_check_payload$parameters$
      candidate_labels,
    c(
      "training",
      "requirement",
      "schedule",
      "skill",
      "other"
    )
  ),
  identical(
    label_check_payload$parameters$
      multi_label,
    FALSE
  )
)

str(label_check_payload)
```

A future model call could flag surprising candidates for review. A model label
would not prove that the augmentation is valid. The same codebook and human
review still apply. An actual request would also require a named model and
provider and would transmit the text under that provider's terms. No model or
provider is selected and no network request occurs here.

## A small edit can reverse the claim

Word replacement is risky when it changes negation or obligation.
`str_to_lower()` makes the keyword checks ignore case. `str_detect()` asks
whether each pattern is present.

```{r}
meaning_check <- tibble(
  version = c("source", "unsafe edit"),
  text = c(
    "Paid training is provided.",
    "Paid training is not provided."
  )
) |>
  mutate(
    lower_text = str_to_lower(text),
    keyword_only_match =
      str_detect(lower_text, fixed("paid")) &
        str_detect(lower_text, fixed("training")) &
        str_detect(lower_text, fixed("provided")),
    negation_aware_offer =
      keyword_only_match &
        !str_detect(lower_text, "\\bnot\\b")
  ) |>
  select(-lower_text)

stopifnot(
  !identical(
    meaning_check$text[[1]],
    meaning_check$text[[2]]
  ),
  identical(
    meaning_check$keyword_only_match,
    c(TRUE, TRUE)
  ),
  identical(
    meaning_check$negation_aware_offer,
    c(TRUE, FALSE)
  )
)

knitr::kable(
  meaning_check,
  col.names = c(
    "Version",
    "Text",
    "Keyword-only match",
    "Negation-aware offer rule"
  ),
  caption = "One added word reverses the training offer",
  row.names = FALSE
)
```

Both sentences satisfy the keyword-only check. The narrow negation rule
separates this pair, but it is not a general language-understanding method.
Label preservation still requires review at the meaning level.

## Augmentation can amplify a mistake

Check:

1. whether the label and important details remain true;
2. whether the wording is grammatical in the intended language;
3. whether duplicates cross into the evaluation set;
4. whether templates make one class unnaturally easy to recognize;
5. whether generated text adds stereotypes or private information;
6. whether one language or job family receives most synthetic rows; and
7. how conclusions change when synthetic examples are removed.

Large language models can generate many variants, but scale does not settle
these questions. Repeated training on generated text can also narrow a model's
representation of the underlying distribution.

## What to remember

- Augmented text is synthetic data, not a newly observed case.
- Record the source, method, label, and synthetic status.
- A familiar word does not guarantee that meaning stayed the same.
- Keep augmented rows out of evaluation data derived from their source.
- Compare results with and without synthetic examples.

The Lab ends this stage with five reviewable candidates and one rejected edit.
Their synthetic status and method remain attached. Keyword rules are the final
shortcut. They face both the designed data and cases with negation or misleading
word matches.

## Sources

- [EDA: Easy Data Augmentation Techniques](https://aclanthology.org/D19-1670/)
- [Improving Neural Machine Translation Models with Monolingual Data](https://aclanthology.org/P16-1009/)
- [Self-Instruct: Aligning Language Models with Self-Generated Instructions](https://aclanthology.org/2023.acl-long.754/)
- [AI models collapse when trained on recursively generated data](https://doi.org/10.1038/s41586-024-07566-y)
- [huggingfaceR documentation](https://farach.github.io/huggingfaceR/)
- [CMap dataset](https://doi.org/10.5281/zenodo.15260189)
