---
title: "Creating additional training examples"
subtitle: "Augment text without assuming the label survived"
author: "Alex Farach"
description: "Learn how to create workforce text variants and test whether they still support the intended label."
categories: [training data, augmentation, workforce research]
---

The reviewer comparison left two boundary cases visible rather than treating a
majority as truth. A different risk appears when the Riverton Workforce Lab
creates synthetic examples: the row count can grow without adding observed
evidence. The team can write additional sentences that follow known patterns.
More rows may help a model learn, but a changed sentence can lose the meaning
that justified its label. For this exercise, the team writes five fictional
variants about training and schedules. They cannot tell a worker that a real
course meets at a given time or provides paid training.

**Textual data augmentation** creates new training examples by transforming or
generating text from existing examples. Augmentation changes a dataset; it does
not collect new evidence about the labor market.

::: {.callout-note}
All sentences and organizations in this lesson are fictional. Augmented rows
are synthetic and remain marked as such.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- distinguish a new text variant from a new observation;
- create controlled template variants;
- use cmapr job titles without importing unsupported claims;
- prepare a classification-check payload with huggingfaceR;
- reject candidates whose label no longer holds; and
- test duplication, label balance, and whether augmentation changed anything.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Start with a narrow purpose

The team wants more examples of explicit training offers. It will not claim
that the examples describe real employers or estimate how often training is
offered. The next chunk loads package helpers for files, tables, strings, CMap
examples, and model-check payloads, then reads the file with declared column
types into a tibble.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(cmapr)
library(huggingfaceR)

workforce_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)
training_examples <- workforce_sentences |>
  filter(reference_label == "training")

knitr::kable(
  training_examples |>
    select(sentence_id, text),
  col.names = c("Sentence ID", "Reviewed training example"),
  caption = "The seven training examples before augmentation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(training_examples), 7L),
  inherits(workforce_sentences, "tbl_df"),
  all(training_examples$reference_label == "training")
)
```

The source labels were assigned under the manual annotation codebook. The
augmentation step must preserve that label definition.

## Use job titles as template values

cmapr includes a ten-row teaching fixture with job-title transitions. We use
three titles as vocabulary only. We do not carry over transition weights or
interpret a title change as a promotion.

```{r}
cmap_titles <- read_csv(
  cmap_example_data(),
  col_types = cols(
    sector = col_character(),
    region = col_character(),
    job_title_from = col_character(),
    job_title_to = col_character(),
    transition_weighted_count = col_double(),
    si_from = col_double(),
    si_to = col_double()
  )
)
selected_titles <- cmap_titles$job_title_from[c(3L, 5L, 8L)]

selected_titles
```

```{r}
#| include: false

stopifnot(
  identical(
    selected_titles,
    c(
      "Data Analyst",
      "Medical Assistant",
      "Junior Developer"
    )
  )
)
```

The titles come from the cmapr package's teaching fixture. That file does not
contain upstream record identifiers, so this lesson does not claim that the
three strings are authenticated extracts from the full CMap dataset. The
sentences created next are fictional.

## Create controlled variants

The template states the same kind of claim for each title.

```{r}
source("R/workforce-codebook.R")
augmentation_codebook <- workforce_codebook()

template_variants <- tibble(
  source_sentence_id = "s001",
  source_family_id = "training-offer-s001",
  template_id = "training-pathway-v1",
  vocabulary_source = "cmapr example_transitions.csv",
  vocabulary_row = c(3L, 5L, 8L),
  vocabulary_commit =
    "eb2d0b3810e07782aab4a81fba1f539133f2ba6a",
  text = sprintf(
    "The %s pathway includes paid training.",
    str_to_lower(selected_titles)
  ),
  label = "training",
  synthetic = TRUE,
  method = "controlled template",
  codebook_version = augmentation_codebook$version,
  codebook_hash = augmentation_codebook$hash,
  review_status = "requires_human_review",
  split = "training_candidate"
)

schedule_variants <- tibble(
  source_sentence_id = "s003",
  source_family_id = "schedule-s003",
  template_id = "class-schedule-v1",
  vocabulary_source = "author-created template",
  vocabulary_row = NA_integer_,
  vocabulary_commit = NA_character_,
  text = sprintf(
    "%s schedules are available.",
    c("Morning", "Weekend")
  ),
  label = "schedule",
  synthetic = TRUE,
  method = "controlled template",
  codebook_version = augmentation_codebook$version,
  codebook_hash = augmentation_codebook$hash,
  review_status = "requires_human_review",
  split = "training_candidate"
)

augmented_rows <- bind_rows(
  template_variants,
  schedule_variants
)
augmentation_label_order <- c(
  "training",
  "requirement",
  "schedule",
  "skill",
  "other"
)
augmentation_counts <- tibble(label = augmentation_label_order) |>
  left_join(
    augmented_rows |>
      count(label, name = "n"),
    by = join_by(label)
  ) |>
  mutate(n = coalesce(n, 0L))
source_families <- augmented_rows |>
  distinct(source_family_id) |>
  arrange(source_family_id)

knitr::kable(
  augmented_rows,
  col.names = c(
    "Source sentence",
    "Source family",
    "Template",
    "Vocabulary source",
    "Vocabulary row",
    "Vocabulary commit",
    "Synthetic text",
    "Proposed label",
    "Synthetic",
    "Method",
    "Codebook version",
    "Codebook hash",
    "Review status",
    "Intended split"
  ),
  caption = "Five controlled augmentation candidates",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(augmented_rows), 5L),
  !anyDuplicated(augmented_rows$text),
  !any(augmented_rows$text %in% workforce_sentences$text),
  all(augmented_rows$synthetic),
  all(augmented_rows$review_status == "requires_human_review"),
  all(augmented_rows$split == "training_candidate"),
  all(augmented_rows$codebook_hash == augmentation_codebook$hash),
  identical(template_variants$vocabulary_row, c(3L, 5L, 8L)),
  all(
    template_variants$vocabulary_commit ==
      "eb2d0b3810e07782aab4a81fba1f539133f2ba6a"
  ),
  identical(
    source_families$source_family_id,
    c("schedule-s003", "training-offer-s001")
  ),
  identical(augmentation_counts$n, c(3L, 0L, 2L, 0L, 0L))
)
```

The rows carry their source family, template, second vocabulary source,
package commit when applicable, codebook, synthetic status, intended split,
and pending review status. Any evaluation example derived from `s001` or `s003`
must stay outside the evaluation set with its whole family. The five rows
remain candidates until a person checks that each is natural, useful, and
consistent with the codebook.

## Review the candidates and reject some

A generator that produces only usable output is a generator nobody checked. The
Lab added two more candidates that a template would happily produce, then sent
all seven to a reviewer.

```{r}
risky_variants <- tibble(
  source_sentence_id = c("s001", "s015"),
  source_family_id = c(
    "training-offer-s001",
    "training-offer-s015"
  ),
  template_id = c(
    "training-pathway-v1",
    "training-restatement-v1"
  ),
  vocabulary_source = c(
    "cmapr example_transitions.csv",
    "author-created template"
  ),
  vocabulary_row = c(6L, NA_integer_),
  vocabulary_commit = c(
    "eb2d0b3810e07782aab4a81fba1f539133f2ba6a",
    NA_character_
  ),
  text = c(
    "The financial analyst pathway may include paid training.",
    "On-the-job training is provided."
  ),
  label = "training",
  synthetic = TRUE,
  method = "controlled template",
  codebook_version = augmentation_codebook$version,
  codebook_hash = augmentation_codebook$hash,
  review_status = "requires_human_review",
  split = "training_candidate"
)

candidate_rows <- bind_rows(augmented_rows, risky_variants)

review_decisions <- tibble(
  text = candidate_rows$text,
  decision = c(
    rep("accept", 5L),
    "reject",
    "reject"
  ),
  reason = c(
    rep("States an offer without hedging; label holds", 3L),
    rep("States when schedules are available; label holds", 2L),
    "\"may include\" removes the explicit offer the label requires",
    "Repeats a reviewed sentence word for word"
  )
)

reviewed_candidates <- candidate_rows |>
  left_join(review_decisions, by = join_by(text)) |>
  mutate(
    duplicates_reviewed_text = text %in%
      workforce_sentences$text,
    hedged = str_detect(text, "\\b(may|might|could)\\b"),
    review_status = case_when(
      decision == "accept" ~ "accepted",
      decision == "reject" ~ "rejected",
      TRUE ~ review_status
    )
  )

accepted_rows <- reviewed_candidates |>
  filter(decision == "accept")
rejected_rows <- reviewed_candidates |>
  filter(decision == "reject")

knitr::kable(
  reviewed_candidates |>
    select(text, label, decision, reason),
  col.names = c(
    "Candidate text",
    "Proposed label",
    "Decision",
    "Reason"
  ),
  caption = "Seven candidates, five accepted, two rejected",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(candidate_rows), 7L),
  identical(nrow(accepted_rows), 5L),
  identical(nrow(rejected_rows), 2L),
  identical(
    accepted_rows$text,
    augmented_rows$text
  ),
  all(!accepted_rows$duplicates_reviewed_text),
  all(!accepted_rows$hedged),
  all(accepted_rows$review_status == "accepted"),
  all(rejected_rows$review_status == "rejected"),
  identical(sum(rejected_rows$hedged), 1L),
  identical(
    sum(rejected_rows$duplicates_reviewed_text),
    1L
  )
)
```

Both rejections are the kind a template cannot catch on its own.

The hedged sentence is grammatical, plausible, and wrong for the label it
carries. Under the codebook, `training` requires an explicit offer, and "may
include" states a possibility. Training a model on it teaches that hedged
language marks a training offer, which is close to the opposite of the rule the
annotators applied.

The repeated sentence is worse in a quieter way. It matches a reviewed row word
for word, so it adds no information, and if its copy landed in training while
the original sat in evaluation, the evaluation would measure recall of a string
the model had already seen.

Two rejections out of seven is a rejection rate, not a defect. A generator with
no rejections usually means the review was skipped.

## Ask whether the accepted rows changed anything

The five accepted rows are safe. Safe is not the same as useful, and the way to
find out is to run the downstream task twice.

The task below is deliberately crude: decide whether a sentence is a training
offer, using terms that appear in training examples and never in the others. It
learns from the five complete postings `J001` through `J005`, and it is measured
on the ten units from `J006` and the flyer, which no augmented row was derived
from.

```{r}
labeled_pool <- workforce_sentences |>
  filter(document_id %in% sprintf("J%03d", 1:5)) |>
  select(text, label = reference_label)
held_out <- workforce_sentences |>
  filter(document_id %in% c("J006", "F001")) |>
  select(sentence_id, text, label = reference_label)
accepted_source_ids <- unique(accepted_rows$source_sentence_id)

to_words <- function(text) {
  text |>
    str_to_lower() |>
    str_replace_all("[^a-z0-9 ]", " ") |>
    str_squish() |>
    str_split(" ")
}

distinctive_terms <- function(examples) {
  training_words <- unique(unlist(
    to_words(examples$text[examples$label == "training"])
  ))
  other_words <- unique(unlist(
    to_words(examples$text[examples$label != "training"])
  ))

  sort(setdiff(
    training_words[str_length(training_words) > 2],
    other_words
  ))
}

predict_training <- function(terms, text) {
  map_lgl(
    to_words(text),
    \(row_words) any(row_words %in% terms)
  )
}

without_augmentation <- distinctive_terms(labeled_pool)
with_augmentation <- distinctive_terms(
  bind_rows(
    labeled_pool,
    accepted_rows |>
      select(text, label)
  )
)

is_training <- held_out$label == "training"
ablation <- tibble(
  condition = c("without augmentation", "with augmentation"),
  distinctive_terms = c(
    length(without_augmentation),
    length(with_augmentation)
  ),
  correct = c(
    sum(predict_training(without_augmentation, held_out$text) == is_training),
    sum(predict_training(with_augmentation, held_out$text) == is_training)
  ),
  items = nrow(held_out)
)

new_terms <- setdiff(with_augmentation, without_augmentation)

knitr::kable(
  ablation,
  col.names = c(
    "Condition",
    "Distinctive terms",
    "Correct decisions",
    "Items judged"
  ),
  caption = "The same held-out units judged with and without the augmented rows",
  row.names = FALSE
)

new_terms
```

```{r}
#| include: false

stopifnot(
  identical(nrow(held_out), 10L),
  identical(
    intersect(accepted_source_ids, held_out$sentence_id),
    character()
  ),
  identical(sum(is_training), 2L),
  identical(ablation$distinctive_terms, c(13L, 18L)),
  identical(ablation$correct, c(10L, 10L)),
  identical(length(new_terms), 5L),
  identical(
    new_terms,
    c("analyst", "assistant", "developer", "junior", "pathway")
  ),
  identical(
    predict_training(without_augmentation, held_out$text),
    predict_training(with_augmentation, held_out$text)
  )
)
```

The augmented rows added five terms to the vocabulary and changed nothing. Every
held-out decision is identical under both conditions, and both get all ten
right.

The five new terms explain why. Four of them are job titles and one is the word
`pathway`, all of which arrived from the template rather than from any new
observation about how employers describe training. The word doing the work,
`training`, was already there.

The honest reading of this comparison is modest in both directions. It does not
show that augmentation is useless; ten items with two positive cases could not
detect a small improvement, and a stronger model with more data might use the
extra rows differently. It does show that these five rows earned nothing here,
which is exactly the finding a report should carry when a training pool grows by
more than a quarter and the result does not move.

## Prepare a model check without claiming a result

huggingfaceR can prepare the payload for a zero-shot classification request.
This code records the candidate labels and texts, but it does not call a model
or return predictions.

```{r}
label_check_payload <-
  hf_zero_shot_classification_payload(
    augmented_rows$text,
    candidate_labels = c(
      "training",
      "requirement",
      "schedule",
      "skill",
      "other"
    ),
    multi_label = FALSE
  )

str(label_check_payload)
```

```{r}
#| include: false

stopifnot(
  identical(
    label_check_payload$inputs,
    augmented_rows$text
  ),
  identical(
    label_check_payload$parameters$
      candidate_labels,
    c(
      "training",
      "requirement",
      "schedule",
      "skill",
      "other"
    )
  ),
  identical(
    label_check_payload$parameters$
      multi_label,
    FALSE
  )
)
```

A future model call could flag surprising candidates for review. A model label
would not prove that the augmentation is valid. The same codebook and human
review still apply. An actual request would also require a named model and
provider and would transmit the text under that provider's terms. No model or
provider is selected and no network request occurs here.

## A small edit can reverse the claim

Word replacement is risky when it changes negation or obligation.
`str_to_lower()` makes the keyword checks ignore case. `str_detect()` asks
whether each pattern is present.

```{r}
meaning_check <- tibble(
  version = c("source", "unsafe edit"),
  text = c(
    "Paid training is provided.",
    "Paid training is not provided."
  )
) |>
  mutate(
    lower_text = str_to_lower(text),
    keyword_only_match =
      str_detect(lower_text, fixed("paid")) &
        str_detect(lower_text, fixed("training")) &
        str_detect(lower_text, fixed("provided")),
    negation_aware_offer =
      keyword_only_match &
        !str_detect(lower_text, "\\bnot\\b")
  ) |>
  select(-lower_text)

knitr::kable(
  meaning_check,
  col.names = c(
    "Version",
    "Text",
    "Keyword-only match",
    "Negation-aware offer rule"
  ),
  caption = "One added word reverses the training offer",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  !identical(
    meaning_check$text[[1]],
    meaning_check$text[[2]]
  ),
  identical(
    meaning_check$keyword_only_match,
    c(TRUE, TRUE)
  ),
  identical(
    meaning_check$negation_aware_offer,
    c(TRUE, FALSE)
  )
)
```

Both sentences satisfy the keyword-only check. The narrow negation rule
separates this pair, but it is not a general language-understanding method.
Label preservation still requires review at the meaning level.

## Augmentation can amplify a mistake

Check:

1. whether the label and important details remain true;
2. whether the wording is grammatical in the intended language;
3. whether duplicates cross into the evaluation set;
4. whether templates make one class unnaturally easy to recognize;
5. whether generated text adds stereotypes or private information;
6. whether one language or job family receives most synthetic rows; and
7. how conclusions change when synthetic examples are removed.

Large language models can generate many variants, but scale does not settle
these questions. Repeated training on generated text can also narrow a model's
representation of the underlying distribution.

## What to remember

- Augmented text is synthetic data, not a newly observed case.
- Record the source, method, label, and synthetic status.
- A familiar word does not guarantee that meaning stayed the same.
- Review every candidate and keep the rejected ones with their reasons.
- Keep augmented rows out of evaluation data derived from their source.
- Compare results with and without synthetic examples, and report a flat result
  as readily as a favorable one.

The Lab ends this stage with five accepted rows, two rejected ones, and a
comparison that found no downstream difference. Their synthetic status and
method remain attached. Keyword rules are the final shortcut. They face both the
designed data and cases with negation or misleading word matches.

## Sources

- [EDA: Easy Data Augmentation Techniques](https://aclanthology.org/D19-1670/)
- [Improving Neural Machine Translation Models with Monolingual Data](https://aclanthology.org/P16-1009/)
- [Self-Instruct: Aligning Language Models with Self-Generated Instructions](https://aclanthology.org/2023.acl-long.754/)
- [AI models collapse when trained on recursively generated data](https://doi.org/10.1038/s41586-024-07566-y)
- [huggingfaceR documentation](https://farach.github.io/huggingfaceR/)
- [CMap dataset](https://doi.org/10.5281/zenodo.15260189)
