---
title: "Labeling text with written rules"
subtitle: "Measure rule coverage, conflicts, and mistakes"
author: "Alex Farach"
description: "Learn how rule-based weak labels can expand workforce training data without becoming ground truth."
categories: [training data, weak supervision, workforce research]
---

Five fictional sentence variants remain marked as synthetic. For the rule
test, the Riverton Workforce Lab returns to its 28 reviewed fictional text
units. Several phrases appear repeatedly: `training`, `required`, `shift`,
and `skill`. The team turns those patterns into rules, then checks whether
familiar words preserve distinctions a worker may use when comparing notices.

**Rule-based training data** uses written conditions to assign labels.
These proposed labels are often called **weak labels** because a rule is an
imperfect source of supervision. The team will measure where its rules label,
abstain, conflict, and disagree with human review.

::: {.callout-note}
The sentences and team are fictional. The rules were designed after inspecting
this teaching dataset, so results on these rows are not an independent
performance estimate.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- express labeling rules as text patterns;
- keep each rule's vote instead of hiding conflicts;
- calculate coverage, conflict, and abstention;
- compare usable weak labels with reference labels; and
- stress-test rules on cases they were not designed to fit.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Write rules that can abstain

Each rule searches for words tied to one codebook category. A rule returns
`TRUE` when it votes for its category and `FALSE` otherwise.

```{r}
workforce_sentences <- read.csv(
  "data/workforce/workforce_sentences.csv",
  colClasses = "character",
  check.names = FALSE
)

rule_patterns <- c(
  training = paste(
    "training",
    "certificat",
    "apprenticeship",
    "mentor",
    sep = "|"
  ),
  requirement = paste(
    "required",
    "preferred",
    "license",
    "diploma",
    "portfolio",
    "must be able",
    sep = "|"
  ),
  schedule = "schedule|shift|classes",
  skill = "skill|spreadsheet|communication"
)

rule_votes <- vapply(
  rule_patterns,
  function(pattern) {
    grepl(
      pattern,
      tolower(workforce_sentences$text),
      perl = TRUE
    )
  },
  logical(nrow(workforce_sentences))
)

stopifnot(
  identical(
    dim(rule_votes),
    c(28L, 4L)
  ),
  identical(
    colnames(rule_votes),
    names(rule_patterns)
  )
)

head(rule_votes, 8)
```

The matrix preserves one column per rule. Combining the votes too early would
hide why a label was proposed.

## Mark abstentions and conflicts

A row can receive:

- one vote, which produces a usable weak label;
- no votes, which produces an abstention; or
- several votes, which produces a conflict.

```{r}
rule_count <- rowSums(rule_votes)
first_vote <- colnames(rule_votes)[
  max.col(rule_votes, ties.method = "first")
]
weak_label <- ifelse(
  rule_count == 0L,
  "abstain",
  ifelse(
    rule_count > 1L,
    "conflict",
    first_vote
  )
)

weak_results <- data.frame(
  sentence_id = workforce_sentences$sentence_id,
  text = workforce_sentences$text,
  reference_label =
    workforce_sentences$reference_label,
  rule_count = rule_count,
  weak_label = weak_label,
  stringsAsFactors = FALSE
)

result_counts <- table(weak_results$weak_label)

stopifnot(
  identical(
    unname(
      as.integer(
        result_counts[
          c(
            "abstain",
            "conflict",
            "requirement",
            "schedule",
            "skill",
            "training"
          )
        ]
      )
    ),
    c(3L, 2L, 8L, 4L, 3L, 8L)
  )
)

knitr::kable(
  weak_results[
    weak_results$weak_label %in%
      c("abstain", "conflict"),
  ],
  col.names = c(
    "Sentence ID",
    "Text",
    "Reference label",
    "Rule votes",
    "Weak-label result"
  ),
  caption = "The five sentences that abstain or conflict",
  row.names = FALSE
)
```

Three sentences receive no vote. Two receive conflicting votes. Sentence
`s013`, for example, contains both `certificate` and `preferred`, so the
training and requirement rules disagree.

## Report coverage before accuracy

**Coverage** is the share of rows receiving at least one vote. **Conflict
rate** is the share receiving more than one. **Usable coverage** counts rows
with exactly one vote.

```{r}
rule_summary <- data.frame(
  measure = c(
    "coverage",
    "conflict rate",
    "usable coverage"
  ),
  value = c(
    mean(rule_count > 0L),
    mean(rule_count > 1L),
    mean(rule_count == 1L)
  )
)

stopifnot(
  isTRUE(
    all.equal(
      rule_summary$value,
      c(25 / 28, 2 / 28, 23 / 28)
    )
  )
)

knitr::kable(
  transform(
    rule_summary,
    value = round(value, 3)
  ),
  col.names = c("Measure", "Share of 28 text units"),
  caption = "Coverage and conflict in the teaching dataset",
  row.names = FALSE
)
```

Coverage alone does not measure correctness. A rule can label every row and be
wrong every time.

## Compare usable labels with human review

The 23 rows with one rule vote include two errors. We can summarize that
comparison with foundryR.

```{r}
usable_rows <- weak_results$rule_count == 1L
agreement_data <- weak_results[
  usable_rows,
  c("weak_label", "reference_label")
]
rule_agreement <- foundryR::foundry_agreement(
  agreement_data,
  estimate = "weak_label",
  truth = "reference_label"
)

stopifnot(
  identical(nrow(agreement_data), 23L),
  isTRUE(
    all.equal(
      rule_agreement$value[
        rule_agreement$metric == "accuracy"
      ],
      21 / 23
    )
  )
)

knitr::kable(
  rule_agreement,
  col.names = c("Measure", "Value", "Compared sentences"),
  caption = "In-sample agreement for usable weak labels",
  row.names = FALSE
)
```

The rule labels 21 of the 23 usable rows the same way as the reference
annotation. It incorrectly labels `RIVERTON SKILLS OPEN HOUSE` as `skill` and
`DATA SUPPORT CERTIFICATE` as `training`. The trigger words are present, but
the headings alone do not make either claim.

The full accounting keeps that error beside unresolved rows.

```{r}
all_row_accounting <- data.frame(
  outcome = c(
    "usable and matches reference",
    "usable and differs from reference",
    "unresolved: abstain or conflict"
  ),
  rows = c(
    sum(
      usable_rows &
        weak_results$weak_label ==
          weak_results$reference_label
    ),
    sum(
      usable_rows &
        weak_results$weak_label !=
          weak_results$reference_label
    ),
    sum(!usable_rows)
  )
)

stopifnot(
  identical(
    all_row_accounting$rows,
    c(21L, 2L, 5L)
  ),
  identical(
    sum(all_row_accounting$rows),
    28L
  )
)

knitr::kable(
  all_row_accounting,
  col.names = c("Outcome", "Text units"),
  caption = "Accounting for all 28 rule results",
  row.names = FALSE
)
```

These are in-sample results. The rules were written after reading the
sentences, so they may encode the examples rather than the broader concept.

## Try cases the rules did not shape

Before applying the rules to another batch, the team writes four fictional
adversarial unit tests. Their expected labels were chosen to expose known
shortcuts after the rules were written. They are not an independent evaluation
sample.

```{r}
stress_cases <- data.frame(
  case_id = sprintf("challenge-%d", 1:4),
  text = c(
    "The training department is hiring an accountant.",
    "Employees learn from experienced coworkers.",
    "Certification is not required.",
    "Flexible hours are available."
  ),
  reference_label = c(
    "other",
    "training",
    "requirement",
    "schedule"
  ),
  reference_status = "designed challenge expectation",
  stringsAsFactors = FALSE
)

stress_votes <- vapply(
  rule_patterns,
  function(pattern) {
    grepl(
      pattern,
      tolower(stress_cases$text),
      perl = TRUE
    )
  },
  logical(nrow(stress_cases))
)
stress_count <- rowSums(stress_votes)
stress_first_vote <- colnames(stress_votes)[
  max.col(stress_votes, ties.method = "first")
]
stress_label <- ifelse(
  stress_count == 0L,
  "abstain",
  ifelse(
    stress_count > 1L,
    "conflict",
    stress_first_vote
  )
)

stress_results <- data.frame(
  stress_cases,
  weak_label = stress_label
)

stopifnot(
  identical(
    stress_results$weak_label,
    c("training", "abstain", "conflict", "abstain")
  ),
  !any(
    stress_results$weak_label ==
      stress_results$reference_label
  )
)

knitr::kable(
  stress_results,
  col.names = c(
    "Challenge ID",
    "Challenge text",
    "Reference label",
    "Reference status",
    "Rule result"
  ),
  caption = "Four designed adversarial failures",
  row.names = FALSE
)
```

`training department` triggers a false positive. `Employees learn` and
`Flexible hours` use words the rules do not cover. `Certification is not
required` creates a conflict. These failures reveal missing context, synonyms,
and negation.

## Keep weak labels weak

When rules scale:

1. version every pattern and codebook;
2. preserve each rule vote and abstention;
3. measure coverage and conflict by source, language, and group;
4. reserve independently reviewed data for evaluation;
5. sample apparent successes for human audit;
6. test negation, polysemy, and unseen wording; and
7. compare conclusions with and without weak labels.

Rules can provide useful training signals. They do not create facts, repair a
biased source, or remove the need for people who understand the domain.

## What to remember

- A weak label is a rule's proposal, not ground truth.
- Rules should be allowed to abstain and expose conflicts.
- Coverage, conflict, and agreement answer different questions.
- In-sample success can reflect rules written around the examples.
- Designed challenge cases expose known shortcuts; independent evaluation asks
  whether they generalize.

The investigation ends with the questions that opened it: what does each
fictional notice explicitly say about training, schedules, requirements, and
skills? The Lab can trace an answer through source capture, OCR review, a
codebook version, individual labels, synthetic status, and rule votes. None of
that makes the notices representative or the rules ready for new data. It
makes the limits and unresolved cases inspectable.

## Sources

- [Snorkel: Rapid Training Data Creation with Weak Supervision](https://www.vldb.org/pvldb/vol11/p269-ratner.pdf)
- [Data Programming: Creating Large Training Sets, Quickly](https://proceedings.neurips.cc/paper/2016/hash/6709e8d64a5f47269ed5cea9f625f7ab-Abstract.html)
- [A Survey on Programmatic Weak Supervision](https://arxiv.org/abs/2202.05433)
- [foundryR agreement documentation](https://farach.github.io/foundryR/reference/foundry_agreement.html)
