---
title: "Labeling text with written rules"
subtitle: "Measure rule coverage, conflicts, and mistakes"
author: "Alex Farach"
description: "Learn how rule-based weak labels can expand workforce training data without becoming ground truth."
categories: [training data, weak supervision, workforce research]
---

Five fictional sentence variants remain marked as synthetic. For the rule
test, the Riverton Workforce Lab returns to its 28 reviewed fictional text
units. Several phrases appear repeatedly: `training`, `required`, `shift`,
and `skill`. The team turns those patterns into rules, then checks whether
familiar words preserve distinctions a worker may use when comparing notices.

**Rule-based training data** uses written conditions to assign labels.
These proposed labels are often called **weak labels** because a rule is an
imperfect source of supervision. The team will measure where its rules label,
abstain, conflict, and disagree with human review.

::: {.callout-note}
The sentences and team are fictional. The rules were designed after inspecting
this teaching dataset, so results on these rows are not an independent
performance estimate.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- express labeling rules as text patterns;
- keep each rule's vote and avoid hiding conflicts;
- calculate coverage, conflict, and abstention;
- report each rule's own coverage and error;
- compare usable weak labels with reference labels; and
- stress-test rules on cases they were not designed to fit.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Write rules that can abstain

Each rule searches for words tied to one codebook category. A rule returns
`TRUE` when it votes for its category and `FALSE` otherwise. The next chunk
loads package helpers for files, tables, reshaping, repeated rule checks,
strings, and agreement, then reads the file with declared column types. Inside
the chunk, `map()` makes one TRUE/FALSE list per rule and `as_tibble()` turns
that list into columns.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(foundryR)

workforce_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

rule_patterns <- c(
  training = str_c(
    "training",
    "certificat",
    "apprenticeship",
    "mentor",
    sep = "|"
  ),
  requirement = str_c(
    "required",
    "preferred",
    "license",
    "diploma",
    "portfolio",
    "must be able",
    sep = "|"
  ),
  schedule = "schedule|shift|classes",
  skill = "skill|spreadsheet|communication"
)
rule_labels <- names(rule_patterns)
sentence_text_lower <- str_to_lower(workforce_sentences$text)

rule_votes <- bind_cols(
  tibble(sentence_id = workforce_sentences$sentence_id),
  as_tibble(
    map(rule_patterns, \(pattern) {
      str_detect(sentence_text_lower, pattern)
    })
  )
)

head(rule_votes, 8)
```

```{r}
#| include: false

stopifnot(
  identical(
    dim(rule_votes[rule_labels]),
    c(28L, 4L)
  ),
  identical(
    names(rule_votes[rule_labels]),
    rule_labels
  )
)
```

The rule-vote tibble preserves one column per rule. Combining the votes too
early would hide why a label was proposed.

## Mark abstentions and conflicts

A row can receive:

- one vote, which produces a usable weak label;
- no votes, which produces an abstention; or
- several votes, which produces a conflict.

```{r}
rule_counts <- rule_votes |>
  mutate(
    rule_count = as.integer(rowSums(rule_votes[rule_labels]))
  ) |>
  select(sentence_id, rule_count)
voted_labels <- rule_votes |>
  pivot_longer(
    cols = all_of(rule_labels),
    names_to = "weak_label",
    values_to = "rule_vote"
  ) |>
  filter(rule_vote) |>
  summarise(
    voted_label = first(weak_label),
    .by = sentence_id
  )

weak_results <- workforce_sentences |>
  select(sentence_id, text, reference_label) |>
  left_join(rule_counts, by = join_by(sentence_id)) |>
  left_join(voted_labels, by = join_by(sentence_id)) |>
  mutate(
    weak_label = case_when(
      rule_count == 0L ~ "abstain",
      rule_count > 1L ~ "conflict",
      TRUE ~ voted_label
    )
  ) |>
  select(-voted_label)

result_order <- c(
  "abstain",
  "conflict",
  "requirement",
  "schedule",
  "skill",
  "training"
)
result_counts <- tibble(weak_label = result_order) |>
  left_join(
    weak_results |>
      count(weak_label, name = "rows"),
    by = join_by(weak_label)
  ) |>
  mutate(rows = coalesce(rows, 0L))

knitr::kable(
  weak_results |>
    filter(weak_label %in% c("abstain", "conflict")),
  col.names = c(
    "Sentence ID",
    "Text",
    "Reference label",
    "Rule votes",
    "Weak-label result"
  ),
  caption = "The five sentences that abstain or conflict",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    result_counts$rows,
    c(3L, 2L, 8L, 4L, 3L, 8L)
  )
)
```

Three sentences receive no vote. Two receive conflicting votes. Sentence
`s013`, for example, contains both `certificate` and `preferred`, so the
training and requirement rules disagree.

## Report coverage before accuracy

**Coverage** is the share of rows receiving at least one vote. **Conflict
rate** is the share receiving more than one. **Usable coverage** counts rows
with exactly one vote.

```{r}
rule_summary <- tibble(
  measure = c(
    "coverage",
    "conflict rate",
    "usable coverage"
  ),
  value = c(
    mean(weak_results$rule_count > 0L),
    mean(weak_results$rule_count > 1L),
    mean(weak_results$rule_count == 1L)
  )
)

knitr::kable(
  rule_summary |>
    mutate(
      value = round(value, 3)
    ),
  col.names = c("Measure", "Share of 28 text units"),
  caption = "Coverage and conflict in the teaching dataset",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  isTRUE(
    all.equal(
      rule_summary$value,
      c(25 / 28, 2 / 28, 23 / 28)
    )
  )
)
```

Coverage alone does not measure correctness. A rule can label every row and be
wrong every time.

## Compare usable labels with human review

The 23 rows with one rule vote include two errors. We can summarize that
comparison with foundryR.

```{r}
agreement_data <- weak_results |>
  filter(rule_count == 1L) |>
  select(weak_label, reference_label)
error_sentences <- weak_results |>
  filter(
    rule_count == 1L,
    weak_label != reference_label
  )
error_sentence_ids <- error_sentences$sentence_id
rule_agreement <- foundryR::foundry_agreement(
  agreement_data,
  estimate = "weak_label",
  truth = "reference_label"
)

knitr::kable(
  rule_agreement,
  col.names = c("Measure", "Value", "Compared sentences"),
  caption = "In-sample agreement for usable weak labels",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(agreement_data), 23L),
  isTRUE(
    all.equal(
      filter(rule_agreement, metric == "accuracy")$value,
      21 / 23
    )
  ),
  identical(error_sentence_ids, c("s023", "s024"))
)
```

The rule labels 21 of the 23 usable rows the same way as the reference
annotation. It incorrectly labels `RIVERTON SKILLS OPEN HOUSE` as `skill` and
`DATA SUPPORT CERTIFICATE` as `training`. The trigger words are present, but
the headings alone do not make either claim.

The full accounting keeps that error beside unresolved rows.

```{r}
all_row_accounting <- tibble(
  outcome = c(
    "usable and matches reference",
    "usable and differs from reference",
    "unresolved: abstain or conflict"
  ),
  rows = c(
    weak_results |>
      filter(
        rule_count == 1L,
        weak_label == reference_label
      ) |>
      nrow(),
    weak_results |>
      filter(
        rule_count == 1L,
        weak_label != reference_label
      ) |>
      nrow(),
    weak_results |>
      filter(rule_count != 1L) |>
      nrow()
  )
)

knitr::kable(
  all_row_accounting,
  col.names = c("Outcome", "Text units"),
  caption = "Accounting for all 28 rule results",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    all_row_accounting$rows,
    c(21L, 2L, 5L)
  ),
  identical(
    sum(all_row_accounting$rows),
    28L
  )
)
```

These are in-sample results. The rules were written after reading the
sentences, so they may encode the examples rather than the broader concept.

## Look at each rule on its own

The 21-of-23 figure is a pooled number, and pooled numbers hide which rule is
carrying the work. Each rule has its own coverage and its own error, and the
error looks different depending on whether the rule fired alone or alongside
another.

```{r}
reference_labels <- workforce_sentences$reference_label
votes_per_row <- as.integer(
  rowSums(rule_votes[rule_labels])
)

rule_diagnostics <- map(
  rule_labels,
  \(rule_name) {
    fired <- rule_votes[[rule_name]]
    fired_alone <- fired & votes_per_row == 1L
    matches_reference <- reference_labels == rule_name

    tibble(
      rule = rule_name,
      rows_fired = sum(fired),
      rows_fired_alone = sum(fired_alone),
      alone_and_correct = sum(fired_alone & matches_reference),
      correct_whenever_fired = sum(fired & matches_reference)
    )
  }
) |>
  list_rbind() |>
  mutate(
    precision_when_alone = alone_and_correct / rows_fired_alone
  )

knitr::kable(
  rule_diagnostics |>
    mutate(
      precision_when_alone = round(
        precision_when_alone,
        2
      )
    ) |>
    select(
      rule,
      rows_fired,
      rows_fired_alone,
      alone_and_correct,
      precision_when_alone
    ),
  col.names = c(
    "Rule",
    "Rows it fired on",
    "Fired alone",
    "Alone and correct",
    "Precision when alone"
  ),
  caption = "Each rule's coverage and error, separated",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    rule_diagnostics$rows_fired,
    c(9L, 10L, 5L, 3L)
  ),
  identical(
    rule_diagnostics$rows_fired_alone,
    c(8L, 8L, 4L, 3L)
  ),
  identical(
    rule_diagnostics$alone_and_correct,
    c(7L, 8L, 4L, 2L)
  ),
  identical(
    sum(rule_diagnostics$rows_fired_alone),
    23L
  ),
  identical(
    sum(rule_diagnostics$alone_and_correct),
    21L
  ),
  isTRUE(all.equal(
    rule_diagnostics$precision_when_alone,
    c(7 / 8, 1, 1, 2 / 3)
  ))
)
```

The pooled figure of 21 out of 23 becomes four different stories. The
`requirement` and `schedule` rules were right every time they spoke alone. The
`training` rule missed once in eight. The `skill` rule, which fired on only three
rows, was wrong on one of them, and a precision of two thirds on three rows is a
number to treat as a warning rather than an estimate.

A rule that fires rarely is where in-sample results are least trustworthy. There
is no way to distinguish a rule that is genuinely reliable from one that has not
yet met the case that breaks it.

## Separate conflicting rules from correlated support

Rules can fire on the same row because they conflict or because they repeat the
same evidence. Those cases have different consequences. The example below
contains opposing-label conflicts. It does not estimate correlation or show
confidence inflation.

```{r}
vote_matrix <- as.matrix(rule_votes[rule_labels])
rule_pairs <- expand.grid(
  first = rule_labels,
  second = rule_labels,
  stringsAsFactors = FALSE
) |>
  as_tibble() |>
  filter(first < second) |>
  mutate(
    fired_together = map2_int(
      first,
      second,
      \(left, right) {
        sum(vote_matrix[, left] & vote_matrix[, right])
      }
    )
  ) |>
  arrange(desc(fired_together), first, second)

knitr::kable(
  rule_pairs,
  col.names = c("Rule", "Other rule", "Rows where both fired"),
  caption = "How often each pair of rules spoke about the same row",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(rule_pairs), 6L),
  identical(
    sum(rule_pairs$fired_together),
    2L
  ),
  identical(
    rule_pairs$fired_together[
      rule_pairs$first == "requirement" &
        rule_pairs$second == "training"
    ],
    1L
  ),
  identical(
    rule_pairs$fired_together[
      rule_pairs$first == "requirement" &
        rule_pairs$second == "schedule"
    ],
    1L
  )
)
```

Only two pairs ever overlap here, and both overlaps are the conflicts already
seen. `s013` triggers `training` and `requirement` because a certificate is a
credential and a credential looks like training. `s020` triggers `requirement`
and `schedule` because a required shift is both a demand and a time.

These are conflict overlaps, not duplicate supporting votes. Correlated support
would require two rules that tend to cast the same label because they reuse the
same signal, such as two training rules built from nearly identical keyword
lists. A label model can estimate rule accuracy and dependence under stronger
data and modeling assumptions, but this table does not supply that evidence.

This lesson does not fit such a model: four rules, 28 rows, and two conflicting
overlaps do not contain enough information to estimate accuracies and
correlations. The scope here stops at measuring where the rules agree, where
they collide, and where they say nothing.

### What a conflict policy would cost

A team that needs a label for every row usually writes a priority order. Here is
one, and here is the price it charges.

```{r}
priority_order <- c(
  "schedule",
  "requirement",
  "training",
  "skill"
)

resolve_by_priority <- function(row_votes) {
  chosen <- priority_order[priority_order %in% rule_labels[row_votes]]

  if (length(chosen) == 0L) {
    NA_character_
  } else {
    chosen[1]
  }
}

conflict_rows <- which(votes_per_row > 1L)
conflict_outcomes <- tibble(
  sentence_id = rule_votes$sentence_id[conflict_rows],
  reference_label = reference_labels[conflict_rows],
  resolved_label = map_chr(
    conflict_rows,
    \(row) resolve_by_priority(vote_matrix[row, ])
  )
) |>
  mutate(policy_agrees = resolved_label == reference_label)

abstain_rows <- which(votes_per_row == 0L)
abstain_outcomes <- tibble(
  sentence_id = rule_votes$sentence_id[abstain_rows],
  reference_label = reference_labels[abstain_rows]
)

knitr::kable(
  conflict_outcomes,
  col.names = c(
    "Sentence ID",
    "Reference label",
    "Priority policy label",
    "Policy agrees"
  ),
  caption = "What a priority order does with the two conflicts",
  row.names = FALSE
)

knitr::kable(
  abstain_outcomes,
  col.names = c("Sentence ID", "Reference label"),
  caption = "What the three silent rows actually were",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(conflict_outcomes), 2L),
  identical(
    conflict_outcomes$sentence_id,
    c("s013", "s020")
  ),
  identical(
    conflict_outcomes$resolved_label,
    c("requirement", "schedule")
  ),
  all(conflict_outcomes$policy_agrees),
  identical(nrow(abstain_outcomes), 3L),
  all(abstain_outcomes$reference_label == "other")
)
```

The policy gets both conflicts right, and that is close to meaningless as
evidence. The order was chosen by someone who had read the codebook rule saying
timing takes precedence, and it was applied to two rows. Two successes chosen
after the fact do not establish that the order generalizes.

The three silent rows are more informative. All three carry the reference label
`other`, which is the one category no rule was written for. Silence here is
correct behavior expressed badly: the rules abstain because they have nothing to
say, and a fifth rule for `other` would be worse, since `other` is defined by
the absence of the other four rather than by any words of its own.

Every one of the 28 rows now has an account. Twenty-one usable and correct, two
usable and wrong, two in conflict and resolved by a stated policy, and three
abstentions whose true label was outside the rule set. Nothing is left in a
category called "the rest".

## Try cases the rules did not shape

Before applying the rules to another batch, the team writes four fictional
adversarial unit tests. Their expected labels were chosen to expose known
shortcuts after the rules were written. They are not an independent evaluation
sample.

```{r}
stress_cases <- tibble(
  case_id = sprintf("challenge-%d", 1:4),
  text = c(
    "The training department is hiring an accountant.",
    "Employees learn from experienced coworkers.",
    "Certification is not required.",
    "Flexible hours are available."
  ),
  reference_label = c(
    "other",
    "training",
    "requirement",
    "schedule"
  ),
  reference_status = "designed challenge expectation"
)

stress_text_lower <- str_to_lower(stress_cases$text)
stress_votes <- bind_cols(
  tibble(case_id = stress_cases$case_id),
  as_tibble(
    map(rule_patterns, \(pattern) {
      str_detect(stress_text_lower, pattern)
    })
  )
)
stress_counts <- stress_votes |>
  mutate(
    rule_count = as.integer(rowSums(stress_votes[rule_labels]))
  ) |>
  select(case_id, rule_count)
stress_voted_labels <- stress_votes |>
  pivot_longer(
    cols = all_of(rule_labels),
    names_to = "weak_label",
    values_to = "rule_vote"
  ) |>
  filter(rule_vote) |>
  summarise(
    voted_label = first(weak_label),
    .by = case_id
  )

stress_results <- stress_cases |>
  left_join(stress_counts, by = join_by(case_id)) |>
  left_join(stress_voted_labels, by = join_by(case_id)) |>
  mutate(
    weak_label = case_when(
      rule_count == 0L ~ "abstain",
      rule_count > 1L ~ "conflict",
      TRUE ~ voted_label
    )
  ) |>
  select(-rule_count, -voted_label)

knitr::kable(
  stress_results,
  col.names = c(
    "Challenge ID",
    "Challenge text",
    "Reference label",
    "Reference status",
    "Rule result"
  ),
  caption = "Four designed adversarial failures",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    stress_results$weak_label,
    c("training", "abstain", "conflict", "abstain")
  ),
  !any(
    stress_results$weak_label == stress_results$reference_label
  )
)
```

`training department` triggers a false positive. `Employees learn` and
`Flexible hours` use words the rules do not cover. `Certification is not
required` creates a conflict. These failures reveal missing context, synonyms,
and negation.

## Keep weak labels weak

When rules scale:

1. version every pattern and codebook;
2. preserve each rule vote and abstention;
3. measure coverage and conflict by source, language, and group;
4. reserve independently reviewed data for evaluation;
5. sample apparent successes for human audit;
6. test negation, polysemy, and unseen wording; and
7. compare conclusions with and without weak labels.

Rules can provide useful training signals. They do not create facts, repair a
biased source, or remove the need for people who understand the domain.

## What to remember

- A weak label is a rule's proposal, not ground truth.
- Rules should be allowed to abstain and expose conflicts.
- Coverage, conflict, and agreement answer different questions.
- Report each rule separately; a pooled score hides the weak one.
- Rules that share trigger words are not independent votes.
- In-sample success can reflect rules written around the examples.
- Designed challenge cases expose known shortcuts; independent evaluation asks
  whether they generalize.

The investigation ends with the questions that opened it: what does each
fictional notice explicitly say about training, schedules, requirements, and
skills? The Lab can trace an answer through source capture, OCR review, a
codebook version, individual labels, synthetic status, and rule votes. None of
that makes the notices representative or the rules ready for new data. It
makes the limits and unresolved cases inspectable.

## Sources

- [Snorkel: Rapid Training Data Creation with Weak Supervision](https://www.vldb.org/pvldb/vol11/p269-ratner.pdf)
- [Data Programming: Creating Large Training Sets, Quickly](https://proceedings.neurips.cc/paper/2016/hash/6709e8d64a5f47269ed5cea9f625f7ab-Abstract.html)
- [A Survey on Programmatic Weak Supervision](https://arxiv.org/abs/2202.05433)
- [foundryR agreement documentation](https://farach.github.io/foundryR/reference/foundry_agreement.html)
