---
title: "Counting nearby words with n-grams"
subtitle: "See why phrase counts get sparse fast"
description: "Learn how bigrams and trigrams count adjacent tokens in the Riverton sentences, why most of them appear only once, and why a PMI score without a count floor ranks unique pairs first."
categories: [phrases and entities, n-grams, workforce research]
---

Repeated words such as `training` stand out in the Riverton Workforce Lab's
short job-board lines and flyer headings. The harder question is whether
neighbouring word pairs tell a clearer story.

The coordinator tries a simple phrase count before trusting it. If most phrases
appear once, the list may describe these 28 sentences closely while helping less
with new documents.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- define an n-gram as a run of adjacent tokens;
- build bigrams and trigrams with tidytext;
- count repeated n-grams in the Riverton sentences;
- measure how many n-grams appear once;
- score repeated bigrams with pointwise mutual information; and
- explain why phrase vocabularies can generalise badly from small data.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the sentences

This setup chunk reads the sentence file, prepares small tables, builds n-grams
with tidytext, and draws one summary chart. A **tibble** is a table that prints
its size and column types.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(stringr)
library(tidytext)
library(ggplot2)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

source_counts <- sentences |>
  mutate(source = if_else(document_id == "F001", "training flyer", "job details")) |>
  count(source, name = "rows")

knitr::kable(
  source_counts,
  col.names = c("Source", "Rows"),
  caption = "The 28 Riverton sentences by source",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(
  c("readr", "dplyr", "tibble", "tidyr", "stringr", "tidytext", "ggplot2") %in%
    loadedNamespaces()
))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(sum(sentences$document_id == "F001"), 6L),
  identical(sum(sentences$document_id != "F001"), 22L),
  identical(sentences$text[1], "Paid 12-week training is provided."),
  identical(sentences$text[28], "Apply by October 15")
)
```

The file is small enough to inspect by eye. That convenience is also the reason
its phrase counts should be treated as local examples.

## Build adjacent-word phrases

An **n-gram** is a run of `n` adjacent tokens. A unigram has one token, a bigram
has two, and a trigram has three. These counts use the same token rule as Lesson
15: lowercase the text and drop punctuation. Because each row here is one
sentence, no n-gram spans two sentences; a file with whole documents in rows
would need a different check.

```{r}
make_ngrams <- function(n) {
  if (identical(n, 1L)) {
    sentences |>
      select(sentence_id, text) |>
      unnest_tokens(output = ngram, input = text, token = "words") |>
      filter(!is.na(ngram))
  } else {
    sentences |>
      select(sentence_id, text) |>
      unnest_tokens(output = ngram, input = text, token = "ngrams", n = n) |>
      filter(!is.na(ngram))
  }
}

short_for_trigrams <- make_ngrams(1L) |>
  count(sentence_id, name = "tokens") |>
  filter(tokens < 3L)

bigrams <- make_ngrams(2L) |>
  count(ngram, name = "frequency", sort = TRUE)
trigrams <- make_ngrams(3L) |>
  count(ngram, name = "frequency", sort = TRUE)

top_phrases <- bind_rows(
  bigrams |>
    slice_head(n = 8) |>
    mutate(kind = "bigram"),
  trigrams |>
    slice_head(n = 8) |>
    mutate(kind = "trigram")
) |>
  select(kind, ngram, frequency)

knitr::kable(
  top_phrases,
  col.names = c("N-gram type", "N-gram", "Frequency"),
  caption = "The most frequent bigrams and trigrams in the 28 sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(bigrams), 112L),
  identical(nrow(trigrams), 96L),
  identical(short_for_trigrams$sentence_id, "s026"),
  identical(short_for_trigrams$tokens, 2L),
  identical(bigrams$ngram[1], "is required"),
  identical(bigrams$frequency[1], 4L),
  identical(trigrams$ngram[1], "training is provided"),
  identical(trigrams$frequency[1], 2L)
)
```

The sentence `Evening classes` has only two tokens, so it has no trigram. The
code drops that empty result before counting. The most frequent bigram is `is
required`, with 4 appearances, and the most frequent trigram is `training is
provided`, with 2.

## Measure the sparsity

**Sparsity** means that many possible items have no examples or only one example.
The general reason is arithmetic, not this particular table. Each new word adds
at most one new word type, but it can also add a new pair and a new triple; the
set of pairs a vocabulary could form grows faster than the vocabulary itself.
The Riverton counts below illustrate that argument rather than prove it.

```{r}
unigrams <- make_ngrams(1L) |>
  count(ngram, name = "frequency", sort = TRUE)

ngram_counts <- list(unigrams, bigrams, trigrams)

ngram_summary <- tibble(
  n = 1:3,
  name = c("unigrams", "bigrams", "trigrams"),
  distinct_items = vapply(ngram_counts, nrow, integer(1)),
  seen_once = vapply(
    ngram_counts,
    \(counts) sum(counts$frequency == 1L),
    integer(1)
  ),
  seen_once_label = str_c(seen_once, "/", distinct_items)
)

knitr::kable(
  ngram_summary,
  col.names = c("n", "Unit", "Distinct units", "Seen once", "Seen-once fraction"),
  caption = "Distinct n-grams and one-time n-grams in the Riverton sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(ngram_summary$distinct_items, c(96L, 112L, 96L)),
  identical(ngram_summary$seen_once, c(72L, 102L, 95L)),
  identical(ngram_summary$seen_once_label, c("72/96", "102/112", "95/96"))
)
```

The 28 sentences contain 96 distinct unigrams, 112 distinct bigrams, and 96
distinct trigrams. The seen-once fractions are 72/96 for unigrams, 102/112 for
bigrams, and 95/96 for trigrams. These numbers describe only this teaching
corpus.

```{r}
#| label: fig-ngram-sparsity
#| fig-cap: "Share of distinct n-grams seen once in the 28 Riverton sentences."
#| fig-alt: "Bar chart with three bars for unigrams, bigrams, and trigrams. The seen-once share is high for every n-gram size and highest for trigrams."

plot_data <- ngram_summary |>
  transmute(
    name,
    seen_once_share = seen_once / distinct_items,
    seen_once_label
  )

ggplot(plot_data, aes(x = name, y = seen_once_share)) +
  geom_col(fill = "#4C78A8") +
  geom_text(aes(label = seen_once_label), vjust = -0.4) +
  labs(
    x = NULL,
    y = "Share seen once"
  ) +
  scale_y_continuous(labels = \(value) str_c(round(value * 100), "%"), limits = c(0, 1)) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  all(plot_data$seen_once_share > 0),
  all(plot_data$seen_once_share < 1),
  identical(plot_data$seen_once_label[plot_data$name == "trigrams"], "95/96")
)
```

The chart focuses on the quantity the table asks the reader to compare: how much
of each list appears only once.

## Score pairs against chance

Raw frequency likes common words. **Pointwise mutual information**, or **PMI**,
asks whether two adjacent words appear more often than they would if each word
were chosen independently. For a pair `(w1, w2)`:

`PMI = log2( P(w1, w2) / (P(w1) * P(w2)) )`

The probabilities here are token shares in these 28 sentences: 153 unigram
tokens and 125 bigram tokens. A pair of words that each appear once, and that
appear together once, gets a high score because the denominator is tiny. That
is a property of the formula, not evidence that the pair is a stable phrase.

```{r}
unigram_tokens <- make_ngrams(1L)
bigram_tokens <- make_ngrams(2L)
unigram_n <- nrow(unigram_tokens)
bigram_n <- nrow(bigram_tokens)

unigram_counts <- unigram_tokens |>
  count(ngram, name = "word_n")

pmi_bigrams <- bigram_tokens |>
  separate_wider_delim(
    ngram,
    delim = " ",
    names = c("word1", "word2"),
    too_many = "merge"
  ) |>
  count(word1, word2, name = "pair_n") |>
  left_join(unigram_counts, by = join_by(word1 == ngram)) |>
  rename(word1_n = word_n) |>
  left_join(unigram_counts, by = join_by(word2 == ngram)) |>
  rename(word2_n = word_n) |>
  mutate(
    ngram = str_c(word1, word2, sep = " "),
    pmi = log2(
      (pair_n / bigram_n) /
        ((word1_n / unigram_n) * (word2_n / unigram_n))
    )
  )

hapax_singletons <- pmi_bigrams |>
  filter(pair_n == 1L, word1_n == 1L, word2_n == 1L) |>
  arrange(ngram)

repeated_pmi <- pmi_bigrams |>
  filter(pair_n >= 2L) |>
  arrange(desc(pmi), desc(pair_n), ngram)

hapax_singleton_pmi <- log2(
  (1 / bigram_n) / ((1 / unigram_n) * (1 / unigram_n))
)

is_required <- pmi_bigrams |>
  filter(ngram == "is required")

knitr::kable(
  hapax_singletons |>
    slice_head(n = 3) |>
    transmute(
      ngram,
      pair_n,
      pmi = round(pmi, 2)
    ),
  col.names = c("Bigram", "Pair count", "PMI"),
  caption = "Three one-time bigrams whose words also appear only once",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(unigram_n, 153L),
  identical(bigram_n, 125L),
  identical(nrow(pmi_bigrams), 112L),
  identical(sum(pmi_bigrams$pair_n == 1L), 102L),
  identical(nrow(repeated_pmi), 10L),
  identical(nrow(hapax_singletons), 37L),
  abs(hapax_singletons$pmi[1] - hapax_singleton_pmi) < 1e-10,
  identical(hapax_singletons$ngram[1], "50 pounds"),
  identical(repeated_pmi$ngram[1], "no prior"),
  identical(is_required$pair_n, 4L),
  is_required$pmi < hapax_singleton_pmi,
  is_required$pmi < repeated_pmi$pmi[1]
)
```

Thirty-seven of the 102 one-time bigrams are pairs of words that each appear once.
They all receive PMI 7.55. The most frequent bigram, `is required` (4 times),
scores 3.04 because `is` appears 13 times and `required` appears 7 times. A
floor of two pair counts leaves ten bigrams.

```{r}
repeated_pmi_table <- repeated_pmi |>
  transmute(
    ngram,
    pair_n,
    pmi = round(pmi, 2)
  )

knitr::kable(
  repeated_pmi_table,
  col.names = c("Bigram", "Pair count", "PMI"),
  caption = "Bigrams seen at least twice, ranked by PMI",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(repeated_pmi_table), 10L),
  identical(repeated_pmi_table$ngram[1], "no prior"),
  identical(repeated_pmi_table$pair_n[repeated_pmi_table$ngram == "is required"], 4L),
  identical(repeated_pmi_table$pmi[1], 6.55),
  identical(
    repeated_pmi_table$pmi[repeated_pmi_table$ngram == "is required"],
    3.04
  )
)
```

With that floor, `no prior` leads and `is required` falls to ninth. Both lists
still describe 28 teaching sentences. PMI does not repair sparsity; it changes
which sparse pairs look interesting. For matches in running text, Lesson 25 uses
a concordance instead of a score.

## What to remember

- An n-gram counts adjacent tokens after a chosen token rule.
- The Riverton rows contain 112 distinct bigrams and 96 distinct trigrams.
- A two-token sentence contributes no trigram.
- PMI without a count floor ranks unique word pairs first.
- Counts from 28 sentences illustrate sparsity; they do not estimate phrase use in workforce writing.

Report the repeated phrases as local clues, then stop. The useful result here is
a small list of repeats paired with a warning about sparse phrase counts.

## Sources

- [tidytext documentation for `unnest_tokens()`](https://juliasilge.github.io/tidytext/reference/unnest_tokens.html)
- [tokenizers documentation for n-gram tokenizers](https://docs.ropensci.org/tokenizers/reference/ngram-tokenizers.html)
- [Church and Hanks (1990), Word association norms, mutual information, and lexicography](https://aclanthology.org/J90-1003/)
- [ggplot2 documentation](https://ggplot2.tidyverse.org/)
