---
title: "Finding phrases with readable rules"
subtitle: "Score a credential pattern before trusting it"
author: "Alex Farach"
description: "Learn how simple string rules find credential phrases, how concordances help inspection, and why rules need maintenance."
categories: [phrases and entities, rule-based matching, workforce research]
---

Credential requirements are easy to miss in a short job-board feed. The
Riverton Workforce Lab starts with a readable rule, then checks whether the rule
answers the same question a reviewer would ask.

That readability is useful only if the rule is checked. A pattern can find real
mentions, miss wording it did not include, and catch lines that use the same word
for a different purpose.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

After this lesson, you should be able to:

- write small phrase rules with stringr;
- extract the matched credential word;
- apply a written hand-marking rule for required credentials;
- count false positives and false negatives without treating the counts as a rate; and
- use a concordance to inspect matches in context.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Write the first rules

The setup chunk reads the Riverton sentences, builds rule-result tables, handles
regular expressions, and prepares a concordance. In stringr, regular expressions
use the ICU engine, which defines how boundaries and letter classes behave.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(stringr)
library(quanteda)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

credential_terms <- c("certificate", "certification", "licen[cs]e")
credential_pattern <- regex(
  str_c("\\b(", str_c(credential_terms, collapse = "|"), ")\\b"),
  ignore_case = TRUE
)

rule_results <- sentences |>
  mutate(
    rule_match = str_detect(text, credential_pattern),
    matched_word = str_extract(text, credential_pattern),
    hand_required_credential = sentence_id %in% c("s005", "s017")
  )

matched_rows <- rule_results |>
  filter(rule_match | hand_required_credential) |>
  select(sentence_id, text, matched_word, rule_match, hand_required_credential)

knitr::kable(
  matched_rows,
  col.names = c("Sentence ID", "Text", "Matched word", "Rule matched", "Hand-marked required credential"),
  caption = "Rule matches and hand markings for required credentials",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(
  c("readr", "dplyr", "tibble", "tidyr", "stringr", "quanteda") %in%
    loadedNamespaces()
))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(sum(rule_results$rule_match), 4L),
  identical(sum(rule_results$hand_required_credential), 2L),
  identical(
    matched_rows$sentence_id,
    c("s005", "s006", "s013", "s017", "s024")
  ),
  identical(
    as.character(na.omit(matched_rows$matched_word)),
    c("certification", "certificate", "license", "CERTIFICATE")
  )
)
```

The rule searches for `certificate`, `certification`, and either spelling of
`license`. The lesson author hand-marked each sentence against this written
definition: mark the sentence if it says the employer requires a formal
qualification issued by a school, licensing body, or certifying organisation,
and do not mark a sentence that merely mentions one. Under that rule, `s005`
and `s017` qualify. One reviewer applying one rule to 28 sentences gives a
worked example, not a reference standard.

## Score the rule

A **false positive** is a sentence the rule flags but the hand marking rejects. A
**false negative** is a sentence the hand marking accepts but the rule misses.
The rule and the hand marking are not asking the same question: the pattern
looks for a credential word, while the hand marking asks whether a credential is
required. Nothing in the pattern tests for requirement, so it can fire on
`preferred` and on a heading.

```{r}
score_counts <- rule_results |>
  summarise(
    true_positive = sum(rule_match & hand_required_credential),
    false_positive = sum(rule_match & !hand_required_credential),
    false_negative = sum(!rule_match & hand_required_credential),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "result",
    values_to = "sentences"
  )

false_positive_ids <- rule_results |>
  filter(rule_match & !hand_required_credential) |>
  pull(sentence_id)
false_negative_ids <- rule_results |>
  filter(!rule_match & hand_required_credential) |>
  pull(sentence_id)

problem_examples <- tibble(
  result = c("false positive", "false negative"),
  sentence_id = c("s006", "s005"),
  explanation = c(
    "The line says the employer pays for certification training, not that certification is required.",
    "The line says a diploma is required, but diploma was not in the rule."
  )
)

knitr::kable(
  score_counts,
  col.names = c("Result", "Sentences"),
  caption = "Sentence-level counts for the first credential rule",
  row.names = FALSE
)

knitr::kable(
  problem_examples,
  col.names = c("Result", "Sentence ID", "Explanation"),
  caption = "One false positive and one false negative",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(score_counts$sentences, c(1L, 3L, 1L)),
  identical(false_positive_ids, c("s006", "s013", "s024")),
  identical(false_negative_ids, "s005"),
  problem_examples$sentence_id[problem_examples$result == "false positive"] %in% false_positive_ids,
  problem_examples$sentence_id[problem_examples$result == "false negative"] %in% false_negative_ids,
  identical(rule_results$matched_word[rule_results$sentence_id == "s017"], "license")
)
```

This is a bad rule for required credentials: 3 of its 4 hits are wrong, and it
misses 1 of the 2 hand-marked positives. Two positives cannot measure how well a
rule works. These counts only show failure modes: `certificate` and
`certification` can appear without a requirement, while `diploma` is absent from
the pattern.

## Inspect matches in context

A **concordance** is a list of every place a word occurs in a text. The usual
layout, called keyword in context, or **KWIC**, puts the matched word in the
middle with a few words on each side.

```{r}
riverton_corpus <- corpus(setNames(sentences$text, sentences$sentence_id))
riverton_tokens <- tokens(riverton_corpus, remove_punct = TRUE)

credential_kwic <- kwic(
  riverton_tokens,
  pattern = credential_terms,
  window = 4,
  valuetype = "regex",
  case_insensitive = TRUE
) |>
  as.data.frame() |>
  as_tibble() |>
  select(docname, pre, keyword, post)

knitr::kable(
  credential_kwic,
  col.names = c("Sentence ID", "Before", "Keyword", "After"),
  caption = "Concordance lines for the credential rule",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(credential_kwic), 4L),
  identical(credential_kwic$docname, c("s006", "s013", "s017", "s024")),
  identical(str_to_lower(credential_kwic$keyword), c("certification", "certificate", "license", "certificate")),
  identical(credential_kwic$post[credential_kwic$docname == "s006"], "training")
)
```

The KWIC display makes the maintenance problem visible. The token-level regex
uses the same credential terms as the scoring pattern: it finds `certification
training`, a `certificate` that is preferred, a `license` that is required, and a
flyer heading.

## What to remember

- A readable rule can still answer the wrong question.
- This first credential rule has 1 true positive, 3 false positives, and 1 false negative.
- Two hand-marked positives are enough for a worked example, not for a performance estimate.
- KWIC lines help reviewers see why a pattern fired.

Use this pattern only as a review aid. Its next revision should test requirement
language before widening the credential word list.

## Sources

- [stringr regular expression documentation](https://stringr.tidyverse.org/articles/regular-expressions.html)
- [quanteda `kwic()` documentation](https://quanteda.io/reference/kwic.html)
- [ICU regular expression documentation](https://unicode-org.github.io/icu/userguide/strings/regexp.html)
