---
title: "Finding noun chunks from grammar links"
subtitle: "Compare spaCy phrases with small dependency chunks"
author: "Alex Farach"
description: "Learn how noun chunks group nouns with attached words, using spaCy and a local UDPipe dependency parse."
categories: [phrases and entities, noun chunks, workforce research]
---

After counting adjacent phrases, the Riverton Workforce Lab has a practical
choice to make: should `basic spreadsheet skills` become one review item or
three separate words?

A grammar-aware method offers one convention for making that choice. It can
group candidate phrases for review, but it does not create a single correct
inventory of English phrases.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

Use this lesson to practice how to:

- define a noun chunk in plain language;
- extract noun chunks with a trained spaCy pipeline;
- build simple chunks from a UDPipe dependency parse;
- explain why chunk definitions differ across tools;
- separate agreement from accuracy; and
- explain why parser errors affect chunk rules.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Ask a trained pipeline for noun chunks

This setup chunk reads the sentence file, runs spaCy, and later supports a small
UDPipe-based rule. A **noun chunk** is a short, flat phrase built around one
noun: the noun plus the determiners, numbers, compounds, and describing words
directly in front of it. Chunks do not nest and do not reach past the noun, so
`evening classes in October` becomes `evening classes` and `October`, not one
long phrase. Where a chunk stops is a convention, and tools draw the line in
different places.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(udpipe)
library(spacyr)
source("R/use-spacy.R")

stopifnot(all(
  c("readr", "dplyr", "tibble", "purrr", "stringr", "tokenizers", "udpipe", "spacyr") %in%
    loadedNamespaces()
))

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

pipeline <- use_project_spacy()
pipeline_info <- spacy_pipeline_version()
parsed_spacy <- spacy_parse(
  setNames(sentences$text, sentences$sentence_id),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
)
spacy_phrases <- nounphrase_extract(parsed_spacy) |>
  as_tibble() |>
  transmute(doc_id, chunk = str_replace_all(nounphrase, "_", " "))

example_text <- "Riverton Skills Centre will run evening classes in October."
example_parsed <- spacy_parse(
  c(example = example_text),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
)
example_phrases <- nounphrase_extract(example_parsed) |>
  as_tibble() |>
  transmute(chunk = str_replace_all(nounphrase, "_", " "))

spacy_finalize()

pipeline_table <- tibble(
  field = names(pipeline_info),
  value = unlist(pipeline_info, use.names = FALSE)
)

stopifnot(
  identical(nrow(sentences), 28L),
  identical(pipeline$model, "en_core_web_sm"),
  identical(pipeline_info$version, "3.8.0"),
  identical(pipeline_info$license, "MIT"),
  identical(pipeline_info$spacy, "3.8.7"),
  identical(nrow(parsed_spacy), 179L),
  identical(nrow(spacy_phrases), 40L),
  identical(spacy_phrases$chunk[spacy_phrases$doc_id == "s001"], "Paid 12-week training"),
  identical(spacy_phrases$chunk[spacy_phrases$doc_id == "s017"], "A valid driver's license"),
  identical(
    example_phrases$chunk,
    c("Riverton Skills Centre", "evening classes", "October")
  )
)

knitr::kable(
  pipeline_table,
  col.names = c("Pipeline field", "Value"),
  caption = "spaCy pipeline used for noun chunks",
  row.names = FALSE
)

knitr::kable(
  spacy_phrases |>
    slice_head(n = 12),
  col.names = c("Sentence ID", "spaCy noun chunk"),
  caption = "First spaCy noun chunks from the Riverton sentences",
  row.names = FALSE
)
```

The released spaCy pipeline returns 40 noun chunks for the 28 sentences. For the
check sentence, it returns three flat chunks: `Riverton Skills Centre`, `evening
classes`, and `October`.

## Build chunks from dependency links

The UDPipe model used here was trained for this site on 500 sentences. On 13
held-out political weblog posts from 2003 to 2006, its unlabelled attachment
score was 0.7235 and its labelled attachment score was 0.6561. spaCy's pipeline
is a released production model trained on far more text. This comparison is a
teaching device, not a contest between equals.

The local UDPipe model has no tokenizer, so the lesson gives it one token per
line. In the code, `\(tokens)` is R shorthand for a small function that handles
one sentence's tokens at a time.

This classroom rule starts from each token tagged as `NOUN` or `PROPN`. It then
collects direct dependents whose relation is `det`, `amod`, `compound`, or
`nummod`: a determiner such as `a` or `the`, a describing word, a noun working
with another noun, or a number modifying a noun.

```{r}
parser <- udpipe_load_model("data/treebank/en_ewt-500-parser.udpipe")

vertical <- sentences$text |>
  tokenize_words(lowercase = FALSE, strip_punct = FALSE) |>
  map_chr(\(tokens) paste(tokens, collapse = "\n"))

annotated <- udpipe_annotate(
  parser,
  x = vertical,
  doc_id = sentences$sentence_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "default"
) |>
  as.data.frame() |>
  as_tibble()

token_count_gap <- parsed_spacy |>
  count(doc_id, name = "spacy_tokens") |>
  inner_join(
    annotated |>
      count(doc_id, name = "udpipe_tokens"),
    by = "doc_id"
  ) |>
  filter(spacy_tokens != udpipe_tokens)

chunk_relations <- c("det", "amod", "compound", "nummod")

collect_chunk <- function(current_doc_id, current_token_id) {
  annotated |>
    filter(
      doc_id == current_doc_id,
      token_id == current_token_id |
        (head_token_id == current_token_id & dep_rel %in% chunk_relations)
    ) |>
    arrange(as.integer(token_id)) |>
    pull(token) |>
    str_c(collapse = " ")
}

udpipe_chunks <- annotated |>
  filter(upos %in% c("NOUN", "PROPN")) |>
  transmute(
    doc_id,
    token_id,
    chunk = map2_chr(doc_id, token_id, collect_chunk)
  ) |>
  select(doc_id, chunk)

stopifnot(
  identical(nrow(annotated), 178L),
  identical(token_count_gap$doc_id, "s017"),
  identical(token_count_gap$spacy_tokens, 8L),
  identical(token_count_gap$udpipe_tokens, 7L),
  identical(nrow(udpipe_chunks), 56L),
  identical(udpipe_chunks$chunk[1:5], c("Paid", "Paid week", "week training", "No prior data experience", "Evening schedules")),
  identical(annotated$upos[annotated$doc_id == "s001" & annotated$token == "Paid"], "PROPN"),
  identical(udpipe_chunks$chunk[udpipe_chunks$doc_id == "s027"], "No prior experience")
)

knitr::kable(
  token_count_gap,
  col.names = c("Sentence ID", "spaCy tokens", "UDPipe tokens"),
  caption = "The one token-count difference between the two parses",
  row.names = FALSE
)

knitr::kable(
  udpipe_chunks |>
    slice_head(n = 12),
  col.names = c("Sentence ID", "UDPipe-built chunk"),
  caption = "First noun chunks built from dependency links",
  row.names = FALSE
)
```

The two systems report different row counts because they split words
differently. spaCy separates `driver's` in `s017` into `driver` and `'s`; the
tokenizer feeding UDPipe keeps it as one token, which accounts for the 179 versus
178 row counts. The printed spaCy chunk strings still keep `Paid 12-week
training` and `A valid driver's license`; the gap is in token rows, not those
displayed chunk strings.

This rule emits one chunk for every noun-like tag, so chunks can overlap. `Paid`,
`Paid week`, and `week training` all come from the same four words in `s001`.
The rule also pastes a head to its dependents without checking that the words sit
next to each other, so `Paid week` skips over `12-` and is not a string in the
sentence. `Paid` appears because the small tagger labelled it `PROPN`, which is
wrong here: it describes the training. A chunk rule inherits tagging mistakes
before it makes its own.

## Compare the two lists

Exact agreement is strict: the sentence ID and the chunk text must both match.
That makes agreement a string match, not accuracy. Neither list is ground truth,
and spaCy chunks are non-overlapping while the UDPipe-built list can overlap.

```{r}
agreed_chunks <- inner_join(
  distinct(spacy_phrases),
  distinct(udpipe_chunks),
  by = c("doc_id", "chunk")
)

comparison_summary <- tibble(
  measure = c("spaCy chunks", "UDPipe-built chunks", "exact agreements"),
  count = c(nrow(spacy_phrases), nrow(udpipe_chunks), nrow(agreed_chunks))
)

different_examples <- tibble(
  method = c("spaCy", "UDPipe-built"),
  sentence_id = c("s005", "s005"),
  chunk = c("A high school diploma", "A high school"),
  explanation = c(
    "spaCy keeps the full education phrase.",
    "The dependency rule attaches words to school and leaves diploma outside this chunk."
  )
)

spreadsheet_examples <- bind_rows(
  spacy_phrases |>
    filter(doc_id == "s004", str_detect(chunk, "spreadsheet")) |>
    mutate(method = "spaCy"),
  udpipe_chunks |>
    filter(doc_id == "s004", str_detect(chunk, "spreadsheet")) |>
    mutate(method = "UDPipe-built")
) |>
  select(method, doc_id, chunk)

stopifnot(
  identical(nrow(spacy_phrases), 40L),
  identical(nrow(udpipe_chunks), 56L),
  identical(nrow(agreed_chunks), 25L),
  identical(spacy_phrases$chunk[spacy_phrases$doc_id == "s005"], "A high school diploma"),
  identical(udpipe_chunks$chunk[udpipe_chunks$doc_id == "s005"], "A high school"),
  identical(agreed_chunks$chunk[1], "No prior data experience"),
  identical(
    spreadsheet_examples$chunk,
    c("basic spreadsheet skills", "spreadsheet", "need basic spreadsheet skills")
  )
)

knitr::kable(
  comparison_summary,
  col.names = c("Measure", "Count"),
  caption = "String matches across two chunking conventions",
  row.names = FALSE
)

knitr::kable(
  different_examples,
  col.names = c("Method", "Sentence ID", "Chunk", "Explanation"),
  caption = "One sentence where the two chunk methods differ",
  row.names = FALSE
)

knitr::kable(
  spreadsheet_examples,
  col.names = c("Method", "Sentence ID", "Chunk"),
  caption = "How each method handles the spreadsheet-skills phrase",
  row.names = FALSE
)
```

The two lists have 25 exact string matches. That is agreement, not accuracy, and
the counts 40 and 56 are not two measurements of the same thing. For the opening
question, spaCy groups `basic spreadsheet skills`, while the UDPipe-built rule
also emits `spreadsheet` and overreaches to `need basic spreadsheet skills`. A
bigram takes neighbouring words even across phrase boundaries; a chunk groups
words the parser links together. Whether that helps depends on whether the
parser is right.

## What to remember

- Noun chunks are flat, non-overlapping spans under a tool-specific convention.
- spaCy returns 40 noun chunks for these 28 sentences.
- The UDPipe-built rule returns 56 overlapping chunks from a deliberately weak teaching model.
- The 25 exact matches are agreement, not accuracy.
- `Paid week` shows how a rule can manufacture text by pasting non-adjacent tokens.
- Tokenization alone explains the 179 versus 178 row-count gap.

Review chunk lists as prompts for human inspection. When the two tools disagree,
check the sentence, the tokenization, and the parse before drawing any conclusion
about the phrase.

## Sources

- [spaCy documentation for linguistic features](https://spacy.io/usage/linguistic-features)
- [spacyr package documentation](https://spacyr.quanteda.io/)
- [UDPipe R package reference](https://bnosac.github.io/udpipe/docs/doc0.html)
- [Universal Dependencies relation documentation](https://universaldependencies.org/u/dep/)
