---
title: "Checking named entities against domain text"
subtitle: "See what a trained NER model gets right and wrong"
author: "Alex Farach"
description: "Learn how spaCy labels named entities in the Riverton sentences and why invented local names are hard."
categories: [phrases and entities, named entity recognition, workforce research]
---

Names in the Riverton job-board and flyer text need labels, not just highlights.
The coordinator wants to know whether a span names an organisation, a date, a
time, a quantity, or something else.

A trained named-entity model can supply those labels quickly. The question is
whether its labels survive contact with local names that did not appear in its
training examples.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define named entity recognition in plain language;
- run spaCy's small English pipeline through spacyr;
- read spaCy entity labels such as PERSON, ORG, DATE, TIME, QUANTITY, and GPE;
- hand-check entity spans from the 28 Riverton sentences; and
- explain why local invented names need domain review.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Run the released pipeline

The setup chunk reads the sentences, runs the released spaCy pipeline, and
formats entity spans for inspection. **Named entity recognition**, or **NER**,
finds names and assigns each one a type.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)
library(spacyr)
source("R/use-spacy.R")

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

pipeline <- use_project_spacy()
pipeline_info <- spacy_pipeline_version()
parsed <- spacy_parse(
  setNames(sentences$text, sentences$sentence_id),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
)
entities <- entity_extract(parsed, type = "all") |>
  as_tibble() |>
  transmute(
    sentence_id = doc_id,
    entity = str_replace_all(entity, "_", " "),
    entity_type
  )

example_text <- "Riverton Skills Centre will run evening classes in October."
example_parsed <- spacy_parse(
  c(example = example_text),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
)
example_entities <- entity_extract(example_parsed, type = "all") |>
  as_tibble() |>
  transmute(
    entity = str_replace_all(entity, "_", " "),
    predicted_type = entity_type,
    actual_type = c("ORG", "TIME", "DATE"),
    correct = predicted_type == actual_type
  )

title_case_heading <- "Riverton Skills Open House"
title_case_parsed <- spacy_parse(
  c(title_case = title_case_heading),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
)
title_case_entities <- entity_extract(title_case_parsed, type = "all") |>
  as_tibble() |>
  transmute(
    entity = str_replace_all(entity, "_", " "),
    entity_type
  )

spacy_finalize()

pipeline_table <- tibble(
  field = names(pipeline_info),
  value = unlist(pipeline_info, use.names = FALSE)
)

knitr::kable(
  pipeline_table,
  col.names = c("Pipeline field", "Value"),
  caption = "spaCy pipeline used for named entity recognition",
  row.names = FALSE
)

knitr::kable(
  example_entities,
  col.names = c("Entity", "Predicted type", "Actual type", "Correct"),
  caption = "NER output for a Riverton Skills Centre check sentence",
  row.names = FALSE
)

knitr::kable(
  title_case_entities,
  col.names = c("Entity", "Entity type"),
  caption = "spaCy output for a title-case version of the flyer heading",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(
  c("readr", "dplyr", "tibble", "stringr", "spacyr") %in%
    loadedNamespaces()
))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(pipeline$model, "en_core_web_sm"),
  identical(pipeline_info$version, "3.8.0"),
  identical(pipeline_info$license, "MIT"),
  identical(pipeline_info$spacy, "3.8.7"),
  identical(nrow(parsed), 179L),
  identical(example_entities$entity, c("Riverton Skills Centre", "evening", "October")),
  identical(example_entities$predicted_type, c("PERSON", "TIME", "DATE")),
  identical(example_entities$correct, c(FALSE, TRUE, TRUE)),
  identical(title_case_entities$entity, "Riverton Skills Open House"),
  identical(title_case_entities$entity_type, "PERSON")
)
```

The check sentence exposes the main lesson. spaCy labels `Riverton Skills Centre`
as `PERSON`, which is wrong in this passage; it names the provider, so the hand
check treats it as `ORG`. A facility reading would point toward `FAC`. The
invented name explains the domain mismatch, but it does not make the output
correct. A title-case version of the all-capital heading becomes one `PERSON`
span, a small reminder that surface form can change the model output.

## Read the labels

These type names come from OntoNotes, one annotation project's category list.
They are conventions, not natural divisions of the world. `PERSON` means a
person. `ORG` covers organisations such as companies, agencies, and
institutions. `DATE` covers dates and periods such as months or years. `TIME`
marks a time of day or similar expression. `QUANTITY` marks an amount with a
unit. `GPE` covers a geopolitical place such as a city, state, or country. The
same word can be a `GPE` in `Riverton raised the training budget` and part of an
`ORG` in `Riverton Skills Centre`, so the sentence matters.

```{r}
entity_type_counts <- entities |>
  count(entity_type, name = "entities", sort = TRUE)

knitr::kable(
  entity_type_counts,
  col.names = c("Entity type", "Entities"),
  caption = "Entity types returned for the 28 Riverton sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(entities), 10L),
  identical(entity_type_counts$entity_type, c("DATE", "TIME", "ORG", "QUANTITY")),
  identical(entity_type_counts$entities, c(6L, 2L, 1L, 1L)),
  !any(entities$entity_type %in% c("CARDINAL", "GPE")),
  identical(entities$entity[entities$sentence_id == "s023"], "HOUSE")
)
```

The model finds 10 entity spans in the 28 sentences. Six are `DATE`, two are
`TIME`, one is `ORG`, and one is `QUANTITY`. It returns nothing for `RIVERTON`
in the all-capital heading. That is defensible: here `RIVERTON` is part of an
event name rather than a reference to the town. The mistake in that sentence is
`HOUSE`: the model takes one word out of a four-word event heading and calls it
an organisation, so the span and the type are wrong.

## Hand-check the Riverton entities

The table below marks the model's returned spans by hand. The hand labels are
teaching judgments for these 28 sentences, not a public benchmark.

```{r}
entity_answers <- tibble(
  sentence_id = c("s001", "s003", "s008", "s010", "s011", "s014", "s019", "s020", "s023", "s028"),
  entity = c("12 - week", "Evening", "50 pounds", "Six months", "two days", "daytime", "Two years", "Weekend", "HOUSE", "October 15"),
  expected_span = c("12 - week", "Evening", "50 pounds", "Six months", "two days", "daytime", "Two years", "Weekend", "RIVERTON SKILLS OPEN HOUSE", "October 15"),
  expected_type = c("DATE", "TIME", "QUANTITY", "DATE", "DATE", "TIME", "DATE", "DATE", "EVENT", "DATE"),
  note = c(
    rep("The span and type fit this sentence.", 8),
    "The event heading is four words; `HOUSE` alone is the wrong span and `ORG` is the wrong type.",
    "The span and type fit this sentence."
  )
)

entity_judgments <- entities |>
  left_join(entity_answers, by = c("sentence_id", "entity")) |>
  mutate(
    correct = entity == expected_span & entity_type == expected_type
  )

entity_score <- entity_judgments |>
  summarise(
    returned_spans = n(),
    correct_spans = sum(correct),
    wrong_spans = sum(!correct),
    .groups = "drop"
  )

knitr::kable(
  entity_judgments,
  col.names = c("Sentence ID", "Entity", "Predicted type", "Expected span", "Expected type", "Note", "Correct"),
  caption = "Hand check of all spaCy entity spans returned from the 28 sentences",
  row.names = FALSE
)

knitr::kable(
  entity_score,
  col.names = c("Returned spans", "Correct spans", "Wrong spans"),
  caption = "Hand-check counts for returned entity spans",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(entity_score$returned_spans, 10L),
  identical(entity_score$correct_spans, 9L),
  identical(entity_score$wrong_spans, 1L),
  !any(is.na(entity_judgments$expected_type)),
  identical(entity_judgments$correct[entity_judgments$sentence_id == "s023"], FALSE),
  identical(entity_judgments$expected_type[entity_judgments$sentence_id == "s023"], "EVENT"),
  identical(entity_judgments$expected_type[entity_judgments$sentence_id == "s028"], "DATE")
)
```

Nine of the ten returned spans hold up, but that is not an evaluation. Eight of
the correct spans are dates, durations, or times of day, and one is a weight.
The only returned organisation-like case is the one the model gets wrong. A
single figure hides that mix, and the check says nothing about names the model
missed.

## Keep domain mismatch visible

spaCy `en_core_web_sm` 3.8.0 is a released production pipeline. The local
Riverton names are invented for this lesson, so the model cannot rely on having
seen them in training. It has to generalise from surface cues, word shape, and
surrounding words. That is domain mismatch, not proof that the pipeline is weak.

## What to remember

- NER labels spans under an annotation scheme such as OntoNotes.
- The spaCy run returned 10 entity spans from the 28 Riverton sentences.
- Ten returned spans are too few, and too uneven by type, to evaluate a model.
- Local invented names need direct review because domain mismatch can change outputs.

For Riverton work, keep NER in the triage step: collect candidates, inspect the
local names, and report the model version beside the examples.

## Sources

- [spaCy named entity recognition documentation](https://spacy.io/usage/linguistic-features#named-entities)
- [spaCy model documentation for English pipelines](https://spacy.io/models/en)
- [spacyr package documentation](https://spacyr.quanteda.io/)
