---
title: "Finding abbreviations without trusting capitals"
subtitle: "Compare all-capital tokens with parenthetical definitions"
description: "Learn why capital letters create false abbreviation hits and how parenthetical long-form rules behave on constructed examples."
categories: [phrases and entities, abbreviations, workforce research]
---

Short forms only help the Riverton Workforce Lab if they can be tied back to
their long forms. A tempting shortcut is to look for all-capital tokens, then
ask whether any of them are real abbreviations.

The flyer immediately tests that shortcut. Its headings are printed in capitals,
so a rule that treats capitals as an abbreviation signal may confuse design with
meaning.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

In this lesson, you will practice how to:

- find all-capital tokens with a regular expression;
- hand-check capitalised heading hits;
- test what a dictionary-and-length guard removes;
- find `Long Form (SHORT)` definitions; and
- separate real data from constructed examples.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Find all-capital tokens

The setup chunk reads the sentence file, expands regex matches, and checks words
against hunspell. stringr regular expressions use the ICU engine, but the
`[A-Z]` range below is still an ASCII capital-letter rule; it will not match a
token such as `ÉCOLE`.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(hunspell)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

all_caps_pattern <- "\\b[A-Z]{2,}\\b"
all_caps_hits <- sentences |>
  transmute(
    sentence_id,
    text,
    token = str_extract_all(text, all_caps_pattern)
  ) |>
  unnest(token)

knitr::kable(
  all_caps_hits,
  col.names = c("Sentence ID", "Text", "All-capital token"),
  caption = "All-capital tokens found in the Riverton text",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(
  c("readr", "dplyr", "tibble", "tidyr", "purrr", "stringr", "hunspell") %in%
    loadedNamespaces()
))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(all_caps_hits), 7L),
  identical(
    all_caps_hits$token,
    c("RIVERTON", "SKILLS", "OPEN", "HOUSE", "DATA", "SUPPORT", "CERTIFICATE")
  ),
  identical(unique(all_caps_hits$sentence_id), c("s023", "s024"))
)
```

An **abbreviation** is any shortened written form, including `Dr.` and
`approx.`. An **initialism** is built from initial letters and read letter by
letter, like `NLP`. An **acronym** is built from initial letters and read as a
word, like `NATO`. This rule finds none of those categories directly; it finds
runs of two or more ASCII capital letters, a different thing that sometimes
overlaps. Here it finds 7 tokens, all from the two flyer headings `RIVERTON
SKILLS OPEN HOUSE` and `DATA SUPPORT CERTIFICATE`. The Riverton text contains no
genuine short-form abbreviation in those headings.

## Score and guard the capital rule

The lesson author hand-checked every all-capital hit in the Riverton text and
found no true short-form abbreviations. The guard below keeps only tokens that
are five characters or shorter and absent from the dictionary.

```{r}
scored_caps <- all_caps_hits |>
  mutate(
    true_abbreviation = FALSE,
    token_lower = str_to_lower(token),
    in_dictionary = hunspell_check(token_lower),
    longer_than_five = str_length(token) > 5L,
    guarded_match = !longer_than_five & !in_dictionary,
    removal_reason = case_when(
      in_dictionary & longer_than_five ~ "dictionary word; longer than five",
      in_dictionary ~ "dictionary word",
      longer_than_five ~ "longer than five",
      TRUE ~ "kept"
    )
  )

capital_score <- tibble(
  rule = c("all-capital token", "short and not in dictionary"),
  candidates = c(nrow(scored_caps), sum(scored_caps$guarded_match)),
  hand_rejected = c(
    sum(!scored_caps$true_abbreviation),
    sum(scored_caps$guarded_match & !scored_caps$true_abbreviation)
  ),
  true_abbreviations_found = c(
    sum(scored_caps$true_abbreviation),
    sum(scored_caps$guarded_match & scored_caps$true_abbreviation)
  )
)

guard_details <- scored_caps |>
  select(sentence_id, token, in_dictionary, longer_than_five, guarded_match, removal_reason)

guard_failure_examples <- tibble(token = c("UNESCO", "IT")) |>
  mutate(
    token_lower = str_to_lower(token),
    in_dictionary = hunspell_check(token_lower),
    longer_than_five = str_length(token) > 5L,
    guarded_match = !longer_than_five & !in_dictionary
  )

knitr::kable(
  capital_score,
  col.names = c("Rule", "Candidates kept", "Hand-rejected hits", "True abbreviations found"),
  caption = "Hand review of the capital-letter abbreviation rule before and after a guard",
  row.names = FALSE
)

knitr::kable(
  guard_details,
  col.names = c("Sentence ID", "Token", "In dictionary", "Longer than five", "Kept by guard", "Removal reason"),
  caption = "Dictionary guard details for the all-capital tokens",
  row.names = FALSE
)

knitr::kable(
  guard_failure_examples,
  col.names = c("Token", "Lowercase token", "In dictionary", "Longer than five", "Kept by guard"),
  caption = "Two real short forms the guard would reject",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(capital_score$candidates, c(7L, 0L)),
  identical(capital_score$hand_rejected, c(7L, 0L)),
  identical(capital_score$true_abbreviations_found, c(0L, 0L)),
  identical(sum(scored_caps$in_dictionary), 6L),
  identical(sum(scored_caps$longer_than_five), 4L),
  identical(any(scored_caps$guarded_match), FALSE),
  identical(scored_caps$in_dictionary[scored_caps$token == "RIVERTON"], FALSE),
  identical(scored_caps$longer_than_five[scored_caps$token == "RIVERTON"], TRUE),
  identical(guard_failure_examples$guarded_match, c(FALSE, FALSE))
)
```

The guard keeps zero candidates. In this file that means it removes 7
hand-rejected heading words and finds 0 true abbreviations. A filter that removes
every hit is not evidence of a good filter: six tokens are dictionary words, four
are longer than five characters, and `RIVERTON` is removed only by the length
limit. The same guard rejects `UNESCO` on length and `IT` because `it` is a
dictionary word. The dictionary result also depends on the hunspell dictionary
available at run time.

## Look for parenthetical definitions

The Riverton text has no true `Long Form (SHORT)` examples, so this section uses
a tiny author-created passage. It is constructed only to show how the pattern
works.

In the code, `\(match_table, current_sentence_id)` is R shorthand for a small
function that handles one sentence's match table and ID.

```{r}
definition_pattern <- regex(
  "\\b((?:[A-Z][a-z]+\\s+){1,}[A-Z][a-z]+)\\s*\\(([A-Z]{2,})\\)"
)

extract_definitions <- function(text, sentence_id) {
  match_tables <- str_match_all(text, definition_pattern)
  pieces <- map2(match_tables, sentence_id, \(match_table, current_sentence_id) {
    if (nrow(match_table) == 0L) {
      return(tibble(
        sentence_id = character(),
        long_form = character(),
        short_form = character()
      ))
    }
    tibble(
      sentence_id = current_sentence_id,
      long_form = str_squish(match_table[, 2]),
      short_form = match_table[, 3]
    )
  })
  list_rbind(pieces)
}

riverton_definitions <- extract_definitions(sentences$text, sentences$sentence_id)
riverton_parenthetical_score <- tibble(
  source = "Riverton text",
  definitions_found = nrow(riverton_definitions)
)

constructed_passage <- tibble(
  sentence_id = c("c001", "c002"),
  text = c(
    "The note defines Natural Language Processing (NLP) for the workshop.",
    "The class lists a Data Support Certificate (DSC) for applicants."
  )
)

definitions <- extract_definitions(
  constructed_passage$text,
  constructed_passage$sentence_id
)

shape_only_example <- extract_definitions(
  "Riverton Skills Centre (NLP)",
  "c003"
)

parenthetical_score <- tibble(
  true_definitions_in_constructed_passage = 2L,
  definitions_found = nrow(definitions)
)

knitr::kable(
  riverton_parenthetical_score,
  col.names = c("Source", "Definitions found"),
  caption = "Parenthetical definition rule on the Riverton text",
  row.names = FALSE
)

knitr::kable(
  definitions,
  col.names = c("Sentence ID", "Long form", "Short form"),
  caption = "Constructed parenthetical abbreviation definitions",
  row.names = FALSE
)

knitr::kable(
  parenthetical_score,
  col.names = c("True definitions in constructed passage", "Definitions found"),
  caption = "Parenthetical rule on the constructed positive cases",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(riverton_definitions), 0L),
  identical(riverton_parenthetical_score$definitions_found, 0L),
  identical(nrow(definitions), 2L),
  identical(definitions$long_form, c("Natural Language Processing", "Data Support Certificate")),
  identical(definitions$short_form, c("NLP", "DSC")),
  identical(parenthetical_score$definitions_found, 2L),
  identical(shape_only_example$long_form, "Riverton Skills Centre"),
  identical(shape_only_example$short_form, "NLP")
)
```

The parenthetical pattern finds zero definitions in the Riverton text. It finds
`Natural Language Processing (NLP)` and `Data Support Certificate (DSC)` in the
constructed passage, which contains only positive cases. That run cannot show
how often the pattern fires where it should not, or what it misses. The pattern
only checks shapes: capitalised words followed by a bracketed run of capitals.
It does not check that the short-form letters come from the long form, so
`Riverton Skills Centre (NLP)` also matches. The Schwartz and Hearst algorithm
listed under Sources checks those letter links and allows lowercase long forms;
this rule is much smaller.

## What to remember

- All-capital text is only a rough short-form signal.
- The 7 capital-letter hits in the Riverton headings contain 0 true abbreviations.
- A guard that keeps 0 of 7 hits has not proved that it protects true abbreviations.
- The parenthetical rule finds 2 constructed positive cases and 0 Riverton definitions.
- Shape-only parenthetical rules need a letter-link check before they can be trusted.

Reject the all-capital shortcut for these flyer headings. For future text, keep
candidate short forms visible until a rule has been tested on both true examples
and counter-examples.

## Sources

- [stringr regular expression documentation](https://stringr.tidyverse.org/articles/regular-expressions.html)
- [hunspell R package documentation](https://docs.ropensci.org/hunspell/)
- [Schwartz and Hearst abbreviation algorithm paper](https://psb.stanford.edu/psb-online/proceedings/psb03/schwartz.pdf)
