---
title: "Chopping word endings"
subtitle: "Use Snowball stems without mistaking them for words"
description: "Learn how stemming groups some workforce word forms and creates non-word stems."
categories: [word processing, stemming, workforce research]
---

A short vocabulary list from the job-board and flyer sentences is the team's
next goal. The team sees `shifts`, `skills`, and `schedules` and asks
whether plural endings should count as separate words.

That choice affects a simple summary. If every surface spelling stays separate,
the list grows. If endings are chopped too aggressively, different meanings can
collapse into one code.

**Stemming** removes or changes word endings using fixed rules. In linguistics,
a stem is the part of a word that an ending attaches to. In this lesson, a
**stem** is whatever string these rules return. It is a
comparison key. It may or may not be a word.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The stems come from SnowballC running locally.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- describe what a stemmer does;
- find stems that are not dictionary words;
- identify collisions, where several words share one stem; and
- name related words that a suffix rule leaves apart.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build the Riverton vocabulary

The vocabulary build uses readr, dplyr, tibble, tidytext, SnowballC, hunspell,
purrr, and stringr. A word type is one distinct lowercased word form in these 28
sentences using the same `tokenizers::tokenize_words()` rule as Lesson 15, so
the count would change under a different tokenizer.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tidytext)
library(SnowballC)
library(hunspell)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

vocabulary <- sentences |>
  select(sentence_id, text) |>
  unnest_tokens(word, text) |>
  distinct(word) |>
  arrange(word) |>
  pull(word)

known_to_hunspell <- \(words) map_lgl(hunspell(words), \(result) length(result) == 0L)

stemmed_vocabulary <- tibble(
  word = vocabulary,
  stem = wordStem(vocabulary, language = "en")
) |>
  mutate(
    changed = word != stem,
    stem_in_dictionary = known_to_hunspell(stem)
  )

knitr::kable(
  stemmed_vocabulary |>
    filter(word %in% c("shifts", "skills", "required", "provided", "training")) |>
    arrange(word),
  col.names = c("Word type", "Stem", "Changed", "Stem in dictionary"),
  caption = "Five Riverton word types after stemming",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(length(vocabulary), 96L),
  identical(stemmed_vocabulary$stem[stemmed_vocabulary$word == "shifts"], "shift"),
  identical(stemmed_vocabulary$stem[stemmed_vocabulary$word == "required"], "requir"),
  identical(sum(stemmed_vocabulary$changed), 43L),
  identical(sum(!stemmed_vocabulary$changed), 53L)
)
```

The stemmer changes 43 of the 96 word types. `shifts` becomes `shift`, a useful
result for a count that should group singular and plural forms. `required`
becomes `requir`, which is not a word a reader should see.

## Watch the rule on a nearby word

SnowballC's English stemmer, often called Porter2, uses a fixed rule system. The
word `requires` is not in the Riverton vocabulary, but it shows the same kind of
ending chop as `required`.

```{r}
example_stems <- tibble(
  word = c("shifts", "requires", "required"),
  stem = wordStem(word, language = "en")
)

knitr::kable(
  example_stems,
  col.names = c("Word", "Snowball stem"),
  caption = "A rule can produce useful stems and non-word stems",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(example_stems$stem, c("shift", "requir", "requir")),
  !"requires" %in% vocabulary,
  identical(
    example_stems$stem[example_stems$word == "requires"],
    example_stems$stem[example_stems$word == "required"]
  )
)
```

Treat the returned stem as a machine comparison string. It should not be shown to
a reader as if it were the dictionary form of the word.

Hunspell checks a spelling dictionary and affix rules. SnowballC does not ask a
dictionary before returning a stem, so it can return `requir`, `provid`, and
`schedul`.

## Count stems that are words and stems that are not

The next table separates changed stems that are dictionary words from changed
stems that hunspell flags as unknown.

```{r}
stem_quality <- stemmed_vocabulary |>
  summarise(
    word_types = n(),
    changed_by_stemmer = sum(changed),
    unchanged_by_stemmer = sum(!changed),
    changed_to_dictionary_word = sum(changed & stem_in_dictionary),
    changed_to_non_word = sum(changed & !stem_in_dictionary),
    .groups = "drop"
  )

non_word_examples <- stemmed_vocabulary |>
  filter(changed, !stem_in_dictionary) |>
  arrange(word) |>
  slice_head(n = 8)

knitr::kable(
  stem_quality,
  col.names = c(
    "Word types",
    "Changed",
    "Unchanged",
    "Changed to dictionary word",
    "Changed to non-word"
  ),
  caption = "Stemming outcomes for the Riverton vocabulary",
  row.names = FALSE
)

knitr::kable(
  non_word_examples,
  col.names = c("Word type", "Stem", "Changed", "Stem in dictionary"),
  caption = "Examples of non-word stems in the Riverton vocabulary",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(stem_quality$word_types, 96L),
  identical(stem_quality$changed_by_stemmer, 43L),
  identical(stem_quality$unchanged_by_stemmer, 53L),
  identical(stem_quality$changed_to_dictionary_word, 21L),
  identical(stem_quality$changed_to_non_word, 22L),
  identical(non_word_examples$word[1], "able"),
  identical(non_word_examples$stem[8], "essenti")
)
```

Among these 96 word types, 21 changed forms lead to dictionary words and 22 lead
to strings hunspell does not know. A dictionary-word stem can still be the wrong
grouping key for a particular workforce question.

## Find collisions

A **collision** happens when different word types reduce to the same stem. A
collision can help when the words are close enough for the question. It can also
hide a useful distinction.

```{r}
collision_groups <- stemmed_vocabulary |>
  group_by(stem) |>
  filter(n() > 1L) |>
  summarise(
    word_types = paste(word, collapse = ", "),
    n_words = n(),
    .groups = "drop"
  ) |>
  arrange(stem)

knitr::kable(
  collision_groups,
  col.names = c("Stem", "Word types that share it", "Word types"),
  caption = "Snowball collisions in the Riverton vocabulary",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(collision_groups), 4L),
  identical(sum(collision_groups$n_words), 8L),
  identical(
    collision_groups$stem,
    c("certif", "includ", "schedul", "skill")
  ),
  str_detect(collision_groups$word_types[1], fixed("certificate"))
)
```

The four collision groups contain eight word types. `schedule` and `schedules`
are a helpful group. `certificate` and `certification` may or may not be close
enough, depending on whether the Lab is counting credentials or training topics.

## Find misses

A **miss** is a related pair that the stemmer leaves apart. The team picked three
pairs it already suspected the stemmer would keep apart, so the table is not a
survey. It is a set of examples chosen to show what a suffix rule cannot see.

```{r}
miss_pairs <- tribble(
  ~word_a, ~word_b, ~reason,
  "applicants", "apply", "Both point to applying, but their stems differ.",
  "paid", "pays", "Snowball strips suffixes. 'Paid' has no suffix to strip, so no rule can reach 'pay'.",
  "work", "workers", "'-er' names the person doing the activity. Whether to group them depends on the question."
) |>
  mutate(
    stem_a = wordStem(word_a, language = "en"),
    stem_b = wordStem(word_b, language = "en"),
    same_stem = stem_a == stem_b
  )

knitr::kable(
  miss_pairs,
  col.names = c("Word A", "Word B", "Why the pair was checked", "Stem A", "Stem B", "Same stem"),
  caption = "Related Riverton word pairs that Snowball keeps apart",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(miss_pairs$stem_a, c("applic", "paid", "work")),
  identical(miss_pairs$stem_b, c("appli", "pay", "worker")),
  identical(sum(!miss_pairs$same_stem), 3L),
  all(c(miss_pairs$word_a, miss_pairs$word_b) %in% vocabulary)
)
```

All three selected pairs stay apart. The result says what these examples were
chosen to show: suffix rules can miss irregular forms and relations whose meaning
comes from more than an ending.

## What to remember

- Stemming chops word endings with fixed rules.
- The Snowball English stemmer is a Porter2 rule system.
- SnowballC changed 43 of the 96 Riverton word types.
- In this vocabulary, 22 changed stems are non-words.
- Collisions and selected misses both need human judgment.

Stems are useful backstage keys for rough grouping. The public-facing list should
show words a reader recognizes, with the disputed groups reviewed by a person.

## Sources

- [SnowballC package reference](https://cran.r-project.org/package=SnowballC)
- [Snowball stemming algorithms](https://snowballstem.org/)
- [hunspell package reference](https://docs.ropensci.org/hunspell/reference/hunspell.html)
