---
title: "Choosing dictionary forms"
subtitle: "Compare teaching lemmas with a released pipeline"
author: "Alex Farach"
description: "Learn how lemmas differ from stems and why lemmatizers built in different ways disagree on workforce text."
categories: [word processing, lemmatization, workforce research]
---

Stemmed forms shorten the vocabulary, but strings such as `requir` and `provid`
give the team pause. Those strings help a computer compare
forms, yet they look broken on a public report.

The team wants a cleaner label for words such as `is`, `provided`, and
`classes`. It needs to know when a model's dictionary form is better than a
rule-chopped stem, and when a model can still guess badly.

A **lemma** is the dictionary form of a word, the form a reader might look up.
**Lemmatization** chooses that form in context. This lesson compares a tiny
UDPipe teaching model with a released spaCy English pipeline.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. Both models run locally; the lesson does not
download anything while it runs.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define a lemma in plain language;
- compare a lemma with a Snowball stem;
- see how tokenization changes the comparison set;
- name what a dictionary proxy can and cannot measure; and
- separate raw dictionary flags from broken-string counts.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load both lemmatizers

Two local lemmatizers need the same sentence file and a shared set of table
tools. The code loads the CSV reader, tokenizers, the UDPipe teaching model, the
spaCy bridge, SnowballC stems, and hunspell dictionary checks. UDPipe receives
one token per line and returns lemmas from its trained tagger, which combines a
model with dictionary and guesser components. spaCy receives the raw sentences
through the project helper `use_project_spacy()`.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tidytext)
library(tokenizers)
library(udpipe)
library(spacyr)
library(SnowballC)
library(hunspell)
source("R/use-spacy.R")

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

pipeline <- use_project_spacy()
spacy_info <- spacy_pipeline_version()
spacy_info_table <- tibble(
  item = names(spacy_info),
  value = unname(unlist(spacy_info))
)

udpipe_parser <- udpipe_load_model(
  "data/treebank/en_ewt-500-parser.udpipe"
)

vertical <- sentences$text |>
  tokenize_words(lowercase = FALSE, strip_punct = FALSE) |>
  map_chr(\(tokens) paste(tokens, collapse = "\n"))

udpipe_parsed <- udpipe_annotate(
  udpipe_parser,
  x = vertical,
  doc_id = sentences$sentence_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "default"
) |>
  as.data.frame() |>
  as_tibble()

spacy_text <- sentences$text
names(spacy_text) <- sentences$sentence_id

spacy_parsed <- spacy_parse(
  spacy_text,
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  nounphrase = TRUE
) |>
  as_tibble()

stopifnot(
  identical(nrow(sentences), 28L),
  identical(pipeline$model, "en_core_web_sm"),
  identical(spacy_info$version, "3.8.0"),
  identical(spacy_info$license, "MIT"),
  identical(spacy_info$spacy, "3.8.7"),
  identical(nrow(udpipe_parsed), 178L),
  identical(nrow(spacy_parsed), 179L),
  identical(udpipe_parsed$lemma[udpipe_parsed$doc_id == "s001" & udpipe_parsed$token == "provided"], "provid"),
  identical(spacy_parsed$lemma[spacy_parsed$doc_id == "s001" & spacy_parsed$token == "provided"], "provide")
)

knitr::kable(
  spacy_info_table,
  col.names = c("Pipeline field", "Value"),
  caption = "Local spaCy pipeline used by this lesson",
  row.names = FALSE
)
```

The spaCy table records the exact local pipeline: `core_web_sm` version 3.8.0
under the MIT license, running with spaCy 3.8.7. The two tools return slightly
different token counts for the same 28 sentences: 178 from UDPipe and 179 from
spaCy. The extra comparison problem is `driver's`, which spaCy splits into
`driver` and `'s` while the vertical tokenizer keeps it whole. That word type has
no spaCy lemma to compare, and the tables below drop it.

## Compare lemmas with stems

The next table uses word types that appear in the Riverton sentences and that
both parsers emit as tokens. A stem and a lemma can agree, but they answer
different questions.

```{r}
known_to_hunspell <- \(words) map_lgl(hunspell(words), \(result) length(result) == 0L)

vocabulary_words <- sentences |>
  select(sentence_id, text) |>
  unnest_tokens(word, text) |>
  distinct(word) |>
  arrange(word)

udpipe_lookup <- udpipe_parsed |>
  filter(upos != "PUNCT") |>
  transmute(
    word = str_to_lower(token),
    udpipe_lemma = str_to_lower(lemma)
  ) |>
  group_by(word) |>
  summarise(udpipe_lemma = first(udpipe_lemma), .groups = "drop")

spacy_lookup <- spacy_parsed |>
  filter(pos != "PUNCT") |>
  transmute(
    word = str_to_lower(token),
    spacy_lemma = str_to_lower(lemma)
  ) |>
  group_by(word) |>
  summarise(spacy_lemma = first(spacy_lemma), .groups = "drop")

same_word_lemmas <- vocabulary_words |>
  left_join(udpipe_lookup, by = "word") |>
  left_join(spacy_lookup, by = "word")

comparison_words <- c(
  "is",
  "provided",
  "required",
  "classes",
  "shifts",
  "applicants",
  "training",
  "schedules",
  "paid",
  "pays"
)

lemma_stem_comparison <- same_word_lemmas |>
  filter(word %in% comparison_words) |>
  mutate(
    stem = wordStem(word, language = "en"),
    udpipe_known = known_to_hunspell(udpipe_lemma),
    spacy_known = known_to_hunspell(spacy_lemma)
  ) |>
  arrange(match(word, comparison_words))

stopifnot(
  identical(nrow(vocabulary_words), 96L),
  identical(sum(is.na(same_word_lemmas$spacy_lemma)), 1L),
  identical(same_word_lemmas$word[is.na(same_word_lemmas$spacy_lemma)], "driver's"),
  identical(lemma_stem_comparison$udpipe_lemma[1:4], c("be", "provid", "requi", "classe")),
  identical(lemma_stem_comparison$spacy_lemma[1:4], c("be", "provide", "require", "class")),
  identical(lemma_stem_comparison$stem[1:4], c("is", "provid", "requir", "class")),
  identical(nrow(lemma_stem_comparison), 10L)
)

knitr::kable(
  lemma_stem_comparison,
  col.names = c(
    "Word",
    "UDPipe lemma",
    "spaCy lemma",
    "Snowball stem",
    "UDPipe lemma known",
    "spaCy lemma known"
  ),
  caption = "Lemmas and stems for selected Riverton word types",
  row.names = FALSE
)
```

`is` shows why a lemma can be better than a stem. A stemmer leaves `is` alone,
while both parsers choose `be`. The sharper contrast is `provided`: the UDPipe
teaching model returns `provid`, while spaCy returns `provide`. For `classes`,
UDPipe returns `classe`, while spaCy returns `class`.

The table is deliberately simple, but it hides context. A word type can receive
different lemmas in different sentences, which is why lemmatization is contextual
in the first place. The next check shows the type rows where at least one tool
gave more than one lemma.

```{r}
lemma_variation <- bind_rows(
  udpipe_parsed |>
    filter(upos != "PUNCT") |>
    transmute(system = "UDPipe", word = str_to_lower(token), lemma = str_to_lower(lemma)),
  spacy_parsed |>
    filter(pos != "PUNCT") |>
    transmute(system = "spaCy", word = str_to_lower(token), lemma = str_to_lower(lemma))
) |>
  group_by(system, word) |>
  summarise(
    distinct_lemmas = n_distinct(lemma),
    lemmas = paste(sort(unique(lemma)), collapse = ", "),
    occurrences = n(),
    .groups = "drop"
  ) |>
  filter(distinct_lemmas > 1L) |>
  arrange(system, word)

stopifnot(
  identical(lemma_variation$system, c("UDPipe", "UDPipe", "UDPipe", "spaCy")),
  identical(lemma_variation$word, c("evening", "paid", "skills", "skills")),
  identical(lemma_variation$distinct_lemmas, c(2L, 2L, 2L, 2L))
)

knitr::kable(
  lemma_variation,
  col.names = c("System", "Word type", "Distinct lemmas", "Lemmas seen", "Occurrences"),
  caption = "Word types whose lemmas vary across Riverton occurrences",
  row.names = FALSE
)
```

This is not an equal contest. The UDPipe model was trained in this project on a
500-sentence excerpt for teaching. The spaCy pipeline is a released English
pipeline packaged for general use. The methods differ too: UDPipe lemmatizes
through its trained tagger together with dictionary and guesser machinery, while
spaCy's English lemmatizer uses lookup tables and part-of-speech rules. The
tools, tokenization, and training setup differ, so this lesson cannot turn this
contrast into a causal story about training size alone.

## Count dictionary failures

A cheap screen asks whether each lemma appears in hunspell's dictionary. It
answers one narrow question: did the model return a string that is not an English
word? It cannot tell a correct lemma that the dictionary happens not to hold,
such as `riverton`, from a genuine error. It is also blind to a wrong lemma that
is a real word. If a model returned `leave` as the lemma of `left` in the phrase
`the left column`, this check would pass it. The counts below measure broken
strings, not correctness.

```{r}
compared_word_lemmas <- same_word_lemmas |>
  filter(!is.na(udpipe_lemma), !is.na(spacy_lemma)) |>
  mutate(
    udpipe_known = known_to_hunspell(udpipe_lemma),
    spacy_known = known_to_hunspell(spacy_lemma),
    udpipe_flagged = !udpipe_known,
    spacy_flagged = !spacy_known,
    coverage_or_case_artifact = word %in% c("october", "riverton"),
    udpipe_broken_string = udpipe_flagged & !coverage_or_case_artifact,
    spacy_broken_string = spacy_flagged & !coverage_or_case_artifact
  )

lemma_quality <- tibble(
  system = c("UDPipe teaching model", "spaCy released pipeline"),
  word_types_compared = c(nrow(compared_word_lemmas), nrow(compared_word_lemmas)),
  raw_dictionary_flags = c(
    sum(compared_word_lemmas$udpipe_flagged),
    sum(compared_word_lemmas$spacy_flagged)
  ),
  coverage_or_case_artifacts = c(
    sum(compared_word_lemmas$udpipe_flagged & compared_word_lemmas$coverage_or_case_artifact),
    sum(compared_word_lemmas$spacy_flagged & compared_word_lemmas$coverage_or_case_artifact)
  ),
  broken_string_flags = c(
    sum(compared_word_lemmas$udpipe_broken_string),
    sum(compared_word_lemmas$spacy_broken_string)
  )
)

flagged_comparison <- compared_word_lemmas |>
  filter(udpipe_flagged | spacy_flagged) |>
  select(
    word,
    udpipe_lemma,
    spacy_lemma,
    udpipe_flagged,
    spacy_flagged,
    coverage_or_case_artifact,
    udpipe_broken_string,
    spacy_broken_string
  )

stopifnot(
  identical(nrow(compared_word_lemmas), 95L),
  identical(lemma_quality$raw_dictionary_flags, c(8L, 2L)),
  identical(lemma_quality$coverage_or_case_artifacts, c(2L, 2L)),
  identical(lemma_quality$broken_string_flags, c(6L, 0L)),
  identical(
    flagged_comparison$word,
    c("classes", "diploma", "includes", "need", "october", "provided", "required", "riverton")
  ),
  identical(
    flagged_comparison$udpipe_lemma,
    c("classe", "diploman", "includ", "ne", "october", "provid", "requi", "riverton")
  ),
  identical(flagged_comparison$spacy_flagged, c(FALSE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, TRUE))
)

knitr::kable(
  lemma_quality,
  col.names = c(
    "System",
    "Word types compared",
    "Raw dictionary flags",
    "Coverage or case artifacts",
    "Broken-string flags"
  ),
  caption = "Dictionary proxy counts for lemmas on the same Riverton word types",
  row.names = FALSE
)

knitr::kable(
  flagged_comparison,
  col.names = c(
    "Word",
    "UDPipe lemma",
    "spaCy lemma",
    "UDPipe flagged",
    "spaCy flagged",
    "Coverage or case artifact",
    "UDPipe broken string",
    "spaCy broken string"
  ),
  caption = "Where either model's lemma is flagged by hunspell",
  row.names = FALSE
)
```

On the 95 word types both parsers return a lemma for, hunspell flags 8 UDPipe
lemmas and 2 spaCy lemmas. Two flags are the same on both sides and are not model
errors: `riverton` is a made-up place name, and `october` is flagged because this
lesson lowercases lemmas before checking. spaCy returned `October`, which
hunspell knows. Setting those aside, UDPipe produced six broken strings:
`classe`, `diploman`, `includ`, `ne`, `provid`, and `requi`. spaCy produced none.
Six word types is a small count from 28 sentences of one kind of text.

## Decide when the cost is justified

For a public table, a checked lemma is usually safer than a Snowball stem.
`is` and `are` can group under `be`, and `provided` can become `provide` rather
than `provid`.

Stemming is cheaper and often enough for rough private matching. It is a poor
choice for display because strings such as `requir` and `schedul` are not reader
words. Lemmatization is cleaner when it works, but the UDPipe `provid` example
shows that a lemma column still needs checks.

```{r}
final_check <- tibble(
  model = c("UDPipe", "spaCy"),
  provided_lemma = c(
    lemma_stem_comparison$udpipe_lemma[lemma_stem_comparison$word == "provided"],
    lemma_stem_comparison$spacy_lemma[lemma_stem_comparison$word == "provided"]
  )
)

stopifnot(
  identical(final_check$provided_lemma, c("provid", "provide")),
  identical(spacy_info$version, "3.8.0"),
  identical(spacy_info$spacy, "3.8.7")
)

knitr::kable(
  final_check,
  col.names = c("Model", "Lemma for provided"),
  caption = "A final check on the clearest contrast",
  row.names = FALSE
)

invisible(suppressMessages(spacy_finalize()))
```

## What to remember

- A lemma is a dictionary form chosen for a word in context.
- A stem is a rule-chopped comparison string.
- Tokenizers changed the comparison: `driver's` has no spaCy word-type match.
- The dictionary screen flags unknown strings, not lemma correctness.
- Raw hunspell flags are 8 for UDPipe and 2 for spaCy on 95 word types.
- After coverage and case artifacts are set aside, the broken-string count is 6
  for UDPipe and 0 for spaCy.

Use lemmas for reader-facing grouping only after checking the rows that look
broken or surprising. The important comparison here is six small-model lemma
guesses to inspect, not a causal story about training size.

## Sources

- [spaCy models and languages](https://spacy.io/usage/models)
- [spaCy lemmatizer documentation](https://spacy.io/api/lemmatizer)
- [spacyr package reference](https://spacyr.quanteda.io/)
- [UDPipe R package reference](https://bnosac.github.io/udpipe/docs/doc0.html)
- [UDPipe User's Manual](https://ufal.mff.cuni.cz/udpipe/1/users-manual)
- [hunspell package reference](https://docs.ropensci.org/hunspell/reference/hunspell.html)
