---
title: "Making text comparable"
subtitle: "Normalize carefully and keep the display text intact"
author: "Alex Farach"
description: "Learn how case, spacing, punctuation, and Unicode normalization change workforce text."
categories: [word processing, normalization, workforce research]
---

Repeated phrases across the 28 sentences do not quite line up. One flyer
heading is all capitals, `RIVERTON SKILLS OPEN HOUSE`, while the job
details use ordinary case. The team also expects text pasted from other systems
to arrive with stray spacing and accented characters, so it wants to know what
each cleanup step costs before applying any of them.

Before counting matches, the team has to decide which small differences to
ignore. Each edit can make comparison easier and also erase information.

**Normalization** is the set of small edits made before comparing text. Common
steps include changing case, trimming spaces, removing punctuation, and putting
Unicode characters into a consistent form.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The text files are local and fixed.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain why text is normalized before comparison;
- see how case folding and space repair change examples;
- test punctuation removal with stringr's ICU regex engine; and
- compare two Unicode forms of the same visible word.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load text with an accent check

readr opens the CSV files, dplyr and tibble keep the examples in compact tables,
stringr handles text operations, stringi handles Unicode normalization, and
purrr repeats checks. The inputs are the Riverton sentences and a feedback file
that contains the word `café`.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(stringr)
library(stringi)
library(purrr)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

feedback <- read_csv(
  "data/customer_feedback.csv",
  na = character(),
  col_types = cols(
    feedback_id = col_character(),
    submitted_at = col_character(),
    channel = col_character(),
    rating = col_character(),
    comment = col_character()
  )
)

stopifnot(
  identical(nrow(sentences), 28L),
  identical(sentences$text[23], "RIVERTON SKILLS OPEN HOUSE"),
  identical(nrow(feedback), 5L),
  identical(feedback$comment[2], "The café listing was easy to find.")
)
```

The accent check matters because normalization can change text that looks stable
on screen. The visible word may stay the same while the stored code points
change.

## Fold case and repair spaces

**Case folding** means converting letters to one case before comparison. The
next table uses one real Riverton string, one copy of `s002` with its spaces
tripled to stand in for pasted text, and one short warning pair, `US` and `us`.

```{r}
padded_requirement <- str_replace_all(
  sentences$text[sentences$sentence_id == "s002"],
  " ",
  "   "
)
padded_requirement <- paste0("  ", padded_requirement, "  ")

case_space_examples <- tibble(
  example = c(
    "Flyer heading",
    "Padded requirement sentence",
    "Warning pair, capital",
    "Warning pair, lower"
  ),
  raw_text = c(
    sentences$text[sentences$sentence_id == "s023"],
    padded_requirement,
    "US",
    "us"
  ),
  lowercased = str_to_lower(raw_text),
  squished = str_squish(raw_text)
)

stopifnot(
  identical(case_space_examples$lowercased[1], "riverton skills open house"),
  identical(case_space_examples$squished[2], "No prior data experience is required."),
  identical(case_space_examples$lowercased[3], case_space_examples$lowercased[4]),
  identical(str_length(case_space_examples$raw_text[2]), 51L)
)

knitr::kable(
  case_space_examples,
  col.names = c("Example", "Raw text", "Lowercased", "Squished"),
  caption = "Case folding and spacing changes",
  row.names = FALSE
)
```

Lowercasing makes `RIVERTON SKILLS OPEN HOUSE` easier to compare with ordinary
case. It also makes `US` and `us` the same string. `str_squish()` removes leading
and trailing spaces and turns repeated spaces into one space.

## Remove punctuation only when it is safe

Punctuation removal can make matching easier, but it can damage units such as
`12-week`. The next table uses Riverton strings that contain hyphens,
apostrophes, and periods.

```{r}
punctuation_examples <- tibble(
  sentence_id = c("s001", "s015", "s017"),
  raw_text = sentences$text[match(sentence_id, sentences$sentence_id)],
  without_punctuation = str_remove_all(raw_text, "[[:punct:]]")
)

unicode_punctuation_check <- str_remove_all("“paid—training”", "[[:punct:]]")

stopifnot(
  identical(
    punctuation_examples$without_punctuation[1],
    "Paid 12week training is provided"
  ),
  identical(
    punctuation_examples$without_punctuation[2],
    "Onthejob training is provided"
  ),
  identical(
    punctuation_examples$without_punctuation[3],
    "A valid drivers license is required"
  ),
  identical(unicode_punctuation_check, "paidtraining")
)

knitr::kable(
  punctuation_examples,
  col.names = c("Sentence ID", "Raw text", "After punctuation removal"),
  caption = "Punctuation removal can erase useful distinctions",
  row.names = FALSE
)
```

Removing punctuation destroys the hyphen in `12-week` and the apostrophe in
`driver's`. `str_remove_all()` uses stringr's ICU engine, where `[[:punct:]]`
means punctuation across Unicode, including curly quotes and em dashes pasted
from word processors.

## Normalize Unicode forms

Unicode stores text using code points. The word `café` can be stored as one code
point for `é`, or as `e` plus a combining accent. NFC keeps the composed form;
NFD splits the accent into its own code point.

```{r}
cafe_word <- str_extract(feedback$comment[2], "café")
cafe_nfc <- stri_trans_nfc(cafe_word)
cafe_nfd <- stri_trans_nfd(cafe_word)

unicode_forms <- tibble(
  form = c("NFC", "NFD"),
  text = c(cafe_nfc, cafe_nfd),
  code_point_count = str_length(text),
  bytes_utf8 = nchar(text, type = "bytes"),
  code_points = map_chr(text, \(value) paste(utf8ToInt(value), collapse = ", "))
)

stopifnot(
  identical(cafe_word, "café"),
  identical(cafe_nfc, "café"),
  !identical(cafe_nfc, cafe_nfd),
  identical(unicode_forms$code_point_count, c(4L, 5L)),
  identical(unicode_forms$bytes_utf8, c(5L, 6L)),
  identical(unicode_forms$code_points[2], "99, 97, 102, 101, 769")
)

knitr::kable(
  unicode_forms,
  col.names = c("Form", "Text", "Code points", "Bytes as UTF-8", "Code point values"),
  caption = "Two Unicode forms of the visible word 'café'",
  row.names = FALSE
)
```

The two forms look the same in the table, but `identical()` says they are not the
same stored string. Both forms show four letters on screen. The NFC form stores
them as 4 code points and, as UTF-8, 5 bytes. The NFD form stores them as 5 code
points and 6 bytes, because the accent is a separate code point after the `e`.

## Keep two columns

The safe pattern is to store at least two versions: the raw text for display and
a normalized text for comparison. The normalized column can be lowercased,
squished, or changed to NFC as the task requires.

```{r}
display_and_match <- sentences |>
  filter(sentence_id %in% c("s001", "s023")) |>
  transmute(
    sentence_id,
    display_text = text,
    comparison_text = text |>
      stri_trans_nfc() |>
      str_squish() |>
      str_to_lower()
  )

stopifnot(
  identical(display_and_match$display_text[1], "Paid 12-week training is provided."),
  identical(display_and_match$comparison_text[1], "paid 12-week training is provided."),
  identical(display_and_match$comparison_text[2], "riverton skills open house"),
  !identical(display_and_match$display_text[2], display_and_match$comparison_text[2])
)

knitr::kable(
  display_and_match,
  col.names = c("Sentence ID", "Display text", "Comparison text"),
  caption = "Keeping display text separate from comparison text",
  row.names = FALSE
)
```

Normalize for comparison, keep the raw text for display. These edits run one
way. Once `US` has been lowercased to `us`, nothing in the normalized column can
tell you which one it was. Pick NFC for the comparison column because it is the
form recommended for web text, so strings arriving from different systems settle
on the same spelling.

## What to remember

- Normalization makes text easier to compare.
- Lowercasing can merge strings that mean different things.
- ICU punctuation classes reach beyond keyboard punctuation.
- Unicode forms can look the same while storing different code points.
- Keep the display text and comparison text separate.
- Raw text is the fallback when a cleanup step erases information.

A normalized column is a matching aid, not display copy. The Riverton sentences
that readers see should remain the raw sentences the Lab collected.

## Sources

- [stringr reference for case and whitespace helpers](https://stringr.tidyverse.org/reference/)
- [stringi Unicode normalization reference](https://stringi.gagolewski.com/rapi/stri_trans_nf.html)
- [Unicode normalization forms](https://unicode.org/reports/tr15/)
