---
title: "Checking dictionary unknowns"
subtitle: "Measure what a spell checker catches and what it cannot see"
description: "Learn how hunspell flags unknown workforce words and how to test constructed spelling errors."
categories: [word processing, spell checking, workforce research]
---

A vocabulary list is almost ready for review. The team wants to know whether a
spell checker can clean obvious typing errors before a
person reads the list.

That sounds simple until the team runs into proper names and real-word typos. A
dictionary can say which strings it does not know. It cannot know what the writer
meant.

**Spell checking** here means asking hunspell for words that are absent from its
dictionary. The result is a list of unknown words, not a list of all errors.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The deliberately misspelled words below are
constructed examples made from Riverton words.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- run `hunspell()` on the Riverton vocabulary;
- count word types the dictionary flags;
- build a small constructed spelling test;
- see why a test built from known non-words cannot fail; and
- explain why real-word errors escape a dictionary.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Check the real vocabulary

For the vocabulary check, readr brings in the sentences, dplyr and tibble shape
the word list, purrr repeats the dictionary call, tidytext creates tokens, and
hunspell supplies dictionary flags. It builds the same lowercased 96-word
Riverton vocabulary as the stemming lesson, using the same tokenization rule.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(tidytext)
library(hunspell)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

vocabulary <- sentences |>
  select(sentence_id, text) |>
  unnest_tokens(word, text) |>
  distinct(word) |>
  arrange(word) |>
  pull(word)

known_to_hunspell <- \(words) map_lgl(hunspell(words), \(result) length(result) == 0L)

flagged_types <- vocabulary[!known_to_hunspell(vocabulary)] |>
  sort()

flagged_table <- tibble(flagged_word_type = flagged_types)

knitr::kable(
  flagged_table,
  col.names = "Flagged word type",
  caption = "Riverton word types flagged by hunspell",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(length(vocabulary), 96L),
  identical(length(flagged_types), 2L),
  identical(flagged_types, c("october", "riverton"))
)
```

Hunspell flags 2 of the 96 real word types: `october` and `riverton`. In this
small vocabulary, the dictionary flags a month word under the lowercased form
and a fictional place name. Those are unknown to this dictionary setting, not
proved spelling mistakes.

## Build a constructed error test

The next table uses eight deliberate misspellings of real Riverton word types.
This is a constructed test, not a claim about how often people make these
errors.

```{r}
constructed_errors <- tribble(
  ~correct_word, ~misspelled_word,
  "training", "traning",
  "provided", "provied",
  "experience", "experiance",
  "required", "requiered",
  "workers", "wrokers",
  "schedules", "scheduels",
  "communication", "communciation",
  "certificate", "certficate"
)

suggestions <- hunspell_suggest(constructed_errors$misspelled_word)

constructed_results <- constructed_errors |>
  mutate(
    flagged = !known_to_hunspell(misspelled_word),
    correct_word_wrongly_flagged = !known_to_hunspell(correct_word),
    suggestion_count = map_int(suggestions, length),
    suggestions = map_chr(suggestions, \(result) paste(result, collapse = ", "))
  )

spell_summary <- tibble(
  real_errors_caught = sum(constructed_results$flagged),
  constructed_errors = nrow(constructed_results),
  correct_words_wrongly_flagged = sum(constructed_results$correct_word_wrongly_flagged),
  correct_words_checked = nrow(constructed_results),
  caught_fraction = paste0(real_errors_caught, "/", constructed_errors),
  false_flag_fraction = paste0(correct_words_wrongly_flagged, "/", correct_words_checked)
)

knitr::kable(
  constructed_results,
  col.names = c("Correct Riverton word", "Constructed misspelling", "Flagged", "Correct word flagged", "Suggestions returned", "Suggestions"),
  caption = "Constructed spelling errors and hunspell suggestions",
  row.names = FALSE
)

knitr::kable(
  spell_summary,
  col.names = c(
    "Errors caught",
    "Constructed errors",
    "Correct words wrongly flagged",
    "Correct words checked",
    "Caught fraction",
    "False flag fraction"
  ),
  caption = "Counts from the constructed spelling test",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(constructed_results$flagged, rep(TRUE, 8)),
  identical(constructed_results$correct_word_wrongly_flagged, rep(FALSE, 8)),
  all(constructed_results$suggestion_count > 0L),
  identical(spell_summary$real_errors_caught, 8L),
  identical(spell_summary$correct_words_wrongly_flagged, 0L),
  identical(spell_summary$caught_fraction, "8/8"),
  identical(spell_summary$false_flag_fraction, "0/8")
)
```

Hunspell flags 8 of 8. That number was settled before the test ran: all eight
constructed misspellings are strings that are not English words, and flagging
non-words is the dictionary's job. It falsely flags 0 of the 8 source words,
which was also settled in advance because those words came from a vocabulary
already shown to hold only two dictionary unknowns. This is a wiring check, not a
measurement of spelling quality. The next section shows the case the eight left
out.

## Show the real-word limit

A dictionary will not flag a typo if the typo is also a real word. `form` and
`from` are both dictionary words.

```{r}
real_word_error <- tibble(
  intended_word = "from",
  typed_word = "form",
  intended_flagged = !known_to_hunspell(intended_word),
  typed_flagged = !known_to_hunspell(typed_word)
)

knitr::kable(
  real_word_error,
  col.names = c("Intended word", "Typed word", "Intended word flagged", "Typed word flagged"),
  caption = "A real-word typo that dictionary lookup cannot catch",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(real_word_error$intended_flagged, FALSE),
  identical(real_word_error$typed_flagged, FALSE),
  known_to_hunspell("form"),
  known_to_hunspell("from")
)
```

If a writer meant `from` and typed `form`, this dictionary stays quiet. Both
strings are familiar to it, so context has to do the work.

## What to remember

- `hunspell()` returns dictionary unknowns.
- Unknown does not always mean misspelled.
- Checking each of the 96 word types on its own flags `october` and `riverton`.
- The constructed 8/8 check is a demonstration because it contains only non-word
  errors.
- Real-word errors need context as well as a dictionary.

Use hunspell to make a review queue, then read the queue. A dictionary can sort
strings into familiar and unfamiliar; people still decide whether the sentence is
right.

## Sources

- [hunspell package reference](https://docs.ropensci.org/hunspell/reference/hunspell.html)
- [Hunspell project documentation](https://hunspell.github.io/)
