---
title: "Finding what a sentence denies"
subtitle: "Use parser links to test a simple negation rule"
description: "Learn why negation breaks keyword matching and how a small parser-based rule performs on workforce text."
categories: [word processing, negation recognition, workforce research]
---

A search for experience requirements begins with an uneasy hit. The team tries a
quick match for `experience` followed by `required` and finds a
sentence that says, `No prior data experience is required.`

That match is dangerous. The words are present, but the sentence says experience
is not required. A keyword search has found the opposite of the claim the team
cares about.

**Negation recognition** looks for language that reverses or blocks an ordinary
positive reading. A real system needs a **scope**, the part of the sentence that
the negating word controls.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The hand markings below are teaching judgments
for these 28 sentences.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- show why a keyword search can misread negation;
- find `no` in this corpus and name trigger words it does not test;
- define a small scope rule using dependency links;
- read a development-set score without treating it as performance; and
- explain why trigger words alone are not enough.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Parse the sentences

The parser starts with the CSV, then readr reads it, dplyr and tibble shape the
tables, purrr repeats scope work, stringr matches text, tokenizers fixes the
word split, and udpipe applies the local model. Each sentence is passed to
udpipe as one vertical block of tokens.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(udpipe)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

parser <- udpipe_load_model(
  "data/treebank/en_ewt-500-parser.udpipe"
)

vertical <- sentences$text |>
  tokenize_words(lowercase = FALSE, strip_punct = FALSE) |>
  map_chr(\(tokens) paste(tokens, collapse = "\n"))

parsed <- udpipe_annotate(
  parser,
  x = vertical,
  doc_id = sentences$sentence_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "default"
) |>
  as.data.frame() |>
  as_tibble()
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(parsed), 178L),
  identical(sentences$text[sentences$sentence_id == "s002"], "No prior data experience is required."),
  identical(parsed$lemma[parsed$doc_id == "s002" & parsed$token == "No"], "no")
)
```

The parser supplies lemmas, relation labels, and head word IDs. Those columns
let the team move beyond a plain string match.

## Start with the tempting keyword search

The naive search below lowercases each sentence and looks for `experience`
before `required`.

```{r}
keyword_hits <- sentences |>
  mutate(
    lowered = str_to_lower(text),
    keyword_match = str_detect(lowered, "experience.*required")
  ) |>
  filter(keyword_match) |>
  select(sentence_id, text)

knitr::kable(
  keyword_hits,
  col.names = c("Sentence ID", "Text"),
  caption = "Keyword matches for experience before required",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(keyword_hits$sentence_id, c("s002", "s019", "s027")),
  identical(nrow(keyword_hits), 3L),
  str_detect(keyword_hits$text[1], fixed("No prior data experience"))
)
```

The keyword search returns three sentences. Two of them start with `No`, so the
plain match is not enough for the requirement question.

## Mark negation by hand

Before showing the rule, the team records a judgment for every sentence. For
this teaching example, a sentence counts as negated when a word in it reverses
the truth of the sentence's main claim. These are single-author teaching labels,
so they are open to dispute.

```{r}
no_sentence_count <- sum(str_detect(str_to_lower(sentences$text), "\\bno\\b"))

hand_judgments <- sentences |>
  transmute(
    sentence_id,
    text,
    hand_negated = sentence_id %in% c("s002", "s027"),
    reason = case_when(
      sentence_id == "s002" ~ "No negates prior data experience; experience is not required.",
      sentence_id == "s027" ~ "No negates prior experience in this flyer headline.",
      TRUE ~ "No negating word changes the main claim."
    )
  )

knitr::kable(
  hand_judgments,
  col.names = c("Sentence ID", "Text", "Hand-marked negated", "Reason"),
  caption = "Hand judgments for negation in the 28 Riverton sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(hand_judgments), 28L),
  all(hand_judgments$sentence_id %in% sentences$sentence_id),
  identical(no_sentence_count, 2L),
  identical(
    hand_judgments$sentence_id[hand_judgments$hand_negated],
    sentences$sentence_id[str_detect(str_to_lower(sentences$text), "\\bno\\b")]
  )
)
```

Only 2 of the 28 sentences are hand-marked as negated. Both use `No`, but they
have different shapes: one is a full sentence and one is a flyer headline.

## Define a parser-based scope rule

The rule has three parts. First, find a trigger whose lemma is `no`, `not`, or
`never`. A surface-form backup also checks `advmod` and `det` tokens; on these 28
sentences it never has to add a trigger. Second, build a scope from the trigger's
head, that head's clause head, and their direct dependents. Third, mark the
sentence negated when that scope contains a form of `be`. The rule was shaped
after reading `s002`, so this page treats the score as a demonstration.
This trigger-and-scope shape echoes NegEx, a rule system built for clinical notes
in 2001; the idea is broader than that setting.

```{r}
negating_words <- c("no", "not", "never")

parsed_with_triggers <- parsed |>
  mutate(
    token_lower = str_to_lower(token),
    lemma_lower = str_to_lower(lemma),
    negation_trigger = lemma_lower %in% negating_words |
      (dep_rel %in% c("advmod", "det") & token_lower %in% negating_words)
  )

surface_only_triggers <- parsed_with_triggers |>
  filter(
    !(lemma_lower %in% negating_words),
    dep_rel %in% c("advmod", "det"),
    token_lower %in% negating_words
  )

trigger_inventory <- parsed_with_triggers |>
  filter(token_lower %in% negating_words) |>
  distinct(token_lower) |>
  arrange(token_lower)

triggers <- parsed_with_triggers |>
  filter(negation_trigger) |>
  transmute(
    sentence_id = doc_id,
    trigger_id = token_id,
    trigger = token,
    trigger_head_id = head_token_id
  )

scope_for_trigger <- function(sentence_id, trigger_id, trigger, trigger_head_id) {
  current_sentence_id <- sentence_id
  current_trigger_id <- trigger_id
  current_trigger <- trigger
  current_trigger_head_id <- trigger_head_id

  sentence_tokens <- parsed_with_triggers |>
    filter(doc_id == current_sentence_id)

  clause_head_id <- sentence_tokens$head_token_id[
    match(current_trigger_head_id, sentence_tokens$token_id)
  ]

  if (is.na(clause_head_id) || identical(clause_head_id, "0")) {
    clause_head_id <- current_trigger_head_id
  }

  sentence_tokens |>
    filter(
      token_id %in% c(current_trigger_id, current_trigger_head_id, clause_head_id) |
        head_token_id %in% c(current_trigger_head_id, clause_head_id)
    ) |>
    transmute(sentence_id = doc_id, trigger = current_trigger, token, lemma, upos, dep_rel)
}

trigger_scopes <- pmap(triggers, scope_for_trigger) |>
  list_rbind()

rule_predictions <- trigger_scopes |>
  group_by(sentence_id) |>
  summarise(
    rule_negated = any(lemma == "be" & upos == "AUX"),
    scope_tokens = paste(token, collapse = " "),
    .groups = "drop"
  ) |>
  right_join(sentences |> select(sentence_id), by = "sentence_id") |>
  mutate(
    rule_negated = if_else(is.na(rule_negated), FALSE, rule_negated),
    scope_tokens = if_else(is.na(scope_tokens), "", scope_tokens)
  )

knitr::kable(
  trigger_scopes,
  col.names = c("Sentence ID", "Trigger", "Scoped token", "Lemma", "POS tag", "Relation"),
  caption = "Tokens inside the simple negation scope",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(surface_only_triggers), 0L),
  identical(trigger_inventory$token_lower, "no"),
  identical(triggers$sentence_id, c("s002", "s027")),
  identical(triggers$trigger, c("No", "No")),
  identical(
    trigger_scopes$token[trigger_scopes$sentence_id == "s002"],
    c("No", "prior", "data", "experience", "is", "required", ".")
  ),
  identical(
    trigger_scopes$token[trigger_scopes$sentence_id == "s027"],
    c("No", "prior", "experience", "required")
  ),
  identical(
    rule_predictions$sentence_id[rule_predictions$rule_negated],
    "s002"
  )
)
```

The corpus contains `No` but not `not` or `never`; those words are in the rule so
the pattern is explicit, not because this batch tests them. The surface-form
backup adds no triggers here. The rule catches the full sentence because the
scope contains `is`, and it misses the headline because the headline has no
helper verb.

## Show the development-set score

The scoring table compares the parser-based rule with the hand judgments on the
same sentences that shaped the rule. It is a demonstration, not a held-out test.

```{r}
negation_scored <- hand_judgments |>
  select(sentence_id, hand_negated) |>
  left_join(rule_predictions, by = "sentence_id") |>
  mutate(
    result = case_when(
      hand_negated & rule_negated ~ "true positive",
      !hand_negated & rule_negated ~ "false positive",
      hand_negated & !rule_negated ~ "false negative",
      TRUE ~ "true negative"
    )
  )

score_summary <- negation_scored |>
  summarise(
    true_positive = sum(result == "true positive"),
    false_positive = sum(result == "false positive"),
    false_negative = sum(result == "false negative"),
    true_negative = sum(result == "true negative"),
    correct = sum(hand_negated == rule_negated),
    total = n(),
    correct_fraction = paste0(correct, "/", total),
    .groups = "drop"
  )

mistakes <- negation_scored |>
  filter(hand_negated != rule_negated) |>
  left_join(sentences |> select(sentence_id, text), by = "sentence_id") |>
  select(sentence_id, text, hand_negated, rule_negated, result, scope_tokens)

knitr::kable(
  score_summary,
  col.names = c(
    "True positives",
    "False positives",
    "False negatives",
    "True negatives",
    "Correct",
    "Total",
    "Correct fraction"
  ),
  caption = "Parser-rule score against hand negation judgments",
  row.names = FALSE
)

knitr::kable(
  mistakes,
  col.names = c("Sentence ID", "Text", "Hand negated", "Rule negated", "Result", "Scope tokens"),
  caption = "The case the simple rule gets wrong",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(score_summary$true_positive, 1L),
  identical(score_summary$false_positive, 0L),
  identical(score_summary$false_negative, 1L),
  identical(score_summary$true_negative, 26L),
  identical(score_summary$correct_fraction, "27/28"),
  identical(mistakes$sentence_id, "s027")
)
```

The rule agrees with the hand marking on 27 of 28 sentences. That number should
not be read as performance. Twenty-six sentences contain no negation, so a rule
that never fired would score 26 of 28. The rule was written after looking at the
two negated sentences and was shaped to catch `s002`, so `s002` is not a test of
it. One true positive and one false negative is a recall count from two cases,
and zero false positives in twenty-six negatives is still consistent with false
alarms in a larger set. The score also belongs to the parser: every trigger,
scope, and lemma comes from the 500-sentence model. Keep the error: `No prior
experience required` has no helper verb, and this rule requires one.

## What to remember

- Negation can reverse a keyword match.
- This corpus tests `no`; it does not test `not` or `never`.
- A scope says which words the trigger controls.
- The simple parser rule catches `s002` and misses `s027`.
- Trigger words are only one way English says no: `Experience is unnecessary`
  has no trigger word and the same shape as `Experience is required`.
- The 27/28 score is a demonstration on the development set.

Use the rule as a triage note beside the sentence, not as an automatic label. The
missed headline is the part to carry forward into the next design.

## Sources

- [Chapman et al. (2001), A Simple Algorithm for Identifying Negated Findings and Diseases in Discharge Summaries](https://doi.org/10.1006/jbin.2001.1029)
- [Universal Dependencies relation `det`](https://universaldependencies.org/u/dep/det.html)
- [Universal Dependencies relation `advmod`](https://universaldependencies.org/u/dep/advmod.html)
- [UDPipe R package reference](https://bnosac.github.io/udpipe/docs/doc0.html)
