---
title: "Training models"
subtitle: "Fit a text model only after the example set is large enough"
author: "Alex Farach"
description: "Learn what model training means for text, why a 28-sentence pilot set is too small, and how a penalized logistic model learns from inaugural paragraphs."
categories: [model development, training models, text classification]
---

Maya has a folder of sentence labels that did its first job well. It helped a
small team agree on what counted as training, scheduling, skills, requirements,
and other workforce details.

Then someone asks for a predictor. The folder has the same labels, the same
care, and a new problem: the material that is useful for learning a codebook can
be far too small for measuring a model.

This lesson makes that break visible. The Riverton teaching set stops at a pilot
check. The modeling work moves to a larger public corpus, presidential inaugural
paragraphs bundled with quanteda.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The inaugural addresses are works of the United
States government and are treated here as public domain text packaged by
quanteda.
:::

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- tell when a labeled set is too small for model testing;
- move from a labeling pilot to a larger teaching corpus;
- build a text classification workflow with tokens, stop words, and tf-idf;
- fit a penalized logistic regression model; and
- report a live score with a baseline and limits attached.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Stop at the pilot set

A **training set** is the labeled material a model learns from. A **test set** is
labeled material held back until scoring. If the test set is tiny, one right or
wrong row can swing the score by a large amount.

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tibble)
  library(rsample)
  library(recipes)
  library(textrecipes)
  library(parsnip)
  library(workflows)
  library(yardstick)
  library(glmnet)
})

riverton_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

set.seed(4300)
riverton_split <- initial_split(
  riverton_sentences,
  prop = 0.75
)
riverton_training <- training(riverton_split)
riverton_testing <- testing(riverton_split)

riverton_label_counts <- riverton_sentences |>
  count(reference_label, name = "sentences") |>
  arrange(desc(sentences), reference_label)

split_size <- tibble(
  part = c("training", "test"),
  rows = c(nrow(riverton_training), nrow(riverton_testing))
)

stopifnot(
  identical(nrow(riverton_sentences), 28L),
  identical(n_distinct(riverton_sentences$reference_label), 5L),
  identical(nrow(riverton_training), 21L),
  identical(nrow(riverton_testing), 7L),
  identical(sum(riverton_label_counts$sentences), 28L)
)

knitr::kable(
  riverton_label_counts,
  col.names = c("Label", "Sentences"),
  caption = "The Riverton pilot labels are too few for a model score",
  row.names = FALSE
)

knitr::kable(
  split_size,
  col.names = c("Split part", "Rows"),
  caption = "A 75 percent split leaves only seven test sentences",
  row.names = FALSE
)
```

The split is deterministic because the seed is `4300`. It leaves
`r nrow(riverton_testing)` test sentences across `r n_distinct(riverton_sentences$reference_label)` labels.
A single changed answer would move accuracy by about
`r round(100 / nrow(riverton_testing), 1)` percentage points. That is not enough
room to learn how far off an automatic answer might be.

## Change the material

The larger corpus comes from `quanteda::data_corpus_inaugural`, reached through
the shared helper. Each row is one paragraph of at least 25 words. The outcome
for this lesson is **era**, which asks whether a paragraph came from an address
before 1900 or from 1900 or later.

```{r}
source("R/inaugural-corpus.R")

paragraphs <- inaugural_paragraphs()

corpus_summary <- tibble(
  measure = c("paragraphs", "speeches", "people", "surnames"),
  value = c(
    nrow(paragraphs),
    n_distinct(paragraphs$speech_id),
    n_distinct(paragraphs$president),
    n_distinct(paragraphs$surname)
  )
)

era_counts <- paragraphs |>
  group_by(era) |>
  summarise(
    paragraphs = n(),
    median_words = median(paragraph_words),
    .groups = "drop"
  )

stopifnot(
  identical(nrow(paragraphs), 1377L),
  identical(n_distinct(paragraphs$speech_id), 60L),
  identical(n_distinct(paragraphs$president), 40L),
  identical(n_distinct(paragraphs$surname), 36L),
  "paragraph_words" %in% names(paragraphs),
  identical(as.integer(era_counts$paragraphs), c(485L, 892L)),
  identical(as.numeric(era_counts$median_words), c(112, 54)),
  identical(levels(paragraphs$era), c("before 1900", "1900 or later"))
)

knitr::kable(
  corpus_summary,
  col.names = c("Measure", "Value"),
  caption = "The inaugural paragraph corpus used for model training",
  row.names = FALSE
)

knitr::kable(
  era_counts,
  col.names = c("Era", "Paragraphs", "Median paragraph words"),
  caption = "Paragraphs by era",
  row.names = FALSE
)
```

The material has `r format(nrow(paragraphs), big.mark = ",")` paragraphs from
`r n_distinct(paragraphs$speech_id)` speeches by `r n_distinct(paragraphs$president)` people.
The helper keeps `president` as the full name and also keeps `surname`, because a
surname-only grouping would merge separate people who share a family name. The
word-count column matters too: paragraphs before 1900 have median length
`r era_counts$median_words[era_counts$era == "before 1900"]`, while later
paragraphs have median length
`r era_counts$median_words[era_counts$era == "1900 or later"]`.

That size does not make the task general to all political speech or all
historical writing. It does give the lesson enough held-out paragraphs to test a
specific claim on this corpus.

## Fit one speech-held-out model

The split keeps each speech whole. A paragraph from one inaugural address cannot
sit in training while another paragraph from the same address sits in testing.
The seed is `4301` for the split and `4302` for the fit.

The recipe turns text into model columns. **Tf-idf** means term frequency times
inverse document frequency. In plain language, a word counts for more in a
paragraph when it is common there and rare elsewhere. Stop words such as `the`
and `of` are removed first, and the recipe keeps the 500 most frequent remaining
tokens.

The model is **logistic regression**, a linear classifier for categories. The
penalty pulls large coefficients toward zero so the model does not chase every
small accident in the training text. Here the penalty is fixed at `0.01` so the
page can run quickly and exactly.

```{r}
baselines <- read_csv(
  "data/inaugural/baselines.csv",
  na = character(),
  col_types = cols(
    task = col_character(),
    outcome = col_character(),
    rows = col_integer(),
    speeches = col_integer(),
    people = col_integer(),
    largest_class = col_character(),
    baseline_accuracy = col_double()
  )
)

set.seed(4301)
era_split <- group_initial_split(
  paragraphs,
  group = speech_id,
  prop = 0.75
)
era_training <- training(era_split)
era_testing <- testing(era_split)

text_recipe <- recipe(era ~ paragraph, data = era_training) |>
  step_tokenize(paragraph) |>
  step_stopwords(paragraph) |>
  step_tokenfilter(paragraph, max_tokens = 500) |>
  step_tfidf(paragraph)

text_model <- logistic_reg(penalty = 0.01, mixture = 0) |>
  set_engine("glmnet")

era_workflow <- workflow() |>
  add_recipe(text_recipe) |>
  add_model(text_model)

set.seed(4302)
era_fitted <- fit(era_workflow, data = era_training)

era_predictions <- predict(era_fitted, era_testing) |>
  bind_cols(era_testing |> select(era))

live_accuracy <- accuracy(
  era_predictions,
  truth = era,
  estimate = .pred_class
) |>
  pull(.estimate)

corpus_baseline <- baselines |>
  filter(task == "era") |>
  pull(baseline_accuracy)

training_majority <- era_training |>
  count(era, sort = TRUE) |>
  slice(1) |>
  pull(era)

split_baseline_accuracy <- mean(era_testing$era == training_majority)

length_cut <- median(era_training$paragraph_words)
length_levels <- levels(era_training$era)
best_length_accuracy <- -Inf
for (flip in c(FALSE, TRUE)) {
  low_label <- if (flip) length_levels[2] else length_levels[1]
  high_label <- if (flip) length_levels[1] else length_levels[2]
  training_length_pred <- if_else(
    era_training$paragraph_words <= length_cut,
    low_label,
    high_label
  )
  training_length_accuracy <- mean(training_length_pred == as.character(era_training$era))
  if (training_length_accuracy > best_length_accuracy) {
    best_length_accuracy <- training_length_accuracy
    best_length_direction <- c(low = low_label, high = high_label)
  }
}
length_predictions <- factor(
  if_else(
    era_testing$paragraph_words <= length_cut,
    best_length_direction[["low"]],
    best_length_direction[["high"]]
  ),
  levels = length_levels
)
length_rule_accuracy <- mean(length_predictions == era_testing$era)

score_table <- tibble(
  measure = c(
    "Live speech-held-out model",
    "Same-split training-majority rule",
    "Same-split paragraph-length rule",
    "Corpus largest-class baseline"
  ),
  accuracy = c(
    live_accuracy,
    split_baseline_accuracy,
    length_rule_accuracy,
    corpus_baseline
  )
)

stopifnot(
  identical(nrow(era_training), 1042L),
  identical(nrow(era_testing), 335L),
  identical(n_distinct(era_training$speech_id), 46L),
  identical(n_distinct(era_testing$speech_id), 14L),
  identical(round(live_accuracy, 4), 0.8209),
  identical(round(split_baseline_accuracy, 4), 0.4179),
  identical(round(length_rule_accuracy, 4), 0.6388),
  identical(round(corpus_baseline, 4), 0.6478),
  all(era_predictions$.pred_class %in% levels(paragraphs$era))
)

knitr::kable(
  score_table |>
    mutate(accuracy = round(accuracy, 4)),
  col.names = c("Measure", "Accuracy"),
  caption = "One live model score with same-split baselines",
  row.names = FALSE
)
```

The live model scores `r round(live_accuracy, 4)` on this speech-held-out test
set. The majority label learned from training would score
`r round(split_baseline_accuracy, 4)` on this exact test set, and a one-number
rule based only on paragraph length would score `r round(length_rule_accuracy, 4)`.
The corpus-wide largest-class baseline is `r round(corpus_baseline, 4)`, but the
same-split comparisons are the ones a deployable rule could use.

## What the fit did not prove

A fitted model is a frozen recipe plus learned weights. It can label new rows in
the same shape as the training rows. It cannot prove that the label is well
chosen, that the corpus stands for other text, or that the score will hold under
a harder split. The next lesson tests those questions with repeated splits.

## What to remember

- A pilot corpus can teach labeling without being large enough for testing.
- A test set with seven rows cannot measure much.
- A speech-held-out split keeps one address from leaking across both sides.
- Tf-idf gives more weight to words that are common in one paragraph and rare in
  the rest.
- A model score needs a baseline beside it.

A trained classifier is only the start of the measurement. The useful question is
whether its score survives a fairer test.

## Sources

- [quanteda inaugural corpus documentation](https://quanteda.io/reference/data_corpus_inaugural.html)
- [textrecipes reference](https://textrecipes.tidymodels.org/)
- [glmnet documentation](https://glmnet.stanford.edu/)
- [rsample group splits](https://rsample.tidymodels.org/reference/group_initial_split.html)
