---
title: "Training models"
subtitle: "Fit a text model only after the example set is large enough"
description: "Learn what model training means for text, why a 28-sentence pilot set is too small, and how a penalized logistic model learns from inaugural paragraphs."
categories: [model development, training models, text classification]
---

Maya has a folder of sentence labels that did its first job well. It helped a
small team agree on what counted as training, scheduling, skills, requirements,
and other workforce details.

Then someone asks for a predictor. The folder has the same labels, the same
care, and a new problem: the material that is useful for learning a codebook can
be far too small for measuring a model.

This lesson makes that break visible. The Riverton teaching set stops at a pilot
check. The modeling work moves to a larger public corpus, presidential inaugural
paragraphs bundled with quanteda.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional
and were created for teaching. The inaugural addresses are works of the United
States government and are treated here as public domain text packaged by
quanteda.
:::

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- tell when a labeled set is too small for model testing;
- move from a labeling pilot to a larger teaching corpus;
- distinguish a recipe, model specification, workflow, fit, and prediction;
- compare a few candidate models without touching the final test set;
- fit the selected penalized logistic regression model; and
- report a live score with a baseline and limits attached.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Stop at the pilot set

A **training set** is the labeled material a model learns from. A **test set** is
labeled material held back until scoring. If the test set is tiny, one right or
wrong row can swing the score by a large amount.

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tibble)
  library(rsample)
  library(recipes)
  library(textrecipes)
  library(parsnip)
  library(workflows)
  library(yardstick)
  library(glmnet)
})

riverton_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

set.seed(4300)
riverton_split <- initial_split(
  riverton_sentences,
  prop = 0.75
)
riverton_training <- training(riverton_split)
riverton_testing <- testing(riverton_split)

riverton_label_counts <- riverton_sentences |>
  count(reference_label, name = "sentences") |>
  arrange(desc(sentences), reference_label)

split_size <- tibble(
  part = c("training", "test"),
  rows = c(nrow(riverton_training), nrow(riverton_testing))
)

knitr::kable(
  riverton_label_counts,
  col.names = c("Label", "Sentences"),
  caption = "The Riverton pilot labels are too few for a model score",
  row.names = FALSE
)

knitr::kable(
  split_size,
  col.names = c("Split part", "Rows"),
  caption = "A 75 percent split leaves only seven test sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(riverton_sentences), 28L),
  identical(n_distinct(riverton_sentences$reference_label), 5L),
  identical(nrow(riverton_training), 21L),
  identical(nrow(riverton_testing), 7L),
  identical(sum(riverton_label_counts$sentences), 28L)
)
```

The split is deterministic because the seed is `4300`. It leaves
`r nrow(riverton_testing)` test sentences across `r n_distinct(riverton_sentences$reference_label)` labels.
A single changed answer would move accuracy by about
`r round(100 / nrow(riverton_testing), 1)` percentage points. That is not enough
room to learn how far off an automatic answer might be.

## Change the material

The larger corpus comes from `quanteda::data_corpus_inaugural`, reached through
the shared helper. Each row is one paragraph of at least 25 words. The outcome
for this lesson is **era**, which asks whether a paragraph came from an address
before 1900 or from 1900 or later.

```{r}
source("R/inaugural-corpus.R")

paragraphs <- inaugural_paragraphs()

corpus_summary <- tibble(
  measure = c("paragraphs", "speeches", "people", "surnames"),
  value = c(
    nrow(paragraphs),
    n_distinct(paragraphs$speech_id),
    n_distinct(paragraphs$president),
    n_distinct(paragraphs$surname)
  )
)

era_counts <- paragraphs |>
  group_by(era) |>
  summarise(
    paragraphs = n(),
    median_words = median(paragraph_words),
    .groups = "drop"
  )

knitr::kable(
  corpus_summary,
  col.names = c("Measure", "Value"),
  caption = "The inaugural paragraph corpus used for model training",
  row.names = FALSE
)

knitr::kable(
  era_counts,
  col.names = c("Era", "Paragraphs", "Median paragraph words"),
  caption = "Paragraphs by era",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(paragraphs), 1377L),
  identical(n_distinct(paragraphs$speech_id), 60L),
  identical(n_distinct(paragraphs$president), 40L),
  identical(n_distinct(paragraphs$surname), 36L),
  "paragraph_words" %in% names(paragraphs),
  identical(as.integer(era_counts$paragraphs), c(485L, 892L)),
  identical(as.numeric(era_counts$median_words), c(112, 54)),
  identical(levels(paragraphs$era), c("before 1900", "1900 or later"))
)
```

The material has `r format(nrow(paragraphs), big.mark = ",")` paragraphs from
`r n_distinct(paragraphs$speech_id)` speeches by `r n_distinct(paragraphs$president)` people.
The helper keeps `president` as the full name and also keeps `surname`, because a
surname-only grouping would merge separate people who share a family name. The
word-count column matters too: paragraphs before 1900 have median length
`r era_counts$median_words[era_counts$era == "before 1900"]`, while later
paragraphs have median length
`r era_counts$median_words[era_counts$era == "1900 or later"]`.

That size does not make the task general to all political speech or all
historical writing. It does give the lesson enough held-out paragraphs to test a
specific claim on this corpus.

## Keep the final test set out of model selection

The split keeps each speech whole. A paragraph from one inaugural address cannot
sit in training while another paragraph from the same address sits in testing.
The seed is `4301` for the split.

The test speeches get one job in this worked example: estimate performance
after the model choice is finished. Choosing a model from its test score would
spend that evidence before the final check. This is still an internal holdout
from a corpus used throughout the section, not confirmation on new text from
another source.

Inside the training set, **cross-validation** makes several temporary
train-and-check splits. Each temporary check is called a **fold**. The folds are
often called **CV folds**. They are grouped by speech here, so every paragraph
from one address stays together. Five folds put 9 or 10 speeches in each
assessment set while leaving both era labels represented.

```{r}
baselines <- read_csv(
  "data/inaugural/baselines.csv",
  na = character(),
  col_types = cols(
    task = col_character(),
    outcome = col_character(),
    rows = col_integer(),
    speeches = col_integer(),
    people = col_integer(),
    largest_class = col_character(),
    baseline_accuracy = col_double()
  )
)

model_comparison <- read_csv(
  "data/inaugural/model-comparison.csv",
  na = "",
  col_types = cols(
    rank = col_integer(),
    candidate = col_character(),
    family = col_character(),
    penalty = col_double(),
    mtry = col_integer(),
    min_n = col_integer(),
    mean_bal_accuracy = col_double(),
    std_err_bal_accuracy = col_double(),
    mean_accuracy = col_double(),
    std_err_accuracy = col_double(),
    selected = col_logical(),
    selection_reason = col_character()
  )
)

fold_summary <- read_csv(
  "data/inaugural/model-comparison-folds.csv",
  na = character(),
  col_types = cols(
    fold = col_character(),
    analysis_speeches = col_integer(),
    assessment_speeches = col_integer(),
    assessment_paragraphs = col_integer(),
    before_1900 = col_integer(),
    later_1900 = col_integer()
  )
)

set.seed(4301)
era_split <- group_initial_split(
  paragraphs,
  group = speech_id,
  prop = 0.75
)
era_training <- training(era_split)
era_testing <- testing(era_split)

set.seed(4310)
era_folds <- group_vfold_cv(
  era_training,
  group = speech_id,
  v = 5,
  strata = era,
  balance = "groups"
)

fold_group_overlap <- vapply(
  era_folds$splits,
  \(split) {
    length(intersect(
      unique(analysis(split)$speech_id),
      unique(assessment(split)$speech_id)
    ))
  },
  integer(1)
)

knitr::kable(
  fold_summary,
  col.names = c(
    "Fold",
    "Training speeches",
    "Assessment speeches",
    "Assessment paragraphs",
    "Before 1900",
    "1900 or later"
  ),
  caption = "Five grouped folds inside the training speeches",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(era_training), 1042L),
  identical(nrow(era_testing), 335L),
  identical(n_distinct(era_training$speech_id), 46L),
  identical(n_distinct(era_testing$speech_id), 14L),
  identical(nrow(era_folds), 5L),
  all(fold_group_overlap == 0L),
  identical(fold_summary$assessment_speeches, c(10L, 9L, 9L, 9L, 9L)),
  all(fold_summary$before_1900 > 0L),
  all(fold_summary$later_1900 > 0L)
)
```

Stratification distributes early and later speeches across the folds; it does
not equalize paragraph counts. The table shows that the paragraph-level class
mix still varies, which is one reason balanced accuracy is the primary measure.

## Compare a small set of candidates

This is not a search across every model R can fit. The comparison asks whether a
different kind of decision rule changes the answer:

- **Ridge logistic regression** keeps every word feature but pulls all fitted
  weights toward zero.
- **Lasso logistic regression** uses the same linear model but can set some word
  weights exactly to zero.
- A **random forest** combines many branching rules and can learn relationships
  that are not linear.

The **tuning parameters** control how each candidate learns. Ridge tests four
penalties from `0.01` to `10`; lasso tests four from `0.0001` to `0.1`.
Their fitted penalty paths span different scales, so copying one grid across
both would waste settings on duplicate or intercept-only fits. The forest tries
six combinations of features per split (`mtry`) and the minimum rows needed in
a branch (`min_n`). The builder in `data-raw/build-model-comparison.R` fits all
14 settings: four ridge, four lasso, and six forest. It never reads the final
test speeches.

Balanced accuracy is the primary selection measure because it gives the two era
classes equal weight. Raw accuracy appears beside it so the effect of the
uneven class counts stays visible.

The table shows the two tied ridge settings and the leading lasso and forest
settings. The standard error describes how each score moved across these five
folds; five related folds do not provide a confidence interval for other
corpora.

```{r}
comparison_display <- model_comparison |>
  filter(selected | rank %in% c(2L, 3L, 5L)) |>
  mutate(
    setting = case_when(
      candidate == "random forest" ~
        paste0("mtry ", mtry, "; min_n ", min_n),
      TRUE ~ paste0("penalty ", penalty)
    ),
    chosen = if_else(selected, "yes", "no")
  ) |>
  select(
    candidate,
    setting,
    mean_bal_accuracy,
    std_err_bal_accuracy,
    mean_accuracy,
    chosen
  )

selected_candidate <- model_comparison |>
  filter(selected)

knitr::kable(
  comparison_display |>
    mutate(
      across(
        c(
          mean_bal_accuracy,
          std_err_bal_accuracy,
          mean_accuracy
        ),
        \(value) round(value, 4)
      )
    ),
  col.names = c(
    "Candidate",
    "Setting",
    "Mean balanced accuracy",
    "Standard error",
    "Mean accuracy",
    "Chosen"
  ),
  caption = "Leading settings from 14 candidates compared on training folds",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(model_comparison), 14L),
  identical(nrow(comparison_display), 4L),
  identical(sum(model_comparison$selected), 1L),
  identical(selected_candidate$candidate, "ridge logistic"),
  identical(selected_candidate$penalty, 0.01),
  identical(round(selected_candidate$mean_bal_accuracy, 4), 0.8008),
  identical(
    round(
      model_comparison$mean_bal_accuracy[
        model_comparison$candidate == "ridge logistic" &
          model_comparison$penalty == 0.1
      ],
      4
    ),
    0.7901
  ),
  identical(
    round(
      model_comparison$mean_bal_accuracy[
        model_comparison$candidate == "lasso logistic" &
          model_comparison$penalty == 0.001
      ],
      4
    ),
    0.7809
  )
)
```

Ridge with penalty `0.01` has the highest mean balanced accuracy,
`r round(selected_candidate$mean_bal_accuracy, 4)`. The fold-to-fold standard
errors overlap the nearby candidates, so the table does not establish a
universal winner. We keep ridge because it leads this comparison and leaves a
linear model whose coefficients can be inspected in lesson 45.

The random forest is a useful check rather than a required upgrade. Its best raw
accuracy is competitive, but its lower balanced accuracy shows that more correct
paragraphs can still mean less even performance across classes.

## Build the selected workflow

A tidymodels pipeline has four parts:

1. A **recipe** is a set of instructions that turns raw text into numeric model
   columns. Each `step_*()` line adds one instruction.
2. A **model specification** names the kind of model and its settings before it
   sees data.
3. An **engine** names the package that performs the calculations. Here that
   package is glmnet.
4. A **workflow** keeps the recipe and model specification together. **Fitting**
   the workflow hands it training data so it can estimate model weights.

The recipe turns text into model columns. **Tf-idf** means term frequency times
inverse document frequency. In plain language, a word counts for more in a
paragraph when it is common there and rare elsewhere. Stop words such as `the`
and `of` are removed first, and the recipe keeps the 500 most frequent remaining
tokens.

The selected model is **logistic regression**, a linear classifier for
categories. The penalty pulls large coefficients toward zero so the model does
not chase every small accident in the training text.

```{r}
text_recipe <- recipe(era ~ paragraph, data = era_training) |>
  step_tokenize(paragraph) |>
  step_stopwords(paragraph) |>
  step_tokenfilter(paragraph, max_tokens = 500) |>
  step_tfidf(paragraph)

text_model <- logistic_reg(penalty = 0.01, mixture = 0) |>
  set_engine("glmnet")

era_workflow <- workflow() |>
  add_recipe(text_recipe) |>
  add_model(text_model)

set.seed(4302)
era_fitted <- fit(era_workflow, data = era_training)

era_predictions <- predict(era_fitted, era_testing) |>
  bind_cols(era_testing |> select(era))

live_metrics <- metric_set(accuracy, bal_accuracy)(
  era_predictions,
  truth = era,
  estimate = .pred_class
)
live_accuracy <- live_metrics |>
  filter(.metric == "accuracy") |>
  pull(.estimate)
live_bal_accuracy <- live_metrics |>
  filter(.metric == "bal_accuracy") |>
  pull(.estimate)

corpus_baseline <- baselines |>
  filter(task == "era") |>
  pull(baseline_accuracy)

training_majority <- era_training |>
  count(era, sort = TRUE) |>
  slice(1) |>
  pull(era)

split_baseline_accuracy <- mean(era_testing$era == training_majority)

score_table <- tibble(
  measure = c(
    "Final test accuracy",
    "Final test balanced accuracy",
    "Same-split training-majority rule",
    "Corpus largest-class baseline"
  ),
  value = c(
    live_accuracy,
    live_bal_accuracy,
    split_baseline_accuracy,
    corpus_baseline
  )
)

knitr::kable(
  score_table |>
    mutate(value = round(value, 4)),
  col.names = c("Measure", "Value"),
  caption = "One held-out test after candidate selection",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(round(live_accuracy, 4), 0.8209),
  identical(round(live_bal_accuracy, 4), 0.8361),
  identical(round(split_baseline_accuracy, 4), 0.4179),
  identical(round(corpus_baseline, 4), 0.6478),
  all(era_predictions$.pred_class %in% levels(paragraphs$era))
)
```

Only after the candidate and penalty are fixed do we use the 14 final test
speeches. The model scores `r round(live_accuracy, 4)` accuracy and
`r round(live_bal_accuracy, 4)` balanced accuracy. The majority label learned
from training would score `r round(split_baseline_accuracy, 4)` on this exact
test set. The corpus-wide largest-class baseline is
`r round(corpus_baseline, 4)`, but the same-split comparison is the one a
deployable rule could use.

## What the fit did not prove

A fitted model is a frozen recipe plus learned weights. It can label new rows in
the same shape as the training rows. It cannot prove that the label is well
chosen, that the corpus stands for other text, or that the score will hold under
a harder split. The next lesson tests those questions with repeated splits.

## What to remember

- A pilot corpus can teach labeling without being large enough for testing.
- A test set with seven rows cannot measure much.
- A speech-held-out split keeps one address from leaking across both sides.
- Grouped folds compare candidates without spending the final test set.
- A recipe, model specification, engine, and workflow have different jobs.
- Tf-idf gives more weight to words that are common in one paragraph and rare in
  the rest.
- A model score needs a baseline beside it.

A trained classifier is only the start of the measurement. The useful question is
whether its score survives a fairer test.

## Sources

- [quanteda inaugural corpus documentation](https://quanteda.io/reference/data_corpus_inaugural.html)
- [textrecipes reference](https://textrecipes.tidymodels.org/)
- [glmnet documentation](https://glmnet.stanford.edu/)
- [rsample group splits](https://rsample.tidymodels.org/reference/group_initial_split.html)
