---
title: "Evaluating models"
subtitle: "Read a score only after asking what it was compared with"
description: "Learn why model evaluation needs a baseline, repeated splits, and a split scheme that matches the claim."
categories: [model development, evaluation, text classification]
---

A program manager sees one number on a slide: 61 percent accurate. It sounds
better than guessing, so the room starts to relax.

Then she asks what the model would have scored if it had always chosen the
larger class. The answer changes the meeting. The score was not evidence of a
working model after all.

This lesson reads a committed split study for three related text models. The era
model is useful against cheap alternatives. The party model, with these features
and this fixed penalty, does not learn a reliable party signal.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- separate candidate selection, final testing, and robustness checks;
- compare every score with a deployable largest-class baseline;
- read the spread across repeated train/test splits;
- explain what paragraph, speech, and person splits measure;
- use a confusion matrix and balanced accuracy; and
- say when a specific model did not learn the target.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Start with the baseline

A **baseline** is the simple rule a model has to beat. The largest-class baseline
always predicts the most common label in training data. It is crude, which is
exactly why it is a useful check. If a model cannot beat that rule, its score
should not be sold as success.

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(stringr)
  library(ggplot2)
  library(rsample)
  library(recipes)
  library(textrecipes)
  library(parsnip)
  library(workflows)
  library(yardstick)
  library(broom)
  library(glmnet)
})

study <- read_csv(
  "data/inaugural/split-study.csv",
  na = c("", "NA"),
  col_types = cols(
    task = col_character(),
    split_scheme = col_character(),
    replicate = col_integer(),
    accuracy = col_double(),
    bal_accuracy = col_double(),
    train_majority_accuracy = col_double(),
    test_majority_rate = col_double(),
    length_rule_accuracy = col_double(),
    test_rows = col_integer(),
    test_speeches = col_integer()
  )
)

baselines <- read_csv(
  "data/inaugural/baselines.csv",
  na = character(),
  col_types = cols(
    task = col_character(),
    outcome = col_character(),
    rows = col_integer(),
    speeches = col_integer(),
    people = col_integer(),
    largest_class = col_character(),
    baseline_accuracy = col_double()
  )
)

split_levels <- c("paragraph", "speech", "president")
task_levels <- c("era", "era_party_rows", "party")

task_labels <- c(
  era = "era",
  era_party_rows = "era on party rows",
  party = "party"
)

split_summary <- study |>
  mutate(
    task = factor(task, levels = task_levels),
    split_scheme = factor(split_scheme, levels = split_levels)
  ) |>
  group_by(task, split_scheme) |>
  summarise(
    replicates = n(),
    mean_accuracy = mean(accuracy),
    mean_bal_accuracy = mean(bal_accuracy, na.rm = TRUE),
    bal_accuracy_missing = sum(is.na(bal_accuracy)),
    mean_train_majority = mean(train_majority_accuracy),
    mean_test_majority = mean(test_majority_rate),
    mean_length_rule = mean(length_rule_accuracy),
    low_accuracy = min(accuracy),
    high_accuracy = max(accuracy),
    low_train_majority = min(train_majority_accuracy),
    high_train_majority = max(train_majority_accuracy),
    low_length_rule = min(length_rule_accuracy),
    high_length_rule = max(length_rule_accuracy),
    mean_test_speeches = mean(test_speeches),
    .groups = "drop"
  ) |>
  arrange(task, split_scheme)

baseline_display <- baselines |>
  mutate(task = factor(task, levels = task_levels)) |>
  arrange(task) |>
  transmute(
    task = recode(as.character(task), !!!task_labels),
    rows,
    speeches,
    people,
    largest_class,
    baseline_accuracy
  )

knitr::kable(
  baseline_display |>
    mutate(baseline_accuracy = round(baseline_accuracy, 4)),
  col.names = c("Task", "Rows", "Speeches", "People", "Largest class", "Corpus baseline"),
  caption = "Corpus-level baselines and task sizes",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(study), 90L),
  identical(nrow(baselines), 3L),
  identical(sort(unique(study$replicate)), 1:10),
  all(split_summary$replicates == 10L),
  identical(round(baseline_display$baseline_accuracy, 4), c(0.6478, 0.7452, 0.6065)),
  identical(baseline_display$people, c(40L, 31L, 31L)),
  identical(baseline_display$speeches, c(60L, 47L, 47L)),
  identical(baseline_display$largest_class, c("1900 or later", "1900 or later", "Republican"))
)
```

The era task has `r baseline_display$rows[baseline_display$task == "era"]` rows,
`r baseline_display$speeches[baseline_display$task == "era"]` speeches, and
`r baseline_display$people[baseline_display$task == "era"]` people. The party
model can train only on rows with a Democratic or Republican label, so it uses
`r baseline_display$rows[baseline_display$task == "party"]` rows from
`r baseline_display$speeches[baseline_display$task == "party"]` speeches by
`r baseline_display$people[baseline_display$task == "party"]` people. The
remaining rows come from other party labels. Each task also selects its own 500
tokens from its own training data.

## Selection, testing, and robustness do different jobs

Lesson 43 compared candidate settings with five grouped folds made only from
the training speeches. It then used the 14 untouched test speeches once, after
the candidate and penalty were fixed. That is the final test for that modeling
decision inside this worked example. A new corpus would be needed to test
whether the result transfers to another collection.

The repeated study below asks a different question. It redraws the split 10
times to show how scores move when different speeches land on each side. These
repeated holdouts are a **robustness check**, not 10 new untouched test sets,
and they were not used to choose the candidate in lesson 43.

**Accuracy** is the share of rows with the right label. When classes are uneven,
a model can get many rows right by favoring the larger class. A **confusion
matrix** shows which labels were confused, and **balanced accuracy** averages the
accuracy within each class so the smaller class still counts.

## Repeated splits change the answer

The study runs 10 replicates for each task and split scheme. A **replicate** is a
fresh split made by the same rule. The mean tells the center of the results, and
the low and high values show how much the score moved across the 10 runs. Unlike
five-fold cross-validation, a repeated holdout does not guarantee that every
speech takes one turn in assessment. Its purpose here is sensitivity to the
split, not model selection.

```{r}
score_table <- split_summary |>
  transmute(
    task = recode(as.character(task), !!!task_labels),
    split_scheme = as.character(split_scheme),
    mean_accuracy = round(mean_accuracy, 4),
    mean_bal_accuracy = round(mean_bal_accuracy, 4),
    train_majority_accuracy = round(mean_train_majority, 4),
    test_majority_rate = round(mean_test_majority, 4),
    length_rule_accuracy = round(mean_length_rule, 4),
    low_accuracy = round(low_accuracy, 4),
    high_accuracy = round(high_accuracy, 4),
    missing_bal_accuracy = bal_accuracy_missing
  )

get_summary <- function(task_name, scheme_name) {
  split_summary |>
    filter(task == task_name, split_scheme == scheme_name)
}

era_speech <- get_summary("era", "speech")
era_president <- get_summary("era", "president")
era_party_speech <- get_summary("era_party_rows", "speech")
era_party_president <- get_summary("era_party_rows", "president")
party_paragraph <- get_summary("party", "paragraph")
party_speech <- get_summary("party", "speech")
party_president <- get_summary("party", "president")

na_replicate <- study |>
  filter(task == "era_party_rows", split_scheme == "president", is.na(bal_accuracy))

era_speech_range_width <- era_speech$high_accuracy - era_speech$low_accuracy

era_speech_length_gap <- era_speech$mean_accuracy - era_speech$mean_length_rule

era_party_speech_gap <- era_party_speech$mean_accuracy - era_party_speech$mean_train_majority

paired_comparisons <- bind_rows(
  study |>
    filter(task == "era", split_scheme == "speech") |>
    summarise(
      comparison = "era, all rows",
      model_wins = sum(accuracy > train_majority_accuracy),
      ties = sum(accuracy == train_majority_accuracy),
      model_losses = sum(accuracy < train_majority_accuracy)
    ),
  study |>
    filter(task == "era_party_rows", split_scheme == "speech") |>
    summarise(
      comparison = "era, party rows",
      model_wins = sum(accuracy > train_majority_accuracy),
      ties = sum(accuracy == train_majority_accuracy),
      model_losses = sum(accuracy < train_majority_accuracy)
    ),
  study |>
    filter(task == "party", split_scheme == "paragraph") |>
    summarise(
      comparison = "party, paragraph split",
      model_wins = sum(accuracy > train_majority_accuracy),
      ties = sum(accuracy == train_majority_accuracy),
      model_losses = sum(accuracy < train_majority_accuracy)
    )
)

party_president_bal_range <- range(
  study$bal_accuracy[
    study$task == "party" &
      study$split_scheme == "president"
  ]
)

knitr::kable(
  score_table,
  col.names = c(
    "Task", "Split", "Model accuracy", "Balanced accuracy",
    "Training-majority rule", "Test-majority rate", "Length rule",
    "Model low", "Model high", "Missing balanced accuracy"
  ),
  caption = "Ten split replicates for each task and split scheme",
  row.names = FALSE
)

knitr::kable(
  paired_comparisons,
  col.names = c("Comparison", "Model wins", "Ties", "Model losses"),
  caption = "Paired model-versus-training-majority results across 10 replicates",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(score_table$missing_bal_accuracy, c(rep(0L, 5L), 1L, rep(0L, 3L))),
  identical(round(era_speech$mean_accuracy, 4), 0.8457),
  identical(round(era_speech$mean_bal_accuracy, 4), 0.8516),
  identical(round(era_speech$mean_train_majority, 4), 0.6282),
  identical(round(era_speech$mean_length_rule, 4), 0.6827),
  identical(round(era_speech$low_accuracy, 4), 0.7970),
  identical(round(era_speech$high_accuracy, 4), 0.9039),
  identical(round(era_speech$low_train_majority, 4), 0.3576),
  identical(round(era_speech$high_train_majority, 4), 0.8318),
  identical(round(era_speech$low_length_rule, 4), 0.5255),
  identical(round(era_speech$high_length_rule, 4), 0.8498),
  identical(round(era_party_speech$mean_accuracy, 4), 0.8240),
  identical(round(era_party_speech$mean_train_majority, 4), 0.8086),
  identical(round(party_paragraph$mean_accuracy, 4), 0.6060),
  identical(round(party_paragraph$mean_train_majority, 4), 0.6067),
  identical(round(party_speech$mean_accuracy, 4), 0.5514),
  identical(round(party_speech$mean_train_majority, 4), 0.6563),
  identical(round(party_president$mean_accuracy, 4), 0.4997),
  identical(round(party_president$mean_bal_accuracy, 4), 0.5090),
  identical(paired_comparisons$model_wins, c(10L, 6L, 3L)),
  identical(paired_comparisons$ties, c(0L, 0L, 1L)),
  identical(paired_comparisons$model_losses, c(0L, 4L, 6L)),
  identical(round(party_president_bal_range, 4), c(0.4769, 0.5595)),
  identical(nrow(na_replicate), 1L),
  identical(na_replicate$replicate, 3L),
  identical(na_replicate$test_majority_rate, 1)
)
```

The speech split is the main era test. The model averages
`r round(era_speech$mean_accuracy, 4)`, while the deployable majority rule
averages `r round(era_speech$mean_train_majority, 4)` and the paragraph-length
rule averages `r round(era_speech$mean_length_rule, 4)`. The model's lowest
speech-split replicate is `r round(era_speech$low_accuracy, 4)`, which is still
higher than the best mean reached by either cheap rule. Across the 10 speech
replicates, the model ranges from `r round(era_speech$low_accuracy, 4)` to
`r round(era_speech$high_accuracy, 4)`. The deployable rule ranges from
`r round(era_speech$low_train_majority, 4)` to
`r round(era_speech$high_train_majority, 4)`, and the length rule ranges from
`r round(era_speech$low_length_rule, 4)` to
`r round(era_speech$high_length_rule, 4)`.

A score is meaningful only against the best cheap alternative you bothered to
try. Here the length rule matters because paragraph length is confounded with
era. The model still beats that one-number rule by about
`r round(100 * era_speech_length_gap, 1)` points at the speech split, so the text
features are doing real work.

```{r}
#| label: fig-split-study
#| fig-cap: "Accuracy over 10 replicates for the model and cheap rules."
#| fig-alt: "A dot and line chart faceted by task. In the era panel, model accuracy stays above both the training-majority rule and the paragraph-length rule at all three split schemes: paragraph, speech, and president. In the party panel, model accuracy sits near or below the training-majority rule at all three split schemes."

plot_data <- study |>
  filter(task %in% c("era", "party")) |>
  mutate(
    task = recode(task, !!!task_labels),
    split_scheme = factor(split_scheme, levels = split_levels)
  ) |>
  select(task, split_scheme, replicate, accuracy, train_majority_accuracy, length_rule_accuracy) |>
  pivot_longer(
    cols = c(accuracy, train_majority_accuracy, length_rule_accuracy),
    names_to = "measure",
    values_to = "score"
  ) |>
  mutate(
    measure = recode(
      measure,
      accuracy = "model",
      train_majority_accuracy = "training-majority rule",
      length_rule_accuracy = "paragraph-length rule"
    )
  )

# The y scale is left to the data. Fixing the lower bound at 0.3 would drop the
# two replicates where the training-majority rule scores 0.27, and those
# are exactly the cases worth seeing: the rule collapses when the class that was
# largest in training is the smaller one in test.
ggplot(plot_data, aes(x = split_scheme, y = score, color = measure)) +
  geom_point(alpha = 0.55, position = position_jitter(width = 0.08, height = 0)) +
  stat_summary(fun = mean, geom = "line", aes(group = measure), linewidth = 0.8) +
  stat_summary(fun = mean, geom = "point", size = 2.4) +
  facet_wrap(~task) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.01)) +
  labs(
    x = "Split scheme",
    y = "Accuracy",
    color = "Measure"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  identical(nrow(plot_data), 180L),
  identical(
    plot_data |>
      count(task, split_scheme, measure) |>
      pull(n),
    rep(10L, 18L)
  ),
  all(plot_data$score >= 0 & plot_data$score <= 1),
  identical(round(min(plot_data$score), 4), 0.2749)
)
```

The party model does not beat its deployable majority rule on average at the
paragraph split: `r round(party_paragraph$mean_accuracy, 4)` for the model
against `r round(party_paragraph$mean_train_majority, 4)` for the rule. Across
the 10 paired replicates, the model wins 3, ties 1, and loses 6. At the speech
split, the model averages `r round(party_speech$mean_accuracy, 4)` against
`r round(party_speech$mean_train_majority, 4)`. Under the person split, accuracy
is `r round(party_president$mean_accuracy, 4)` and mean balanced accuracy is
`r round(party_president$mean_bal_accuracy, 4)`. Its replicate values range from
`r round(party_president_bal_range[[1]], 4)` to
`r round(party_president_bal_range[[2]], 4)`, straddling the theoretical
two-class chance value of 0.5. The conclusion is model-level: this penalty,
recipe, and row set did not learn party reliably.

## The controlled comparison is smaller

The era and party tasks are not a clean pair by default. Party keeps only the
Democratic and Republican speeches, which leaves
`r baseline_display$rows[baseline_display$task == "party"]` of the full
`r baseline_display$rows[baseline_display$task == "era"]` paragraphs. The dropped
rows are from other party labels, and those rows are concentrated in the earliest
speeches with longer paragraphs.

The controlled comparison is `era_party_rows`, which runs the era label on the
same rows as the party model. On the full era corpus, the model beats the
deployable rule in all 10 speech-level replicates. On the party-row subset, the
model averages `r round(era_party_speech$mean_accuracy, 4)` against
`r round(era_party_speech$mean_train_majority, 4)` for the training-majority
rule, but it wins 6 paired replicates and loses 4. The mean gap is only
`r round(100 * era_party_speech_gap, 1)` points because this subset is much more
imbalanced, and its direction is not stable across splits. Era is learnable on
the full rows here and party is not learned by this model, but the controlled
subset does not support the same strong era conclusion.

For `era_party_rows` under the person split, replicate
`r na_replicate$replicate[[1]]` has undefined balanced accuracy because the test
set has only one class. Its test-majority rate is
`r round(na_replicate$test_majority_rate[[1]], 4)`, so the accuracy for that one
replicate is not a two-class result. The summary table uses `na.rm = TRUE` for
balanced accuracy and reports the missing count so the case stays visible.

## Read the final confusion matrix

**Sensitivity** is the share of `before 1900` paragraphs assigned to that first
factor level. **Specificity** is the share of `1900 or later` paragraphs assigned
to the second level. Naming the event level matters because reversing the factor
levels reverses which class each measure describes.

```{r}
source("R/inaugural-corpus.R")
paragraphs <- inaugural_paragraphs()

set.seed(4301)
era_split <- group_initial_split(paragraphs, group = speech_id, prop = 0.75)
era_training <- training(era_split)
era_testing <- testing(era_split)

era_recipe <- recipe(era ~ paragraph, data = era_training) |>
  step_tokenize(paragraph) |>
  step_stopwords(paragraph) |>
  step_tokenfilter(paragraph, max_tokens = 500) |>
  step_tfidf(paragraph)

era_model <- logistic_reg(penalty = 0.01, mixture = 0) |>
  set_engine("glmnet")

era_workflow <- workflow() |>
  add_recipe(era_recipe) |>
  add_model(era_model)

set.seed(4302)
era_fitted <- fit(era_workflow, data = era_training)

era_predictions <- predict(era_fitted, era_testing) |>
  bind_cols(era_testing |> select(era))

final_training_majority <- era_training |>
  count(era, sort = TRUE) |>
  slice(1) |>
  pull(era)
final_split_baseline <- mean(era_testing$era == final_training_majority)

confusion <- conf_mat(
  era_predictions,
  truth = era,
  estimate = .pred_class
)
confusion_table <- as.data.frame(confusion$table) |>
  as_tibble()

metric_table <- metric_set(accuracy, bal_accuracy, sens, spec)(
  era_predictions,
  truth = era,
  estimate = .pred_class
) |>
  select(.metric, .estimate) |>
  bind_rows(
    tibble(
      .metric = "training_majority_accuracy",
      .estimate = final_split_baseline
    )
  )

knitr::kable(
  confusion_table,
  col.names = c("Prediction", "Reference label", "Paragraphs"),
  caption = "Confusion matrix for one speech-held-out era model",
  row.names = FALSE
)

knitr::kable(
  metric_table |>
    mutate(.estimate = round(.estimate, 4)),
  col.names = c("Measure", "Value"),
  caption = "Model measures and the same-split deployable baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(era_testing), 335L),
  identical(as.character(final_training_majority), "1900 or later"),
  identical(round(final_split_baseline, 4), 0.4179),
  identical(confusion_table$Freq, c(145, 50, 10, 130)),
  identical(
    round(metric_table$.estimate, 4),
    c(0.8209, 0.8361, 0.7436, 0.9286, 0.4179)
  )
)
```

The model gets `r confusion_table$Freq[[1]]` early-era paragraphs right and
misses `r confusion_table$Freq[[2]]` of them. It gets
`r confusion_table$Freq[[4]]` later-era paragraphs right and misses
`r confusion_table$Freq[[3]]` of them. Balanced accuracy is
`r round(metric_table$.estimate[metric_table$.metric == "bal_accuracy"], 4)`,
which gives both classes equal weight. The model accuracy is
`r round(metric_table$.estimate[metric_table$.metric == "accuracy"], 4)`;
the deployable training-majority rule scores
`r round(final_split_baseline, 4)` on these same rows.

## What else could produce this score

Era is a deterministic function of `year`, and `year` is constant within a
speech. The era study has `r baseline_display$speeches[baseline_display$task == "era"]`
independent speech units, not `r baseline_display$rows[baseline_display$task == "era"]`
independent paragraphs. The speech split averages about
`r round(era_speech$mean_test_speeches, 1)` speeches per test set. That is why the
model range across speech splits is `r round(era_speech_range_width, 4)`, not the
tighter range one might expect from hundreds of independent rows. The 10
replicates describe resampling variability inside one corpus; they are not a
confidence interval for political speech beyond this collection.

The next lesson prints this model's largest coefficients. Read them when you get
there, because they cut both ways. The terms pushing toward the earlier label are
mostly formal-register words, the vocabulary of a certain kind of nineteenth-
century public address rather than a certain kind of subject. The terms pushing
the other way include words like `industrial`, `economic`, `international`, and
`federal`, which are closer to genuine changes in what the speeches are about.
So part of this score is a style detector and part of it is a subject detector,
and nothing here separates the two. Nineteenth- and twenty-first-century texts
can also differ because of transcription and typesetting conventions, which no
measurement on this page would distinguish from either.

This lesson supports the limited claim that era is learnable from this text. It
does not show that the model learned what a period was about. A features-only
stylometric model, or a subsample matched on paragraph length, would be needed to
pull period vocabulary apart from period register.

## What to remember

- A score with no baseline is hard to interpret.
- Candidate selection must finish before the final test set is opened.
- Group splits need same-split baselines, not only corpus-wide baselines.
- One split gives one number; repeated splits show the spread.
- The split scheme decides what the score is measuring.
- This text model learned era well. This model, with this penalty and these
  features, did not learn party, which is a smaller claim than saying the task
  is impossible.

A useful evaluation does not flatter the model. It tells you which answer you are
allowed to trust.

## Sources

- [yardstick metrics reference](https://yardstick.tidymodels.org/)
- [rsample initial splits](https://rsample.tidymodels.org/reference/initial_split.html)
- [rsample grouped splits](https://rsample.tidymodels.org/reference/group_initial_split.html)
- [rsample grouped V-fold cross-validation](https://rsample.tidymodels.org/reference/group_vfold_cv.html)
- [tidymodels tuning](https://www.tidymodels.org/start/tuning/)
- [tidymodels workflows](https://workflows.tidymodels.org/)
