---
title: "Explaining models"
subtitle: "Inspect the words a fitted text model learned to weight"
author: "Alex Farach"
description: "Learn how to read linear model coefficients and why a plausible explanation can still belong to a weak model."
categories: [model development, explanation, text classification]
---

A reviewer asks why the model called a paragraph modern. The answer cannot be
"because the computer said so." She needs to see which words pulled the score.

That request is reasonable. It is also dangerous if the score has not been
checked. A weak model can still hand back a tidy list of words that looks like an
explanation.

This lesson fits the same kind of text model twice. The era model has useful
evaluation evidence. The party model, with these features and this fixed penalty,
does not.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- extract coefficients from a fitted workflow;
- read positive and negative coefficients in plain language;
- inspect terms weighted by an era classifier;
- mask terms from a weak political classifier; and
- keep explanation separate from evidence that a model is right.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Fit the era model for inspection

A **coefficient** is a learned weight. In this linear model, a positive
coefficient pushes a paragraph toward the second class, `1900 or later`. A
negative coefficient pushes it toward the first class, `before 1900`. The size of
the coefficient shows the fitted weight after the recipe has created tf-idf
columns.

Ranking coefficients by raw magnitude does not prove exactly what the model
relied on. The features are correlated, and ridge regression spreads weight
across related words. A different penalty could change the ranking. Read the
list as terms this fitted model weighted heavily, not as a stable dictionary of
historical meaning.

There is a second gap worth naming before any table appears. The model examined
below is fitted on every paragraph in the corpus, because an explanation is
usually wanted for the model a team would actually ship, and that model is
trained on everything available. It therefore has no score. The 0.8457 reported
in the evaluation lesson belongs to models fitted on three quarters of the
speeches and tested on the rest. Those are different fits. Nothing on this page
measures how well the model explained here performs, and the coefficients would
look broadly similar whether it performed well or not.

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tibble)
  library(stringr)
  library(recipes)
  library(textrecipes)
  library(parsnip)
  library(workflows)
  library(broom)
  library(glmnet)
})

source("R/inaugural-corpus.R")
paragraphs <- inaugural_paragraphs()

make_text_workflow <- function(formula, training_data) {
  recipe(formula, data = training_data) |>
    step_tokenize(paragraph) |>
    step_stopwords(paragraph) |>
    step_tokenfilter(paragraph, max_tokens = 500) |>
    step_tfidf(paragraph) |>
    workflow() |>
    add_model(
      logistic_reg(penalty = 0.01, mixture = 0) |>
        set_engine("glmnet")
    )
}

set.seed(4500)
era_fitted <- make_text_workflow(era ~ paragraph, paragraphs) |>
  fit(data = paragraphs)

era_coefficients <- tidy(extract_fit_parsnip(era_fitted)) |>
  mutate(
    word = str_remove(term, "^tfidf_paragraph_"),
    direction = if_else(estimate >= 0, "1900 or later", "before 1900")
  )

check_words <- c("attention", "object", "effect", "placed", "parts", "opinion")
checked_coefficients <- era_coefficients |>
  filter(word %in% check_words) |>
  arrange(word)

knitr::kable(
  checked_coefficients |>
    select(word, estimate, direction) |>
    mutate(estimate = round(estimate, 2)),
  col.names = c("Term", "Coefficient", "Pushes toward"),
  caption = "Selected large era-model coefficients",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(paragraphs), 1377L),
  identical(n_distinct(paragraphs$president), 40L),
  identical(nrow(era_coefficients), 501L),
  identical(sort(checked_coefficients$word), sort(check_words)),
  all(abs(checked_coefficients$estimate) > 7),
  identical(
    checked_coefficients$direction,
    rep("before 1900", length(check_words))
  )
)
```

The fitted era model has `r nrow(era_coefficients)` terms, including the
intercept. The listed words push toward `before 1900` in this fit. They are
formal-register signals in this corpus, not summaries of what any period was
about.

## Read both directions

The largest positive coefficients push toward `1900 or later`; the largest
negative coefficients push toward `before 1900`. This table is a map of the
model's fitted weights, not a dictionary of what the words mean.

```{r}
era_extremes <- bind_rows(
  era_coefficients |>
    filter(term != "(Intercept)") |>
    slice_max(estimate, n = 8) |>
    arrange(desc(estimate)),
  era_coefficients |>
    filter(term != "(Intercept)") |>
    slice_min(estimate, n = 8) |>
    arrange(estimate)
) |>
  mutate(estimate = round(estimate, 2)) |>
  select(word, direction, estimate)

sensitive_terms <- c("slavery", "slave", "race", "war")

knitr::kable(
  era_extremes,
  col.names = c("Term", "Pushes toward", "Coefficient"),
  caption = "Strongest positive and negative coefficients in the era model",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(era_extremes), 16L),
  max(era_extremes$estimate) > 8,
  min(era_extremes$estimate) < -15,
  all(era_extremes$direction %in% levels(paragraphs$era)),
  !any(str_to_lower(era_extremes$word) %in% sensitive_terms)
)
```

For a paragraph with the same other terms, a larger tf-idf value on a positive
term increases the model's pull toward `1900 or later`. A larger value on a
negative term increases the pull toward `before 1900`. The model is linear, so it
adds these pushes together.

## Explain the weak model too

The evaluation study found that the party model did not beat the deployable
largest-class rule at the paragraph or speech split, and its balanced accuracy
sat near chance. Still, the same extraction code produces coefficients. That is
the trap: explanation output can be neat even when the model is not useful.

```{r}
study <- read_csv(
  "data/inaugural/split-study.csv",
  na = c("", "NA"),
  col_types = cols(
    task = col_character(),
    split_scheme = col_character(),
    replicate = col_integer(),
    accuracy = col_double(),
    bal_accuracy = col_double(),
    train_majority_accuracy = col_double(),
    test_majority_rate = col_double(),
    length_rule_accuracy = col_double(),
    test_rows = col_integer(),
    test_speeches = col_integer()
  )
)

party_rows <- paragraphs |>
  filter(party %in% c("Democratic", "Republican")) |>
  mutate(party = factor(party))

set.seed(4503)
party_fitted <- make_text_workflow(party ~ paragraph, party_rows) |>
  fit(data = party_rows)

party_coefficients <- tidy(extract_fit_parsnip(party_fitted)) |>
  mutate(
    word = str_remove(term, "^tfidf_paragraph_"),
    direction = if_else(estimate >= 0, levels(party_rows$party)[2], levels(party_rows$party)[1])
  )

party_extremes <- party_coefficients |>
  filter(term != "(Intercept)") |>
  slice_max(abs(estimate), n = 10) |>
  arrange(desc(abs(estimate))) |>
  mutate(
    term = paste("term", row_number()),
    estimate = round(estimate, 2)
  ) |>
  select(term, direction, estimate)

party_summary <- study |>
  filter(task == "party") |>
  summarise(
    mean_accuracy = mean(accuracy),
    mean_bal_accuracy = mean(bal_accuracy),
    mean_train_majority = mean(train_majority_accuracy),
    .by = split_scheme
  )

party_paragraph <- party_summary |>
  filter(split_scheme == "paragraph")
party_speech <- party_summary |>
  filter(split_scheme == "speech")
party_president <- party_summary |>
  filter(split_scheme == "president")

knitr::kable(
  party_extremes,
  col.names = c("Masked term", "Pushes toward", "Coefficient"),
  caption = "Large coefficients from the party model that did not pass evaluation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(party_rows), 1197L),
  identical(n_distinct(party_rows$president), 31L),
  identical(nrow(party_coefficients), 501L),
  identical(nrow(party_extremes), 10L),
  identical(party_extremes$term, paste("term", 1:10)),
  !any(party_extremes$term %in% party_coefficients$word),
  identical(round(party_paragraph$mean_accuracy, 4), 0.6060),
  identical(round(party_paragraph$mean_train_majority, 4), 0.6067),
  identical(round(party_speech$mean_accuracy, 4), 0.5514),
  identical(round(party_speech$mean_train_majority, 4), 0.6563),
  identical(round(party_president$mean_accuracy, 4), 0.4997),
  identical(round(party_president$mean_bal_accuracy, 4), 0.5090),
  max(abs(party_coefficients$estimate)) > 13
)
```

The party terms are withheld deliberately. The model scores
`r round(party_speech$mean_accuracy, 4)` at the speech split against a deployable
baseline of `r round(party_speech$mean_train_majority, 4)`, and its person-split
balanced accuracy is `r round(party_president$mean_bal_accuracy, 4)`. Showing the
words would give a screenshottable political story from a model that did not earn
one. Masking costs nothing because the teaching point is the table's shape: the
output looks convincing regardless of whether the model works.

## What to remember

- A coefficient is a learned push toward one class or another.
- Coefficients show fitted weights after preprocessing, not proof of causation.
- A coefficient is not proof that the model is right.
- The era explanation matters because the era model beat cheap baselines.
- The party explanation is a warning because this model did not learn party.

If the score fails, the explanation should make you more cautious, not more
comfortable.

## Sources

- [broom documentation](https://broom.tidymodels.org/)
- [glmnet vignette](https://glmnet.stanford.edu/articles/glmnet.html)
- [tidymodels model fitting](https://www.tidymodels.org/start/models/)
- [textrecipes tf-idf step](https://textrecipes.tidymodels.org/reference/step_tfidf.html)
