---
title: "Deploying models"
subtitle: "Turn a fitted workflow into a callable service file"
author: "Alex Farach"
description: "Learn how vetiver records a fitted text model, writes an API file, and keeps evaluation details attached before deployment."
categories: [model development, deployment, vetiver]
---

A data team has a model that scores well enough for review. The next request is
practical: can another system send it a paragraph and get back a label?

That request is about deployment. Deployment does not make a model smarter. It
packages the fitted object, the expected input shape, and the code needed to call
it later.

The safest deployment record also carries the evaluation that justified the
model. Without that record, an API can make a weak answer look official.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- wrap a fitted tidymodels workflow with vetiver;
- pin the model to a local board;
- inspect the prototype input saved with the model;
- write the Plumber API file without starting a server; and
- name the evaluation details that must travel with a deployed model.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Fit the model to be packaged

A deployment needs the same fitted object that evaluation approved. Here the
object is the era classifier from a speech-held-out split with seed `4301` and
fit seed `4302`.

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tibble)
  library(stringr)
  library(rsample)
  library(recipes)
  library(textrecipes)
  library(parsnip)
  library(workflows)
  library(yardstick)
  library(vetiver)
  library(pins)
  library(glmnet)
})

source("R/inaugural-corpus.R")
paragraphs <- inaugural_paragraphs()

study <- read_csv(
  "data/inaugural/split-study.csv",
  na = c("", "NA"),
  col_types = cols(
    task = col_character(),
    split_scheme = col_character(),
    replicate = col_integer(),
    accuracy = col_double(),
    bal_accuracy = col_double(),
    train_majority_accuracy = col_double(),
    test_majority_rate = col_double(),
    length_rule_accuracy = col_double(),
    test_rows = col_integer(),
    test_speeches = col_integer()
  )
)

set.seed(4301)
era_split <- group_initial_split(paragraphs, group = speech_id, prop = 0.75)
era_training <- training(era_split)
era_testing <- testing(era_split)

era_recipe <- recipe(era ~ paragraph, data = era_training) |>
  step_tokenize(paragraph) |>
  step_stopwords(paragraph) |>
  step_tokenfilter(paragraph, max_tokens = 500) |>
  step_tfidf(paragraph)

era_model <- logistic_reg(penalty = 0.01, mixture = 0) |>
  set_engine("glmnet")

era_workflow <- workflow() |>
  add_recipe(era_recipe) |>
  add_model(era_model)

set.seed(4302)
era_fitted <- fit(era_workflow, data = era_training)

era_predictions <- predict(era_fitted, era_testing) |>
  bind_cols(era_testing |> select(era))

live_accuracy <- accuracy(
  era_predictions,
  truth = era,
  estimate = .pred_class
) |>
  pull(.estimate)

training_majority <- era_training |>
  count(era, sort = TRUE) |>
  slice(1) |>
  pull(era)
live_majority_accuracy <- mean(era_testing$era == training_majority)

speech_summary <- study |>
  filter(task == "era", split_scheme == "speech") |>
  summarise(
    model_accuracy = mean(accuracy),
    train_majority_accuracy = mean(train_majority_accuracy),
    length_rule_accuracy = mean(length_rule_accuracy),
    .groups = "drop"
  )

evaluation_date <- as.Date("2026-08-29")
deployment_record <- tibble(
  record = c(
    "data source",
    "split scheme",
    "training paragraphs",
    "test paragraphs",
    "live test accuracy",
    "live train-majority baseline",
    "speech-split mean accuracy",
    "speech-split mean train-majority baseline",
    "speech-split mean length baseline",
    "evaluation date"
  ),
  value = c(
    "inaugural paragraphs",
    "speech_id group split",
    as.character(nrow(era_training)),
    as.character(nrow(era_testing)),
    as.character(round(live_accuracy, 4)),
    as.character(round(live_majority_accuracy, 4)),
    as.character(round(speech_summary$model_accuracy, 4)),
    as.character(round(speech_summary$train_majority_accuracy, 4)),
    as.character(round(speech_summary$length_rule_accuracy, 4)),
    as.character(evaluation_date)
  )
)

stopifnot(
  identical(nrow(era_training), 1042L),
  identical(nrow(era_testing), 335L),
  identical(round(live_accuracy, 4), 0.8209),
  identical(round(live_majority_accuracy, 4), 0.4179),
  identical(round(speech_summary$model_accuracy, 4), 0.8457),
  identical(round(speech_summary$train_majority_accuracy, 4), 0.6282),
  identical(round(speech_summary$length_rule_accuracy, 4), 0.6827),
  identical(evaluation_date, as.Date("2026-08-29")),
  identical(nrow(deployment_record), 10L)
)

knitr::kable(
  deployment_record,
  col.names = c("Record item", "Value"),
  caption = "Evaluation details kept beside the deployable model",
  row.names = FALSE
)
```

Deployment freezes a model at one point in its life. The record above says which
data split and score go with the object, and it records the baseline that makes
the score interpretable. The repeated speech-split study matters because one live
split can have an unusual class mix.

## Pin the fitted workflow

`vetiver_model()` wraps the fitted workflow with metadata. A **prototype** is the
empty input table the model expects. The local **pins board** is a storage place
for versioned objects; this lesson writes to a project-local scratch board and
removes it after reading the generated files.

```{r}
make_api_artifacts <- function(fitted_model) {
  board_dir <- file.path("9_model_development", ".lesson-46-board")
  api_file <- file.path("9_model_development", ".lesson-46-api.R")
  unlink(c(board_dir, api_file), recursive = TRUE, force = TRUE)
  dir.create(board_dir, recursive = TRUE, showWarnings = FALSE)
  on.exit(unlink(c(board_dir, api_file), recursive = TRUE, force = TRUE), add = TRUE)

  vetiver_object <- suppressMessages(
    vetiver_model(fitted_model, "inaugural-era-workflow")
  )
  board <- board_folder(board_dir, versioned = TRUE)
  capture.output(
    suppressMessages(vetiver_pin_write(board, vetiver_object))
  )
  versions <- pin_versions(board, "inaugural-era-workflow")
  pinned_model <- vetiver_pin_read(
    board,
    "inaugural-era-workflow",
    version = versions$version[[1]]
  )

  suppressMessages(
    vetiver_write_plumber(
      board,
      "inaugural-era-workflow",
      version = versions$version[[1]],
      file = api_file
    )
  )
  api_lines <- read_lines(api_file)
  board_path <- str_match(api_lines[str_detect(api_lines, "board_folder")], 'path = "([^"]+)"')[, 2]
  if (length(board_path) == 0L || is.na(board_path[[1]])) {
    board_path <- ""
  }
  display_lines <- api_lines
  if (nzchar(board_path)) {
    display_lines <- display_lines |>
      str_replace_all(stringr::fixed(board_path), "<local board path>")
  }
  display_lines <- display_lines |>
    str_replace_all(stringr::fixed(versions$version[[1]]), "<pinned version>")

  list(
    vetiver_object = vetiver_object,
    versions = versions,
    pinned_model = pinned_model,
    api_lines = api_lines,
    display_lines = display_lines
  )
}

api_artifacts <- make_api_artifacts(era_fitted)
vetiver_object <- api_artifacts$vetiver_object
versions <- api_artifacts$versions
pinned_model <- api_artifacts$pinned_model
prototype_table <- tibble(
  columns = paste(names(vetiver_object$prototype), collapse = ", "),
  rows = nrow(vetiver_object$prototype)
)

stopifnot(
  identical(vetiver_object$model_name, "inaugural-era-workflow"),
  identical(names(vetiver_object$prototype), "paragraph"),
  identical(nrow(vetiver_object$prototype), 0L),
  identical(prototype_table$columns, "paragraph"),
  identical(prototype_table$rows, 0L),
  identical(nrow(versions), 1L),
  identical(pinned_model$model_name, vetiver_object$model_name)
)

knitr::kable(
  prototype_table,
  col.names = c("Prototype columns", "Prototype rows"),
  caption = "Input shape stored with the vetiver model",
  row.names = FALSE
)
```

The version lets a later system ask for this exact object. The prototype says the
API expects a table with a `paragraph` column. The pinned board is the place the
API reads from when it starts.

## Write the API file without starting it

`vetiver_write_plumber()` writes an R file for a Plumber API. This lesson only
writes and reads that file. It does not start a server and it makes no network
request.

```{r}
#| output: asis

api_lines <- api_artifacts$api_lines
display_lines <- api_artifacts$display_lines

stopifnot(
  length(api_lines) >= 20L,
  any(str_detect(api_lines, stringr::fixed("vetiver_api(v)"))),
  any(str_detect(api_lines, stringr::fixed("vetiver_pin_read"))),
  any(str_detect(api_lines, stringr::fixed("inaugural-era-workflow"))),
  any(str_detect(api_lines, stringr::fixed("@plumber"))),
  any(str_detect(api_lines, stringr::fixed("library(plumber)"))),
  !any(str_detect(display_lines, stringr::fixed(versions$version[[1]]))),
  any(str_detect(display_lines, stringr::fixed("<pinned version>")))
)

cat("```r\n", paste(display_lines, collapse = "\n"), "\n```\n", sep = "")
```

The file shown above was generated by vetiver. The local board path and version
string are masked in the display so the rendered lesson stays stable, but the
assertions read the generated file before masking it.

A deployment record must carry the data source, split scheme, score, baseline,
and date. For this model, the live record is: inaugural paragraphs,
speech-held-out split, accuracy `r round(live_accuracy, 4)`, same-split
training-majority baseline `r round(live_majority_accuracy, 4)`, evaluation date
`r as.character(evaluation_date)`, and fixed seeds `4301` and `4302`.

The party model with this recipe and penalty did not pass evaluation. It should
not be deployed by wrapping it in an API. Packaging makes access easier; it does
not repair a failed measurement.

## What to remember

- Deployment packages a fitted object so another system can call it.
- A vetiver prototype records the input columns the model expects.
- A pins board stores model versions.
- Do not deploy a model that did not pass evaluation.
- Keep the data, split, score, baseline, and date with the deployed object.

An API turns a model into an answer people can request. It should also make the
limits hard to lose.

## Sources

- [vetiver documentation](https://vetiver.co/)
- [pins documentation](https://pins.rstudio.com/)
- [Plumber documentation](https://www.rplumber.io/)
- [Model cards paper](https://doi.org/10.1145/3287560.3287596)
