---
title: "Monitoring models"
subtitle: "Watch a deployed text model after the first score"
description: "Learn how to check model performance across time slices and why changing labels can make an old score misleading."
categories: [model development, monitoring, model drift]
---

A dashboard turns green on launch day. The model passed its test, the API is
ready, and a queue of paragraphs can finally be scored.

Three months later, the queue is different. The words are familiar, the file
format still works, and the old test score is still printed in the deployment
record. None of that proves the model is doing the same job.

Monitoring asks what changed after deployment. It can watch the text arriving,
and when checked labels arrive, it can watch the score.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- score a fitted text model across time slices;
- read held-out slice scores without treating them as a new train/test study;
- check input drift when labels are late or missing;
- define concept drift in plain language;
- explain why a fixed party text model needs fresh evidence; and
- name what monitoring can and cannot see.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Refit the era model

For a monitoring lesson, we replay the corpus as if labeled paragraphs arrived
from different periods. The model is the same penalized logistic regression with
tf-idf features. The split seed is `4301`, and the fit seed is `4302`.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tibble)
  library(stringr)
  library(tidyr)
  library(stopwords)
  library(ggplot2)
  library(rsample)
  library(recipes)
  library(textrecipes)
  library(parsnip)
  library(workflows)
  library(yardstick)
  library(broom)
  library(glmnet)
})

source("R/inaugural-corpus.R")
paragraphs <- inaugural_paragraphs()

set.seed(4301)
era_split <- group_initial_split(paragraphs, group = speech_id, prop = 0.75)
era_training <- training(era_split)
era_testing <- testing(era_split)

era_recipe <- recipe(era ~ paragraph, data = era_training) |>
  step_tokenize(paragraph) |>
  step_stopwords(paragraph) |>
  step_tokenfilter(paragraph, max_tokens = 500) |>
  step_tfidf(paragraph)

era_model <- logistic_reg(penalty = 0.01, mixture = 0) |>
  set_engine("glmnet")

era_workflow <- workflow() |>
  add_recipe(era_recipe) |>
  add_model(era_model)

set.seed(4302)
era_fitted <- fit(era_workflow, data = era_training)

held_out_predictions <- predict(era_fitted, era_testing) |>
  bind_cols(era_testing |> select(era))

held_out_accuracy <- accuracy(
  held_out_predictions,
  truth = era,
  estimate = .pred_class
) |>
  pull(.estimate)

era_length_summary <- paragraphs |>
  group_by(era) |>
  summarise(
    paragraphs = n(),
    median_words = median(paragraph_words),
    .groups = "drop"
  )

knitr::kable(
  tibble(
    item = c("paragraphs", "held-out paragraphs", "held-out accuracy"),
    value = c(nrow(paragraphs), nrow(era_testing), round(held_out_accuracy, 4))
  ),
  col.names = c("Item", "Value"),
  caption = "Model and corpus size for the monitoring replay",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(paragraphs), 1377L),
  identical(n_distinct(paragraphs$president), 40L),
  identical(nrow(era_testing), 335L),
  identical(round(held_out_accuracy, 4), 0.8209),
  identical(as.integer(era_length_summary$paragraphs), c(485L, 892L)),
  identical(as.numeric(era_length_summary$median_words), c(112, 54))
)
```

The held-out score is still the evaluation score. Any slice score below uses
only those `r nrow(era_testing)` held-out paragraphs, never rows from the
training side.

## Score held-out time slices

A **monitoring slice** is a subset watched on its own, such as a month, region,
source, or time period. Here the slices are half-centuries. This is already a
trap, because the label `era` is derived from the year. A time slice contains
almost only one label, so its score is one-class recall rather than overall
performance.

```{r}
period_breaks <- c(1780, 1849, 1899, 1949, 1999, 2030)
period_levels <- c(
  "1789-1849",
  "1850-1899",
  "1900-1949",
  "1950-1999",
  "2000-2025"
)

held_out_by_period <- held_out_predictions |>
  bind_cols(era_testing |> select(paragraph_id, year, paragraph)) |>
  mutate(
    period = cut(
      year,
      breaks = period_breaks,
      labels = period_levels,
      right = TRUE
    )
  )

label_counts <- held_out_by_period |>
  count(period, era, name = "labels") |>
  complete(
    period = factor(period_levels, levels = period_levels),
    era = levels(paragraphs$era),
    fill = list(labels = 0L)
  )

period_accuracy <- label_counts |>
  group_by(period) |>
  summarise(
    held_out_paragraphs = sum(labels),
    before_1900 = sum(labels[era == "before 1900"]),
    later_1900 = sum(labels[era == "1900 or later"]),
    .groups = "drop"
  ) |>
  left_join(
    held_out_by_period |>
      group_by(period) |>
      summarise(
        correct = sum(.pred_class == era),
        accuracy = mean(.pred_class == era),
        .groups = "drop"
      ),
    by = join_by(period)
  ) |>
  mutate(
    correct = if_else(is.na(correct), 0L, as.integer(correct)),
    score_kind = case_when(
      held_out_paragraphs == 0L ~ "no held-out rows",
      before_1900 == 0L | later_1900 == 0L ~ "one-class recall",
      TRUE ~ "two-class accuracy"
    )
  )

lowest_slice <- period_accuracy |>
  filter(held_out_paragraphs > 0L) |>
  slice_min(accuracy, n = 1, with_ties = FALSE)

knitr::kable(
  period_accuracy |>
    select(
      period,
      held_out_paragraphs,
      before_1900,
      later_1900,
      correct,
      score_kind,
      accuracy
    ) |>
    mutate(accuracy = round(accuracy, 3)),
  col.names = c(
    "Period", "Held-out paragraphs", "Before 1900 labels",
    "1900 or later labels", "Correct", "Score type", "Accuracy"
  ),
  caption = "Held-out era scores by period with class composition",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(period_accuracy), 5L),
  identical(period_accuracy$held_out_paragraphs, c(67L, 128L, 23L, 117L, 0L)),
  identical(period_accuracy$before_1900, c(67L, 128L, 0L, 0L, 0L)),
  identical(period_accuracy$later_1900, c(0L, 0L, 23L, 117L, 0L)),
  identical(period_accuracy$correct, c(58L, 87L, 17L, 113L, 0L)),
  identical(
    round(period_accuracy$accuracy[!is.na(period_accuracy$accuracy)], 3),
    c(0.866, 0.680, 0.739, 0.966)
  ),
  identical(sum(period_accuracy$held_out_paragraphs), 335L),
  identical(as.character(lowest_slice$period), "1850-1899"),
  identical(lowest_slice$held_out_paragraphs, 128L)
)
```

The lowest held-out slice is
`r as.character(lowest_slice$period)`, with `r lowest_slice$held_out_paragraphs`
paragraphs and accuracy `r round(lowest_slice$accuracy, 3)`. It is not the slice
around 1900. The deeper problem is the same in every nonempty row: each slice has
one class. The table is a warning about circular monitoring, not a clean accuracy
report.

The `2000-2025` row is empty because all speeches from that period landed in
training under seed `4301`. Grouped splitting keeps whole speeches together, so
a period can have no held-out examples even when the full corpus contains them.

## Watch input drift without waiting for labels

Labels often arrive late, and sometimes they never arrive. One thing a team can
watch without labels is whether the incoming text is moving away from the text
the model saw during fitting. The code below uses the 500 model features as the
training vocabulary, then asks what share of non-stopword tokens in each period
falls outside that vocabulary. It also reports median paragraph length.

```{r}
training_vocabulary <- tidy(extract_fit_parsnip(era_fitted)) |>
  filter(term != "(Intercept)") |>
  transmute(token = str_remove(term, "^tfidf_paragraph_")) |>
  pull(token)

period_paragraphs <- paragraphs |>
  mutate(
    period = cut(
      year,
      breaks = period_breaks,
      labels = period_levels,
      right = TRUE
    )
  )

period_lengths <- period_paragraphs |>
  group_by(period) |>
  summarise(
    paragraphs = n(),
    median_words = median(paragraph_words),
    .groups = "drop"
  )

model_stop_words <- tibble(
  word = stopwords("en", source = "snowball")
)

drift_table <- period_paragraphs |>
  select(paragraph_id, period, paragraph) |>
  # Tokenise with an explicit pattern rather than `unnest_tokens()`. The default
  # word tokeniser uses ICU word boundaries, and ICU versions differ between
  # machines, so the same text can yield slightly different token counts on a
  # Linux runner than on a laptop. A stated pattern makes the count reproducible.
  mutate(token = str_extract_all(tolower(paragraph), "[a-z']+")) |>
  select(-paragraph) |>
  unnest_longer(token) |>
  anti_join(model_stop_words, by = join_by(token == word)) |>
  group_by(period) |>
  summarise(
    tokens = n(),
    absent_tokens = sum(!(token %in% training_vocabulary)),
    absent_share = absent_tokens / tokens,
    .groups = "drop"
  ) |>
  left_join(period_lengths, by = join_by(period)) |>
  select(period, paragraphs, median_words, tokens, absent_tokens, absent_share)

knitr::kable(
  drift_table |>
    mutate(absent_share = round(absent_share, 3)),
  col.names = c(
    "Period", "Paragraphs", "Median words",
    "Non-stopword tokens", "Tokens outside 500 features",
    "Share outside 500 features"
  ),
  caption = "Input drift checks by period using the training vocabulary",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(length(training_vocabulary), 500L),
  identical(nrow(drift_table), 5L),
  identical(drift_table$paragraphs, c(220L, 265L, 317L, 371L, 204L)),
  identical(drift_table$median_words, c(149.5, 92, 67, 53, 43)),
  identical(drift_table$tokens, c(19192L, 14562L, 14388L, 10650L, 5579L)),
  identical(drift_table$absent_tokens, c(9900L, 7390L, 7014L, 4972L, 2800L)),
  identical(round(drift_table$absent_share, 3), c(0.516, 0.507, 0.487, 0.467, 0.502)),
  max(drift_table$median_words) - min(drift_table$median_words) > 100
)
```

The stop-word list now matches `step_stopwords()` in the fitted recipe: the
Snowball English list supplied by the stopwords package.

The vocabulary drift measure is close to flat partly by construction. It counts
tokens outside a fixed 500-token vocabulary, and about half of the non-stopword
tokens fall outside that short list. The share ranges from
`r round(min(drift_table$absent_share), 3)` to
`r round(max(drift_table$absent_share), 3)` across
`r max(paragraphs$year) - min(paragraphs$year)` years.

Median paragraph length is the clearer signal. It falls from
`r max(drift_table$median_words)` words in the earliest period to
`r min(drift_table$median_words)` in the latest. The same variable carried part
of the era result in the evaluation lesson: before-1900 paragraphs have median
length `r era_length_summary$median_words[era_length_summary$era == "before 1900"]`,
while later paragraphs have median length
`r era_length_summary$median_words[era_length_summary$era == "1900 or later"]`.
A team watching only the vocabulary check would have missed that shift.

Neither check requires labels, which is why they can run while a team waits for
review samples. Neither one tells you the model is still correct. They tell you
whether the incoming text still looks like the text the model was fitted to, and
a model can be badly wrong on inputs that look perfectly familiar.

```{r}
#| label: fig-input-drift
#| fig-cap: "Input drift checks by period."
#| fig-alt: "Two small line charts by period. One shows the share of non-stopword tokens outside the training features. The other shows median paragraph length falling over time."

drift_plot <- drift_table |>
  transmute(
    period,
    `share outside training features` = absent_share,
    `median paragraph words` = median_words
  ) |>
  pivot_longer(
    cols = -period,
    names_to = "measure",
    values_to = "value"
  )

ggplot(drift_plot, aes(x = period, y = value, group = measure)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  facet_wrap(~measure, scales = "free_y", ncol = 1) +
  labs(
    x = "Period",
    y = NULL
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  identical(nrow(drift_plot), 10L),
  identical(
    sort(unique(drift_plot$measure)),
    c("median paragraph words", "share outside training features")
  ),
  max(drift_table$median_words) - min(drift_table$median_words) > 100
)
```

The length pattern is much stronger than the absent-token pattern here. In a live
system, either could trigger review. Neither says the model is wrong by itself.
It says the input no longer looks the same in a way a person should inspect.

## Watch labels as well as words

**Concept drift** means the relationship between inputs and labels changes over
time. Text changes when writers choose new words. Labels can change too.

The party column in this corpus spans the full address range. A label such as
`Democratic` appears across many decades. That span is not evidence that one
fixed paragraph model can predict party across the collection. The evaluation
result for this recipe and penalty is a measurement warning, not a political
finding.

```{r}
party_range <- paragraphs |>
  summarise(
    first_year = min(year),
    last_year = max(year),
    years_spanned = last_year - first_year,
    parties = n_distinct(party),
    .groups = "drop"
  )

democratic_range <- paragraphs |>
  filter(party == "Democratic") |>
  summarise(
    first_year = min(year),
    last_year = max(year),
    paragraphs = n(),
    speeches = n_distinct(speech_id),
    .groups = "drop"
  )

party_rows <- paragraphs |>
  filter(party %in% c("Democratic", "Republican"))

knitr::kable(
  bind_rows(
    party_range |>
      transmute(label = "all party labels", first_year, last_year, paragraphs = nrow(paragraphs)),
    democratic_range |>
      transmute(label = "Democratic", first_year, last_year, paragraphs)
  ),
  col.names = c("Label scope", "First year", "Last year", "Paragraphs"),
  caption = "Long time spans in the party labels",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(party_range$first_year, 1789L),
  identical(party_range$last_year, 2025L),
  identical(party_range$years_spanned, 236L),
  identical(party_range$parties, 6L),
  identical(democratic_range$first_year, 1829L),
  identical(democratic_range$last_year, 2021L),
  identical(democratic_range$paragraphs, 471L),
  identical(democratic_range$speeches, 22L),
  identical(nrow(party_rows), 1197L),
  identical(n_distinct(party_rows$president), 31L)
)
```

Monitoring can see the input distribution moving. With checked labels, it can
see a score fall. It cannot see confident errors on questions nobody checks.
That blind spot is why monitoring needs sampled review, not only dashboards.

## What to remember

- Monitoring starts after a model is packaged or put into routine use.
- Slice scores can show where performance changes.
- A slice score is not the same as a new held-out evaluation.
- Concept drift can affect words, labels, or the relationship between them.
- Monitoring cannot find errors in rows that no one checks.

A green dashboard is a prompt to keep looking, not a permanent certificate.

## Sources

- [yardstick metrics reference](https://yardstick.tidymodels.org/)
- [NIST AI Risk Management Framework](https://www.nist.gov/itl/ai-risk-management-framework)
- [quanteda inaugural corpus documentation](https://quanteda.io/reference/data_corpus_inaugural.html)
- [Evidently guide to data drift](https://www.evidentlyai.com/ml-system-design/data-drift)
