---
title: "Finding sentence boundaries"
subtitle: "Compare three ways to split text into sentences"
author: "Alex Farach"
description: "Learn why sentence segmentation can fail on abbreviations, decimals, initials, titles, and headings."
categories: [sentences and paragraphs, sentence segmentation, workforce research]
---

One copied page mixes a flyer line, a pay line, and a job-board note before the
coordinator tries to count its sentences. The count has to line up with the 28
labelled Riverton rows.

The count looks simple until the page reaches `Dr.`, `a.m.`, `$18.50`, and a
heading with no period. A sentence boundary is the place where one sentence ends
and the next begins. The team has to choose a rule before it can trust the
count.

::: {.callout-note}
The Riverton page and CSV rows are fictional teaching data.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain why sentence segmentation is harder than splitting on periods;
- compare an author-marked sentence count with three splitters;
- inspect the fragments each splitter creates;
- use spaCy sentence identifiers from `spacy_parse()`; and
- explain why the Riverton CSV already contains an upstream boundary choice.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build a passage with traps

Here, readr brings in the Riverton CSV, dplyr and tibble build tables, purrr and
stringr run the small splitters, tokenizers supplies a package splitter, and
spacyr calls a trained parser. A **splitter** is a rule or model that marks
boundaries in text.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(spacyr)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

passage <- str_c(
  "RIVERTON SKILLS OPEN HOUSE\n",
  "Dr. Vale said shifts start at 6 a.m.\n",
  "Pay is $18.50 per hour.\n",
  "J. Rivera can apply by Friday\n",
  "Bring a photo ID."
)

expected_sentences <- c(
  "RIVERTON SKILLS OPEN HOUSE",
  "Dr. Vale said shifts start at 6 a.m.",
  "Pay is $18.50 per hour.",
  "J. Rivera can apply by Friday",
  "Bring a photo ID."
)

knitr::kable(
  tibble(
    expected_number = seq_along(expected_sentences),
    sentence = expected_sentences
  ),
  col.names = c("Expected sentence", "Author-marked text"),
  caption = "The five expected sentence boundaries",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(c(
  "readr", "dplyr", "tibble", "purrr", "stringr", "tokenizers", "spacyr"
) %in% loadedNamespaces()))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(length(expected_sentences), 5L),
  any(str_detect(sentences$text, fixed("Apply by October 15")))
)
```

The expected-answer list has five sentences. Because that list and the examples
share an author, agreement with it is a lesson check rather than an accuracy
estimate. The first sentence is a flyer-style heading with no punctuation. The
next three contain a title, an abbreviation, a decimal, and an initial.

## Compare three splitters

The naive rule below splits wherever it sees a full stop. `tokenizers` uses a
sentence-splitting rule from an R package. spaCy uses a trained English pipeline;
`spacy_parse()` returns one row per token, and its `sentence_id` column records
the boundary spaCy chose.

```{r}
split_on_full_stop <- function(text) {
  text |>
    str_split(fixed(".")) |>
    pluck(1) |>
    str_trim() |>
    discard(\(piece) identical(piece, ""))
}

join_spacy_tokens <- function(tokens) {
  str_c(tokens, collapse = " ") |>
    str_squish() |>
    str_replace_all(fixed(" ."), ".") |>
    str_replace_all(fixed(" ,"), ",") |>
    str_replace_all(fixed("$ "), "$")
}

naive_pieces <- split_on_full_stop(passage)
tokenizer_pieces <- tokenize_sentences(passage)[[1]]

source("R/use-spacy.R")
pipeline <- use_project_spacy()
spacy_metadata <- spacy_pipeline_version()
spacy_tokens <- spacy_parse(
  setNames(passage, "test_passage"),
  pos = TRUE,
  lemma = TRUE
)
spacy_finalize()

spacy_pieces <- spacy_tokens |>
  as_tibble() |>
  summarise(piece = join_spacy_tokens(token), .by = sentence_id) |>
  pull(piece)

split_results <- bind_rows(
  tibble(splitter = "split on full stops", piece = naive_pieces),
  tibble(splitter = "tokenizers::tokenize_sentences()", piece = tokenizer_pieces),
  tibble(splitter = "spaCy sentence_id", piece = spacy_pieces)
) |>
  mutate(
    splitter = factor(
      splitter,
      levels = c(
        "split on full stops",
        "tokenizers::tokenize_sentences()",
        "spaCy sentence_id"
      )
    )
  ) |>
  arrange(splitter) |>
  group_by(splitter) |>
  mutate(
    piece_number = row_number(),
    exact_expected_sentence = piece %in% expected_sentences
  ) |>
  ungroup()

splitter_scores <- split_results |>
  summarise(
    pieces_found = n(),
    exact_expected_pieces = sum(exact_expected_sentence),
    .by = splitter
  ) |>
  mutate(
    expected_pieces = length(expected_sentences),
    count_difference = pieces_found - expected_pieces,
    matches_expected_count = pieces_found == expected_pieces
  )

knitr::kable(
  splitter_scores,
  col.names = c(
    "Splitter",
    "Pieces found",
    "Exact pieces",
    "Expected pieces",
    "Difference",
    "Count matches"
  ),
  caption = "Splitter counts compared with the five expected sentences",
  row.names = FALSE
)

knitr::kable(
  split_results |>
    mutate(exact_expected_sentence = if_else(exact_expected_sentence, "yes", "no")),
  col.names = c("Splitter", "Piece", "Piece number", "Exact expected sentence"),
  caption = "Where each splitter puts the boundaries",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(spacy_metadata$version, "3.8.0"),
  identical(spacy_metadata$license, "MIT"),
  identical(spacy_metadata$spacy, "3.8.7"),
  identical(length(naive_pieces), 7L),
  identical(length(tokenizer_pieces), 5L),
  identical(length(spacy_pieces), 2L),
  identical(splitter_scores$pieces_found, c(7L, 5L, 2L)),
  identical(splitter_scores$exact_expected_pieces, c(0L, 1L, 0L)),
  identical(splitter_scores$count_difference, c(2L, 0L, -3L))
)
```

The naive rule finds seven pieces and none is one of the five expected
sentences. It cuts `Dr.`, `a.m.`, `$18.50`, and `J.` apart. The package splitter
gets the count right, but only one piece exactly matches an expected sentence. It
attaches `Dr.` to the heading and splits `J.` away from the name.

spaCy returns two pieces. It keeps `Dr. Vale` and `a.m.` together, then runs the
heading, pay line, and two closing lines into fewer sentences than the layout
implies. The pipeline segments from syntax and does not treat a line break as a
boundary, which is what the flyer format relies on.

The spaCy pieces are rebuilt from tokens, so spaces and line breaks do not
survive. Read the exact-match column for spaCy as a comparison against a
reconstruction, not against the copied page. Comparing piece counts is also a
weak test. Boundary positions matter, because a splitter can return the right
number of pieces and put the boundaries in the wrong places.

## Check the Riverton rows

The 28 Riverton rows are already one sentence each by construction. That is an
upstream boundary choice: someone made a CSV row out of each sentence or flyer
line before this lesson reads the file.

```{r}
sentence_piece_counts <- tokenize_sentences(sentences$text) |>
  map_int(length)

row_boundary_summary <- sentences |>
  mutate(sentence_pieces = sentence_piece_counts) |>
  summarise(
    rows = n(),
    tokenizer_pieces = sum(sentence_pieces),
    .by = document_id
  )

flyer_rows <- sentences |>
  filter(document_id == "F001") |>
  nrow()

job_rows <- sentences |>
  filter(document_id != "F001") |>
  nrow()

knitr::kable(
  row_boundary_summary,
  col.names = c("Document ID", "Rows", "Sentence pieces"),
  caption = "Sentence splitting agrees with the stored Riverton rows",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(sentence_piece_counts, rep(1L, 28L)),
  identical(sum(row_boundary_summary$rows), 28L),
  identical(sum(row_boundary_summary$tokenizer_pieces), 28L),
  identical(flyer_rows, 6L),
  identical(job_rows, 22L)
)
```

For these stored rows, the package splitter returns one sentence piece for every
row. That agreement does not prove the rows are the only possible boundaries. It
only shows that this file and this splitter line up on this small teaching set.

## What to remember

- Sentence segmentation marks where one sentence ends and another begins.
- A full stop can appear inside titles, initials, abbreviations, and decimals.
- In the test passage, the three splitters return 7, 5, and 2 pieces.
- A correct count can still hide wrong boundaries.
- Rebuilding model output from tokens can change the text being scored.
- The 28 Riverton rows already contain one upstream boundary choice.

For the copied page, five expected sentences are a human layout choice. For the
CSV, 28 one-row sentence units are an upstream data choice. Both counts depend on
the rule named beside them.

## Sources

- [tokenizers documentation for sentence tokenizers](https://docs.ropensci.org/tokenizers/reference/basic-tokenizers.html)
- [spaCy documentation on linguistic features](https://spacy.io/usage/linguistic-features)
- [spacyr package documentation](https://cran.r-project.org/package=spacyr)
