---
title: "Finding paragraph boundaries"
subtitle: "Keep layout units before text conversion removes them"
author: "Alex Farach"
description: "Learn how blank lines and HTML block elements define paragraphs in different file formats."
categories: [sentences and paragraphs, paragraph segmentation, workforce research]
---

Help articles and job-board HTML reach the coordinator as readable text, yet
their paragraph breaks live in different places. Comparing them requires keeping
those boundaries before cleanup erases them.

In a plain-text help article, a blank line can mark a new paragraph. HTML gives
the team block elements it can use as paragraph boundaries. The standard's own
definition of a paragraph is looser than a `<p>` element, so this is a working
rule for the lesson rather than the specification's answer.

::: {.callout-note}
The help articles and job board are local fictional fixtures.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define a paragraph as a layout unit;
- split plain text on blank lines;
- collect paragraph-like blocks from HTML with rvest;
- compare paragraph counts across file formats; and
- explain why block boundaries should be captured before tag removal.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Split plain text on blank lines

This section reads files with readr, repeats file work with purrr, splits text
with stringr, expands list columns with tidyr, shapes tables with dplyr and
tibble, and asks rvest for HTML. A **paragraph** is a block of text grouped by
layout. It may contain one sentence or several.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tidyr)
library(rvest)

stopifnot(all(c(
  "readr", "dplyr", "tibble", "purrr", "stringr", "tidyr", "rvest"
) %in% loadedNamespaces()))

help_files <- list.files(
  "data/help_articles",
  pattern = "[.]txt$",
  full.names = TRUE
) |>
  sort()

split_blank_lines <- function(text) {
  str_split(str_trim(text), "\\r?\\n\\s*\\r?\\n+")[[1]] |>
    str_trim() |>
    discard(\(piece) identical(piece, ""))
}

help_articles <- tibble(
  file = basename(help_files),
  text = map_chr(help_files, read_file)
) |>
  mutate(paragraph = map(text, split_blank_lines))

help_paragraphs <- help_articles |>
  select(file, paragraph) |>
  unnest_longer(paragraph) |>
  group_by(file) |>
  mutate(paragraph_number = row_number()) |>
  ungroup()

help_counts <- help_paragraphs |>
  count(file, name = "paragraphs")

total_help_paragraphs <- sum(help_counts$paragraphs)
help_blank_line_counts <- str_count(help_articles$text, "\\r?\\n\\s*\\r?\\n+")

stopifnot(
  identical(length(help_files), 3L),
  identical(help_counts$paragraphs, c(1L, 1L, 1L)),
  identical(total_help_paragraphs, 3L),
  identical(help_blank_line_counts, c(0L, 0L, 0L)),
  str_detect(help_paragraphs$paragraph[1], fixed("password reset link"))
)

knitr::kable(
  help_counts,
  col.names = c("Help article", "Paragraphs from blank-line rule"),
  caption = "Paragraph counts in the plain-text help articles",
  row.names = FALSE
)

knitr::kable(
  help_paragraphs,
  col.names = c("Help article", "Paragraph", "Paragraph number"),
  caption = "Plain-text paragraphs after splitting on blank lines",
  row.names = FALSE
)
```

None of these help files contains a blank line, so the rule has nothing to split
on and returns each file whole. That is a fair result but an empty demonstration.
The HTML section below is where the rule has to make a choice.

## Take blocks from HTML

HTML records layout with elements. The code below reads the job-board page and
keeps headings, paragraphs, and list items. The selector `h1, h2, p, li` names
the block types the team wants to treat as paragraph-like units.

```{r}
job_page <- read_html("data/workforce/job-board.html")
block_nodes <- html_elements(job_page, "h1, h2, p, li")

html_blocks <- tibble(
  tag = html_name(block_nodes),
  text = html_text2(block_nodes)
) |>
  mutate(block_number = row_number())

html_tag_counts <- html_blocks |>
  count(tag, name = "blocks") |>
  arrange(tag)

expected_tag_counts <- tibble(
  tag = c("h1", "h2", "li", "p"),
  blocks = c(1L, 6L, 22L, 12L)
)

stopifnot(
  identical(nrow(html_blocks), 41L),
  identical(html_tag_counts, expected_tag_counts),
  identical(html_blocks$text[1], "Riverton workforce opportunities"),
  identical(html_blocks$text[5], "Paid 12-week training is provided."),
  identical(sum(html_blocks$tag == "li"), 22L)
)

knitr::kable(
  html_tag_counts,
  col.names = c("HTML tag", "Blocks captured"),
  caption = "Paragraph-like blocks captured from the job-board HTML",
  row.names = FALSE
)

knitr::kable(
  html_blocks |>
    slice_head(n = 8),
  col.names = c("HTML tag", "Text", "Block number"),
  caption = "The first eight block elements from the job board",
  row.names = FALSE
)
```

The HTML rule finds 41 blocks: 1 page heading, 6 job headings, 12 paragraph
elements, and 22 list items. The job-detail sentences in the Riverton CSV came
from those list items.

## Show what tag removal loses

A common conversion step removes tags and keeps only the visible words. That can
be useful for word counts, but it is too late for paragraph segmentation if the
block marks are gone.

```{r}
html_source <- read_file("data/workforce/job-board.html")
stripped_before_squish <- html_source |>
  str_remove_all("<[^>]+>")

paragraphs_after_tag_removal <- split_blank_lines(stripped_before_squish)
paragraphs_after_squish <- stripped_before_squish |>
  str_squish() |>
  split_blank_lines()

boundary_loss <- tibble(
  source = c(
    "HTML block elements",
    "Tags removed, whitespace preserved",
    "Tags removed, whitespace squished"
  ),
  paragraph_like_units = c(
    nrow(html_blocks),
    length(paragraphs_after_tag_removal),
    length(paragraphs_after_squish)
  )
)

stopifnot(
  identical(length(paragraphs_after_tag_removal), 14L),
  identical(length(paragraphs_after_squish), 1L),
  identical(boundary_loss$paragraph_like_units, c(41L, 14L, 1L)),
  str_starts(paragraphs_after_squish[1], "Riverton workforce opportunities"),
  str_ends(paragraphs_after_squish[1], "Clear written communication is an essential skill.")
)

knitr::kable(
  boundary_loss,
  col.names = c("Rule", "Paragraph-like units"),
  caption = "Whitespace cleanup collapses the tag-stripped text",
  row.names = FALSE
)
```

Tag removal alone leaves 14 blank-line chunks because the source file still has
line breaks and indentation. Those chunks are not paragraphs; they are source
layout around titles, job cards, and lists. `str_squish()` removes the remaining
blank-line boundaries and leaves one long text string. For this file, paragraph
structure has to be captured before whitespace cleanup.

## What to remember

- A paragraph is a layout block, so the right rule depends on the file format.
- These help files have no blank lines, so the blank-line rule cannot show much.
- The chosen HTML selector captures 41 block-like units, including headings and
  list items.
- Squishing whitespace after tag removal collapses the text to one unit.

The help files and the job-board HTML need different boundary rules. Here the
reported counts are layout counts, not a grammar claim about where paragraphs
must begin.

## Sources

- [rvest documentation](https://rvest.tidyverse.org/)
- [HTML Living Standard section on grouping content](https://html.spec.whatwg.org/multipage/grouping-content.html)
- [readr documentation for reading files](https://readr.tidyverse.org/reference/read_file.html)
