---
title: "Checking small grammar rules"
subtitle: "Score three checks and keep the limits visible"
author: "Alex Farach"
description: "Learn how small grammar checks can find planted errors and why they are not a full grammar checker."
categories: [sentences and paragraphs, grammar checking, workforce research]
---

Correctly spelled sentences can still make a public summary sound broken, which
is why the coordinator hesitates. Real words are not enough when the words do not
fit together.

Spelling asks whether a word exists. **Grammar** asks whether the words fit
together in a sentence. This lesson builds three small checks and scores them on
sentences where the planted errors are known.

This project does not install a grammar checker, so the lesson builds three
small checks by hand and treats them as a demonstration. Offline means the check
runs from local files, without sending text to a web service.

::: {.callout-note}
The Riverton sentences and marked grammar examples are invented for this lesson.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain the difference between spelling and grammar checks;
- use UDPipe dependency output to compare a subject and a verb;
- find doubled words with a stringr pattern;
- test a simple `a` and `an` rule; and
- explain why a demonstration check is not a working grammar checker.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Make a marked test set

The setup reads the Riverton CSV, prepares words with tokenizers and purrr, runs
two pattern checks with stringr, and sends sentences to a local udpipe dependency
parser. dplyr and tibble organise the results. A **dependency parser** labels how
words relate to one another, such as which noun is the subject of a verb.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(udpipe)

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

marked_sentences <- tibble(
  example_id = sprintf("g%02d", 1:14),
  text = c(
    "The worker checks the list.",
    "The worker check the list.",
    "The applicants submit forms.",
    "The applicants submits forms.",
    "Bring a badge to orientation.",
    "Bring a envelope to orientation.",
    "Apply for an evening class.",
    "Apply for a evening class.",
    "The session begins in an hour.",
    "A university partner sends trainers.",
    "The coordinator checks the the form.",
    "The team needs needs support.",
    "The employee uses a spreadsheet.",
    "The employee use a spreadsheet."
  ),
  planted_check = c(
    "none",
    "subject-verb number",
    "none",
    "subject-verb number",
    "none",
    "article letters",
    "none",
    "article letters",
    "none",
    "none",
    "doubled word",
    "doubled word",
    "none",
    "subject-verb number"
  )
) |>
  mutate(has_planted_error = planted_check != "none")

knitr::kable(
  marked_sentences,
  col.names = c("Example ID", "Sentence", "Planted check", "Has planted error"),
  caption = "Author-marked sentences before running the checks",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(c(
  "readr", "dplyr", "tibble", "purrr", "stringr", "tokenizers", "udpipe"
) %in% loadedNamespaces()))

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(marked_sentences), 14L),
  identical(sum(marked_sentences$has_planted_error), 7L),
  identical(sum(!marked_sentences$has_planted_error), 7L),
  identical(sum(marked_sentences$planted_check == "subject-verb number"), 3L),
  identical(sum(marked_sentences$planted_check == "doubled word"), 2L),
  identical(sum(marked_sentences$planted_check == "article letters"), 2L)
)
```

Seven of the 14 test sentences are marked as correct, and seven contain exactly
one planted error. These examples describe only themselves. A checker that works
on them has not solved grammar checking.

## Run the three checks

The subject-verb check uses the UDPipe parser's `feats` column, which can contain
values such as `Number=Sing` or `Number=Plur`. It looks for an `nsubj` relation,
short for nominal subject, pointing to a verb. In the code, `\(tokens)` is R
shorthand for a small function applied to one sentence's tokens.

```{r}
extract_number <- function(feats) {
  str_match(coalesce(feats, ""), "Number=([^|]+)")[, 2]
}

find_subject_verb_flags <- function(annotated) {
  annotated_with_number <- annotated |>
    mutate(number = extract_number(feats))

  subjects <- annotated_with_number |>
    filter(dep_rel == "nsubj") |>
    transmute(
      doc_id,
      head_token_id,
      subject_token = token,
      subject_number = number
    )

  verbs <- annotated_with_number |>
    filter(upos %in% c("VERB", "AUX"), !is.na(number)) |>
    transmute(
      doc_id,
      verb_token_id = token_id,
      verb_token = token,
      verb_number = number
    )

  verbs |>
    inner_join(subjects, by = c("doc_id", "verb_token_id" = "head_token_id")) |>
    mutate(flag = !is.na(subject_number) & subject_number != verb_number)
}

score_check <- function(flag, expected) {
  tibble(
    true_positive = sum(flag & expected),
    false_positive = sum(flag & !expected),
    false_negative = sum(!flag & expected),
    true_negative = sum(!flag & !expected)
  )
}

parser <- udpipe_load_model("data/treebank/en_ewt-500-parser.udpipe")

marked_vertical <- marked_sentences$text |>
  tokenize_words(lowercase = FALSE, strip_punct = FALSE) |>
  map_chr(\(tokens) paste(tokens, collapse = "\n"))

marked_annotated <- udpipe_annotate(
  parser,
  x = marked_vertical,
  doc_id = marked_sentences$example_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "default"
) |>
  as.data.frame() |>
  as_tibble()

agreement_pairs <- find_subject_verb_flags(marked_annotated)
agreement_flags <- agreement_pairs |>
  summarise(agreement_flag = any(flag), .by = doc_id)

doubled_pattern <- regex("\\b([[:alpha:]]+)\\s+\\1\\b", ignore_case = TRUE)
article_pattern <- regex(
  "\\b(a)\\s+[aeiou][[:alpha:]]*|\\b(an)\\s+[^aeiou\\W][[:alpha:]]*",
  ignore_case = TRUE
)

scored_examples <- marked_sentences |>
  left_join(agreement_flags, by = c("example_id" = "doc_id")) |>
  mutate(
    agreement_flag = coalesce(agreement_flag, FALSE),
    doubled_flag = str_detect(text, doubled_pattern),
    article_flag = str_detect(text, article_pattern)
  )

check_scores <- bind_rows(
  score_check(
    scored_examples$agreement_flag,
    scored_examples$planted_check == "subject-verb number"
  ) |>
    mutate(check = "subject-verb number"),
  score_check(
    scored_examples$doubled_flag,
    scored_examples$planted_check == "doubled word"
  ) |>
    mutate(check = "doubled word"),
  score_check(
    scored_examples$article_flag,
    scored_examples$planted_check == "article letters"
  ) |>
    mutate(check = "article letters")
) |>
  select(check, everything())

expected_scores <- tibble(
  check = c("subject-verb number", "doubled word", "article letters"),
  true_positive = c(3L, 2L, 2L),
  false_positive = c(0L, 0L, 2L),
  false_negative = c(0L, 0L, 0L),
  true_negative = c(11L, 12L, 10L)
)

knitr::kable(
  agreement_pairs |>
    select(doc_id, subject_token, subject_number, verb_token, verb_number, flag),
  col.names = c("Example ID", "Subject", "Subject number", "Verb", "Verb number", "Flagged"),
  caption = "Subject-verb pairs found by the dependency parser",
  row.names = FALSE
)

knitr::kable(
  check_scores,
  col.names = c("Check", "True positives", "False positives", "False negatives", "True negatives"),
  caption = "Scores for the three grammar checks on marked examples",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    scored_examples$example_id[scored_examples$agreement_flag],
    c("g02", "g04", "g14")
  ),
  identical(
    scored_examples$example_id[scored_examples$doubled_flag],
    c("g11", "g12")
  ),
  identical(
    scored_examples$example_id[scored_examples$article_flag],
    c("g06", "g08", "g09", "g10")
  ),
  identical(check_scores, expected_scores)
)
```

The checks and expected answers share an author, so the score is a teaching
check rather than an accuracy estimate. A true positive is a planted error that a
check flags. A false positive is a correct sentence that a check flags. A false
negative is a planted error that a check misses. A true negative is a correct
sentence it leaves alone.

The subject-verb number check catches all three planted number errors in this
small set. The doubled-word check catches both repeated-word examples. The
article check catches the two planted `a` mistakes, but it also flags two correct
sentences: `an hour` and `A university`. That rule uses the first letter, while
English article choice depends on sound. `Hour` begins with a silent h, so it
sounds like a vowel. `University` begins with a y sound, so it sounds like a
consonant. The rule reads letters and cannot hear either word.

A real grammar checker uses thousands of rules, a trained model, or both. These
three checks are a demonstration of the idea. The UDPipe model was trained here
on 500 sentences and is deliberately weak, so every subject-verb result inherits
its mistakes. The article check is wrong by design so the failure is visible.

## Try the Riverton sentences

The same three checks can be run over the 28 Riverton sentences. The output is a
screening list, not a proof that every unflagged sentence is correct.

```{r}
riverton_vertical <- sentences$text |>
  tokenize_words(lowercase = FALSE, strip_punct = FALSE) |>
  map_chr(\(tokens) paste(tokens, collapse = "\n"))

riverton_annotated <- udpipe_annotate(
  parser,
  x = riverton_vertical,
  doc_id = sentences$sentence_id,
  tokenizer = "vertical",
  tagger = "default",
  parser = "default"
) |>
  as.data.frame() |>
  as_tibble()

riverton_agreement_pairs <- find_subject_verb_flags(riverton_annotated)
riverton_agreement_ids <- riverton_agreement_pairs |>
  filter(flag) |>
  pull(doc_id)
riverton_pair_sentence_count <- n_distinct(riverton_agreement_pairs$doc_id)

riverton_flags <- sentences |>
  transmute(
    sentence_id,
    text,
    agreement_flag = sentence_id %in% riverton_agreement_ids,
    doubled_flag = str_detect(text, doubled_pattern),
    article_flag = str_detect(text, article_pattern)
  ) |>
  mutate(any_flag = agreement_flag | doubled_flag | article_flag)

riverton_flag_summary <- tibble(
  check = c("subject-verb number", "doubled word", "article letters", "any check"),
  flagged_sentences = c(
    sum(riverton_flags$agreement_flag),
    sum(riverton_flags$doubled_flag),
    sum(riverton_flags$article_flag),
    sum(riverton_flags$any_flag)
  )
)

riverton_coverage <- tibble(
  measure = c("sentences with subject-verb pair", "sentences without subject-verb pair"),
  sentences = c(
    riverton_pair_sentence_count,
    nrow(sentences) - riverton_pair_sentence_count
  )
)

knitr::kable(
  riverton_flag_summary,
  col.names = c("Check", "Flagged Riverton sentences"),
  caption = "Grammar-check flags in the 28 Riverton sentences",
  row.names = FALSE
)

knitr::kable(
  riverton_coverage,
  col.names = c("Coverage measure", "Sentences"),
  caption = "Subject-verb coverage in the Riverton sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(riverton_agreement_pairs), 4L),
  identical(riverton_coverage$sentences, c(4L, 24L)),
  identical(riverton_flag_summary$flagged_sentences, c(0L, 0L, 0L, 0L)),
  identical(sum(riverton_flags$any_flag), 0L),
  identical(nrow(riverton_flags), 28L)
)
```

The three checks flag no Riverton sentence. The subject-verb check found only
four subject-and-verb pairs in all 28 sentences, because it needs the tagger to
mark a number feature on the verb and English verbs rarely carry one. On the
other 24 sentences the check could not fire. A zero mostly measures coverage, not
correctness.

## Grammar checking in production

The three custom checks above demonstrate the concept, but real offline grammar checking requires a mature system. In the NLP ecosystem, open-source tools like LanguageTool are standard. Production grammar systems distinguish between:

- **Detection vs. correction**: Highlighting a possible error (detection) is much safer than automatically rewriting the text (correction), especially when the engine's coverage is low.
- **Normative context and auditability**: What counts as "correct" grammar depends on the intended style (e.g., formal business English vs. informal messaging). Rule-based systems provide an auditable explanation for why text was flagged, whereas deep learning models can suggest fluent corrections but often struggle to explain their normative reasoning.
- **Rule/model/error-type comparison**: Rule-based engines are highly precise for local stylistic and typographical errors (like doubled words or wrong articles). Neural models are generally used for catching semantic confusion where the grammar technically parses but the meaning is broken.

## What to remember

- Spelling checks words; grammar checks how words fit together.
- UDPipe's `feats` and `nsubj` labels can support a small subject-verb check.
- A doubled-word pattern is simple and useful for one common typo.
- The article-letter rule falsely flags `an hour` and `A university`.
- These three checks are demonstrations, not an offline grammar checker for R.
- No Riverton flags here means low coverage, not clean grammar.

The three checks can produce a review queue. They cannot approve public text on
their own, especially when the most technical check fires on only four Riverton
sentences.

## Sources

- [UDPipe R package reference](https://bnosac.github.io/udpipe/docs/doc0.html)
- [Universal Dependencies relation `nsubj`](https://universaldependencies.org/u/dep/nsubj.html)
- [Universal Dependencies feature `Number`](https://universaldependencies.org/u/feat/Number.html)
