---
title: "Scoring readability formulas"
subtitle: "Compute Flesch scores and measure the syllable shortcut"
author: "Alex Farach"
description: "Learn how readability formulas use sentence, word, and syllable counts, and why their limits are severe."
categories: [sentences and paragraphs, readability scoring, workforce research]
---

Plain public help text is ready for review, but the coordinator still asks
whether the job board is harder to read than the help articles. A single score
would be convenient, but the score has to come from counted parts.

**Readability formulas** estimate text difficulty from surface counts such as
sentences, words, and syllables. A **syllable** is a spoken beat inside a word.
The formulas in this lesson can be computed by hand, which makes their limits
visible.

::: {.callout-note}
The job board and help articles scored here are fictional teaching texts.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define the counts used by Flesch Reading Ease;
- define the counts used by Flesch-Kincaid Grade Level;
- write a small syllable-counting heuristic;
- measure that heuristic against hand-counted job-board words; and
- explain why a readability score is not a reading level for a person.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Count syllables with a visible shortcut

This lesson reads local files, extracts HTML with rvest, tokenises text, and uses
dplyr, tibble, purrr, and stringr for the count checks. The syllable counter
below is only a shortcut: clean the word, drop a final silent `e` unless the word
ends in consonant plus `le`, then count groups of vowel letters including `y`.
In the code, `\(word)` is R shorthand for a small function applied to one word.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(tokenizers)
library(rvest)

job_page <- read_html("data/workforce/job-board.html")

job_all_blocks <- job_page |>
  html_elements("h1, h2, p, li") |>
  html_text2()

job_detail_sentences <- job_page |>
  html_elements("li.job-detail") |>
  html_text2()

job_context_blocks <- job_page |>
  html_elements("h1, h2, p") |>
  html_text2()

help_text <- list.files(
  "data/help_articles",
  pattern = "[.]txt$",
  full.names = TRUE
) |>
  sort() |>
  map_chr(read_file) |>
  str_c(collapse = "\n\n")

count_syllables <- function(words) {
  clean_words <- words |>
    str_to_lower() |>
    str_replace_all("[^a-z]", "")

  map_int(clean_words, \(word) {
    if (identical(word, "")) {
      return(0L)
    }

    adjusted_word <- if (str_detect(word, "[^aeiou]le$")) {
      word
    } else {
      str_remove(word, "e$")
    }

    max(1L, str_count(adjusted_word, "[aeiouy]+"))
  })
}

syllable_check <- tibble(
  word = c(
    "paid",
    "training",
    "provided",
    "prior",
    "data",
    "experience",
    "required",
    "evening",
    "schedules",
    "available",
    "applicants",
    "basic",
    "spreadsheet",
    "skills",
    "employer",
    "certification",
    "rotating",
    "workers",
    "portfolio",
    "remote",
    "receives",
    "mentor",
    "medical",
    "certificate",
    "preferred",
    "daytime",
    "apprenticeship",
    "valid",
    "license",
    "customer"
  ),
  hand_syllables = c(
    1L, 2L, 3L, 2L, 2L, 4L, 2L, 3L, 2L, 4L,
    3L, 2L, 2L, 1L, 3L, 5L, 3L, 2L, 4L, 2L,
    2L, 2L, 3L, 4L, 2L, 2L, 4L, 2L, 2L, 3L
  )
) |>
  mutate(
    heuristic_syllables = count_syllables(word),
    comparison = case_when(
      heuristic_syllables == hand_syllables ~ "exact",
      heuristic_syllables > hand_syllables ~ "overcount",
      TRUE ~ "undercount"
    )
  )

syllable_summary <- syllable_check |>
  count(comparison, name = "words")

job_detail_words <- tokenize_words(
  job_detail_sentences,
  lowercase = TRUE,
  strip_punct = TRUE
) |>
  unlist(use.names = FALSE)

expected_syllable_summary <- tibble(
  comparison = c("exact", "overcount", "undercount"),
  words = c(21L, 5L, 4L)
)

knitr::kable(
  syllable_check,
  col.names = c("Word", "Hand count", "Heuristic count", "Comparison"),
  caption = "Hand-counted job-board syllables compared with the vowel-group heuristic",
  row.names = FALSE
)

knitr::kable(
  syllable_summary,
  col.names = c("Comparison", "Words"),
  caption = "Syllable heuristic results for 30 job-board words",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(all(c(
  "readr", "dplyr", "tibble", "purrr", "stringr", "tokenizers", "rvest"
) %in% loadedNamespaces()))

stopifnot(
  identical(length(job_all_blocks), 41L),
  identical(length(job_detail_sentences), 22L),
  identical(nrow(syllable_check), 30L),
  all(syllable_check$word %in% job_detail_words),
  identical(syllable_summary, expected_syllable_summary),
  identical(sum(syllable_check$hand_syllables), 78L),
  identical(sum(syllable_check$heuristic_syllables), 79L)
)
```

On these 30 job-board words, the heuristic matches 21 of 30, overcounts 5 of 30,
and undercounts 4 of 30. The hand counts sum to 78 syllables, while the heuristic
sums to 79. That one-syllable difference is small here, but the error is still
real and it enters the readability formulas directly. The failures show three
common traps: adjacent vowels at a syllable break, final `es` endings, and `y`
joining a vowel group as in `employer`.

## Compute the formulas directly

Flesch Reading Ease uses this formula:

`206.835 - 1.015 * (words / sentences) - 84.6 * (syllables / words)`

Flesch-Kincaid Grade Level uses this formula:

`0.39 * (words / sentences) + 11.8 * (syllables / words) - 15.59`

The code below uses those definitions. It does not call a readability package to
hide the arithmetic.

```{r}
score_from_counts <- function(document, sentence_count, word_count, syllable_count) {
  tibble(
    document = document,
    sentences = sentence_count,
    words = word_count,
    syllables = syllable_count,
    words_per_sentence = round(word_count / sentence_count, 1),
    syllables_per_word = round(syllable_count / word_count, 1),
    flesch_reading_ease = round(
      206.835 - 1.015 * (word_count / sentence_count) -
        84.6 * (syllable_count / word_count)
    ),
    flesch_kincaid_grade = round(
      0.39 * (word_count / sentence_count) +
        11.8 * (syllable_count / word_count) - 15.59
    )
  )
}

score_readability <- function(text, document) {
  sentence_count <- tokenize_sentences(text) |>
    map_int(length) |>
    sum()
  word_tokens <- tokenize_words(
    text,
    lowercase = TRUE,
    strip_punct = TRUE
  ) |>
    unlist(use.names = FALSE)
  word_count <- length(word_tokens)
  syllable_count <- sum(count_syllables(word_tokens))

  score_from_counts(document, sentence_count, word_count, syllable_count)
}

preprocessing_counts <- tibble(
  text_set = c(
    "all HTML blocks collapsed",
    "job-detail list items",
    "headings, employers, and locations"
  ),
  blocks = c(
    length(job_all_blocks),
    length(job_detail_sentences),
    length(job_context_blocks)
  ),
  sentences = c(
    length(tokenize_sentences(str_c(job_all_blocks, collapse = " "))[[1]]),
    sum(map_int(tokenize_sentences(job_detail_sentences), length)),
    sum(map_int(tokenize_sentences(job_context_blocks), length))
  ),
  words = c(
    length(unlist(tokenize_words(str_c(job_all_blocks, collapse = " "), lowercase = TRUE, strip_punct = TRUE), use.names = FALSE)),
    length(job_detail_words),
    length(unlist(tokenize_words(job_context_blocks, lowercase = TRUE, strip_punct = TRUE), use.names = FALSE))
  )
)

readability_scores <- bind_rows(
  score_readability(job_detail_sentences, "job detail sentences"),
  score_readability(help_text, "help articles")
)

collapsed_score <- score_readability(str_c(job_all_blocks, collapse = " "), "job board collapsed")

job_score <- readability_scores |>
  filter(document == "job detail sentences")

help_score <- readability_scores |>
  filter(document == "help articles")

net_bias_ratio <- sum(syllable_check$heuristic_syllables) / sum(syllable_check$hand_syllables)
adjusted_job_syllables <- round(job_score$syllables / net_bias_ratio)

syllable_sensitivity <- bind_rows(
  job_score,
  score_from_counts(
    "job detail sentences, net-bias adjusted",
    job_score$sentences,
    job_score$words,
    adjusted_job_syllables
  )
)

knitr::kable(
  preprocessing_counts,
  col.names = c("Text set", "Blocks", "Sentence pieces", "Words"),
  caption = "What is included before the readability formulas run",
  row.names = FALSE
)

knitr::kable(
  readability_scores,
  col.names = c(
    "Document",
    "Sentences",
    "Words",
    "Syllables",
    "Words per sentence",
    "Syllables per word",
    "Flesch Reading Ease, rounded",
    "Flesch-Kincaid grade, rounded"
  ),
  caption = "Readability formulas computed directly from counted parts",
  row.names = FALSE
)

knitr::kable(
  syllable_sensitivity |>
    select(document, syllables, flesch_reading_ease, flesch_kincaid_grade),
  col.names = c("Document", "Syllables", "Flesch Reading Ease, rounded", "Flesch-Kincaid grade, rounded"),
  caption = "Effect of applying the observed net syllable bias to the job-board score",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(preprocessing_counts$sentences, c(22L, 22L, 19L)),
  identical(preprocessing_counts$words, c(177L, 133L, 44L)),
  identical(job_score$sentences, 22L),
  identical(job_score$words, 133L),
  identical(job_score$syllables, 230L),
  isTRUE(all.equal(job_score$words_per_sentence, 6.0)),
  isTRUE(all.equal(job_score$syllables_per_word, 1.7)),
  isTRUE(all.equal(job_score$flesch_reading_ease, 54)),
  isTRUE(all.equal(job_score$flesch_kincaid_grade, 7)),
  identical(collapsed_score$sentences, 22L),
  identical(collapsed_score$words, 177L),
  identical(collapsed_score$flesch_reading_ease, 33),
  identical(help_score$sentences, 6L),
  identical(help_score$words, 56L),
  identical(help_score$syllables, 89L),
  isTRUE(all.equal(help_score$words_per_sentence, 9.3)),
  isTRUE(all.equal(help_score$syllables_per_word, 1.6)),
  isTRUE(all.equal(help_score$flesch_reading_ease, 63)),
  isTRUE(all.equal(help_score$flesch_kincaid_grade, 7)),
  identical(syllable_sensitivity$flesch_reading_ease, c(54, 56))
)
```

The code scores only the 22 job-detail list items as job-board sentences. The 19
headings, employer names, and location labels contain 44 words, but they are not
running sentences. If all 41 HTML blocks are glued into one string, those labels
run into the list items and the Flesch score rounds to 33. That is a preprocessing
artefact, not a reading result.

With the list items kept separate, the job-board detail sentences have 22
sentences, 133 words, and 230 heuristic syllables. The help articles have 6
sentences, 56 words, and 89 syllables. The rounded scores are 54 and 7 for the
job details, and 63 and 7 for the help articles. On common Flesch bands, the
collapsed score of 33 falls in the difficult range; treating short job-board
sentences as college-level prose would be implausible. The split score is still
driven by the syllable term, especially long nouns such as `certification` and
`apprenticeship`.

Those values compare two teaching texts under one set of counting rules. The
formulas were fitted decades ago for particular populations (e.g., US military personnel or schoolchildren) and specific tasks. They count
surface features and know nothing about meaning, layout, stakes, or whether a
reader already knows the subject. A readability formula is a surface heuristic, never a true comprehension model. In practice, calculating these formulas reliably requires heavily tested ecosystem tools like `quanteda.textstats` in R or `textstat` in Python rather than custom arithmetic.

The 30-word audit found errors on nearly a
third of tested words, and the net bias alone moves the job-detail Flesch score
from 54 to 56. The help article sample is only 56 words. Whole-number scores are
already more precision than the inputs deserve.

## What to remember

- Readability formulas use sentence, word, and syllable counts.
- Preprocessing can change the sentence and word counts before the formula runs.
- The syllable shortcut matched 21 of 30 job-board words and missed 9.
- Whole-number scores are still rough because the syllable input is rough.

The help articles score a little easier than the job-detail sentences here, but
the samples are small and the syllable counter is noisy. The numbers are a prompt
to read the text, not a substitute for reading it.

## Sources

- [Flesch, A new readability yardstick](https://doi.org/10.1037/h0057532)
- [Kincaid and colleagues, Derivation of new readability formulas](https://eric.ed.gov/?id=ED149038)
- [tokenizers documentation for word and sentence tokenizers](https://docs.ropensci.org/tokenizers/reference/basic-tokenizers.html)
