---
title: "Catching scam messages"
subtitle: "Read a spam rule, train a small filter, and choose which mistake costs more"
description: "Learn how spam detection turns text into decisions while keeping false alarms and missed scams separate."
categories: [classification, spam detection, workforce research]
---

Maya opens the Riverton Workforce Lab inbox before the morning appointments.
Most messages ask ordinary questions about applications, passwords, and classes.
Mixed into the same list are offers that ask jobseekers to send money.

If the Lab blocks too much, a real request can disappear. If it blocks too
little, a scam can reach someone who is looking for work. A spam filter is a
text classifier, a tool that assigns a label to a piece of writing.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

A dataset written by the same person who writes the classifier cannot measure
whether the method works. This invented inbox checks that the code runs and
makes the trade-off visible; it does not estimate performance for real spam
filters.

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- read a transparent spam rule;
- train a small text classifier;
- compare the rule and model where they disagree;
- read a confusion matrix; and
- explain how a decision threshold trades false alarms against missed scams.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the invented inbox

The inbox file was generated once by `data-raw/build-riverton-inbox.R` and then
committed. The lesson reads that file; it does not write examples while the page
renders. readr opens the CSV files, dplyr and tidyr shape tables, stringr reads
patterns, ggplot2 draws the chart, tidymodels fits the classifier, textrecipes
turns words into features, glmnet fits the model, and digest checks the file.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(tidymodels)
library(textrecipes)
library(glmnet)
library(digest)

inbox_path <- "data/riverton/riverton-inbox.csv"
inbox <- read_csv(
  inbox_path,
  na = character(),
  col_types = cols(
    message_id = col_character(),
    text = col_character(),
    is_spam = col_logical(),
    intent = col_character(),
    author_note = col_character()
  )
)

inbox_metadata <- read_csv(
  "data/riverton/riverton-inbox-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    purpose = col_character(),
    rows = col_integer(),
    spam_class_counts = col_character(),
    intent_class_counts = col_character(),
    fingerprint = col_character()
  )
)

spam_counts <- inbox |>
  count(is_spam, name = "messages") |>
  mutate(class = if_else(is_spam, "job scam", "genuine message"))

inbox_hash <- digest(
  paste(read_lines(inbox_path), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)

knitr::kable(
  spam_counts |>
    select(class, messages),
  col.names = c("Recorded class", "Messages"),
  caption = "Invented Riverton inbox labels",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "readr", "dplyr", "tibble", "tidyr", "purrr", "stringr",
    "ggplot2", "tidymodels", "textrecipes", "glmnet", "digest"
  ) %in% loadedNamespaces()),
  identical(nrow(inbox), 60L),
  identical(spam_counts$messages, c(44L, 16L)),
  identical(inbox_metadata$rows, 60L),
  identical(
    inbox_hash,
    "b71b06e353918e7a421327883950c7409ea149c34b6ba26396fc5bc8c27df216"
  ),
  identical(inbox_hash, inbox_metadata$fingerprint)
)
```

The file has 60 invented messages: 44 genuine messages and 16 job scams. Those
counts describe the fixture only. They do not say how much scam traffic reaches
a workforce office.

Hand-written fixtures can fail before any model runs. The classes may separate
on grammar rather than on the intended idea. A **superficial feature** is a cue
such as punctuation or pronoun use that can predict a label without reading the
substance of the message. The check below tests a bad shortcut: mark a message as
spam when it has no first-person pronoun and does not end with a question mark.

```{r}
first_person_pattern <- regex(
  "\\b(I|I'm|I’ve|I’d|I’ll|me|my|mine|we|we're|we’ve|us|our|ours)\\b",
  ignore_case = TRUE
)

surface_rule <- inbox |>
  mutate(
    first_person_pronoun = str_detect(text, first_person_pattern),
    terminal_question_mark = str_detect(text, "\\?\\s*$"),
    surface_spam = !first_person_pronoun & !terminal_question_mark,
    truth = factor(if_else(is_spam, "spam", "genuine"), levels = c("spam", "genuine")),
    estimate = factor(
      if_else(surface_spam, "spam", "genuine"),
      levels = c("spam", "genuine")
    )
  )

surface_confusion <- yardstick::conf_mat(
  surface_rule,
  truth = truth,
  estimate = estimate
)
surface_matrix <- as.matrix(surface_confusion$table)

surface_scores <- tibble(
  shortcut = "No first-person pronoun and no terminal question mark",
  accuracy = yardstick::accuracy(surface_rule, truth = truth, estimate = estimate) |>
    pull(.estimate),
  scam_recall = yardstick::sens(
    surface_rule,
    truth = truth,
    estimate = estimate,
    event_level = "first"
  ) |>
    pull(.estimate),
  genuine_recall = yardstick::spec(
    surface_rule,
    truth = truth,
    estimate = estimate,
    event_level = "first"
  ) |>
    pull(.estimate)
)

knitr::kable(
  surface_scores |>
    mutate(across(c(accuracy, scam_recall, genuine_recall), \(value) round(value, 4))),
  col.names = c("Shortcut", "Accuracy", "Scam recall", "Genuine recall"),
  caption = "A grammar-only shortcut no longer solves the invented inbox",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(
    surface_matrix ==
      matrix(
        c(5L, 9L, 11L, 35L),
        nrow = 2,
        byrow = TRUE,
        dimnames = dimnames(surface_matrix)
      )
  ),
  identical(round(surface_scores$accuracy, 4), 0.6667),
  identical(round(surface_scores$scam_recall, 4), 0.3125),
  identical(round(surface_scores$genuine_recall, 4), 0.7955)
)
```

This shortcut scores 0.6667 accuracy and 0.3125 scam recall. That weak result is
the point. When you write examples yourself, the two classes can separate on a
pattern you did not mean to teach. The first check on any hand-built dataset is
whether a stupid feature already solves it.

One shortcut is not the whole audit, so the next chunk sweeps several. For each
feature it tries both orientations, marking a message as a scam when the feature
is present and again when it is absent, and keeps whichever scores better. That
is deliberately generous to the shortcut, because the question is whether any
cheap cue could substitute for reading.

```{r}
superficial_features <- inbox |>
  transmute(
    is_spam,
    `first-person pronoun` = str_detect(text, first_person_pattern),
    `ends with a question mark` = str_detect(text, "\\?\\s*$"),
    `longer than the median` = str_count(text, "\\S+") >= median(str_count(text, "\\S+")),
    `exclamation mark` = str_detect(text, "!"),
    `contains a digit` = str_detect(text, "[0-9]")
  ) |>
  pivot_longer(-is_spam, names_to = "feature", values_to = "present")

shortcut_sweep <- superficial_features |>
  reframe(
    spam_when_feature_is = c("present", "absent"),
    accuracy = c(
      mean(present == is_spam),
      mean(present != is_spam)
    ),
    scam_recall = c(
      mean(present[is_spam]),
      mean(!present[is_spam])
    ),
    .by = feature
  ) |>
  arrange(feature, desc(accuracy), desc(scam_recall)) |>
  slice_head(n = 1, by = feature) |>
  arrange(desc(accuracy), feature)

shortcut_display <- bind_rows(
  shortcut_sweep,
  tibble(
    feature = "always genuine baseline",
    spam_when_feature_is = "never",
    accuracy = mean(!inbox$is_spam),
    scam_recall = 0
  )
) |>
  arrange(desc(accuracy), feature)

length_feature_counts <- superficial_features |>
  filter(feature == "longer than the median") |>
  summarise(
    scams_above_median = sum(present & is_spam),
    genuine_above_median = sum(present & !is_spam)
  )

knitr::kable(
  shortcut_display |>
    mutate(across(where(is.numeric), \(value) round(value, 4))),
  col.names = c(
    "Superficial feature",
    "Predict spam when feature is",
    "Accuracy",
    "Scam recall"
  ),
  caption = "Each shortcut uses one orientation and is compared with the majority baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(shortcut_sweep), 5L),
  identical(nrow(shortcut_display), 6L),
  identical(shortcut_sweep$feature[[1]], "exclamation mark"),
  identical(
    round(
      shortcut_sweep$accuracy[
        shortcut_sweep$feature == "exclamation mark"
      ],
      4
    ),
    0.7667
  ),
  identical(
    round(
      shortcut_sweep$scam_recall[
        shortcut_sweep$feature == "exclamation mark"
      ],
      4
    ),
    0.125
  ),
  identical(
    round(
      shortcut_sweep$accuracy[
        shortcut_sweep$feature == "contains a digit"
      ],
      4
    ),
    0.7333
  ),
  identical(
    round(
      shortcut_display$accuracy[
        shortcut_display$feature == "always genuine baseline"
      ],
      4
    ),
    0.7333
  ),
  identical(length_feature_counts$scams_above_median, 13L),
  identical(length_feature_counts$genuine_above_median, 22L)
)
```

No cue improves accuracy much beyond always predicting `genuine`. Exclamation
marks improve accuracy by two messages, but catch only 2 of 16 scams. The digit
cue exactly matches the 0.7333 majority baseline. Message length exposes another
fixture artifact: 13 of 16 scams, but also 22 of 44 genuine messages, are longer
than the median. That cue catches many scams while producing too many false
alarms to classify well. These are diagnostics of how the hand-written file was
made, not evidence that any cue will transfer. The way to find out is to test
text that did not shape the examples or the rule.

The build script for this file recomputes the orientation-preserving sweep before
writing the CSV and refuses to write if any single feature reaches 0.80 accuracy
or 0.95 scam recall.

## Start with a rule a person can read

A rule-based spam filter searches for known warning signs. This one marks a
message when it mentions fees, gift cards, crypto payments, bank details,
guaranteed placement, deposits, or similar language. The rule is easy to inspect,
which also means it is easy for a scammer to route around.

```{r}
spam_terms <- c(
  "fee", "payment", "gift card", "gift cards", "crypto",
  "wire transfer", "deposit", "bank", "card number",
  "guarantee", "guaranteed", "starter kit", "activation fee",
  "Social Security", "secret shopper", "premium number",
  "pay now", "\\$[0-9]+"
)

spam_pattern <- regex(
  paste(spam_terms, collapse = "|"),
  ignore_case = TRUE
)

term_hits <- tibble(term = spam_terms) |>
  mutate(
    messages_matched = map_int(
      term,
      \(term_text) sum(str_detect(inbox$text, regex(term_text, ignore_case = TRUE)))
    )
  )

rule_results <- inbox |>
  mutate(
    truth = factor(
      if_else(is_spam, "spam", "genuine"),
      levels = c("spam", "genuine")
    ),
    estimate = factor(
      if_else(str_detect(text, spam_pattern), "spam", "genuine"),
      levels = c("spam", "genuine")
    )
  )

rule_confusion <- yardstick::conf_mat(
  rule_results,
  truth = truth,
  estimate = estimate
)

rule_matrix <- as.matrix(rule_confusion$table)

knitr::kable(
  as.data.frame(rule_confusion$table),
  col.names = c("Rule prediction", "Recorded label", "Messages"),
  caption = "Rule-based spam filter on the full invented inbox",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(sum(term_hits$messages_matched == 1L), 11L),
  all(
    rule_matrix ==
      matrix(
        c(14L, 2L, 2L, 42L),
        nrow = 2,
        byrow = TRUE,
        dimnames = dimnames(rule_matrix)
      )
  )
)
```

The rule marks 14 of the 16 invented scams and also blocks two genuine messages.
It misses two scams. In this fixture, the easy cases are the ones that say
`fee`, `deposit`, or `crypto` plainly.

The word list was written after reading these messages. Eleven of its 18 terms
match exactly one message, so the full 60-row table is a lookup table scored on
the same rows that shaped it. The two blocked genuine messages were written with
scam vocabulary on purpose to show that a readable rule can fail on reports about
scams. The comparison below applies both methods to the same 19 rows, but that
does not make the rule out-of-sample: its terms were chosen after reading all 60
messages, including those 19.

## Train a small filter

A trained model learns word weights from labelled examples. The seed for the
train-test split is 4801, and the seed before fitting is 4802. The split is
stratified, meaning each side keeps both recorded classes.

```{r}
set.seed(4801)
spam_data <- inbox |>
  mutate(
    spam_label = factor(
      if_else(is_spam, "spam", "genuine"),
      levels = c("spam", "genuine")
    )
  )

spam_split <- initial_split(
  spam_data,
  prop = 0.7,
  strata = spam_label
)
spam_train <- training(spam_split)
spam_test <- testing(spam_split)

spam_recipe <- recipe(spam_label ~ text, data = spam_train) |>
  step_tokenize(text) |>
  step_tokenfilter(text, max_tokens = 80) |>
  step_tfidf(text)

spam_model <- logistic_reg(penalty = 0.01, mixture = 1) |>
  set_engine("glmnet") |>
  set_mode("classification")

set.seed(4802)
spam_fit <- workflow() |>
  add_recipe(spam_recipe) |>
  add_model(spam_model) |>
  fit(data = spam_train)

probability_column <- predict(spam_fit, spam_test, type = "prob") |>
  names() |>
  keep(\(column) str_detect(column, "spam$"))

spam_scored <- spam_test |>
  select(message_id, text, is_spam, spam_label) |>
  bind_cols(predict(spam_fit, spam_test, type = "prob")) |>
  mutate(
    spam_probability = .data[[probability_column]],
    model_spam = spam_probability >= 0.5,
    rule_spam = str_detect(text, spam_pattern),
    model_label = factor(
      if_else(model_spam, "spam", "genuine"),
      levels = c("spam", "genuine")
    )
  )

model_confusion <- yardstick::conf_mat(
  spam_scored,
  truth = spam_label,
  estimate = model_label
)
model_matrix <- as.matrix(model_confusion$table)

test_rule_confusion <- yardstick::conf_mat(
  spam_scored |>
    mutate(
      rule_label = factor(
        if_else(rule_spam, "spam", "genuine"),
        levels = c("spam", "genuine")
      )
    ),
  truth = spam_label,
  estimate = rule_label
)
test_rule_matrix <- as.matrix(test_rule_confusion$table)

test_comparison <- tibble(
  method = c("Readable rule", "Trained model", "Always genuine"),
  correct = c(
    sum(spam_scored$rule_spam == spam_scored$is_spam),
    sum(spam_scored$model_spam == spam_scored$is_spam),
    sum(!spam_scored$is_spam)
  ),
  rows = nrow(spam_scored),
  accuracy = correct / rows,
  evidence_status = c(
    "in-sample: terms came from all 60 messages",
    "out-of-sample for the fitted model",
    "not fitted"
  )
)

knitr::kable(
  as.data.frame(test_rule_confusion$table),
  col.names = c("Rule prediction", "Recorded label", "Messages"),
  caption = "Rule-based spam filter on the same invented test split",
  row.names = FALSE
)

knitr::kable(
  as.data.frame(model_confusion$table),
  col.names = c("Model prediction", "Recorded label", "Messages"),
  caption = "Model confusion matrix on the invented test split",
  row.names = FALSE
)

knitr::kable(
  test_comparison |>
    mutate(accuracy = round(accuracy, 4)),
  col.names = c("Method", "Correct", "Rows", "Accuracy", "Evidence status"),
  caption = "The model is compared with its same-split majority baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(spam_train), 41L),
  identical(nrow(spam_test), 19L),
  identical(as.integer(count(spam_test, spam_label)$n), c(5L, 14L)),
  identical(test_comparison$correct, c(16L, 14L, 14L)),
  identical(
    round(test_comparison$accuracy, 4),
    c(0.8421, 0.7368, 0.7368)
  ),
  all(
    test_rule_matrix ==
      matrix(
        c(3L, 1L, 2L, 13L),
        nrow = 2,
        byrow = TRUE,
        dimnames = dimnames(test_rule_matrix)
      )
  ),
  all(
    model_matrix ==
      matrix(
        c(3L, 3L, 2L, 11L),
        nrow = 2,
        byrow = TRUE,
        dimnames = dimnames(model_matrix)
      )
  ),
  identical(
    round(
      spam_scored |>
        filter(message_id == "M004") |>
        pull(spam_probability),
      3
    ),
    0.972
  )
)
```

On the same 19 rows, the rule labels 3 of the 5 invented scams as spam and blocks
one genuine message. Its 0.8421 accuracy is in-sample because the terms were
written after reading all 60 messages. The model also catches 3 scams, but blocks
three genuine messages. Its out-of-sample accuracy is 0.7368, exactly the
always-genuine baseline on this split. The rule and model therefore cannot be
ranked as held-out competitors. The table does show why a score needs a baseline:
without one, a fitted model can look informative while matching a trivial rule.

## Look at the disagreements

The rule and the model do not mark the same messages. Disagreements are useful
because a person can inspect a short list and ask which tool failed in a way that
matters.

```{r}
disagreements <- spam_scored |>
  filter(rule_spam != model_spam) |>
  arrange(message_id) |>
  mutate(
    recorded_label = if_else(is_spam, "spam", "genuine"),
    spam_probability = round(spam_probability, 3)
  ) |>
  select(
    message_id,
    recorded_label,
    spam_probability,
    rule_spam,
    model_spam,
    text
  )

knitr::kable(
  disagreements,
  col.names = c(
    "Message",
    "Recorded label",
    "Model spam score",
    "Rule says spam",
    "Model says spam",
    "Text"
  ),
  caption = "Where the readable rule and the trained model disagree",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(disagreements$message_id, c("M004", "M030", "M034", "M046")),
  identical(disagreements$recorded_label, c("genuine", "genuine", "genuine", "genuine")),
  identical(disagreements$spam_probability, c(0.972, 0.869, 0.010, 0.998))
)
```

The message about documents reaching the office looks genuine to the recorded
label but receives a high model score. The message about a bank password is
genuine because it reports a suspicious listing, yet the rule catches the word
`bank`. The two callback and eligibility messages show the model's own false
alarms. Those are different kinds of errors.

## Move the cut point

The model produces a score between 0 and 1. The cut point turns that score into a
decision. A lower cut point catches more possible scams and risks blocking more
genuine messages. A higher cut point does the reverse.

This curve uses the same 19 rows as the model table. Choosing a cut point from
the curve and then reporting its errors on these rows would spend the test set
twice. A deployed threshold should be chosen on separate calibration data before
the final test.

```{r}
#| label: fig-spam-threshold
#| fig-cap: "The spam cut point shifts false alarms first, then missed scams, on the invented test split."
#| fig-alt: "Line chart with thresholds from 0.05 to 0.95. Genuine messages blocked drops from four to two. Scams missed rises from one to three."

threshold_table <- tibble(threshold = seq(0.05, 0.95, by = 0.05)) |>
  mutate(
    results = map(threshold, \(cut_point) {
      spam_scored |>
        summarise(
          genuine_blocked = sum(spam_probability >= cut_point & !is_spam),
          scams_missed = sum(spam_probability < cut_point & is_spam),
          sent_to_review = sum(spam_probability >= cut_point),
          .groups = "drop"
        )
    })
  ) |>
  unnest(results)

threshold_plot <- threshold_table |>
  select(threshold, genuine_blocked, scams_missed) |>
  pivot_longer(
    cols = c(genuine_blocked, scams_missed),
    names_to = "error_type",
    values_to = "messages"
  ) |>
  mutate(
    error_type = recode(
      error_type,
      genuine_blocked = "Genuine messages blocked",
      scams_missed = "Scams missed"
    )
  )

knitr::kable(
  threshold_table,
  col.names = c(
    "Spam score cut point",
    "Genuine messages blocked",
    "Scams missed",
    "Messages sent to spam review"
  ),
  caption = "Threshold trade-off on the invented test split",
  row.names = FALSE
)

ggplot(threshold_plot, aes(x = threshold, y = messages, color = error_type)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = threshold_table$threshold) +
  labs(
    x = "Decision threshold",
    y = "Messages",
    color = "Error type"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  identical(threshold_table$genuine_blocked, c(4L, rep(3L, 16), 2L, 2L)),
  identical(threshold_table$scams_missed, c(rep(1L, 5), rep(2L, 13), 3L)),
  identical(threshold_table$sent_to_review, c(8L, rep(7L, 4), rep(6L, 12), 5L, 4L))
)
```

For this invented split, thresholds from 0.30 through 0.85 have the same error
counts: three genuine messages blocked and two scams missed. Only the low and
high ends move the counts. That is a number about this fixture, not about the
world. A company inbox might prefer to block anything suspicious. A workforce
inbox serving jobseekers may need a review queue because deleting a genuine
posting can cost someone an opportunity, while letting a scam through can cost
money or personal information.

Real spam also adapts. A static test set cannot measure what happens after
senders learn the filter's habits and change their wording.

## What to remember

- Spam detection is text classification with unequal errors.
- A readable rule helps inspection, but it is brittle.
- A trained model still needs a human-chosen threshold.
- Every score needs a baseline and a named test set.
- The Riverton inbox proves only that the lesson code runs on an invented file.

## Sources

- [glmnet package documentation](https://glmnet.stanford.edu/articles/glmnet.html)
- [tidymodels textrecipes reference](https://textrecipes.tidymodels.org/)
- [yardstick confusion matrix reference](https://yardstick.tidymodels.org/reference/conf_mat.html)
