---
title: "Scoring customer feeling"
subtitle: "Compare a word list with labels from the domain"
author: "Alex Farach"
description: "Learn why sentiment scores from word lists and trained labels can disagree, especially with short feedback."
categories: [classification, sentiment analysis, emotion analysis]
---

At the end of the week, Nia receives five customer comments and one blank rating
cell. The dashboard wants a quick mood score, but the comments are too short to
hide the weak spots.

One row thanks the service. Another says a parcel arrived late. A third says the
sender could not reset a password. The same column holds praise, complaint, and
language the English word list will miss.

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- score comments with the Bing sentiment lexicon;
- count how many comments receive no lexicon match;
- identify context failures such as negation and domain wording;
- contrast a lexicon score with labels from the same domain; and
- explain why a word score is not a claim about a person's inner state.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the feedback file

The feedback file is fictional. It was written for teaching and contains no real
customers, parcels, accounts, or service records.

readr opens the CSV, dplyr and tibble shape the rows, tidyr handles missing
cells, purrr repeats scoring, stringr checks words, tidytext supplies the
lexicon, ggplot2 gives a small plot, and quanteda supplies the bundled inaugural
corpus through the shared helper. The file keeps the blank rating as a string
first because the project does not let `read_csv()` guess missing values.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(tidytext)
library(ggplot2)
library(quanteda)

feedback <- read_csv(
  "data/customer_feedback.csv",
  na = character(),
  col_types = cols(
    feedback_id = col_integer(),
    submitted_at = col_date(),
    channel = col_character(),
    rating = col_character(),
    comment = col_character()
  )
) |>
  mutate(
    rating = if_else(rating == "", NA_integer_, as.integer(rating))
  )

stopifnot(
  all(c(
    "readr", "dplyr", "tibble", "tidyr", "purrr", "stringr",
    "tidytext", "ggplot2", "quanteda"
  ) %in% loadedNamespaces()),
  identical(nrow(feedback), 5L),
  identical(feedback$feedback_id, 1001:1005),
  identical(feedback$rating, c(5L, 4L, NA_integer_, 2L, 5L))
)

knitr::kable(
  feedback,
  col.names = c("ID", "Submitted", "Channel", "Rating", "Comment"),
  caption = "Customer feedback used in this lesson",
  row.names = FALSE
)
```

The blank rating stays visible. Turning it into a zero would invent a rating the
customer did not give.

## Count sentiment words

A **sentiment lexicon** is a word list with labels attached. The Bing Liu
opinion lexicon in tidytext comes from product-review research by Minqing Hu and
Bing Liu. It has 6,786 English words, each marked positive or negative. The
source page asks users to cite the work and states research-use terms rather than
an open-data licence. A lexicon score counts matches and gives positive words +1
and negative words -1.

```{r}
bing <- tidytext::get_sentiments("bing")

token_scores <- feedback |>
  tidytext::unnest_tokens(word, comment) |>
  left_join(
    bing,
    by = join_by(word),
    multiple = "all"
  ) |>
  mutate(
    token_score = case_when(
      sentiment == "positive" ~ 1L,
      sentiment == "negative" ~ -1L,
      TRUE ~ 0L
    )
  )

lexicon_scores <- token_scores |>
  summarise(
    matched_words = sum(token_score != 0L),
    lexicon_score = sum(token_score),
    .by = feedback_id
  ) |>
  right_join(feedback, by = join_by(feedback_id)) |>
  mutate(
    matched_words = replace_na(matched_words, 0L),
    lexicon_score = replace_na(lexicon_score, 0L)
  ) |>
  arrange(feedback_id)

zero_match_feedback_ids <- lexicon_scores |>
  filter(matched_words == 0L) |>
  pull(feedback_id)

stopifnot(
  identical(nrow(bing), 6786L),
  identical(
    as.integer(count(bing, sentiment) |> arrange(sentiment) |> pull(n)),
    c(4781L, 2005L)
  ),
  identical(lexicon_scores$matched_words, c(1L, 1L, 0L, 0L, 0L)),
  identical(lexicon_scores$lexicon_score, c(1L, 1L, 0L, 0L, 0L)),
  identical(zero_match_feedback_ids, c(1003L, 1004L, 1005L))
)

knitr::kable(
  lexicon_scores |>
    select(feedback_id, rating, matched_words, lexicon_score, comment),
  col.names = c(
    "ID",
    "Rating",
    "Matched sentiment words",
    "Lexicon score",
    "Comment"
  ),
  caption = "Bing lexicon scores for the feedback rows",
  row.names = FALSE
)
```

In this five-row walkthrough, rows 1003, 1004, and 1005 contain no Bing word at
all. Treat those as inspectable examples, not as a rate. The English lexicon
misses the Spanish praise in row 1005. It also gives the late parcel complaint no
negative word.

## Measure lexicon silence on a larger corpus

To ask how often the lexicon has nothing to say, use a larger text source. The
shared inaugural helper returns 1,377 public-domain paragraphs after a 25-word
minimum-length filter. That filter matters because longer documents have more
chances to contain a word from any lexicon. The same Bing word list is applied to
each paragraph, and the code counts paragraphs with no positive or negative
match. This is also a domain-mismatch example: a product-review lexicon is being
applied to political oratory, so any score must be read as a warning sign, not as
validated sentiment.

```{r}
source("R/inaugural-corpus.R")
paragraphs <- inaugural_paragraphs() |>
  mutate(word_count = str_count(paragraph, "\\S+"))

inaugural_lexicon <- paragraphs |>
  select(paragraph_id, paragraph, word_count) |>
  tidytext::unnest_tokens(word, paragraph) |>
  semi_join(bing, by = join_by(word)) |>
  count(paragraph_id, name = "matched_words") |>
  right_join(
    paragraphs |> select(paragraph_id, word_count),
    by = join_by(paragraph_id)
  ) |>
  mutate(matched_words = replace_na(matched_words, 0L))

inaugural_silence <- inaugural_lexicon |>
  summarise(
    paragraphs = n(),
    zero_match_paragraphs = sum(matched_words == 0L),
    zero_match_share = mean(matched_words == 0L),
    median_matched_words = median(matched_words),
    .groups = "drop"
  )

coverage_by_length <- inaugural_lexicon |>
  mutate(
    length_bin = cut(
      word_count,
      breaks = c(0, 24, 49, 99, Inf),
      labels = c("1-24 words", "25-49 words", "50-99 words", "100+ words"),
      right = TRUE
    )
  ) |>
  summarise(
    paragraphs = n(),
    zero_match_paragraphs = sum(matched_words == 0L),
    zero_match_share = mean(matched_words == 0L),
    median_words = median(word_count),
    .by = length_bin
  ) |>
  arrange(length_bin)

stopifnot(
  identical(inaugural_silence$paragraphs, 1377L),
  identical(inaugural_silence$zero_match_paragraphs, 41L),
  identical(round(inaugural_silence$zero_match_share, 4), 0.0298),
  isTRUE(all.equal(as.numeric(inaugural_silence$median_matched_words), 6)),
  min(paragraphs$word_count) >= 25L,
  identical(as.character(coverage_by_length$length_bin), c("25-49 words", "50-99 words", "100+ words")),
  identical(coverage_by_length$paragraphs, c(441L, 497L, 439L)),
  identical(coverage_by_length$zero_match_paragraphs, c(33L, 8L, 0L)),
  identical(round(coverage_by_length$zero_match_share, 4), c(0.0748, 0.0161, 0))
)

knitr::kable(
  inaugural_silence |>
    mutate(zero_match_share = round(zero_match_share, 4)),
  col.names = c(
    "Paragraphs",
    "Paragraphs with no Bing match",
    "Share with no Bing match",
    "Median matched words"
  ),
  caption = "Bing lexicon coverage on inaugural paragraphs",
  row.names = FALSE
)

knitr::kable(
  coverage_by_length |>
    mutate(zero_match_share = round(zero_match_share, 4)),
  col.names = c(
    "Paragraph length bin",
    "Paragraphs",
    "Paragraphs with no Bing match",
    "Share with no Bing match",
    "Median words"
  ),
  caption = "Bing lexicon silence falls as paragraph length rises",
  row.names = FALSE
)
```

On the larger corpus, 41 of 1,377 paragraphs have no Bing match, a share of
0.0298 after the 25-word filter. The median paragraph has 6 matched words. The
length table gives the more honest finding: 7.48 percent of 25-49 word paragraphs
are silent, 1.61 percent of 50-99 word paragraphs are silent, and none of the 439
paragraphs with at least 100 words are silent. That explains why 3 of the 5 short
feedback comments are silent while long paragraphs almost never are.

## Read the failures before trusting the score

The lexicon sees words, not the situation around them. Negation means a word such
as `helpful` can be flipped by nearby `not`. Domain wording means a plain
customer-service problem, such as a password reset, may carry no sentiment word.
Sarcasm is harder still because the same surface words can be sincere or barbed.

```{r}
failure_cases <- lexicon_scores |>
  mutate(
    issue = case_when(
      feedback_id == 1003L ~ "account wording with no matched sentiment word",
      feedback_id == 1004L ~ "domain complaint with no matched word",
      feedback_id == 1005L ~ "non-English praise",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(issue)) |>
  select(feedback_id, rating, lexicon_score, issue, comment)

thank_tokens <- token_scores |>
  filter(feedback_id == 1001L, word == "thank") |>
  select(word, sentiment, token_score)

negation_examples <- tibble(
  phrase = c("helpful", "not helpful")
) |>
  tidytext::unnest_tokens(word, phrase, drop = FALSE) |>
  left_join(bing, by = join_by(word), multiple = "all") |>
  mutate(
    token_score = case_when(
      sentiment == "positive" ~ 1L,
      sentiment == "negative" ~ -1L,
      TRUE ~ 0L
    )
  ) |>
  summarise(
    matched_words = sum(token_score != 0L),
    lexicon_score = sum(token_score),
    .by = phrase
  )

stopifnot(
  identical(failure_cases$feedback_id, c(1003L, 1004L, 1005L)),
  identical(failure_cases$lexicon_score, c(0L, 0L, 0L)),
  identical(thank_tokens$sentiment, "positive"),
  identical(thank_tokens$token_score, 1L),
  identical(negation_examples$phrase, c("helpful", "not helpful")),
  identical(negation_examples$lexicon_score, c(1L, 1L))
)

knitr::kable(
  failure_cases,
  col.names = c("ID", "Rating", "Lexicon score", "Failure type", "Comment"),
  caption = "Rows where the word list misses the practical reading",
  row.names = FALSE
)

knitr::kable(
  negation_examples,
  col.names = c("Phrase", "Matched sentiment words", "Lexicon score"),
  caption = "A bag-of-words lexicon does not flip the score for negation",
  row.names = FALSE
)
```

Row 1003 is an account-access sentence with no matched sentiment word. The
separate `not helpful` example shows a real negation failure: it scores the same
as `helpful` because `helpful` is positive and `not` has no flipping rule here.
Row 1001 shows the opposite risk: `thank` is positive, but a word list cannot
tell from the word alone whether thanks are warm, routine, or sarcastic.

## Train on labels from the setting

A trained approach starts with labels from the setting that matters. Here the
rating field supplies three tiny labels for demonstration: ratings of 5 become
`satisfied`, rating 2 becomes `frustrated`, and rating 4 is left out because it
is mild. The seed before this deterministic fit is 4901.

```{r}
set.seed(4901)
labeled_feedback <- feedback |>
  filter(!is.na(rating), rating != 4L) |>
  mutate(
    sentiment_label = factor(
      if_else(rating >= 5L, "satisfied", "frustrated"),
      levels = c("frustrated", "satisfied")
    )
  )

training_tokens <- labeled_feedback |>
  select(feedback_id, sentiment_label, comment) |>
  tidytext::unnest_tokens(word, comment) |>
  count(sentiment_label, word, name = "label_word_count")

vocabulary <- training_tokens |>
  distinct(word) |>
  arrange(word)

label_totals <- training_tokens |>
  summarise(total_words = sum(label_word_count), .by = sentiment_label)

word_rates <- expand_grid(
  sentiment_label = levels(labeled_feedback$sentiment_label),
  word = vocabulary$word
) |>
  left_join(training_tokens, by = join_by(sentiment_label, word)) |>
  mutate(label_word_count = replace_na(label_word_count, 0L)) |>
  left_join(label_totals, by = join_by(sentiment_label)) |>
  mutate(
    word_rate = (label_word_count + 1) /
      (total_words + nrow(vocabulary))
  )

score_with_labels <- function(comment) {
  words <- tibble(comment = comment) |>
    tidytext::unnest_tokens(word, comment) |>
    semi_join(vocabulary, by = join_by(word))

  if (nrow(words) == 0L) {
    return(0)
  }

  joined <- words |>
    count(word, name = "n") |>
    left_join(word_rates, by = join_by(word), multiple = "all")

  totals <- joined |>
    summarise(
      log_score = sum(n * log(word_rate)),
      .by = sentiment_label
    )

  satisfied <- totals |>
    filter(sentiment_label == "satisfied") |>
    pull(log_score)
  frustrated <- totals |>
    filter(sentiment_label == "frustrated") |>
    pull(log_score)

  satisfied - frustrated
}

trained_scores <- feedback |>
  mutate(
    training_role = case_when(
      feedback_id %in% labeled_feedback$feedback_id ~ "fitting row",
      feedback_id == 1002L ~ "held-out rated row",
      TRUE ~ "unrated row"
    ),
    trained_score = map_dbl(comment, score_with_labels),
    trained_label = case_when(
      trained_score > 0 ~ "satisfied",
      trained_score < 0 ~ "frustrated",
      TRUE ~ "no seen training words"
    )
  ) |>
  select(feedback_id, rating, training_role, comment, trained_score, trained_label) |>
  left_join(
    lexicon_scores |> select(feedback_id, lexicon_score),
    by = join_by(feedback_id)
  ) |>
  relocate(lexicon_score, .before = trained_score)

stopifnot(
  identical(nrow(labeled_feedback), 3L),
  identical(
    as.character(labeled_feedback$sentiment_label),
    c("satisfied", "frustrated", "satisfied")
  ),
  identical(nrow(vocabulary), 15L),
  identical(
    trained_scores$training_role,
    c("fitting row", "held-out rated row", "unrated row", "fitting row", "fitting row")
  ),
  identical(
    trained_scores$trained_label,
    c(
      "satisfied",
      "frustrated",
      "no seen training words",
      "frustrated",
      "satisfied"
    )
  )
)

knitr::kable(
  trained_scores |>
    mutate(trained_score = round(trained_score, 3)),
  col.names = c(
    "ID",
    "Rating",
    "Training role",
    "Comment",
    "Lexicon score",
    "Rating-trained score",
    "Rating-trained label"
  ),
  caption = "A tiny rating-trained scorer compared with the lexicon",
  row.names = FALSE
)
```

This trained scorer learns from only three labelled rows, so it is a mechanics
example rather than an evaluation. Rows 1001, 1004, and 1005 are fitting rows,
and reporting success on them would be resubstitution. Row 1002 is the only rated
row held away from fitting, and this scorer sends it to `frustrated`. The table
therefore shows inspection output, not accuracy. A larger trained model can use
context if the training labels record the domain you care about. Its cost is that
labels must be collected, checked, and kept in step with the way people write.

## Emotion needs stronger evidence

Sentiment is usually a coarse positive-negative score. **Emotion analysis** asks
for labels such as anger, fear, sadness, joy, or surprise. That claim is much
stronger.

```{r}
#| label: fig-feedback-score-counts
#| fig-cap: "Only two Bing lexicon scores appear in the five fictional feedback comments."
#| fig-alt: "Bar chart showing three comments with lexicon score 0 and two comments with lexicon score 1."

score_range <- lexicon_scores |>
  summarise(
    lowest = min(lexicon_score),
    highest = max(lexicon_score),
    distinct_scores = n_distinct(lexicon_score),
    .groups = "drop"
  )

score_plot <- lexicon_scores |>
  count(lexicon_score, name = "comments")

stopifnot(
  identical(score_range$lowest, 0L),
  identical(score_range$highest, 1L),
  identical(score_range$distinct_scores, 2L),
  identical(score_plot$comments, c(3L, 2L))
)

knitr::kable(
  score_plot,
  col.names = c("Lexicon score", "Comments"),
  caption = "Only two lexicon scores appear in the feedback file",
  row.names = FALSE
)

ggplot(score_plot, aes(x = factor(lexicon_score), y = comments)) +
  geom_col(fill = "#4C78A8") +
  labs(
    x = "Lexicon score",
    y = "Comments"
  ) +
  theme_minimal()
```

A positive-word count is not an emotion. Self-reported emotion and observed
emotion can differ, and neither appears in this file. A score such as 0.62 would
be a number about words or a model output, not a claim about a person's inner
state.

## What to remember

- A sentiment lexicon counts words it recognizes.
- On the 25-word-filtered inaugural corpus, 41 of 1,377 paragraphs have no Bing
  match, and silence falls as paragraph length rises.
- Negation, language, domain wording, and sarcasm can break a word-count score.
- A trained approach needs labels from the setting it will serve.
- Emotion claims require evidence beyond positive and negative words.

## Sources

- [tidytext sentiment lexicons](https://juliasilge.github.io/tidytext/reference/sentiments.html)
- [Bing Liu opinion lexicon](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html),
  created for product-review sentiment research by Minqing Hu and Bing Liu; the
  source page requests citation and states research-use terms.
- [tidytext text mining book](https://www.tidytextmining.com/sentiment.html)
