---
title: "Turning documents into categories"
subtitle: "Follow the text-classification recipe on inaugural paragraphs"
author: "Alex Farach"
description: "Learn how a text classifier represents documents, fits a model, returns probabilities, and needs a cut point."
categories: [classification, text classification, inaugural corpus]
---

Elena has a box of labelled speeches and a simple question for the archive
search page. Could a paragraph's vocabulary help sort it into a broad time
period?

The page must be careful. A classifier can find differences in wording without
explaining history, policy, or people. This lesson treats the inaugural corpus as
a technical corpus for vocabulary practice and nothing more.

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- name the four steps in a text-classification recipe;
- compare token counts with term frequency-inverse document frequency;
- read a baseline beside a model score;
- inspect a probability distribution; and
- explain why a cut point is a separate choice.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the shared corpus

The inaugural addresses come bundled with quanteda and are works of the United
States government. The shared helper creates 1,377 paragraphs from 60 speeches.
readr opens the committed score files, dplyr and tibble shape tables, ggplot2
draws the probability chart, and tidymodels with textrecipes and glmnet fits the
models.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(tidymodels)
library(textrecipes)
library(glmnet)
library(quanteda)

source("R/inaugural-corpus.R")
paragraphs <- inaugural_paragraphs()

split_study <- read_csv(
  "data/inaugural/split-study.csv",
  na = "NA",
  col_types = cols(
    task = col_character(),
    split_scheme = col_character(),
    replicate = col_integer(),
    accuracy = col_double(),
    bal_accuracy = col_double(),
    train_majority_accuracy = col_double(),
    test_majority_rate = col_double(),
    length_rule_accuracy = col_double(),
    test_rows = col_integer(),
    test_speeches = col_integer()
  )
)

baselines <- read_csv(
  "data/inaugural/baselines.csv",
  na = character(),
  col_types = cols(
    task = col_character(),
    outcome = col_character(),
    rows = col_integer(),
    speeches = col_integer(),
    people = col_integer(),
    largest_class = col_character(),
    baseline_accuracy = col_double()
  )
)

era_counts <- paragraphs |>
  count(era, name = "paragraphs")

era_evidence <- split_study |>
  filter(task == "era", split_scheme == "speech") |>
  summarise(
    speech_split_accuracy = mean(accuracy),
    split_minimum = min(accuracy),
    split_maximum = max(accuracy),
    split_spread = split_maximum - split_minimum,
    deployable_baseline = mean(train_majority_accuracy),
    length_rule_accuracy = mean(length_rule_accuracy),
    replicates = n(),
    .groups = "drop"
  )

era_baseline_metadata <- baselines |>
  filter(task == "era")

stopifnot(
  all(c(
    "readr", "dplyr", "tibble", "tidyr", "purrr", "stringr",
    "ggplot2", "tidymodels", "textrecipes", "glmnet", "quanteda"
  ) %in% loadedNamespaces()),
  identical(nrow(paragraphs), 1377L),
  identical(era_counts$paragraphs, c(485L, 892L)),
  identical(n_distinct(paragraphs$speech_id), 60L),
  identical(n_distinct(paragraphs$president), 40L),
  identical(round(era_evidence$speech_split_accuracy, 4), 0.8457),
  identical(round(era_evidence$split_minimum, 4), 0.7970),
  identical(round(era_evidence$split_maximum, 4), 0.9039),
  identical(round(era_evidence$split_spread, 4), 0.1069),
  identical(round(era_evidence$deployable_baseline, 4), 0.6282),
  identical(round(era_evidence$length_rule_accuracy, 4), 0.6827),
  identical(era_baseline_metadata$people, 40L),
  all(c("era", "era_party_rows", "party") %in% unique(split_study$task))
)

knitr::kable(
  era_counts,
  col.names = c("Era label", "Paragraphs"),
  caption = "Paragraph labels created by the shared inaugural helper",
  row.names = FALSE
)

knitr::kable(
  era_evidence |>
    mutate(across(where(is.double), \(value) round(value, 4))),
  col.names = c(
    "Mean speech-level accuracy",
    "Minimum replicate",
    "Maximum replicate",
    "Replicate spread",
    "Train-majority baseline",
    "Paragraph-length rule",
    "Replicates"
  ),
  caption = "Committed evidence for the era classifier",
  row.names = FALSE
)
```

The committed study reports 0.8457 mean accuracy across 10 speech-level splits.
Those 10 replicate scores run from 0.7970 to 0.9039, so a single split can land
above or below the mean. The deployable train-majority baseline averages 0.6282:
it chooses the majority label from the training split, then applies that choice
to the test split. A paragraph-length-only rule averages 0.6827. The corpus has
60 speeches by 40 people. Lesson 44 covers evaluation methodology; this page
focuses on how representation changes the model's input.

## Represent the text

Text classification follows four steps: represent the text, fit a model, predict
a score, then decide what score becomes a label. A **feature** is an input column
for the model. Here the features are words after tokenization.

Two common representations are token counts and **term frequency-inverse document
frequency**, often shortened to **tf-idf**. A count records how often a word
appears. tf-idf gives less weight to words that appear in many documents.

```{r}
set.seed(5101)
speech_split <- group_initial_split(
  paragraphs,
  group = speech_id,
  prop = 0.75,
  strata = era
)
era_train <- training(speech_split)
era_test <- testing(speech_split)

fit_era_model <- function(use_tfidf = TRUE, max_tokens = 500L) {
  model_recipe <- recipe(era ~ paragraph, data = era_train) |>
    step_tokenize(paragraph) |>
    step_tokenfilter(paragraph, max_tokens = max_tokens)

  if (use_tfidf) {
    model_recipe <- model_recipe |>
      step_tfidf(paragraph)
  } else {
    model_recipe <- model_recipe |>
      step_tf(paragraph)
  }

  workflow() |>
    add_recipe(model_recipe) |>
    add_model(
      logistic_reg(penalty = 0.01, mixture = 1) |>
        set_engine("glmnet") |>
        set_mode("classification")
    ) |>
    fit(data = era_train)
}

set.seed(5102)
era_tfidf_fit <- fit_era_model(use_tfidf = TRUE, max_tokens = 500L)

set.seed(5103)
era_count_fit <- fit_era_model(use_tfidf = FALSE, max_tokens = 500L)

score_era_model <- function(fitted_workflow) {
  probability_table <- predict(fitted_workflow, era_test, type = "prob")
  late_column <- names(probability_table)[str_detect(names(probability_table), "later")]

  era_test |>
    select(paragraph_id, era) |>
    bind_cols(probability_table, predict(fitted_workflow, era_test)) |>
    mutate(prob_1900_or_later = .data[[late_column]])
}

tfidf_scores <- score_era_model(era_tfidf_fit)
count_scores <- score_era_model(era_count_fit)

feature_comparison <- tibble(
  representation = c("tf-idf, 500 tokens", "token counts, 500 tokens"),
  accuracy = c(
    yardstick::accuracy(tfidf_scores, truth = era, estimate = .pred_class) |>
      pull(.estimate),
    yardstick::accuracy(count_scores, truth = era, estimate = .pred_class) |>
      pull(.estimate)
  )
)

feature_gap <- abs(diff(feature_comparison$accuracy))

score_feature_split <- function(split_seed) {
  set.seed(split_seed)
  replicate_split <- group_initial_split(
    paragraphs,
    group = speech_id,
    prop = 0.75,
    strata = era
  )
  replicate_train <- training(replicate_split)
  replicate_test <- testing(replicate_split)

  fit_replicate_model <- function(use_tfidf, fit_seed) {
    replicate_recipe <- recipe(era ~ paragraph, data = replicate_train) |>
      step_tokenize(paragraph) |>
      step_tokenfilter(paragraph, max_tokens = 500)

    if (use_tfidf) {
      replicate_recipe <- replicate_recipe |>
        step_tfidf(paragraph)
    } else {
      replicate_recipe <- replicate_recipe |>
        step_tf(paragraph)
    }

    set.seed(fit_seed)
    workflow() |>
      add_recipe(replicate_recipe) |>
      add_model(
        logistic_reg(penalty = 0.01, mixture = 1) |>
          set_engine("glmnet") |>
          set_mode("classification")
      ) |>
      fit(data = replicate_train)
  }

  tfidf_fit <- fit_replicate_model(use_tfidf = TRUE, fit_seed = split_seed + 1000L)
  count_fit <- fit_replicate_model(use_tfidf = FALSE, fit_seed = split_seed + 2000L)

  bind_rows(
    tibble(
      representation = "tf-idf, 500 tokens",
      accuracy = yardstick::accuracy(
        bind_cols(replicate_test |> select(era), predict(tfidf_fit, replicate_test)),
        truth = era,
        estimate = .pred_class
      ) |>
        pull(.estimate)
    ),
    tibble(
      representation = "token counts, 500 tokens",
      accuracy = yardstick::accuracy(
        bind_cols(replicate_test |> select(era), predict(count_fit, replicate_test)),
        truth = era,
        estimate = .pred_class
      ) |>
        pull(.estimate)
    )
  ) |>
    mutate(split_seed = split_seed, test_rows = nrow(replicate_test))
}

feature_replicates <- bind_rows(
  feature_comparison |>
    mutate(split_seed = 5101L, test_rows = nrow(era_test)),
  map(5102:5105, score_feature_split) |>
    list_rbind()
) |>
  relocate(split_seed)

feature_summary <- feature_replicates |>
  summarise(
    mean_accuracy = mean(accuracy),
    minimum = min(accuracy),
    maximum = max(accuracy),
    spread = maximum - minimum,
    .by = representation
  )

feature_gaps <- feature_replicates |>
  select(split_seed, representation, accuracy) |>
  pivot_wider(names_from = representation, values_from = accuracy) |>
  mutate(gap = abs(`token counts, 500 tokens` - `tf-idf, 500 tokens`))

test_counts <- era_test |>
  count(era, name = "paragraphs")

stopifnot(
  identical(nrow(era_train), 1044L),
  identical(nrow(era_test), 333L),
  identical(test_counts$paragraphs, c(111L, 222L)),
  identical(round(feature_comparison$accuracy, 4), c(0.8799, 0.8949)),
  identical(round(feature_gap, 4), 0.0150),
  identical(round(feature_replicates$accuracy, 4), c(
    0.8799, 0.8949, 0.7632, 0.7690, 0.8764,
    0.8966, 0.8347, 0.8320, 0.8365, 0.8459
  )),
  isTRUE(all.equal(round(feature_summary$spread, 4), c(0.1167, 0.1275))),
  max(feature_gaps$gap) < min(feature_summary$spread)
)

knitr::kable(
  feature_comparison |>
    mutate(accuracy = round(accuracy, 4)),
  col.names = c("Representation", "Accuracy on this split"),
  caption = "Changing features changes the result on the same speech-level split",
  row.names = FALSE
)

knitr::kable(
  feature_summary |>
    mutate(across(c(mean_accuracy, minimum, maximum, spread), \(value) round(value, 4))),
  col.names = c("Representation", "Mean accuracy", "Minimum", "Maximum", "Spread"),
  caption = "Five same-configuration speech-level splits for each representation",
  row.names = FALSE
)
```

On this split, token counts score 0.8949 and tf-idf scores 0.8799. The gap is
0.0150. With the same model settings across five speech-level splits, tf-idf
ranges from 0.7632 to 0.8799, and token counts range from 0.7690 to 0.8966. The
displayed gap is inside those same-configuration spreads, so this lesson is not
claiming that token counts beat tf-idf. The point is that the model does not read
paragraphs directly; it reads the feature table we give it.

A linear model over word features also loses word order. It can learn that a word
is common in one label and rare in another, but it cannot tell whether two words
appeared beside each other unless we create features for that pattern.

## Inspect the probability scores

The model returns probabilities. A probability is still not a final action. The
usual 0.50 cut point is a convention, and a project can choose a different point
if the two errors have different costs.

```{r}
#| label: fig-era-probabilities
#| fig-cap: "Predicted probabilities for the '1900 or later' era label on the speech-level test split."
#| fig-alt: "Histogram of predicted probabilities. Before-1900 paragraphs cluster mostly near 0, while 1900-or-later paragraphs cluster mostly near 1 with overlap between the groups."

probability_summary <- tfidf_scores |>
  summarise(
    minimum = min(prob_1900_or_later),
    tenth = quantile(prob_1900_or_later, 0.10),
    median = median(prob_1900_or_later),
    ninetieth = quantile(prob_1900_or_later, 0.90),
    maximum = max(prob_1900_or_later),
    .groups = "drop"
  )

stopifnot(
  isTRUE(all.equal(round(as.numeric(probability_summary$minimum), 4), 0.0240)),
  isTRUE(all.equal(round(as.numeric(probability_summary$tenth), 4), 0.1394)),
  isTRUE(all.equal(round(as.numeric(probability_summary$median), 4), 0.8629)),
  isTRUE(all.equal(round(as.numeric(probability_summary$ninetieth), 4), 0.9928)),
  isTRUE(all.equal(round(as.numeric(probability_summary$maximum), 4), 1.0000))
)

knitr::kable(
  probability_summary |>
    mutate(across(everything(), \(value) round(value, 4))),
  col.names = c("Minimum", "10th percentile", "Median", "90th percentile", "Maximum"),
  caption = "Distribution of predicted probabilities for the later era label",
  row.names = FALSE
)

ggplot(tfidf_scores, aes(x = prob_1900_or_later, fill = era)) +
  geom_histogram(binwidth = 0.05, boundary = 0, color = "white") +
  labs(
    x = "Predicted probability for 1900 or later",
    y = "Paragraphs",
    fill = "Recorded era"
  ) +
  theme_minimal()
```

Many scores sit near 0 or 1, and the median is 0.8629 because the later era is
the larger class in this test split. Class imbalance matters: the model sees
more examples from the larger class, and a naive baseline already gets many rows
right by guessing that class.

## Choose the cut point

Changing the cut point changes which side absorbs the doubtful cases. This is a
decision about how the classifier will be used, not something the fitted model
settles alone.

```{r}
era_thresholds <- tibble(threshold = c(0.40, 0.50, 0.60)) |>
  mutate(
    results = map(threshold, \(cut_point) {
      tfidf_scores |>
        summarise(
          before_1900_sent_late =
            sum(prob_1900_or_later >= cut_point & era == "before 1900"),
          late_sent_before =
            sum(prob_1900_or_later < cut_point & era == "1900 or later"),
          .groups = "drop"
        )
    })
  ) |>
  unnest(results)

stopifnot(
  identical(era_thresholds$before_1900_sent_late, c(31L, 28L, 17L)),
  identical(era_thresholds$late_sent_before, c(7L, 12L, 22L))
)

knitr::kable(
  era_thresholds,
  col.names = c(
    "Cut point",
    "Earlier paragraphs sent to later label",
    "Later paragraphs sent to earlier label"
  ),
  caption = "Cut-point trade-off for the tf-idf model",
  row.names = FALSE
)
```

At 0.40, seven later-era test paragraphs fall below the cut point. At 0.60, that
count rises to 22 while fewer earlier paragraphs move to the later label. A
classifier returns a score; a person chooses the cut point that matches the
task.

## What to remember

- Text classification needs labelled documents and a fixed set of categories.
- The representation step decides which text evidence the model can see.
- A baseline gives the model score a reference point.
- Class imbalance can make easy guesses look better than they are.
- Word-count models ignore order unless order is built into the features.

## Sources

- [quanteda inaugural corpus documentation](https://quanteda.io/reference/data_corpus_inaugural.html)
- [glmnet package documentation](https://glmnet.stanford.edu/articles/glmnet.html)
- [textrecipes package documentation](https://textrecipes.tidymodels.org/)
