---
title: "One label or several labels"
subtitle: "Separate multi-class choices from multi-label tags"
author: "Alex Farach"
description: "Learn the difference between multi-class and multi-label text classification using inaugural paragraphs."
categories: [classification, multi-class classification, multi-label classification]
---

Ravi is building filters for an archive search page. One control asks for a time
period, where each paragraph can belong to only one period. Another asks for
themes, where a paragraph can mention work, time, people, all three, or none.

Those two controls look similar on the screen, but they ask different questions.
One forces a single choice. The other lets several tags sit on the same
paragraph.

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- distinguish multi-class from multi-label classification;
- build a four-way period label from years;
- read a confusion matrix for a multi-class model;
- create theme labels with a small quanteda dictionary; and
- explain why label-level accuracy can mislead in multi-label work.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load paragraphs and build periods

**Multi-class classification** means one document receives one label from several
mutually exclusive choices. The four periods below are equal-width year bins:
1789-1847, 1848-1906, 1907-1965, and 1966-2025. They are technical bins for this
lesson, not claims about political history.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(ggplot2)
library(tidymodels)
library(textrecipes)
library(glmnet)
library(quanteda)

source("R/inaugural-corpus.R")
paragraphs <- inaugural_paragraphs()

periods <- paragraphs |>
  mutate(
    period = factor(
      cut(
        year,
        breaks = c(1788, 1847, 1906, 1965, 2025),
        labels = c("1789-1847", "1848-1906", "1907-1965", "1966-2025"),
        right = TRUE
      ),
      levels = c("1789-1847", "1848-1906", "1907-1965", "1966-2025")
    )
  )

period_counts <- periods |>
  count(period, name = "paragraphs")

knitr::kable(
  period_counts,
  col.names = c("Period label", "Paragraphs"),
  caption = "Four mutually exclusive period labels",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "readr", "dplyr", "tibble", "tidyr", "purrr", "stringr",
    "ggplot2", "tidymodels", "textrecipes", "glmnet", "quanteda"
  ) %in% loadedNamespaces()),
  identical(nrow(periods), 1377L),
  identical(as.character(period_counts$period), levels(periods$period)),
  identical(period_counts$paragraphs, c(210L, 291L, 423L, 453L)),
  identical(sum(period_counts$paragraphs), 1377L)
)
```

Each paragraph receives exactly one period label. That single-label structure is
what makes the task multi-class rather than multi-label.

## Fit the four-way classifier

The split keeps whole speeches together. The seed for the split is 5201, and the
seed before fitting is 5202. The model uses tf-idf features and a multinomial
glmnet classifier.

```{r}
set.seed(5201)
period_split <- group_initial_split(
  periods,
  group = speech_id,
  prop = 0.75,
  strata = period
)
period_train <- training(period_split)
period_test <- testing(period_split)

period_recipe <- recipe(period ~ paragraph, data = period_train) |>
  step_tokenize(paragraph) |>
  step_tokenfilter(paragraph, max_tokens = 600) |>
  step_tfidf(paragraph)

period_model <- multinom_reg(penalty = 0.01, mixture = 1) |>
  set_engine("glmnet") |>
  set_mode("classification")

set.seed(5202)
period_fit <- workflow() |>
  add_recipe(period_recipe) |>
  add_model(period_model) |>
  fit(data = period_train)

period_predictions <- period_test |>
  select(paragraph_id, period) |>
  bind_cols(predict(period_fit, period_test))

period_accuracy <- yardstick::accuracy(
  period_predictions,
  truth = period,
  estimate = .pred_class
) |>
  pull(.estimate)

period_confusion <- yardstick::conf_mat(
  period_predictions,
  truth = period,
  estimate = .pred_class
)
period_matrix <- as.matrix(period_confusion$table)

test_period_counts <- period_test |>
  count(period, name = "paragraphs")

period_baseline <- max(test_period_counts$paragraphs) / nrow(period_test)

period_scores <- tibble(
  measure = c("Model accuracy", "Largest-class baseline"),
  value = c(period_accuracy, period_baseline)
)

knitr::kable(
  period_scores |>
    mutate(value = round(value, 4)),
  col.names = c("Measure", "Value"),
  caption = "Four-way model accuracy beside the largest-class baseline",
  row.names = FALSE
)

knitr::kable(
  as.data.frame(period_confusion$table),
  col.names = c("Model prediction", "Recorded period", "Paragraphs"),
  caption = "Confusion matrix for the four-way period classifier",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(period_train), 1025L),
  identical(nrow(period_test), 352L),
  identical(test_period_counts$paragraphs, c(48L, 88L, 99L, 117L)),
  identical(round(period_accuracy, 4), 0.5938),
  identical(round(period_baseline, 4), 0.3324),
  all(
    period_matrix ==
    matrix(
      c(
        19L, 13L, 14L, 3L,
        23L, 63L, 30L, 2L,
        5L, 9L, 47L, 32L,
        1L, 3L, 8L, 80L
      ),
      nrow = 4,
      byrow = TRUE,
      dimnames = dimnames(period_matrix)
    )
  )
)
```

The overall accuracy is 0.5938 on this speech-level split, compared with a
0.3324 largest-class baseline from always predicting the most common test period.
That single model number still hides which periods are hard. The confusion
matrix shows, for example, that many 1907-1965 test paragraphs are sent to the
1966-2025 label.

## Check recall by class

**Recall** asks, within one recorded class, how many rows returned to that class.
For a multi-class task, recall by class is often more useful than one accuracy
number.

```{r}
period_recall <- period_predictions |>
  summarise(
    support = n(),
    correct = sum(.pred_class == period),
    recall = correct / support,
    .by = period
  ) |>
  arrange(period)

knitr::kable(
  period_recall |>
    mutate(recall = round(recall, 3)),
  col.names = c("Recorded period", "Test paragraphs", "Correct", "Recall"),
  caption = "Recall varies by period label",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(period_recall$support, c(48L, 88L, 99L, 117L)),
  identical(period_recall$correct, c(19L, 63L, 47L, 80L)),
  identical(round(period_recall$recall, 3), c(0.396, 0.716, 0.475, 0.684))
)
```

The 1789-1847 period has recall of 0.396, while the 1848-1906 period has recall
of 0.716. The model's mistakes are uneven, so the confusion matrix is required.

## Build several theme labels

**Multi-label classification** means a document may receive zero, one, or several
labels at the same time. A small quanteda dictionary marks three vocabulary
themes. These labels are the dictionary's own output. Agreement with them
measures agreement with the dictionary, not agreement with human themes or with
the corpus itself.

```{r}
theme_dictionary <- dictionary(list(
  work = c("work*", "labor", "employment", "industry", "business"),
  time = c("year*", "day*", "time*", "today", "future"),
  people = c("people", "citizen*", "family", "families", "children")
))

theme_dfm <- corpus(paragraphs, text_field = "paragraph") |>
  tokens(remove_punct = TRUE) |>
  tokens_tolower() |>
  dfm() |>
  dfm_lookup(dictionary = theme_dictionary)

theme_counts <- convert(theme_dfm, to = "data.frame") |>
  as_tibble() |>
  select(-doc_id)

theme_labels <- bind_cols(
  paragraphs |> select(paragraph_id),
  theme_counts > 0
)

label_count_table <- theme_labels |>
  mutate(label_count = rowSums(across(-paragraph_id))) |>
  count(label_count, name = "paragraphs")

knitr::kable(
  label_count_table,
  col.names = c("Theme labels on a paragraph", "Paragraphs"),
  caption = "Paragraphs can carry zero, one, or several dictionary labels",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(as.integer(colSums(theme_counts > 0)), c(219L, 501L, 581L)),
  identical(label_count_table$label_count, c(0, 1, 2, 3)),
  identical(label_count_table$paragraphs, c(449L, 605L, 273L, 50L)),
  identical(sum(label_count_table$paragraphs), 1377L)
)
```

The memorable distinction is this: multi-class chooses one box from several;
multi-label lets the same document carry several tags.

## Accuracy can reward saying no

Most possible theme labels are absent. A do-nothing system that predicts every
label as absent therefore gets many cells right while finding no present labels.
That is why multi-label reporting needs per-label precision and recall.

```{r}
label_long <- theme_labels |>
  pivot_longer(
    cols = -paragraph_id,
    names_to = "label",
    values_to = "present"
  )

always_absent <- label_long |>
  mutate(predicted = FALSE)

per_label_metrics <- always_absent |>
  summarise(
    true_present = sum(present),
    true_absent = sum(!present),
    prevalence = mean(present),
    predicted_present = sum(predicted),
    true_positive = sum(present & predicted),
    false_positive = sum(!present & predicted),
    false_negative = sum(present & !predicted),
    accuracy = mean(predicted == present),
    precision = if_else(
      true_positive + false_positive == 0L,
      NA_real_,
      true_positive / (true_positive + false_positive)
    ),
    recall = if_else(
      true_positive + false_negative == 0L,
      NA_real_,
      true_positive / (true_positive + false_negative)
    ),
    .by = label
  ) |>
  arrange(label)

overall_label_accuracy <- mean(always_absent$predicted == always_absent$present)
exact_empty_accuracy <- theme_labels |>
  mutate(
    label_count = rowSums(across(-paragraph_id)),
    predicted_empty_matches = label_count == 0L
  ) |>
  summarise(value = mean(predicted_empty_matches), .groups = "drop") |>
  pull(value)

knitr::kable(
  per_label_metrics |>
    mutate(
      accuracy = round(accuracy, 3),
      prevalence = round(prevalence, 3),
      precision = if_else(is.na(precision), NA_real_, round(precision, 3)),
      recall = round(recall, 3)
    ) |>
    select(label, true_present, true_absent, prevalence, accuracy, precision, recall),
  col.names = c(
    "Label",
    "Present",
    "Absent",
    "Prevalence",
    "Accuracy if always absent",
    "Precision",
    "Recall"
  ),
  caption = "Per-label metrics expose the no-label baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(per_label_metrics$true_present, c(581L, 501L, 219L)),
  identical(per_label_metrics$true_absent, c(796L, 876L, 1158L)),
  identical(round(per_label_metrics$prevalence, 3), c(0.422, 0.364, 0.159)),
  identical(per_label_metrics$predicted_present, c(0L, 0L, 0L)),
  all(is.na(per_label_metrics$precision)),
  identical(per_label_metrics$recall, c(0, 0, 0)),
  identical(round(overall_label_accuracy, 4), 0.6851),
  identical(round(exact_empty_accuracy, 4), 0.3261)
)
```

The always-absent baseline gets 0.6851 label-cell accuracy because absent labels
are common. The label prevalences are 0.422 for people, 0.364 for time, and 0.159
for work. Its recall is 0 for every label, so it finds none of the 219 work, 501
time, or 581 people dictionary labels. Exact empty-set accuracy is 0.3261, the
share of paragraphs with no dictionary theme at all.

## What to remember

- Multi-class means one label from more than two choices.
- Multi-label means zero, one, or several labels can apply at once.
- Multi-class accuracy can hide weak classes.
- Multi-label accuracy can be high when absent labels dominate.
- Per-class or per-label precision and recall tell the reader where the errors
  are.

## Sources

- [quanteda dictionary documentation](https://quanteda.io/reference/dictionary.html)
- [yardstick multiclass metrics](https://yardstick.tidymodels.org/reference/accuracy.html)
- [glmnet multinomial models](https://glmnet.stanford.edu/articles/glmnet.html)
