---
title: "Predicting the next token"
subtitle: "Count context, choose on validation data, and test once"
description: "Learn how a language model assigns probabilities to the next token and how to evaluate an inspectable count model."
categories: [natural language generation, next-token prediction, language models]
---

Nia is building an autocomplete demonstration for an archive. After a reader
types `we`, the tool must decide which token could come next.

A **language model** assigns probabilities to token sequences. In
**next-token prediction**, the input is the context already seen and the output
is a probability for each possible next token. This is one language-model
training objective. It is not a definition of every language model or every
text-generation system.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- keep training, validation, and test speeches separate;
- build next-token rows without crossing paragraph boundaries;
- estimate unigram, bigram, and trigram count models;
- compare greedy choice with probability-weighted sampling;
- select context length and interpolation strength with validation perplexity; and
- explain what held-out perplexity does not measure.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the split and study results

The inaugural corpus contains 60 public-domain speeches. The committed study
uses 37 speeches for training, 10 for model selection, and 13 for the final
test. A speech belongs to one split only. The split is stratified by era so both
eras appear in every part. The split holds out speeches, not speakers:
presidents with more than one inaugural address can appear in multiple splits.

```{r}
library(digest)
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(tidytext)
library(purrr)
library(readr)
library(knitr)

source("R/inaugural-corpus.R")

split_path <- "data/inaugural/next-token-splits.csv"
validation_path <- "data/inaugural/next-token-validation.csv"
test_path <- "data/inaugural/next-token-test.csv"
metadata_path <- "data/inaugural/next-token-study-metadata.csv"

split_assignments <- read_csv(
  split_path,
  na = character(),
  col_types = cols(
    speech_id = col_character(),
    era = col_character(),
    split = col_character()
  )
)
validation_results <- read_csv(
  validation_path,
  na = c("", "NA"),
  col_types = cols(
    split = col_character(),
    model = col_character(),
    context_tokens = col_integer(),
    interpolation_strength = col_double(),
    next_token_rows = col_integer(),
    oov_target_share = col_double(),
    seen_context_share = col_double(),
    mean_log_loss = col_double(),
    perplexity = col_double(),
    selected = col_logical(),
    selection_reason = col_character()
  )
)
test_results <- read_csv(
  test_path,
  na = c("", "NA"),
  col_types = cols(
    split = col_character(),
    model = col_character(),
    context_tokens = col_integer(),
    interpolation_strength = col_double(),
    next_token_rows = col_integer(),
    oov_target_share = col_double(),
    seen_context_share = col_double(),
    mean_log_loss = col_double(),
    perplexity = col_double(),
    top_1_accuracy = col_double(),
    role = col_character()
  )
)
test_by_speech <- read_csv(
  "data/inaugural/next-token-test-by-speech.csv",
  na = character(),
  col_types = cols(
    speech_id = col_character(),
    next_token_rows = col_integer(),
    model = col_character(),
    perplexity = col_double()
  )
)
study_metadata <- read_csv(
  metadata_path,
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    settings = col_character(),
    source = col_character(),
    license = col_character(),
    built_on = col_character(),
    fingerprint = col_character()
  )
)

split_counts <- split_assignments |>
  mutate(
    split = factor(
      split,
      levels = c("training", "validation", "test")
    ),
    era = factor(
      era,
      levels = c("before 1900", "1900 or later")
    )
  ) |>
  count(split, era, name = "speeches") |>
  arrange(split, era)

kable(
  split_counts,
  col.names = c("Split", "Era", "Speeches"),
  caption = "Speech-level assignments used for next-token modeling",
  row.names = FALSE
)
```

```{r}
#| include: false

hash_lines <- function(path) {
  digest(
    paste(read_lines(path), collapse = "\n"),
    algo = "sha256",
    serialize = FALSE
  )
}

artifact_paths <- file.path(
  "data/inaugural",
  study_metadata$artifact
)

stopifnot(
  all(c(
    "digest", "dplyr", "tidyr", "tibble", "stringr",
    "tidytext", "purrr", "readr", "knitr"
  ) %in% loadedNamespaces()),
  identical(nrow(split_assignments), 60L),
  identical(nrow(validation_results), 13L),
  identical(nrow(test_results), 3L),
  identical(nrow(test_by_speech), 26L),
  identical(split_counts$speeches, c(17L, 20L, 5L, 5L, 6L, 7L)),
  !anyDuplicated(split_assignments$speech_id),
  all(file.exists(artifact_paths)),
  identical(map_chr(artifact_paths, hash_lines), study_metadata$fingerprint),
  identical(
    validation_results$model[validation_results$selected],
    "trigram"
  ),
  identical(
    validation_results$interpolation_strength[validation_results$selected],
    100
  )
)
```

The validation set is used to choose among candidate models. The test speeches
stay closed until that choice is fixed. Calling both sets "held out" would hide
their different jobs.

## Build paragraph-bounded next-token rows

The study uses lowercase ASCII alphabetic tokens. It keeps the 2,499 most common
training tokens and maps every other token to `<unk>`, giving a 2,500-token
model vocabulary.

Each prediction row keeps its paragraph and speech ID. `previous_1` is the token
immediately before the target. `previous_2` is one more step back. The first two
tokens in each paragraph have no full two-token context, so they are not
evaluation rows. No transition joins the end of one paragraph to the beginning
of another, and this model does not learn how to start a paragraph.

```{r}
paragraphs <- inaugural_paragraphs()

training_tokens <- paragraphs |>
  select(paragraph_id, speech_id, paragraph) |>
  inner_join(
    split_assignments |>
      filter(split == "training") |>
      select(speech_id),
    by = join_by(speech_id)
  ) |>
  unnest_tokens(token, paragraph) |>
  filter(str_detect(token, "^[a-z]+$")) |>
  mutate(position = row_number(), .by = paragraph_id)

training_vocabulary <- training_tokens |>
  count(token, sort = TRUE) |>
  slice_head(n = 2499L) |>
  pull(token)
model_vocabulary <- c(training_vocabulary, "<unk>")

training_next_tokens <- training_tokens |>
  mutate(
    model_token = if_else(
      token %in% training_vocabulary,
      token,
      "<unk>"
    )
  ) |>
  arrange(paragraph_id, position) |>
  mutate(
    previous_1 = lag(model_token, 1L),
    previous_2 = lag(model_token, 2L),
    .by = paragraph_id
  ) |>
  filter(position >= 3L) |>
  transmute(
    paragraph_id,
    speech_id,
    previous_2,
    previous_1,
    next_token = model_token
  )

training_summary <- tibble(
  measure = c(
    "training speeches",
    "training paragraphs",
    "training tokens",
    "next-token rows",
    "model vocabulary"
  ),
  value = c(
    n_distinct(training_next_tokens$speech_id),
    n_distinct(training_tokens$paragraph_id),
    nrow(training_tokens),
    nrow(training_next_tokens),
    length(model_vocabulary)
  )
)

kable(
  training_summary,
  col.names = c("Training item", "Value"),
  caption = "Training data for the inspectable count model",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    training_summary$value,
    c(37L, 867L, 83458L, 81724L, 2500L)
  ),
  all(!is.na(training_next_tokens$paragraph_id)),
  all(!is.na(training_next_tokens$speech_id)),
  all(training_next_tokens$paragraph_id %in% paragraphs$paragraph_id)
)
```

## Count what followed each token

A **unigram** ignores context and uses overall token frequency. A **bigram**
uses one previous token. A **trigram** uses two. More context can help, but each
longer context appears less often.

Raw counts assign probability zero to a next token that never followed the
context in training. The study avoids zero probabilities through
**interpolation**: every estimate mixes the longer-context counts with the
shorter model. The **interpolation strength** controls how much weight the
shorter model receives, so a larger value leans more on that model. Six values
are compared on validation speeches; the selected trigram uses 100.

```{r}
selected_validation <- validation_results |>
  filter(selected)
interpolation_strength <- selected_validation$interpolation_strength

unigram_counts <- training_tokens |>
  mutate(
    model_token = if_else(
      token %in% training_vocabulary,
      token,
      "<unk>"
    )
  ) |>
  count(model_token, name = "unigram_count")
training_token_total <- sum(unigram_counts$unigram_count)

unigram_probabilities <- tibble(
  model_token = model_vocabulary
) |>
  left_join(unigram_counts, by = join_by(model_token)) |>
  mutate(
    unigram_count = replace_na(unigram_count, 0L),
    unigram_probability = (unigram_count + 1) /
      (training_token_total + length(model_vocabulary))
  )

bigram_counts <- training_next_tokens |>
  count(previous_1, next_token, name = "bigram_count")
trigram_counts <- training_next_tokens |>
  count(
    previous_2,
    previous_1,
    next_token,
    name = "trigram_count"
  )

predict_after_one <- function(previous_token) {
  context_counts <- bigram_counts |>
    filter(previous_1 == previous_token)

  tibble(next_token = model_vocabulary) |>
    left_join(
      unigram_probabilities,
      by = join_by(next_token == model_token)
    ) |>
    left_join(context_counts, by = join_by(next_token)) |>
    mutate(
      bigram_count = replace_na(bigram_count, 0L),
      context_count = sum(bigram_count),
      bigram_probability = (
        bigram_count +
          interpolation_strength * unigram_probability
      ) / (
        context_count +
          interpolation_strength
      )
    ) |>
    arrange(desc(bigram_probability), next_token)
}

predict_after_two <- function(previous_2_token, previous_1_token) {
  context_counts <- trigram_counts |>
    filter(
      previous_2 == previous_2_token,
      previous_1 == previous_1_token
    )

  predict_after_one(previous_1_token) |>
    left_join(context_counts, by = join_by(next_token)) |>
    mutate(
      trigram_count = replace_na(trigram_count, 0L),
      context_count = sum(trigram_count),
      probability = (
        trigram_count +
          interpolation_strength * bigram_probability
      ) / (
        context_count +
          interpolation_strength
      ),
      context = str_c(
        previous_2_token,
        previous_1_token,
        sep = " "
      )
    ) |>
    arrange(desc(probability), next_token)
}

context_predictions <- bind_rows(
  predict_after_two("we", "must") |>
    slice_head(n = 5),
  predict_after_two("the", "united") |>
    slice_head(n = 5)
) |>
  select(context, next_token, trigram_count, probability)

kable(
  context_predictions |>
    mutate(probability = round(probability, 4)),
  col.names = c("Previous two tokens", "Possible next token", "Training count", "Probability"),
  caption = "Highest-probability next tokens from the selected trigram",
  row.names = FALSE
)
```

```{r}
#| include: false

we_must_distribution <- predict_after_two("we", "must")
the_united_distribution <- predict_after_two("the", "united")

stopifnot(
  identical(selected_validation$model, "trigram"),
  identical(interpolation_strength, 100),
  identical(
    context_predictions$next_token[
      context_predictions$context == "we must"
    ],
    c("be", "<unk>", "not", "do", "show")
  ),
  identical(
    context_predictions$next_token[
      context_predictions$context == "the united"
    ],
    c("states", "nations", "<unk>", "and", "the")
  ),
  abs(sum(we_must_distribution$probability) - 1) < 1e-12,
  abs(sum(the_united_distribution$probability) - 1) < 1e-12,
  identical(
    round(
      context_predictions$probability[
        context_predictions$context == "the united" &
          context_predictions$next_token == "states"
      ],
      3
    ),
    0.719
  )
)
```

After `the united`, the model assigns about 0.72 probability to `states`. That
reflects a repeated phrase in US inaugural addresses. It does not show a general
English rule. After `we must`, several continuations remain plausible.

The `<unk>` row stands for any token outside the training vocabulary. Predicting
that bucket does not identify the original word.

## Choose or sample a next token

The probability table does not decide how text is generated. **Greedy decoding**
takes the highest-probability token. **Sampling** draws according to the full
distribution. The seed below makes five draws reproducible.

```{r}
greedy_next <- we_must_distribution$next_token[[1]]

set.seed(6303)
sampled_next <- sample(
  we_must_distribution$next_token,
  size = 5,
  replace = TRUE,
  prob = we_must_distribution$probability
)

choice_table <- tibble(
  method = c("greedy", "five seeded samples"),
  output = c(
    greedy_next,
    str_c(sampled_next, collapse = ", ")
  )
)

kable(
  choice_table,
  col.names = c("Choice rule", "Next-token output after 'we must'"),
  caption = "One probability distribution can support different choice rules",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(greedy_next, "be"),
  identical(length(sampled_next), 5L),
  all(sampled_next %in% model_vocabulary),
  identical(choice_table$method, c("greedy", "five seeded samples"))
)
```

Greedy choice repeats the most common continuation. Sampling can select less
common tokens and will change when the seed changes. Neither rule checks whether
a continuation is factual, useful, or appropriate for a reader.

## Select context length and smoothing before testing

The builder compares six interpolation strengths for both the bigram and trigram, plus
the unigram baseline, on validation speeches.
**Log loss** penalizes probability assigned away from the observed next token.
**Perplexity** is the exponentiated mean log loss; lower is better when the
tokenizer, vocabulary, and evaluation rows are identical.

```{r}
validation_display <- validation_results |>
  filter(
    model == "unigram" |
      (model == "bigram" & interpolation_strength == 100) |
      (
        model == "trigram" &
          interpolation_strength %in% c(10, 100, 1000)
      )
  ) |>
  arrange(context_tokens, interpolation_strength) |>
  transmute(
    model,
    context_tokens,
    interpolation_strength = if_else(
      is.na(interpolation_strength),
      "not applicable",
      as.character(interpolation_strength)
    ),
    seen_context = if_else(
      is.na(seen_context_share),
      "not applicable",
      sprintf("%.1f%%", 100 * seen_context_share)
    ),
    perplexity = round(perplexity, 1),
    selected = if_else(selected, "yes", "no")
  )

kable(
  validation_display,
  col.names = c(
    "Candidate",
    "Context tokens",
    "Interpolation strength",
    "Validation contexts seen in training",
    "Validation perplexity",
    "Selected"
  ),
  caption = "Validation settings around the selected trigram",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(validation_results), 13L),
  identical(
    validation_display$model,
    c("unigram", "bigram", "trigram", "trigram", "trigram")
  ),
  identical(
    validation_display$interpolation_strength,
    c("not applicable", "100", "10", "100", "1000")
  ),
  identical(
    validation_display$perplexity,
    c(238.0, 134.8, 194.4, 133.3, 143.6)
  ),
  identical(
    validation_display$selected,
    c("no", "no", "no", "yes", "no")
  ),
  identical(
    round(
      validation_results$seen_context_share[
        validation_results$model == "trigram" &
          validation_results$interpolation_strength == 100
      ],
      3
    ),
    0.749
  ),
  identical(
    round(validation_results$oov_target_share[[1]], 3),
    0.117
  )
)
```

The trigram with interpolation strength 100 has the lowest validation perplexity,
133.3. The best bigram is close at 134.8. The tested strengths on both sides of
100 score worse, so the selected value is not a grid boundary. Only 74.9 percent
of the trigram's two-token validation contexts appeared in training. When a
context is unseen, its estimate relies entirely on the bigram and then the
unigram. The bigram's 100 percent seen-context rate follows from mapping every
token into the training vocabulary, including `<unk>`; it is not evidence that
the model generalizes to every word.

## Open the test speeches once

After selecting the trigram and its interpolation strength, the study scores it,
the predeclared unigram baseline, and a uniform 2,500-token floor on the 13 test
speeches.

```{r}
test_display <- test_results |>
  transmute(
    model,
    role,
    next_token_rows,
    oov_targets = sprintf("%.1f%%", 100 * oov_target_share),
    perplexity = round(perplexity, 1),
    top_1_accuracy = if_else(
      is.na(top_1_accuracy),
      "not applicable",
      sprintf("%.1f%%", 100 * top_1_accuracy)
    )
  )

perplexity_reduction <- 1 -
  test_results$perplexity[test_results$model == "trigram"] /
    test_results$perplexity[test_results$model == "unigram"]

test_comparison_by_speech <- test_by_speech |>
  select(speech_id, model, perplexity) |>
  pivot_wider(
    names_from = model,
    values_from = perplexity
  )

speech_people <- paragraphs |>
  distinct(speech_id, president)
training_presidents <- split_assignments |>
  filter(split == "training") |>
  inner_join(speech_people, by = join_by(speech_id)) |>
  pull(president)
test_speaker_overlap <- split_assignments |>
  filter(split == "test") |>
  inner_join(speech_people, by = join_by(speech_id)) |>
  mutate(
    president_seen_in_training = president %in% training_presidents
  )

paired_test <- test_comparison_by_speech |>
  summarise(
    test_speeches = n(),
    trigram_wins = sum(trigram < unigram),
    ties = sum(trigram == unigram),
    trigram_losses = sum(trigram > unigram)
  )

speaker_overlap_test <- test_speaker_overlap |>
  summarise(
    repeated_president_speeches = sum(president_seen_in_training),
    other_speeches = sum(!president_seen_in_training)
  )
non_overlap_test <- test_comparison_by_speech |>
  inner_join(
    test_speaker_overlap |>
      select(speech_id, president_seen_in_training),
    by = join_by(speech_id)
  ) |>
  filter(!president_seen_in_training) |>
  summarise(
    speeches = n(),
    trigram_wins = sum(trigram < unigram)
  )

kable(
  test_display,
  col.names = c(
    "Model",
    "Role",
    "Test next-token rows",
    "Targets mapped to <unk>",
    "Test perplexity",
    "Top-1 accuracy"
  ),
  caption = "Untouched-test result for the selected model and baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(test_results$model, c("uniform", "unigram", "trigram")),
  identical(test_display$next_token_rows, rep(24337L, 3)),
  identical(test_display$perplexity, c(2500.0, 240.7, 145.3)),
  identical(
    test_display$top_1_accuracy,
    c("not applicable", "11.7%", "19.2%")
  ),
  identical(round(perplexity_reduction, 3), 0.396),
  identical(round(test_results$oov_target_share, 3), rep(0.117, 3)),
  identical(
    as.integer(unlist(paired_test)),
    c(13L, 13L, 0L, 0L)
  ),
  identical(
    as.integer(unlist(speaker_overlap_test)),
    c(8L, 5L)
  ),
  identical(
    as.integer(unlist(non_overlap_test)),
    c(5L, 5L)
  ),
  identical(
    unigram_probabilities$model_token[
      which.max(unigram_probabilities$unigram_probability)
    ],
    "<unk>"
  )
)
```

The uniform model has perplexity 2,500 because it gives every vocabulary item
the same probability. The selected trigram reduces test perplexity by about
`r sprintf("%.1f%%", 100 * perplexity_reduction)` relative to the unigram
baseline and wins the paired comparison in all
`r paired_test$trigram_wins` test speeches.

Its highest-probability guess is right on 19.2 percent of test rows. The unigram
scores 11.7 percent because its most likely token is `<unk>`, the bucket that
collects many rare training words. A correct unknown-bucket guess does not
recover the original word. The trigram also leans on this bucket: roughly one
third of its correct guesses are unknown-bucket matches rather than recovered
words.

Eight of the 13 test speeches were delivered by a president who also appears in
training. The test is speech-held-out, not speaker-held-out; repeated presidents
can carry phrasing across splits. The trigram still beats the unigram on all five
test speeches whose president is absent from training.

Perplexity would change under another vocabulary, tokenizer, boundary rule, or
test set. The values here cannot be compared with the LDA perplexity in Lesson
55 or a published model score from another tokenization.

The result does not show that the model understands a speech, writes a good
report, or produces facts. The next lesson uses these same measurements to show
how a report can stay tied to evidence.

## What to remember

- Next-token prediction assigns probabilities after a context.
- Paragraph boundaries prevent invented transitions between source records.
- Validation data select context length and interpolation strength; test data
  check the fixed choice.
- The trigram and interpolation strength are selected together on validation data.
- The selected trigram beats the unigram on all 13 test speeches here, but its
  top choice is still wrong on about four out of five rows.
- Perplexity is comparable only under the same tokens, vocabulary, and test rows.
- A likely next token is not necessarily factual or useful.

## Sources

- [Jurafsky and Martin, Speech and Language Processing, chapter on n-gram language models](https://web.stanford.edu/~jurafsky/slp3/3.pdf)
- [Bengio et al., A Neural Probabilistic Language Model](https://www.jmlr.org/papers/v3/bengio03a.html)
- [Chen and Goodman, An Empirical Study of Smoothing Techniques for Language Modeling](https://aclanthology.org/P96-1041/)
- [Hugging Face documentation on perplexity for fixed-length models](https://huggingface.co/docs/transformers/perplexity)
- [Meister and Cotterell, Language Model Evaluation Beyond Perplexity](https://aclanthology.org/2021.acl-long.414/)
- [quanteda inaugural-address corpus documentation](https://quanteda.io/reference/data_corpus_inaugural.html)
