---
title: "Writing a report from checked claims"
subtitle: "Choose evidence before generating sentences"
description: "Learn how to turn verified results into a short report while keeping every factual sentence tied to its source."
categories: [natural language generation, report writing, data to text]
---

Nia has finished the next-token study. The archive team wants a short report,
not a folder of CSV files and not a guess about why one model scored better.

**Report writing** is a natural language generation application. A
**data-to-text** workflow chooses supported content, orders the claims, turns
them into sentences, and checks the result against its sources. The sentences
can be produced by templates or drafted by a language model. Neither method
makes a claim true.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- define a report's audience, purpose, and exclusions;
- keep one evidence row for each reported value;
- separate content selection, document planning, and sentence realization;
- generate a deterministic report with `glue`;
- reject fluent claims that the evidence does not support; and
- keep factual checks separate from human prose review.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Start with the reporting decision

The audience is the archive team. They need to decide whether the
validation-selected trigram improves on the unigram baseline on the test
speeches. They do not need a claim about how people write, why one context
length won, or how another corpus would behave.

That brief limits the report before any sentence is generated:

- include the validation choice and untouched-test comparison;
- state the tokenizer, corpus, and unknown-token limit; and
- exclude causes, transfer claims, and claims about factual writing quality.

## Build an evidence ledger

An **evidence ledger** gives each usable result an ID. The report can then name
which records support each sentence instead of attaching a source only after
the prose is written.

```{r}
library(dplyr)
library(tibble)
library(stringr)
library(readr)
library(glue)
library(knitr)

validation_results <- read_csv(
  "data/inaugural/next-token-validation.csv",
  na = c("", "NA"),
  col_types = cols(
    split = col_character(),
    model = col_character(),
    context_tokens = col_integer(),
    interpolation_strength = col_double(),
    next_token_rows = col_integer(),
    oov_target_share = col_double(),
    seen_context_share = col_double(),
    mean_log_loss = col_double(),
    perplexity = col_double(),
    selected = col_logical(),
    selection_reason = col_character()
  )
)
test_results <- read_csv(
  "data/inaugural/next-token-test.csv",
  na = c("", "NA"),
  col_types = cols(
    split = col_character(),
    model = col_character(),
    context_tokens = col_integer(),
    interpolation_strength = col_double(),
    next_token_rows = col_integer(),
    oov_target_share = col_double(),
    seen_context_share = col_double(),
    mean_log_loss = col_double(),
    perplexity = col_double(),
    top_1_accuracy = col_double(),
    role = col_character()
  )
)
test_by_speech <- read_csv(
  "data/inaugural/next-token-test-by-speech.csv",
  na = character(),
  col_types = cols(
    speech_id = col_character(),
    next_token_rows = col_integer(),
    model = col_character(),
    perplexity = col_double()
  )
)

validation_trigram <- validation_results |>
  filter(selected)
test_uniform <- test_results |>
  filter(model == "uniform")
test_unigram <- test_results |>
  filter(model == "unigram")
test_trigram <- test_results |>
  filter(model == "trigram")

perplexity_reduction <- 1 -
  test_trigram$perplexity / test_unigram$perplexity

paired_test <- test_by_speech |>
  select(speech_id, model, perplexity) |>
  tidyr::pivot_wider(
    names_from = model,
    values_from = perplexity
  ) |>
  summarise(
    test_speeches = n(),
    trigram_wins = sum(trigram < unigram),
    ties = sum(trigram == unigram),
    trigram_losses = sum(trigram > unigram)
  )

evidence_ledger <- tribble(
  ~evidence_id, ~source_artifact, ~split, ~measure, ~value, ~unit, ~limit,
  "E1", "next-token-validation.csv", "validation", "trigram perplexity",
  validation_trigram$perplexity, "perplexity",
  "trigram with interpolation strength 100; minimum across 13 candidates",
  "E2", "next-token-test.csv", "test", "uniform perplexity",
  test_uniform$perplexity, "perplexity",
  "2,500 equally likely vocabulary items",
  "E3", "next-token-test.csv", "test", "unigram perplexity",
  test_unigram$perplexity, "perplexity",
  "predeclared baseline under the same tokens",
  "E4", "next-token-test.csv", "test", "trigram perplexity",
  test_trigram$perplexity, "perplexity",
  "selected on validation speeches",
  "E5", "next-token-test.csv", "test", "trigram top-1 accuracy",
  test_trigram$top_1_accuracy, "share",
  "one highest-probability guess per row",
  "E6", "next-token-test-by-speech.csv", "test", "speech-level wins",
  paired_test$trigram_wins, "speeches",
  "paired against the unigram on the same test speeches",
  "E7", "next-token-test.csv", "test", "targets mapped to <unk>",
  test_trigram$oov_target_share, "share",
  "the original unseen token is not identified",
  "E8", "data-raw/build-next-token-study.R", "study", "token rule",
  NA_real_, "setup",
  "lowercase ASCII alphabetic tokens; training-only vocabulary",
  "E9", "R/inaugural-corpus.R", "study", "corpus grain",
  NA_real_, "setup",
  "reconstructed inaugural paragraphs; paragraph-bounded contexts"
)

kable(
  evidence_ledger |>
    mutate(
      value = case_when(
        unit == "share" ~ sprintf("%.1f%%", 100 * value),
        unit == "speeches" ~ as.character(as.integer(value)),
        unit == "setup" ~ "documented",
        TRUE ~ sprintf("%.1f", value)
      )
    ),
  col.names = c(
    "Evidence ID",
    "Source artifact",
    "Split",
    "Measure",
    "Value",
    "Unit",
    "Limit"
  ),
  caption = "Evidence records available to the report",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(evidence_ledger$evidence_id, paste0("E", 1:9)),
  identical(validation_trigram$model, "trigram"),
  identical(validation_trigram$interpolation_strength, 100),
  identical(
    validation_trigram$perplexity,
    min(validation_results$perplexity)
  ),
  identical(round(validation_trigram$perplexity, 1), 133.3),
  identical(round(test_uniform$perplexity, 1), 2500.0),
  identical(round(test_unigram$perplexity, 1), 240.7),
  identical(round(test_trigram$perplexity, 1), 145.3),
  identical(round(test_trigram$top_1_accuracy, 3), 0.192),
  identical(paired_test$test_speeches, 13L),
  identical(paired_test$trigram_wins, 13L),
  identical(round(perplexity_reduction, 3), 0.396),
  identical(round(test_trigram$oov_target_share, 3), 0.117),
  identical(
    evidence_ledger$source_artifact[8:9],
    c("data-raw/build-next-token-study.R", "R/inaugural-corpus.R")
  ),
  all(is.na(evidence_ledger$value[8:9]))
)
```

The IDs carry more than numbers. They also record the split, unit, role, and
limit. E1 cannot support a final test claim. E2 through E4 are comparable
because they use the same test rows, tokenizer, and vocabulary.

## Decide which claims belong

**Content determination** chooses the facts the report will include. A fluent
sentence can still fail here.

```{r}
candidate_claims <- tribble(
  ~claim_id, ~claim, ~evidence_ids, ~status, ~reason,
  "C1",
  glue(
    "The trigram with interpolation strength 100 had the lowest ",
    "validation perplexity ({round(validation_trigram$perplexity, 1)})."
  ),
  "E1",
  "supported",
  "describes the recorded selection result",
  "C2",
  glue(
    "On test speeches, trigram perplexity was ",
    "{round(test_trigram$perplexity, 1)} versus ",
    "{round(test_unigram$perplexity, 1)} for the unigram baseline."
  ),
  "E3;E4",
  "supported",
  "compares the same unit and test rows",
  "C3",
  glue(
    "The trigram's highest-probability token matched ",
    "{sprintf('%.1f%%', 100 * test_trigram$top_1_accuracy)} ",
    "of test rows."
  ),
  "E5",
  "supported",
  "reports the decision rule separately from perplexity",
  "C4",
  glue(
    "The trigram reduced test perplexity by ",
    "{sprintf('%.1f%%', 100 * perplexity_reduction)} ",
    "relative to the unigram baseline."
  ),
  "E3;E4",
  "supported",
  "derives a comparison from two test perplexities",
  "C5",
  glue(
    "The trigram had lower perplexity in ",
    "{paired_test$trigram_wins} of {paired_test$test_speeches} ",
    "paired test speeches."
  ),
  "E6",
  "supported",
  "uses the paired speech-level comparison",
  "C6",
  paste(
    "The study uses lowercase alphabetic tokens from reconstructed",
    "inaugural paragraphs."
  ),
  "E8;E9",
  "supported",
  "states the token rule and source grain",
  "C7",
  glue(
    "{sprintf('%.1f%%', 100 * test_trigram$oov_target_share)} ",
    "of test targets were mapped to an unknown-token bucket."
  ),
  "E7",
  "supported",
  "states the recorded unknown-token limit",
  "U1",
  "The trigram understands presidential language.",
  NA_character_,
  "reject",
  "perplexity does not measure understanding",
  "U2",
  "Using a trigram will improve report writing.",
  NA_character_,
  "reject",
  "the study did not evaluate reports"
)

kable(
  candidate_claims,
  col.names = c("Claim ID", "Candidate sentence", "Evidence IDs", "Decision", "Reason"),
  caption = "Candidate claims are checked before document planning",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    candidate_claims$status,
    c(rep("supported", 7), "reject", "reject")
  ),
  identical(
    candidate_claims$evidence_ids[1:7],
    c("E1", "E3;E4", "E5", "E3;E4", "E6", "E8;E9", "E7")
  ),
  all(is.na(candidate_claims$evidence_ids[8:9])),
  str_detect(candidate_claims$claim[[1]], fixed("133.3")),
  str_detect(candidate_claims$claim[[2]], fixed("145.3")),
  str_detect(candidate_claims$claim[[2]], fixed("240.7")),
  str_detect(candidate_claims$claim[[3]], fixed("19.2%")),
  str_detect(candidate_claims$claim[[4]], fixed("39.6%")),
  str_detect(candidate_claims$claim[[5]], fixed("13 of 13")),
  str_detect(candidate_claims$claim[[6]], fixed("reconstructed")),
  str_detect(candidate_claims$claim[[7]], fixed("11.7%"))
)
```

The rejected sentences sound reasonable. They still exceed the evidence.
Perplexity measures next-token probability under one setup. The study contains
no test of understanding and no written-report outcome.

## Plan the document before writing it

**Document planning** sets the order and purpose of the sentences.
**Realization** turns the checked values into wording. The plan below starts
with the decision, gives the untouched-test result, and ends with the limit.

```{r}
report_plan <- tribble(
  ~sentence_id, ~role, ~claim_ids, ~evidence_ids,
  "S1", "selection", "C1", "E1",
  "S2", "test comparison", "C2;C4;C5", "E3;E4;E6",
  "S3", "top-choice result", "C3", "E5",
  "S4", "scope and unknown-token limit", "C6;C7", "E7;E8;E9"
)

report_sentences <- report_plan |>
  mutate(
    text = case_when(
      sentence_id == "S1" ~ glue(
        "A trigram with interpolation strength 100 was selected because it had ",
        "the lowest validation perplexity ",
        "({round(validation_trigram$perplexity, 1)})."
      ),
      sentence_id == "S2" ~ glue(
        "On {paired_test$test_speeches} untouched test speeches, ",
        "its perplexity was ",
        "{round(test_trigram$perplexity, 1)}, compared with ",
        "{round(test_unigram$perplexity, 1)} for the unigram baseline, ",
        "a reduction of {sprintf('%.1f%%', 100 * perplexity_reduction)}; ",
        "it had lower perplexity in {paired_test$trigram_wins} of ",
        "{paired_test$test_speeches} speeches."
      ),
      sentence_id == "S3" ~ glue(
        "Its highest-probability token matched ",
        "{sprintf('%.1f%%', 100 * test_trigram$top_1_accuracy)} ",
        "of test rows."
      ),
      sentence_id == "S4" ~ glue(
        "These results describe lowercase alphabetic tokens from reconstructed ",
        "inaugural paragraphs; {sprintf('%.1f%%', 100 * test_trigram$oov_target_share)} ",
        "of test targets were mapped to an unknown-token bucket."
      )
    )
  )

report_text <- str_c(
  report_sentences$text,
  collapse = "\n\n"
)

kable(
  report_sentences |>
    select(sentence_id, role, claim_ids, evidence_ids, text),
  col.names = c("Sentence", "Role", "Claim IDs", "Evidence IDs", "Generated text"),
  caption = "Every factual report sentence retains its claim and evidence IDs",
  row.names = FALSE
)
```

```{r}
#| include: false

split_id_string <- function(value) {
  value |>
    str_split(";") |>
    unlist(use.names = FALSE) |>
    unique() |>
    sort()
}

used_claim_ids <- split_id_string(report_sentences$claim_ids)
used_evidence_ids <- split_id_string(report_sentences$evidence_ids)
supported_claims <- candidate_claims |>
  filter(status == "supported")

plan_trace_resolves <- vapply(
  seq_len(nrow(report_plan)),
  function(row) {
    plan_claim_ids <- split_id_string(report_plan$claim_ids[[row]])
    plan_evidence_ids <- split_id_string(
      report_plan$evidence_ids[[row]]
    )
    claim_evidence_ids <- supported_claims |>
      filter(claim_id %in% plan_claim_ids) |>
      pull(evidence_ids) |>
      split_id_string()

    all(plan_claim_ids %in% supported_claims$claim_id) &&
      identical(claim_evidence_ids, plan_evidence_ids)
  },
  logical(1)
)

claim_and_evidence_ids_resolve <-
  setequal(used_claim_ids, supported_claims$claim_id) &&
    all(used_evidence_ids %in% evidence_ledger$evidence_id) &&
    all(plan_trace_resolves)

numbers_match_evidence <- all(c(
  str_detect(
    report_text,
    fixed(as.character(round(validation_trigram$perplexity, 1)))
  ),
  str_detect(
    report_text,
    fixed(as.character(round(test_trigram$perplexity, 1)))
  ),
  str_detect(
    report_text,
    fixed(as.character(round(test_unigram$perplexity, 1)))
  ),
  str_detect(
    report_text,
    fixed(sprintf("%.1f%%", 100 * perplexity_reduction))
  ),
  str_detect(
    report_text,
    fixed(sprintf("%.1f%%", 100 * test_trigram$top_1_accuracy))
  ),
  str_detect(
    report_text,
    fixed(sprintf("%.1f%%", 100 * test_trigram$oov_target_share))
  ),
  str_detect(
    report_text,
    fixed(as.character(glue(
      "{paired_test$trigram_wins} of ",
      "{paired_test$test_speeches} speeches"
    )))
  )
))

unsupported_claims_excluded <- all(c(
  !str_detect(
    report_text,
    regex("\\bunderstands?\\b", ignore_case = TRUE)
  ),
  !str_detect(
    report_text,
    regex("\\bcaused?\\b", ignore_case = TRUE)
  ),
  !str_detect(report_text, fixed("improve report writing"))
))

automated_checks <- c(
  "claim and evidence IDs resolve" =
    claim_and_evidence_ids_resolve,
  "numbers match evidence" = numbers_match_evidence,
  "unsupported claims excluded" = unsupported_claims_excluded
)

stopifnot(
  identical(report_sentences$sentence_id, paste0("S", 1:4)),
  identical(length(plan_trace_resolves), 4L),
  all(automated_checks)
)
```

The trace table stays with the report draft. If a value changes, the sentence
can be found by claim or evidence ID rather than by searching for a number
copied into a paragraph.

## Render the checked report

This chunk writes the four planned sentences as report prose. The generation
is deterministic: the same evidence and templates produce the same text.

```{r}
#| output: asis

cat(
  "### Next-token study report\n\n",
  report_text,
  "\n",
  sep = ""
)
```

```{r}
#| include: false

expected_report <- str_c(
  paste0(
    "A trigram with interpolation strength 100 was selected because it had the ",
    "lowest validation perplexity (133.3)."
  ),
  paste0(
    "On 13 untouched test speeches, its perplexity was 145.3, compared with ",
    "240.7 for the unigram baseline, a reduction of 39.6%; it had lower ",
    "perplexity in 13 of 13 speeches."
  ),
  paste0(
    "Its highest-probability token matched 19.2% of test rows."
  ),
  paste0(
    "These results describe lowercase alphabetic tokens from reconstructed ",
    "inaugural paragraphs; 11.7% of test targets were mapped to an ",
    "unknown-token bucket."
  ),
  sep = "\n\n"
)

stopifnot(
  identical(report_text, expected_report)
)
```

## Check facts and prose separately

The automated checks can verify known numbers, claim and evidence IDs, and
prohibited claims. They cannot decide whether the report is clear enough for the
archive team or whether that team should act on it.

```{r}
review_status <- tibble(
  review = c(
    names(automated_checks),
    "human factual review",
    "human prose review"
  ),
  status = c(
    unname(if_else(automated_checks, "passed", "failed")),
    "pending",
    "pending"
  )
)

kable(
  review_status,
  col.names = c("Review gate", "Status"),
  caption = "Automated checks do not substitute for human approval",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    review_status$review[1:3],
    names(automated_checks)
  ),
  identical(
    review_status$status,
    c("passed", "passed", "passed", "pending", "pending")
  ),
  sum(review_status$status == "passed") == 3L,
  sum(review_status$status == "pending") == 2L
)
```

The factual reviewer follows each sentence back to the source records and checks
that the wording does not add a cause, population, or claim of transfer. The
prose reviewer checks order, clarity, tone, and usefulness for the named
audience. Passing one review says nothing about the other.

## Where a language model fits

A language model can draft alternate wording after the evidence ledger and
report plan are fixed. Save the prompt, model, version, and output if the draft
is used. Reject any new number, entity, cause, citation, or evidence ID that is
not in the ledger, then run factual and human review again.

This lesson does not call a live model. Credentials, model updates, and
stochastic output would make the rendered page hard to reproduce. More
important, a plausible continuation would still not supply evidence for its own
claims.

## What to remember

- Define the audience and decision before generating prose.
- Select supported claims before arranging sentences.
- Keep evidence IDs, units, splits, and limits attached.
- Templates improve reproducibility; they do not establish truth.
- A language model may draft wording, but new claims require evidence.
- Automated factual checks and human prose review answer different questions.

## Sources

- [Reiter and Dale, Building Natural Language Generation Systems](https://doi.org/10.1017/CBO9780511519857)
- [Gatt and Krahmer, Survey of the State of the Art in Natural Language Generation](https://arxiv.org/abs/1703.09902)
- [Wiseman, Shieber, and Rush on challenges in data-to-document generation](https://aclanthology.org/D17-1239/)
- [Thomson and Reiter on automatic evaluation of generated text](https://aclanthology.org/2020.inlg-1.22/)
- [NIST AI 600-1, Generative Artificial Intelligence Profile](https://doi.org/10.6028/NIST.AI.600-1)
- [glue package documentation](https://glue.tidyverse.org/articles/glue.html)
