---
title: "Moving a notice from English to French"
subtitle: "Run one fixed translation direction and inspect what survived"
description: "Learn how a local machine-translation checkpoint changes language while preserving the source message."
categories: [natural language generation, machine translation, evaluation]
---

An archive sends short notices to English- and French-speaking visitors. A
translation must change the language while preserving the people, quantities,
dates, permissions, and limits in the source.

The model in this lesson has one direction: English to French. That direction
belongs to the checkpoint. It is not a language choice discovered from each
input, and it is not changed by adding a target-language label to the output.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- load a revision-pinned English-to-French model and tokenizer locally;
- keep constructed source records and generated translations tied to IDs;
- count source tokens against the checkpoint's 512-token input limit;
- compare translation with a source-copy baseline;
- inspect names, numbers, and negation without calling them adequacy scores; and
- reserve meaning and register judgments for a bilingual reviewer.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load one local translation direction

`huggingfaceR::hf_load_tokenizer()` and
`huggingfaceR::hf_load_pipeline()` load the prepared local snapshot through the
course helper. Offline mode prevents a render-time download. The model and
tokenizer files came from the same immutable revision.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(huggingfaceR)
  library(jsonlite)
  library(knitr)
  library(purrr)
  library(reticulate)
  library(stringr)
  library(tibble)
  library(tidyr)
})

source("R/use-nlg.R")

translation_model <- load_nlg_pipeline(
  "opus_en_fr",
  "translation"
)

model_record <- translation_model$metadata |>
  select(model_id, revision, license, task_scope)

kable(
  model_record,
  col.names = c("Model", "Revision", "License", "Use in this course"),
  caption = "Pinned local checkpoint for English-to-French translation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "dplyr", "huggingfaceR", "jsonlite", "knitr", "purrr",
    "reticulate", "stringr", "tibble", "tidyr"
  ) %in% loadedNamespaces()),
  identical(
    model_record$model_id,
    "Helsinki-NLP/opus-mt-en-fr"
  ),
  identical(
    model_record$revision,
    "dd7f6540a7a48a7f4db59e5c0b9c42c8eea67f18"
  ),
  identical(model_record$license, "Apache-2.0")
)
```

The package also exports `hf_translate()`, which calls a remote inference
service. This lesson does not use it. Generation runs through the local model
object returned by the pinned pipeline.

## Read direction from the checkpoint

The local tokenizer configuration names English as the source and French as
the target. A different direction requires a different checkpoint.

```{r}
tokenizer_config_path <- file.path(
  "data-raw",
  ".cache",
  "nlg-models",
  translation_model$metadata$local_directory,
  "tokenizer_config.json"
)
generation_config_path <- file.path(
  "data-raw",
  ".cache",
  "nlg-models",
  translation_model$metadata$local_directory,
  "generation_config.json"
)

tokenizer_config <- fromJSON(tokenizer_config_path)
generation_config <- fromJSON(generation_config_path)

direction <- tibble(
  checkpoint = translation_model$metadata$model_id,
  source_language = tokenizer_config$source_lang,
  target_language = tokenizer_config$target_lang,
  runtime_language_switch = FALSE,
  checkpoint_beams = generation_config$num_beams
)

kable(
  direction,
  col.names = c(
    "Checkpoint",
    "Source",
    "Target",
    "Runtime direction switch",
    "Checkpoint beams"
  ),
  caption = "Translation direction recorded in the tokenizer configuration",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  file.exists(tokenizer_config_path),
  file.exists(generation_config_path),
  identical(direction$source_language, "en"),
  identical(direction$target_language, "fr"),
  identical(direction$runtime_language_switch, FALSE),
  identical(direction$checkpoint_beams, 4L)
)
```

## Prepare constructed notices

These four English notices were written for the lesson. They are demonstrations,
not human reference translations, a held-out test set, or a population quality
benchmark.

```{r}
translation_inputs <- tribble(
  ~source_id, ~source_text, ~protected_name, ~requires_negation,
  "MT-01",
  "Maya Chen will open Room 5 on 14 September.",
  "Maya Chen",
  FALSE,
  "MT-02",
  "The library did not cancel the workshop.",
  "",
  TRUE,
  "MT-03",
  "Visitors must bring 2 forms.",
  "",
  FALSE,
  "MT-04",
  "The volunteers may use 12 computers on Tuesday.",
  "",
  FALSE
)

kable(
  translation_inputs,
  col.names = c(
    "Source ID",
    "Constructed English source",
    "Protected name",
    "Negation required"
  ),
  caption = "Constructed records for the translation demonstration",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(translation_inputs$source_id, paste0("MT-0", 1:4)),
  !anyDuplicated(translation_inputs$source_id),
  identical(sum(translation_inputs$requires_negation), 1L)
)
```

## Count tokens before calling the model

Marian tokenizes words into subword units. Its configured input ceiling is 512
tokens. Truncation is not enabled here because dropping the end of a notice can
drop its meaning.

```{r}
translation_budget <- translation_inputs |>
  mutate(
    input_tokens = map_int(
      source_text,
      \(text) nlg_token_count(translation_model$tokenizer, text)
    ),
    model_input_limit = 512L,
    fits_without_truncation = input_tokens <= model_input_limit
  )

kable(
  translation_budget |>
    select(source_id, input_tokens, model_input_limit, fits_without_truncation),
  col.names = c(
    "Source ID",
    "Tokenizer units",
    "Input limit",
    "Fits without truncation"
  ),
  caption = "Input budgets are checked before translation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(translation_budget$input_tokens > 0L),
  all(translation_budget$fits_without_truncation),
  all(translation_budget$model_input_limit == 512L)
)
```

For a longer source, setting `truncation = TRUE` would discard source tokens and
become a lossy data decision. A production workflow would section the document
or stop for review.

## Generate French candidates

The source text goes directly to the encoder. The checkpoint ships with a
four-beam generation default. The course helper overrides it with
`num_beams = 1` and `do_sample = FALSE`, a deterministic greedy choice used to
keep this teaching workflow consistent. It is not a claim that the checkpoint's
four-beam setting is nondeterministic or inferior.

The output limit is 80 newly generated tokens. The returned record separately
identifies whether the decoder emitted its end-of-sequence token and whether it
used the full output budget. Both can be true when the checkpoint forces EOS at
the cap. Neither field proves that the translation is complete for a reader.

```{r}
translation_results <- translation_budget |>
  mutate(
    generation = map(
      source_text,
      \(text) {
        nlg_generate(
          translation_model,
          text,
          max_new_tokens = 80L
        )
      }
    ),
    french_candidate = map_chr(generation, "text"),
    output_tokens = map_int(generation, "output_tokens"),
    ended_by_eos = map_lgl(generation, "ended_by_eos"),
    hit_token_cap = map_lgl(generation, "hit_token_cap"),
    source_copy_baseline = source_text
  )

kable(
  translation_results |>
    select(source_id, source_text, french_candidate, source_copy_baseline),
  col.names = c(
    "Source ID",
    "English source",
    "Generated French candidate",
    "Untranslated copy baseline"
  ),
  caption = "Real local translations beside a no-translation baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(nzchar(translation_results$french_candidate)),
  all(translation_results$output_tokens > 0L),
  all(translation_results$output_tokens <= 80L),
  all(
    translation_results$ended_by_eos |
      translation_results$hit_token_cap
  ),
  any(
    translation_results$french_candidate !=
      translation_results$source_copy_baseline
  )
)
```

The copy baseline preserves the English source exactly and performs no
translation. It prevents a changed-looking string from being treated as
automatic evidence that the task succeeded.

## Inspect visible details

The next checks compare exact Arabic numerals, look for the protected person
name, and look for a common French negation marker when the English source
contains `did not`. These checks can find some clear failures. They cannot
decide whether the sentence is grammatical, whether `must` and `may` have the
right force, or whether the register fits the audience.

```{r}
extract_numbers <- function(text) {
  str_extract_all(text, "\\b\\d+\\b")[[1]] |>
    sort()
}

translation_checks <- translation_results |>
  mutate(
    numerals_preserved = map2_lgl(
      source_text,
      french_candidate,
      \(source, output) {
        identical(
          extract_numbers(source),
          extract_numbers(output)
        )
      }
    ),
    name_present = map2_lgl(
      protected_name,
      french_candidate,
      \(name, output) {
        if (!nzchar(name)) {
          NA
        } else {
          str_detect(output, fixed(name))
        }
      }
    ),
    negation_visible = map2_lgl(
      requires_negation,
      french_candidate,
      \(required, output) {
        if (!required) {
          NA
        } else {
          str_detect(
            output,
            regex("\\b(pas|jamais|aucun)\\b", ignore_case = TRUE)
          )
        }
      }
    )
  ) |>
  select(
    source_id,
    numerals_preserved,
    name_present,
    negation_visible
  )

translation_check_display <- translation_checks |>
  mutate(
    across(
      c(numerals_preserved, name_present, negation_visible),
      nlg_screen_label
    )
  )

kable(
  translation_check_display,
  col.names = c(
    "Source ID",
    "Arabic-numeral screen",
    "Protected-name screen",
    "French-negation screen"
  ),
  caption = "Conservative surface screens for the constructed notices",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(translation_checks$source_id, translation_inputs$source_id),
  all(vapply(translation_checks[-1], is.logical, logical(1))),
  identical(nrow(translation_checks), 4L),
  all(is.na(
    translation_checks$name_present[
      !nzchar(translation_inputs$protected_name)
    ]
  )),
  all(is.na(
    translation_checks$negation_visible[
      !translation_inputs$requires_negation
    ]
  ))
)
```

No BLEU or chrF number is reported. Those metrics require a documented
reference corpus, preprocessing signature, and corpus-level interpretation.
Even then, one reference cannot represent every acceptable French rendering,
and automatic overlap does not replace bilingual judgment.

`Not flagged by screen` means only that this finite string check found no
problem. It does not verify translation adequacy. `Not applicable` means the
source contained no protected name or negation condition to test.

## Leave approval to a bilingual reader

```{r}
translation_screen_counts <- translation_checks |>
  rowwise() |>
  mutate(
    applicable_screens = sum(!is.na(c_across(c(
      numerals_preserved,
      name_present,
      negation_visible
    )))),
    flagged_screens = sum(c_across(c(
      numerals_preserved,
      name_present,
      negation_visible
    )) %in% FALSE)
  ) |>
  ungroup()

translation_review <- tibble(
  review = c(
    "local model execution",
    "source within the 512-token input limit",
    "decoder stopped at EOS or the output cap",
    "surface preservation checks",
    "bilingual meaning and register review"
  ),
  status = c(
    sprintf(
      "completed: %d candidates",
      sum(nzchar(translation_results$french_candidate))
    ),
    sprintf(
      "within budget: %d/%d candidates",
      sum(translation_budget$fits_without_truncation),
      nrow(translation_budget)
    ),
    sprintf(
      "recorded: %d/%d candidates",
      sum(
        translation_results$ended_by_eos |
          translation_results$hit_token_cap
      ),
      nrow(translation_results)
    ),
    sprintf(
      "scan completed: %d flags across %d applicable checks",
      sum(translation_screen_counts$flagged_screens),
      sum(translation_screen_counts$applicable_screens)
    ),
    "pending"
  )
)

kable(
  translation_review,
  col.names = c("Review gate", "Status"),
  caption = "Automatic checks do not grant bilingual approval",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  !any(translation_review$status == "passed"),
  identical(tail(translation_review$status, 1L), "pending"),
  all(translation_screen_counts$flagged_screens >= 0L)
)
```

The lesson contains no human ratings. A bilingual reviewer still has to compare
each pair for meaning, tone, register, and natural French.

## What to remember

- Translation changes language while trying to preserve the source message.
- This checkpoint's English-to-French direction is fixed in its model files.
- Count model tokens before generation and treat truncation as data loss.
- Keep an untranslated copy baseline beside generated candidates.
- Names, numbers, and negation checks can reveal errors but cannot prove
  adequacy.
- Human bilingual review remains necessary before publication.

## Sources

- [Helsinki-NLP English-to-French model card](https://huggingface.co/Helsinki-NLP/opus-mt-en-fr)
- [OPUS-MT: open translation models for many languages](https://aclanthology.org/2020.eamt-1.61/)
- [Papineni et al. on corpus-level BLEU](https://aclanthology.org/P02-1040/)
- [Post on reproducible machine-translation evaluation](https://aclanthology.org/W18-6319/)
- [Hossain et al. on negation in machine translation](https://aclanthology.org/2020.findings-emnlp.345/)
