---
title: "Moving a notice from English to French"
subtitle: "Run one fixed translation direction and inspect what survived"
description: "Learn how a local machine-translation checkpoint changes language while preserving the source message."
categories: [natural language generation, machine translation, evaluation]
---

An archive sends short notices to English- and French-speaking visitors. A
translation must change the language while preserving the people, quantities,
dates, permissions, and limits in the source.

The model in this lesson has one direction: English to French. That direction
belongs to the checkpoint. It is not a language choice discovered from each
input, and it is not changed by adding a target-language label to the output.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- load a revision-pinned English-to-French model and tokenizer locally;
- keep constructed source records and generated translations tied to IDs;
- count source tokens against the checkpoint's 512-token input limit;
- compare translation with a source-copy baseline;
- inspect names, numbers, and negation without calling them adequacy scores; and
- reserve meaning and register judgments for a bilingual reviewer.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load one local translation direction

`huggingfaceR::hf_load_tokenizer()` and
`huggingfaceR::hf_load_pipeline()` load the prepared local snapshot through the
course helper. Offline mode prevents a render-time download. The model and
tokenizer files came from the same immutable revision.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(huggingfaceR)
  library(jsonlite)
  library(knitr)
  library(purrr)
  library(reticulate)
  library(stringr)
  library(tibble)
  library(tidyr)
})

source("R/use-nlg.R")

translation_model <- load_nlg_pipeline(
  "opus_en_fr",
  "translation"
)

model_record <- translation_model$metadata |>
  select(model_id, revision, license, task_scope)

kable(
  model_record,
  col.names = c("Model", "Revision", "License", "Use in this course"),
  caption = "Pinned local checkpoint for English-to-French translation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "dplyr", "huggingfaceR", "jsonlite", "knitr", "purrr",
    "reticulate", "stringr", "tibble", "tidyr"
  ) %in% loadedNamespaces()),
  identical(
    model_record$model_id,
    "Helsinki-NLP/opus-mt-en-fr"
  ),
  identical(
    model_record$revision,
    "dd7f6540a7a48a7f4db59e5c0b9c42c8eea67f18"
  ),
  identical(model_record$license, "Apache-2.0")
)
```

The package also exports `hf_translate()`, which calls a remote inference
service. This lesson does not use it. Generation runs through the local model
object returned by the pinned pipeline.

## Read direction from the checkpoint

The local tokenizer configuration names English as the source and French as
the target. A different direction requires a different checkpoint.

```{r}
tokenizer_config_path <- file.path(
  "data-raw",
  ".cache",
  "nlg-models",
  translation_model$metadata$local_directory,
  "tokenizer_config.json"
)

tokenizer_config <- fromJSON(tokenizer_config_path)

direction <- tibble(
  checkpoint = translation_model$metadata$model_id,
  source_language = tokenizer_config$source_lang,
  target_language = tokenizer_config$target_lang,
  runtime_language_switch = FALSE
)

kable(
  direction,
  col.names = c(
    "Checkpoint",
    "Source",
    "Target",
    "Runtime direction switch"
  ),
  caption = "Translation direction recorded in the tokenizer configuration",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  file.exists(tokenizer_config_path),
  identical(direction$source_language, "en"),
  identical(direction$target_language, "fr"),
  identical(direction$runtime_language_switch, FALSE)
)
```

## Prepare constructed notices

These four English notices were written for the lesson. They are demonstrations,
not human reference translations, a held-out test set, or a population quality
benchmark.

```{r}
translation_inputs <- tribble(
  ~source_id, ~source_text, ~protected_name, ~requires_negation,
  "MT-01",
  "Maya Chen will open Room 5 on 14 September.",
  "Maya Chen",
  FALSE,
  "MT-02",
  "The library did not cancel the workshop.",
  "",
  TRUE,
  "MT-03",
  "Visitors must bring 2 forms.",
  "",
  FALSE,
  "MT-04",
  "The volunteers may use 12 computers on Tuesday.",
  "",
  FALSE
)

kable(
  translation_inputs,
  col.names = c(
    "Source ID",
    "Constructed English source",
    "Protected name",
    "Negation required"
  ),
  caption = "Constructed records for the translation demonstration",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(translation_inputs$source_id, paste0("MT-0", 1:4)),
  !anyDuplicated(translation_inputs$source_id),
  identical(sum(translation_inputs$requires_negation), 1L)
)
```

## Count tokens before calling the model

Marian tokenizes words into subword units. Its configured input ceiling is 512
tokens. Truncation is not enabled here because dropping the end of a notice can
drop its meaning.

```{r}
translation_budget <- translation_inputs |>
  mutate(
    input_tokens = map_int(
      source_text,
      \(text) nlg_token_count(translation_model$tokenizer, text)
    ),
    model_input_limit = 512L,
    fits_without_truncation = input_tokens <= model_input_limit
  )

kable(
  translation_budget |>
    select(source_id, input_tokens, model_input_limit, fits_without_truncation),
  col.names = c(
    "Source ID",
    "Tokenizer units",
    "Input limit",
    "Fits without truncation"
  ),
  caption = "Input budgets are checked before translation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(translation_budget$input_tokens > 0L),
  all(translation_budget$fits_without_truncation),
  all(translation_budget$model_input_limit == 512L)
)
```

For a longer source, setting `truncation = TRUE` would discard source tokens and
become a lossy data decision. A production workflow would section the document
or stop for review.

## Generate French candidates

The source text goes directly to the encoder. Greedy decoding uses one beam and
a limit of 80 new tokens. The returned record identifies whether the decoder
emitted its end-of-sequence token or exhausted the configured output budget.

```{r}
translation_results <- translation_budget |>
  mutate(
    generation = map(
      source_text,
      \(text) {
        nlg_generate(
          translation_model,
          text,
          max_new_tokens = 80L
        )
      }
    ),
    french_candidate = map_chr(generation, "text"),
    output_tokens = map_int(generation, "output_tokens"),
    ended_by_eos = map_lgl(generation, "ended_by_eos"),
    hit_token_cap = map_lgl(generation, "hit_token_cap"),
    source_copy_baseline = source_text
  )

kable(
  translation_results |>
    select(source_id, source_text, french_candidate, source_copy_baseline),
  col.names = c(
    "Source ID",
    "English source",
    "Generated French candidate",
    "Untranslated copy baseline"
  ),
  caption = "Real local translations beside a no-translation baseline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(nzchar(translation_results$french_candidate)),
  all(translation_results$output_tokens > 0L),
  all(translation_results$output_tokens <= 80L),
  all(
    translation_results$ended_by_eos |
      translation_results$hit_token_cap
  ),
  any(
    translation_results$french_candidate !=
      translation_results$source_copy_baseline
  )
)
```

The copy baseline preserves the English source exactly and performs no
translation. It prevents a changed-looking string from being treated as
automatic evidence that the task succeeded.

## Inspect visible details

The next checks compare exact Arabic numerals, look for the protected person
name, and look for a common French negation marker when the English source
contains `did not`. These checks can find some clear failures. They cannot
decide whether the sentence is grammatical, whether `must` and `may` have the
right force, or whether the register fits the audience.

```{r}
extract_numbers <- function(text) {
  str_extract_all(text, "\\b\\d+\\b")[[1]] |>
    sort()
}

translation_checks <- translation_results |>
  mutate(
    numerals_preserved = map2_lgl(
      source_text,
      french_candidate,
      \(source, output) {
        identical(
          extract_numbers(source),
          extract_numbers(output)
        )
      }
    ),
    name_present = map2_lgl(
      protected_name,
      french_candidate,
      \(name, output) {
        !nzchar(name) || str_detect(output, fixed(name))
      }
    ),
    negation_visible = map2_lgl(
      requires_negation,
      french_candidate,
      \(required, output) {
        !required ||
          str_detect(
            output,
            regex("\\b(pas|jamais|aucun)\\b", ignore_case = TRUE)
          )
      }
    )
  ) |>
  select(
    source_id,
    numerals_preserved,
    name_present,
    negation_visible
  )

kable(
  translation_checks,
  col.names = c(
    "Source ID",
    "Numerals match",
    "Protected name present",
    "French negation marker present"
  ),
  caption = "Surface preservation checks for the constructed notices",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(translation_checks$source_id, translation_inputs$source_id),
  all(vapply(translation_checks[-1], is.logical, logical(1))),
  identical(nrow(translation_checks), 4L)
)
```

No BLEU or chrF number is reported. Those metrics require a documented
reference corpus, preprocessing signature, and corpus-level interpretation.
Even then, one reference cannot represent every acceptable French rendering,
and automatic overlap does not replace bilingual judgment.

## Leave approval to a bilingual reader

```{r}
translation_review <- tibble(
  review = c(
    "local model execution",
    "source within the 512-token input limit",
    "decoder stopped at EOS or the output cap",
    "surface preservation checks",
    "bilingual meaning and register review"
  ),
  status = c(
    "passed",
    "passed",
    "passed",
    "passed",
    "pending"
  )
)

kable(
  translation_review,
  col.names = c("Review gate", "Status"),
  caption = "Automatic checks do not grant bilingual approval",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    translation_review$status,
    c("passed", "passed", "passed", "passed", "pending")
  )
)
```

The lesson contains no human ratings. A bilingual reviewer still has to compare
each pair for meaning, tone, register, and natural French.

## What to remember

- Translation changes language while trying to preserve the source message.
- This checkpoint's English-to-French direction is fixed in its model files.
- Count model tokens before generation and treat truncation as data loss.
- Keep an untranslated copy baseline beside generated candidates.
- Names, numbers, and negation checks can reveal errors but cannot prove
  adequacy.
- Human bilingual review remains necessary before publication.

## Sources

- [Helsinki-NLP English-to-French model card](https://huggingface.co/Helsinki-NLP/opus-mt-en-fr)
- [OPUS-MT: open translation models for many languages](https://aclanthology.org/2020.eamt-1.61/)
- [Papineni et al. on corpus-level BLEU](https://aclanthology.org/P02-1040/)
- [Post on reproducible machine-translation evaluation](https://aclanthology.org/W18-6319/)
- [Hossain et al. on negation in machine translation](https://aclanthology.org/2020.findings-emnlp.345/)
