---
title: "Writing a shorter account"
subtitle: "Generate a summary, then check what it kept and changed"
description: "Learn how abstractive summarization differs from sentence selection and why shorter text still needs source review."
categories: [natural language generation, abstractive summarization, evaluation]
---

Rosa prepares the morning brief for a neighborhood archive. Two service notes
have arrived, and each is longer than the one sentence her colleagues need.
The shorter sentence still has to respect the source.

**Abstractive summarization** generates wording for a shorter account of a
source. The output may reuse some source phrases, but it is not limited to
selecting whole sentences. Lesson 54 used sentence selection. This lesson asks
a local language model to write a new sentence, then checks the candidate
against the original record.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- keep source records, prompts, and outputs tied to durable IDs;
- count the model's input tokens before generation;
- compare generated summaries with a first-sentence baseline;
- separate compression from source support and useful coverage;
- inspect names, numbers, and explicit limits without calling them proof; and
- leave factual and editorial approval to a recorded human review.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the local model

The model files were downloaded before rendering at the revision recorded in
`data/nlg-models.csv`. The code below uses `huggingfaceR` to load that local
snapshot. Offline mode is set by `R/use-nlg.R`, so a missing model stops the
lesson instead of starting a network download.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(huggingfaceR)
  library(knitr)
  library(purrr)
  library(reticulate)
  library(stringr)
  library(tibble)
  library(tidyr)
})

source("R/use-nlg.R")

summary_model <- load_nlg_pipeline(
  "qwen_1_5b_instruct",
  "text-generation"
)

model_record <- summary_model$metadata |>
  select(model_id, revision, license, task_scope)

kable(
  model_record,
  col.names = c("Model", "Revision", "License", "Use in this course"),
  caption = "Pinned local model used for the generated summaries",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "dplyr", "huggingfaceR", "knitr", "purrr",
    "reticulate", "stringr", "tibble", "tidyr"
  ) %in% loadedNamespaces()),
  identical(model_record$model_id, "Qwen/Qwen2.5-1.5B-Instruct"),
  identical(
    model_record$revision,
    "989aa7980e4cf806f80c7fef2b1adb7bc71aa306"
  ),
  identical(model_record$license, "Apache-2.0")
)
```

The model card documents an instruction-tuned causal language model. That
description and license do not establish that its summaries are accurate.

## Keep the source beside the brief

These records are fictional teaching examples. They were written for this
lesson and are not a random sample, a held-out benchmark, or evidence about
archive services elsewhere.

Each brief states information that matters to the reader. SUM-01 needs the
service, schedule, and boundary on volunteer help. SUM-02 needs the event,
registration rule, and restricted-storage limit.

```{r}
summary_sources <- tribble(
  ~source_id, ~brief, ~source_text,
  "SUM-01",
  "One sentence for staff; retain the service and its boundary.",
  paste(
    "The Riverton library opens a new job-search room on Monday.",
    "The room has 12 computers.",
    "Volunteers offer free help on Tuesdays, but they do not complete",
    "applications for visitors.",
    "The pilot runs for six weeks."
  ),
  "SUM-02",
  "One sentence for visitors; retain the event and both access conditions.",
  paste(
    "The neighborhood archive will open Room 5 on 14 September for a",
    "catalog workshop led by Maya Chen.",
    "Registration is free, but visitors must bring photo identification.",
    "The session does not include access to restricted storage."
  )
)

kable(
  summary_sources,
  col.names = c("Source ID", "Brief", "Source text"),
  caption = "Constructed source records and their summary briefs",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(summary_sources$source_id, c("SUM-01", "SUM-02")),
  !anyDuplicated(summary_sources$source_id),
  all(str_count(summary_sources$source_text, "[.!?]") >= 3L)
)
```

## Check the input budget before generation

A model token is a tokenizer unit, not necessarily a word. The prompt includes
the instruction, brief, and source text, so all three count against the input
budget. This demonstration uses a 256-token ceiling and refuses to truncate.

```{r}
summary_system <- paste(
  "Write one sentence for the stated brief.",
  "Use only information in the source.",
  "Do not add a cause, recommendation, or new event."
)

summary_inputs <- summary_sources |>
  mutate(
    user_prompt = paste0(
      "BRIEF: ", brief,
      "\nSOURCE ID: ", source_id,
      "\nSOURCE: ", source_text
    ),
    model_prompt = map2_chr(
      summary_system,
      user_prompt,
      \(system, user) {
        nlg_chat_prompt(
          summary_model$tokenizer,
          system,
          user
        )
      }
    ),
    input_tokens = map_int(
      model_prompt,
      \(prompt) nlg_token_count(summary_model$tokenizer, prompt)
    )
  )

kable(
  summary_inputs |>
    select(source_id, input_tokens),
  col.names = c("Source ID", "Complete prompt tokens"),
  caption = "Tokenizer counts before summarization",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(summary_inputs), 2L),
  all(summary_inputs$input_tokens > 0L),
  all(summary_inputs$input_tokens <= 256L),
  all(str_detect(summary_inputs$model_prompt, "SOURCE ID"))
)
```

No text is cut to fit. A longer document would need a stated policy such as
sectioning or hierarchical summarization before the model call.

## Generate the candidate summaries

Greedy decoding chooses the highest-scoring next token at each step. The output
limit is 72 new tokens. The call returns the generated text and token-level
stopping information from the local model.

```{r}
summary_results <- summary_inputs |>
  mutate(
    generation = map(
      model_prompt,
      \(prompt) {
        nlg_generate(
          summary_model,
          prompt,
          max_new_tokens = 72L,
          repetition_penalty = 1.05
        )
      }
    ),
    generated_summary = map_chr(generation, "text"),
    output_tokens = map_int(generation, "output_tokens"),
    ended_by_eos = map_lgl(generation, "ended_by_eos"),
    hit_token_cap = map_lgl(generation, "hit_token_cap"),
    first_sentence_baseline = map_chr(
      str_split(source_text, "(?<=[.!?])\\s+"),
      \(sentences) sentences[[1]]
    )
  )

kable(
  summary_results |>
    select(source_id, generated_summary, first_sentence_baseline),
  col.names = c(
    "Source ID",
    "Generated abstractive candidate",
    "First-sentence teaching baseline"
  ),
  caption = "Generated summaries beside an extractive comparator",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(nzchar(summary_results$generated_summary)),
  all(summary_results$output_tokens > 0L),
  all(summary_results$output_tokens <= 72L),
  all(summary_results$ended_by_eos | summary_results$hit_token_cap),
  any(
    summary_results$generated_summary !=
      summary_results$first_sentence_baseline
  )
)
```

The first-sentence baseline is available before any model output is read. It is
not a fair contest with a matched length budget; it shows the difference
between selecting source wording and generating a new account.

## Measure shortening without calling it truth

The next table counts whitespace-separated words. That rule is easy to inspect,
but it is different from the model tokenizer used for the generation budget.
A compression ratio below 1 means the candidate is shorter. It says nothing
about whether the candidate kept the needed facts.

```{r}
count_words <- function(text) {
  str_count(str_squish(text), "\\S+")
}

compression <- summary_results |>
  transmute(
    source_id,
    source_words = map_int(source_text, count_words),
    generated_words = map_int(generated_summary, count_words),
    first_sentence_words = map_int(
      first_sentence_baseline,
      count_words
    ),
    generated_to_source = generated_words / source_words
  )

kable(
  compression,
  digits = 2,
  col.names = c(
    "Source ID",
    "Source words",
    "Generated words",
    "Baseline words",
    "Generated/source ratio"
  ),
  caption = "Length diagnostics use whitespace-separated words",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(compression$source_words > compression$generated_words),
  all(compression$generated_to_source > 0),
  all(compression$generated_to_source < 1)
)
```

The generated sentence can be short because it chose the right content, because
it omitted necessary content, or both. Length cannot separate those cases.

## Inspect facts chosen before generation

Rosa writes a checklist before reading the output. The regular expressions
below flag visible strings related to that checklist. A missing flag points to
an omission worth reading. A present flag cannot prove that the name, number,
or negative statement has the right role or scope.

```{r}
summary_checks <- tribble(
  ~source_id, ~check_id, ~needed_information, ~pattern,
  "SUM-01", "F1", "12 computers", "\\b12\\b",
  "SUM-01", "F2", "Tuesday help", "\\bTuesday",
  "SUM-01", "F3", "volunteers do not complete applications",
  "\\b(not|no|never|without|do not|does not)\\b",
  "SUM-01", "F4", "six-week pilot", "\\b(six|6)[ -]week",
  "SUM-02", "F5", "Room 5", "\\bRoom 5\\b",
  "SUM-02", "F6", "14 September", "\\b14\\b.*\\bSeptember\\b",
  "SUM-02", "F7", "Maya Chen", "\\bMaya Chen\\b",
  "SUM-02", "F8", "photo identification", "\\b(photo|identification|ID)\\b",
  "SUM-02", "F9", "no restricted-storage access",
  "\\b(not|no|without|does not)\\b"
)

summary_flags <- summary_checks |>
  left_join(
    summary_results |>
      select(source_id, generated_summary),
    by = "source_id"
  ) |>
  mutate(
    text_flag = str_detect(
      generated_summary,
      regex(pattern, ignore_case = TRUE)
    )
  ) |>
  select(source_id, check_id, needed_information, text_flag)

summary_actions <- summary_flags |>
  group_by(source_id) |>
  summarise(
    flagged_items = sum(!text_flag),
    action = if_else(
      flagged_items > 0L,
      "revise",
      "send to human source review"
    ),
    .groups = "drop"
  )

kable(
  summary_flags,
  col.names = c(
    "Source ID",
    "Check",
    "Information chosen before generation",
    "Text flag present"
  ),
  caption = "Surface flags for facts the brief asked the summary to cover",
  row.names = FALSE
)

kable(
  summary_actions,
  col.names = c("Source ID", "Missing surface flags", "Next action"),
  caption = "Missing brief items send a generated summary back for revision",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(summary_flags$check_id, paste0("F", 1:9)),
  all(summary_flags$source_id %in% summary_sources$source_id),
  is.logical(summary_flags$text_flag),
  length(summary_flags$text_flag) == 9L,
  any(!summary_flags$text_flag),
  all(summary_actions$action %in% c(
    "revise",
    "send to human source review"
  ))
)
```

This check is intentionally incomplete. It cannot detect a changed relation, a
number attached to the wrong thing, an unsupported cause, or a fluent
generalization. Those require reading the source and candidate together.
In this render, the first candidate changes the boundary on volunteer help.
The second uses negative wording while changing the relationship to restricted
storage. The surface table catches part of the problem, and source comparison
catches the rest.
Likewise, n-gram overlap metrics such as ROUGE measure reference overlap under a
specified normalization rule. They do not measure the share of claims supported
by the source.

## Record the review boundary

Execution can establish that a model ran, stayed inside the configured budgets,
and returned text. It cannot grant factual or editorial approval.

```{r}
summary_review <- tibble(
  review = c(
    "local model execution",
    "prompt input within 256 tokens",
    "output stopped at EOS or token cap",
    "human source-support review",
    "human usefulness and prose review"
  ),
  status = c(
    "passed",
    "passed",
    "passed",
    "pending",
    "pending"
  )
)

kable(
  summary_review,
  col.names = c("Review gate", "Status"),
  caption = "Automatic execution checks remain separate from human review",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    summary_review$status,
    c("passed", "passed", "passed", "pending", "pending")
  )
)
```

A publishable summary needs a reader to trace each claim to the source, decide
whether the brief's essential information survived, and reject unsupported
additions. A fluent one-sentence result is only a candidate.

## What to remember

- Abstractive summarization generates wording; extractive summarization selects
  source text.
- Count the complete prompt with the model tokenizer and do not hide truncation.
- Compression describes length, not coverage or source support.
- Check omissions and additions against a brief written before generation.
- Names, numbers, and negation flags are prompts for review, not semantic proof.
- Keep model execution separate from human factual and editorial approval.

## Sources

- [Maynez et al. on faithfulness and factuality in abstractive summarization](https://aclanthology.org/2020.acl-main.173/)
- [Lin on ROUGE and reference n-gram overlap](https://aclanthology.org/W04-1013/)
- [Pagnoni et al. on classes of factual errors in summaries](https://aclanthology.org/2021.naacl-main.383/)
- [Qwen2.5-1.5B-Instruct model card](https://huggingface.co/Qwen/Qwen2.5-1.5B-Instruct)
- [Transformers text-generation controls](https://huggingface.co/docs/transformers/v4.57.1/en/main_classes/text_generation)
