---
title: "Changing the wording without changing the instruction"
subtitle: "Separate a real rewrite from a meaning-preservation check"
description: "Learn how to generate paraphrases and inspect names, numbers, negation, and modality without treating word overlap as equivalence."
categories: [natural language generation, paraphrasing, evaluation]
---

June edits public notices for an archive. Repeating the same sentence in every
channel sounds stiff, but a rewrite that changes who may enter or what visitors
must bring creates a new instruction.

For this lesson, a **paraphrase** changes the wording while preserving the
participants, quantities, negation, scope, and obligations that matter for the
notice. Linguistic research sometimes uses a looser idea of approximate
meaning. An operational notice needs a stricter rule because readers may act on
one word such as `may`, `must`, or `not`.

The primary aim of paraphrasing is to preserve meaning while changing wording;
summarization primarily compresses or selects content, and simplification
primarily reduces complexity or register. A real output can serve more than one
aim, so these are not mutually exclusive classes.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- generate paraphrase candidates with a pinned local model;
- keep an unchanged-input baseline beside the generated wording;
- measure visible wording change without calling it semantic equivalence;
- probe names, numbers, negation, and modality;
- preserve model failures instead of filtering them away; and
- record human meaning review as pending.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the same fixed checkpoint

The checkpoint is shared with the summarization lesson, but this page gives it
a different task, prompt, baseline, and review rule. The model is loaded from
the prepared local directory through `huggingfaceR`.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(huggingfaceR)
  library(purrr)
  library(reticulate)
  library(stringr)
  library(tibble)
  library(tidyr)
})

source("R/use-nlg.R")

paraphrase_model <- load_nlg_pipeline(
  "qwen_1_5b_instruct",
  "text-generation"
)

model_record <- paraphrase_model$metadata |>
  select(model_id, revision, license)
```

```{r}
#| echo: false

knitr::kable(
  model_record,
  col.names = c("Model", "Revision", "License"),
  caption = "Pinned local model used for paraphrase generation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "dplyr", "huggingfaceR", "purrr",
    "reticulate", "stringr", "tibble", "tidyr"
  ) %in% loadedNamespaces()),
  identical(model_record$model_id, "Qwen/Qwen2.5-1.5B-Instruct"),
  identical(
    model_record$revision,
    "989aa7980e4cf806f80c7fef2b1adb7bc71aa306"
  ),
  identical(model_record$license, "Apache-2.0")
)
```

## Write the preservation rule first

The four notices below are constructed behavioral probes. They were chosen to
expose common changes in operational meaning. They are not a random test set,
and no population accuracy will be estimated from them.

`protected_name`, `modality_class`, and `modality_pattern` are blank when a
probe does not need that check. A copy of the source is kept as the baseline.
Copying preserves the original wording and facts, but it does not perform the
requested rewrite.

```{r}
paraphrase_inputs <- tribble(
  ~probe_id, ~source_text, ~protected_name, ~requires_negation,
  ~modality_class, ~modality_pattern,
  "PAR-01",
  "The clinic sent 24 reminder letters on Monday.",
  "",
  FALSE,
  "",
  "",
  "PAR-02",
  "Visitors may bring one notebook to the reading room.",
  "",
  FALSE,
  "permission",
  "\\b(may|can|allowed|permitted)\\b",
  "PAR-03",
  paste(
    "Maya Chen did not cancel the 14 September workshop;",
    "she moved it from Room 2 to Room 5."
  ),
  "Maya Chen",
  TRUE,
  "",
  "",
  "PAR-04",
  "Only archive staff must sign the restricted-room log.",
  "",
  FALSE,
  "obligation",
  "\\b(must|required|obliged|has to)\\b"
)
```

```{r}
#| echo: false

knitr::kable(
  paraphrase_inputs,
  col.names = c(
    "Probe ID",
    "Source notice",
    "Protected name",
    "Negation required",
    "Modality class",
    "Accepted surface markers"
  ),
  caption = "Constructed paraphrase probes and predeclared invariants",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(paraphrase_inputs$probe_id, paste0("PAR-0", 1:4)),
  !anyDuplicated(paraphrase_inputs$probe_id),
  identical(sum(paraphrase_inputs$requires_negation), 1L),
  identical(
    paraphrase_inputs$modality_class,
    c("", "permission", "", "obligation")
  )
)
```

## Generate candidates without silent truncation

The complete chat-formatted prompt must stay below 192 input tokens. Greedy
decoding and a 72-token output cap make the call repeatable, but deterministic
generation can still produce a wrong paraphrase.

```{r}
paraphrase_system <- paste(
  "Rewrite the notice in different words.",
  "Preserve every participant, quantity, date, negation, scope,",
  "permission, and obligation.",
  "Return the rewritten notice only."
)

paraphrase_requests <- paraphrase_inputs |>
  mutate(
    user_prompt = paste0(
      "PROBE ID: ", probe_id,
      "\nNOTICE: ", source_text
    ),
    model_prompt = map_chr(
      user_prompt,
      \(user) {
        nlg_chat_prompt(
          paraphrase_model$tokenizer,
          paraphrase_system,
          user
        )
      }
    ),
    input_tokens = map_int(
      model_prompt,
      \(prompt) nlg_token_count(paraphrase_model$tokenizer, prompt)
    )
  )
```

```{r}
#| echo: false

knitr::kable(
  paraphrase_requests |>
    select(probe_id, input_tokens),
  col.names = c("Probe ID", "Complete prompt tokens"),
  caption = "Paraphrase input budgets checked before generation",
  row.names = FALSE
)
```

```{r}
#| include: false

nlg_assert_input_budget(
  paraphrase_requests$input_tokens,
  max_input_tokens = 192L,
  item_ids = paraphrase_requests$probe_id
)
stopifnot(all(paraphrase_requests$input_tokens > 0L))
```

Only after that assertion succeeds does the lesson make any generation call.

```{r}
paraphrase_results <- paraphrase_requests |>
  mutate(
    generation = map(
      model_prompt,
      \(prompt) {
        nlg_generate(
          paraphrase_model,
          prompt,
          max_new_tokens = 72L,
          repetition_penalty = 1.05
        )
      }
    ),
    candidate = map_chr(generation, "text"),
    candidate_display = str_replace_all(
      candidate,
      "\\s*\\n+\\s*",
      " / "
    ),
    output_tokens = map_int(generation, "output_tokens"),
    ended_by_eos = map_lgl(generation, "ended_by_eos"),
    hit_token_cap = map_lgl(generation, "hit_token_cap"),
    copy_baseline = source_text
  )
```

```{r}
#| echo: false

knitr::kable(
  paraphrase_results |>
    select(probe_id, source_text, candidate_display, copy_baseline),
  col.names = c(
    "Probe ID",
    "Source",
    "Generated candidate",
    "Unchanged-input baseline"
  ),
  caption = "Generated paraphrases retain failures and copies for inspection",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(paraphrase_results$input_tokens <= 192L),
  all(nzchar(paraphrase_results$candidate)),
  all(paraphrase_results$output_tokens > 0L),
  all(paraphrase_results$output_tokens <= 72L),
  all(
    paraphrase_results$ended_by_eos |
      paraphrase_results$hit_token_cap
  ),
  any(paraphrase_results$candidate != paraphrase_results$source_text)
)
```

The copy baseline should pass every literal preservation check. It also fails
the user's request for different wording. The generated column has to solve
both parts.

## Measure wording change

This diagnostic lowercases each string, extracts letter and number sequences
with the ICU regular-expression engine used by `stringr`, and computes Jaccard
overlap on the unique token sets. An overlap of 1 means the sets match. A lower
value means some visible wording changed.

```{r}
token_set <- function(text) {
  str_extract_all(
    str_to_lower(text, locale = "en"),
    "[\\p{L}\\p{N}]+"
  )[[1]] |>
    unique()
}

jaccard_overlap <- function(left, right) {
  left_tokens <- token_set(left)
  right_tokens <- token_set(right)
  length(intersect(left_tokens, right_tokens)) /
    length(union(left_tokens, right_tokens))
}

wording_diagnostics <- paraphrase_results |>
  transmute(
    probe_id,
    exact_copy = candidate == source_text,
    unique_token_jaccard = map2_dbl(
      source_text,
      candidate,
      jaccard_overlap
    )
  )
```

```{r}
#| echo: false

knitr::kable(
  wording_diagnostics,
  digits = 2,
  col.names = c(
    "Probe ID",
    "Exact copy",
    "Unique-token Jaccard overlap"
  ),
  caption = "Wording diagnostics do not decide whether meaning survived",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(
    wording_diagnostics$unique_token_jaccard >= 0 &
      wording_diagnostics$unique_token_jaccard <= 1
  ),
  any(!wording_diagnostics$exact_copy)
)
```

A low overlap can come from a good synonym or a damaging change. A high overlap
can hide a removed `not`. The score describes token sets, not equivalence.

## Check visible invariants

The code now looks for exact Arabic numerals, the protected name, an explicit
negation marker, and a short list of surface markers for the predeclared
modality class. These conservative screens catch some preventable mistakes.
They also miss valid alternatives: `24` may become `twenty-four`, and `must`
may become `have to`. A screen result cannot establish whether the meaning
survived.

```{r}
extract_numbers <- function(text) {
  str_extract_all(text, "\\b\\d+\\b")[[1]] |>
    sort()
}

has_negation <- function(text) {
  str_detect(
    text,
    regex(
      "\\b(no|not|never|without|cannot|can't|didn't|did not)\\b",
      ignore_case = TRUE
    )
  )
}

invariant_checks <- paraphrase_results |>
  mutate(
    numbers_preserved = map2_lgl(
      source_text,
      candidate,
      \(source, output) {
        source_numbers <- extract_numbers(source)
        if (length(source_numbers) == 0L) {
          NA
        } else {
          identical(source_numbers, extract_numbers(output))
        }
      }
    ),
    name_present = map2_lgl(
      protected_name,
      candidate,
      \(name, output) {
        if (!nzchar(name)) {
          NA
        } else {
          str_detect(output, fixed(name))
        }
      }
    ),
    negation_visible = map2_lgl(
      requires_negation,
      candidate,
      \(required, output) {
        if (!required) NA else has_negation(output)
      }
    ),
    modal_visible = map2_lgl(
      modality_pattern,
      candidate,
      \(pattern, output) {
        if (!nzchar(pattern)) {
          NA
        } else {
          str_detect(
            output,
            regex(pattern, ignore_case = TRUE)
          )
        }
      }
    )
  ) |>
  select(
    probe_id,
    numbers_preserved,
    name_present,
    negation_visible,
    modal_visible
  )

invariant_check_display <- invariant_checks |>
  mutate(
    across(
      c(
        numbers_preserved,
        name_present,
        negation_visible,
        modal_visible
      ),
      nlg_screen_label
    )
  )

screen_actions <- invariant_checks |>
  rowwise() |>
  mutate(
    applicable_screens = sum(!is.na(c_across(c(
      numbers_preserved,
      name_present,
      negation_visible,
      modal_visible
    )))),
    flagged_screens = sum(c_across(c(
      numbers_preserved,
      name_present,
      negation_visible,
      modal_visible
    )) %in% FALSE)
  ) |>
  ungroup() |>
  select(probe_id, applicable_screens, flagged_screens)

paraphrase_actions <- wording_diagnostics |>
  left_join(screen_actions, by = "probe_id") |>
  mutate(
    action = case_when(
      exact_copy ~ "revise wording",
      flagged_screens > 0L ~ "hold for human meaning review",
      TRUE ~ "send to human meaning review"
    )
  ) |>
  select(probe_id, flagged_screens, action)
```

```{r}
#| echo: false

knitr::kable(
  invariant_check_display,
  col.names = c(
    "Probe ID",
    "Arabic-numeral screen",
    "Protected-name screen",
    "Negation screen",
    "Modality screen"
  ),
  caption = "Conservative screens for predeclared paraphrase invariants",
  row.names = FALSE
)

knitr::kable(
  paraphrase_actions,
  col.names = c("Probe ID", "Screen flags", "Next action"),
  caption = "Screen flags route candidates to human meaning review",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(invariant_checks$probe_id, paraphrase_inputs$probe_id),
  all(vapply(invariant_checks[-1], is.logical, logical(1))),
  identical(nrow(invariant_checks), 4L),
  any(paraphrase_actions$flagged_screens > 0L),
  all(is.na(
    invariant_checks$name_present[
      !nzchar(paraphrase_inputs$protected_name)
    ]
  )),
  all(is.na(
    invariant_checks$modal_visible[
      !nzchar(paraphrase_inputs$modality_pattern)
    ]
  ))
)
```

Any applicable `FALSE` value is a reason to hold the candidate for human
review, not proof that its meaning changed. In this render, reading source and
output together shows that the stress probe echoes prompt labels and drops Room
2. The obligation probe replaces `must sign` with `authorized to sign`, which
raises a substantive obligation concern. Its modal screen also flags the row,
but the same finite screen would miss or flag some equivalent phrasings, so the
human comparison carries the conclusion.

## Keep the decision honest

The model has completed generation, and the automatic checks have completed
their limited jobs. No human has approved the candidates.

```{r}
paraphrase_review <- tibble(
  review = c(
    "local model execution",
    "complete prompt within 192 tokens",
    "wording-change diagnostic",
    "surface invariant checks",
    "human meaning-preservation review",
    "human prose review"
  ),
  status = c(
    sprintf(
      "completed: %d candidates",
      sum(nzchar(paraphrase_results$candidate))
    ),
    sprintf(
      "within budget: %d/%d candidates",
      sum(paraphrase_results$input_tokens <= 192L),
      nrow(paraphrase_results)
    ),
    sprintf(
      "computed: %d/%d candidates changed wording",
      sum(!wording_diagnostics$exact_copy),
      nrow(wording_diagnostics)
    ),
    sprintf(
      "held: %d/%d candidates flagged",
      sum(paraphrase_actions$flagged_screens > 0L),
      nrow(paraphrase_actions)
    ),
    "pending",
    "pending"
  )
)
```

```{r}
#| echo: false

knitr::kable(
  paraphrase_review,
  col.names = c("Review gate", "Status"),
  caption = "A generated rewrite is not an approved paraphrase",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  !any(paraphrase_review$status == "passed"),
  identical(tail(paraphrase_review$status, 2L), c("pending", "pending"))
)
```

June can accept a candidate only after reading it beside the notice and applying
the rule stated at the start. Generation, visible wording change, and invariant
flags are evidence for that review. None replaces it.

## What to remember

- A paraphrase changes wording while preserving the information that matters.
- Copying is a useful preservation baseline, but it is not a successful rewrite.
- Token overlap measures wording, not semantic equivalence.
- Check names, numbers, negation, scope, permission, and obligation separately.
- Keep failed model outputs visible when they reveal a broken invariant.
- Record human meaning and prose review apart from automatic execution.

## Sources

- [Bhagat and Hovy on definitions and varieties of paraphrase](https://aclanthology.org/J13-3001/)
- [Qwen2.5-1.5B-Instruct model card](https://huggingface.co/Qwen/Qwen2.5-1.5B-Instruct)
- [Transformers text-generation controls](https://huggingface.co/docs/transformers/v4.57.1/en/main_classes/text_generation)
- [Transformers tokenizer documentation](https://huggingface.co/docs/transformers/v4.57.1/en/main_classes/tokenizer)
