---
title: "Building a document one section at a time"
subtitle: "Plan evidence, carry continuity, inspect stopping, and hold for review"
description: "Learn a three-section local generation workflow with explicit evidence scopes, token budgets, continuity state, and publication gates."
categories: [natural language generation, long-text generation, evaluation]
---

Kai needs a short public update about a fictional seed-sharing pilot. Asking a
small model to "write the whole report" would combine document planning,
evidence selection, continuity, and prose generation in one opaque call.

This lesson separates those jobs. A human-written plan assigns evidence to
three sections. The local model generates each section. A compact,
evidence-checked continuity state moves forward without copying the whole draft
back into the prompt.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- plan a three-section document before generating prose;
- give each section a bounded evidence ledger and output budget;
- account for prompt and output tokens in a decoder-only context window;
- carry compact, checked continuity state between section calls;
- distinguish end-of-sequence stopping from a token-cap stop;
- inspect repetition, unsupported numbers, and evidence coverage; and
- reject the assembled draft for publication until human review is complete.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the local causal model

The course helper loads a prepared local snapshot through `huggingfaceR`. The
model is a decoder-only causal language model, so prompt tokens and newly
generated tokens share the same context window.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(huggingfaceR)
  library(jsonlite)
  library(knitr)
  library(purrr)
  library(reticulate)
  library(stringr)
  library(tibble)
  library(tidyr)
})

source("R/use-nlg.R")

long_model <- load_nlg_pipeline(
  "qwen_1_5b_instruct",
  "text-generation"
)

model_config_path <- file.path(
  "data-raw",
  ".cache",
  "nlg-models",
  long_model$metadata$local_directory,
  "config.json"
)
model_config <- fromJSON(model_config_path)
context_limit <- as.integer(model_config$max_position_embeddings)

model_record <- long_model$metadata |>
  transmute(
    model = model_id,
    revision,
    license,
    architecture = "decoder-only causal language model",
    context_limit
  )

kable(
  model_record,
  col.names = c(
    "Model",
    "Revision",
    "License",
    "Architecture",
    "Context limit"
  ),
  caption = "Pinned model and its shared causal context window",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "dplyr", "huggingfaceR", "jsonlite", "knitr", "purrr",
    "reticulate", "stringr", "tibble", "tidyr"
  ) %in% loadedNamespaces()),
  identical(
    model_record$model,
    "Qwen/Qwen2.5-1.5B-Instruct"
  ),
  identical(
    model_record$revision,
    "989aa7980e4cf806f80c7fef2b1adb7bc71aa306"
  ),
  identical(model_record$license, "Apache-2.0"),
  file.exists(model_config_path),
  context_limit >= 1024L
)
```

For this architecture, enforce
`prompt_tokens + max_new_tokens <= context_limit`.

An encoder-decoder model has separate encoder-input and decoder-output limits.
That distinction matters when moving the workflow to another checkpoint.

## Plan before writing

The evidence below is fictional demonstration material. It is not a survey
estimate, a program evaluation, or a claim about a real seed library.

Each evidence row receives an ID. The plan assigns those IDs to one section and
states the section's job before any prose is generated.

```{r}
evidence_ledger <- tribble(
  ~evidence_id, ~evidence,
  "E01", "The pilot registered 48 residents.",
  "E02", "The pilot completed 36 pickup appointments in April.",
  "E03", "12 volunteers staffed the collection table.",
  "E04", "Volunteers did not guarantee that any seed would grow.",
  "E05", "29 participants requested an evening pickup.",
  "E06",
  paste(
    "Staff will decide whether to add one Thursday evening;",
    "no decision has been made."
  )
)

document_plan <- tribble(
  ~section_id, ~heading, ~purpose, ~evidence_ids, ~max_new_tokens,
  ~continuity_in, ~continuity_out,
  "SEC-01",
  "Pilot activity",
  "Describe participation, completed pickups, and staffing.",
  c("E01", "E02", "E03"),
  120L,
  "This is the opening section.",
  paste(
    "The pilot registered 48 residents, completed 36 April pickups,",
    "and used 12 volunteers."
  ),
  "SEC-02",
  "Participant request",
  "Describe the request for an evening pickup without claiming a decision.",
  c("E05"),
  100L,
  paste(
    "The pilot registered 48 residents, completed 36 April pickups,",
    "and used 12 volunteers."
  ),
  "Twenty-nine participants requested an evening pickup.",
  "SEC-03",
  "Decision and limit",
  "State the pending Thursday decision and the volunteer guarantee limit.",
  c("E04", "E06"),
  120L,
  "Twenty-nine participants requested an evening pickup.",
  paste(
    "A Thursday evening remains undecided, and volunteers did not",
    "guarantee seed growth."
  )
)

kable(
  document_plan |>
    select(
      section_id,
      heading,
      purpose,
      evidence_ids,
      max_new_tokens
    ) |>
    mutate(evidence_ids = map_chr(evidence_ids, paste, collapse = ", ")),
  col.names = c(
    "Section ID",
    "Heading",
    "Section purpose",
    "Evidence IDs",
    "Output budget"
  ),
  caption = "A three-section document plan fixes scope before generation",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(document_plan$section_id, paste0("SEC-0", 1:3)),
  !anyDuplicated(evidence_ledger$evidence_id),
  !anyDuplicated(document_plan$section_id),
  setequal(unlist(document_plan$evidence_ids), paste0("E0", 1:6)),
  identical(
    document_plan$continuity_in[-1],
    document_plan$continuity_out[-nrow(document_plan)]
  )
)
```

The continuity state is a checked restatement of assigned evidence. It is not a
model-generated summary of the previous section. This prevents an unsupported
detail in one section from becoming accepted context for the next.

## Build section prompts and budgets

Each prompt contains one section purpose, only its assigned evidence, and the
compact incoming state. Raw previous prose is not concatenated.

```{r}
long_system <- paste(
  "Write one section of a public program update in 2 or 3 sentences.",
  "Use only the listed evidence.",
  "Do not add examples, causes, recommendations, people, places, or outcomes.",
  "Do not write a heading."
)

section_requests <- document_plan |>
  mutate(
    scoped_evidence = map(
      evidence_ids,
      \(ids) {
        evidence_ledger |>
          filter(evidence_id %in% ids) |>
          transmute(line = paste0(evidence_id, ": ", evidence)) |>
          pull(line) |>
          paste(collapse = "\n")
      }
    ),
    user_prompt = pmap_chr(
      list(section_id, purpose, continuity_in, scoped_evidence),
      \(section_id, purpose, continuity_in, scoped_evidence) {
        paste0(
          "SECTION ID: ", section_id,
          "\nPURPOSE: ", purpose,
          "\nCONTINUITY STATE: ", continuity_in,
          "\nSCOPED EVIDENCE:\n", scoped_evidence
        )
      }
    ),
    model_prompt = map_chr(
      user_prompt,
      \(user) {
        nlg_chat_prompt(
          long_model$tokenizer,
          long_system,
          user
        )
      }
    ),
    prompt_tokens = map_int(
      model_prompt,
      \(prompt) nlg_token_count(long_model$tokenizer, prompt)
    ),
    planned_total_tokens = prompt_tokens + max_new_tokens,
    fits_context = planned_total_tokens <= context_limit
  )

kable(
  section_requests |>
    select(
      section_id,
      prompt_tokens,
      max_new_tokens,
      planned_total_tokens,
      fits_context
    ),
  col.names = c(
    "Section ID",
    "Prompt tokens",
    "Maximum new tokens",
    "Planned total",
    "Fits context"
  ),
  caption = "Decoder-only token accounting for every section call",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(section_requests$prompt_tokens > 0L),
  all(section_requests$max_new_tokens > 0L),
  all(section_requests$fits_context),
  all(
    section_requests$planned_total_tokens ==
      section_requests$prompt_tokens +
        section_requests$max_new_tokens
  )
)
```

## Generate all three sections

The loop below makes three separate neural generation calls. Greedy decoding
keeps this demonstration repeatable. Literature documents repetitive
degeneration as a risk of likelihood-maximizing decoding in open-ended tasks;
it is an empirical tendency, not a guarantee that every greedy output loops.

```{r}
section_generations <- section_requests |>
  mutate(
    generation = map2(
      model_prompt,
      max_new_tokens,
      \(prompt, budget) {
        nlg_generate(
          long_model,
          prompt,
          max_new_tokens = budget,
          repetition_penalty = 1.05
        )
      }
    ),
    generated_text = map_chr(generation, "text"),
    output_tokens = map_int(generation, "output_tokens"),
    ended_by_eos = map_lgl(generation, "ended_by_eos"),
    hit_token_cap = map_lgl(generation, "hit_token_cap")
  )

kable(
  section_generations |>
    select(section_id, heading, generated_text),
  col.names = c("Section ID", "Heading", "Generated section"),
  caption = "Three separately generated sections",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(nzchar(section_generations$generated_text)),
  all(section_generations$output_tokens > 0L),
  all(
    section_generations$output_tokens <=
      section_generations$max_new_tokens
  ),
  all(
    section_generations$ended_by_eos |
      section_generations$hit_token_cap
  )
)
```

## Inspect architectural stopping

A final period is a surface character. It does not reveal why generation
stopped. The helper checks the final generated token ID against the model's EOS
configuration and separately records whether the output reached the token cap.

```{r}
stopping_diagnostics <- section_generations |>
  transmute(
    section_id,
    output_tokens,
    output_budget = max_new_tokens,
    ended_by_eos,
    hit_token_cap,
    ends_with_punctuation = str_detect(
      str_trim(generated_text),
      "[.!?]$"
    )
  )

kable(
  stopping_diagnostics,
  col.names = c(
    "Section ID",
    "Output tokens",
    "Output budget",
    "EOS emitted",
    "Hit token cap",
    "Ends with punctuation"
  ),
  caption = "Token-level stopping evidence beside a surface heuristic",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(
    stopping_diagnostics$ended_by_eos |
      stopping_diagnostics$hit_token_cap
  ),
  !any(
    stopping_diagnostics$ended_by_eos &
      stopping_diagnostics$hit_token_cap
  )
)
```

EOS says the model selected a termination token. It does not establish that the
section is complete, accurate, or ready to publish.

## Check repetition and scoped evidence

The distinct trigram ratio divides unique three-token sequences by all
three-token sequences after lowercasing and extracting letter and number
sequences. Values nearer 1 show fewer exact repeated trigrams. No universal
publication threshold is asserted.

The evidence flags are simpler: each planned fact has a surface pattern. A
missing pattern is a reason to revise. A present pattern cannot prove that the
fact has the correct relation or scope.

```{r}
distinct_ngram_ratio <- function(text, n = 3L) {
  tokens <- str_extract_all(
    str_to_lower(text, locale = "en"),
    "[\\p{L}\\p{N}]+"
  )[[1]]

  if (length(tokens) < n) {
    return(1)
  }

  starts <- seq_len(length(tokens) - n + 1L)
  ngrams <- map_chr(
    starts,
    \(start) paste(tokens[start:(start + n - 1L)], collapse = " ")
  )
  length(unique(ngrams)) / length(ngrams)
}

evidence_patterns <- tribble(
  ~evidence_id, ~pattern,
  "E01", "\\b48\\b",
  "E02", "\\b36\\b.*\\bApril\\b|\\bApril\\b.*\\b36\\b",
  "E03", "\\b(12|twelve)\\b.*\\bvolunteer",
  "E04", "\\b(not|no|never|without|did not)\\b.*\\bguarantee",
  "E05",
  paste(
    "\\b(29|twenty-nine|twenty nine)\\b.*\\bevening",
    "|\\bevening\\b.*\\b(29|twenty-nine|twenty nine)\\b"
  ),
  "E06", "\\bThursday\\b.*\\b(no|not|undecided|pending)"
)

section_diagnostics <- section_generations |>
  transmute(
    section_id,
    distinct_trigram_ratio = map_dbl(
      generated_text,
      distinct_ngram_ratio
    ),
    source_numbers = map(
      scoped_evidence,
      \(text) sort(str_extract_all(text, "\\b\\d+\\b")[[1]])
    ),
    output_numbers = map(
      generated_text,
      \(text) sort(str_extract_all(text, "\\b\\d+\\b")[[1]])
    ),
    unsupported_number = map2_lgl(
      source_numbers,
      output_numbers,
      \(source, output) !all(output %in% source)
    ),
    unsupported_evaluation = str_detect(
      generated_text,
      regex(
        "\\b(successfully|effective|improved|successful)\\b",
        ignore_case = TRUE
      )
    )
  )

coverage_diagnostics <- document_plan |>
  select(section_id, evidence_ids) |>
  unnest_longer(evidence_ids, values_to = "evidence_id") |>
  left_join(evidence_patterns, by = "evidence_id") |>
  left_join(
    section_generations |>
      select(section_id, generated_text),
    by = "section_id"
  ) |>
  mutate(
    surface_flag = str_detect(
      generated_text,
      regex(pattern, ignore_case = TRUE)
    )
  ) |>
  select(section_id, evidence_id, surface_flag)

kable(
  section_diagnostics |>
    select(
      section_id,
      distinct_trigram_ratio,
      unsupported_number,
      unsupported_evaluation
    ),
  digits = 2,
  col.names = c(
    "Section ID",
    "Distinct trigram ratio",
    "Unsupported numeral present",
    "Unsupported evaluative wording"
  ),
  caption = "Repetition and numeral diagnostics",
  row.names = FALSE
)

kable(
  coverage_diagnostics,
  col.names = c("Section ID", "Evidence ID", "Surface flag present"),
  caption = "Evidence coverage flags for each generated section",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(
    section_diagnostics$distinct_trigram_ratio >= 0 &
      section_diagnostics$distinct_trigram_ratio <= 1
  ),
  setequal(coverage_diagnostics$evidence_id, paste0("E0", 1:6)),
  is.logical(coverage_diagnostics$surface_flag),
  is.logical(section_diagnostics$unsupported_number),
  is.logical(section_diagnostics$unsupported_evaluation),
  any(!coverage_diagnostics$surface_flag),
  any(section_diagnostics$unsupported_evaluation)
)
```

These are structural hygiene checks. They do not detect an invented plant
example, an unsupported causal explanation, a shifted obligation, or a
contradiction written without a new numeral.
This render still exposes two concrete failures. `Successfully` adds an
evaluation absent from the ledger, and the final section does not state that
volunteers made no guarantee about seed growth. Both send the draft back for
revision.

## Assemble the draft and make a decision

The headings come from the human plan. The prose under each heading comes from
the three model calls above.

```{r}
assembled_document <- section_generations |>
  transmute(
    section_id,
    section = paste0("## ", heading, "\n\n", generated_text)
  ) |>
  pull(section) |>
  paste(collapse = "\n\n")

cat(assembled_document)
```

```{r}
#| include: false

stopifnot(
  str_count(
    assembled_document,
    regex("^## ", multiline = TRUE)
  ) == 3L,
  all(str_detect(assembled_document, document_plan$heading))
)
```

The automatic findings determine whether the draft already needs revision.
They still cannot authorize publication. Until a person checks every generated
claim, continuity across sections, audience, and tone, the explicit publication
decision is **reject**.

```{r}
automatic_revision_needed <- any(
  stopping_diagnostics$hit_token_cap,
  section_diagnostics$unsupported_number,
  section_diagnostics$unsupported_evaluation,
  !coverage_diagnostics$surface_flag
)

document_decision <- tibble(
  automated_recommendation = if_else(
    automatic_revision_needed,
    "revise before human review",
    "send to human review"
  ),
  publication_decision = "reject",
  reason = paste(
    "Human factual, continuity, tone, and audience review is pending;",
    "automatic checks cannot approve the document."
  )
)

kable(
  document_decision,
  col.names = c(
    "Automated recommendation",
    "Publication decision",
    "Reason"
  ),
  caption = "The generated document is not approved for publication",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  document_decision$automated_recommendation %in%
    c("revise before human review", "send to human review"),
  identical(document_decision$publication_decision, "reject")
)
```

A future editor may revise and accept the document. This render records no such
approval.

## What to remember

- Long-text generation starts with document planning and section-scoped
  evidence.
- Decoder-only prompts and outputs share one context window.
- Carry compact, evidence-checked continuity state instead of unbounded raw
  history.
- Check EOS and token-cap status directly; punctuation is only a surface clue.
- Repetition, numeral, and evidence flags are preliminary filters.
- Reject generated prose for publication until a human editor reviews the
  complete document.

## Sources

- [Reiter and Dale on document planning in applied NLG](https://doi.org/10.1017/S1351324997001681)
- [Holtzman et al. on neural text degeneration](https://arxiv.org/abs/1904.09751)
- [Fan et al. on hierarchical long-form generation](https://aclanthology.org/P18-1082/)
- [See et al. on multi-paragraph generation and discourse failures](https://aclanthology.org/K19-1079/)
- [Transformers text-generation controls](https://huggingface.co/docs/transformers/v4.57.1/en/main_classes/text_generation)
- [Qwen2.5-1.5B-Instruct model card](https://huggingface.co/Qwen/Qwen2.5-1.5B-Instruct)
