---
title: "02 - Manual Typewriting"
author: "Alex Farach"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    toc: TRUE
    toc_depth: 3
    toc_float:
      collapsed: FALSE
      smooth_scroll: TRUE
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = FALSE, cache.lazy = FALSE, warning = FALSE,
                      message = FALSE, echo = TRUE, dpi = 180, fig.width = 8,
                      fig.height = 5)
```

# Background

Task 2 in the [Periodic Table of NLP
Tasks](https://www.innerdoc.com/periodic-table-of-nlp-tasks/) has to do with
manual typewriting.

> Typing your own text gives you more confidence when testing your code or demo.
>
> Why not just type your own text and process it in your NLP task. Creating your
> own content is the best test in demo's. Hopefully it increases your confidence
> in the NLP-task that you are testing and lets you insert all the rare cases
> you can think of.

Whenever I'm trying to figure out how to attack an NLP problem this is exactly
this, use some manually created examples to better understand what I'm trying to
test. In the last NLP task, "[01 - Bits to Character
Encoding](https://github.com/farach/periodic_table_nlp/blob/main/01---Bits-to-Character-Encoding.md)"
I selected a handful of words to test out different functions. By manually
creating a small set of examples I was able to better understand what exactly
was going on in each of the functions.

Let's do this again to get learn about something new that is foundational in the
natural language processing field: regular expressions (or regex).

# Regular expressions history

So what is "regex"? Regex is a sequence of characters that specifies a search
pattern in text. The concept of regular expressions began in the 1940's when a
Warren McCulloch, a neuroscience, and Stephen Kleene, a logician, were trying to
understand how the human brain could produce complex patterns using simple cells
bound together.

In the last section of this Periodic Table of NLP Tasks in R series we went over
character encoding and highlighted how simple binary switches could be combined
to encode increasingly complex language symbols. The leap from ASCII to Unicode
was a particularly large jump. By changing the way we encoded language without
changing the fundamental binary switch mechanism we could go from encoding
hundreds of language symbols to millions.

In a 1951 paper Kleene thought through the types of patterns neural networks
were able to detect by applying them to simple toy languages - so-called
"regular languages". For example, given a language whose "grammar" allows only
the letters "A" and "B", is there a neural network that can detect whether an
arbitrary string of letters is valid within the "A/B" grammar or not?

This is why Kleene developed an algebraic notation for encapsulating these
"regular grammars". For example, a\*b\* in the case of the "A/B" language.

Regex is a kind of algebraic shorthand that allows you to define text patterns
which can be used for a variety of natural language processing (NLP) tasks. The
most common of which is likely the searching for or extracting of words or parts
of words from longer string patterns like a document.

Quick note, regular expressions are not a programming language. Yes, it is a
formal language that a computer understands but there are no variables, they
don't set values, they don't contain instructions, and they can't make
decisions.

# Regular expressions in R

Let's jump right in and begin understanding the various patterns this algebraic
shorthand uses to define text patterns.

In base R the [`grep()`
function](https://stat.ethz.ch/R-manual/R-devel/library/base/html/grep.html) is
used to do pattern matching and replacement. In order to understand this
function we need to match to *something*. This is where manual typewriting will
help!

```{r}
grep(pattern = "Alex", x = "Alex like to play guitar", useBytes = TRUE)
```

# References

[Why is this interesting? The Regular Expression
Edition](whyisthisinteresting.substack.com/p/the-regular-expression-edition)
