---
title: "Finding patterns in text"
subtitle: "Learn regular expressions with small, hand-written examples"
author: "Alex Farach"
description: "Find dates, email addresses, and order numbers in text with R."
categories: [source data loading, regular expressions, testing]
---

Near the end of a support shift, a worker needs to connect incoming messages
with the right orders. One message may contain two order numbers, another may
contain an email address, and most contain neither. Reading a small inbox is
easy. Repeating the job across hundreds of messages makes missed and mistaken
matches more likely.

The worker starts with five hand-written messages rather than the full inbox.
Every expected answer fits on the screen, including the examples that should
fail. This small rehearsal set will help us build a text pattern, test its
limits, and hide an address before the messages are shared.

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- search for an exact word or phrase;
- describe a family of values such as order numbers or dates;
- extract and replace matching text; and
- turn hand-written examples into checks that catch mistakes.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

The first code chunk loads the packages used for text patterns and tables before checking the rehearsal messages.

## Make a small rehearsal set

The five messages below are varied enough to pose real questions and short
enough to check by eye. In R, `c()` combines values into a collection called a
**vector**. The arrow `<-` saves that vector under the name `messages`.

```{r}
library(stringr)
library(tibble)
library(dplyr)
library(purrr)

messages <- c(
  "Alex likes to play guitar.",
  "Email support@example.com before 17:30.",
  "Order IDs: R-1042 and R-1043.",
  "The café opens on 2026-08-28.",
  "No identifiers here."
)

stopifnot(
  identical(length(messages), 5L),
  identical(messages[3], "Order IDs: R-1042 and R-1043.")
)

messages
```

An **assertion** is a check that stops R when a result no longer matches what
the lesson says it should be. That is what `stopifnot()` does. Here the checks
confirm the number of messages and the text
of the third one. R then prints all five messages. The number in square
brackets at the left marks the position where that line of output begins.

## Search for exact text

The simplest first search is the narrowest one that solves the problem. Here we
ask for the exact text `"café"` before introducing any flexible pattern.

```{r}
cafe_message <- str_subset(messages, fixed("café"))

stopifnot(
  identical(cafe_message, "The café opens on 2026-08-28.")
)

cafe_message
```

`str_subset()` from stringr keeps the messages that match a pattern. `messages`
says where to look, and `fixed("café")` says to treat `"café"` as ordinary text
rather than as a regular expression.

`fixed()` is a good choice when the exact word, phrase, label, or punctuation
is already known. It prevents punctuation from taking on a special
pattern-matching meaning.

### Ask a yes-or-no question

The support worker will eventually need one answer for every message.
`str_detect()` asks whether each message contains a match, returning `TRUE` for
yes and `FALSE` for no. We will display the answer as a tibble, a table that
prints its size and each column's type.

```{r}
has_order_prefix <- str_detect(messages, fixed("R-"))

stopifnot(
  identical(has_order_prefix, c(FALSE, FALSE, TRUE, FALSE, FALSE))
)

knitr::kable(
  tibble(
    message = messages,
    contains_R_dash = has_order_prefix
  ),
  col.names = c("Message", "Contains R-"),
  caption = "Which messages contain R-?",
  row.names = FALSE
)
```

Only the third message contains `"R-"`. We have found the likely message, but
the rule is still too loose. It would also accept `"R-"` with no digits after
it.

## Describe a family of matches

The two visible order numbers share a shape that we can state one piece at a
time:

- the letter `R`;
- a hyphen; and
- exactly four digits.

A **regular expression**, often shortened to **regex**, is a compact
description of a text pattern. The next expression asks for the whole order ID
rather than its first two characters.

```{r}
order_id_pattern <- r"(\bR-[0-9]{4}\b)"

order_ids_by_message <- str_extract_all(
  messages,
  order_id_pattern
)

stopifnot(
  identical(order_ids_by_message[[3]], c("R-1042", "R-1043")),
  identical(sum(lengths(order_ids_by_message)), 2L)
)

order_ids_by_message
```

The output stays grouped by message. An empty entry means that no order number
was found there. The third entry contains exactly the two IDs the worker could
see in the rehearsal set.

This pattern can be read from left to right:

| Part | Meaning |
|---|---|
| `\b` | Start or end at a word boundary. |
| `R-` | Match these two characters exactly. |
| `[0-9]` | Match one ASCII digit from 0 through 9. |
| `{4}` | Require the previous item four times. |
| `\b` | Stop at another word boundary. |

: Parts of the order ID pattern

R's raw string notation, written here as `r"(...)"`, keeps the backslashes
readable. Without it, the same pattern would be written as
`"\\bR-[0-9]{4}\\b"`.

The order system defines its IDs with ASCII digits. In the ICU engine used by
stringr, `\d` also matches decimal digits from other writing systems. Using
`[0-9]` makes the rule explicit.

## Find several kinds of information

Order IDs are only one kind of information mixed into the messages. The same
rehearsal set also contains a date and an email address, so each gets a
separate pattern.

```{r}
all_text <- str_c(messages, collapse = " ")

date_pattern <- r"(\b[0-9]{4}-[0-9]{2}-[0-9]{2}\b)"
email_pattern <- r"([[:alnum:]._%+-]+@[[:alnum:].-]+\.[[:alpha:]]{2,})"

found <- list(
  order_ids = str_extract_all(
    all_text,
    order_id_pattern
  )[[1]],
  date = str_extract(all_text, date_pattern),
  email = str_extract(all_text, email_pattern)
)

stopifnot(
  identical(found$order_ids, c("R-1042", "R-1043")),
  identical(found$date, "2026-08-28"),
  identical(found$email, "support@example.com")
)

found
```

`str_extract()` returns the first match, while `str_extract_all()` returns
every match. All three expected pieces are present.

The email pattern is intentionally modest. It recognizes the address in this
example, but it does not validate every email address allowed by internet
standards. Finding text that resembles an email and proving that an address is
valid are different tasks.

## Use the start and end of a string

A loose match can occur in the wrong place. Some labels matter only at the
start of a text value, and some values must occupy the whole string. In these
single-line examples, `^` means "the start of the string" and `$` means "the
end of the string."

```{r}
log_lines <- c(
  "ERROR: file not found",
  "INFO: job started",
  "ERRORISH: not an error record"
)

error_lines <- str_subset(log_lines, "^ERROR:")

stopifnot(
  identical(error_lines, "ERROR: file not found")
)

error_lines
```

The pattern requires the complete label `ERROR:` at the start. It leaves out
`ERRORISH`, whose opening text and punctuation do not meet that rule.

Using both anchors checks a complete value from beginning to end. The next
pattern also uses parentheses to capture the year, month, and day separately.

```{r}
complete_date_pattern <- r"(^([0-9]{4})-([0-9]{2})-([0-9]{2})$)"
date_inputs <- c(
  "2026-08-28",
  "2026-8-28",
  "due 2026-08-28"
)

date_parts <- str_match(
  date_inputs,
  complete_date_pattern
)

stopifnot(
  identical(
    unname(date_parts[1, 2:4]),
    c("2026", "08", "28")
  ),
  all(is.na(date_parts[2:3, 1]))
)

knitr::kable(
  tibble(
    input = date_inputs,
    complete_match = date_parts[, 1],
    year = date_parts[, 2],
    month = date_parts[, 3],
    day = date_parts[, 4]
  ),
  col.names = c("Input", "Complete match", "Year", "Month", "Day"),
  caption = "Checking complete dates",
  row.names = FALSE
)
```

`NA` means "missing" in R. The last two inputs do not have the required
four-digit year, two-digit month, and two-digit day as a complete value.

This pattern checks the arrangement of digits. It does not check the calendar.
For example, `2026-99-99` has the expected shape but is not a real date.

## Replace matching text

The worker may need to share an example message while keeping the sender's
address private. A pattern can replace matching text before the example is
shared. The following code substitutes `[email]` for the address.

```{r}
redacted_messages <- str_replace_all(
  messages,
  email_pattern,
  "[email]"
)

stopifnot(
  identical(
    redacted_messages[2],
    "Email [email] before 17:30."
  ),
  !any(str_detect(redacted_messages, fixed("@")))
)

redacted_messages
```

The address is gone, while the rest of the sentence stays unchanged. One
successful example is not enough to establish that a redaction method protects
private information. Test varied cases carefully, inspect the result, and
follow the privacy rules that apply to the real records.

## Turn examples into safety checks

Before searching the full inbox, the worker adds close failures beside the
clear matches. The order ID must contain one `R`, one hyphen, and exactly four
digits. `mutate()` from dplyr adds the actual test result as a new column.

```{r}
id_cases <- tibble(
  input = c(
    "R-0001",
    "prefix R-1234 suffix",
    "R-12",
    "RR-1234",
    "R-12345"
  ),
  expected = c(TRUE, TRUE, FALSE, FALSE, FALSE)
)

id_cases <- id_cases |>
  mutate(actual = str_detect(input, order_id_pattern))

stopifnot(
  identical(id_cases$actual, id_cases$expected)
)

knitr::kable(
  id_cases,
  col.names = c("Input", "Expected", "Actual"),
  caption = "Expected and actual order ID results",
  row.names = FALSE
)
```

The `stopifnot()` line compares the actual answers with the expected answers.
If one differs, R stops with an error and the page is not published. A near
miss such as `R-12345` therefore remains visible during testing rather than
becoming a mistaken match later.

## Reuse a tested pattern

The checked extraction will be used more than once, so we place it in a
**function**. A function is a named set of steps that accepts an input and
returns a result.

```{r}
extract_order_ids <- function(text) {
  str_extract_all(text, order_id_pattern) |>
    list_c() |>
    unique()
}

extracted_ids <- extract_order_ids(messages)

stopifnot(
  identical(extracted_ids, c("R-1042", "R-1043"))
)

extracted_ids
```

The function returns the two expected order numbers and lists each one only
once, even if it appears repeatedly in the input. `list_c()` from purrr combines
the message-by-message matches into one character vector.

## State which text the pattern accepts

Support messages may use accented letters or writing systems beyond English.
In ICU, `\p{L}` means a code point classified as a letter and `\p{M}` means a
combining mark. The second category matters because a visible accented letter
can be stored as a base letter followed by a separate mark.

```{r}
words <- c(
  "cafe",
  "café",
  "naïve",
  "李雷",
  "e\u0301",
  "hello-world"
)
letters_and_marks_pattern <- r"(^(?:\p{L}\p{M}*)+$)"

letters_and_marks <- str_detect(
  words,
  letters_and_marks_pattern
)

stopifnot(
  identical(
    letters_and_marks,
    c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE)
  )
)

knitr::kable(
  tibble(
    text = words,
    accepted_by_pattern = letters_and_marks
  ),
  col.names = c("Text", "Accepted by this pattern"),
  caption = "Checking the stated letter-and-mark rule",
  row.names = FALSE
)
```

The first five examples follow the stated rule. The fifth displays as `é` but
uses `e` plus a combining accent. `hello-world` does not match because a hyphen
is punctuation. This check answers a narrow text question; it is not a rule
for deciding whether a person's name or a word is valid.

::: {.callout-note}
## Technical detail: R has more than one pattern engine

The software that interprets a regular expression is called a **pattern
engine**. Different engines do not agree on every feature.

Base R uses POSIX extended regular expressions by default and PCRE2 when
`perl = TRUE`. The stringr package uses the ICU engine through stringi. These
engines share common syntax, but some Unicode rules and advanced features
differ.

The examples in this lesson use stringr for regular expressions. Test a pattern
again if you move it to a different engine. Avoid `useBytes = TRUE` for normal
text because byte-level matching can split one UTF-8 character into several
pieces.
:::

## Carry the tested pattern into real work

The support worker now has a repeatable process rather than a pattern copied
straight into the full inbox:

1. Write a few short examples that should match.
2. Add similar examples that must not match.
3. Use `fixed()` when you only need exact text.
4. Build a regular expression one piece at a time.
5. Inspect the text you extract before replacing anything.
6. Record the expected answers with an assertion.
7. Add a new example whenever the pattern surprises you.

## What to remember

- Exact searches are simpler than regular expressions and should be used when
  they solve the problem.
- A regular expression describes a family of possible matches.
- Small hand-written examples make a pattern easier to understand.
- Expected failures are as useful as expected matches.
- A published example should stop the build when its result changes.

The five-message rehearsal ends with both order IDs found and the address
hidden in the shared copy. Keep the same two small piles beside any new
pattern: examples it must accept and examples it must refuse.

## Sources

- [R pattern matching documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/grep.html)
- [R regular-expression documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/regex.html)
- [stringr regular-expression guide](https://stringr.tidyverse.org/articles/regular-expressions.html)
- [`stringr::str_detect()` documentation](https://stringr.tidyverse.org/reference/str_detect.html)
- [ICU regular-expression guide](https://unicode-org.github.io/icu/userguide/strings/regexp.html)
- [Unicode character properties, UAX #44](https://www.unicode.org/reports/tr44/)
