---
title: "Loading a structured data file"
subtitle: "Open a CSV file and check what R received"
author: "Alex Farach"
description: "Learn how rows, columns, missing values, and text encodings affect a CSV file in R."
categories: [source data loading, CSV, data quality]
---

Before a community team's weekly meeting, a coordinator receives one CSV file
containing feedback from email, web forms, and live chat. The group plans to
review low ratings. First, the coordinator has to find out what the five
records contain and whether their structure survived the move into R.

A file can open without an error while a column is missing, an accent mark is
damaged, or an identification number has changed format. A blank rating can
also be mistaken for a score if its meaning is not checked. We will follow the
coordinator's arrival check from the file's raw lines to the one low rating the
table can support.

::: {.callout-note}
The feedback records in this lesson are fictional and were created for
teaching. They do not describe real people or customers.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain how a structured data file organizes information;
- load a CSV file into R;
- check rows, columns, and data types;
- find missing values; and
- select records that meet a stated condition.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## What makes a file structured?

The first uncertainty is basic: what does each part of the file represent? A
**structured data file** follows a consistent arrangement. In a CSV file:

- each **row** usually represents one record;
- each **column** represents one kind of information; and
- commas separate one column from the next.

CSV stands for **comma-separated values**. The first line often contains the
column names. Text containing a comma is placed inside quotation marks so that
the comma remains part of the text. A quoted field can also contain a line
break, so one physical line does not always equal one record. A quotation mark
inside a quoted field is usually written twice.

Before asking R to make a table, the coordinator reads the file as plain lines.
This exposes what is actually stored and keeps the example small enough to
count by eye.

```{r}
file_path <- "data/customer_feedback.csv"
file_lines <- readLines(
  file_path,
  encoding = "UTF-8",
  warn = TRUE
)

stopifnot(
  identical(length(file_lines), 6L),
  identical(
    file_lines[1],
    "feedback_id,submitted_at,channel,rating,comment"
  ),
  grepl(
    '"Delivery was quick, thank you!"',
    file_lines[2],
    fixed = TRUE
  )
)

head(file_lines, 3)
```

The `stopifnot()` lines are **assertions**, checks that make R stop if an
expected result changes. They verify the line count, header, and quoted first
comment. There are six lines in all: one header and five records. The first
comment contains a comma, and its quotation marks keep the full sentence in one
field rather than splitting it across columns. This line count works because
the teaching file has no quoted line breaks.

## Load the CSV into R

The line count and header are as expected, so the next step is to organize the
records. `read.csv()` turns the file into a **data frame**, R's table-like
structure. We specify UTF-8, a common way to store characters from many
languages, so R reads the accented text using the file's encoding.

```{r}
feedback <- read.csv(
  file = file_path,
  fileEncoding = "UTF-8",
  na.strings = "",
  colClasses = c(
    feedback_id = "character",
    submitted_at = "character",
    channel = "character",
    rating = "integer",
    comment = "character"
  ),
  check.names = FALSE
)

stopifnot(
  identical(nrow(feedback), 5L),
  identical(ncol(feedback), 5L),
  identical(
    names(feedback),
    c("feedback_id", "submitted_at", "channel", "rating", "comment")
  ),
  identical(feedback$feedback_id, as.character(1001:1005)),
  identical(
    feedback$comment[2],
    "The café listing was easy to find."
  )
)

knitr::kable(
  feedback,
  col.names = c("Feedback ID", "Submitted", "Channel", "Rating", "Comment"),
  caption = "Customer feedback loaded from the CSV file",
  row.names = FALSE
)
```

All five records appear in five columns. The comma in the first comment stayed
inside the comment, and the accented letters in `café` display correctly. The
final comment, `"Muy útil y fácil de usar"`, is Spanish for "Very useful and
easy to use."

The teaching file's data dictionary defines a blank rating as missing.
`na.strings = ""` applies that rule while loading. A real project should use
the missing-value rules supplied by its data provider. The literal text
`"NA"`, zero, and a blank field must not be treated as equivalent without such
evidence.

Those checks show that the visible layout survived the trip into R. We still
need to check how R interpreted each column.

### Why choose the column types?

The `colClasses` setting tells R which **data type** to use for each column. A
data type records whether values should behave as text, whole numbers, dates,
or another kind of information.

- identification numbers are text because they label records;
- dates begin as text so we can convert them deliberately;
- ratings are whole numbers; and
- channels and comments are text.

Treating an ID as text protects leading zeros. For example, the label `"0012"`
should remain different from `"12"` when that difference exists in the source.

We can ask R to report the type it used for every column.

```{r}
column_types <- vapply(
  feedback,
  function(column) class(column)[1],
  character(1)
)

stopifnot(
  identical(
    unname(column_types),
    c("character", "character", "character", "integer", "character")
  )
)

knitr::kable(
  data.frame(
    column = names(column_types),
    type = unname(column_types)
  ),
  col.names = c("Column", "Type used by R"),
  caption = "The type assigned to each column",
  row.names = FALSE
)
```

R reports `character` for text and `integer` for whole numbers. Here,
`character` is R's name for text; a value may contain an entire comment.
Choosing these types deliberately prevents R from turning record labels into
quantities.

## Convert the date column

The submission dates have arrived as text. `as.Date()` converts them to R's
date type so the coordinator can sort and compare them as calendar dates.

```{r}
feedback$submitted_at <- as.Date(
  feedback$submitted_at,
  format = "%Y-%m-%d"
)

stopifnot(
  inherits(feedback$submitted_at, "Date"),
  identical(
    range(feedback$submitted_at),
    as.Date(c("2026-08-21", "2026-08-25"))
  )
)

knitr::kable(
  feedback[c("feedback_id", "submitted_at", "channel")],
  col.names = c("Feedback ID", "Submitted", "Channel"),
  caption = "Submission dates after conversion",
  row.names = FALSE
)
```

The format `%Y-%m-%d` means four-digit year, two-digit month, and two-digit day.
Writing it explicitly records the order expected in this file instead of
leaving that order to a guess.

## Find missing information

With the rows, columns, and types accounted for, the coordinator can ask which
values are absent. R uses `NA` to represent a **missing value**. The CSV
contains one blank rating, which `read.csv()` changed to `NA`.

```{r}
missing_by_column <- colSums(is.na(feedback))

stopifnot(
  identical(
    unname(as.integer(missing_by_column)),
    c(0L, 0L, 0L, 1L, 0L)
  ),
  is.na(feedback$rating[3])
)

knitr::kable(
  data.frame(
    column = names(missing_by_column),
    missing_values = as.integer(missing_by_column)
  ),
  col.names = c("Column", "Missing values"),
  caption = "Missing values in each column",
  row.names = FALSE
)
```

The rating column has one missing value. A blank tells us only that no rating
is recorded in this file. It is different from a rating of zero and should not
be replaced without evidence about what the blank means.

## Select a record that needs attention

Only now does the coordinator select feedback with a rating of two or lower.
The expression `!is.na(feedback$rating)` keeps the unknown rating out of the
comparison rather than treating it as high or low.

```{r}
low_ratings <- feedback[
  !is.na(feedback$rating) & feedback$rating <= 2,
  c("feedback_id", "rating", "comment")
]

stopifnot(
  identical(nrow(low_ratings), 1L),
  identical(low_ratings$feedback_id, "1004"),
  identical(low_ratings$rating, 2L)
)

knitr::kable(
  low_ratings,
  col.names = c("Feedback ID", "Rating", "Comment"),
  caption = "Feedback with a rating of two or lower",
  row.names = FALSE
)
```

One record meets the condition: feedback `1004`, which says that the parcel
arrived late. The missing rating remains separate because the file does not
tell us how that person would have scored the experience.

## Questions to ask about any structured file

The five-row arrival check is complete, but technical tidiness cannot answer
every question about the feedback. Before analyzing any structured file, ask:

1. What does one row represent?
2. What does each column mean?
3. Which encoding does the file use?
4. How are missing values recorded?
5. Are dates and identification numbers stored correctly?
6. Who created the file, and what information did they leave out?

A tidy table cannot tell you whether the collection process was fair or
complete. These rows represent only the feedback that reached the recorded
channels and was included in this file. The file's origin, exclusions, and
limits belong alongside the technical checks.

## What to remember

- A structured file uses a consistent arrangement of rows and columns.
- Quotation marks protect text that contains a separator such as a comma.
- Choose important column types instead of relying on guesses.
- Check missing values before filtering or summarizing.
- Opening a file successfully is only the beginning of checking it.

At the meeting, the coordinator can report five records, one missing rating,
and one recorded rating of two. The blank remains a blank, and the file's reach
remains an open question. That is the useful discipline of an arrival check:
say exactly what is recorded, what is blank, and what the file cannot settle.

## Sources

- [R documentation for `read.csv()` and `read.table()`](https://stat.ethz.ch/R-manual/R-devel/library/utils/html/read.table.html)
- [R Data Import/Export manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html)
- [RFC 4180: Common format and MIME type for CSV files](https://www.rfc-editor.org/rfc/rfc4180)
