---
title: "Loading a structured data file"
subtitle: "Open a CSV file and check what R received"
author: "Alex Farach"
description: "Learn how rows, columns, missing values, and text encodings affect a CSV file in R."
categories: [source data loading, CSV, data quality]
---

Before a community team's weekly meeting, a coordinator receives one CSV file
containing feedback from email, web forms, and live chat. The group plans to
review low ratings. First, the coordinator has to find out what the five
records contain and whether their structure survived the move into R.

A file can open without an error while a column is missing, an accent mark is
damaged, or an identification number has changed format. A blank rating can
also be mistaken for a score if its meaning is not checked. We will follow the
coordinator's arrival check from the file's raw lines to the one low rating the
table can support, then repeat the check on a second export that arrives with a
repeated identifier, an impossible rating, and a row that has too many fields.

::: {.callout-note}
The feedback records in this lesson are fictional and were created for
teaching. They do not describe real people or customers.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain how a structured data file organizes information;
- load a CSV file into R;
- check rows, columns, and data types;
- find missing values;
- inspect the parse problems, duplicate keys, and out-of-range values in a
  flawed file; and
- select records that meet a stated condition.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

The first code chunk loads the packages used for reading, checking, and displaying the feedback file.

## What makes a file structured?

The first uncertainty is basic: what does each part of the file represent? A
**structured data file** follows a consistent arrangement. In a CSV file:

- each **row** usually represents one record;
- each **column** represents one kind of information; and
- commas separate one column from the next.

CSV stands for **comma-separated values**. The first line often contains the
column names. Text containing a comma is placed inside quotation marks so that
the comma remains part of the text. A quoted field can also contain a line
break, so one physical line does not always equal one record. A quotation mark
inside a quoted field is usually written twice.

Before asking R to make a table, the coordinator reads the file as plain lines.
`read_lines()` from readr returns one text value for each physical line. This
exposes what is actually stored and keeps the example small enough to count by
eye.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)

file_path <- "data/customer_feedback.csv"
file_lines <- read_lines(file_path)

stopifnot(
  identical(length(file_lines), 6L),
  identical(
    file_lines[1],
    "feedback_id,submitted_at,channel,rating,comment"
  ),
  str_detect(file_lines[2], fixed('"Delivery was quick, thank you!"'))
)

head(file_lines, 3)
```

The `stopifnot()` lines are **assertions**, checks that make R stop if an
expected result changes. They verify the line count, header, and quoted first
comment. There are six lines in all: one header and five records. The first
comment contains a comma, and its quotation marks keep the full sentence in one
field rather than splitting it across columns. This line count works because
the teaching file has no quoted line breaks.

## Load the CSV into R

The line count and header are as expected, so the next step is to organize the
records. `read_csv()` from readr turns the file into a **tibble**, a table that
prints its size and each column's type. readr treats text as UTF-8 by default,
a common way to store characters from many languages, so the accented text is
read using that expected encoding.

```{r}
feedback <- read_csv(
  file = file_path,
  na = "",
  col_types = cols(
    feedback_id = col_character(),
    submitted_at = col_character(),
    channel = col_character(),
    rating = col_integer(),
    comment = col_character()
  )
)

stopifnot(
  inherits(feedback, "tbl_df"),
  identical(nrow(feedback), 5L),
  identical(ncol(feedback), 5L),
  identical(
    names(feedback),
    c("feedback_id", "submitted_at", "channel", "rating", "comment")
  ),
  identical(feedback$feedback_id, as.character(1001:1005)),
  identical(
    feedback$comment[2],
    "The café listing was easy to find."
  )
)

knitr::kable(
  feedback,
  col.names = c("Feedback ID", "Submitted", "Channel", "Rating", "Comment"),
  caption = "Customer feedback loaded from the CSV file",
  row.names = FALSE
)
```

All five records appear in five columns. The comma in the first comment stayed
inside the comment, and the accented letters in `café` display correctly. The
final comment, `"Muy útil y fácil de usar"`, is Spanish for "Very useful and
easy to use."

The teaching file's data dictionary defines a blank rating as missing.
`na = ""` applies that rule while loading. A real project should use the
missing-value rules supplied by its data provider. The literal text `"NA"`,
zero, and a blank field must not be treated as equivalent without such evidence.

Those checks show that the visible layout survived the trip into R. We still
need to check how R interpreted each column.

### Why choose the column types?

The `col_types = cols(...)` setting tells readr which **data type** to use for
each column. A data type records whether values should behave as text, whole
numbers, dates, or another kind of information.

- identification numbers are text because they label records;
- dates begin as text so we can convert them deliberately;
- ratings are whole numbers; and
- channels and comments are text.

Treating an ID as text protects leading zeros. For example, the label `"0012"`
should remain different from `"12"` when that difference exists in the source.

We can ask R to report the type it used for every column. `map_chr()` from
purrr applies the same check to each column and returns a character vector.
The shorthand `\(column)` means "for each column, run the small calculation that follows."

```{r}
column_types <- map_chr(
  feedback,
  \(column) class(column)[1]
)

stopifnot(
  identical(
    unname(column_types),
    c("character", "character", "character", "integer", "character")
  )
)

knitr::kable(
  tibble(
    column = names(column_types),
    type = unname(column_types)
  ),
  col.names = c("Column", "Type used by R"),
  caption = "The type assigned to each column",
  row.names = FALSE
)
```

R reports `character` for text and `integer` for whole numbers. Here,
`character` is R's name for text; a value may contain an entire comment.
Choosing these types deliberately prevents R from turning record labels into
quantities.

## Convert the date column

The submission dates have arrived as text. `mutate()` from dplyr changes a
column while keeping the rest of the table. Here `as.Date()` converts the text
to R's date type so the coordinator can sort and compare the values as calendar
dates.

```{r}
feedback <- feedback |>
  mutate(
    submitted_at = as.Date(
      submitted_at,
      format = "%Y-%m-%d"
    )
  )

stopifnot(
  inherits(feedback$submitted_at, "Date"),
  identical(
    range(feedback$submitted_at),
    as.Date(c("2026-08-21", "2026-08-25"))
  )
)

knitr::kable(
  feedback |>
    select(feedback_id, submitted_at, channel),
  col.names = c("Feedback ID", "Submitted", "Channel"),
  caption = "Submission dates after conversion",
  row.names = FALSE
)
```

The format `%Y-%m-%d` means four-digit year, two-digit month, and two-digit day.
Writing it explicitly records the order expected in this file rather than
leaving that order to a guess.

## Find missing information

With the rows, columns, and types accounted for, the coordinator can ask which
values are absent. R uses `NA` to represent a **missing value**. The CSV
contains one blank rating, which `read_csv()` changed to `NA`. `map_int()`
returns one whole-number count for each column.

```{r}
missing_counts <- map_int(
  feedback,
  \(column) sum(is.na(column))
)

stopifnot(
  identical(unname(missing_counts), c(0L, 0L, 0L, 1L, 0L)),
  is.na(feedback$rating[3])
)

knitr::kable(
  tibble(
    column = names(missing_counts),
    missing_values = unname(missing_counts)
  ),
  col.names = c("Column", "Missing values"),
  caption = "Missing values in each column",
  row.names = FALSE
)
```

The rating column has one missing value. A blank tells us only that no rating
is recorded in this file. It is different from a rating of zero and should not
be replaced without evidence about what the blank means.

## Select a record that needs attention

Only now does the coordinator select feedback with a rating of two or lower.
`filter()` from dplyr keeps rows that meet a condition, and `select()` keeps the
columns needed for the meeting. The expression `!is.na(rating)` keeps the
unknown rating out of the comparison rather than treating it as high or low.

```{r}
low_ratings <- feedback |>
  filter(!is.na(rating), rating <= 2) |>
  select(feedback_id, rating, comment)

stopifnot(
  identical(nrow(low_ratings), 1L),
  identical(low_ratings$feedback_id, "1004"),
  identical(low_ratings$rating, 2L)
)

knitr::kable(
  low_ratings,
  col.names = c("Feedback ID", "Rating", "Comment"),
  caption = "Feedback with a rating of two or lower",
  row.names = FALSE
)
```

One record meets the condition: feedback `1004`, which says that the parcel
arrived late. The missing rating remains separate because the file does not
tell us how that person would have scored the experience.

## When the next file is not clean

Everything above worked because the file was well formed. That is the
uninteresting case. The next export from the same web form is written out
below, inside the lesson rather than saved to disk, so the damage is visible
line by line.

It contains six records and six problems: a repeated `feedback_id`, a rating
typed as a word, a row with an extra field, a rating outside the documented
range of one to five, a blank rating, and a comment that runs across two
physical lines inside its quotation marks.

```{r}
messy_text <- paste(
  "feedback_id,submitted_at,channel,rating,comment",
  "1006,2026-08-26,email,4,\"Arrived on time\"",
  "1006,2026-08-26,email,4,\"Second row with the same ID\"",
  "1007,2026-08-27,chat,five,\"Rating typed as a word\"",
  "1008,2026-08-27,web,3,\"Extra field\",unexpected",
  "1009,2026-08-28,web,9,\"Rating outside the documented range\"",
  "1010,2026-08-28,email,,\"Comment that runs\non to a second line\"",
  sep = "\n"
)

messy_lines <- read_lines(I(messy_text))
messy_feedback <- suppressWarnings(
  read_csv(
    I(messy_text),
    na = "",
    col_types = cols(
      feedback_id = col_character(),
      submitted_at = col_character(),
      channel = col_character(),
      rating = col_integer(),
      comment = col_character()
    )
  )
)

stopifnot(
  identical(length(messy_lines), 8L),
  identical(nrow(messy_feedback), 6L),
  identical(
    names(messy_feedback),
    names(feedback)
  )
)

tibble(
  measure = c("Physical lines", "Records read"),
  count = c(length(messy_lines), nrow(messy_feedback))
)
```

Eight lines produced six records. The quoted line break is the reason, and it is
why the arrival check earlier in this lesson could only compare a line count
with a record count on a file known to have none.

`suppressWarnings()` appears because this page turns warnings into errors.
Suppressing the warning does not discard the evidence: readr keeps a structured
report of everything it could not parse.

### Read the parse problems

```{r}
parse_problems <- problems(messy_feedback) |>
  select(row, col, expected, actual)

stopifnot(
  inherits(parse_problems, "tbl_df"),
  identical(nrow(parse_problems), 2L),
  identical(
    parse_problems$expected,
    c("an integer", "5 columns")
  ),
  identical(
    parse_problems$actual,
    c("five", "6 columns")
  )
)

knitr::kable(
  parse_problems,
  col.names = c("Row", "Column", "Expected", "Found"),
  caption = "What readr could not parse",
  row.names = FALSE
)
```

readr counts the header line, so the reported row 4 is the third record, the one
whose rating was typed as `five`. That value became `NA`. The row with an extra
field kept its data by folding the stray text into the comment, which is worth
knowing: a parse problem does not always produce a missing value, and a missing
value does not always come from a parse problem.

The transferable idea is not the function name. Declaring column types and
missing-value rules at read time, and keeping the parser's complaint, is the
same discipline that `dtype` and `na_values` serve in pandas or an explicit
schema serves in Arrow. A parser that guesses quietly is the hazard in every
language.

### Check the things a parser cannot know

Parsing succeeds when the shape is right. It says nothing about whether the
identification numbers are unique or the ratings are possible. Those rules come
from the data dictionary, and each one can be written as a check that fails
loudly.

```{r}
documented_rating_range <- 1:5

row_report <- messy_feedback |>
  mutate(
    record_number = row_number(),
    duplicate_id = feedback_id %in%
      feedback_id[duplicated(feedback_id)],
    rating_missing = is.na(rating),
    rating_out_of_range = !rating_missing &
      !(rating %in% documented_rating_range),
    extra_field_text = str_detect(comment, fixed("unexpected")),
    status = case_when(
      duplicate_id ~ "quarantine: repeated feedback_id",
      rating_out_of_range ~ "quarantine: rating outside 1 to 5",
      extra_field_text ~ "quarantine: extra field folded into comment",
      rating_missing ~ "keep: rating recorded as missing",
      TRUE ~ "keep"
    )
  )

status_counts <- row_report |>
  count(status, name = "records")

quarantined <- row_report |>
  filter(str_starts(status, "quarantine"))

stopifnot(
  identical(sum(status_counts$records), 6L),
  identical(nrow(quarantined), 4L),
  identical(
    quarantined$feedback_id,
    c("1006", "1006", "1008", "1009")
  ),
  identical(
    sum(row_report$rating_missing),
    2L
  ),
  identical(
    n_distinct(messy_feedback$feedback_id),
    5L
  )
)

knitr::kable(
  row_report |>
    select(record_number, feedback_id, rating, status),
  col.names = c("Record", "Feedback ID", "Rating", "Decision"),
  caption = "Every record accounted for, including the ones held back",
  row.names = FALSE
)
```

Four of six records are held back and the reasons differ. Two share an ID, so
neither can be trusted as the record for `1006` until the source explains the
repeat. One folded an extra field into its comment. One carries a rating of
nine, which the dictionary does not allow.

The two remaining records are kept, and one of them has no rating at all. A
blank that the dictionary defines as missing is not an error.

The important discipline is the accounting: six records in, six records
classified, nothing dropped in silence. A pipeline that filtered to the clean
rows and moved on would have reported two tidy records and lost the fact that a
duplicate ID exists in the source.

## Questions to ask about any structured file

The five-row arrival check is complete, but technical tidiness cannot answer
every question about the feedback. Before analyzing any structured file, ask:

1. What does one row represent?
2. What does each column mean?
3. Which encoding does the file use?
4. How are missing values recorded?
5. Are dates and identification numbers stored correctly?
6. Who created the file, and what information did they leave out?

A tidy table cannot tell you whether the collection process was fair or
complete. These rows represent only the feedback that reached the recorded
channels and was included in this file. The file's origin, exclusions, and
limits belong alongside the technical checks.

## What to remember

- A structured file uses a consistent arrangement of rows and columns.
- Quotation marks protect text that contains a separator such as a comma.
- Quoted line breaks make physical lines and records different counts.
- Choose important column types rather than relying on guesses.
- Read the parser's problem report instead of only its output table.
- Uniqueness and value ranges come from the data dictionary, not the parser.
- Check missing values before filtering or summarizing.
- Opening a file successfully is only the beginning of checking it.

At the meeting, the coordinator can report five records, one missing rating,
and one recorded rating of two. The second file is a different report: six
records, two usable, four held back with a stated reason each. The blank
remains a blank, and the file's reach remains an open question. That is the
useful discipline of an arrival check: say exactly what is recorded, what is
blank, and what the file cannot settle.

## Sources

- [readr documentation for `read_csv()`](https://readr.tidyverse.org/reference/read_delim.html)
- [readr documentation for `problems()`](https://readr.tidyverse.org/reference/problems.html)
- [Tibble documentation](https://tibble.tidyverse.org/)
- [R Data Import/Export manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html)
- [RFC 4180: Common format and MIME type for CSV files](https://www.rfc-editor.org/rfc/rfc4180)
