---
title: "Building a text corpus"
subtitle: "Turn separate documents into an organized collection"
author: "Alex Farach"
description: "Learn how to collect text, attach useful labels, and check a corpus before analysis."
categories: [source data loading, corpus, data quality]
---

A person locked out of an account needs the right help article. The
help-center editor has three separate files about deliveries, returns, and
account access. The password-reset instructions are somewhere in those files,
but each search result also needs to retain its topic and source. That becomes
harder as loose documents accumulate.

The editor needs an organized collection called a **corpus**. Its plural is
**corpora**. A corpus keeps text documents together and usually stores
information describing each one. We will build a three-document corpus, check
for empty or duplicate records, and use it to find the password-reset article.

::: {.callout-note}
The help articles in this lesson were written for this example. They do not
come from a real support system.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain what a corpus is;
- read several text files into R;
- decide what counts as one document;
- state the sampling frame and what was excluded;
- keep source information with each document;
- find near-duplicate documents and keep their families together; and
- check a corpus for common problems.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

The first code chunk loads the packages needed to read files, build tibbles, repeat work, and handle text. A tibble is a table that prints its size and each column's type.

## Decide what one document means

Before collecting anything, the editor must choose the **document unit**, the
amount of text that will count as one document. Depending on the question, it
might be:

- one help article;
- one customer message;
- one news story;
- one interview; or
- one paragraph within a longer report.

The right unit depends on the question. Because the editor wants to compare and
search help articles, each complete article will be one document. Treating each
sentence as a separate document would detach the sentence from the rest of its
article.

Before loading real documents, confirm the right to use them. Record the
license or permission, collection method, expected audience, privacy risk, and
whether the people represented could reasonably expect this use. The files
below are synthetic, so they contain no personal records or third-party text.

The example uses three **plain-text files**, files that store characters
without page layout or styling. `list.files()` asks R to find file names ending
in `.txt`.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(digest)

article_paths <- sort(
  list.files(
    path = "data/help_articles",
    pattern = "[.]txt$",
    full.names = TRUE
  )
)

expected_files <- c(
  "account-access.txt",
  "delivery.txt",
  "returns.txt"
)

stopifnot(
  identical(basename(article_paths), expected_files)
)

basename(article_paths)
```

The `stopifnot()` line is an **assertion**, a check that makes R stop if an
expected result changes. It confirms that R found all three file names in
alphabetical order. We keep those names as the first labels in the collection
so each result can still be traced to the file that supplied its text.

## Read every document

Finding the file names does not yet load their contents. `read_lines()` reads
the text inside one file. These files use UTF-8, a common way to store
characters from many languages. A **function** is a named set of steps. The
small function below preserves line breaks in the source copy.

```{r}
expected_line_counts <- c(
  "account-access.txt" = 1L,
  "delivery.txt" = 1L,
  "returns.txt" = 1L
)

read_document <- function(path) {
  lines <- read_lines(path)
  file_name <- basename(path)
  stopifnot(identical(length(lines), unname(expected_line_counts[file_name])))

  str_c(lines, collapse = "\n")
}

source_documents <- article_paths |>
  set_names(basename(article_paths)) |>
  map_chr(read_document)

analysis_documents <- str_replace_all(
  source_documents,
  "[\r\n]+",
  " "
)

stopifnot(
  identical(length(source_documents), 3L),
  identical(names(source_documents), expected_files),
  identical(unname(expected_line_counts), c(1L, 1L, 1L)),
  all(str_length(source_documents) > 0),
  identical(
    names(analysis_documents),
    expected_files
  ),
  str_detect(
    analysis_documents[1],
    fixed("password reset link")
  )
)

source_documents
```

`map_chr()` runs `read_document()` once for each path and returns a character
vector. The result contains three source copies, and the file names remain
attached. A separate `analysis_documents` object replaces line breaks with
spaces. The source copy
is left untouched so paragraph boundaries can be recovered later.

The check using `str_length()` confirms that none of the documents is
empty. An empty file might mean that loading failed or that the source itself
was blank. Either case should be investigated before analysis.

## Put text and labels together

The text is present, but the values alone have little room for context. Each
document also needs **metadata**, information about the document rather than
its main text. A source file, topic, publication date, or author can be
metadata.

We will place the text and its labels in a tibble, with one row per document.

```{r}
hash_document <- function(path) {
  lines <- read_lines(path)
  file_name <- basename(path)
  stopifnot(identical(length(lines), unname(expected_line_counts[file_name])))

  digest::digest(
    paste(lines, collapse = "\n"),
    algo = "md5",
    serialize = FALSE
  )
}

corpus <- tibble(
  document_id = str_remove(
    basename(article_paths),
    "[.]txt$"
  ),
  topic = c("account access", "delivery", "returns"),
  source_file = basename(article_paths),
  source_hash = map_chr(
    article_paths,
    hash_document
  ),
  source_text = unname(source_documents),
  analysis_text = unname(analysis_documents)
)

stopifnot(
  inherits(corpus, "tbl_df"),
  identical(nrow(corpus), 3L),
  identical(
    corpus$document_id,
    c("account-access", "delivery", "returns")
  ),
  identical(
    corpus$source_file,
    expected_files
  ),
  all(str_detect(corpus$source_hash, "^[0-9a-f]{32}$"))
)

knitr::kable(
  corpus |>
    select(document_id, topic, source_file),
  col.names = c("Document ID", "Topic", "Source file"),
  caption = "The documents and their metadata",
  row.names = FALSE
)
```

The `document_id` gives every row a stable label. The topic identifies the
article's subject without requiring someone to open the full text. The source
file records where that text came from. `source_hash` is a fingerprint that
changes if a file changes; it helps detect accidental edits but is not proof of
authorship or trustworthiness. Keeping these fields on the same row is the
corpus equivalent of keeping a label attached to its folder.

## Say what the collection is a sample of

A corpus is always a sample of something, even when nobody chose a sample. The
honest version of that sentence names the frame: the population the collection
was drawn from, the rule used to draw it, and what the rule left out.

For these three articles the frame is small and the answer is unusual. The
collection is not a sample at all. It is every article this fictional help
center had published as of the stated date, which makes it a census of a tiny
frame rather than a sample of the support questions people actually ask.

That distinction changes what the corpus can support later. A count over a
census of published articles describes what the help center wrote. It does not
describe what readers needed and did not find.

```{r}
collection_record <- tibble(
  field = c(
    "frame",
    "selection rule",
    "collected on",
    "excluded by rule",
    "known coverage gap",
    "language",
    "unit"
  ),
  value = c(
    "articles published by one fictional help center",
    "all published articles, no sampling",
    "2026-08-28",
    "drafts, archived articles, and reader messages",
    "questions people asked that no article answers",
    "English (en)",
    "one complete article"
  )
)

stopifnot(
  identical(nrow(collection_record), 7L),
  all(str_length(collection_record$value) > 0),
  identical(
    collection_record$value[
      collection_record$field == "selection rule"
    ],
    "all published articles, no sampling"
  )
)

knitr::kable(
  collection_record,
  col.names = c("Field", "Recorded value"),
  caption = "What the collection covers and what it leaves out",
  row.names = FALSE
)
```

Write this record before the analysis, not after it. Once results exist, the
frame tends to be described in whatever terms make the results sound broader.

## Add a simple measurement

A quick count can reveal a document that loaded as empty or unexpectedly long.
The function below separates text wherever it finds whitespace, then counts the
resulting pieces. Here, whitespace means spaces, tabs, line breaks, and
non-breaking spaces matched by `[[:space:]]+`.

```{r}
count_whitespace_tokens <- function(text) {
  tokens <- str_split_1(str_trim(text), "[[:space:]]+")

  sum(str_length(tokens) > 0)
}

corpus <- corpus |>
  mutate(
    whitespace_token_count = map_int(
      analysis_text,
      count_whitespace_tokens
    )
  )

stopifnot(
  identical(count_whitespace_tokens(""), 0L),
  identical(count_whitespace_tokens("   "), 0L),
  identical(count_whitespace_tokens("one\u00A0two three"), 3L),
  identical(
    corpus$whitespace_token_count,
    c(18L, 16L, 21L)
  )
)

knitr::kable(
  corpus |>
    select(document_id, topic, whitespace_token_count),
  col.names = c("Document ID", "Topic", "Whitespace-token count"),
  caption = "Whitespace-token counts for the three documents",
  row.names = FALSE
)
```

The three articles contain 18, 16, and 21 whitespace-separated pieces. These
are often called tokens, not universal word counts. Languages without spaces
between words, contractions, punctuation, and specialized tokenizers can
produce different results. The counts are quick checks, not quality scores.

## Check the corpus before using it

The editor now has rows, labels, and text. Before trusting the collection, we
confirm that:

- every source file still exists;
- every document has text;
- document IDs are unique;
- no two documents contain exactly the same text; and
- the topic and source file are present.

```{r}
quality_checks <- tibble(
  check = c(
    "source_files_exist",
    "no_empty_text",
    "unique_document_ids",
    "no_exact_duplicate_text",
    "metadata_is_complete"
  ),
  passed = c(
    all(file.exists(article_paths)),
    all(str_length(str_trim(corpus$source_text)) > 0),
    n_distinct(corpus$document_id) == nrow(corpus),
    n_distinct(corpus$source_text) == nrow(corpus),
    !any(
      is.na(
        corpus |>
          select(topic, source_file, source_hash)
      )
    )
  )
)

stopifnot(
  inherits(quality_checks, "tbl_df"),
  identical(nrow(quality_checks), 5L),
  all(quality_checks$passed)
)

quality_checks
```

Every `passed` value is `TRUE`, so the corpus passes these structural checks.
They do not establish that the articles are accurate, useful, or
representative. They show only that the three expected files produced three
labeled, nonempty, distinct records.

## Use the corpus as a collection

The collection is now ready for the editor's original question. Because every
document shares one structure, the same search can be applied to all three.
This one looks for the exact phrase `"password reset"`.

```{r}
matching_documents <- corpus |>
  filter(
    str_detect(
      analysis_text,
      fixed("password reset")
    )
  ) |>
  select(document_id, topic, analysis_text)

stopifnot(
  inherits(matching_documents, "tbl_df"),
  identical(nrow(matching_documents), 1L),
  identical(
    matching_documents$document_id,
    "account-access"
  )
)

knitr::kable(
  matching_documents,
  col.names = c("Document ID", "Topic", "Text"),
  caption = "The document that mentions a password reset",
  row.names = FALSE
)
```

The account-access article is the single match. Its topic, document ID, and
text arrive together, so the editor can inspect the result and return to the
source file. That connection becomes easier to lose as a corpus grows.

::: {.callout-note}
## Technical detail: a corpus can have a special R class

An R **class** tells R what kind of object it is handling and which operations
belong with it.

Packages such as quanteda and tm provide dedicated corpus objects with tools
for metadata and text processing. A tibble is enough for this lesson because
its rows and columns are visible and easy to inspect. The same records can be
converted to a package-specific corpus later.
:::

## A near duplicate is not an exact duplicate

A fourth file arrives. The support team copied the account-access article into
a seasonal FAQ, changed the capitalization, replaced a hyphen, and added a
courtesy line. To a reader it is the same article. To an exact comparison it is
a new document.

The candidate below is written into the lesson so its wording is visible.

```{r}
candidate_text <- paste(
  "Use the password reset link on the sign in page.",
  "The link expires after 30 minutes for your security.",
  "Contact the help desk if the message does not arrive."
)

incoming <- tibble(
  document_id = c(corpus$document_id, "seasonal-faq-account"),
  analysis_text = c(corpus$analysis_text, candidate_text)
)

exact_duplicate_found <- n_distinct(incoming$analysis_text) < nrow(incoming)

normalize_for_matching <- function(text) {
  text |>
    str_to_lower() |>
    str_replace_all("[^a-z0-9]+", " ") |>
    str_squish()
}

normalized <- normalize_for_matching(incoming$analysis_text)
similarity <- function(left, right) {
  distance <- as.integer(adist(left, right))
  1 - distance / max(str_length(left), str_length(right))
}
containment <- function(left, right) {
  left_words <- unique(str_split_1(left, " "))
  right_words <- unique(str_split_1(right, " "))
  shorter <- if (length(left_words) <= length(right_words)) {
    left_words
  } else {
    right_words
  }
  longer <- if (length(left_words) <= length(right_words)) {
    right_words
  } else {
    left_words
  }

  mean(shorter %in% longer)
}

pair_scores <- tibble(
  left = incoming$document_id[
    c(1L, 1L, 1L, 2L, 2L, 3L)
  ],
  right = incoming$document_id[
    c(2L, 3L, 4L, 3L, 4L, 4L)
  ],
  similarity = map2_dbl(
    normalized[c(1L, 1L, 1L, 2L, 2L, 3L)],
    normalized[c(2L, 3L, 4L, 3L, 4L, 4L)],
    similarity
  ),
  containment = map2_dbl(
    normalized[c(1L, 1L, 1L, 2L, 2L, 3L)],
    normalized[c(2L, 3L, 4L, 3L, 4L, 4L)],
    containment
  )
)

stopifnot(
  identical(exact_duplicate_found, FALSE),
  identical(nrow(pair_scores), 6L),
  all(pair_scores$similarity >= 0),
  all(pair_scores$similarity <= 1),
  all(pair_scores$containment >= 0),
  all(pair_scores$containment <= 1),
  identical(
    pair_scores$right[which.max(pair_scores$similarity)],
    "seasonal-faq-account"
  ),
  identical(
    pair_scores$left[which.max(pair_scores$similarity)],
    "account-access"
  ),
  sum(pair_scores$similarity > 0.6) == 1L,
  sum(pair_scores$containment > 0.9) == 1L
)

knitr::kable(
  pair_scores |>
    mutate(
      similarity = round(similarity, 3),
      containment = round(containment, 3)
    ) |>
    arrange(desc(similarity)),
  col.names = c(
    "Document",
    "Compared with",
    "Character similarity",
    "Word containment"
  ),
  caption = "All six document pairs, most similar first",
  row.names = FALSE
)
```

The exact-duplicate check returns `FALSE`: no two documents are byte-identical.
The two comparisons that follow both single out the same pair, and they
disagree about how strong the resemblance is.

Character similarity reaches 0.65 for the reprint. The added courtesy line
holds it down, because every extra character counts as a difference. Word
containment asks a narrower question, whether one document's words all appear in
the other, and returns 1.00. When a copy is extended rather than edited,
containment is the measure that keeps working.

The thresholds used here, 0.6 and 0.9, are decisions rather than discoveries.
Any threshold belongs in the record with the measure that produced it, because
the same pair can be a near duplicate under one measure and a distinct document
under another. Both comparisons are quadratic in the number of documents, which
is invisible at four and prohibitive at four million.

### Keep families together, then report what left

A near duplicate creates a second problem. If one copy trains a model and the
other tests it, the test measures memorization. The fix is to group related
documents into a **family** and let a split move the whole family at once.

```{r}
families <- incoming |>
  mutate(
    family_id = if_else(
      document_id %in% c("account-access", "seasonal-faq-account"),
      "account-access-family",
      document_id
    ),
    kept = document_id != "seasonal-faq-account"
  )

family_split <- families |>
  distinct(family_id) |>
  mutate(
    split = if_else(
      family_id == "returns",
      "evaluation",
      "development"
    )
  )

assigned <- families |>
  left_join(family_split, by = join_by(family_id))
removed <- assigned |>
  filter(!kept)

split_is_family_consistent <- assigned |>
  summarise(splits = n_distinct(split), .by = family_id) |>
  pull(splits) |>
  max()

stopifnot(
  identical(nrow(assigned), 4L),
  identical(sum(assigned$kept), 3L),
  identical(nrow(removed), 1L),
  identical(split_is_family_consistent, 1L),
  identical(
    n_distinct(assigned$family_id),
    3L
  ),
  identical(
    removed$family_id,
    "account-access-family"
  )
)

knitr::kable(
  assigned |>
    select(document_id, family_id, split, kept),
  col.names = c("Document", "Family", "Split", "Kept"),
  caption = "Four candidates, three kept, one removed with its family recorded",
  row.names = FALSE
)

removed$analysis_text
```

Four candidates went in, three came out, and the removed text is printed rather
than summarized. Reading what a cleaning step deleted is the only way to notice
that it deleted the wrong thing. Here the discarded copy carries one sentence
the original does not, about contacting the help desk, so a careful editor might
merge that line instead of dropping the file.

`split_is_family_consistent` is the falsifiable part. It counts the distinct
splits any family received, and the corpus is only safe to evaluate on when that
number is one.

## A corpus is a set of choices

The successful search does not make the collection neutral. Someone decided
which files belong, where a document ends, which metadata to keep, and which
text to exclude. Every later result inherits those choices.

Record:

1. where the documents came from;
2. when they were collected;
3. what one row represents;
4. which frame the collection covers and which rule selected from it;
5. why any documents were excluded;
6. what each metadata field means;
7. what license, permission, or consent supports the use;
8. which privacy risks were checked;
9. whether people or communities are missing from the collection;
10. which near-duplicate rule and threshold were applied; and
11. which fingerprint identifies each untouched source file.

## What to remember

- A corpus is an organized collection of text documents.
- The document unit should match the question being asked.
- Name the frame the collection came from, even when it is a census.
- Metadata keeps each document connected to its source and context.
- Preserve source text separately from text changed for analysis.
- Empty files, duplicate IDs, and duplicate text should be checked early.
- Near duplicates survive exact-match checks and need a stated threshold.
- Related documents belong to one family and move between splits together.
- A technically clean corpus can still be incomplete or unrepresentative.

The password-reset answer is now findable without losing the article around
it. The seasonal reprint is out of the collection, its family is recorded, and
the sentence it added is still on the page for someone to judge. For a larger
corpus, keep the same small label beside every document: what this text is,
where it came from, and why it belongs in the collection.

## Sources

- [readr documentation for reading text lines](https://readr.tidyverse.org/reference/read_lines.html)
- [purrr documentation for mapping over vectors](https://purrr.tidyverse.org/reference/map.html)
- [quanteda documentation for creating a corpus](https://quanteda.io/reference/corpus.html)
- [Introduction to the tm text mining package](https://cran.r-project.org/web/packages/tm/vignettes/tm.pdf)
- [Unicode Text Segmentation, UAX #29](https://www.unicode.org/reports/tr29/)
- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
- [Datasheets for Datasets](https://doi.org/10.1145/3458723)
