---
title: "Loading data from an API"
subtitle: "Ask an online service for information and check its reply"
author: "Alex Farach"
description: "Learn how an API request becomes a checked data table in R."
categories: [source data loading, API, JSON]
---

A librarian gathering possible books about natural language processing could
copy search results from a catalog into a table. Search results are candidates,
not recommendations. The librarian still needs to judge whether each record is
relevant, accurate, and useful.

An **application programming interface**, usually called an **API**, gives
software a defined way to ask another service for information. This lesson
reconstructs a query sent to the Open Library Search API and checks the saved
reply. The displayed table comes from a dated response so the published output
can be reproduced even when the service is unavailable or its ranking changes.

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain the request-and-response pattern used by an API;
- add search choices to an API request;
- check whether the request succeeded;
- turn a JSON response into a table in R;
- replay a rate limit, a second page, and a changed schema from saved
  responses; and
- handle changing results, service limits, and secret API keys.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

The first code chunk loads the packages needed to build the request, read files, shape tibbles, and protect displayed text. A tibble is a table that prints its size and each column's type.

## Start with a request and a reply

Every API exchange in this lesson has two parts:

1. R sends a **request** to a web address.
2. The service sends a **response** back.

The request records the librarian's question. The response contains a status
code, descriptive details called headers, and a body holding the requested
data.

Our request uses this **endpoint**, the web address assigned to book searches:
`https://openlibrary.org/search.json`.

## Build the request

We build the request before sending it so each choice remains visible. The
httr2 package adds those choices one step at a time.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(httr2)
library(digest)
library(jsonlite)
library(htmltools)

book_request <- httr2::request(
  "https://openlibrary.org/search.json"
) |>
  httr2::req_url_query(
    q = "natural language processing",
    fields = str_c(
      c("key", "title", "author_name", "first_publish_year"),
      collapse = ","
    ),
    limit = 5
  ) |>
  httr2::req_user_agent(
    str_c(
      "periodic-table-nlp-r/1.0 ",
      "(https://github.com/farach/periodic_table_nlp)"
    )
  ) |>
  httr2::req_timeout(20) |>
  httr2::req_retry(max_tries = 3)

stopifnot(
  inherits(book_request, "httr2_request"),
  str_detect(
    book_request$url,
    fixed("openlibrary.org/search.json")
  )
)

book_request
```

Those `stopifnot()` calls are **assertions**. Each one stops the page from
building if the result it names ever changes. They confirm that httr2 built a
request for the
intended address. The printed request begins with `GET`, the standard method
used here to ask a service to send information. Printing it lets us inspect the
destination and search choices without contacting the service again.

The choices added to the address are called **query parameters**:

| Parameter | What it asks for |
|---|---|
| `q` | Books related to the words "natural language processing." |
| `fields` | Only the four pieces of information used in this lesson. |
| `limit` | No more than five results. |

: Query parameters used in the Open Library request

The user-agent identifies the project making the request. The 20-second timeout
keeps R from waiting indefinitely. The retry setting permits up to three
attempts when a temporary connection problem interrupts the exchange; it does
not guarantee that an unavailable service will respond.

## Understand what happened when the request was sent

The request was sent once on the retrieval date recorded below. httr2's
`req_perform()` function performed the exchange and would have stopped on an
unsuccessful HTTP status such as 404 or 500. The response returned status `200`,
which means the service completed that request successfully.

**JSON**, short for JavaScript Object Notation, is a text format that can hold
named values, lists, and records. Many programming languages can read it, which
makes it common in API responses.

## Keep a dated response

One successful request cannot guarantee that the service will be available
tomorrow or return the same ranking. The repository keeps the JSON response
with its retrieval time, request address, HTTP status, content type, record
level, licensing reference, and fingerprint.

```{r}
fixture_path <- "data/openlibrary-nlp-search.json"
metadata_path <- "data/openlibrary-nlp-search-metadata.csv"

fixture_metadata <- read_csv(
  metadata_path,
  col_types = cols(
    retrieved_at_utc = col_character(),
    request_url = col_character(),
    http_status = col_character(),
    content_type = col_character(),
    record_level = col_character(),
    api_documentation = col_character(),
    licensing_reference = col_character(),
    md5 = col_character()
  )
)
fixture_lines <- read_lines(fixture_path)
fixture_md5 <- digest::digest(
  paste(fixture_lines, collapse = "\n"),
  algo = "md5",
  serialize = FALSE
)

stopifnot(
  inherits(fixture_metadata, "tbl_df"),
  identical(nrow(fixture_metadata), 1L),
  identical(length(fixture_lines), 59L),
  identical(fixture_metadata$md5, fixture_md5),
  str_detect(
    fixture_metadata$retrieved_at_utc,
    "^2026-08-28T[0-9]{2}:[0-9]{2}:[0-9]{2}Z$"
  ),
  identical(
    fixture_metadata$record_level,
    "work-level catalog record"
  ),
  identical(fixture_metadata$http_status, "200"),
  identical(
    fixture_metadata$content_type,
    "application/json"
  ),
  identical(
    fixture_metadata$request_url,
    book_request$url
  )
)

knitr::kable(
  tibble(
    detail = c(
      "Retrieved at",
      "HTTP status",
      "Content type",
      "Record level",
      "File fingerprint"
    ),
    value = c(
      fixture_metadata$retrieved_at_utc,
      fixture_metadata$http_status,
      fixture_metadata$content_type,
      fixture_metadata$record_level,
      fixture_md5
    )
  ),
  col.names = c("Fixture detail", "Recorded value"),
  caption = "Provenance for the saved API response",
  row.names = FALSE
)
```

The MD5 fingerprint detects an accidental change to the saved file. It is not
proof that the source is accurate or trustworthy. The request address and API
documentation remain in the metadata file for a later audit.

## Turn the saved response into a table

Open Library places search results inside a JSON section named `docs`.
`jsonlite::fromJSON()` translates that section into rows and columns because
this response has the same fields for each catalog record.

```{r}
fixture_data <- jsonlite::fromJSON(
  fixture_path,
  simplifyVector = TRUE
)
book_records <- as_tibble(fixture_data$docs)

format_authors <- function(author_names) {
  if (length(author_names) == 0 || all(is.na(author_names))) {
    return("Not listed")
  }

  str_c(author_names, collapse = ", ")
}

books <- book_records |>
  transmute(
    title = title,
    authors = map_chr(author_name, format_authors),
    first_published = first_publish_year,
    catalog_url = str_c("https://openlibrary.org", key)
  )

stopifnot(
  is.data.frame(book_records),
  inherits(books, "tbl_df"),
  identical(nrow(books), 5L),
  all(str_length(books$title) > 0),
  all(
    str_detect(
     books$catalog_url,
     fixed("https://openlibrary.org/works/")
    )
  )
)

escape_html <- function(value) {
  as.character(htmltools::htmlEscape(value))
}

catalog_links <- map2_chr(
  books$catalog_url,
  books$title,
  \(url, title) {
    safe_url <- as.character(
      htmltools::htmlEscape(url, attribute = TRUE)
    )
    safe_title <- as.character(
      htmltools::htmlEscape(
        title,
        attribute = TRUE
      )
    )
    sprintf(
      '<a href="%s" aria-label="View Open Library record for %s">View record</a>',
      safe_url,
      safe_title
    )
  }
)

display_books <- books |>
  transmute(
    title = escape_html(title),
    authors = escape_html(authors),
    first_published = first_published,
    catalog_link = catalog_links
  )

knitr::kable(
  display_books,
  col.names = c(
    "Title",
    "Author or authors",
    "First published",
    "Open Library record"
  ),
  caption = str_c(
    "Five candidate work-level catalog records retrieved",
    " ",
    fixture_metadata$retrieved_at_utc
  ),
  row.names = FALSE,
  escape = FALSE
)
```

Open Library returns work-level catalog records by default. A row can represent
many editions of the same work. The table contains candidate search results,
not a checked reading list or a judgment about a book's quality.

The authors and titles are escaped before display so API text cannot become
page markup. Only the validated Open Library addresses are rendered as links.

## Live data can change

Open Library may add records, correct dates, or reorder results. Nothing in this
page contacts the service while it builds, so the displayed table comes from the
saved response and stays stable.

When the request is sent again, the checks worth applying describe a bounded
structure rather than five specific titles:

- the service returned a successful status;
- the response is JSON;
- the response contains between one and five records; and
- every returned record has a title.

The saved fixture makes the displayed table repeatable. Its metadata records
when and how it was retrieved. Open Library states that the Internet Archive
does not assert new proprietary rights over the database, while warning that
existing rights may vary by contribution and jurisdiction. Reusers should read
the current licensing statement rather than assume a blanket license.

Respect the service's rate limit and stated preference for low-volume,
human-facing requests.

## Replay the awkward responses

A successful reply is the easy case. The three awkward ones are a refusal, a
result set too large for one reply, and a reply whose fields have moved. Each is
reconstructed below from the status codes and body shapes the service documents,
built inside the lesson with `httr2::response()`. No request leaves the machine.

```{r}
saved_response <- function(status, body, extra_headers = list()) {
  httr2::response(
    status_code = status,
    url = book_request$url,
    method = "GET",
    headers = c(
      list("content-type" = "application/json"),
      extra_headers
    ),
    body = charToRaw(body)
  )
}

make_docs <- function(keys, titles) {
  str_c(
    sprintf(
      '{"key": "%s", "title": "%s"}',
      keys,
      titles
    ),
    collapse = ", "
  )
}

rate_limited <- saved_response(
  429L,
  '{"error": "too many requests"}',
  list("retry-after" = "30")
)

page_one <- saved_response(
  200L,
  str_c(
    '{"numFound": 7, "start": 0, "docs": [',
    make_docs(
      sprintf("/works/OL%d", 1:5),
      sprintf("Candidate title %d", 1:5)
    ),
    ']}'
  )
)

page_two <- saved_response(
  200L,
  str_c(
    '{"numFound": 7, "start": 5, "docs": [',
    make_docs(
      c("/works/OL5", "/works/OL6"),
      c("Candidate title 5", "Candidate title 6")
    ),
    ']}'
  )
)

field_renamed <- saved_response(
  200L,
  str_c(
    '{"numFound": 1, "start": 0, "docs": [',
    '{"key": "/works/OL7", "name": "Title moved to a new field"}',
    ']}'
  )
)

stopifnot(
  identical(httr2::resp_status(rate_limited), 429L),
  identical(httr2::resp_status(page_one), 200L),
  identical(
    httr2::resp_header(rate_limited, "retry-after"),
    "30"
  ),
  identical(
    httr2::resp_content_type(page_one),
    "application/json"
  )
)

httr2::resp_status(rate_limited)
```

The reader now has four saved replies and one function that has to survive all
of them.

```{r}
required_fields <- c("key", "title")

read_saved_page <- function(response, page_size) {
  status <- httr2::resp_status(response)

  if (identical(status, 429L)) {
    return(list(
      outcome = "rate limited",
      wait_seconds = as.integer(
        httr2::resp_header(response, "retry-after")
      ),
      reported_total = NA_integer_,
      records = tibble(
        key = character(),
        title = character()
      )
    ))
  }

  if (!identical(status, 200L)) {
    return(list(
      outcome = "http error",
      wait_seconds = NA_integer_,
      reported_total = NA_integer_,
      records = tibble(
        key = character(),
        title = character()
      )
    ))
  }

  body <- httr2::resp_body_json(response)
  documents <- body$docs
  absent <- required_fields[
    !map_lgl(
      required_fields,
      \(field) all(map_lgl(documents, \(record) field %in% names(record)))
    )
  ]

  if (length(absent) > 0) {
    return(list(
      outcome = str_c(
        "schema failure: missing ",
        str_c(absent, collapse = ", ")
      ),
      wait_seconds = NA_integer_,
      reported_total = as.integer(body$numFound),
      records = tibble(
        key = character(),
        title = character()
      )
    ))
  }

  list(
    outcome = if (length(documents) < page_size) {
      "last page"
    } else {
      "full page"
    },
    wait_seconds = NA_integer_,
    reported_total = as.integer(body$numFound),
    records = tibble(
      key = map_chr(documents, "key"),
      title = map_chr(documents, "title")
    )
  )
}

page_size <- 5L
refused <- read_saved_page(rate_limited, page_size)
first_page <- read_saved_page(page_one, page_size)
second_page <- read_saved_page(page_two, page_size)
drifted <- read_saved_page(field_renamed, page_size)

collected <- bind_rows(
  first_page$records,
  second_page$records
)
distinct_records <- collected |>
  distinct(key, .keep_all = TRUE)
shortfall <- first_page$reported_total - nrow(distinct_records)

stopifnot(
  identical(refused$outcome, "rate limited"),
  identical(refused$wait_seconds, 30L),
  identical(nrow(refused$records), 0L),
  identical(first_page$outcome, "full page"),
  identical(second_page$outcome, "last page"),
  identical(drifted$outcome, "schema failure: missing title"),
  identical(nrow(drifted$records), 0L),
  identical(nrow(collected), 7L),
  identical(nrow(distinct_records), 6L),
  identical(first_page$reported_total, 7L),
  identical(shortfall, 1L)
)

knitr::kable(
  tibble(
    saved_reply = c(
      "429 with Retry-After",
      "page 1 of results",
      "page 2 of results",
      "title field renamed"
    ),
    outcome = c(
      refused$outcome,
      first_page$outcome,
      second_page$outcome,
      drifted$outcome
    ),
    records_returned = c(
      nrow(refused$records),
      nrow(first_page$records),
      nrow(second_page$records),
      nrow(drifted$records)
    )
  ),
  col.names = c("Saved reply", "Outcome", "Records used"),
  caption = "Four saved replies and what the reading function did with each",
  row.names = FALSE
)
```

Four things happened, and none of them was a crash or a quietly empty table.

The refusal returned no records and a waiting time taken from the
`Retry-After` header, which is the service telling the client how long to pause.
Guessing that number, or retrying immediately, is how a temporary limit becomes
a ban.

The first page filled its five slots, which is the signal that more results
exist. The second page returned two, fewer than the page size, which is the
signal to stop asking.

The renamed field is the quiet one. The record still parsed as JSON and still
had a key. Only the explicit list of required fields turned it into a reported
failure instead of a table with a missing column.

### Do not let the arithmetic pass unexamined

The service reported seven matching records. Seven rows arrived across two
pages. Six of them were distinct, because one work appeared on both pages, which
happens when ranking shifts between two requests.

```{r}
pagination_report <- tibble(
  measure = c(
    "reported by the service",
    "rows received",
    "distinct works kept",
    "unexplained shortfall"
  ),
  value = c(
    first_page$reported_total,
    nrow(collected),
    nrow(distinct_records),
    shortfall
  )
)

stopifnot(
  identical(pagination_report$value, c(7L, 7L, 6L, 1L)),
  identical(
    distinct_records$key,
    sprintf("/works/OL%d", 1:6)
  )
)

knitr::kable(
  pagination_report,
  col.names = c("Measure", "Count"),
  caption = "Reported, received, and kept records after paging",
  row.names = FALSE
)
```

The shortfall of one is the useful number. A collection script that reported six
records without it would look complete. Record the totals the service claimed,
the rows you received, and the rows you kept, and treat any gap as a question
about the collection rather than a rounding detail.

## Keep API keys private

Open Library's search endpoint does not require a key. Many other APIs do. An
**API key** is a secret value that identifies an account and may permit
billable or private access.

Never place a real key in:

- an R or Quarto file;
- a public repository;
- a screenshot; or
- printed output.

Store secrets outside the project, usually in an **environment variable**, a
named setting that R can read at run time. If a key is exposed, revoke it and
create a replacement.

## When an API request fails

A failed request leaves the librarian without a checked list. Turning that
failure into an empty table would hide the difference between "no books found"
and "the service did not answer." Report the failure and keep the reported
error. Then ask:

1. Is the device connected to the internet?
2. Is the endpoint address correct?
3. Does the service require a key or other permission?
4. Did the request exceed a rate limit?
5. Did the response format change?

The code in this lesson stops when the request fails or the response lacks the
expected structure. The interruption is visible rather than being published as
an apparently complete result.

## What to remember

- An API request asks a service for structured information.
- Query parameters describe what to return and how much.
- Check the status and response type before using the body.
- Some JSON structures can be reshaped into rows and columns in R.
- A refusal, a second page, and a renamed field can all be rehearsed from saved
  replies before any request is sent.
- Honor the waiting time a service sends instead of guessing one.
- Compare the total a service reports with the rows you actually kept.
- Live API results may change even when the code stays the same.
- A dated fixture can preserve the output used in a published example.
- Keys and other secrets never belong in published code.

The librarian ends with five candidate records, not five recommendations. The
request can be tried again, while the dated fixture preserves the reply shown
here. Human judgment still decides which books belong on the reading list.

## Sources

- [Open Library Search API documentation](https://openlibrary.org/dev/docs/api/search)
- [Open Library API usage and rate limits](https://openlibrary.org/developers/api)
- [Open Library licensing statement](https://openlibrary.org/developers/licensing)
- [RFC 8259: The JSON Data Interchange Format](https://www.rfc-editor.org/rfc/rfc8259)
- [httr2 request documentation](https://httr2.r-lib.org/reference/request.html)
- [httr2 request performance and error handling](https://httr2.r-lib.org/reference/req_perform.html)
- [httr2 JSON response documentation](https://httr2.r-lib.org/reference/resp_body_raw.html)
