---
title: "Collecting text from a web page"
subtitle: "Scrape a saved job board without losing its source"
author: "Alex Farach"
description: "Learn how to extract job titles and training details from structured HTML with R."
categories: [source data loading, scraping, workforce research]
---

The fictional Riverton Workforce Lab begins with six job postings saved from a
job board. A worker comparing the notices would need to keep each training
offer, schedule, experience requirement, and named skill attached to the right
job. Copying the lines into one block would make omissions and source mix-ups
hard to audit.

**Web scraping** uses code to collect information from web pages. The team will
scrape the saved page, keep the job and sentence identifiers, and check every
record before it becomes research data. This first stage collects what the
notices say; it does not classify the sentences or generalize from them.

::: {.callout-note}
The employers, postings, and Riverton Workforce Lab are fictional. The HTML
file was written for this lesson and contains no personal information.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain how HTML gives parts of a page names and structure;
- select repeated job cards with rvest;
- keep source identifiers beside extracted text;
- compare scraped output with an expected fixture; and
- identify legal, ethical, and technical checks needed before live scraping.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Begin with an archived page

Live websites change. A saved source file lets another researcher inspect the
same page later. The team records the file path, size, and fingerprint before
extracting anything.

```{r}
job_page_path <- "data/workforce/job-board.html"
job_page_lines <- readLines(
  job_page_path,
  encoding = "UTF-8",
  warn = TRUE
)
job_page_fingerprint <- digest::digest(
  paste(job_page_lines, collapse = "\n"),
  algo = "md5",
  serialize = FALSE
)
job_page_metadata <- read.csv(
  "data/workforce/job-board-metadata.csv",
  colClasses = "character",
  check.names = FALSE
)

stopifnot(
  file.exists(job_page_path),
  file.info(job_page_path)$size > 1000,
  identical(nrow(job_page_metadata), 1L),
  identical(
    job_page_fingerprint,
    job_page_metadata$md5
  ),
  identical(
    job_page_metadata$retrieval_status,
    "synthetic archived page"
  ),
  grepl(
    "^[0-9a-f]{32}$",
    job_page_fingerprint
  )
)

data.frame(
  source_file = job_page_path,
  bytes = file.info(job_page_path)$size,
  fingerprint = job_page_fingerprint
)
```

The fingerprint is compared with a stored value. It changes if the archived
HTML changes and can reveal an accidental edit. It cannot prove that the
postings are true or complete.

## Read the page structure

HTML uses elements such as headings, paragraphs, and lists to organize a page.
The example page places every posting inside an element whose class is
`job-card`. A **CSS selector** is a short pattern that names the elements to
collect.

```{r}
job_page <- rvest::read_html(job_page_path)
job_cards <- rvest::html_elements(
  job_page,
  ".job-card"
)

stopifnot(
  identical(length(job_cards), 6L)
)

length(job_cards)
```

R found six job cards. This count is the first signal that the selector matches
the archived page we expected.

## Extract one field at a time

Each card contains a job ID, posting date, title, employer, and location. The
small helper below extracts one text field from each card.

```{r}
card_text <- function(cards, selector) {
  vapply(
    cards,
    function(card) {
      elements <- card |>
        rvest::html_elements(selector)
      stopifnot(identical(length(elements), 1L))

      elements |>
        rvest::html_text2()
    },
    character(1)
  )
}

jobs <- data.frame(
  job_id = rvest::html_attr(
    job_cards,
    "data-job-id"
  ),
  posted = as.Date(
    rvest::html_attr(
      job_cards,
      "data-posted"
    )
  ),
  title = card_text(job_cards, ".job-title"),
  employer = card_text(job_cards, ".employer"),
  location = card_text(job_cards, ".location"),
  stringsAsFactors = FALSE
)

stopifnot(
  identical(jobs$job_id, sprintf("J%03d", 1:6)),
  !anyDuplicated(jobs$job_id),
  all(!is.na(jobs$posted)),
  all(nzchar(jobs$title)),
  all(nzchar(jobs$employer)),
  all(nzchar(jobs$location)),
  identical(jobs$title[1], "Data analyst trainee")
)

knitr::kable(
  jobs,
  col.names = c(
    "Job ID",
    "Posted",
    "Title",
    "Employer",
    "Location"
  ),
  caption = "Six postings extracted from the archived job board",
  row.names = FALSE
)
```

The job ID remains attached to every title. Later, a schedule or experience
requirement must still point back to the posting that stated it.

## Keep sentences connected to their postings

The details in each card are list items with the class `job-detail`. We extract
them in page order and add stable sentence IDs.

```{r}
details_by_job <- lapply(
  seq_along(job_cards),
  function(card_number) {
    details <- job_cards[[card_number]] |>
      rvest::html_elements(".job-detail") |>
      rvest::html_text2()

    data.frame(
      document_id = jobs$job_id[card_number],
      text = details,
      stringsAsFactors = FALSE
    )
  }
)

scraped_sentences <- do.call(
  rbind,
  details_by_job
)
scraped_sentences$sentence_id <- sprintf(
  "s%03d",
  seq_len(nrow(scraped_sentences))
)
scraped_sentences <- scraped_sentences[
  c("sentence_id", "document_id", "text")
]

expected_sentences <- read.csv(
  "data/workforce/workforce_sentences.csv",
  colClasses = "character",
  check.names = FALSE
)
expected_job_sentences <- expected_sentences[
  expected_sentences$document_id != "F001",
  c("sentence_id", "document_id", "text")
]

stopifnot(
  identical(nrow(scraped_sentences), 22L),
  identical(
    scraped_sentences,
    expected_job_sentences
  )
)

knitr::kable(
  head(scraped_sentences, 8),
  col.names = c("Sentence ID", "Job ID", "Text"),
  caption = "The first eight extracted job-detail sentences",
  row.names = FALSE
)
```

The exact comparison succeeds: all 22 sentences match the fixture in
the same order. This check would stop the build if a selector silently dropped,
duplicated, or reordered text.

Both the HTML and expected table were written for this lesson, so the
comparison tests parser behavior rather than independent truth. A live
collection would need a separately captured audit sample.

## A saved page is still incomplete evidence

The archived HTML shows what was captured, not every posting that existed.
Before scraping a live site, the team must check:

1. the site's terms and robots policy;
2. whether collection is lawful and expected in the relevant jurisdiction;
3. whether personal or sensitive information appears;
4. how often requests may be sent;
5. whether page content is loaded later by JavaScript;
6. which dates, filters, and geographic limits shaped the results; and
7. how the original HTML and retrieval time will be preserved.

The Robots Exclusion Protocol communicates crawler preferences. It is not a
grant of legal permission, and permission does not settle every ethical
question. A public page can still contain information that should not be
collected or republished.

## What to remember

- Scraping turns page structure into data; it does not establish that the data
  is true or representative.
- Archive the source and record when it was collected.
- Keep source IDs attached to extracted text.
- Assert expected counts and content so selectors cannot fail silently.
- Review terms, rate limits, privacy, and research ethics before live
  collection.

The team can now inspect 22 sentences without detaching them from the six
notices. One source remains outside that table: a fictional training flyer
saved as an image. That image shifts the investigation from HTML structure to
OCR. Its source must still be preserved, and the extracted wording must be
checked rather than treated as certain.

## Sources

- [rvest overview](https://rvest.tidyverse.org/)
- [rvest selector documentation](https://rvest.tidyverse.org/reference/html_element.html)
- [RFC 9309: Robots Exclusion Protocol](https://www.rfc-editor.org/rfc/rfc9309)
- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
