---
title: "Collecting text from a web page"
subtitle: "Scrape a saved job board without losing its source"
author: "Alex Farach"
description: "Learn how to extract job titles and training details from structured HTML with R."
categories: [source data loading, scraping, workforce research]
---

The fictional Riverton Workforce Lab begins with six job postings saved from a
job board. A worker comparing the notices would need to keep each training
offer, schedule, experience requirement, and named skill attached to the right
job. Copying the lines into one block would make omissions and source mix-ups
hard to audit.

**Web scraping** uses code to collect information from web pages. The team will
scrape the saved page, keep the job and sentence identifiers, and check every
record before it becomes research data. This first stage collects what the
notices say; it does not classify the sentences or generalize from them.

The whole lesson works on static archived HTML: a file whose text is already in
the markup when it is saved. That is the scope of what follows, and the last
sections show two ordinary pages where the method stops working.

::: {.callout-note}
The employers, postings, and Riverton Workforce Lab are fictional. The HTML
file was written for this lesson and contains no personal information.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- explain how HTML gives parts of a page names and structure;
- select repeated job cards with rvest;
- keep source identifiers beside extracted text;
- compare scraped output with an expected fixture;
- notice a renamed class and a page whose text is not in the markup; and
- identify legal, ethical, and technical checks needed before live scraping.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

The first code chunk loads the packages needed to read the archived page, extract HTML fields, and build tibbles. A tibble is a table that prints its size and each column's type.

## Begin with an archived page

Live websites change. A saved source file lets another researcher inspect the
same page later. The team records the file path, size, and fingerprint before
extracting anything.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(purrr)
library(stringr)
library(digest)
library(rvest)

job_page_path <- "data/workforce/job-board.html"
job_page_lines <- read_lines(job_page_path)
job_page_fingerprint <- digest::digest(
  paste(job_page_lines, collapse = "\n"),
  algo = "md5",
  serialize = FALSE
)
job_page_metadata <- read_csv(
  "data/workforce/job-board-metadata.csv",
  col_types = cols(
    source_file = col_character(),
    created_for_lesson = col_character(),
    retrieval_status = col_character(),
    records = col_character(),
    md5 = col_character(),
    rights_note = col_character()
  )
)
# A size read from disk counts the line endings the local copy happens to
# use, so it changes between systems. Measuring the newline-joined text, the
# same text the fingerprint covers, describes the content instead.
job_page_size <- nchar(
  paste(job_page_lines, collapse = "\n"),
  type = "bytes"
)

tibble(
  source_file = job_page_path,
  bytes = job_page_size,
  fingerprint = job_page_fingerprint
)
```

```{r}
#| include: false

stopifnot(
  file.exists(job_page_path),
  identical(job_page_size, 3441L),
  identical(length(job_page_lines), 82L),
  inherits(job_page_metadata, "tbl_df"),
  identical(nrow(job_page_metadata), 1L),
  identical(
    job_page_fingerprint,
    job_page_metadata$md5
  ),
  identical(
    job_page_metadata$retrieval_status,
    "synthetic archived page"
  ),
  str_detect(
    job_page_fingerprint,
    "^[0-9a-f]{32}$"
  )
)
```

The fingerprint is compared with a stored value. It changes if the archived
HTML changes and can reveal an accidental edit. It cannot prove that the
postings are true or complete.

## Read the page structure

HTML uses elements such as headings, paragraphs, and lists to organize a page.
The example page places every posting inside an element whose class is
`job-card`. A **CSS selector** is a short pattern that names the elements to
collect.

```{r}
job_page <- rvest::read_html(job_page_path)
job_cards <- rvest::html_elements(
  job_page,
  ".job-card"
)

length(job_cards)
```

```{r}
#| include: false

stopifnot(
  identical(length(job_cards), 6L)
)
```

R found six job cards. This count is the first signal that the selector matches
the archived page we expected.

## Extract one field at a time

Each card contains a job ID, posting date, title, employer, and location. The
small helper below extracts one text field from each card.

```{r}
card_text <- function(cards, selector) {
  map_chr(
    cards,
    \(card) {
      elements <- card |>
        rvest::html_elements(selector)
      stopifnot(identical(length(elements), 1L))

      elements |>
        rvest::html_text2()
    }
  )
}

jobs <- tibble(
  job_id = rvest::html_attr(
    job_cards,
    "data-job-id"
  ),
  posted = as.Date(
    rvest::html_attr(
      job_cards,
      "data-posted"
    )
  ),
  title = card_text(job_cards, ".job-title"),
  employer = card_text(job_cards, ".employer"),
  location = card_text(job_cards, ".location")
)

knitr::kable(
  jobs,
  col.names = c(
    "Job ID",
    "Posted",
    "Title",
    "Employer",
    "Location"
  ),
  caption = "Six postings extracted from the archived job board",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  inherits(jobs, "tbl_df"),
  identical(jobs$job_id, sprintf("J%03d", 1:6)),
  n_distinct(jobs$job_id) == nrow(jobs),
  all(!is.na(jobs$posted)),
  all(str_length(jobs$title) > 0),
  all(str_length(jobs$employer) > 0),
  all(str_length(jobs$location) > 0),
  identical(jobs$title[1], "Data analyst trainee")
)
```

The job ID remains attached to every title. Later, a schedule or experience
requirement must still point back to the posting that stated it.

## Keep sentences connected to their postings

The details in each card are list items with the class `job-detail`. We extract
them in page order and add stable sentence IDs.

```{r}
scraped_sentences <- map2(
  job_cards,
  jobs$job_id,
  \(card, job_id) {
    details <- card |>
      rvest::html_elements(".job-detail") |>
      rvest::html_text2()

    tibble(
      document_id = job_id,
      text = details
    )
  }
) |>
  list_rbind() |>
  mutate(
    sentence_id = sprintf("s%03d", row_number()),
    .before = document_id
  )

expected_sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)
expected_job_sentences <- expected_sentences |>
  filter(document_id != "F001") |>
  select(sentence_id, document_id, text) |>
  as_tibble()

knitr::kable(
  scraped_sentences |>
    slice_head(n = 8),
  col.names = c("Sentence ID", "Job ID", "Text"),
  caption = "The first eight extracted job-detail sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  inherits(scraped_sentences, "tbl_df"),
  inherits(expected_job_sentences, "tbl_df"),
  identical(nrow(scraped_sentences), 22L),
  identical(
    scraped_sentences,
    expected_job_sentences
  )
)
```

The exact comparison succeeds: all 22 sentences match the fixture in
the same order. This check would stop the build if a selector silently dropped,
duplicated, or reordered text.

Both the HTML and expected table were written for this lesson, so the
comparison tests parser behavior rather than independent truth. A live
collection would need a separately captured audit sample.

## When the page changes shape

The archived file is stable because it is a file. A live board is maintained by
people who redesign it. The two pages below are written into the lesson as short
HTML strings, and both would break the code above in ways worth telling apart.

The first is a redesign. The employer's new template renames `job-card` to
`posting` and moves the title into a heading.

```{r}
redesigned_html <- str_c(
  "<html><body>",
  "<div class='posting' data-job-id='J007'>",
  "<h3 class='posting-title'>Warehouse trainee</h3>",
  "<ul><li class='posting-detail'>Paid training is provided.</li></ul>",
  "</div>",
  "</body></html>"
)

redesigned_page <- rvest::read_html(redesigned_html)
old_selector_hits <- length(
  rvest::html_elements(redesigned_page, ".job-card")
)
new_selector_hits <- length(
  rvest::html_elements(redesigned_page, ".posting")
)

collect_cards <- function(page, selector, expected) {
  cards <- rvest::html_elements(page, selector)
  if (!identical(length(cards), expected)) {
    stop(
      str_c(
        "selector ",
        selector,
        " found ",
        length(cards),
        " cards; expected ",
        expected
      ),
      call. = FALSE
    )
  }

  list(
    selector = selector,
    found = length(cards),
    expected = expected,
    status = "counts agree"
  )
}

drift_error <- tryCatch(
  collect_cards(redesigned_page, ".job-card", 1L),
  error = conditionMessage
)

knitr::kable(
  tibble(
    page = c("archived board", "redesigned page"),
    selector = ".job-card",
    cards_found = c(
      length(job_cards),
      old_selector_hits
    ),
    cards_expected = c(6L, 1L)
  ),
  col.names = c("Page", "Selector", "Cards found", "Cards expected"),
  caption = "The same selector against two versions of one site",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(old_selector_hits, 0L),
  identical(new_selector_hits, 1L),
  identical(
    drift_error,
    "selector .job-card found 0 cards; expected 1"
  ),
  identical(
    collect_cards(job_page, ".job-card", 6L)$status,
    "counts agree"
  )
)
```

Zero matches is the dangerous result, because zero matches is also what an
empty page produces. Without the expected count, the scraper would write an
empty table and the day's collection would look like a day with no postings.
An expected count turns a silent redesign into a stop.

The second page is harder. Its visible container has no job text. The same file
stores a JSON payload in a script element that application code could read and
render later. This fixture does not execute that JavaScript; it separates
visible markup from embedded data.

```{r}
dynamic_html <- str_c(
  "<html><body>",
  "<div id='board'></div>",
  "<script type='application/json' id='board-data'>",
  '{"jobs": [{"id": "J008", "title": "Night stocker"}]}',
  "</script>",
  "</body></html>"
)

dynamic_page <- rvest::read_html(dynamic_html)
visible_container_text <- rvest::html_element(
  dynamic_page,
  "#board"
) |>
  rvest::html_text2()
payload_text <- rvest::html_element(
  dynamic_page,
  "#board-data"
) |>
  rvest::html_text2()

tibble(
  source = c("visible container", "script payload"),
  characters = c(
    str_length(visible_container_text),
    str_length(payload_text)
  ),
  mentions_the_job = c(
    str_detect(visible_container_text, fixed("Night stocker")),
    str_detect(payload_text, fixed("Night stocker"))
  )
)
```

```{r}
#| include: false

stopifnot(
  identical(
    length(rvest::html_elements(dynamic_page, ".job-card")),
    0L
  ),
  identical(visible_container_text, ""),
  str_detect(payload_text, fixed("Night stocker")),
  !str_detect(visible_container_text, fixed("Night stocker"))
)
```

The container holds no characters. The job title exists only in the embedded
script payload, which rvest can read as text but does not render into the
container.

This example is deliberately gentle because the payload is already in the
file. Many sites fetch data separately after the page loads. A saved copy of
their initial HTML may contain nothing to extract, and an HTML parser cannot
recover a response that was never saved.

Three routes exist from there, and they differ in cost and in what they promise.
A documented API returns the same data with stated terms, and is the first thing
to look for. The underlying data request the page itself makes can sometimes be
called directly, which is faster but relies on an interface nobody promised to
keep. A headless browser, driven from R through chromote or from Python through
Playwright or Selenium, runs the page as a browser would; it is the general
answer and by far the most expensive to run and maintain.

Whichever route is chosen, record it. Text recovered from a rendered page and
text read from a saved file are not the same evidence, and a reader deserves to
know which one produced a number.


## A saved page is still incomplete evidence

The archived HTML shows what was captured, not every posting that existed.
Before scraping a live site, the team must check:

1. the site's terms and robots policy;
2. whether collection is lawful and expected in the relevant jurisdiction;
3. whether personal or sensitive information appears;
4. how often requests may be sent;
5. whether page content is loaded later by JavaScript;
6. which dates, filters, and geographic limits shaped the results; and
7. how the saved HTML and retrieval time will be preserved.

The Robots Exclusion Protocol communicates crawler preferences. It is not a
grant of legal permission, and permission does not settle every ethical
question. A public page can still contain information that should not be
collected or republished.

## What to remember

- Scraping turns page structure into data; it does not establish that the data
  is true or representative.
- Archive the source and record when it was collected.
- Keep source IDs attached to extracted text.
- Assert expected counts and content so selectors cannot fail silently.
- Zero matches and an empty page look identical without an expected count.
- Some pages hold their text outside the markup, where an HTML parser cannot
  reach it.
- Review terms, rate limits, privacy, and research ethics before live
  collection.

The team can now inspect 22 sentences without detaching them from the six
notices. One source remains outside that table: a fictional training flyer
saved as an image. That image shifts the investigation from HTML structure to
OCR. Its source must still be preserved, and the extracted wording must be
checked rather than treated as certain.

## Sources

- [rvest overview](https://rvest.tidyverse.org/)
- [rvest selector documentation](https://rvest.tidyverse.org/reference/html_element.html)
- [RFC 9309: Robots Exclusion Protocol](https://www.rfc-editor.org/rfc/rfc9309)
- [Data Statements for Natural Language Processing](https://aclanthology.org/Q18-1041/)
