---
title: "Creating a small text app"
subtitle: "Wrap a keyword-in-context search without embedding a dead form"
description: "Learn how a Shiny app separates its visible controls from server logic and how to test the text search without running the app on the lesson page."
categories: [information visualization, Shiny, interactive apps]
---

Leila helps a local history class inspect inaugural speeches. The class wants a
search box: type a word, see the nearby words, and decide whether the passages
are worth reading.

That request calls for **interactive app creation**: building a small interface
so another person can use an NLP function without writing R code. In Shiny, the
**UI** is the page people see. The **server** is the R function that reads the
input and computes the output.

This lesson builds the app object and tests its server logic during render. It
does not run the app inside this static page. A printed text box here would look
interactive, but there would be no Shiny process behind it.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- define a small Shiny UI, server function, and app object;
- wrap a keyword-in-context search from quanteda;
- trim and validate text input on the server;
- test Shiny server logic with `testServer()`;
- show UI HTML as text without putting live controls on the page; and
- state what is and is not checked by a server-only test.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build the search job first

The app will search paragraphs from the inaugural-address corpus bundled with
quanteda. A **keyword-in-context** display, or **KWIC**, puts the matched word in
the middle and a few nearby words on each side. Lesson 25 introduced KWIC as an
inspection tool; here it is the function placed behind an app.

```{r}
library(dplyr)
library(knitr)
library(quanteda)
library(shiny)
library(stringr)
library(tibble)

source("R/inaugural-corpus.R")

paragraphs <- inaugural_paragraphs()
inaugural_tokens <- paragraphs |>
  select(paragraph_id, paragraph) |>
  tibble::deframe() |>
  corpus() |>
  tokens(remove_punct = TRUE)

find_in_context <- function(word, window = 5L) {
  kwic(
    inaugural_tokens,
    pattern = word,
    window = window,
    valuetype = "fixed",
    case_insensitive = TRUE
  ) |>
    as.data.frame() |>
    as_tibble() |>
    transmute(
      paragraph_id = docname,
      before = pre,
      keyword,
      after = post
    )
}

search_setup <- tibble(
  item = c(
    "paragraphs",
    "speeches",
    "Shiny version",
    "matching rule"
  ),
  value = c(
    as.character(nrow(paragraphs)),
    as.character(n_distinct(paragraphs$speech_id)),
    as.character(packageVersion("shiny")),
    "fixed, case-insensitive token matching"
  )
)

kable(
  search_setup,
  col.names = c("Item", "Value"),
  caption = "Inputs and package version for the small app",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c("dplyr", "knitr", "quanteda", "shiny", "stringr", "tibble") %in%
    loadedNamespaces()),
  identical(nrow(paragraphs), 1377L),
  identical(n_distinct(paragraphs$speech_id), 60L),
  identical(as.character(packageVersion("shiny")), "1.14.0"),
  identical(nrow(find_in_context("liberty")), 118L),
  identical(nrow(find_in_context("Liberty")), 118L),
  identical(nrow(find_in_context("*")), 0L),
  identical(nrow(find_in_context("(")), 0L),
  identical(nrow(find_in_context("self-government")), 28L)
)
```

The important choice is `valuetype = "fixed"`. Without it, quanteda treats the
pattern as glob syntax by default. In this app, a typed asterisk is a literal
asterisk, not a request to match every token.

## Define the app without running it

The UI has one visible label, one text input, a short status line, and a table
for matches. The server trims the typed word before `req()` sees it, then uses
`validate(need())` for invalid input that deserves a visible message.

```{r}
max_word_chars <- 30L
max_rows_shown <- 8L

ui <- tagList(
  tags$title("Inaugural KWIC explorer"),
  fluidPage(
    title = "Inaugural KWIC explorer",
    lang = "en",
    h2("Inaugural KWIC explorer"),
    p(
      "Type one word, then press Enter. Hyphenated words are allowed; spaces are not."
    ),
    textInput(
      inputId = "word",
      label = "Word to find (one word, 30 characters or fewer)",
      value = "",
      placeholder = "liberty",
      updateOn = "blur"
    ),
    textOutput("status"),
    tableOutput("matches")
  ),
)
attr(ui, "lang") <- "en"

server <- function(input, output, session) {
  context_rows <- reactive({
    word <- str_trim(input$word %||% "")
    req(nzchar(word))
    validate(
      need(nchar(word) <= max_word_chars, "Enter at most 30 characters."),
      need(!str_detect(word, "\\s"), "Enter one word, with no spaces.")
    )

    find_in_context(word)
  })

  output$status <- renderText({
    rows <- context_rows()
    word <- str_trim(input$word %||% "")

    if (nrow(rows) == 0L) {
      paste0(
        "No token in the punctuation-free search matches '",
        word,
        "'."
      )
    } else {
      shown <- min(nrow(rows), max_rows_shown)
      paste0(
        "Found ",
        nrow(rows),
        if_else(nrow(rows) == 1L, " line", " lines"),
        " for '",
        word,
        "'; showing the first ",
        shown,
        "."
      )
    }
  })

  output$matches <- renderTable({
    context_rows() |>
      slice_head(n = max_rows_shown)
  })
}

app <- shinyApp(ui = ui, server = server)
```

```{r}
#| include: false

server_source <- paste(deparse(body(server)), collapse = "\n")
search_source <- paste(deparse(body(find_in_context)), collapse = "\n")
forbidden_server_calls <- c(
  "\\bHTML\\s*\\(",
  "\\brenderUI\\s*\\(",
  "\\bprint\\s*\\(",
  "\\bmessage\\s*\\(",
  "\\bcat\\s*\\(",
  "\\bwrite\\s*\\(",
  "\\bwriteLines\\s*\\(",
  "\\bsaveRDS\\s*\\(",
  "\\bbindCache\\s*\\("
)
ui_html_for_checks <- paste(
  c(
    "<html lang=\"en\">",
    as.character(ui),
    "</html>"
  ),
  collapse = "\n"
)

stopifnot(
  inherits(ui, "shiny.tag.list"),
  is.function(server),
  inherits(app, "shiny.appobj"),
  all(!str_detect(server_source, regex(str_c(forbidden_server_calls, collapse = "|")))),
  str_detect(server_source, fixed("renderText")),
  str_detect(server_source, fixed("renderTable")),
  str_detect(search_source, fixed("valuetype = \"fixed\"")),
  identical(attr(ui, "lang"), "en"),
  str_detect(ui_html_for_checks, fixed("<title>Inaugural KWIC explorer</title>")),
  str_detect(ui_html_for_checks, fixed("<html lang=\"en\">"))
)
```

`reactive()` wraps the KWIC calculation so both outputs can use the same current
rows after the input changes. `%||%` turns a missing input into empty text.
`req()` stops without a displayed message until a word exists, and
`validate(need())` stops with a message the reader can see when the word breaks
the app's rules. `renderText()` and `renderTable()` turn R values into page text
and a table.

Typed text goes only through `renderText()` and `renderTable()`, not through raw
HTML, so markup-like input is treated as text. The length and spacing limits
live on the server because browser-side checks can be bypassed.

The server keeps the typed word only in the current Shiny session's memory. It
writes nothing to disk, prints nothing, logs nothing, and uses no cache. That
does not make every hosted app private. A hosting service still has process logs,
and a developer could add logging later. The boundary here is narrower: this
example app does not store or log the user's word.

## Show the UI as text

The lesson page must not contain a live Shiny form. The UI HTML is printed as
plain console output, so the browser shows text, not a working input.

```{r}
ui_html <- paste(as.character(ui), collapse = "\n")
ui_html <- paste(
  c(
    "<html lang=\"en\">",
    ui_html,
    "</html>"
  ),
  collapse = "\n"
)
cat(ui_html)
```

```{r}
#| include: false

label_for <- str_match(ui_html, "for=\"([^\"]+)\"")[, 2]
input_id <- str_match(ui_html, "<input[^>]+id=\"([^\"]+)\"")[, 2]

stopifnot(
  identical(label_for, "word"),
  identical(input_id, "word"),
  identical(label_for, input_id),
  str_detect(ui_html, fixed("Word to find")),
  str_detect(ui_html, fixed("data-update-on=\"blur\""))
)
```

The text output lets a reader inspect the generated label and input. In the
printed HTML, the label's `for="word"` matches the input's `id="word"`; that
link is how a screen reader names the box, and the placeholder `liberty` is
only a hint. It is not an invitation to type in this page.

## Test server behavior

`testServer()` runs the server function without starting a browser. The
`session$setInputs()` calls below play the part of a person typing into the app.

```{r}
status_result <- function(expr) {
  tryCatch(
    {
      list(
        status = "output",
        value = force(expr),
        condition_class = NA_character_
      )
    },
    error = function(error) {
      list(
        status = "condition",
        value = conditionMessage(error),
        condition_class = class(error)[[1]]
      )
    }
  )
}

case_results <- NULL
liberty_rows_from_server <- NULL

shiny::testServer(server, {
  session$setInputs(word = "liberty")
  liberty_status <- status_result(output$status)
  liberty_rows <- context_rows()
  liberty_rows_from_server <<- liberty_rows |>
    slice_head(n = max_rows_shown)
  stopifnot(
    identical(nrow(liberty_rows), 118L),
    identical(
      liberty_status$value,
      "Found 118 lines for 'liberty'; showing the first 8."
    )
  )

  session$setInputs(word = "")
  empty_status <- status_result(output$status)
  stopifnot(
    identical(empty_status$condition_class, "shiny.silent.error"),
    identical(empty_status$value, "")
  )

  session$setInputs(word = "   ")
  space_status <- status_result(output$status)
  stopifnot(
    identical(space_status$condition_class, "shiny.silent.error"),
    identical(space_status$value, "")
  )

  session$setInputs(word = "free people")
  two_word_status <- status_result(output$status)
  stopifnot(
    identical(
      two_word_status$value,
      "Enter one word, with no spaces."
    )
  )

  session$setInputs(word = strrep("a", 31L))
  long_status <- status_result(output$status)
  stopifnot(
    identical(
      long_status$value,
      "Enter at most 30 characters."
    )
  )

  session$setInputs(word = "*")
  star_status <- status_result(output$status)
  star_rows <- context_rows()
  stopifnot(
    identical(nrow(star_rows), 0L),
    identical(
      star_status$value,
      "No token in the punctuation-free search matches '*'."
    )
  )

  session$setInputs(word = "<b>")
  markup_status <- status_result(output$status)
  markup_rows <- context_rows()
  stopifnot(
    identical(nrow(markup_rows), 0L),
    identical(
      markup_status$value,
      "No token in the punctuation-free search matches '<b>'."
    )
  )

  session$setInputs(word = "liberty,")
  punctuation_status <- status_result(output$status)
  punctuation_rows <- context_rows()
  stopifnot(
    identical(nrow(punctuation_rows), 0L),
    identical(
      punctuation_status$value,
      "No token in the punctuation-free search matches 'liberty,'."
    )
  )

  session$setInputs(word = "zzzzzz")
  no_match_status <- status_result(output$status)
  no_match_rows <- context_rows()
  stopifnot(
    identical(nrow(no_match_rows), 0L),
    identical(
      no_match_status$value,
      "No token in the punctuation-free search matches 'zzzzzz'."
    )
  )

  case_results <<- tibble(
    case = c(
      "known word",
      "empty text",
      "three spaces",
      "two words",
      "over 30 characters",
      "literal asterisk",
      "markup-like text",
      "punctuated word",
      "no matching word"
    ),
    typed_text = c(
      "liberty",
      "\"\"",
      "\"   \"",
      "free people",
      strrep("a", 31L),
      "*",
      "<b>",
      "liberty,",
      "zzzzzz"
    ),
    rows = c(
      nrow(liberty_rows),
      NA_integer_,
      NA_integer_,
      NA_integer_,
      NA_integer_,
      nrow(star_rows),
      nrow(markup_rows),
      nrow(punctuation_rows),
      nrow(no_match_rows)
    ),
    result_type = c(
      liberty_status$status,
      empty_status$status,
      space_status$status,
      two_word_status$status,
      long_status$status,
      star_status$status,
      markup_status$status,
      punctuation_status$status,
      no_match_status$status
    ),
    condition_class = c(
      liberty_status$condition_class,
      empty_status$condition_class,
      space_status$condition_class,
      two_word_status$condition_class,
      long_status$condition_class,
      star_status$condition_class,
      markup_status$condition_class,
      punctuation_status$condition_class,
      no_match_status$condition_class
    ),
    server_result = c(
      liberty_status$value,
      empty_status$value,
      space_status$value,
      two_word_status$value,
      long_status$value,
      star_status$value,
      markup_status$value,
      punctuation_status$value,
      no_match_status$value
    )
  )
})
```

The test returns `NULL` when it passes. Its value is not a report; the checks
are the important part.

```{r}
#| include: false

stopifnot(
  identical(status_result("ok")$value, "ok"),
  identical(nrow(find_in_context("zzzzzz")), 0L),
  identical(nrow(case_results), 9L),
  identical(nrow(liberty_rows_from_server), 8L),
  identical(
    case_results$server_result[case_results$case == "known word"],
    "Found 118 lines for 'liberty'; showing the first 8."
  ),
  identical(
    case_results$server_result[case_results$case == "punctuated word"],
    "No token in the punctuation-free search matches 'liberty,'."
  )
)
```

The recorded server cases below are printed with an escaping HTML table. That
matters for hostile-looking input: markup characters should be shown as text,
not read as page markup.

```{r}
kable(
  case_results,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Case",
    "Typed text",
    "Rows",
    "Result type",
    "Condition class",
    "Server result"
  ),
  caption = "Server cases recorded by testServer during the render",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(case_results$typed_text[case_results$case == "literal asterisk"], "*"),
  identical(case_results$typed_text[case_results$case == "markup-like text"], "<b>"),
  str_detect(case_results$server_result[case_results$case == "two words"], fixed("one word")),
  str_detect(case_results$server_result[case_results$case == "over 30 characters"], fixed("30 characters"))
)
```

The table below is the first app table for `liberty`, captured from the same
server test. It shows the first eight KWIC rows even though the full result has
118 lines.

```{r}
kable(
  liberty_rows_from_server,
  col.names = c("Paragraph ID", "Before", "Keyword", "After"),
  caption = "First eight server-returned KWIC rows for liberty",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(liberty_rows_from_server), 8L),
  identical(liberty_rows_from_server$keyword, rep("liberty", 8L)),
  identical(liberty_rows_from_server, find_in_context("liberty") |> slice_head(n = 8L))
)
```

## What this page cannot test

Server tests do not exercise layout, browser JavaScript, focus order, keyboard
use, or screen-reader output. They also cannot check the live-region behavior
Shiny adds in a running browser. A real app needs a browser test and manual
accessibility review before anyone treats it as ready for public use.

To try the app locally, run the code chunks in "Build the search job first" and
"Define the app without running it" from the project root, then run
`shiny::runApp(app)`. The app will run only while that R session is serving it.

## What question does this app answer?

The app helps a reader inspect context lines. It does not say a word is
important, estimate a topic, or deploy a model. KWIC counts depend on the token
rules, punctuation handling, and case policy. Here the unit is the token stream
created from reconstructed paragraphs, and matching is fixed and
case-insensitive.

## What to remember

- A Shiny app has a UI object, a server function, and an app object.
- A static lesson page should not print a live-looking Shiny input with no
  server behind it.
- Trim text before `req()` so whitespace-only input pauses instead of searching.
- Use server-side `need()` checks for invalid values such as spaces or too many
  characters.
- Fixed matching keeps `*` and `(` from becoming pattern syntax.
- `testServer()` checks server logic, not the browser experience.

The static lesson page shows how the app is built and checked; the running app
belongs in a local R session.

## Sources

- [Mastering Shiny: Your first Shiny app](https://mastering-shiny.org/basic-app.html)
- [Mastering Shiny: Testing](https://mastering-shiny.org/scaling-testing.html)
- [Shiny reference for `req()`](https://shiny.posit.co/r/reference/shiny/latest/req.html)
- [Shiny reference for `validate()` and `need()`](https://shiny.posit.co/r/reference/shiny/latest/validate.html)
- [Shiny article on server function testing](https://shiny.posit.co/r/articles/improve/server-function-testing/)
- [Mastering Shiny: Security](https://mastering-shiny.org/scaling-security.html)
- [OWASP Logging Cheat Sheet](https://cheatsheetseries.owasp.org/cheatsheets/Logging_Cheat_Sheet.html)
- [W3C Web Content Accessibility Guidelines 2.2](https://www.w3.org/TR/WCAG22/)
- [W3C WAI forms tutorial on labels and instructions](https://www.w3.org/WAI/tutorials/forms/instructions/)
- [UK Government Digital Service accessibility tools audit](https://accessibility.blog.gov.uk/2017/02/24/what-we-found-when-we-tested-tools-on-the-worlds-least-accessible-webpage/)
- [quanteda `kwic()` reference](https://quanteda.io/reference/kwic.html)
