---
title: "Drawing labels on text spans"
subtitle: "Use offsets, escaping, and visible labels for annotated text"
description: "Learn how to draw safe inline annotations from spaCy offsets and why a table of spans must stay beside the highlights."
categories: [information visualization, annotated text, named entities]
---

Sam is reviewing short job-board and flyer sentences. A table of entity rows is
accurate, but it is hard to see which words were labeled without reading the
same sentence twice.

**Annotated text visualization** draws labels on spans of text while leaving the
source text unchanged. A **span** is a stretch of text located by character
offsets. The display can help readers inspect labels quickly, but the offsets,
escaping, and overlap rules have to be honest.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- use spaCy token offsets instead of searching for entity strings;
- convert 0-based, end-exclusive offsets for R's `substr()`;
- draw labels with htmltools tag objects that escape text;
- keep annotation layers separate when spans overlap;
- provide a captioned table with every span; and
- record the spaCy and model versions beside the display.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Start from offsets, not searched text

The spaCy pipeline returns each token's `idx`, the 0-based character position
where that token starts in the original string. The end of a span is the first
character after it. R's `substr()` starts counting at 1 and includes its end
position, so the slice uses `start + 1` and `end`.

```{r}
library(dplyr)
library(htmltools)
library(purrr)
library(readr)
library(spacyr)
library(stringr)
library(tibble)

source("R/use-spacy.R")

sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

display_sentences <- sentences |>
  filter(sentence_id %in% c("s001", "s023", "s028")) |>
  transmute(
    sentence_id,
    source_text = text
  )

pipeline <- use_project_spacy()
pipeline_info <- spacy_pipeline_version()

parsed <- spacy_parse(
  setNames(display_sentences$source_text, display_sentences$sentence_id),
  pos = TRUE,
  entity = TRUE,
  dependency = TRUE,
  additional_attributes = "idx"
) |>
  as_tibble()

entity_tokens <- parsed |>
  filter(nzchar(entity)) |>
  mutate(
    label = str_remove(entity, "_[BI]$"),
    begins_span = str_detect(entity, "_B$")
  ) |>
  group_by(doc_id) |>
  mutate(span_number = cumsum(begins_span)) |>
  ungroup()

model_spans <- entity_tokens |>
  left_join(
    display_sentences,
    by = c("doc_id" = "sentence_id")
  ) |>
  group_by(doc_id, span_number, source_text, label) |>
  summarise(
    start = min(idx),
    end = max(idx + nchar(token, type = "chars")),
    .groups = "drop"
  ) |>
  transmute(
    sentence_id = doc_id,
    layer = "spaCy model",
    start,
    end,
    text = substr(source_text, start + 1L, end),
    label,
    reviewed = "model output"
  )
```

```{r}
#| echo: false

knitr::kable(
  tibble(
    field = c("spacyr", names(pipeline_info)),
    value = c(
      as.character(packageVersion("spacyr")),
      unname(unlist(pipeline_info))
    )
  ),
  col.names = c("Field", "Value"),
  caption = "spaCy runtime and English model used for this display",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "dplyr", "htmltools", "purrr", "readr",
    "spacyr", "stringr", "tibble"
  ) %in% loadedNamespaces()),
  identical(nrow(display_sentences), 3L),
  identical(pipeline$model, "en_core_web_sm"),
  identical(pipeline_info$version, "3.8.0"),
  identical(pipeline_info$spacy, "3.8.7"),
  identical(pipeline_info$license, "MIT"),
  identical(nrow(model_spans), 3L),
  identical(model_spans$text, c("12-week", "HOUSE", "October 15")),
  identical(model_spans$label, c("DATE", "ORG", "DATE")),
  identical(model_spans$start, c(5L, 21L, 9L)),
  identical(model_spans$end, c(12L, 26L, 19L))
)
```

The model labels `HOUSE` as `ORG`. That wrong span stays visible. Lesson 27
checks the named-entity output; this lesson only shows how to draw labels
without losing the source positions.

The entity tags use `B` for the first token in a span and `I` for later tokens
inside the same span. In `s001`, the three tagged tokens start at character 5,
7, and 8; together they make the span from 5 up to, but not including, 12.

```{r}
s001_token_offsets <- parsed |>
  filter(doc_id == "s001") |>
  transmute(
    token,
    start = idx,
    end = idx + nchar(token, type = "chars"),
    entity_tag = if_else(nzchar(entity), entity, "outside")
  )
```

```{r}
#| echo: false

knitr::kable(
  s001_token_offsets,
  format = "html",
  escape = TRUE,
  col.names = c("Token", "Start", "End", "Entity tag"),
  caption = "spaCy token offsets and entity tags for sentence s001",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(s001_token_offsets$token[2:4], c("12", "-", "week")),
  identical(s001_token_offsets$start[2:4], c(5L, 7L, 8L)),
  identical(s001_token_offsets$end[2:4], c(7L, 8L, 12L)),
  identical(s001_token_offsets$entity_tag[2:4], c("DATE_B", "DATE_I", "DATE_I"))
)
```

Searching for an entity string rebuilt with spaces can fail even when the
offsets are right.

```{r}
s001_source <- display_sentences |>
  filter(sentence_id == "s001") |>
  pull(source_text)

offset_slice <- model_spans |>
  filter(sentence_id == "s001") |>
  transmute(
    search_text = "12 - week",
    search_found = !is.na(str_locate(s001_source, fixed(search_text))[, 1]),
    offset_slice = substr(s001_source, start + 1L, end)
  )
```

```{r}
#| echo: false

knitr::kable(
  offset_slice,
  format = "html",
  escape = TRUE,
  col.names = c("Searched text", "Search found", "Offset slice"),
  caption = "Offset slicing recovers the span when a searched string fails",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  isFALSE(offset_slice$search_found[[1]]),
  identical(offset_slice$offset_slice, "12-week")
)
```

## Check code-point units with an accented name

Offsets need a counting rule. spaCy's Python strings count Unicode code points,
and R's `substr()` here slices by character positions, not bytes. The sentence
below includes accented characters so the byte count differs from the character
count.

```{r}
accent_text <- "Renée Díaz runs the Tools & Dies class at Riverton Skills Centre in October."
accent_parsed <- spacy_parse(
  c(accent = accent_text),
  pos = TRUE,
  entity = TRUE,
  dependency = TRUE,
  additional_attributes = "idx"
) |>
  as_tibble()

accent_person <- accent_parsed |>
  filter(str_detect(entity, "^PERSON")) |>
  summarise(
    start = min(idx),
    end = max(idx + nchar(token, type = "chars")),
    text = substr(accent_text, start + 1L, end),
    .groups = "drop"
  )

byte_offset <- function(text, position) {
  if (position == 0L) {
    return(0L)
  }

  nchar(substr(text, 1L, position), type = "bytes")
}

person_byte_start <- byte_offset(accent_text, accent_person$start)
person_byte_end <- byte_offset(accent_text, accent_person$end)
wrong_byte_as_character_slice <- substr(
  accent_text,
  accent_person$start + 1L,
  person_byte_end
)

accent_units <- tibble(
  example = "accented constructed sentence",
  characters = nchar(accent_text, type = "chars"),
  bytes = nchar(accent_text, type = "bytes"),
  person_start = accent_person$start,
  person_end = accent_person$end,
  byte_start = person_byte_start,
  byte_end = person_byte_end,
  wrong_slice_if_bytes_are_positions = wrong_byte_as_character_slice,
  sliced_text = accent_person$text
)
```

```{r}
#| echo: false

knitr::kable(
  accent_units,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Example",
    "Characters",
    "Bytes",
    "Start",
    "End",
    "Byte start",
    "Byte end",
    "Wrong slice if bytes are used as positions",
    "Sliced text"
  ),
  caption = "Character offsets still slice the accented name correctly",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(accent_units$characters, 76L),
  identical(accent_units$bytes, 78L),
  identical(accent_units$person_start, 0L),
  identical(accent_units$person_end, 10L),
  identical(accent_units$byte_start, 0L),
  identical(accent_units$byte_end, 12L),
  identical(accent_units$wrong_slice_if_bytes_are_positions, "Renée Díaz r"),
  identical(accent_units$sliced_text, "Renée Díaz"),
  accent_units$bytes > accent_units$characters
)

riverton_span <- accent_parsed |>
  filter(token %in% c("Riverton", "Skills", "Centre")) |>
  summarise(
    start = min(idx),
    end = max(idx + nchar(token, type = "chars")),
    byte_start = byte_offset(accent_text, start),
    byte_end = byte_offset(accent_text, end),
    text = substr(accent_text, start + 1L, end),
    .groups = "drop"
  )

stopifnot(
  identical(riverton_span$text, "Riverton Skills Centre"),
  identical(riverton_span$start, 42L),
  identical(riverton_span$end, 64L),
  identical(riverton_span$byte_start, 44L),
  identical(riverton_span$byte_end, 66L)
)
```

Do not use byte offsets as character offsets. In this example, the byte offsets
would take two extra characters (`Renée Díaz r`), and every later span in the
sentence would slide to the right.

## Draw one layer only when spans do not overlap

The renderer below accepts one sentence and one layer of non-overlapping spans.
It builds the line with htmltools tags, so source text is escaped by
construction. The label appears as visible text inside each mark; the background
color is only a backup cue. The colors and borders come from two classes in the
site's stylesheet, so the code only names them.

```{r}
has_layer_overlap <- function(spans) {
  ordered <- spans |>
    arrange(start, end)

  if (nrow(ordered) < 2L) {
    return(FALSE)
  }

  any(ordered$start[-1] < ordered$end[-nrow(ordered)])
}

render_annotation_layer <- function(sentence_id, layer_name, spans, source_text) {
  if (has_layer_overlap(spans)) {
    stop("This layer has overlapping spans; show it as a table.", call. = FALSE)
  }

  ordered <- spans |>
    arrange(start, end)
  cursor <- 0L
  parts <- list(
    tags$strong(paste(layer_name, ": ", sep = ""))
  )
  plain_parts <- character()

  for (row_index in seq_len(nrow(ordered))) {
    span <- ordered[row_index, ]

    if (span$start > cursor) {
      before <- substr(source_text, cursor + 1L, span$start)
      parts <- append(parts, list(before))
      plain_parts <- c(plain_parts, before)
    }

    span_text <- substr(source_text, span$start + 1L, span$end)
    label_text <- paste("[", span$label, "]", sep = "")

    parts <- append(
      parts,
      list(
        tags$mark(
          class = "annotation-mark",
          .noWS = "outside",
          span_text,
          tags$span(class = "annotation-label", label_text)
        )
      )
    )
    plain_parts <- c(plain_parts, span_text)
    cursor <- span$end
  }

  if (cursor < nchar(source_text, type = "chars")) {
    after <- substr(source_text, cursor + 1L, nchar(source_text, type = "chars"))
    parts <- append(parts, list(after))
    plain_parts <- c(plain_parts, after)
  }

  list(
    sentence_id = sentence_id,
    layer = layer_name,
    plain_text = paste(plain_parts, collapse = ""),
    html = tags$p(
      class = "annotated-text-example",
      do.call(tagList, parts)
    )
  )
}

model_lines <- display_sentences |>
  mutate(
    spans = map(
      sentence_id,
      \(current_id) {
        model_spans |>
          filter(sentence_id == current_id, layer == "spaCy model")
      }
    ),
    rendered = pmap(
      list(sentence_id, spans, source_text),
      \(sentence_id, spans, source_text) {
        render_annotation_layer(
          sentence_id = sentence_id,
          layer_name = "spaCy model",
          spans = spans,
          source_text = source_text
        )
      }
    )
  )

tagList(map(model_lines$rendered, "html"))
```

```{r}
#| include: false

plain_model_text <- map_chr(model_lines$rendered, "plain_text")

stopifnot(
  identical(plain_model_text, display_sentences$source_text),
  all(map_lgl(model_lines$spans, \(spans) !has_layer_overlap(spans))),
  all(str_detect(
    as.character(tagList(map(model_lines$rendered, "html"))),
    fixed("<mark")
  )),
  all(str_detect(
    as.character(tagList(map(model_lines$rendered, "html"))),
    fixed("[")
  ))
)
```

The marks are useful because the original words stay in order. The table below
is still necessary: it gives the offsets and labels in a form that does not
depend on color or inline layout.

## Keep a table of every span

Sam also wants to show a reviewer layer for a constructed nested example:
`Riverton` can be a `GPE`, spaCy's label for a geopolitical place such as a
town, inside `Riverton Skills Centre`, an organisation name. The two annotations
are kept in separate layers so each line can be drawn without overlap.

```{r}
review_sentence <- tibble(
  sentence_id = "constructed-01",
  source_text = "Riverton Skills Centre will run evening classes in October."
)

review_spans <- tibble(
  sentence_id = c("constructed-01", "constructed-01"),
  layer = c("reviewer organisation", "reviewer place"),
  start = c(0L, 0L),
  end = c(22L, 8L),
  text = c("Riverton Skills Centre", "Riverton"),
  label = c("ORG", "GPE"),
  reviewed = c("author-written layer", "author-written layer")
)

all_visible_spans <- bind_rows(
  model_spans,
  review_spans
)
```

```{r}
#| echo: false

knitr::kable(
  all_visible_spans |>
    arrange(sentence_id, layer, start, end),
  format = "html",
  escape = TRUE,
  col.names = c(
    "Sentence ID",
    "Layer",
    "Start",
    "End",
    "Text",
    "Label",
    "Review status"
  ),
  caption = "Every span used in the inline annotated-text displays",
  row.names = FALSE
)
```

```{r}
#| include: false

span_sources <- bind_rows(display_sentences, review_sentence)

span_slices <- all_visible_spans |>
  left_join(span_sources, by = "sentence_id") |>
  mutate(sliced = substr(source_text, start + 1L, end))

stopifnot(
  identical(nrow(all_visible_spans), 5L),
  identical(span_slices$text, span_slices$sliced),
  all(c("sentence_id", "layer", "start", "end", "label") %in%
    names(all_visible_spans)),
  identical(review_spans$text, c("Riverton Skills Centre", "Riverton"))
)
```

Now each reviewer layer can be drawn on its own line.

```{r}
review_lines <- review_spans |>
  group_by(layer) |>
  group_split() |>
  map(
    \(layer_spans) {
      render_annotation_layer(
        sentence_id = "constructed-01",
        layer_name = layer_spans$layer[[1]],
        spans = layer_spans,
        source_text = review_sentence$source_text[[1]]
      )
    }
  )

tagList(map(review_lines, "html"))
```

```{r}
#| include: false

stopifnot(
  identical(map_chr(review_lines, "plain_text"), rep(review_sentence$source_text, 2L)),
  all(!map_lgl(
    split(review_spans, review_spans$layer),
    has_layer_overlap
  ))
)
```

The two reviewer labels are not a spaCy result. They are an author-written
example showing why layers matter.

## Show overlap instead of forcing it inline

If those two reviewer spans are put into one layer, they overlap. This lesson
does not try to squeeze them into one inline row. It switches to the table view
for that layer.

```{r}
overlap_spans <- review_spans |>
  mutate(layer = "combined reviewer layer")

overlap_report <- overlap_spans |>
  mutate(
    inline_display = if_else(
      has_layer_overlap(overlap_spans),
      "table view only: spans overlap within the layer",
      "safe for inline display"
    )
  )
```

```{r}
#| echo: false

knitr::kable(
  overlap_report,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Sentence ID",
    "Layer",
    "Start",
    "End",
    "Text",
    "Label",
    "Review status",
    "Display decision"
  ),
  caption = "Overlapping spans are kept in a table instead of one inline layer",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  has_layer_overlap(overlap_spans),
  all(str_detect(overlap_report$inline_display, fixed("table view only")))
)

overlap_error <- tryCatch(
  render_annotation_layer(
    sentence_id = "constructed-01",
    layer_name = "combined reviewer layer",
    spans = overlap_spans,
    source_text = review_sentence$source_text[[1]]
  ),
  error = conditionMessage
)

stopifnot(
  identical(overlap_error, "This layer has overlapping spans; show it as a table.")
)
```

Nested spans can be drawn by some specialized tools. Crossing spans are harder
because one tag cannot contain the other. For a first lesson, the safer rule is
simple: draw one non-overlapping layer at a time and keep the complete table.

## Escape source text by construction

The renderer treats source text as text. The example below contains `<b>` and
`&`. They should appear as characters, not as an HTML tag or entity.

```{r}
escape_text <- "Sam wrote <b> & kept it as text."
escape_span <- tibble(
  sentence_id = "constructed-escape",
  layer = "escaping check",
  start = 10L,
  end = 13L,
  text = "<b>",
  label = "TEXT",
  reviewed = "constructed hostile text"
)

escape_line <- render_annotation_layer(
  sentence_id = "constructed-escape",
  layer_name = "escaping check",
  spans = escape_span,
  source_text = escape_text
)

escape_line$html
```

```{r}
#| include: false

escaped_html <- as.character(escape_line$html)

stopifnot(
  identical(substr(escape_text, escape_span$start + 1L, escape_span$end), "<b>"),
  str_detect(escaped_html, fixed("&lt;b&gt;")),
  !str_detect(escaped_html, fixed("<b>")),
  str_detect(escaped_html, fixed("&amp;")),
  identical(escape_line$plain_text, escape_text)
)
```

The same rule applies to tables that show source text. Use an escaping HTML
table so text such as `<b>` remains visible.

```{r}
escape_table <- escape_span |>
  select(sentence_id, layer, start, end, text, label, reviewed)
```

```{r}
#| echo: false

knitr::kable(
  escape_table,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Sentence ID",
    "Layer",
    "Start",
    "End",
    "Text",
    "Label",
    "Review status"
  ),
  caption = "Markup-like source text preserved as table text",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(escape_table$text, "<b>"),
  identical(escape_table$label, "TEXT")
)
```

```{r}
#| include: false

invisible(suppressMessages(spacy_finalize()))
spacy_closed <- TRUE

stopifnot(spacy_closed)
```

## What this display does not prove

The highlights show where the model or reviewer layer placed labels. They do not
prove that the labels are correct. The `HOUSE` example is deliberately left in
place because it is a model-output error from the earlier NER lesson.

The page also does not certify accessibility. The visible label text inside each
mark keeps color from carrying the meaning alone, and the contrast is designed
for the automated scan. Screen-reader behavior and forced-colors behavior still
need manual review.

## What to remember

- An annotated text visualization draws labels on spans while keeping source
  text unchanged.
- spaCy's `idx` gives 0-based token starts; R's `substr()` needs `start + 1`.
- Slice spans from offsets instead of searching for joined entity strings.
- Build inline HTML with tag objects so source text is escaped as text.
- Put visible labels inside the marks; do not rely on color alone.
- Draw one non-overlapping layer at a time and keep a captioned span table.

Sam can now read the highlights and still trace every label back to a sentence
ID, a layer, and an offset pair.

## Sources

- [brat rapid annotation tool paper](https://aclanthology.org/E12-2021/)
- [brat standoff format](https://brat.nlplab.org/standoff.html)
- [W3C Web Annotation Data Model: Text Quote Selector](https://www.w3.org/TR/annotation-model/#text-quote-selector)
- [W3C Web Annotation Data Model: Text Position Selector](https://www.w3.org/TR/annotation-model/#text-position-selector)
- [Finkel and Manning on nested named entity recognition](https://aclanthology.org/D09-1015/)
- [spaCy visualizers](https://spacy.io/usage/visualizers)
- [spaCy token API](https://spacy.io/api/token#attributes)
- [spaCy document API](https://spacy.io/api/doc#char_span)
- [spacyr `spacy_parse()` reference](https://spacyr.quanteda.io/reference/spacy_parse.html)
- [htmltools `htmlEscape()` reference](https://rstudio.github.io/htmltools/reference/htmlEscape.html)
- [OWASP Cross Site Scripting Prevention Cheat Sheet](https://cheatsheetseries.owasp.org/cheatsheets/Cross_Site_Scripting_Prevention_Cheat_Sheet.html)
- [W3C WCAG 2.2 use of color](https://www.w3.org/TR/WCAG22/#use-of-color)
- [Understanding WCAG 2.2 use of color](https://www.w3.org/WAI/WCAG22/Understanding/use-of-color.html)
- [Neves and Seva review of manual annotation tools](https://doi.org/10.1093/bib/bbz130)
