---
title: "Reading a word cloud carefully"
subtitle: "Compare the picture with the count table that made it"
description: "Learn how word clouds encode word counts, why the layout can mislead, and how to check the same words in a sorted bar chart and table."
categories: [information visualization, word clouds, inaugural addresses]
---

Twenty-two words can fill Iris's slide about the inaugural addresses. The table
behind the picture matters more than the shape of the picture itself.

The tempting part is also the risk. A large word can mean a high count, a long
spelling, or a lucky position in the packing algorithm.

**Word clouds** draw words from a count table with larger type for larger
counts. They are useful for a quick look at a vocabulary. They are weak for
exact comparisons, rank order, meaning, and change over time.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- build one corpus-wide word-count table from inaugural paragraphs;
- use a named stop-word list and check what it removes;
- draw a sorted bar chart and a word cloud from the same counts;
- read size, position, and colour in a cloud honestly;
- test the word-length trap with real words from the data; and
- give the cloud a two-part text alternative.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build one count table

The corpus helper returns one row per paragraph. This lesson asks one
corpus-wide question: after a stated cleanup, which words occur most often
across all inaugural paragraphs? Pooling every paragraph gives long speeches
more weight than short ones, so the result is a preview of the vocabulary, not
a claim about importance.

The figures use the site's shared style, `theme_lesson()`, from
`R/lesson-figures.R`. That keeps the plotting code below focused on the choices
that belong to each chart.

```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(tidytext)
library(ggplot2)
library(ggwordcloud)
library(purrr)

source("R/inaugural-corpus.R")
source("R/lesson-figures.R")

paragraphs <- inaugural_paragraphs()

all_word_tokens <- paragraphs |>
  select(paragraph_id, speech_id, paragraph) |>
  unnest_tokens(word, paragraph)

dropped_by_letters_filter <- all_word_tokens |>
  filter(!str_detect(word, "^[a-z]+$"))

word_tokens <- all_word_tokens |>
  filter(str_detect(word, "^[a-z]+$"))

onix_stop_words <- stop_words |>
  filter(lexicon == "onix") |>
  distinct(word)

inaugural_word_counts <- word_tokens |>
  anti_join(onix_stop_words, by = join_by(word)) |>
  count(word, sort = TRUE, name = "count") |>
  mutate(letters = str_length(word))

cloud_words <- inaugural_word_counts |>
  slice_head(n = 22)
```

```{r}
#| echo: false

knitr::kable(
  cloud_words,
  col.names = c("Word", "Count", "Letters"),
  caption = "Top words after removing the onix stop-word list from inaugural paragraphs",
  row.names = FALSE
)
```

```{r}
#| include: false

top_five <- cloud_words |>
  slice_head(n = 5)

bar_alt_text <- paste0(
  "Horizontal bar chart of the 22 most common words after onix stop-word ",
  "removal. The five longest bars are ",
  paste0(top_five$word, " (", top_five$count, ")", collapse = ", "),
  ". The bar for so is highlighted because onix does not list so as a stop word."
)

cloud_alt_seed_7701 <- paste(
  "Word cloud of the same 22 words from the count table.",
  "The largest words are government, used 585 times, and people, used 561 times;",
  "the positions of the words are not data.",
  "The count table above is the long text alternative."
)

stopifnot(
  identical(nrow(paragraphs), 1377L),
  "america's" %in% dropped_by_letters_filter$word,
  n_distinct(word_tokens$paragraph_id) == 1377L,
  "states" %in% onix_stop_words$word,
  !"united" %in% onix_stop_words$word,
  !"so" %in% onix_stop_words$word,
  identical(nrow(cloud_words), 22L),
  !anyDuplicated(cloud_words$word),
  identical(top_five$word, c("government", "people", "so", "country", "world")),
  identical(top_five$count, c(585L, 561L, 384L, 306L, 305L)),
  top_five$count[2] - top_five$count[3] > 150L,
  identical(min(cloud_words$count), 157L),
  identical(
    str_match_all(bar_alt_text, "([a-z]+) \\(([0-9]+)\\)")[[1]][, 2],
    top_five$word
  ),
  identical(
    as.integer(str_match_all(bar_alt_text, "([a-z]+) \\(([0-9]+)\\)")[[1]][, 3]),
    top_five$count
  )
)
```

The onix list removes `states` but keeps `united`, so a phrase such as
`United States` can leave `united` standing alone. A different stop-word list
would produce a different table.

`unnest_tokens()` lowercases the paragraphs and splits punctuation away from
words, except an apostrophe inside a word. The `^[a-z]+$` filter then keeps
only tokens made entirely of letters, so it drops whole tokens such as
`america's`; in this corpus it drops `r nrow(dropped_by_letters_filter)`
tokens. After that filter, `so` survives
`r inaugural_word_counts$count[inaugural_word_counts$word == "so"]` times
because onix does not list it as a stop word.

## Show the counts before the cloud

A sorted bar chart puts the same words on a common scale. People compare
positions along a common scale more accurately than areas, so the chart is the
check Iris should read before the cloud. Cleveland and McGill placed area below
position in their hypothesized ranking of graphical perception tasks; they did
not test word clouds themselves.

Each bar carries its count as a label, so the chart needs no numeric axis. The
one highlighted bar is the stop-word decision the reader should notice.

```{r}
#| label: fig-word-cloud-bars
#| fig-cap: "The 22 words used in the word cloud, sorted by count after onix stop-word removal."
#| fig-alt: !expr bar_alt_text
#| fig-width: 7
#| fig-height: 5.6

bar_data <- cloud_words |>
  mutate(
    kept_by_onix = word == "so",
    bar_label = if_else(
      kept_by_onix,
      str_c(count, ", not an onix stop word"),
      as.character(count)
    )
  )

bar_plot <- ggplot(bar_data, aes(x = count, y = reorder(word, count))) +
  geom_col(aes(fill = kept_by_onix), width = 0.72, show.legend = FALSE) +
  geom_text(
    aes(label = bar_label),
    hjust = 0,
    nudge_x = 6,
    family = lesson_font,
    size = 9.5,
    size.unit = "pt",
    colour = lesson_colours[["body"]]
  ) +
  scale_fill_manual(
    values = c(
      `FALSE` = lesson_colours[["accent"]],
      `TRUE` = lesson_colours[["contrast"]]
    )
  ) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.32))) +
  labs(
    title = "Two words lead the count by a wide margin",
    subtitle = "Uses in 1,377 inaugural paragraphs after removing onix stop words",
    x = NULL,
    y = NULL
  ) +
  theme_lesson(grid = "none") +
  theme(axis.text.x = element_blank())

bar_plot
```

```{r}
#| include: false

extract_bar_layer <- function(plot) {
  built <- ggplot_build(plot)
  bars <- built$data[[1]]
  axis_labels <- built$layout$panel_params[[1]]$y$get_labels()
  tibble(
    word = axis_labels[bars$y],
    built_count = bars$x
  )
}

bar_layer_check <- extract_bar_layer(bar_plot) |>
  left_join(cloud_words, by = join_by(word))

stopifnot(
  identical(nrow(bar_layer_check), 22L),
  all(bar_layer_check$built_count == bar_layer_check$count),
  setequal(bar_layer_check$word, cloud_words$word),
  identical(
    bar_layer_check |>
      arrange(desc(built_count), word) |>
      slice_head(n = 5) |>
      pull(word),
    top_five$word
  ),
  identical(sum(bar_data$kept_by_onix), 1L)
)
```

The bar chart and the table use the same `word` and `count` values. Nothing
about a word's position in the cloud below is evidence of a relationship
between words.

## Draw the same counts as a cloud

The cloud below uses `geom_text_wordcloud()` with one colour and no rotation.
`scale_size_area()` sets the type size so that the area of each letter grows in
step with the count: a word used four times as often gets letters twice as
tall. The scale does nothing about word length, so a long word with a modest
count can still cover more of the slide than a short word used more often.

```{r}
#| label: fig-word-cloud-seed-one
#| fig-cap: "A word cloud drawn from the same 22 counts as the bar chart, using layout seed 7701."
#| fig-alt: !expr cloud_alt_seed_7701
#| fig-width: 7
#| fig-height: 4.4
#| dev: png

draw_cloud <- function(seed) {
  ggplot(cloud_words, aes(label = word, size = count)) +
    # ggwordcloud places each word by first measuring it on a grDevices::png()
    # device. These figures are drawn on that same device, with its default
    # font, so each word is drawn exactly as wide as it was measured.
    geom_text_wordcloud(
      seed = seed,
      colour = lesson_colours[["accent"]],
      angle = 0,
      grid_margin = 2,
      rm_outside = FALSE
    ) +
    scale_size_area(max_size = 16, guide = "none") +
    theme_lesson(grid = "none", base_family = "sans")
}

cloud_plot_seed_7701 <- draw_cloud(seed = 7701) +
  labs(
    title = "The first word drawn takes the centre",
    subtitle = "Type size follows the count. Position, neighbours, and colour carry no meaning."
  )

cloud_plot_seed_7701
```

```{r}
#| include: false

extract_cloud_layer <- function(plot) {
  ggplot_build(plot)$data[[1]] |>
    transmute(
      word = as.character(label),
      plotted_size = size
    ) |>
    left_join(cloud_words, by = join_by(word))
}

cloud_layer_seed_7701 <- extract_cloud_layer(cloud_plot_seed_7701)
cloud_size_ratio_seed_7701 <- cloud_layer_seed_7701$plotted_size^2 /
  cloud_layer_seed_7701$count

stopifnot(
  identical(nrow(cloud_layer_seed_7701), nrow(cloud_words)),
  !anyNA(cloud_layer_seed_7701$count),
  setequal(cloud_layer_seed_7701$word, cloud_words$word),
  max(cloud_size_ratio_seed_7701) - min(cloud_size_ratio_seed_7701) < 1e-10,
  identical(cloud_words$word[1], "government"),
  identical(packageVersion("ggwordcloud"), package_version("0.6.2"))
)
```

The word in the middle is the first row drawn, not the topic of the speeches.
The table is sorted by count, so `government` is drawn first. If the table were
alphabetical, a different word could start in the centre. The packing also
depends on the seed, the drawing device, and the fonts, so the page makes no
claim about where any word should sit.

`area_corr`, an option in ggwordcloud, makes a different trade. It resizes
labels so each word's whole text box follows the count, which removes the
length bias but gives short words taller type. Felix, Franconeri, and Bertini
later found that people judged values more accurately from bar length than
from font size.

The short part of the text alternative is the cloud's alt text:
"`r cloud_alt_seed_7701`" The long part is the sorted count table above, which
lists every word and count used in the cloud.

## Change the seed, not the data

The next cloud calls the same function with a different layout seed. The
counts are unchanged.

```{r}
#| label: fig-word-cloud-seed-two
#| fig-cap: "The same 22 count-table words drawn again with layout seed 7702."
#| fig-alt: "Second word cloud with the same words and counts as the first cloud but a different layout seed. Government is still used 585 times and people 561 times; every change in position is a layout artefact."
#| fig-width: 7
#| fig-height: 4.4
#| dev: png

cloud_plot_seed_7702 <- draw_cloud(seed = 7702) +
  labs(
    title = "A new seed moves every word; the counts stay the same",
    subtitle = "Same 22 words and counts as the previous cloud, layout seed 7702"
  )

cloud_plot_seed_7702
```

```{r}
#| include: false

cloud_layer_seed_7702 <- extract_cloud_layer(cloud_plot_seed_7702)
cloud_size_ratio_seed_7702 <- cloud_layer_seed_7702$plotted_size^2 /
  cloud_layer_seed_7702$count

stopifnot(
  identical(nrow(cloud_layer_seed_7702), nrow(cloud_words)),
  !anyNA(cloud_layer_seed_7702$count),
  setequal(cloud_layer_seed_7702$word, cloud_layer_seed_7701$word),
  max(cloud_size_ratio_seed_7702) - min(cloud_size_ratio_seed_7702) < 1e-10,
  identical(sum(cloud_layer_seed_7702$count), sum(cloud_words$count))
)
```

If the two pictures feel different, that feeling is the warning. The counts did
not change.

## Check preprocessing sensitivity

Stop-word lists are editorial choices. The table below recomputes the top ten
words under three named choices: onix only, snowball only, and all three
tidytext lists together.

```{r}
stopword_choices <- list(
  "onix only" = stop_words |>
    filter(lexicon == "onix") |>
    distinct(word),
  "snowball only" = stop_words |>
    filter(lexicon == "snowball") |>
    distinct(word),
  "all tidytext lexicons" = stop_words |>
    distinct(word)
)

sensitivity_table <- imap(
  stopword_choices,
  \(stop_table, stop_choice) {
    word_tokens |>
      anti_join(stop_table, by = join_by(word)) |>
      count(word, sort = TRUE, name = "count") |>
      slice_head(n = 10) |>
      mutate(stop_choice = stop_choice, rank = row_number(), .before = 1)
  }
) |>
  list_rbind()
```

```{r}
#| echo: false

sensitivity_table |>
  mutate(entry = str_c(word, " (", count, ")")) |>
  select(rank, stop_choice, entry) |>
  pivot_wider(names_from = stop_choice, values_from = entry) |>
  knitr::kable(
    col.names = c("Rank", "Onix only", "Snowball only", "All tidytext lexicons"),
    caption = "Top ten words, with counts, under three stop-word choices",
    row.names = FALSE
  )
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sensitivity_table), 30L),
  identical(
    sensitivity_table$word[sensitivity_table$stop_choice == "onix only"][1:2],
    c("government", "people")
  ),
  "states" %in% sensitivity_table$word[
    sensitivity_table$stop_choice == "snowball only"
  ],
  !"states" %in% sensitivity_table$word[
    sensitivity_table$stop_choice == "onix only"
  ],
  "united" %in% inaugural_word_counts$word
)
```

With snowball only, common words such as `will`, `us`, and `states` remain.
With onix, `states` is removed while `united` remains. A cloud without this
preprocessing note would look more objective than it is.

## Check the word-length trap

The next table compares real words from the count table. The rough space score
is `letters * sqrt(count)`. It is not a model of perception; it only shows why
a long word can take more horizontal room even when its count is lower.
Alexander and coauthors tested word-length effects in word clouds directly.

```{r}
length_pairs <- inaugural_word_counts |>
  filter(word %in% c("constitution", "time", "peace", "government", "people")) |>
  mutate(rough_horizontal_space = round(letters * sqrt(count), 1)) |>
  arrange(desc(rough_horizontal_space))
```

```{r}
#| echo: false

knitr::kable(
  length_pairs,
  col.names = c("Word", "Count", "Letters", "Rough space score"),
  caption = "Long words can cover more space than shorter words with higher counts",
  row.names = FALSE
)
```

```{r}
#| include: false

constitution_space <- length_pairs |>
  filter(word == "constitution") |>
  pull(rough_horizontal_space)
time_space <- length_pairs |>
  filter(word == "time") |>
  pull(rough_horizontal_space)
peace_space <- length_pairs |>
  filter(word == "peace") |>
  pull(rough_horizontal_space)

stopifnot(
  identical(length_pairs$count[length_pairs$word == "constitution"], 207L),
  identical(length_pairs$count[length_pairs$word == "time"], 214L),
  constitution_space > time_space,
  constitution_space > peace_space,
  length_pairs$letters[length_pairs$word == "constitution"] == 12L,
  length_pairs$letters[length_pairs$word == "time"] == 4L
)
```

`constitution` appears 207 times and `time` appears 214 times. The longer word
can still take about `r round(constitution_space / time_space, 1)` times the
horizontal space under this simple geometry check. That is why Iris should not
ask readers to compare close counts from font size.

## What this cloud does not show

The cloud does not show word order, phrases, relationships, distinctiveness,
change over time, or importance. It also hides that `united` and `states` were
once part of a phrase before tokenizing and the stop-word list split them.

Use a cloud as a doorway to the table, not as the table's replacement.

## What to remember

- A word cloud starts from a count table; keep that table visible.
- The onix stop-word list removes `states` but keeps `united` in this corpus.
- The bar chart and both clouds here use the same 22 counts.
- Placement, neighbours, the centre, and colour in a cloud carry no meaning.
- Long words can look stronger than shorter words with similar counts.
- The sorted counts are the long text alternative for the cloud.

Iris keeps the cloud, but the slide title points to the count table: "Common
words after one stated cleanup."

## Sources

- [ggwordcloud vignette](https://lepennec.github.io/ggwordcloud/articles/ggwordcloud.html)
- [tidytext stop-word documentation](https://juliasilge.github.io/tidytext/reference/stop_words.html)
- [Cleveland and McGill on graphical perception](https://doi.org/10.1080/01621459.1984.10478080)
- [Heer and Bostock on crowdsourced graphical perception](https://doi.org/10.1145/1753326.1753357)
- [Alexander et al. on word-length effects in word clouds](https://doi.org/10.1109/TVCG.2017.2723397)
- [Felix, Franconeri, and Bertini on word-cloud magnitude judgments](https://doi.org/10.1109/TVCG.2017.2746018)
- [Hearst and Rosner on tag-cloud limits](https://doi.org/10.1109/HICSS.2008.422)
- [W3C guidance on complex image text alternatives](https://www.w3.org/WAI/tutorials/images/complex/)
