---
title: "Reading a word cloud carefully"
subtitle: "Compare the picture with the count table that made it"
description: "Learn how word clouds encode word counts, why the layout can mislead, and how to check the same words in a sorted table."
categories: [information visualization, word clouds, inaugural addresses]
---

Twenty-two words can fill Iris's slide about the inaugural addresses. The table
behind the picture matters more than the shape of the picture itself.

The tempting part is also the risk. A large word can mean a high count, a long
spelling, or a lucky position in the packing algorithm.

**Word clouds** draw words from a count table with larger type for larger
counts. They are useful for a quick vocabulary preview. They are weak for exact
comparisons, rank order, meaning, and change over time.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- build one corpus-wide word-count table from inaugural paragraphs;
- use a named stop-word lexicon and inspect what it removes;
- draw a sorted bar chart and a word cloud from the same keyed counts;
- keep word-cloud size, layout, and colour claims honest;
- test a word-length trap with real words from the data; and
- provide a two-part text alternative for the cloud.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build one count table

The corpus helper returns one row per paragraph. This lesson asks one
corpus-wide question: after a stated cleanup, which words occur most often
across all inaugural paragraphs? That flattening weights long speeches more
than short speeches, so the output is a vocabulary preview, not a claim about
importance.

```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(tidytext)
library(ggplot2)
library(ggwordcloud)
library(purrr)
library(knitr)

source("R/inaugural-corpus.R")

paragraphs <- inaugural_paragraphs()

all_word_tokens <- paragraphs |>
  select(paragraph_id, speech_id, paragraph) |>
  unnest_tokens(word, paragraph)

dropped_by_letters_filter <- all_word_tokens |>
  filter(!str_detect(word, "^[a-z]+$"))

word_tokens <- all_word_tokens |>
  filter(str_detect(word, "^[a-z]+$"))

onix_stop_words <- stop_words |>
  filter(lexicon == "onix") |>
  distinct(word)

inaugural_word_counts <- word_tokens |>
  anti_join(onix_stop_words, by = join_by(word)) |>
  count(word, sort = TRUE, name = "count") |>
  mutate(
    word_key = word,
    letters = str_length(word)
  )

cloud_words <- inaugural_word_counts |>
  slice_head(n = 22) |>
  mutate(word_for_plot = reorder(word_key, count))

count_table <- cloud_words |>
  transmute(
    word_key,
    count,
    letters
  )

top_five_bars <- count_table |>
  slice_head(n = 5)

bar_alt_text <- sprintf(
  paste(
    "Horizontal bar chart of 22 words.",
    "The first five bars are %s (%s), %s (%s), %s (%s), %s (%s), and %s (%s).",
    "Every bar comes from the same count table used for the cloud."
  ),
  top_five_bars$word_key[1],
  top_five_bars$count[1],
  top_five_bars$word_key[2],
  top_five_bars$count[2],
  top_five_bars$word_key[3],
  top_five_bars$count[3],
  top_five_bars$word_key[4],
  top_five_bars$count[4],
  top_five_bars$word_key[5],
  top_five_bars$count[5]
)

cloud_alt_seed_7701 <- paste(
  "Word cloud of the same 22 count-table words.",
  "The largest words by count are government with 585 and people with 561;",
  "the exact positions are not data.",
  "The full count table above is the long text alternative."
)

stopword_note <- tibble(
  lexicon = "onix",
  removes_states = "states" %in% onix_stop_words$word,
  removes_united = "united" %in% onix_stop_words$word
)

kable(
  count_table,
  col.names = c("Word key", "Count", "Letters"),
  caption = "Top words after removing the onix stop-word lexicon from inaugural paragraphs",
  row.names = FALSE
)

kable(
  stopword_note,
  col.names = c("Stop-word lexicon", "Removes `states`", "Removes `united`"),
  caption = "The named stop-word choice used for the count table",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(paragraphs), 1377L),
  identical(nrow(dropped_by_letters_filter), 369L),
  "america's" %in% dropped_by_letters_filter$word,
  n_distinct(word_tokens$paragraph_id) == 1377L,
  identical(stopword_note$removes_states, TRUE),
  identical(stopword_note$removes_united, FALSE),
  identical(cloud_words$word_key[1], "government"),
  identical(cloud_words$count[1], 585L),
  identical(cloud_words$word_key[2], "people"),
  identical(cloud_words$count[2], 561L),
  identical(top_five_bars$word_key, c("government", "people", "so", "country", "world")),
  identical(top_five_bars$count, c(585L, 561L, 384L, 306L, 305L)),
  str_detect(bar_alt_text, fixed("government (585)")),
  str_detect(bar_alt_text, fixed("people (561)")),
  str_detect(bar_alt_text, fixed("so (384)")),
  str_detect(bar_alt_text, fixed("country (306)")),
  str_detect(bar_alt_text, fixed("world (305)")),
  identical(
    str_match_all(bar_alt_text, "([a-z]+) \\(([0-9]+)\\)")[[1]][, 2],
    top_five_bars$word_key
  ),
  identical(
    as.integer(str_match_all(bar_alt_text, "([a-z]+) \\(([0-9]+)\\)")[[1]][, 3]),
    top_five_bars$count
  ),
  identical(nrow(count_table), 22L),
  !anyDuplicated(count_table$word_key)
)
```

The onix lexicon removes `states` but keeps `united`, so a phrase such as
`United States` can leave `united` by itself. A different stop-word list would
produce a different table.

`unnest_tokens()` lowercases the paragraphs and splits punctuation away from
words. The `^[a-z]+$` filter keeps only tokens made entirely of letters, so it
drops whole tokens such as `america's`; in this corpus it drops
`r nrow(dropped_by_letters_filter)` tokens. After that filter and the onix
stop-word list, `so` survives
`r inaugural_word_counts$count[inaugural_word_counts$word_key == "so"]` times
because `so` is not in the onix list.

## Show the counts before the cloud

A sorted bar chart puts the same words on a common horizontal scale. Position
along a common scale is easier to compare than text area, so the chart is the
check Iris should read before the cloud. Cleveland and McGill placed area below
position in their hypothesized ranking of graphical perception tasks; they did
not run the word-cloud experiment on this page.

```{r}
#| label: fig-word-cloud-bars
#| fig-cap: "The 22 words used in the word cloud, sorted by count after onix stop-word removal."
#| fig-alt: !expr bar_alt_text
#| fig-width: 7
#| fig-height: 5

bar_plot <- ggplot(cloud_words, aes(x = count, y = word_for_plot)) +
  geom_col(fill = "#3B5F8A") +
  labs(
    x = "Count in inaugural paragraphs",
    y = NULL
  ) +
  theme_minimal()

bar_plot
```

```{r}
#| include: false

extract_bar_layer <- function(plot, expected_counts) {
  built <- ggplot_build(plot)
  layer <- built$data[[1]]
  axis_labels <- built$layout$panel_params[[1]]$y$get_labels()
  tibble(
    word_key = axis_labels[layer$y],
    built_count = layer$x
  ) |>
    left_join(
      expected_counts |>
        select(word_key, count),
      by = join_by(word_key)
    )
}

bar_check <- cloud_words |>
  select(word_key, count) |>
  left_join(
    count_table |>
      select(word_key, count),
    by = join_by(word_key),
    suffix = c("_plot", "_table")
  )
bar_layer_check <- extract_bar_layer(bar_plot, count_table)

stopifnot(
  identical(nrow(bar_check), 22L),
  identical(bar_check$count_plot, bar_check$count_table),
  all(bar_layer_check$built_count == bar_layer_check$count),
  setequal(bar_layer_check$word_key, count_table$word_key),
  identical(
    bar_layer_check |>
      arrange(desc(built_count), word_key) |>
      slice_head(n = 5) |>
      pull(word_key),
    top_five_bars$word_key
  ),
  identical(max(cloud_words$count), 585L),
  identical(min(cloud_words$count), 157L)
)
```

The bar chart and the table use the same `word_key` and `count` values. No
position in the later cloud is evidence of a relationship between words.

## Draw the same counts as a cloud

The cloud below uses `geom_text_wordcloud()` with one colour and no rotation.
`scale_size_area()` maps the numeric count through an area scale for the size
aesthetic: a count of zero would map to zero, and the largest count maps to the
largest font size in this plot. Font size grows with the square root of the
count, so each letter's area follows the count, and a word's whole footprint
also grows with its number of letters. It does not divide by word length. A
longer word can still cover more horizontal space than a shorter word with a
similar or even larger count.

```{r}
#| label: fig-word-cloud-seed-one
#| fig-cap: "A word cloud drawn from the same 22 counts as the bar chart, using seed 7701."
#| fig-alt: !expr cloud_alt_seed_7701
#| fig-width: 7
#| fig-height: 4.5

set.seed(7701)
cloud_plot_seed_7701 <- ggplot(
  cloud_words,
  aes(label = word_key, size = count)
) +
  geom_text_wordcloud(
    seed = 7701,
    color = "#3B5F8A",
    angle = 0,
    rm_outside = FALSE
  ) +
  scale_size_area(max_size = 9) +
  labs(size = "Count") +
  theme_minimal()

cloud_plot_seed_7701
```

```{r}
#| include: false

extract_cloud_layer <- function(plot, expected_counts) {
  ggplot_build(plot)$data[[1]] |>
    transmute(
      word_key = as.character(label),
      plotted_size = size
    ) |>
    left_join(
      expected_counts |>
        select(word_key, count),
      by = join_by(word_key)
    )
}

cloud_layer_seed_7701 <- extract_cloud_layer(
  cloud_plot_seed_7701,
  count_table
)
cloud_size_ratio_seed_7701 <- cloud_layer_seed_7701$plotted_size^2 /
  cloud_layer_seed_7701$count
cloud_input_seed_7701 <- cloud_words |>
  arrange(desc(count), word_key) |>
  select(word_key, count)

stopifnot(
  identical(cloud_input_seed_7701, count_table |> select(word_key, count)),
  identical(nrow(cloud_layer_seed_7701), nrow(count_table)),
  !anyNA(cloud_layer_seed_7701$count),
  setequal(cloud_layer_seed_7701$word_key, count_table$word_key),
  max(cloud_size_ratio_seed_7701) - min(cloud_size_ratio_seed_7701) < 1e-10,
  identical(cloud_input_seed_7701$word_key[1], "government"),
  identical(cloud_input_seed_7701$count[1], 585L),
  identical(packageVersion("ggwordcloud"), package_version("0.6.2"))
)
```

The middle word is the first row drawn, not the topic of the speeches. Because
the table is sorted by count, `government` is drawn first. If the table were
alphabetical, a different word could start in the centre. The packing layout
also depends on the seed, device, and fonts, so the page does not claim that any
word belongs at a particular coordinate.

The area-corrected ggwordcloud option, `area_corr`, is a different choice. It
shrinks or grows labels so each whole text box better follows the count, which
trades the length bias for a letter-height bias because short words get taller
type. Cleveland and McGill did not test word clouds; Felix and colleagues later
found that bar length beat font size for judging values.

The short part of the text alternative is the cloud's alt text:
"`r cloud_alt_seed_7701`" The long part is the sorted count table above, which
lists every word and count used in the cloud.

## Change the seed, not the data

The next cloud uses the identical keyed counts and a different seed. It is still
the same input table.

```{r}
#| label: fig-word-cloud-seed-two
#| fig-cap: "The same 22 count-table words drawn again with seed 7702."
#| fig-alt: "Second word cloud with the same words and counts as the first cloud but a different random layout seed. Government still has 585 counts and people has 561; position changes are layout artifacts."
#| fig-width: 7
#| fig-height: 4.5

set.seed(7702)
cloud_plot_seed_7702 <- ggplot(
  cloud_words,
  aes(label = word_key, size = count)
) +
  geom_text_wordcloud(
    seed = 7702,
    color = "#3B5F8A",
    angle = 0,
    rm_outside = FALSE
  ) +
  scale_size_area(max_size = 9) +
  labs(size = "Count") +
  theme_minimal()

cloud_plot_seed_7702
```

```{r}
#| include: false

cloud_input_seed_7702 <- cloud_words |>
  arrange(desc(count), word_key) |>
  select(word_key, count)
cloud_layer_seed_7702 <- extract_cloud_layer(
  cloud_plot_seed_7702,
  count_table
)
cloud_size_ratio_seed_7702 <- cloud_layer_seed_7702$plotted_size^2 /
  cloud_layer_seed_7702$count

stopifnot(
  identical(cloud_input_seed_7702, cloud_input_seed_7701),
  identical(nrow(cloud_layer_seed_7702), nrow(count_table)),
  !anyNA(cloud_layer_seed_7702$count),
  setequal(cloud_layer_seed_7702$word_key, count_table$word_key),
  max(cloud_size_ratio_seed_7702) - min(cloud_size_ratio_seed_7702) < 1e-10,
  identical(setequal(cloud_input_seed_7702$word_key, cloud_input_seed_7701$word_key), TRUE),
  identical(sum(cloud_input_seed_7702$count), sum(cloud_input_seed_7701$count))
)
```

If the two pictures feel different, that feeling is the warning. The counts did
not change.

## Check preprocessing sensitivity

Stop-word lists are editorial choices. The table below recomputes the top words
with three named choices: onix only, snowball only, and all three tidytext
lexicons together.

```{r}
stopword_choices <- list(
  "onix only" = stop_words |>
    filter(lexicon == "onix") |>
    distinct(word),
  "snowball only" = stop_words |>
    filter(lexicon == "snowball") |>
    distinct(word),
  "all tidytext lexicons" = stop_words |>
    distinct(word)
)

sensitivity_table <- imap_dfr(
  stopword_choices,
  \(stop_table, stop_choice) {
    word_tokens |>
      anti_join(stop_table, by = join_by(word)) |>
      count(word, sort = TRUE, name = "count") |>
      slice_head(n = 10) |>
      mutate(stop_choice = stop_choice, rank = row_number(), .before = 1)
  }
)

kable(
  sensitivity_table,
  col.names = c("Stop-word choice", "Rank", "Word", "Count"),
  caption = "Top words change when the stop-word lexicon changes",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sensitivity_table), 30L),
  identical(
    sensitivity_table$word[sensitivity_table$stop_choice == "onix only"][1:2],
    c("government", "people")
  ),
  "states" %in% sensitivity_table$word[
    sensitivity_table$stop_choice == "snowball only"
  ],
  !"states" %in% sensitivity_table$word[
    sensitivity_table$stop_choice == "onix only"
  ],
  "united" %in% inaugural_word_counts$word
)
```

With snowball only, common words such as `will`, `us`, and `states` remain.
With onix, `states` is removed while `united` remains. A cloud without this
preprocessing note would look more objective than it is.

## Check the word-length trap

The next table compares real words from the count table. The rough space score
is `letters * sqrt(count)`. It is not a perception model; it only shows why a
long word can occupy more horizontal room even when its count is lower.
Alexander and coauthors tested word-length effects in word clouds; Felix,
Franconeri, and Bertini tested magnitude judgments for several visual encodings,
not word length.

```{r}
length_pairs <- inaugural_word_counts |>
  filter(word_key %in% c("constitution", "time", "peace", "government", "people")) |>
  transmute(
    word_key,
    count,
    letters,
    rough_horizontal_space = round(letters * sqrt(count), 1)
  ) |>
  arrange(desc(rough_horizontal_space))

kable(
  length_pairs,
  col.names = c("Word", "Count", "Letters", "Rough space score"),
  caption = "Long words can cover more space than shorter words with higher counts",
  row.names = FALSE
)
```

```{r}
#| include: false

constitution_space <- length_pairs |>
  filter(word_key == "constitution") |>
  pull(rough_horizontal_space)
time_space <- length_pairs |>
  filter(word_key == "time") |>
  pull(rough_horizontal_space)
peace_space <- length_pairs |>
  filter(word_key == "peace") |>
  pull(rough_horizontal_space)

stopifnot(
  identical(
    length_pairs$count[length_pairs$word_key == "constitution"],
    207L
  ),
  identical(length_pairs$count[length_pairs$word_key == "time"], 214L),
  constitution_space > time_space,
  constitution_space > peace_space,
  length_pairs$letters[length_pairs$word_key == "constitution"] == 12L,
  length_pairs$letters[length_pairs$word_key == "time"] == 4L
)
```

`constitution` appears 207 times and `time` appears 214 times. The longer word
can still take about `r round(constitution_space / time_space, 1)` times the
horizontal space under this simple geometry check. That is why Iris should not
ask readers to compare close counts from font size.

## What this cloud does not show

The cloud does not show word order, phrases, relationships, distinctiveness,
change over time, or importance. It also does not show that `united` and
`states` were part of a phrase before preprocessing split and filtered them.

Use a cloud as a doorway to the table, not as the table's replacement.

## What to remember

- A word cloud starts from a count table; keep that table visible.
- The onix stop-word lexicon removes `states` but keeps `united` in this corpus.
- The bar chart and both clouds here use the same 22 keyed counts.
- Word-cloud placement, adjacency, centre position, and colour carry no meaning.
- Long words can look stronger than shorter words with similar counts.
- Sorted counts are the long text alternative for the cloud.

Iris keeps the cloud, but the slide title points to the count table: "Common
words after one stated cleanup."

## Sources

- [ggwordcloud vignette](https://lepennec.github.io/ggwordcloud/articles/ggwordcloud.html)
- [tidytext stop-word documentation](https://juliasilge.github.io/tidytext/reference/stop_words.html)
- [Cleveland and McGill on graphical perception](https://doi.org/10.1080/01621459.1984.10478080)
- [Heer and Bostock on crowdsourced graphical perception](https://doi.org/10.1145/1753326.1753357)
- [Alexander et al. on word-length effects in word clouds](https://doi.org/10.1109/TVCG.2017.2723397)
- [Felix, Franconeri, and Bertini on word-cloud magnitude judgments](https://doi.org/10.1109/TVCG.2017.2746018)
- [Hearst and Rosner on tag-cloud limits](https://doi.org/10.1109/HICSS.2008.422)
- [W3C guidance on complex image text alternatives](https://www.w3.org/WAI/tutorials/images/complex/)
