---
title: "Projecting word vectors onto a map"
subtitle: "Measure what the picture keeps and what it loses"
description: "Learn how PCA and t-SNE draw word embeddings in two dimensions while preserving only part of the original neighbourhoods."
categories: [information visualization, word embeddings, dimensionality reduction]
---

Theo wants to show a small word-vector model on one slide. A scatterplot feels
like a map, and maps invite stories.

This picture needs a warning label. Each word begins as 50 numbers. A
two-dimensional drawing can keep some nearby neighbours and lose others.

**Word embedding visualization** draws word vectors as points by projecting many
numeric dimensions into two. The plot is a question to check, not a finding
about topics, clusters, or meaning.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- verify the pinned inaugural word2vec model before loading it;
- scale each word vector to length 1 so straight-line distance ranks
  neighbours the same way as cosine similarity;
- choose a stated set of frequent words to plot;
- mark, in the plot itself, which nearest neighbours a projection keeps;
- compare PCA with t-SNE, and two t-SNE seeds, by neighbour preservation; and
- avoid reading clusters, gaps, sizes, and axis directions as evidence.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load and verify the pinned model

The model comes from lesson 61. It is deliberately small: the metadata records
134,419 cleaned training tokens and a 50-dimensional skip-gram word2vec fit. In
lesson 61, even the easy pair `freedom / liberty` landed below the median of all
pairwise similarities, so this lesson measures whether the picture keeps the
model's neighbours. It does not claim the model learned reliable meaning.

```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(readr)
library(purrr)
library(tidytext)
library(quanteda)
library(word2vec)
library(Rtsne)
library(ggplot2)
library(digest)

source("R/inaugural-corpus.R")
source("R/lesson-figures.R")

metadata <- read_csv(
  "data/word2vec/word2vec-metadata.csv",
  na = character(),
  col_types = cols(
    field = col_character(),
    value = col_character()
  )
)
metadata_values <- setNames(metadata$value, metadata$field)

model_path <- "data/word2vec/inaugural-word2vec.bin"
model_sha256 <- digest(file = model_path, algo = "sha256")
expected_model_sha256 <- metadata_values[["sha256_model"]]

stopifnot(
  identical(model_sha256, expected_model_sha256)
)

word2vec_model <- word2vec::read.word2vec(model_path)
embedding_matrix <- as.matrix(word2vec_model)
```

```{r}
#| echo: false

tibble(
  item = c(
    "training tokens",
    "vocabulary rows",
    "dimensions",
    "algorithm",
    "model SHA-256 matches metadata"
  ),
  value = c(
    metadata_values[["training_tokens"]],
    metadata_values[["vocabulary_rows"]],
    metadata_values[["dimensions"]],
    metadata_values[["algorithm"]],
    as.character(identical(model_sha256, expected_model_sha256))
  )
) |>
  knitr::kable(
    col.names = c("Model check", "Value"),
    caption = "Pinned word2vec model loaded for projection",
    row.names = FALSE
  )
```

```{r}
#| include: false

stopifnot(
  identical(model_sha256, expected_model_sha256),
  identical(model_sha256, "b182100ecb33cb339f16fdf5df5cea783dd113843a3a89e7e65393b921e448dd"),
  identical(nrow(embedding_matrix), as.integer(metadata_values[["vocabulary_rows"]])),
  identical(ncol(embedding_matrix), as.integer(metadata_values[["dimensions"]])),
  identical(metadata_values[["training_tokens"]], "134419"),
  identical(metadata_values[["algorithm"]], "skip-gram")
)
```

The fingerprint check fails closed if the binary changes. Rendering does not
train, download, or update a model.

## Choose the plotted words

The full saved vocabulary has 2,709 rows, including the trainer's `</s>` marker.
This page plots the 300 most frequent training words that also have vectors.
Changing the subset can change the projection.

```{r}
paragraphs <- inaugural_paragraphs()

training_text <- tokens(
  str_to_lower(paragraphs$paragraph, locale = "en"),
  remove_punct = TRUE,
  remove_numbers = TRUE
) |>
  vapply(paste, character(1), collapse = " ")

training_frequencies <- tibble(text = training_text) |>
  unnest_tokens(word, text) |>
  count(word, sort = TRUE, name = "cleaned_paragraph_count") |>
  filter(word %in% rownames(embedding_matrix), word != "</s>")

selected_words <- training_frequencies |>
  slice_head(n = 300)

selected_embeddings <- embedding_matrix[selected_words$word, , drop = FALSE]
normalized_embeddings <- selected_embeddings / sqrt(rowSums(selected_embeddings^2))
```

```{r}
#| echo: false

knitr::kable(
  selected_words |>
    slice_head(n = 12),
  col.names = c("Word", "Count in cleaned paragraphs"),
  caption = "The most frequent plotted words before projection",
  row.names = FALSE
)
```

```{r}
#| include: false

row_lengths <- sqrt(rowSums(normalized_embeddings^2))

stopifnot(
  identical(nrow(selected_words), 300L),
  identical(rownames(normalized_embeddings), selected_words$word),
  identical(ncol(normalized_embeddings), 50L),
  all(abs(row_lengths - 1) < 1e-10),
  identical(selected_words$word[1], "the"),
  identical(max(selected_words$cleaned_paragraph_count), 10003L),
  all(c("government", "people", "freedom", "union", "peace", "constitution") %in%
    selected_words$word)
)
```

The 300 plotted words range from
`r format(min(selected_words$cleaned_paragraph_count), big.mark = ",")` to
`r format(max(selected_words$cleaned_paragraph_count), big.mark = ",")` uses.
The count column is an approximate training count rebuilt from the cleaned
paragraphs with tidytext's tokenizer, not a byte-for-byte count from word2vec's
internal splitter.

Dividing each vector by its length, a step called L2 normalization, makes every
row length equal to 1. On rows of length 1, straight-line (Euclidean) distance
ranks neighbours the same way as cosine similarity, the measure used in lesson
61. The neighbour lists below are computed that way in all 50 dimensions, and
the same kind of distance is then measured in each two-dimensional drawing.

## Decide what counts as a kept neighbour

A plot of words is only useful if nearby points were nearby in the model. For
each word, the code below finds its 10 nearest neighbours among the 300 plotted
words, measured in all 50 dimensions, and later its 10 nearest points in a
two-dimensional drawing. A neighbour is **kept** when it is also among the
word's 10 nearest points in the drawing. A neighbour that is not kept may sit
just outside that top ten or far away; the count does not say which.

```{r}
nearest_by_similarity <- function(matrix, k = 10L) {
  similarity <- tcrossprod(matrix)
  diag(similarity) <- NA_real_
  t(apply(
    similarity,
    1,
    \(values) names(sort(values, decreasing = TRUE))[seq_len(k)]
  ))
}

nearest_by_distance <- function(points, k = 10L) {
  distances <- as.matrix(dist(points))
  diag(distances) <- Inf
  t(apply(
    distances,
    1,
    \(values) names(sort(values))[seq_len(k)]
  ))
}

full_space_neighbours <- nearest_by_similarity(normalized_embeddings, 10L)

label_words <- c(
  "government", "people", "freedom", "union",
  "peace", "constitution", "war", "america"
)
```

```{r}
#| include: false

stopifnot(
  identical(dim(full_space_neighbours), c(300L, 10L)),
  identical(rownames(full_space_neighbours), selected_words$word),
  all(label_words %in% rownames(full_space_neighbours)),
  !any(full_space_neighbours == rownames(full_space_neighbours))
)
```

The eight words in `label_words` get their own panel in each figure below. They
are landmarks for matching the two figures by eye, not cluster names.

## Project with PCA

**PCA**, short for principal component analysis, finds the directions along
which the normalized vectors spread out most. The plot uses the first two.

```{r}
pca_fit <- prcomp(normalized_embeddings, center = TRUE, scale. = FALSE)
pca_variance <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
pca_scores <- pca_fit$x[, 1:2, drop = FALSE]

# A component's sign is arbitrary, so fix it with named words before plotting.
if (pca_scores["government", "PC1"] < 0) {
  pca_scores[, "PC1"] <- -pca_scores[, "PC1"]
}

if (pca_scores["freedom", "PC2"] < 0) {
  pca_scores[, "PC2"] <- -pca_scores[, "PC2"]
}

pca_neighbours <- nearest_by_distance(pca_scores, 10L)
```

```{r}
#| include: false

stopifnot(
  identical(rownames(pca_scores), rownames(full_space_neighbours)),
  identical(rownames(pca_neighbours), rownames(full_space_neighbours)),
  sum(pca_variance[1:2]) > 0.20,
  sum(pca_variance[1:2]) < 0.30,
  identical(round(100 * sum(pca_variance[1:2]), 1), 23.9),
  pca_variance[1] > pca_variance[2]
)
```

The first two components hold
`r sprintf("%.1f%%", 100 * sum(pca_variance[1:2]))` of the variance in the
normalized vectors. The rest is invisible in this plot.

The function below turns a set of two-dimensional coordinates into one panel
per landmark word. In each panel, every 50-dimensional neighbour of the landmark is
marked as kept or not kept, and the panel title counts how many were kept.

```{r}
anchor_panels <- function(points, neighbours_2d) {
  plotted <- as_tibble(points, rownames = "word", .name_repair = "minimal") |>
    setNames(c("word", "x", "y"))

  map(label_words, \(anchor) {
    true_neighbours <- full_space_neighbours[anchor, ]
    kept <- intersect(true_neighbours, neighbours_2d[anchor, ])

    plotted |>
      mutate(
        anchor_word = anchor,
        kept_count = length(kept),
        role = case_when(
          word == anchor ~ "landmark",
          word %in% kept ~ "kept",
          word %in% true_neighbours ~ "not kept",
          TRUE ~ "other"
        )
      )
  }) |>
    list_rbind() |>
    mutate(
      panel = factor(
        str_c(anchor_word, ": ", kept_count, " of 10 kept"),
        levels = unique(str_c(anchor_word, ": ", kept_count, " of 10 kept"))
      )
    )
}

draw_projection <- function(panels) {
  # A square window gives every panel the same shape on every computer. A
  # t-SNE run can come out wider or taller, and coord_equal() alone would then
  # change the panel shape.
  half_span <- max(diff(range(panels$x)), diff(range(panels$y))) / 2
  x_centre <- mean(range(panels$x))
  y_centre <- mean(range(panels$y))

  ggplot(panels, aes(x = x, y = y)) +
    geom_point(
      data = \(d) filter(d, role == "other"),
      colour = lesson_colours[["context"]],
      size = 0.7
    ) +
    geom_point(
      data = \(d) filter(d, role == "not kept"),
      shape = 22,
      fill = lesson_colours[["paper"]],
      colour = lesson_colours[["contrast"]],
      stroke = 0.9,
      size = 2.3
    ) +
    geom_point(
      data = \(d) filter(d, role == "kept"),
      shape = 22,
      fill = lesson_colours[["accent"]],
      colour = lesson_colours[["paper"]],
      stroke = 0.4,
      size = 2.6
    ) +
    geom_point(
      data = \(d) filter(d, role == "landmark"),
      shape = 24,
      fill = lesson_colours[["ink"]],
      colour = lesson_colours[["paper"]],
      stroke = 0.6,
      size = 3.8
    ) +
    facet_wrap(vars(panel), ncol = 4) +
    coord_equal(
      xlim = x_centre + c(-1, 1) * half_span,
      ylim = y_centre + c(-1, 1) * half_span
    ) +
    labs(x = NULL, y = NULL) +
    theme_lesson(grid = "none") +
    theme(
      axis.text = element_blank(),
      panel.background = element_rect(fill = "#F4F6F8", colour = NA)
    )
}

key_subtitle <- str_c(
  "Each panel shows one landmark word (dark triangle) and its 10 nearest ",
  "neighbours<br>among the 300 words, measured in all 50 dimensions. ",
  lesson_key("Filled squares", lesson_text_colours[["accent"]]),
  " are still among<br>its 10 nearest points here; ",
  lesson_key("hollow squares", lesson_text_colours[["contrast"]]),
  " are not."
)

pca_panels <- anchor_panels(pca_scores, pca_neighbours)
```

```{r}
#| include: false

pca_kept <- pca_panels |>
  distinct(anchor_word, kept_count)

pca_alt_text <- str_c(
  "Eight-panel PCA plot, one panel per landmark word. Each panel shows the 300 ",
  "projected words as faint dots, the landmark as a dark triangle, and its 10 ",
  "nearest neighbours in the 50-dimensional vectors as squares, filled when they ",
  "stay among its 10 nearest points in the plot and hollow when they do not. ",
  "The number kept ranges from ",
  min(pca_kept$kept_count), " to ", max(pca_kept$kept_count),
  " of 10 across the eight landmarks."
)

stopifnot(
  identical(nrow(pca_panels), 300L * length(label_words)),
  identical(sum(pca_panels$role == "landmark"), length(label_words)),
  all(
    pca_panels |>
      filter(role %in% c("kept", "not kept")) |>
      count(anchor_word) |>
      pull(n) == 10L
  ),
  identical(pca_kept$anchor_word, label_words),
  all(pca_kept$kept_count <= 5L),
  mean(pca_kept$kept_count) < 3,
  all(is.finite(pca_panels$x)),
  all(is.finite(pca_panels$y))
)
```

```{r}
#| label: fig-word-pca
#| fig-cap: "PCA projection of 300 word vectors, one panel per landmark word. Squares mark the landmark's 10 nearest neighbours among the 300 words, measured in all 50 dimensions."
#| fig-alt: !expr pca_alt_text
#| fig-width: 8
#| fig-height: 5.4

pca_plot <- draw_projection(pca_panels) +
  labs(
    title = "PCA keeps few of each landmark's 10 nearest neighbours",
    subtitle = key_subtitle
  )

pca_plot
```

```{r}
#| include: false

stopifnot(
  identical(nlevels(pca_panels$panel), length(label_words)),
  identical(
    length(unique(ggplot_build(pca_plot)$layout$layout$PANEL)),
    length(label_words)
  )
)
```

PCA axes can flip sign across software builds, so the code fixes the signs with
named words before plotting. The plot has no axis labels because left, right,
up, and down mean nothing here; only nearness inside a panel matters. Each
panel title reports how many of the landmark's 10 nearest neighbours stayed among
its 10 nearest points in the drawing.

## Project with t-SNE

**t-SNE** is a stochastic method that tries to keep local neighbours together in
the drawing. It can also create groups and gaps that are not stable evidence.
Here, perplexity 30 sets roughly how many neighbours each point considers; it
is unrelated to the language-model perplexity in lessons 63 and 64. Theta 0.5
trades a little accuracy for speed, and 750 iterations is the number of
optimization updates. The run also uses seed 7801, no initial PCA step, no
extra normalization, and one thread.

```{r}
run_tsne <- function(seed) {
  set.seed(seed)
  fit <- Rtsne(
    normalized_embeddings,
    dims = 2,
    perplexity = 30,
    theta = 0.5,
    max_iter = 750L,
    pca = FALSE,
    normalize = FALSE,
    check_duplicates = FALSE,
    num_threads = 1,
    verbose = FALSE
  )
  coordinates <- fit$Y
  rownames(coordinates) <- rownames(normalized_embeddings)
  coordinates
}

tsne_seed_7801 <- run_tsne(7801)
tsne_neighbours_7801 <- nearest_by_distance(tsne_seed_7801, 10L)
tsne_panels <- anchor_panels(tsne_seed_7801, tsne_neighbours_7801)
```

```{r}
#| include: false

tsne_kept <- tsne_panels |>
  distinct(anchor_word, kept_count)

tsne_alt_text <- str_c(
  "Eight-panel t-SNE plot using seed 7801, one panel per landmark word, drawn ",
  "the same way as the PCA figure. The number of 50-dimensional neighbours kept ",
  "among the landmark's 10 nearest points ranges from ",
  min(tsne_kept$kept_count), " to ", max(tsne_kept$kept_count),
  " of 10 across the eight landmarks. The preservation table below gives the ",
  "average over all 300 words."
)

stopifnot(
  !any(duplicated(as.data.frame(normalized_embeddings))),
  3 * 30 < nrow(normalized_embeddings) - 1,
  identical(dim(tsne_seed_7801), c(300L, 2L)),
  identical(rownames(tsne_neighbours_7801), rownames(full_space_neighbours)),
  identical(nrow(tsne_panels), 300L * length(label_words)),
  identical(tsne_kept$anchor_word, label_words),
  sum(tsne_kept$kept_count) > sum(pca_kept$kept_count),
  any(tsne_kept$kept_count < 10L),
  all(is.finite(tsne_panels$x)),
  all(is.finite(tsne_panels$y))
)
```

```{r}
#| label: fig-word-tsne
#| fig-cap: "t-SNE projection of the same 300 word vectors with seed 7801, drawn the same way as the PCA figure."
#| fig-alt: !expr tsne_alt_text
#| fig-width: 8
#| fig-height: 5.4

tsne_plot <- draw_projection(tsne_panels) +
  labs(
    title = "t-SNE keeps more neighbours close, but not all of them",
    subtitle = key_subtitle
  )

tsne_plot
```

```{r}
#| include: false

stopifnot(
  identical(nlevels(tsne_panels$panel), length(label_words)),
  identical(
    length(unique(ggplot_build(tsne_plot)$layout$layout$PANEL)),
    length(label_words)
  )
)
```

The panels hold the same landmarks as the PCA figure, so the two pictures can
be compared by eye. The counts in the panel titles are the evidence; the shapes
of the groups are not.

## Measure neighbour preservation

A **10-nearest-neighbour preservation** score averages the panel-title count
over every plotted word, not just the eight landmarks, and divides by 10.
Chance is 10 divided by 299, or
`r sprintf("%.1f%%", 100 * 10 / (nrow(normalized_embeddings) - 1))`. A second
t-SNE run with seed 7802 shows how much of the picture depends on the seed.

```{r}
neighbour_overlap <- function(first, second) {
  mean(vapply(
    seq_len(nrow(first)),
    \(row) sum(first[row, ] %in% second[row, ]) / ncol(first),
    numeric(1)
  ))
}

tsne_seed_7802 <- run_tsne(7802)
tsne_neighbours_7802 <- nearest_by_distance(tsne_seed_7802, 10L)

preservation_table <- tibble(
  comparison = c(
    "PCA versus 50-dimensional neighbours",
    "t-SNE seed 7801 versus 50-dimensional neighbours",
    "t-SNE seed 7802 versus 50-dimensional neighbours",
    "t-SNE seed 7801 versus t-SNE seed 7802",
    "chance level for a random 10-word set"
  ),
  preservation = c(
    neighbour_overlap(full_space_neighbours, pca_neighbours),
    neighbour_overlap(full_space_neighbours, tsne_neighbours_7801),
    neighbour_overlap(full_space_neighbours, tsne_neighbours_7802),
    neighbour_overlap(tsne_neighbours_7801, tsne_neighbours_7802),
    10 / (nrow(normalized_embeddings) - 1)
  )
)
```

```{r}
#| echo: false

knitr::kable(
  preservation_table |>
    mutate(preservation = sprintf("%.1f%%", 100 * preservation)),
  col.names = c("Comparison", "Average 10-neighbour overlap"),
  caption = "Neighbour preservation for two-dimensional projections",
  row.names = FALSE
)
```

```{r}
#| include: false

pca_preservation <- preservation_table$preservation[1]
tsne_preservation_7801 <- preservation_table$preservation[2]
tsne_preservation_7802 <- preservation_table$preservation[3]
cross_seed_agreement <- preservation_table$preservation[4]
chance_preservation <- preservation_table$preservation[5]

display_preservation <- sprintf("%.1f%%", 100 * preservation_table$preservation)

stopifnot(
  identical(rownames(tsne_neighbours_7802), rownames(full_space_neighbours)),
  identical(dim(tsne_neighbours_7802), c(300L, 10L)),
  chance_preservation > 0.03,
  chance_preservation < 0.04,
  pca_preservation > chance_preservation * 4,
  pca_preservation < 0.35,
  tsne_preservation_7801 > pca_preservation * 2,
  tsne_preservation_7802 > pca_preservation * 2,
  tsne_preservation_7801 < 0.90,
  cross_seed_agreement > tsne_preservation_7801,
  cross_seed_agreement < 0.90
)
```

The measured result is modest but useful. PCA keeps about
`r display_preservation[1]` of each word's 10 neighbours. The plotted t-SNE run
keeps about `r display_preservation[2]`, and the second seed keeps about
`r display_preservation[3]`. The two t-SNE runs agree with each other on about
`r display_preservation[4]` of neighbours. That is well above chance, but it is
not perfect stability.

## Inspect labelled words as rows

The table below is the text alternative for the landmark panels. It lists
neighbours from the full vector space and from the two plotted projections, so
no one has to infer them from point positions.

```{r}
neighbour_table <- tibble(word = label_words) |>
  mutate(
    full_space = map_chr(word, \(w) str_c(full_space_neighbours[w, 1:5], collapse = ", ")),
    pca_plot = map_chr(word, \(w) str_c(pca_neighbours[w, 1:5], collapse = ", ")),
    tsne_plot = map_chr(word, \(w) str_c(tsne_neighbours_7801[w, 1:5], collapse = ", "))
  )
```

```{r}
#| echo: false

knitr::kable(
  neighbour_table,
  col.names = c(
    "Landmark word",
    "Nearest in 50 dimensions",
    "Nearest in PCA plot",
    "Nearest in t-SNE plot"
  ),
  caption = "Five nearest neighbours of each landmark word, in the model and in the two projections",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(neighbour_table$word, label_words),
  all(str_count(neighbour_table$full_space, ",") == 4L),
  all(str_count(neighbour_table$pca_plot, ",") == 4L),
  all(str_count(neighbour_table$tsne_plot, ",") == 4L),
  all(!str_detect(neighbour_table$full_space, fixed("</s>")))
)
```

If a nearby label looks interesting, the next step is to read the texts or the
neighbour table. The picture alone is not evidence that the words
form a topic.

## What this does not show

The t-SNE plot does not give meaningful cluster sizes, gap sizes, group
positions, or axis directions. PCA directions are mathematical summaries, but
their signs can flip, and this plot still leaves most of the variance out.

The preservation scores describe how faithfully the projection draws this small
model, not how faithfully the model captures meaning. Small-corpus embeddings
can change their neighbour lists when trained again, which is why this model is
pinned. The same t-SNE seed can also draw a different layout on another
computer, because low-level numerical details can differ, so every count on
this page is computed during the render.

## What to remember

- Each plotted point began as a 50-number word vector.
- Rows were scaled to length 1 before PCA and t-SNE, so distance in the
  50-dimensional space ranks neighbours the same way as cosine similarity; a
  projection can still change that order.
- PCA's first two components explain about
  `r sprintf("%.1f%%", 100 * sum(pca_variance[1:2]))` of the variance here.
- Mark kept and lost neighbours in the plot instead of trusting how groups look.
- t-SNE preserves more local neighbours than PCA in this run, but two seeds
  still disagree on many neighbours.
- Do not interpret clusters, gaps, sizes, positions, or axis directions as
  findings.

Keep the scatterplot beside the preservation table. The table says what the
picture kept.

## Sources

- [Lesson 61, Training word vectors](../12_similarity/61-distributed-word-representations.qmd)
- [van der Maaten and Hinton on t-SNE](https://jmlr.org/papers/v9/vandermaaten08a.html)
- [Rtsne package documentation](https://cran.r-project.org/package=Rtsne)
- [Kobak and Berens on t-SNE art and analysis](https://doi.org/10.1038/s41467-019-13056-x)
- [Wattenberg, Viégas, and Johnson on how to use t-SNE effectively](https://doi.org/10.23915/distill.00002)
- [Antoniak and Mimno on stability of word embeddings](https://aclanthology.org/Q18-1008/)
- [R `prcomp` documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/prcomp.html)
