---
title: "Projecting word vectors onto a map"
subtitle: "Measure what the picture keeps and what it loses"
description: "Learn how PCA and t-SNE draw word embeddings in two dimensions while preserving only part of the original neighbourhoods."
categories: [information visualization, word embeddings, dimensionality reduction]
---

Theo wants to show a small word-vector model on one slide. A scatterplot feels
like a map, and maps invite stories.

This picture needs a warning label. Each word begins as 50 numbers. A
two-dimensional drawing can keep some nearby neighbours and lose others.

**Word embedding visualization** draws word vectors as points by projecting many
numeric dimensions into two. The plot is a question to check, not a finding
about topics, clusters, or meaning.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- verify the pinned inaugural word2vec model before loading it;
- L2-normalize rows so Euclidean projection distances match cosine neighbours;
- select a stated subset of frequent words;
- compare PCA with t-SNE using 10-nearest-neighbour preservation;
- check two t-SNE seeds instead of trusting one run; and
- avoid reading clusters, gaps, sizes, and axis directions as evidence.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load and verify the pinned model

The model comes from lesson 61. It is deliberately small: the metadata records
134,419 cleaned training tokens and a 50-dimensional skip-gram word2vec fit. In
lesson 61, even the easy pair `freedom / liberty` landed below the median of all
pairwise similarities, so this lesson measures whether the picture keeps the
model's neighbours. It does not claim the model learned reliable meaning.

```{r}
library(dplyr)
library(tidyr)
library(tibble)
library(stringr)
library(readr)
library(tidytext)
library(quanteda)
library(word2vec)
library(Rtsne)
library(ggplot2)
library(knitr)
library(digest)

source("R/inaugural-corpus.R")

metadata <- read_csv(
  "data/word2vec/word2vec-metadata.csv",
  na = character(),
  col_types = cols(
    field = col_character(),
    value = col_character()
  )
)
metadata_values <- setNames(metadata$value, metadata$field)

model_path <- "data/word2vec/inaugural-word2vec.bin"
model_sha256 <- digest(file = model_path, algo = "sha256")
expected_model_sha256 <- metadata_values[["sha256_model"]]

stopifnot(
  identical(model_sha256, expected_model_sha256)
)

word2vec_model <- word2vec::read.word2vec(model_path)
embedding_matrix <- as.matrix(word2vec_model)

model_summary <- tibble(
  item = c(
    "training tokens",
    "vocabulary rows",
    "dimensions",
    "algorithm",
    "model SHA-256 matches metadata"
  ),
  value = c(
    metadata_values[["training_tokens"]],
    metadata_values[["vocabulary_rows"]],
    metadata_values[["dimensions"]],
    metadata_values[["algorithm"]],
    as.character(identical(model_sha256, expected_model_sha256))
  )
)

kable(
  model_summary,
  col.names = c("Model check", "Value"),
  caption = "Pinned word2vec model loaded for projection",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(model_sha256, expected_model_sha256),
  identical(model_sha256, "b182100ecb33cb339f16fdf5df5cea783dd113843a3a89e7e65393b921e448dd"),
  identical(nrow(embedding_matrix), as.integer(metadata_values[["vocabulary_rows"]])),
  identical(ncol(embedding_matrix), as.integer(metadata_values[["dimensions"]])),
  identical(metadata_values[["training_tokens"]], "134419"),
  identical(metadata_values[["algorithm"]], "skip-gram")
)
```

The fingerprint check fails closed if the binary changes. Rendering does not
train, download, or update a model.

## Choose the plotted words

The full saved vocabulary has 2,709 rows, including the trainer's `</s>` marker.
This page plots the 300 most frequent training words that also have vectors.
Changing the subset can change the projection.

```{r}
paragraphs <- inaugural_paragraphs()

training_text <- tokens(
  str_to_lower(paragraphs$paragraph, locale = "en"),
  remove_punct = TRUE,
  remove_numbers = TRUE
) |>
  vapply(paste, character(1), collapse = " ")

training_frequencies <- tibble(text = training_text) |>
  unnest_tokens(word, text) |>
  count(word, sort = TRUE, name = "cleaned_paragraph_count") |>
  filter(word %in% rownames(embedding_matrix), word != "</s>")

subset_n <- 300L
selected_words <- training_frequencies |>
  slice_head(n = subset_n)

selected_embeddings <- embedding_matrix[selected_words$word, , drop = FALSE]
embedding_norms <- sqrt(rowSums(selected_embeddings^2))
normalized_embeddings <- selected_embeddings / embedding_norms
rownames(normalized_embeddings) <- selected_words$word

subset_summary <- tibble(
  item = c("plotted words", "minimum cleaned count", "maximum cleaned count"),
  value = c(
    nrow(selected_words),
    min(selected_words$cleaned_paragraph_count),
    max(selected_words$cleaned_paragraph_count)
  )
)

kable(
  selected_words |>
    slice_head(n = 12),
  col.names = c("Word", "Count in cleaned paragraphs"),
  caption = "The most frequent plotted words before projection",
  row.names = FALSE
)

kable(
  subset_summary,
  col.names = c("Subset check", "Value"),
  caption = "Projection subset for this lesson",
  row.names = FALSE
)
```

```{r}
#| include: false

row_lengths <- sqrt(rowSums(normalized_embeddings^2))

stopifnot(
  identical(nrow(selected_words), 300L),
  identical(ncol(normalized_embeddings), 50L),
  all(abs(row_lengths - 1) < 1e-10),
  identical(selected_words$word[1], "the"),
  identical(max(selected_words$cleaned_paragraph_count), 10003L),
  all(c("government", "people", "freedom", "union", "peace", "constitution") %in%
    selected_words$word)
)
```

L2 normalization makes each row length equal to 1. On unit-length rows, Euclidean
distance ranks neighbours the same way as cosine similarity, the measure used in
lesson 61. The count column is an approximate training count rebuilt from the
cleaned paragraphs with tidytext's tokenizer, not a byte-for-byte count from
word2vec's internal splitter.

## Project with PCA

**PCA**, short for principal component analysis, finds directions with the most
spread in the normalized vectors. The table below reports how much variance the
first two components hold. The rest is invisible in this plot.

```{r}
pca_fit <- prcomp(normalized_embeddings, center = TRUE, scale. = FALSE)
pca_variance <- pca_fit$sdev^2 / sum(pca_fit$sdev^2)
pca_scores <- pca_fit$x[, 1:2, drop = FALSE]

if (pca_scores["government", "PC1"] < 0) {
  pca_scores[, "PC1"] <- -pca_scores[, "PC1"]
}

if (pca_scores["freedom", "PC2"] < 0) {
  pca_scores[, "PC2"] <- -pca_scores[, "PC2"]
}

pca_points <- as_tibble(pca_scores, rownames = "word") |>
  rename(x = PC1, y = PC2) |>
  left_join(selected_words, by = join_by(word))

label_words <- c(
  "government", "people", "freedom", "union",
  "peace", "constitution", "war", "america"
)

nearest_by_similarity <- function(matrix, k = 10L) {
  similarity <- tcrossprod(matrix)
  diag(similarity) <- NA_real_
  t(apply(
    similarity,
    1,
    \(values) names(sort(values, decreasing = TRUE))[seq_len(k)]
  ))
}

make_anchor_panel_data <- function(points, anchors, neighbour_matrix) {
  bind_rows(lapply(
    anchors,
    \(anchor) {
      points |>
        mutate(
          anchor_word = anchor,
          panel_role = case_when(
            word == anchor ~ "anchor word",
            word %in% neighbour_matrix[anchor, ] ~ "10 full-space neighbours",
            TRUE ~ "other plotted words"
          )
        )
    }
  )) |>
    mutate(
      anchor_word = factor(anchor_word, levels = anchors),
      panel_role = factor(
        panel_role,
        levels = c(
          "other plotted words",
          "10 full-space neighbours",
          "anchor word"
        )
      )
    )
}

full_space_neighbours <- nearest_by_similarity(normalized_embeddings, 10L)
pca_panel_points <- make_anchor_panel_data(
  pca_points,
  label_words,
  full_space_neighbours
)

pca_variance_table <- tibble(
  component = c("PC1", "PC2", "PC1 plus PC2"),
  variance_explained = c(
    pca_variance[1],
    pca_variance[2],
    sum(pca_variance[1:2])
  )
)

kable(
  pca_variance_table |>
    mutate(variance_explained = sprintf("%.1f%%", 100 * variance_explained)),
  col.names = c("PCA component", "Variance explained"),
  caption = "Variance visible in the two-dimensional PCA plot",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(pca_points), 300L),
  identical(nrow(pca_panel_points), 300L * length(label_words)),
  identical(
    as.integer(table(pca_panel_points$panel_role)["anchor word"]),
    length(label_words)
  ),
  all(
    table(
      pca_panel_points$anchor_word,
      pca_panel_points$panel_role
    )[, "10 full-space neighbours"] == 10L
  ),
  sum(pca_variance[1:2]) > 0.20,
  sum(pca_variance[1:2]) < 0.30,
  identical(round(100 * sum(pca_variance[1:2]), 1), 23.9),
  pca_variance[1] > pca_variance[2],
  all(label_words %in% pca_panel_points$word)
)
```

```{r}
#| label: fig-word-pca
#| fig-cap: "PCA projection shown as one panel per anchor word. In each panel, faint dots are the other 289 plotted words, squares are that anchor's 10 nearest neighbours in the full 50-dimensional vectors, and the large triangle is the anchor word."
#| fig-alt: "Eight-panel PCA small multiple. Each panel is named for one anchor word. Every panel shows the other 289 projected words as faint dots, the anchor word as a larger triangle, and its 10 full-space nearest neighbours as squares; the first two PCA components explain about 23.9 percent of variance."
#| fig-width: 8
#| fig-height: 6.5

ggplot(pca_panel_points, aes(x = x, y = y)) +
  geom_point(
    data = pca_panel_points |>
      filter(panel_role == "other plotted words"),
    color = "#3B5F8A",
    alpha = 0.22,
    size = 0.8
  ) +
  geom_point(
    data = pca_panel_points |>
      filter(panel_role == "10 full-space neighbours"),
    color = "#8A5F3B",
    fill = "white",
    shape = 22,
    stroke = 0.7,
    size = 2.2
  ) +
  geom_point(
    data = pca_panel_points |>
      filter(panel_role == "anchor word"),
    color = "#1F2D3D",
    fill = "white",
    shape = 24,
    stroke = 0.9,
    size = 3.6
  ) +
  facet_wrap(vars(anchor_word), ncol = 4) +
  labs(
    x = "First principal component",
    y = "Second principal component"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  all(is.finite(pca_points$x)),
  all(is.finite(pca_points$y)),
  identical(levels(pca_panel_points$anchor_word), label_words)
)
```

PCA axes can flip sign across software builds, so the code fixes the signs with
named anchor words before plotting. This lesson does not interpret left, right,
up, or down. Each anchor word gets its own panel in both projection plots. The
facet strip names the anchor, the large triangle marks the anchor, and squares
mark all 10 of its nearest neighbours in the full vector space.

## Project with t-SNE

**t-SNE** is a stochastic method that tries to keep local neighbours together in
the drawing. It can make groups and gaps that are not stable evidence. Here,
perplexity 30 is a t-SNE setting for roughly how many neighbours each point
considers; it is unrelated to the language-model perplexity in lessons 63 and
64. Theta 0.5 trades a little accuracy for speed, and 750 iterations is the
number of optimization updates. The run also uses seed 7801, no initial PCA
step, no extra normalization, and one thread.

```{r}
tsne_perplexity <- 30
tsne_iterations <- 750L
tsne_theta <- 0.5
normalized_rows_unique <- !any(duplicated(as.data.frame(normalized_embeddings)))

stopifnot(
  normalized_rows_unique,
  3 * tsne_perplexity < nrow(normalized_embeddings) - 1
)

set.seed(7801)
tsne_seed_7801 <- Rtsne(
  normalized_embeddings,
  dims = 2,
  perplexity = tsne_perplexity,
  theta = tsne_theta,
  max_iter = tsne_iterations,
  pca = FALSE,
  normalize = FALSE,
  check_duplicates = FALSE,
  num_threads = 1,
  verbose = FALSE
)

tsne_points <- as_tibble(tsne_seed_7801$Y, .name_repair = "minimal") |>
  setNames(c("x", "y")) |>
  mutate(word = rownames(normalized_embeddings), .before = 1) |>
  left_join(selected_words, by = join_by(word))

tsne_labels <- tsne_points |>
  filter(word %in% label_words)
tsne_panel_points <- make_anchor_panel_data(
  tsne_points,
  label_words,
  full_space_neighbours
)

tsne_settings <- tibble(
  setting = c("seed", "perplexity", "iterations", "theta", "threads"),
  value = c(
    "7801",
    as.character(tsne_perplexity),
    as.character(tsne_iterations),
    as.character(tsne_theta),
    "1"
  )
)

kable(
  tsne_settings,
  col.names = c("t-SNE setting", "Value"),
  caption = "Settings for the plotted t-SNE run",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(tsne_points), 300L),
  identical(nrow(tsne_labels), length(label_words)),
  identical(nrow(tsne_panel_points), 300L * length(label_words)),
  identical(
    as.integer(table(tsne_panel_points$panel_role)["anchor word"]),
    length(label_words)
  ),
  all(
    table(
      tsne_panel_points$anchor_word,
      tsne_panel_points$panel_role
    )[, "10 full-space neighbours"] == 10L
  ),
  all(is.finite(tsne_points$x)),
  all(is.finite(tsne_points$y)),
  identical(sort(tsne_labels$word), sort(label_words))
)
```

```{r}
#| label: fig-word-tsne
#| fig-cap: "t-SNE projection shown as one panel per anchor word, using seed 7801. In each panel, faint dots are the other 289 plotted words, squares are that anchor's 10 nearest neighbours in the full 50-dimensional vectors, and the large triangle is the anchor word."
#| fig-alt: "Eight-panel t-SNE small multiple using seed 7801. Each panel is named for one anchor word. Every panel shows the other 289 projected words as faint dots, the anchor word as a larger triangle, and its 10 full-space nearest neighbours as squares; the preservation table reports how often those neighbours stay near each other in two dimensions."
#| fig-width: 8
#| fig-height: 6.5

ggplot(tsne_panel_points, aes(x = x, y = y)) +
  geom_point(
    data = tsne_panel_points |>
      filter(panel_role == "other plotted words"),
    color = "#3B5F8A",
    alpha = 0.22,
    size = 0.8
  ) +
  geom_point(
    data = tsne_panel_points |>
      filter(panel_role == "10 full-space neighbours"),
    color = "#8A5F3B",
    fill = "white",
    shape = 22,
    stroke = 0.7,
    size = 2.2
  ) +
  geom_point(
    data = tsne_panel_points |>
      filter(panel_role == "anchor word"),
    color = "#1F2D3D",
    fill = "white",
    shape = 24,
    stroke = 0.9,
    size = 3.6
  ) +
  facet_wrap(vars(anchor_word), ncol = 4) +
  labs(
    x = "t-SNE dimension 1",
    y = "t-SNE dimension 2"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  all(is.finite(tsne_points$x)),
  all(is.finite(tsne_points$y)),
  identical(levels(tsne_panel_points$anchor_word), label_words)
)
```

The panels are landmarks so the two pictures can be matched by eye. They are not
cluster names. Squares mark all 10 full-space neighbours; where they land shows
whether the projection kept them near the anchor. The table below lists every
anchor word.

## Measure neighbour preservation

A **10-nearest-neighbour preservation** score asks: for each word, what share of
its 10 nearest neighbours in the full 50-dimensional normalized vectors also
appear among its 10 nearest neighbours in the two-dimensional drawing? Chance is
10 divided by 299, or `r sprintf("%.1f%%", 100 * 10 / (nrow(normalized_embeddings) - 1))`.

```{r}
nearest_by_distance <- function(points, k = 10L) {
  distances <- as.matrix(dist(points))
  diag(distances) <- Inf
  t(apply(
    distances,
    1,
    \(values) names(sort(values))[seq_len(k)]
  ))
}

neighbour_overlap <- function(first, second) {
  mean(vapply(
    seq_len(nrow(first)),
    \(row) sum(first[row, ] %in% second[row, ]) / ncol(first),
    numeric(1)
  ))
}

pca_matrix <- pca_points |>
  select(x, y) |>
  as.matrix()
rownames(pca_matrix) <- pca_points$word

stopifnot(
  identical(rownames(full_space_neighbours), rownames(pca_matrix))
)

pca_neighbours <- nearest_by_distance(pca_matrix, 10L)
rownames(pca_neighbours) <- pca_points$word

tsne_matrix_7801 <- tsne_points |>
  select(x, y) |>
  as.matrix()
rownames(tsne_matrix_7801) <- tsne_points$word

stopifnot(
  identical(rownames(full_space_neighbours), rownames(tsne_matrix_7801))
)

tsne_neighbours_7801 <- nearest_by_distance(tsne_matrix_7801, 10L)
rownames(tsne_neighbours_7801) <- tsne_points$word

set.seed(7802)
tsne_seed_7802 <- Rtsne(
  normalized_embeddings,
  dims = 2,
  perplexity = tsne_perplexity,
  theta = tsne_theta,
  max_iter = tsne_iterations,
  pca = FALSE,
  normalize = FALSE,
  check_duplicates = FALSE,
  num_threads = 1,
  verbose = FALSE
)

tsne_points_7802 <- as_tibble(tsne_seed_7802$Y, .name_repair = "minimal") |>
  setNames(c("x", "y")) |>
  mutate(word = rownames(normalized_embeddings), .before = 1)

tsne_matrix_7802 <- tsne_points_7802 |>
  select(x, y) |>
  as.matrix()
rownames(tsne_matrix_7802) <- tsne_points_7802$word

stopifnot(
  identical(rownames(full_space_neighbours), rownames(tsne_matrix_7802)),
  identical(rownames(tsne_neighbours_7801), rownames(tsne_matrix_7802))
)

tsne_neighbours_7802 <- nearest_by_distance(tsne_matrix_7802, 10L)
rownames(tsne_neighbours_7802) <- tsne_points_7802$word

preservation_table <- tibble(
  comparison = c(
    "PCA versus full-space neighbours",
    "t-SNE seed 7801 versus full-space neighbours",
    "t-SNE seed 7802 versus full-space neighbours",
    "t-SNE seed 7801 versus t-SNE seed 7802",
    "chance level for a random 10-word set"
  ),
  preservation = c(
    neighbour_overlap(full_space_neighbours, pca_neighbours),
    neighbour_overlap(full_space_neighbours, tsne_neighbours_7801),
    neighbour_overlap(full_space_neighbours, tsne_neighbours_7802),
    neighbour_overlap(tsne_neighbours_7801, tsne_neighbours_7802),
    10 / (nrow(normalized_embeddings) - 1)
  )
)

display_preservation <- preservation_table |>
  mutate(preservation = sprintf("%.1f%%", 100 * preservation))

kable(
  display_preservation,
  col.names = c("Comparison", "Average 10-neighbour overlap"),
  caption = "Neighbour preservation for two-dimensional projections",
  row.names = FALSE
)
```

```{r}
#| include: false

pca_preservation <- preservation_table$preservation[1]
tsne_preservation_7801 <- preservation_table$preservation[2]
tsne_preservation_7802 <- preservation_table$preservation[3]
cross_seed_agreement <- preservation_table$preservation[4]
chance_preservation <- preservation_table$preservation[5]

stopifnot(
  identical(dim(full_space_neighbours), c(300L, 10L)),
  identical(dim(pca_neighbours), c(300L, 10L)),
  identical(dim(tsne_neighbours_7801), c(300L, 10L)),
  identical(dim(tsne_neighbours_7802), c(300L, 10L)),
  chance_preservation > 0.03,
  chance_preservation < 0.04,
  pca_preservation > chance_preservation * 4,
  tsne_preservation_7801 > pca_preservation * 2,
  tsne_preservation_7802 > pca_preservation * 2,
  cross_seed_agreement > tsne_preservation_7801,
  cross_seed_agreement < 0.90
)
```

The measured result is modest but useful. PCA keeps about
`r display_preservation$preservation[1]` of each word's 10 neighbours. The
plotted t-SNE run keeps about `r display_preservation$preservation[2]`, and the
second t-SNE seed keeps about `r display_preservation$preservation[3]`. The two
t-SNE runs agree with each other on about
`r display_preservation$preservation[4]` of neighbours. That is well above
chance, but it is not perfect stability.

## Inspect labelled words as rows

The table below is the text alternative for the labelled points. It reports
neighbours from the full vector space and from the two plotted projections,
without asking the reader to infer them from point positions.

```{r}
neighbour_table <- tibble(word = label_words) |>
  rowwise() |>
  mutate(
    full_space_neighbours = paste(full_space_neighbours[word, 1:5], collapse = ", "),
    pca_neighbours = paste(pca_neighbours[word, 1:5], collapse = ", "),
    tsne_neighbours = paste(tsne_neighbours_7801[word, 1:5], collapse = ", ")
  ) |>
  ungroup()

kable(
  neighbour_table,
  col.names = c(
    "Labelled word",
    "Nearest in 50 dimensions",
    "Nearest in PCA plot",
    "Nearest in t-SNE plot"
  ),
  caption = "Neighbour lists for the labelled words in the two projection figures",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(neighbour_table$word, label_words),
  all(str_count(neighbour_table$full_space_neighbours, ",") == 4L),
  all(str_count(neighbour_table$pca_neighbours, ",") == 4L),
  all(str_count(neighbour_table$tsne_neighbours, ",") == 4L),
  all(!str_detect(neighbour_table$full_space_neighbours, fixed("</s>")))
)
```

If a nearby label looks interesting, the next step is to inspect the underlying
texts or the full-space neighbour table. The picture alone is not evidence that
the words form a topic.

## What this does not show

The t-SNE plot does not give meaningful cluster sizes, gap sizes, group
positions, or axis directions. PCA directions are mathematical summaries, but
their signs can flip, and this plot still leaves most variance out.

The preservation scores are about the projection's fidelity to this small model,
not about fidelity to meaning. Small-corpus embeddings can change their
neighbour lists when trained again, which is why this model is pinned. The same
t-SNE seed can also draw a different layout on another computer because
low-level numerical details can differ.

## What to remember

- Each plotted point began as a 50-number word vector.
- Rows were L2-normalized before PCA and t-SNE so neighbours match cosine
  geometry.
- PCA's first two components explain about
  `r sprintf("%.1f%%", 100 * sum(pca_variance[1:2]))` of variance here.
- t-SNE preserves more local neighbours than PCA in this run, but two seeds
  still disagree on many neighbours.
- Do not interpret clusters, gaps, sizes, positions, or axis directions as
  findings.

Keep the scatterplot beside the preservation table. The table says what the
picture kept.

## Sources

- [Lesson 61, Training word vectors](../12_similarity/61-distributed-word-representations.qmd)
- [van der Maaten and Hinton on t-SNE](https://jmlr.org/papers/v9/vandermaaten08a.html)
- [Rtsne package documentation](https://cran.r-project.org/package=Rtsne)
- [Kobak and Berens on t-SNE art and analysis](https://doi.org/10.1038/s41467-019-13056-x)
- [Wattenberg, Viégas, and Johnson on how to use t-SNE effectively](https://doi.org/10.23915/distill.00002)
- [Antoniak and Mimno on stability of word embeddings](https://aclanthology.org/Q18-1008/)
- [R `prcomp` documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/prcomp.html)
