---
title: "Putting events on a timeline"
subtitle: "Keep document dates, event dates, and uncertain durations apart"
description: "Learn how to draw a small event timeline without turning coarse or missing dates into false points."
categories: [information visualization, timelines, temporal parsing]
---

Nadia is cleaning up a short program log. Some entries report what happened.
Others say what staff planned. A few lines mention a month or a season instead
of a single day.

**Events on a timeline** means placing event records on a time scale only as
precisely as the text allows. A day can be a point. A month, season, or course
length needs an interval. A duration with no start date stays off the chart.

This lesson uses a small constructed log so the dates can be checked by eye. It
builds on [the temporal parsing lesson](../6_entity_enrichment/31-temporal-parsing.qmd),
which explains reference dates for phrases such as "next Friday."

::: {.callout-note}
The Riverton program log in this lesson is a constructed teaching example about an invented service desk; the entries and normalized dates were written by this lesson's author.
:::

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- keep the entry date separate from the event date;
- resolve relative dates from each entry's own date;
- draw seasons, months, and course lengths as intervals;
- mark planned and reported events without relying on color alone;
- keep undated events in a table instead of guessing; and
- compare narrative order with event order.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Build a small event log

An entry date is the date of the note. It is the anchor for relative language,
not automatically the date of the event. The `entry_id` stays with every row so
the timeline can be traced back to the source sentence.

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(purrr)
library(stringr)
library(tibble)
library(tidyr)

program_entries <- tribble(
  ~entry_id, ~entry_date, ~entry_text,
  "R01", "2026-02-20",
  "The spring outreach window will run this spring.",
  "R02", "2026-03-03",
  "The orientation was held on March 2, and registration closes on March 15.",
  "R03", "2026-08-28",
  "The 12-week course starts on September 14.",
  "R04", "2026-09-20",
  "The reminder says the applicant check-in happened last Monday.",
  "R05", "2026-10-05",
  "The pilot lasted six weeks.",
  "R06", "2026-10-10",
  "The October review meeting is planned for October 22.",
  "R07", "2026-11-01",
  "The November report covers work completed in September."
) |>
  mutate(entry_date = as.Date(entry_date))

kable(
  program_entries,
  col.names = c("Entry ID", "Entry date", "Constructed log entry"),
  caption = "Constructed Riverton program log entries",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c("dplyr", "ggplot2", "knitr", "purrr", "stringr", "tibble", "tidyr") %in%
    loadedNamespaces()),
  identical(nrow(program_entries), 7L),
  identical(program_entries$entry_id, sprintf("R%02d", 1:7)),
  all(!is.na(program_entries$entry_date)),
  !any(str_detect(program_entries$entry_text, regex("Sys[.]Date", ignore_case = TRUE)))
)
```

The second entry has two events and two dates. The orientation belongs to March
2. The registration closing belongs to March 15. Reading only the nearest date
would fail on many real sentences, so the attachment is recorded by hand here.

## Normalize only what the text supports

The normalization table names the event trigger, the time expression, and the
rule used to compute dates. The season rule is a local convention: this page uses
Northern Hemisphere meteorological seasons, so spring runs from March 1 through
May 31.

```{r}
season_bounds <- tribble(
  ~season, ~start_month, ~start_day, ~end_month, ~end_day,
  "spring", 3L, 1L, 5L, 31L,
  "summer", 6L, 1L, 8L, 31L,
  "autumn", 9L, 1L, 11L, 30L,
  "winter", 12L, 1L, 2L, 28L
)

event_plan <- tribble(
  ~event_id, ~entry_id, ~trigger, ~time_text, ~granularity, ~status, ~rule,
  "EV01", "R01", "outreach window", "this spring", "season", "planned",
  "entry-year spring",
  "EV02", "R02", "orientation held", "March 2", "day", "reported",
  "entry-year month day",
  "EV03", "R02", "registration closes", "March 15", "day", "planned",
  "entry-year month day",
  "EV04", "R03", "course starts", "September 14 for 12 weeks", "duration",
  "planned", "entry-year month day plus weeks",
  "EV05", "R04", "applicant check-in", "last Monday", "day", "reported",
  "previous weekday from entry date",
  "EV06", "R05", "pilot lasted", "six weeks", "duration", "reported",
  "duration without start",
  "EV07", "R06", "review meeting", "October 22", "day", "planned",
  "entry-year month day",
  "EV08", "R07", "September work", "September", "month", "reported",
  "entry-year month"
)

make_date <- function(year, month, day) {
  as.Date(sprintf("%04d-%02d-%02d", year, month, day))
}

parse_month_day <- function(entry_date, time_text) {
  parts <- str_match(time_text, "\\b([A-Z][a-z]+)\\s+(\\d{1,2})\\b")
  make_date(
    as.integer(format(entry_date, "%Y")),
    match(parts[, 2], month.name),
    as.integer(parts[, 3])
  )
}

parse_month_interval <- function(entry_date, time_text) {
  month_number <- match(time_text, month.name)
  start <- make_date(as.integer(format(entry_date, "%Y")), month_number, 1L)
  end <- seq(start, by = "1 month", length.out = 2L)[2] - 1L
  list(start = start, end = end)
}

previous_weekday <- function(entry_date, weekday) {
  weekday_names <- c(
    "Sunday", "Monday", "Tuesday", "Wednesday",
    "Thursday", "Friday", "Saturday"
  )
  current_day <- as.POSIXlt(entry_date)$wday + 1L
  target_day <- match(weekday, weekday_names)
  days_back <- (current_day - target_day) %% 7L
  if (days_back == 0L) {
    days_back <- 7L
  }
  entry_date - days_back
}

normalize_event <- function(rule, entry_date, time_text) {
  entry_year <- as.integer(format(entry_date, "%Y"))

  if (rule == "entry-year spring") {
    bounds <- season_bounds |>
      filter(season == "spring")
    return(list(
      start = make_date(entry_year, bounds$start_month, bounds$start_day),
      end = make_date(entry_year, bounds$end_month, bounds$end_day),
      placed = TRUE,
      reason_unplaced = NA_character_
    ))
  }

  if (rule == "entry-year month day") {
    start <- parse_month_day(entry_date, time_text)
    return(list(
      start = start,
      end = start,
      placed = TRUE,
      reason_unplaced = NA_character_
    ))
  }

  if (rule == "entry-year month day plus weeks") {
    start <- parse_month_day(entry_date, time_text)
    return(list(
      start = start,
      end = start + 12L * 7L - 1L,
      placed = TRUE,
      reason_unplaced = NA_character_
    ))
  }

  if (rule == "previous weekday from entry date") {
    start <- previous_weekday(entry_date, "Monday")
    return(list(
      start = start,
      end = start,
      placed = TRUE,
      reason_unplaced = NA_character_
    ))
  }

  if (rule == "entry-year month") {
    interval <- parse_month_interval(entry_date, time_text)
    return(list(
      start = interval$start,
      end = interval$end,
      placed = TRUE,
      reason_unplaced = NA_character_
    ))
  }

  list(
    start = as.Date(NA),
    end = as.Date(NA),
    placed = FALSE,
    reason_unplaced = "duration has no start date"
  )
}

event_rows <- event_plan |>
  left_join(program_entries, by = join_by(entry_id)) |>
  mutate(
    normalized = pmap(
      list(rule, entry_date, time_text),
      normalize_event
    ),
    start = as.Date(map_chr(normalized, \(item) as.character(item$start))),
    end = as.Date(map_chr(normalized, \(item) as.character(item$end))),
    placed = map_lgl(normalized, "placed"),
    reason_unplaced = map_chr(
      normalized,
      \(item) {
        if (is.na(item$reason_unplaced)) {
          NA_character_
        } else {
          item$reason_unplaced
        }
      }
    ),
    narrative_order = match(entry_id, program_entries$entry_id)
  ) |>
  select(
    event_id, entry_id, narrative_order, entry_date, trigger, time_text,
    granularity, status, start, end, placed, reason_unplaced
  )

kable(
  event_rows,
  col.names = c(
    "Event ID", "Entry ID", "Narrative order", "Entry date", "Event trigger",
    "Time text", "Granularity", "Status", "Start", "End", "Placed",
    "Reason unplaced"
  ),
  caption = "Normalized event records with separate document dates and event intervals",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(event_rows), 8L),
  identical(sum(event_rows$entry_id == "R02"), 2L),
  identical(event_rows$start[event_rows$event_id == "EV01"], as.Date("2026-03-01")),
  identical(event_rows$end[event_rows$event_id == "EV01"], as.Date("2026-05-31")),
  identical(event_rows$start[event_rows$event_id == "EV02"], as.Date("2026-03-02")),
  identical(event_rows$start[event_rows$event_id == "EV03"], as.Date("2026-03-15")),
  identical(event_rows$start[event_rows$event_id == "EV04"], as.Date("2026-09-14")),
  identical(event_rows$end[event_rows$event_id == "EV04"], as.Date("2026-12-06")),
  identical(event_rows$start[event_rows$event_id == "EV05"], as.Date("2026-09-14")),
  identical(event_rows$start[event_rows$event_id == "EV08"], as.Date("2026-09-01")),
  identical(event_rows$end[event_rows$event_id == "EV08"], as.Date("2026-09-30")),
  identical(sum(event_rows$placed), 7L),
  identical(sum(!event_rows$placed), 1L),
  !any(is.na(event_rows$entry_date))
)
```

Seven events receive dates. One duration does not. The `start` and `end` columns
make the difference visible: a day has the same start and end, while spring, a
month, and a 12-week course span several days.

## Keep unplaced events visible

Undated events should not disappear from the lesson. They also should not reach
`ggplot2`, because missing dates would either create a warning or invite a false
point.

```{r}
unplaced_events <- event_rows |>
  filter(!placed) |>
  select(event_id, entry_id, entry_date, trigger, time_text, reason_unplaced)

kable(
  unplaced_events,
  col.names = c(
    "Event ID", "Entry ID", "Entry date", "Event trigger",
    "Time text", "Reason not drawn"
  ),
  caption = "Unplaced events kept out of the timeline",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(unplaced_events), 1L),
  identical(unplaced_events$event_id, "EV06"),
  identical(unplaced_events$reason_unplaced, "duration has no start date")
)
```

The "pilot lasted six weeks" row gives a length but no position. Nadia can keep
it in the review table, but drawing it would be a guess.

## Draw points only for days

The timeline below filters to placed events first. It uses segments for intervals
and points for day-level dates. Shape and line type mark status, so the planned
and reported distinction remains readable without color.

```{r}
#| label: fig-event-timeline
#| fig-cap: "Timeline of seven placed events from a constructed Riverton log in 2026; intervals show a season, a month, and a 12-week course, while four day-level events are points."
#| fig-alt: "Timeline with seven placed events in chronological order. Three events are intervals: spring outreach from March 1 to May 31 2026, September work from September 1 to September 30 2026, and a 12-week course from September 14 to December 6 2026. Four day-level events are points on March 2, March 15, September 14, and October 22 2026. Each row label includes the event order and source entry number. The six-week pilot duration without a start date is left off the chart."
#| fig-width: 8
#| fig-height: 5.5

placed_events <- event_rows |>
  filter(placed) |>
  arrange(start, end, event_id) |>
  mutate(
    event_order = row_number(),
    y_position = -event_order,
    event_label = sprintf("%s. entry %s: %s", event_order, narrative_order, trigger)
  )

interval_events <- placed_events |>
  filter(granularity != "day")

day_events <- placed_events |>
  filter(granularity == "day")

ggplot() +
  geom_segment(
    data = interval_events,
    aes(
      x = start, xend = end,
      y = y_position, yend = y_position,
      linetype = status
    ),
    linewidth = 2,
    color = "grey35"
  ) +
  geom_point(
    data = day_events,
    aes(x = start, y = y_position, shape = status),
    size = 3,
    color = "black"
  ) +
  scale_linetype_manual(values = c(planned = "22", reported = "solid")) +
  scale_shape_manual(values = c(planned = 17, reported = 16)) +
  scale_x_date(
    date_breaks = "2 months",
    date_labels = "%b %Y",
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  scale_y_continuous(
    breaks = placed_events$y_position,
    labels = placed_events$event_label
  ) +
  labs(
    x = "Event date or interval in 2026",
    y = "Chronological event order with source entry",
    shape = "Day-level status",
    linetype = "Interval status"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  identical(nrow(placed_events), 7L),
  identical(nrow(interval_events), 3L),
  identical(nrow(day_events), 4L),
  !any(is.na(placed_events$start)),
  !any(is.na(placed_events$end)),
  all(interval_events$start < interval_events$end),
  all(day_events$start == day_events$end),
  which(placed_events$event_id == "EV08") < which(placed_events$event_id == "EV07"),
  identical(
    placed_events$event_order,
    seq_len(nrow(placed_events))
  )
)
```

The y-axis is chronological event order, sorted by the resolved start date. Each
row label also includes the source entry number, which shows narrative order.
The November report is labeled `entry 7`, but its September work interval belongs
before the October review meeting. That difference is why a timeline should not
treat row order as time.

The event table is the long description for the figure. It gives the exact dates,
status, and entry ID for each mark.

```{r}
timeline_long_description <- placed_events |>
  transmute(
    event_order,
    narrative_order,
    event_id,
    entry_id,
    trigger,
    time_text,
    granularity,
    status,
    start,
    end
  )

kable(
  timeline_long_description,
  col.names = c(
    "Event order", "Narrative order", "Event ID", "Entry ID",
    "Event trigger", "Time text", "Granularity", "Status", "Start", "End"
  ),
  caption = "Text version of the timeline data",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(timeline_long_description$event_order, seq_len(7L)),
  identical(timeline_long_description$narrative_order[1:3], c(1L, 2L, 2L)),
  identical(
    timeline_long_description$granularity,
    c("season", "day", "day", "month", "day", "duration", "day")
  )
)
```

## Where this fits

TimeML and related standards separate events, time expressions, and the
document's creation time. This page does the same thing in a small table. It
does not train or evaluate a temporal parser, and the constructed normalizations
are not independent labels.

A real-text coda shows why the table matters. In the inaugural corpus,
Lincoln's 1865 phrase "four years ago" can be anchored only to a speech year if
the source records no month or day. The output should keep that year-level
granularity rather than inventing a day.

## What to remember

- The entry date anchors relative phrases, but it is not the event date.
- A season or month is an interval under a stated convention.
- A day-level event can be a point; a coarse event should not be.
- Planned events need a visible status cue.
- A duration without a start date belongs in the unplaced table, not on the
  chart.

Nadia's timeline is useful because it refuses one false mark. The six-week pilot
is still visible, but it stays off the time scale until a start date appears.

## Sources

- [Pustejovsky et al. on TimeML event and temporal expression representation](https://timeml.github.io/site/publications/timeMLpubs/IWCS-v4.pdf)
- [Library of Congress Extended Date/Time Format specification](https://www.loc.gov/standards/datetime/)
- [UzZaman et al. on TempEval-3](https://aclanthology.org/S13-2001/)
- [Brehmer et al. on timeline design](https://doi.org/10.1109/TVCG.2016.2614803)
- [R Date documentation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Dates.html)
