---
title: "Mapping place mentions"
subtitle: "Review state-name matches before drawing them"
description: "Learn how naive place-name matches can mislead a map, and how reviewed readings keep non-place meanings visible."
categories: [information visualization, maps, place names]
---

Ben wants a map from presidential speeches, but the first name he checks is
`Washington`. In these speeches it often means a person, a city, or the federal
government. It does not mean Washington State.

**Locations on a map** means more than finding a name. The reader has to decide
whether the name really refers to a place, what place it means, what unit is
being counted, and whether the basemap can honestly show it.

This lesson uses U.S. inaugural-address paragraphs and the state names bundled
with R. The review rules and labels below are the author's judgement after
reading the matches in this corpus. They are not an independent geoparser.

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- match U.S. state names with case-sensitive, whole-word rules;
- compare naive matches with reviewed readings;
- keep person, river, event, government, and pre-statehood readings off the map;
- count speeches with at least one reviewed state reading;
- draw symbols with area scaling instead of a choropleth; and
- state a basemap's coverage and projection limits.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Match state names before reviewing them

The source data is the shared inaugural-address paragraph table used in earlier
lessons. A paragraph and speech ID travel with every match. The matcher sorts
state names from longest to shortest so `West Virginia` would be selected before
`Virginia` when both could overlap.

```{r}
library(dplyr)
library(ggplot2)
library(knitr)
library(purrr)
library(stringr)
library(tibble)
library(tidyr)

source("R/inaugural-corpus.R")

paragraphs <- inaugural_paragraphs()

state_lookup <- tibble(state = datasets::state.name) |>
  arrange(desc(str_length(state)), state)

empty_matches <- function() {
  tibble(
    state = character(),
    start = integer(),
    end = integer(),
    mention = character()
  )
}

state_pattern <- function(state_name) {
  regex(
    paste0("\\b", str_replace_all(state_name, " ", "\\\\s+"), "\\b"),
    ignore_case = FALSE
  )
}

find_all_state_candidates <- function(text) {
  state_lookup$state |>
    map_dfr(\(state_name) {
      locations <- str_locate_all(text, state_pattern(state_name))[[1]]
      if (nrow(locations) == 0L) {
        return(empty_matches())
      }

      tibble(
        state = state_name,
        start = as.integer(locations[, 1]),
        end = as.integer(locations[, 2]),
        mention = str_sub(text, start, end)
      )
    }) |>
    arrange(start, desc(end - start), state)
}

choose_longest_nonoverlap <- function(candidates, text_length) {
  if (nrow(candidates) == 0L) {
    return(candidates)
  }

  kept <- rep(FALSE, nrow(candidates))
  used_positions <- rep(FALSE, text_length)

  for (row_number in seq_len(nrow(candidates))) {
    span <- candidates$start[row_number]:candidates$end[row_number]
    if (!any(used_positions[span])) {
      kept[row_number] <- TRUE
      used_positions[span] <- TRUE
    }
  }

  candidates[kept, ]
}

find_state_matches <- function(text) {
  find_all_state_candidates(text) |>
    choose_longest_nonoverlap(nchar(text))
}

state_matches <- paragraphs |>
  mutate(matches = map(paragraph, find_state_matches)) |>
  filter(map_int(matches, nrow) > 0L) |>
  select(
    paragraph_id, speech_id, year, president, paragraph_words,
    paragraph, matches
  ) |>
  unnest(matches) |>
  mutate(
    match_id = row_number(),
    context = str_squish(str_sub(
      paragraph,
      pmax(1L, start - 70L),
      pmin(nchar(paragraph), end + 70L)
    ))
  )

naive_state_counts <- state_matches |>
  summarise(
    naive_matches = n(),
    naive_speeches = n_distinct(speech_id),
    .by = state
  ) |>
  arrange(desc(naive_matches), state)

kable(
  naive_state_counts,
  col.names = c("Matched name", "Naive matches", "Speeches with a match"),
  caption = "Naive case-sensitive state-name matches before review",
  row.names = FALSE
)
```

```{r}
#| include: false

inside_longer_state <- function(text, start, end, state) {
  longer_states <- state_lookup |>
    filter(str_length(state) > str_length(.env$state)) |>
    pull(state)

  longer_states |>
    map_lgl(\(longer_state) {
      locations <- str_locate_all(text, state_pattern(longer_state))[[1]]
      if (nrow(locations) == 0L) {
        return(FALSE)
      }
      any(locations[, 1] <= start & locations[, 2] >= end)
    }) |>
    any()
}

synthetic_overlap_check <- find_state_matches(
  "West Virginia appears before Virginia in this check."
)

stopifnot(
  all(c("dplyr", "ggplot2", "knitr", "purrr", "stringr", "tibble", "tidyr") %in%
    loadedNamespaces()),
  !"package:maps" %in% search(),
  identical(nrow(paragraphs), 1377L),
  identical(nrow(state_matches), 67L),
  identical(n_distinct(state_matches$state), 18L),
  identical(n_distinct(state_matches$speech_id), 24L),
  identical(synthetic_overlap_check$state, c("West Virginia", "Virginia")),
  !any(pmap_lgl(
    list(
      state_matches$paragraph, state_matches$start,
      state_matches$end, state_matches$state
    ),
    inside_longer_state
  ))
)
```

The naive table counts strings. It does not know whether `Washington` is a
person, the city, or the state. That is the next step.

## Review what each match means

The categories below were written after reading the 67 matches. Any residue that
the rules do not explain is assigned to `uncertain`, not to the map.

```{r}
category_table <- tribble(
  ~category, ~meaning,
  "person", "the state name is part of a person's name or title",
  "city or federal government", "the name refers to Washington, DC, New York City, or federal power",
  "event name", "the name belongs to a named meeting or conference",
  "river", "the name refers to a river or river valley",
  "territory or republic before statehood", "the text refers to land before it became a state",
  "law or organisation name", "the name appears inside a law, congress, or organisation label",
  "state or its people", "the text refers to the state, its residents, or someone from that state",
  "uncertain", "the author could not assign a safer category after reading the context"
)

review_matches <- state_matches |>
  mutate(
    reading_category = case_when(
      match_id %in% c(
        24L, 26L, 31L, 33L, 34L, 35L, 38L, 43L, 45L,
        49L, 53L, 55L, 57L, 58L, 59L, 60L
      ) ~ "person",
      match_id %in% c(36L, 37L, 44L, 51L, 62L, 66L, 67L) ~
        "city or federal government",
      match_id == 48L ~ "event name",
      match_id %in% c(2L, 4L, 6L, 7L, 9L, 22L, 23L, 46L, 47L) ~ "river",
      match_id %in% c(
        1L, 3L, 5L, 8L, 10L, 11L, 12L, 13L, 14L, 15L,
        16L, 17L, 18L, 19L, 20L, 21L, 27L, 28L, 32L, 41L, 42L
      ) ~ "territory or republic before statehood",
      match_id %in% c(29L, 52L) ~ "law or organisation name",
      match_id %in% c(25L, 30L, 39L, 40L, 50L, 54L, 56L, 61L, 63L, 64L, 65L) ~
        "state or its people",
      TRUE ~ "uncertain"
    ),
    reviewed_state_reading = reading_category == "state or its people"
  )

category_counts <- category_table |>
  left_join(
    review_matches |>
      count(reading_category, name = "matches"),
    by = join_by(category == reading_category)
  ) |>
  mutate(matches = coalesce(matches, 0L))

kable(
  category_counts,
  col.names = c("Review category", "How this lesson uses the category", "Matches"),
  caption = "Author-written review categories for state-name matches",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(category_table), 8L),
  identical(sum(category_counts$matches), 67L),
  identical(sum(review_matches$reviewed_state_reading), 11L),
  identical(category_counts$matches[category_counts$category == "uncertain"], 0L),
  all(review_matches$reading_category %in% category_table$category)
)
```

The category table is a disclosure, not a score. These labels were shaped by
this corpus. A new collection would need a new review.

## Compare naive and reviewed readings

The table below keeps the naive counts beside the reviewed counts. The reviewed
unit is speeches with at least one state reading for a state. Repeated mentions
inside one speech do not add more weight.

```{r}
reviewed_state_counts <- review_matches |>
  filter(reviewed_state_reading) |>
  distinct(state, speech_id) |>
  count(state, name = "reviewed_state_speeches") |>
  arrange(desc(reviewed_state_speeches), state)

naive_review_comparison <- naive_state_counts |>
  left_join(reviewed_state_counts, by = join_by(state)) |>
  mutate(reviewed_state_speeches = coalesce(reviewed_state_speeches, 0L)) |>
  arrange(desc(naive_matches), state)

washington_summary <- review_matches |>
  filter(state == "Washington") |>
  summarise(
    matches = n(),
    speeches = n_distinct(speech_id),
    reviewed_state_speeches = n_distinct(speech_id[reviewed_state_reading])
  )

river_summary <- review_matches |>
  filter(state %in% c("Mississippi", "Ohio", "Missouri")) |>
  summarise(
    matches = n(),
    speeches = n_distinct(speech_id),
    categories = str_c(sort(unique(reading_category)), collapse = ", "),
    .by = state
  ) |>
  arrange(state)

kable(
  naive_review_comparison,
  col.names = c(
    "Matched name", "Naive matches", "Naive speeches",
    "Reviewed state speeches"
  ),
  caption = "Naive matches compared with reviewed state readings",
  row.names = FALSE
)

kable(
  river_summary,
  col.names = c("Matched name", "Matches", "Speeches", "Reviewed category"),
  caption = "River readings checked before mapping",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(washington_summary$matches, 22L),
  identical(washington_summary$speeches, 17L),
  identical(washington_summary$reviewed_state_speeches, 0L),
  identical(river_summary$state, c("Mississippi", "Missouri", "Ohio")),
  identical(river_summary$categories, rep("river", 3L)),
  identical(
    reviewed_state_counts$state,
    c(
      "Virginia", "North Carolina", "Pennsylvania", "California",
      "Iowa", "Louisiana", "Nebraska"
    )
  ),
  identical(reviewed_state_counts$reviewed_state_speeches, c(3L, 2L, 2L, 1L, 1L, 1L, 1L))
)
```

The code finds `Washington` in 17 speeches and the review counts none as
Washington State. It also finds `Mississippi`, `Ohio`, and `Missouri` as river
readings. Those names stay in the review tables and stay off the map.

The full review table is long because every non-state reading remains visible.

```{r}
review_table <- review_matches |>
  transmute(
    match_id,
    state,
    speech_id,
    year,
    president,
    reading_category,
    reviewed_state_reading,
    context
  ) |>
  arrange(match_id)

kable(
  review_table,
  col.names = c(
    "Match ID", "Matched name", "Speech ID", "Year", "President",
    "Reviewed category", "Mapped as state", "Reading context"
  ),
  caption = "All reviewed state-name matches with author-written categories",
  row.names = FALSE,
  format = "html",
  escape = TRUE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(review_table), 67L),
  identical(review_table$match_id, seq_len(67L)),
  identical(sum(review_table$reviewed_state_reading), 11L),
  !any(is.na(review_table$reading_category)),
  !any(is.na(review_table$context))
)
```

## Draw the reviewed state counts

The basemap comes from `ggplot2::map_data("state")`, which uses the `maps`
package state database. That database covers the 48 contiguous states and the
District of Columbia. It does not draw Alaska or Hawaii, and this lesson does
not plot their `state.center` points.

```{r}
lower48_map <- ggplot2::map_data("state")

state_centers <- tibble(
  state = datasets::state.name,
  longitude = datasets::state.center$x,
  latitude = datasets::state.center$y
) |>
  mutate(
    lower48_or_dc = !state %in% c("Alaska", "Hawaii"),
    region = str_to_lower(state)
  )

map_points <- reviewed_state_counts |>
  left_join(state_centers, by = join_by(state)) |>
  filter(lower48_or_dc)

state_count_table <- map_points |>
  select(state, reviewed_state_speeches, longitude, latitude) |>
  arrange(desc(reviewed_state_speeches), state)

kable(
  state_count_table,
  col.names = c(
    "State", "Speeches with reviewed state reading",
    "Approximate center longitude", "Approximate center latitude"
  ),
  caption = "Long description for the reviewed state-count map",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(n_distinct(lower48_map$region), 49L),
  identical(nrow(map_points), 7L),
  !any(map_points$state %in% c("Alaska", "Hawaii")),
  identical(
    state_centers$lower48_or_dc[state_centers$state %in% c("Alaska", "Hawaii")],
    c(FALSE, FALSE)
  ),
  identical(max(map_points$reviewed_state_speeches), 3L),
  identical(sum(map_points$reviewed_state_speeches), 11L),
  !any(str_detect(state_matches$paragraph, fixed("Riverton")))
)
```

The coordinates are approximate state centers. A symbol at a center is a count
marker, not the place where a speech event happened. Pre-statehood references
such as the Republic of Texas were not counted as state readings.

```{r}
#| label: fig-state-reading-map
#| fig-cap: "Reviewed state readings in inaugural speeches on a lower-48 basemap; seven states remain after review."
#| fig-alt: "Lower-48 U.S. map with proportional circles for seven reviewed state readings. Virginia appears in 3 speeches; North Carolina and Pennsylvania appear in 2 each; California, Iowa, Louisiana, and Nebraska appear in 1 each. Washington State is not plotted because its 17 speech matches were reviewed as non-state readings."
#| fig-width: 8
#| fig-height: 5

mean_latitude <- mean(lower48_map$lat, na.rm = TRUE)
fixed_ratio <- 1 / cos(mean_latitude * pi / 180)
legend_breaks <- seq_len(max(map_points$reviewed_state_speeches))

ggplot() +
  geom_polygon(
    data = lower48_map,
    aes(x = long, y = lat, group = group),
    fill = "grey95",
    color = "grey65",
    linewidth = 0.2
  ) +
  geom_point(
    data = map_points,
    aes(x = longitude, y = latitude, size = reviewed_state_speeches),
    shape = 21,
    fill = "white",
    color = "black",
    stroke = 0.8
  ) +
  scale_size_area(
    breaks = legend_breaks,
    max_size = 10,
    name = "Speeches"
  ) +
  coord_fixed(ratio = fixed_ratio) +
  labs(
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_void() +
  theme(legend.position = "right")
```

```{r}
#| include: false

stopifnot(
  identical(legend_breaks, 1:3),
  fixed_ratio > 1,
  fixed_ratio < 2,
  identical(nrow(map_points), nrow(state_count_table)),
  setequal(map_points$state, state_count_table$state),
  !any(map_points$state %in% c("Alaska", "Hawaii")),
  all(map_points$reviewed_state_speeches %in% legend_breaks)
)
```

The coordinate ratio is `1 / cos(mean latitude in radians)`. It stretches the
longitude and latitude plot so one degree east-west and one degree north-south
look similar near the middle of the country. It does not preserve area. The
circle areas encode the speech counts; exact values belong in the table.

## Where this fits

Lesson 30 shows how a gazetteer can turn a place name into coordinates. This
page begins earlier. It asks whether a matched name should become a place record
at all. Names can refer to people, laws, rivers, cities, events, governments, or
places from an earlier political period.

The lower-48 map also has coverage limits. If a reviewed state reading pointed
to Alaska or Hawaii, this page would keep it in the table and not draw it on
this basemap.

## What to remember

- A state-name match is not automatically a state reference.
- Review labels shaped after reading the matches must be disclosed.
- Count speeches, not repeated mentions, when the unit is speeches.
- Rivers, people, events, and pre-statehood references stay off the reviewed
  map.
- `scale_size_area()` maps values to circle area; the table carries exact
  counts.
- A lower-48 basemap cannot show every U.S. state.

Ben can map seven reviewed states, but only after most naive matches have been
kept out of the symbol layer. The table is part of the map because it shows what
the picture refused to draw.

## Sources

- [Gritta et al. on geoparsing limits](https://doi.org/10.1007/s10579-017-9385-8)
- [Markert and Nissim on location metonymy](https://aclanthology.org/S07-1007/)
- [R datasets documentation for state names and centers](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/state.html)
- [ggplot2 size scales](https://ggplot2.tidyverse.org/reference/scale_size.html)
- [ggplot2 map data documentation](https://ggplot2.tidyverse.org/reference/map_data.html)
- [maps package on CRAN](https://cran.r-project.org/package=maps)
