---
title: "Drawing a knowledge graph"
subtitle: "Use IDs, arrows, labels, and tables before reading a layout"
description: "Learn how to draw a small knowledge graph in R while keeping identifiers, edge labels, layout limits, and text alternatives visible."
categories: [information visualization, knowledge graph visualization, networks]
---

Chloe has a tidy table of Riverton facts, but the table is hard to scan when
she wants to follow a path. Which organisation offers the credential? Which
deadline values did the knowledge base keep for review? Which place label might
cause trouble if IDs were ignored?

**Knowledge graph visualization** draws entities as nodes and typed relations as
edges. In this lesson, the edges are directed and labelled because the relation
itself is part of the data.

A drawing can help Chloe see a path. It can also mislead her if she treats the
layout as a measurement. The table of edges remains the source of the graph.

::: {.callout-note}
The Riverton entities and facts in this lesson are fictional teaching records rebuilt from the earlier knowledge base example.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- key graph nodes by entity ID instead of display label;
- show how label-keyed data wrongly merges two Riverton places;
- draw directed, labelled edges with igraph and ggplot2;
- explain what a force-directed layout can and cannot show;
- give a captioned edge table as the text alternative; and
- compare a co-word network with a typed knowledge graph.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Rebuild the graph facts by ID

The graph starts with the populated fact edges named in the knowledge-base
lesson. The two `Riverton` places have different IDs. That difference is the
first thing the drawing must preserve.

```{r}
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(readr)
library(igraph)
library(ggplot2)
library(tidytext)
library(knitr)

source("R/inaugural-corpus.R")

nodes <- tribble(
  ~node_id, ~label, ~node_type,
  "ORG-0002", "Riverton Skills Centre", "organisation",
  "ORG-0003", "Marrow County Transit", "organisation",
  "LOC-0001", "Riverton", "place",
  "LOC-0002", "Riverton", "place",
  "LOC-0003", "Bellhaven", "place",
  "CRD-0002", "Data Support Certificate", "credential",
  "VALUE-October 15", "October 15", "date literal",
  "VALUE-November 1", "November 1", "date literal"
)

edges <- tribble(
  ~from, ~to, ~relation, ~object_value, ~document_ids, ~source,
  "ORG-0002", "CRD-0002", "offers", "CRD-0002", "A001; A002", "populated fact",
  "ORG-0002", "VALUE-October 15", "application_deadline", "October 15", "A003", "populated fact",
  "ORG-0002", "VALUE-November 1", "application_deadline", "November 1", "A004", "populated fact"
)

edge_list <- edges |>
  left_join(
    nodes |> select(from = node_id, from_label = label, from_type = node_type),
    by = "from"
  ) |>
  left_join(
    nodes |> select(to = node_id, to_label = label, to_type = node_type),
    by = "to"
  ) |>
  transmute(
    from_id = from,
    from_label,
    from_type,
    relation = str_replace_all(relation, "_", " "),
    object_value,
    document_ids,
    to_id = to,
    to_label,
    to_type,
    source
  )

knitr::kable(
  edge_list,
  col.names = c(
    "From ID", "From label", "From type", "Relation",
    "Object value", "Document IDs", "To ID", "To label", "To type", "Source"
  ),
  caption = "Every edge drawn in the Riverton knowledge graph",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(nodes), 8L),
  identical(nrow(edges), 3L),
  all(edges$from %in% nodes$node_id),
  all(edges$to %in% nodes$node_id),
  all(edges$source == "populated fact"),
  all(edges$document_ids %in% c("A001; A002", "A003", "A004")),
  !any(str_detect(nodes$label, "2026")),
  sum(nodes$label == "Riverton") == 2L,
  all(c("LOC-0001", "LOC-0002") %in% nodes$node_id)
)
```

The table is the long description of the graph. It lists every drawn edge with
IDs and labels, so the relation can be checked without seeing the picture. The
two deadline values are both kept because the knowledge base flagged them as a
dated conflict rather than choosing one automatically; no year is supplied in
these graph labels because the announcements say only `October 15` and
`November 1`.

## Show the label-keyed mistake

If the graph used labels as node keys, both `Riverton` records would collapse
into one node. That would create a false path through a single place label.

```{r}
id_graph <- igraph::graph_from_data_frame(
  d = edges,
  directed = TRUE,
  vertices = nodes |> select(name = node_id, label, node_type)
)

label_edges <- edges |>
  left_join(nodes |> select(from = node_id, from_label = label), by = "from") |>
  left_join(nodes |> select(to = node_id, to_label = label), by = "to") |>
  transmute(from = from_label, to = to_label, relation)

label_nodes <- nodes |>
  distinct(name = label, node_type)

label_keyed_graph <- igraph::graph_from_data_frame(
  d = label_edges,
  directed = TRUE,
  vertices = label_nodes
)

merge_check <- tibble(
  key_used = c("entity ID", "display label"),
  vertices = c(igraph::vcount(id_graph), igraph::vcount(label_keyed_graph)),
  riverton_vertices = c(
    sum(igraph::V(id_graph)$label == "Riverton"),
    sum(igraph::V(label_keyed_graph)$name == "Riverton")
  )
)

knitr::kable(
  merge_check,
  col.names = c("Key used", "Vertices", "Vertices displayed as Riverton"),
  caption = "Using labels as keys wrongly merges the two Riverton places",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  igraph::vcount(id_graph) == 8L,
  igraph::vcount(label_keyed_graph) == 7L,
  merge_check$vertices[merge_check$key_used == "entity ID"] >
    merge_check$vertices[merge_check$key_used == "display label"],
  merge_check$riverton_vertices[merge_check$key_used == "entity ID"] == 2L,
  merge_check$riverton_vertices[merge_check$key_used == "display label"] == 1L
)
```

The ID-keyed graph has eight vertices. The label-keyed graph has seven because
the two places named `Riverton` have been merged.

```{r}
name_clash <- nodes |>
  count(label, name = "ids_with_label") |>
  filter(ids_with_label > 1L) |>
  left_join(
    nodes |>
      group_by(label) |>
      summarise(
        entity_ids = paste(node_id, collapse = "; "),
        entity_types = paste(node_type, collapse = "; "),
        .groups = "drop"
      ),
    by = "label"
  )

knitr::kable(
  name_clash,
  col.names = c("Shared label", "IDs with label", "Entity IDs", "Entity types"),
  caption = "The name clash is a label problem, not a factual graph edge",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(name_clash$label, "Riverton"),
  identical(name_clash$ids_with_label, 2L),
  str_detect(name_clash$entity_ids, "LOC-0001"),
  str_detect(name_clash$entity_ids, "LOC-0002")
)
```

## Place nodes by hand and draw arrows

The main figure uses fixed deterministic coordinates chosen so the labels are
readable. That keeps the drawing stable while Chloe reads the arrows and labels.
Up, down, left, right, rotation, and mirror image still do not carry data. Count
links in the table if you need a number. Viewers tend to read a centrally drawn
node as more prominent, so the drawing should not be used as an importance
score.

```{r}
#| label: fig-riverton-graph
#| fig-cap: "ID-keyed Riverton knowledge graph with directed, labelled edges."
#| fig-alt: !expr paste0("A node-link diagram of ", nrow(nodes), " nodes, each labelled with a name and an entity ID, and ", nrow(edges), " directed relation edges. The exact edges are listed in the edge table above.")
#| fig-width: 9
#| fig-height: 6.5

layout_from_matrix <- function(graph, coordinates) {
  tibble(
    node_id = igraph::V(graph)$name,
    x = coordinates[, 1],
    y = coordinates[, 2]
  ) |>
    left_join(nodes, by = "node_id")
}

layout_fixed <- tribble(
  ~node_id, ~x, ~y,
  "ORG-0002", 0.0, 0.0,
  "CRD-0002", -2.2, 1.5,
  "VALUE-October 15", 2.2, 1.5,
  "VALUE-November 1", 2.2, -1.1,
  "LOC-0001", -0.2, -2.2,
  "ORG-0003", -2.4, -3.3,
  "LOC-0003", 1.8, -3.3,
  "LOC-0002", -2.4, -0.8
) |>
  left_join(nodes, by = "node_id")

edge_label_positions <- tribble(
  ~from, ~to, ~label_x, ~label_y,
  "ORG-0002", "CRD-0002", -1.25, 0.95,
  "ORG-0002", "VALUE-October 15", 1.30, 0.95,
  "ORG-0002", "VALUE-November 1", 1.35, -0.55
)

shorten_segments <- function(edge_positions, amount = 0.20) {
  edge_positions |>
    mutate(
      x_start = x_from + amount * (x_to - x_from),
      y_start = y_from + amount * (y_to - y_from),
      x_end = x_from + (1 - amount) * (x_to - x_from),
      y_end = y_from + (1 - amount) * (y_to - y_from),
      x_mid = (x_start + x_end) / 2,
      y_mid = (y_start + y_end) / 2
    )
}

edge_positions <- edges |>
  left_join(
    layout_fixed |> select(from = node_id, x_from = x, y_from = y),
    by = "from"
  ) |>
  left_join(
    layout_fixed |> select(to = node_id, x_to = x, y_to = y),
    by = "to"
  ) |>
  left_join(edge_label_positions, by = c("from", "to")) |>
  shorten_segments() |>
  mutate(relation_label = str_replace_all(relation, "_", " "))

node_positions <- layout_fixed |>
  mutate(display_label = paste0(label, "\n", node_id))

graph_plot <- ggplot() +
  geom_segment(
    data = edge_positions,
    aes(x = x_start, y = y_start, xend = x_end, yend = y_end),
    arrow = grid::arrow(type = "closed", length = grid::unit(0.12, "inches")),
    linewidth = 0.4,
    lineend = "round"
  ) +
  geom_label(
    data = edge_positions,
    aes(x = label_x, y = label_y, label = relation_label),
    fill = "white",
    linewidth = 0.15,
    size = 2.8
  ) +
  geom_label(
    data = node_positions,
    aes(x = x, y = y, label = display_label),
    fill = "white",
    linewidth = 0.25,
    size = 3
  ) +
  coord_equal(clip = "off") +
  expand_limits(x = c(-3.2, 3.0), y = c(-3.9, 2.2)) +
  theme_void() +
  theme(plot.margin = margin(15, 15, 15, 15))

graph_plot
```

```{r}
#| include: false

edge_set <- edges |>
  transmute(edge = paste(from, relation, to, sep = "|")) |>
  arrange(edge) |>
  pull(edge)

degree_table <- tibble(
  node_id = igraph::V(id_graph)$name,
  total_degree = igraph::degree(id_graph, mode = "total"),
  out_degree = igraph::degree(id_graph, mode = "out"),
  in_degree = igraph::degree(id_graph, mode = "in")
)

label_boxes <- bind_rows(
  node_positions |>
    transmute(
      label_id = node_id,
      label_kind = "node",
      x,
      y,
      half_width = 0.050 * pmax(nchar(label), nchar(node_id)),
      half_height = 0.28
    ),
  edge_positions |>
    transmute(
      label_id = paste(from, relation, to, sep = "|"),
      label_kind = "edge",
      x = label_x,
      y = label_y,
      half_width = 0.035 * nchar(relation_label),
      half_height = 0.14
    )
)

label_box_pairs <- label_boxes |>
  cross_join(label_boxes) |>
  filter(label_id.x < label_id.y) |>
  mutate(
    x_overlap = abs(x.x - x.y) < (half_width.x + half_width.y),
    y_overlap = abs(y.x - y.y) < (half_height.x + half_height.y),
    boxes_overlap = x_overlap & y_overlap
  )

stopifnot(
  nrow(layout_fixed) == igraph::vcount(id_graph),
  nrow(edge_positions) == igraph::ecount(id_graph),
  nrow(edge_list) == igraph::ecount(id_graph),
  all(!is.na(edge_positions$x_start)),
  all(!is.na(edge_positions$label_x)),
  !any(label_box_pairs$boxes_overlap),
  all(edge_set == sort(edge_set)),
  !("weight" %in% names(igraph::edge_attr(id_graph)))
)
```

The edge labels are text, not color. The arrowheads show direction, while the
caption and table make the same information available outside the image.

## Check seed dependence without reading coordinates

Force-directed layouts are useful for many graphs, but their coordinates depend
on random starting positions. A layout cannot change the graph's edges or
degrees; it only assigns positions. The check below runs two seeded
Fruchterman-Reingold layouts and compares the node IDs and coordinates rather
than coordinate literals.

```{r}
layout_for_seed <- function(graph, seed) {
  set.seed(seed)
  layout_from_matrix(
    graph,
    igraph::layout_with_fr(graph, weights = NA)
  )
}

layout_seed_17 <- layout_for_seed(id_graph, 17L)
layout_seed_29 <- layout_for_seed(id_graph, 29L)

seed_comparison <- tibble(
  check = c("same node IDs", "coordinates identical"),
  result = c(
    setequal(layout_seed_17$node_id, layout_seed_29$node_id),
    identical(
      layout_seed_17 |> arrange(node_id) |> select(x, y),
      layout_seed_29 |> arrange(node_id) |> select(x, y)
    )
  )
)

knitr::kable(
  seed_comparison,
  col.names = c("Layout check", "Result"),
  caption = "Two layout seeds preserve graph structure while changing coordinates",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  seed_comparison$result[seed_comparison$check == "same node IDs"],
  !seed_comparison$result[seed_comparison$check == "coordinates identical"]
)
```

Use the main layout to follow labelled arrows. The seeded force-layout check is
there only to show why coordinates should not be treated as facts.

## Build a co-word network for comparison

A co-word network is a different graph. Its nodes are words. Its edges are
undirected and based on how often two words appear in the same unit. Here the
unit is an inaugural paragraph. The word rule keeps alphabetic tokens of four
or more letters, lowercases them in English, removes tidytext stop words, and
uses paragraph presence rather than raw repeated counts.

```{r}
paragraphs <- inaugural_paragraphs()

paragraph_words <- paragraphs |>
  transmute(paragraph_id, paragraph) |>
  tidytext::unnest_tokens(word, paragraph, token = "regex", pattern = "[^A-Za-z]+") |>
  mutate(word = str_to_lower(word, locale = "en")) |>
  filter(nchar(word) >= 4L) |>
  anti_join(tidytext::stop_words, by = "word") |>
  distinct(paragraph_id, word)

top_words <- paragraph_words |>
  count(word, name = "paragraphs_with_word", sort = TRUE) |>
  slice_head(n = 10)

top_word_presence <- paragraph_words |>
  semi_join(top_words, by = "word")

word_pairs <- top_word_presence |>
  inner_join(top_word_presence, by = "paragraph_id", relationship = "many-to-many") |>
  filter(word.x < word.y) |>
  count(word.x, word.y, name = "observed") |>
  left_join(top_words, by = c("word.x" = "word")) |>
  rename(first_paragraphs = paragraphs_with_word) |>
  left_join(top_words, by = c("word.y" = "word")) |>
  rename(second_paragraphs = paragraphs_with_word) |>
  mutate(
    total_paragraphs = n_distinct(paragraphs$paragraph_id),
    expected = first_paragraphs * second_paragraphs / total_paragraphs,
    observed_expected_ratio = observed / expected
  )

ratio_threshold <- 1.80
coword_edges <- word_pairs |>
  filter(observed_expected_ratio >= ratio_threshold) |>
  arrange(desc(observed_expected_ratio), word.x, word.y)

kept_pair_count <- nrow(coword_edges)
dropped_pair_count <- nrow(word_pairs) - kept_pair_count
undrawn_words <- setdiff(top_words$word, unique(c(coword_edges$word.x, coword_edges$word.y)))

knitr::kable(
  top_words,
  col.names = c("Word", "Paragraphs with word"),
  caption = "Top ten content words after the paragraph-level word rule",
  row.names = FALSE
)

knitr::kable(
  coword_edges |>
    mutate(
      expected = round(expected, 1),
      observed_expected_ratio = round(observed_expected_ratio, 2)
    ),
  col.names = c(
    "First word", "Second word", "Observed paragraphs",
    "First word paragraphs", "Second word paragraphs",
    "Total paragraphs", "Expected under independence",
    "Observed / expected"
  ),
  caption = "Co-word edges kept by the observed-versus-expected threshold",
  row.names = FALSE
)

threshold_summary <- tibble(
  threshold = ratio_threshold,
  kept_pairs = kept_pair_count,
  dropped_pairs = dropped_pair_count,
  undrawn_top_words = paste(undrawn_words, collapse = ", ")
)

knitr::kable(
  threshold_summary,
  col.names = c("Observed / expected threshold", "Kept pairs", "Dropped pairs", "Top words not drawn"),
  caption = "The author's co-word threshold decides which word pairs are drawn",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(top_words), 10L),
  identical(nrow(word_pairs), 45L),
  all(word_pairs$observed > 0L),
  nrow(coword_edges) < nrow(word_pairs),
  ratio_threshold == 1.80,
  identical(kept_pair_count, 3L),
  identical(dropped_pair_count, 42L),
  all(unique(c(coword_edges$word.x, coword_edges$word.y)) %in% top_words$word)
)
```

Raw co-occurrence among these ten frequent words is a complete graph: all 45
pairs appear in at least one paragraph. The threshold is the author's choice
for this lesson. It keeps `r kept_pair_count` pairs and drops
`r dropped_pair_count` pairs by requiring the observed count to be at least
1.8 times the independence expectation
`count(word A) * count(word B) / paragraphs`. The undrawn top-ten words are
`r paste(undrawn_words, collapse = ", ")`.

## Draw the co-word graph

The co-word graph uses the same drawing machinery, but its edges are not typed
facts. They are paragraph-level co-occurrences that passed the threshold.

```{r}
#| label: fig-coword-network
#| fig-cap: "Co-word network for the three top-word pairs that passed the observed-versus-expected threshold."
#| fig-alt: "An undirected co-word diagram whose nodes are content words and whose edges are thresholded paragraph co-occurrences. The kept edges and counts are listed in the table 'Co-word edges kept by the observed-versus-expected threshold'."
#| fig-width: 8
#| fig-height: 5.5

coword_nodes <- tibble(
  node_id = sort(unique(c(coword_edges$word.x, coword_edges$word.y))),
  label = node_id
)

coword_graph <- igraph::graph_from_data_frame(
  d = coword_edges |>
    transmute(from = word.x, to = word.y, relation = "co-occurs"),
  directed = FALSE,
  vertices = coword_nodes |> select(name = node_id, label)
)

coword_layout_matrix <- igraph::layout_in_circle(coword_graph)
coword_layout <- tibble(
  node_id = igraph::V(coword_graph)$name,
  x = coword_layout_matrix[, 1],
  y = coword_layout_matrix[, 2]
) |>
  left_join(coword_nodes, by = "node_id")

coword_positions <- coword_edges |>
  transmute(from = word.x, to = word.y, relation = "co-occurs") |>
  left_join(
    coword_layout |> select(from = node_id, x_from = x, y_from = y),
    by = "from"
  ) |>
  left_join(
    coword_layout |> select(to = node_id, x_to = x, y_to = y),
    by = "to"
  ) |>
  shorten_segments(amount = 0.12)

coword_plot <- ggplot() +
  geom_segment(
    data = coword_positions,
    aes(x = x_start, y = y_start, xend = x_end, yend = y_end),
    linewidth = 0.4,
    lineend = "round"
  ) +
  geom_label(
    data = coword_layout,
    aes(x = x, y = y, label = label),
    fill = "white",
    linewidth = 0.25,
    size = 3
  ) +
  coord_equal(clip = "off") +
  theme_void() +
  theme(plot.margin = margin(15, 15, 15, 15))

coword_plot
```

```{r}
#| include: false

stopifnot(
  igraph::vcount(coword_graph) == nrow(coword_nodes),
  igraph::ecount(coword_graph) == nrow(coword_edges),
  nrow(coword_positions) == nrow(coword_edges),
  !igraph::is_directed(coword_graph),
  !("weight" %in% names(igraph::edge_attr(coword_graph)))
)
```

Unlike the Riverton knowledge graph, this co-word network has no entity IDs, no
relation types, and no direction. It can show which word pairs survived a
preprocessing and threshold decision. It cannot say that the words mean the
same thing or that one caused the other to appear.

## What to remember

- Use stable IDs as node keys; labels can collide.
- The two `Riverton` places merge if labels are treated as IDs.
- A force-directed layout is a computed drawing, not a measurement.
- Arrowheads and relation labels carry the graph meaning.
- The edge-list table is part of the graph's text alternative.
- Co-word edges are thresholded co-occurrences, not typed knowledge-base facts.

Start with the edge table, then use the drawing only as a guide to the paths it
lists.

## Sources

- [Hogan et al. (2021), Knowledge Graphs](https://arxiv.org/abs/2003.02320)
- [Fruchterman and Reingold (1991), Graph Drawing by Force-directed Placement](https://reingold.co/force-directed.pdf)
- [igraph `layout_in_circle()` reference](https://r.igraph.org/reference/layout_in_circle.html)
- [igraph `layout_with_fr()` reference](https://r.igraph.org/reference/layout_with_fr.html)
- [McGrath, Blythe, and Krackhardt (1997), The effect of spatial arrangement on judgments and errors in interpreting graphs](https://doi.org/10.1016/S0378-8733(96)00299-7)
- [W3C WAI tutorial on complex images](https://www.w3.org/WAI/tutorials/images/complex/)
- [WCAG 2.2 Understanding Success Criterion 1.1.1](https://www.w3.org/WAI/WCAG22/Understanding/non-text-content.html)
- [Rip and Courtial (1984), Co-word maps of biotechnology](https://ris.utwente.nl/ws/files/6907221/co-word.pdf)
- [van Eck and Waltman (2009), How to normalize cooccurrence data](https://repub.eur.nl/pub/14528/ERS-2009-001-LIS.pdf)
