---
title: "Drawing a knowledge graph"
subtitle: "Use IDs, arrows, labels, and tables before reading a layout"
description: "Learn how to draw a small knowledge graph in R with ggraph while keeping identifiers, edge labels, layout limits, and text alternatives visible."
categories: [information visualization, knowledge graph visualization, networks]
---

Chloe has a tidy table of Riverton facts, but the table is hard to scan when
she wants to follow a path. Which organisation offers the credential? Which
deadline values did the knowledge base keep for review? Which place label might
cause trouble if IDs were ignored?

**Knowledge graph visualization** draws entities as nodes and typed relations as
edges. In this lesson, the edges are directed and labelled because the relation
itself is part of the data.

A drawing can help Chloe see a path. It can also mislead her if she treats the
layout as a measurement. The table of edges remains the source of the graph.

::: {.callout-note}
The Riverton entities and facts in this lesson are fictional teaching records rebuilt from the earlier knowledge base example.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- key graph nodes by entity ID instead of display label;
- show how label-keyed data wrongly merges two Riverton places;
- draw directed, labelled edges with ggraph, the ggplot2 extension for
  networks;
- explain why layout coordinates can change and are not data;
- give a captioned edge table as the text alternative; and
- compare a co-word network with a typed knowledge graph.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Rebuild the graph facts by ID

The graph starts with the populated fact edges named in the knowledge-base
lesson. The two `Riverton` places have different IDs. That difference is the
first thing the drawing must preserve.

```{r}
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(igraph)
library(tidygraph)
library(ggplot2)
library(ggraph)
library(tidytext)

source("R/inaugural-corpus.R")
source("R/lesson-figures.R")

nodes <- tribble(
  ~node_id, ~label, ~node_type,
  "ORG-0002", "Riverton Skills Centre", "organisation",
  "ORG-0003", "Marrow County Transit", "organisation",
  "LOC-0001", "Riverton", "place",
  "LOC-0002", "Riverton", "place",
  "LOC-0003", "Bellhaven", "place",
  "CRD-0002", "Data Support Certificate", "credential",
  "VALUE-October 15", "October 15", "date literal",
  "VALUE-November 1", "November 1", "date literal"
)

edges <- tribble(
  ~from, ~to, ~relation, ~document_ids,
  "ORG-0002", "CRD-0002", "offers", "A001; A002",
  "ORG-0002", "VALUE-October 15", "application_deadline", "A003",
  "ORG-0002", "VALUE-November 1", "application_deadline", "A004"
)
```

```{r}
#| echo: false

node_names <- nodes |>
  transmute(node_id, name = str_c(label, " (", node_id, ")"))

edges |>
  left_join(node_names, by = join_by(from == node_id)) |>
  rename(from_node = name) |>
  left_join(node_names, by = join_by(to == node_id)) |>
  rename(to_node = name) |>
  transmute(
    from_node,
    relation = str_replace_all(relation, "_", " "),
    to_node,
    document_ids
  ) |>
  knitr::kable(
    col.names = c("From", "Relation", "To", "Supporting documents"),
    caption = "Every edge drawn in the Riverton knowledge graph",
    row.names = FALSE
  )
```

```{r}
#| include: false

stopifnot(
  identical(nrow(nodes), 8L),
  identical(nrow(edges), 3L),
  !anyDuplicated(nodes$node_id),
  all(edges$from %in% nodes$node_id),
  all(edges$to %in% nodes$node_id),
  all(edges$document_ids %in% c("A001; A002", "A003", "A004")),
  !any(str_detect(nodes$label, "2026")),
  sum(nodes$label == "Riverton") == 2L,
  all(c("LOC-0001", "LOC-0002") %in% nodes$node_id)
)
```

The table is the long description of the graph. It lists every drawn edge with
IDs and labels, so each relation can be checked without seeing the picture.
Both deadline values are kept because the knowledge base flagged them as a
dated conflict rather than choosing one automatically. No year appears in these
labels because the announcements say only `October 15` and `November 1`.

## Show the label-keyed mistake

If the graph used labels as node keys, both `Riverton` records would collapse
into one node. That would merge two different places, so any later fact about
one Riverton would appear to belong to the other.

```{r}
id_graph <- tbl_graph(
  nodes = nodes |> rename(name = node_id),
  edges = edges,
  directed = TRUE,
  node_key = "name"
)

label_edges <- edges |>
  left_join(nodes |> select(from = node_id, from_label = label), by = "from") |>
  left_join(nodes |> select(to = node_id, to_label = label), by = "to") |>
  transmute(from = from_label, to = to_label, relation)

label_keyed_graph <- graph_from_data_frame(
  d = label_edges,
  directed = TRUE,
  vertices = nodes |> distinct(name = label)
)

merge_check <- tibble(
  key_used = c("entity ID", "display label"),
  vertices = c(vcount(id_graph), vcount(label_keyed_graph)),
  riverton_vertices = c(
    sum(V(id_graph)$label == "Riverton"),
    sum(V(label_keyed_graph)$name == "Riverton")
  )
)
```

```{r}
#| echo: false

knitr::kable(
  merge_check,
  col.names = c("Key used", "Vertices", "Vertices displayed as Riverton"),
  caption = "Using labels as keys wrongly merges the two Riverton places",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  vcount(id_graph) == 8L,
  ecount(id_graph) == 3L,
  vcount(label_keyed_graph) == 7L,
  identical(merge_check$riverton_vertices, c(2L, 1L)),
  identical(
    sort(V(id_graph)$name),
    sort(nodes$node_id)
  )
)
```

`tbl_graph()` from tidygraph builds the graph from the two tables, with the
entity ID as the node key. The ID-keyed graph has eight vertices. The
label-keyed graph has seven because the two places named `Riverton` have been
merged.

```{r}
name_clash <- nodes |>
  summarise(
    ids_with_label = n(),
    entity_ids = str_c(node_id, collapse = "; "),
    entity_types = str_c(node_type, collapse = "; "),
    .by = label
  ) |>
  filter(ids_with_label > 1L)

unconnected_nodes <- nodes |>
  filter(!node_id %in% c(edges$from, edges$to))
```

```{r}
#| echo: false

knitr::kable(
  name_clash,
  col.names = c("Shared label", "IDs with label", "Entity IDs", "Entity types"),
  caption = "The name clash is a label problem, not a factual graph edge",
  row.names = FALSE
)

knitr::kable(
  unconnected_nodes,
  col.names = c("Node ID", "Label", "Type"),
  caption = "Nodes drawn without a populated-fact edge",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(name_clash$label, "Riverton"),
  identical(name_clash$ids_with_label, 2L),
  str_detect(name_clash$entity_ids, "LOC-0001"),
  str_detect(name_clash$entity_ids, "LOC-0002"),
  setequal(unconnected_nodes$node_id, c("ORG-0003", "LOC-0001", "LOC-0002", "LOC-0003"))
)
```

The figure draws `r nrow(unconnected_nodes)` nodes that have no populated fact
yet. It places them in its bottom row, and the table above names them.

## Place nodes by hand and draw arrows

**ggraph** draws graphs with the same grammar as ggplot2: a layout gives each
node a position, `geom_edge_*()` layers draw the edges, and `geom_node_*()`
layers draw the nodes. Here the layout is written by hand, so the drawing stays
the same on every render while Chloe reads the arrows and labels.

The bottom row holds the nodes with no populated fact. Apart from that row, up,
down, left, right, rotation, and mirror image carry no data. Count links in the
table if you need a number. Viewers tend to read a centrally drawn node as more
important, so the drawing should not be used as an importance score.

```{r}
#| label: fig-riverton-graph
#| fig-cap: "ID-keyed Riverton knowledge graph with directed, labelled edges; the bottom row holds nodes with no populated fact."
#| fig-alt: !expr paste0("A node-link diagram of ", nrow(nodes), " nodes, each labelled with a name and an ID, and ", nrow(edges), " directed relation edges. Riverton Skills Centre (ORG-0002) offers the Data Support Certificate (CRD-0002) and has two application deadlines, October 15 and November 1. A bottom row holds four nodes with no populated fact, including the two places labelled Riverton, which stay separate because they have different IDs. The exact edges are listed in the edge table above.")
#| fig-width: 9
#| fig-height: 6

hand_layout <- tribble(
  ~name, ~x, ~y,
  "ORG-0002", 0.0, 0.0,
  "CRD-0002", -3.2, 1.9,
  "VALUE-October 15", 3.2, 1.9,
  "VALUE-November 1", 3.2, -1.7,
  "LOC-0001", -3.6, -3.6,
  "LOC-0002", -1.2, -3.6,
  "ORG-0003", 1.2, -3.6,
  "LOC-0003", 3.6, -3.6
)

node_type_fills <- c(
  organisation = "#D6E6F4",
  credential = "#D5EFE6",
  "date literal" = "#FBE3CF",
  place = "#E6E9ED"
)

riverton_graph <- id_graph |>
  activate(nodes) |>
  left_join(hand_layout, by = "name") |>
  mutate(display_label = str_c(label, "\n", name)) |>
  activate(edges) |>
  mutate(relation_label = str_replace_all(relation, "_", " "))

graph_plot <- ggraph(
  riverton_graph,
  layout = "manual",
  x = x,
  y = y
) +
  geom_edge_link(
    aes(
      label = relation_label,
      start_cap = label_rect(node1.display_label, fontsize = 11, fontfamily = lesson_font),
      end_cap = label_rect(node2.display_label, fontsize = 11, fontfamily = lesson_font)
    ),
    arrow = arrow(type = "closed", length = unit(3, "mm")),
    angle_calc = "along",
    label_dodge = unit(3.2, "mm"),
    label_size = 3.4,
    family = lesson_font,
    edge_colour = lesson_colours[["body"]],
    edge_width = 0.5
  ) +
  geom_node_label(
    aes(label = display_label, fill = node_type),
    family = lesson_font,
    size = 3.8,
    lineheight = 1.05,
    colour = lesson_colours[["ink"]],
    label.padding = unit(0.35, "lines"),
    label.r = unit(0.2, "lines"),
    key_glyph = "rect"
  ) +
  scale_fill_manual(values = node_type_fills, name = "Node type") +
  coord_cartesian(clip = "off") +
  scale_x_continuous(expand = expansion(add = 1.2)) +
  scale_y_continuous(expand = expansion(add = 0.6)) +
  labs(
    title = "Facts connect IDs, so the two places named Riverton stay apart",
    subtitle = str_c(
      "Arrows point from subject to object and carry the relation. ",
      "The bottom row holds nodes with no populated fact."
    )
  ) +
  theme_lesson(grid = "none") +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    legend.key.size = unit(0.9, "lines")
  )

graph_plot
```

```{r}
#| include: false

graph_nodes <- riverton_graph |>
  activate(nodes) |>
  as_tibble()

graph_edges <- riverton_graph |>
  activate(edges) |>
  as_tibble() |>
  mutate(
    from_id = graph_nodes$name[from],
    to_id = graph_nodes$name[to]
  )

# Rough text boxes in layout units: about 0.07 units per character at this
# figure size. The check only has to catch labels placed on top of each other.
node_boxes <- graph_nodes |>
  transmute(
    box_id = name,
    x,
    y,
    half_width = 0.07 * pmax(nchar(label), nchar(name)) / 2 + 0.2,
    half_height = 0.45
  )

label_box_pairs <- node_boxes |>
  cross_join(node_boxes) |>
  filter(box_id.x < box_id.y) |>
  mutate(
    boxes_overlap = abs(x.x - x.y) < (half_width.x + half_width.y) &
      abs(y.x - y.y) < (half_height.x + half_height.y)
  )

# Each relation label sits in the middle of its edge, tilted along it. It is
# readable only if it fits in the stretch of edge left visible between the two
# node boxes.
box_reach <- function(half_width, half_height, angle) {
  pmin(half_width / abs(cos(angle)), half_height / abs(sin(angle)))
}

edge_label_room <- graph_edges |>
  left_join(node_boxes |> select(from_id = box_id, x_from = x, y_from = y, w_from = half_width, h_from = half_height), by = "from_id") |>
  left_join(node_boxes |> select(to_id = box_id, x_to = x, y_to = y, w_to = half_width, h_to = half_height), by = "to_id") |>
  mutate(
    angle = atan2(y_to - y_from, x_to - x_from),
    edge_length = sqrt((x_to - x_from)^2 + (y_to - y_from)^2),
    visible_length = edge_length -
      box_reach(w_from, h_from, angle) -
      box_reach(w_to, h_to, angle),
    label_length = 0.07 * nchar(relation_label)
  )

unconnected_rows <- graph_nodes |>
  mutate(unconnected = name %in% unconnected_nodes$node_id)

stopifnot(
  !anyNA(graph_nodes$x),
  !anyNA(graph_nodes$y),
  identical(nrow(graph_edges), nrow(edges)),
  identical(
    str_c(graph_edges$from_id, graph_edges$relation, graph_edges$to_id),
    str_c(edges$from, edges$relation, edges$to)
  ),
  !any(label_box_pairs$boxes_overlap),
  identical(nrow(edge_label_room), nrow(edges)),
  all(edge_label_room$visible_length > edge_label_room$label_length),
  !("weight" %in% names(graph_edges)),
  all(unconnected_rows$y[unconnected_rows$unconnected] == min(unconnected_rows$y)),
  all(unconnected_rows$y[!unconnected_rows$unconnected] > min(unconnected_rows$y)),
  setequal(names(node_type_fills), nodes$node_type)
)
```

The relation labels are text, not colour. The arrowheads show direction, and
the caption and table make the same information available outside the image.
The fill colours group node types, and the ID prefix inside every label says
the same thing in words.

## Check that a layout seed changes only the picture

Force-directed layouts are useful for many graphs, but their coordinates depend
on random starting positions. A layout cannot change the graph's edges or
degrees; it only assigns positions. The code below runs two seeded
Fruchterman-Reingold layouts and checks only whether their coordinates are
identical, without printing them.

```{r}
layout_for_seed <- function(graph, seed) {
  set.seed(seed)
  layout_with_fr(graph, weights = NA)
}

coordinates_identical <- identical(
  layout_for_seed(id_graph, 17L),
  layout_for_seed(id_graph, 29L)
)

coordinates_identical
```

```{r}
#| include: false

stopifnot(
  isFALSE(coordinates_identical)
)
```

`FALSE` means the two seeds put the same eight nodes in different places. Use
the hand-placed figure to follow the labelled arrows; the seeded check is here
only to show why coordinates should not be treated as facts.

## Build a co-word network for comparison

A co-word network is a different graph. Its nodes are words. Its edges are
undirected and based on how often two words appear in the same unit. Here the
unit is an inaugural paragraph. The word rule keeps alphabetic tokens of four
or more letters, lowercases them in English, removes tidytext stop words, and
uses paragraph presence rather than raw repeated counts.

```{r}
paragraphs <- inaugural_paragraphs()

paragraph_words <- paragraphs |>
  transmute(paragraph_id, paragraph) |>
  unnest_tokens(word, paragraph, token = "regex", pattern = "[^A-Za-z]+") |>
  mutate(word = str_to_lower(word, locale = "en")) |>
  filter(nchar(word) >= 4L) |>
  anti_join(stop_words, by = "word") |>
  distinct(paragraph_id, word)

top_words <- paragraph_words |>
  count(word, name = "paragraphs_with_word", sort = TRUE) |>
  slice_head(n = 10)

top_word_presence <- paragraph_words |>
  semi_join(top_words, by = "word")

word_pairs <- top_word_presence |>
  inner_join(top_word_presence, by = "paragraph_id", relationship = "many-to-many") |>
  filter(word.x < word.y) |>
  count(word.x, word.y, name = "observed") |>
  left_join(top_words, by = c("word.x" = "word")) |>
  rename(first_paragraphs = paragraphs_with_word) |>
  left_join(top_words, by = c("word.y" = "word")) |>
  rename(second_paragraphs = paragraphs_with_word) |>
  mutate(
    total_paragraphs = n_distinct(paragraphs$paragraph_id),
    expected = first_paragraphs * second_paragraphs / total_paragraphs,
    observed_expected_ratio = observed / expected
  )

ratio_threshold <- 1.80
coword_edges <- word_pairs |>
  filter(observed_expected_ratio >= ratio_threshold) |>
  arrange(desc(observed_expected_ratio), word.x, word.y)

undrawn_words <- setdiff(top_words$word, c(coword_edges$word.x, coword_edges$word.y))
```

```{r}
#| echo: false

knitr::kable(
  top_words,
  col.names = c("Word", "Paragraphs with word"),
  caption = "Top ten content words after the paragraph-level word rule",
  row.names = FALSE
)

knitr::kable(
  coword_edges |>
    mutate(
      expected = round(expected, 1),
      observed_expected_ratio = round(observed_expected_ratio, 2)
    ),
  col.names = c(
    "First word", "Second word", "Observed paragraphs",
    "First word paragraphs", "Second word paragraphs",
    "Total paragraphs", "Expected under independence",
    "Observed / expected"
  ),
  caption = "Co-word edges kept by the observed-versus-expected threshold",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(top_words), 10L),
  identical(nrow(word_pairs), 45L),
  all(word_pairs$observed > 0L),
  ratio_threshold == 1.80,
  identical(nrow(coword_edges), 3L),
  all(c(coword_edges$word.x, coword_edges$word.y) %in% top_words$word)
)
```

Raw co-occurrence among these ten frequent words is a complete graph: all 45
pairs appear in at least one paragraph. The threshold is the author's choice
for this lesson. It keeps `r nrow(coword_edges)` pairs and drops
`r nrow(word_pairs) - nrow(coword_edges)` by requiring the observed count to be
at least 1.8 times the count expected if the two words were independent,
`count(word A) * count(word B) / paragraphs`. The top-ten words left out of the
drawing are `r str_c(undrawn_words, collapse = ", ")`.

## Draw the co-word graph

The co-word graph uses the same ggraph grammar, but its edges are not typed
facts. They are paragraph-level co-occurrences that passed the threshold, so
each edge is labelled with its observed-to-expected ratio instead of a
relation.

```{r}
#| label: fig-coword-network
#| fig-cap: "Co-word network for the three top-word pairs that passed the observed-versus-expected threshold."
#| fig-alt: "An undirected co-word diagram whose nodes are content words and whose edges are thresholded paragraph co-occurrences, each labelled with its observed-to-expected ratio. The kept edges and counts are listed in the table 'Co-word edges kept by the observed-versus-expected threshold'."
#| fig-width: 8
#| fig-height: 4.6

# Listing the words edge by edge puts linked words next to each other on the
# circle, so no two edges cross.
coword_nodes <- tibble(
  name = unique(as.vector(rbind(coword_edges$word.x, coword_edges$word.y)))
)

coword_graph <- tbl_graph(
  nodes = coword_nodes,
  edges = coword_edges |>
    transmute(
      from = word.x,
      to = word.y,
      ratio_label = sprintf("%.1f×", observed_expected_ratio)
    ),
  directed = FALSE
)

coword_plot <- ggraph(coword_graph, layout = "circle") +
  geom_edge_link(
    aes(
      label = ratio_label,
      start_cap = label_rect(node1.name, fontsize = 12, fontfamily = lesson_font),
      end_cap = label_rect(node2.name, fontsize = 12, fontfamily = lesson_font)
    ),
    angle_calc = "along",
    label_dodge = unit(3, "mm"),
    label_size = 3.3,
    family = lesson_font,
    edge_colour = lesson_colours[["body"]],
    edge_width = 0.6
  ) +
  geom_node_label(
    aes(label = name),
    family = lesson_font,
    size = 4.2,
    fill = lesson_colours[["paper"]],
    colour = lesson_colours[["ink"]],
    label.padding = unit(0.3, "lines")
  ) +
  coord_cartesian(clip = "off") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.15)) +
  labs(
    title = str_c(
      "Only ", nrow(coword_edges), " of ", nrow(word_pairs),
      " frequent word pairs pass the threshold"
    ),
    subtitle = "Undirected edges between words that share paragraphs at least 1.8 times as often as expected"
  ) +
  theme_lesson(grid = "none") +
  theme(axis.text = element_blank(), axis.title = element_blank())

coword_plot
```

```{r}
#| include: false

coword_graph_edges <- coword_graph |>
  activate(edges) |>
  as_tibble()

stopifnot(
  gorder(coword_graph) == n_distinct(c(coword_edges$word.x, coword_edges$word.y)),
  gsize(coword_graph) == nrow(coword_edges),
  !is_directed(coword_graph),
  !("weight" %in% names(coword_graph_edges))
)
```

Unlike the Riverton knowledge graph, this co-word network has no entity IDs, no
relation types, and no direction. It can show which word pairs survived a
preprocessing and threshold decision. It cannot say that the words mean the
same thing or that one caused the other to appear.

## What to remember

- Use stable IDs as node keys; labels can collide.
- The two `Riverton` places merge if labels are treated as IDs.
- A force-directed layout is a computed drawing, not a measurement.
- Arrowheads and relation labels carry the graph meaning.
- The edge-list table is part of the graph's text alternative.
- Co-word edges are thresholded co-occurrences, not typed knowledge-base facts.

Start with the edge table, then use the drawing only as a guide to the paths it
lists.

## Sources

- [Hogan et al. (2021), Knowledge Graphs](https://arxiv.org/abs/2003.02320)
- [Fruchterman and Reingold (1991), Graph Drawing by Force-directed Placement](https://reingold.co/force-directed.pdf)
- [ggraph documentation](https://ggraph.data-imaginist.com/)
- [tidygraph documentation](https://tidygraph.data-imaginist.com/)
- [igraph `layout_with_fr()` reference](https://r.igraph.org/reference/layout_with_fr.html)
- [McGrath, Blythe, and Krackhardt (1997), The effect of spatial arrangement on judgments and errors in interpreting graphs](https://doi.org/10.1016/S0378-8733(96)00299-7)
- [W3C WAI tutorial on complex images](https://www.w3.org/WAI/tutorials/images/complex/)
- [WCAG 2.2 Understanding Success Criterion 1.1.1](https://www.w3.org/WAI/WCAG22/Understanding/non-text-content.html)
- [Rip and Courtial (1984), Co-word maps of biotechnology](https://ris.utwente.nl/ws/files/6907221/co-word.pdf)
- [van Eck and Waltman (2009), How to normalize cooccurrence data](https://repub.eur.nl/pub/14528/ERS-2009-001-LIS.pdf)
