---
title: "Finding relations between named things"
subtitle: "Extract directed triples without treating co-occurrence as evidence"
description: "Learn how a small dependency-pattern extractor turns linked Riverton mentions into relation triples with status and evidence spans."
categories: [systems, relation extraction, workforce research]
---

Tomas has a stack of short notices and one question: which organisation offers
which credential, and which credential does a job require? A word search gives
him fragments. It finds `required`, but it cannot tell whether the sentence
says a licence is required, preferred, or not required.

**Relation extraction** finds a statement that two entities stand in a named,
directed relationship. The output is a triple such as `ORG-0002 offers
CRD-0002`, plus a status and a pointer back to the sentence that said it.

This lesson keeps the example small. It links mentions with the Riverton alias
table, inspects the parser links, and compares the extractor with a co-occurrence
baseline that only knows two mentions shared a sentence.

::: {.callout-note}
The Riverton Workforce Lab, its job board, and its training flyer are fictional and were created for teaching.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- define a relation schema before extracting triples;
- keep entity IDs, direction, status, and offsets with each triple;
- explain why two linked mentions in one sentence are not automatically related;
- mark negated and hedged statements separately from asserted ones;
- compare pair detection with triple extraction without mixing their units; and
- say why a same-author reference is only a demonstration.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Define the relation schema

A **schema** says which relation types the extractor is allowed to produce. It
also says which entity type can appear on each side. This matters because
direction is part of the claim: an organisation offers a credential; the
credential does not offer the organisation.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(digest)
library(spacyr)
library(knitr)
source("R/use-spacy.R")

relation_schema <- tribble(
  ~relation, ~subject_type, ~object_type, ~direction, ~status_values, ~what_counts,
  "offers", "organisation", "credential",
  "organisation -> credential", "asserted, negated, hedged",
  "The sentence says the organisation offers, may offer, or does not offer the credential.",
  "requires", "organisation", "credential",
  "organisation -> credential", "asserted, negated, hedged",
  "The sentence says the organisation requires, may require, or does not require the credential."
)

knitr::kable(
  relation_schema,
  col.names = c(
    "Relation", "Subject type", "Object type", "Direction",
    "Status values", "What counts"
  ),
  caption = "Relation schema used before any extraction output",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(relation_schema$relation, c("offers", "requires")),
  all(str_detect(relation_schema$direction, "->")),
  all(str_detect(relation_schema$status_values, "asserted"))
)
```

The schema is small on purpose. It does not try to extract every fact in a job
notice. It only checks two relations that later lessons can put into a knowledge
base.

## Read the source records and aliases

The 28 Riverton sentences are the verbatim source records used in earlier
lessons. The alias table supplies IDs for organisations, places, and credentials.
The extractor uses those IDs for arguments; named entity recognition is not used
to choose the arguments.

```{r}
sentences <- read_csv(
  "data/workforce/workforce_sentences.csv",
  na = character(),
  col_types = cols(
    sentence_id = col_character(),
    document_id = col_character(),
    source_line = col_character(),
    text = col_character(),
    reference_label = col_character(),
    uncertainty = col_character(),
    annotator_id = col_character(),
    rationale = col_character(),
    codebook_version = col_character(),
    codebook_hash = col_character(),
    derived = col_character(),
    transformation = col_character()
  )
)

entities <- read_csv(
  "data/riverton/riverton-entities.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    canonical_name = col_character(),
    entity_type = col_character(),
    description = col_character()
  )
) |>
  select(entity_id, canonical_name, entity_type)

aliases <- read_csv(
  "data/riverton/riverton-aliases.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    alias = col_character()
  )
) |>
  left_join(entities, by = "entity_id")

metadata <- read_csv(
  "data/riverton/riverton-reference-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    fingerprint = col_character()
  )
)

alias_hash <- digest(
  paste(read_lines("data/riverton/riverton-aliases.csv"), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)

knitr::kable(
  aliases |>
    arrange(entity_id, alias) |>
    select(entity_id, canonical_name, entity_type, alias),
  col.names = c("Entity ID", "Canonical name", "Type", "Alias"),
  caption = "Riverton alias table used to find relation arguments",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(sentences), 28L),
  identical(nrow(entities), 8L),
  identical(
    alias_hash,
    metadata$fingerprint[metadata$artifact == "riverton-aliases.csv"]
  ),
  all(c("ORG-0002", "ORG-0003", "CRD-0001", "CRD-0002") %in% aliases$entity_id)
)
```

The table includes two rows for the alias `Riverton`, which is an ambiguity the
lesson leaves unresolved. For relation arguments, the examples below use
organisation and credential aliases that link to one ID.

## Inspect the 28 verbatim sentences once

The job-board and flyer lines are useful for inspection, but they do not contain
two distinct linked entity mentions in one sentence. The table below is the only
place this lesson prints all 28 source sentences. It shows why none of these lines can
produce complete relation triples.

```{r}
escape_regex <- function(value) {
  str_replace_all(value, "([.\\\\|()\\[\\]{}^$*+?])", "\\\\\\1")
}

find_mentions <- function(document_id, sentence_id, text) {
  candidates <- aliases |>
    mutate(normalized_alias = str_to_lower(alias, locale = "en")) |>
    group_by(normalized_alias, alias) |>
    summarise(
      candidate_count = n_distinct(entity_id),
      entity_id = paste(sort(unique(entity_id)), collapse = "; "),
      canonical_name = paste(sort(unique(canonical_name)), collapse = "; "),
      entity_type = paste(sort(unique(entity_type)), collapse = "; "),
      .groups = "drop"
    ) |>
    mutate(alias_length = nchar(alias)) |>
    arrange(desc(alias_length), alias)

  kept <- vector("list", nrow(candidates))
  used <- matrix(numeric(0), ncol = 2)
  kept_index <- 0L

  for (row in seq_len(nrow(candidates))) {
    pattern <- regex(
      paste0("(?<![[:alnum:]])", escape_regex(candidates$alias[[row]]), "(?![[:alnum:]])"),
      ignore_case = TRUE
    )
    locations <- str_locate_all(text, pattern)[[1]]

    if (nrow(locations) == 0L) {
      next
    }

    for (hit in seq_len(nrow(locations))) {
      span <- locations[hit, ]
      overlaps <- nrow(used) > 0L &&
        any(span[1] <= used[, 2] & span[2] >= used[, 1])

      if (!overlaps) {
        kept_index <- kept_index + 1L
        kept[[kept_index]] <- tibble(
          document_id = document_id,
          sentence_id = sentence_id,
          entity_id = candidates$entity_id[[row]],
          canonical_name = candidates$canonical_name[[row]],
          entity_type = candidates$entity_type[[row]],
          link_status = if_else(candidates$candidate_count[[row]] > 1L, "ambiguous", "linked"),
          mention = str_sub(text, span[1], span[2]),
          start = span[1],
          end = span[2]
        )
        used <- rbind(used, span)
      }
    }
  }

  if (kept_index == 0L) {
    return(tibble(
      document_id = character(),
      sentence_id = character(),
      entity_id = character(),
      canonical_name = character(),
      entity_type = character(),
      link_status = character(),
      mention = character(),
      start = integer(),
      end = integer()
    ))
  }

  kept[seq_len(kept_index)] |>
    list_rbind() |>
    arrange(start, end, entity_id)
}

verbatim_mentions <- pmap(
  sentences |>
    select(document_id, sentence_id, text),
  find_mentions
) |>
  list_rbind()

verbatim_inspection <- sentences |>
  mutate(
    linked_mentions = map_chr(
      sentence_id,
      \(id) {
        rows <- verbatim_mentions |>
          filter(sentence_id == id)

        if (nrow(rows) == 0L) {
          "none"
        } else {
          paste(
            paste0(
              rows$mention,
              " (",
              rows$entity_id,
              if_else(rows$link_status == "ambiguous", ", ambiguous", ""),
              ")"
            ),
            collapse = "; "
          )
        }
      }
    ),
    schema_cue = case_when(
      str_detect(str_to_lower(text), "\\boffer|\\bprovided|\\bincluded|\\bpays") ~ "offer-like wording",
      str_detect(str_to_lower(text), "\\brequired|\\bmust|\\bneed") ~ "require-like wording",
      TRUE ~ "none"
    ),
    polarity_read_by_hand = case_when(
      sentence_id %in% c("s002", "s027") ~ "negated",
      TRUE ~ "asserted or not applicable"
    ),
    modality_read_by_hand = case_when(
      sentence_id %in% c("s010", "s013") ~ "preferred, not asserted as required",
      uncertainty == "needs_review" ~ "needs review",
      TRUE ~ "plain"
    )
  ) |>
  select(
    sentence_id,
    document_id,
    text,
    linked_mentions,
    schema_cue,
    polarity_read_by_hand,
    modality_read_by_hand
  )

knitr::kable(
  verbatim_inspection,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Sentence ID", "Document ID", "Text", "Linked mentions",
    "Schema cue", "Polarity read by hand", "Modality read by hand"
  ),
  caption = "Inspection table for all 28 verbatim Riverton sentences",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(verbatim_inspection), 28L),
  identical(verbatim_inspection$text, sentences$text),
  identical(
    verbatim_inspection$polarity_read_by_hand[verbatim_inspection$sentence_id == "s002"],
    "negated"
  ),
  identical(
    verbatim_inspection$polarity_read_by_hand[verbatim_inspection$sentence_id == "s027"],
    "negated"
  ),
  identical(
    verbatim_inspection$linked_mentions[verbatim_inspection$sentence_id == "s023"],
    "RIVERTON (LOC-0001; LOC-0002, ambiguous)"
  ),
  max(count(verbatim_mentions, sentence_id)$n, na.rm = TRUE) < 2L
)
```

The important negative result is concrete: these 28 lines do not give the
extractor complete entity pairs. A page or posting could supply a missing
employer, but that would be a page-structure assumption, not sentence evidence;
this lesson does not use that assumption. The polarity and modality columns are
read by hand so the table can flag lines worth inspecting.

## Write a same-author reference before extraction

The next examples are extractor test sentences. They are constructed to test
the parser patterns and are not claims about the Riverton handbook or knowledge
base. The reference below is written before the extractor runs. The same author
wrote the sentences, the reference, and the rules, so the score is only a
demonstration.

```{r}
constructed_sentences <- tribble(
  ~sentence_id, ~document_id, ~source, ~text, ~teaching_case,
  "c01", "T001", "extractor test sentence", "Riverton Skills Centre offers the Data Support Certificate.", "active offer",
  "c02", "T002", "extractor test sentence", "The Data Support Certificate is offered by Riverton Skills Centre.", "passive offer",
  "c03", "T003", "extractor test sentence", "Riverton Skills Centre does not offer the Forklift Operator Licence.", "verb negation",
  "c05", "T005", "extractor test sentence", "Riverton Skills Centre offers the Data Support Certificate and the Forklift Operator Licence.", "coordinated objects",
  "c06", "T006", "extractor test sentence", "Riverton Skills Centre may offer the Forklift Operator Licence next year.", "hedged offer",
  "c07", "T007", "extractor test sentence", "Marrow County Transit runs a bus to Riverton Skills Centre.", "same-sentence distractor",
  "c08", "T008", "extractor test sentence", "The Workforce Lab counted job ads that require the Forklift Operator Licence.", "relative-clause distractor",
  "c09", "T009", "extractor test sentence", "Marrow County Transit no longer requires the Forklift Operator Licence.", "negation under an adverb",
  "c11", "T011", "extractor test sentence", "No Forklift Operator Licence is required by Marrow County Transit.", "determiner no on argument",
  "c13", "T013", "extractor test sentence", "Marrow County Transit is reported to require the Forklift Operator Licence.", "reported requirement"
)

reference_triples <- tribble(
  ~sentence_id, ~relation, ~subject_id, ~object_id, ~status,
  "c01", "offers", "ORG-0002", "CRD-0002", "asserted",
  "c02", "offers", "ORG-0002", "CRD-0002", "asserted",
  "c03", "offers", "ORG-0002", "CRD-0001", "negated",
  "c05", "offers", "ORG-0002", "CRD-0002", "asserted",
  "c05", "offers", "ORG-0002", "CRD-0001", "asserted",
  "c06", "offers", "ORG-0002", "CRD-0001", "hedged",
  "c09", "requires", "ORG-0003", "CRD-0001", "negated",
  "c11", "requires", "ORG-0003", "CRD-0001", "negated",
  "c13", "requires", "ORG-0003", "CRD-0001", "hedged"
)

knitr::kable(
  constructed_sentences,
  col.names = c("Sentence ID", "Document ID", "Source", "Extractor test sentence", "Teaching case"),
  caption = "Constructed extractor test sentences, not Riverton facts",
  row.names = FALSE
)

knitr::kable(
  reference_triples,
  col.names = c("Sentence ID", "Relation", "Subject ID", "Object ID", "Status"),
  caption = "Same-author reference triples written before the extractor runs",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(constructed_sentences), 10L),
  all(constructed_sentences$source == "extractor test sentence"),
  all(str_starts(constructed_sentences$document_id, "T")),
  identical(nrow(reference_triples), 9L),
  all(reference_triples$status %in% c("asserted", "negated", "hedged")),
  !any(reference_triples$sentence_id %in% c("c07", "c08")),
  any(reference_triples$status == "negated"),
  any(reference_triples$status == "hedged")
)
```

The reference has no row for the bus sentence or the job-ad counting sentence.
Both contain more than one linked mention, but neither states one of the schema
relations between those mentions. The numbering has gaps because scratch parser
checks were removed before this lesson file was committed.

## Parse and align mentions to tokens

spaCy supplies dependency links. The code aligns each alias-table mention to the
tokens covered by its character offsets, then chooses the token whose head points
outside the mention as the mention head.

```{r}
pipeline <- use_project_spacy()
pipeline_info <- spacy_pipeline_version()

parsed <- spacy_parse(
  setNames(constructed_sentences$text, constructed_sentences$sentence_id),
  pos = TRUE,
  lemma = TRUE,
  entity = TRUE,
  dependency = TRUE,
  additional_attributes = "idx"
) |>
  as_tibble() |>
  mutate(
    token_id = as.integer(token_id),
    head_token_id = as.integer(head_token_id),
    token_start = idx + 1L,
    token_end = idx + nchar(token)
  )

constructed_mentions <- pmap(
  constructed_sentences |>
    select(document_id, sentence_id, text),
  find_mentions
) |>
  list_rbind()

head_for_mention <- function(current_sentence_id, current_start, current_end) {
  covered <- parsed |>
    filter(
      doc_id == current_sentence_id,
      token_start >= current_start,
      token_end <= current_end
    )

  external <- covered |>
    filter(!head_token_id %in% covered$token_id)

  if (nrow(external) == 0L) {
    max(covered$token_id)
  } else {
    external$token_id[[1]]
  }
}

mention_heads <- constructed_mentions |>
  mutate(
    head_token_id = pmap_int(
      list(sentence_id, start, end),
      head_for_mention
    )
  ) |>
  ungroup()

mention_head_table <- mention_heads |>
  left_join(
    parsed |>
      transmute(sentence_id = doc_id, head_token_id = token_id, head_token = token, head_dep = dep_rel),
    by = c("sentence_id", "head_token_id")
  ) |>
  select(sentence_id, entity_id, mention, start, end, head_token, head_dep)

knitr::kable(
  mention_head_table,
  col.names = c(
    "Sentence ID", "Entity ID", "Mention", "Start", "End",
    "Mention head", "Head relation"
  ),
  caption = "Alias-table mentions aligned to spaCy dependency heads",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(pipeline$model, "en_core_web_sm"),
  identical(pipeline_info$version, "3.8.0"),
  nrow(parsed) > 0L,
  nrow(mention_heads) >= 18L,
  all(mention_heads$start <= mention_heads$end),
  all(str_sub(
    constructed_sentences$text[
      match(mention_heads$sentence_id, constructed_sentences$sentence_id)
    ],
    mention_heads$start,
    mention_heads$end
  ) == mention_heads$mention)
)
```

The offsets are one-based character positions in the unmodified sentence text.
That lets the triple point back to the exact evidence instead of to a rebuilt
token string.

In the head-relation column, `nsubj` marks an active subject, `dobj` a direct
object, `nsubjpass` a passive subject, `pobj` an object of a preposition, and
`conj` a coordinated item. Lesson 18 introduces these parser links in more
detail.

## Extract directed triples

The extractor looks for `offer` and `require` triggers. It handles active
subjects, passive agents, coordinated objects, `not` on a verb, `No` on an
argument, `no longer` under an adverb, and hedging from `may`. The reported
requirement is a visible miss for this small pattern set.

```{r}
trigger_tokens <- parsed |>
  filter(lemma %in% c("offer", "require"), pos == "VERB") |>
  transmute(
    sentence_id = doc_id,
    trigger_token_id = token_id,
    trigger = token,
    trigger_lemma = lemma,
    trigger_dep = dep_rel,
    trigger_head_token_id = head_token_id
  )

argument_roles <- mention_heads |>
  left_join(
    parsed |>
      transmute(sentence_id = doc_id, token_head_id = head_token_id, head_token_id = token_id, dep_rel),
    by = c("sentence_id", "head_token_id")
  ) |>
  select(
    sentence_id, document_id, entity_id, canonical_name, entity_type,
    mention, start, end, head_token_id, dep_rel, token_head_id
  )

object_heads <- argument_roles |>
  filter(entity_type == "credential") |>
  select(
    sentence_id,
    object_id = entity_id,
    object_name = canonical_name,
    object_mention = mention,
    object_start = start,
    object_end = end,
    object_head_id = head_token_id,
    object_dep = dep_rel,
    object_token_head_id = token_head_id
  )

active_triples <- trigger_tokens |>
  inner_join(
    argument_roles |>
      filter(entity_type == "organisation", dep_rel == "nsubj") |>
      select(
        sentence_id,
        subject_id = entity_id,
        subject_name = canonical_name,
        subject_mention = mention,
        subject_start = start,
        subject_end = end,
        subject_head_id = head_token_id,
        subject_token_head_id = token_head_id
      ),
    by = c("sentence_id", "trigger_token_id" = "subject_token_head_id")
  ) |>
  inner_join(
    object_heads |>
      filter(object_dep %in% c("dobj", "conj")),
    by = "sentence_id",
    relationship = "many-to-many"
  ) |>
  filter(
    pmap_lgl(
      list(sentence_id, object_dep, object_token_head_id, trigger_token_id),
      \(current_sentence_id, current_object_dep, current_object_token_head_id, current_trigger_token_id) {
        direct_object_heads <- object_heads$object_head_id[
          object_heads$sentence_id == current_sentence_id &
            object_heads$object_token_head_id == current_trigger_token_id
        ]

        current_object_token_head_id == current_trigger_token_id ||
          (current_object_dep == "conj" &&
            current_object_token_head_id %in% direct_object_heads)
      }
    )
  )

passive_agents <- parsed |>
  filter(dep_rel == "agent") |>
  select(sentence_id = doc_id, agent_token_id = token_id, trigger_token_id = head_token_id)

passive_triples <- trigger_tokens |>
  inner_join(
    object_heads |>
      filter(object_dep == "nsubjpass"),
    by = c("sentence_id", "trigger_token_id" = "object_token_head_id")
  ) |>
  inner_join(passive_agents, by = c("sentence_id", "trigger_token_id")) |>
  inner_join(
    argument_roles |>
      filter(entity_type == "organisation", dep_rel == "pobj") |>
      select(
        sentence_id,
        subject_id = entity_id,
        subject_name = canonical_name,
        subject_mention = mention,
        subject_start = start,
        subject_end = end,
        subject_head_id = head_token_id,
        subject_token_head_id = token_head_id
      ),
    by = c("sentence_id", "agent_token_id" = "subject_token_head_id")
  )

raw_triples <- bind_rows(
  active_triples,
  passive_triples
) |>
  mutate(
    relation = recode(trigger_lemma, offer = "offers", require = "requires")
  )

status_for_trigger <- function(current_sentence_id, trigger_token_id, subject_head_id, object_head_id) {
  sentence_tokens <- parsed |>
    filter(doc_id == current_sentence_id)

  verb_negated <- any(
    sentence_tokens$head_token_id == trigger_token_id &
      sentence_tokens$dep_rel == "neg"
  )

  advmod_ids <- sentence_tokens$token_id[
    sentence_tokens$head_token_id == trigger_token_id &
      sentence_tokens$dep_rel == "advmod"
  ]
  adverb_negated <- any(
    sentence_tokens$head_token_id %in% advmod_ids &
      sentence_tokens$dep_rel == "neg"
  )

  argument_det_no <- any(
    sentence_tokens$head_token_id %in% c(subject_head_id, object_head_id) &
      sentence_tokens$dep_rel == "det" &
      str_to_lower(sentence_tokens$lemma) == "no"
  )

  hedged_aux <- any(
    sentence_tokens$head_token_id == trigger_token_id &
      sentence_tokens$dep_rel == "aux" &
      str_to_lower(sentence_tokens$lemma) %in% c("may", "might", "could")
  )

  trigger_row <- sentence_tokens |>
    filter(token_id == trigger_token_id)
  reported_parent <- nrow(trigger_row) == 1L &&
    trigger_row$dep_rel == "xcomp" &&
    sentence_tokens$lemma[
      match(trigger_row$head_token_id, sentence_tokens$token_id)
    ] %in% c("report", "say")

  case_when(
    verb_negated | adverb_negated | argument_det_no ~ "negated",
    hedged_aux | reported_parent ~ "hedged",
    TRUE ~ "asserted"
  )
}

extracted_triples <- raw_triples |>
  mutate(
    status = pmap_chr(
      list(sentence_id, trigger_token_id, subject_head_id, object_head_id),
      status_for_trigger
    )
  ) |>
  left_join(
    constructed_sentences |>
      select(sentence_id, document_id, text),
    by = "sentence_id"
  ) |>
  mutate(
    evidence_start = 1L,
    evidence_end = nchar(text),
    evidence_text = str_sub(text, evidence_start, evidence_end)
  ) |>
  select(
    sentence_id,
    document_id,
    relation,
    subject_id,
    object_id,
    status,
    subject_start,
    subject_end,
    object_start,
    object_end,
    evidence_start,
    evidence_end,
    evidence_text
  ) |>
  arrange(sentence_id, relation, subject_id, object_id, status)

knitr::kable(
  extracted_triples,
  col.names = c(
    "Sentence ID", "Document ID", "Relation", "Subject ID", "Object ID",
    "Status", "Subject start", "Subject end", "Object start", "Object end",
    "Evidence start", "Evidence end", "Evidence text"
  ),
  caption = "Extracted triples with status and evidence offsets",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  nrow(raw_triples) > 0L,
  nrow(extracted_triples) >= 8L,
  all(!is.na(extracted_triples$document_id)),
  all(extracted_triples$status %in% c("asserted", "negated", "hedged")),
  all(extracted_triples$subject_start <= extracted_triples$subject_end),
  all(extracted_triples$object_start <= extracted_triples$object_end),
  any(extracted_triples$sentence_id == "c09" & extracted_triples$status == "negated"),
  any(extracted_triples$sentence_id == "c11" & extracted_triples$status == "negated"),
  any(extracted_triples$sentence_id == "c03" & extracted_triples$status == "negated"),
  any(extracted_triples$sentence_id == "c06" & extracted_triples$status == "hedged"),
  !any(extracted_triples$sentence_id == "c13"),
  !any(
    extracted_triples$sentence_id %in% c("s002", "s027") &
      extracted_triples$relation == "requires" &
      extracted_triples$status == "asserted"
  ),
  all(str_detect(
    extracted_triples$evidence_text[extracted_triples$status == "negated"],
    regex("\\b(no|not)\\b", ignore_case = TRUE)
  )),
  all(str_detect(
    extracted_triples$evidence_text[extracted_triples$status == "hedged"],
    regex("\\b(may|might|could|reported)\\b", ignore_case = TRUE)
  ))
)
```

The negation check is deliberately broader than "look for `not` under the verb."
It catches `no longer` through the adverb and `No Forklift Operator Licence`
through the determiner on the credential argument.

## Compare pairs separately from triples

The co-occurrence baseline pairs every two distinct linked mentions in a
constructed sentence. It does not assign a relation type or direction, so those
fields are not applicable. The extractor is scored twice: once for entity-pair
detection and once for labelled triples.

```{r}
cooccurrence_pairs <- constructed_mentions |>
  select(sentence_id, document_id, entity_id) |>
  distinct() |>
  group_by(sentence_id, document_id) |>
  filter(n() >= 2L) |>
  summarise(
    pairs = list(combn(sort(entity_id), 2, simplify = FALSE)),
    .groups = "drop"
  ) |>
  unnest_longer(pairs) |>
  transmute(
    sentence_id,
    document_id,
    first_id = map_chr(pairs, 1),
    second_id = map_chr(pairs, 2),
    relation = "not applicable",
    direction = "not applicable"
  )

reference_pairs <- reference_triples |>
  transmute(
    sentence_id,
    first_id = pmin(subject_id, object_id),
    second_id = pmax(subject_id, object_id)
  ) |>
  distinct()

extractor_pairs <- extracted_triples |>
  transmute(
    sentence_id,
    first_id = pmin(subject_id, object_id),
    second_id = pmax(subject_id, object_id)
  ) |>
  distinct()

score_pairs <- function(system_pairs, system_name) {
  full_join(
    system_pairs |> mutate(predicted = TRUE),
    reference_pairs |> mutate(reference = TRUE),
    by = c("sentence_id", "first_id", "second_id")
  ) |>
    mutate(
      predicted = coalesce(predicted, FALSE),
      reference = coalesce(reference, FALSE),
      system = system_name,
      result = case_when(
        predicted & reference ~ "true positive",
        predicted & !reference ~ "false positive",
        !predicted & reference ~ "false negative",
        TRUE ~ "true negative"
      )
    ) |>
    filter(result != "true negative")
}

pair_results <- bind_rows(
  score_pairs(cooccurrence_pairs |> select(sentence_id, first_id, second_id), "co-occurrence"),
  score_pairs(extractor_pairs, "dependency extractor")
) |>
  arrange(system, sentence_id, first_id, second_id)

triple_results <- full_join(
  extracted_triples |>
    select(sentence_id, relation, subject_id, object_id, status) |>
    mutate(extractor = TRUE),
  reference_triples |>
    mutate(reference = TRUE),
  by = c("sentence_id", "relation", "subject_id", "object_id", "status")
) |>
  mutate(
    extractor = coalesce(extractor, FALSE),
    reference = coalesce(reference, FALSE),
    cooccurrence_relation = "not applicable",
    cooccurrence_direction = "not applicable",
    result = case_when(
      extractor & reference ~ "true positive",
      extractor & !reference ~ "false positive",
      !extractor & reference ~ "false negative",
      TRUE ~ "true negative"
    )
  ) |>
  filter(result != "true negative") |>
  arrange(sentence_id, relation, subject_id, object_id, status)

knitr::kable(
  cooccurrence_pairs,
  col.names = c(
    "Sentence ID", "Document ID", "First ID", "Second ID",
    "Relation", "Direction"
  ),
  caption = "Co-occurrence pairs have no relation type or direction",
  row.names = FALSE
)

knitr::kable(
  pair_results,
  col.names = c(
    "Sentence ID", "First ID", "Second ID", "Predicted",
    "Reference", "System", "Result"
  ),
  caption = "Unlabelled entity-pair results for both systems",
  row.names = FALSE
)

knitr::kable(
  triple_results,
  col.names = c(
    "Sentence ID", "Relation", "Subject ID", "Object ID", "Status",
    "Extractor", "Reference", "Co-occurrence relation",
    "Co-occurrence direction", "Result"
  ),
  caption = "Labelled triple results for the dependency extractor",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(cooccurrence_pairs$relation == "not applicable"),
  all(cooccurrence_pairs$direction == "not applicable"),
  any(pair_results$system == "co-occurrence" & pair_results$result == "false positive"),
  !all(pair_results$result[pair_results$system == "co-occurrence"] == "true positive"),
  all(triple_results$cooccurrence_relation == "not applicable"),
  all(triple_results$cooccurrence_direction == "not applicable"),
  any(triple_results$result == "true positive"),
  any(triple_results$sentence_id == "c13" & triple_results$result == "false negative"),
  !any(triple_results$sentence_id %in% c("s002", "s027"))
)
```

The pair table and triple table answer different questions. Co-occurrence can
ask whether the right two IDs appeared together. It cannot answer which relation
holds, which way the arrow points, or whether the sentence denied the claim.
The reported-verb test sentence is also visible as a false negative for this
small dependency-pattern extractor.

## Where modern systems fit

Current relation-extraction papers often use large language models, prompts, or
trained classifiers. Those systems still need the same schema, argument spans,
status policy, and evidence checks. Published comparisons disagree by domain and
scoring method, so this page does not make a performance claim about them.

## What to remember

- Relation extraction produces typed, directed triples, not loose word matches.
- The argument IDs in this lesson come from the alias table, not from NER.
- `No` on an argument and `no longer` under an adverb can change triple status.
- Co-occurrence pairs have no relation type or direction.
- This same-author score is a demonstration, not a benchmark.
- Every triple needs sentence, document, and offset evidence.

The safe habit is to treat candidate pairs as leads for review and schema
triples as claims that need sentence evidence.

## Sources

- [ACE English relation annotation guidelines v5.8.3](https://www.ldc.upenn.edu/sites/www.ldc.upenn.edu/files/english-relations-guidelines-v5.8.3.pdf)
- [Jurafsky and Martin, Speech and Language Processing, chapter 21](https://web.stanford.edu/~jurafsky/slp3/21.pdf)
- [spacyr `spacy_parse()` documentation](https://spacyr.quanteda.io/reference/spacy_parse.html)
- [Zhang et al. (2017), TACRED](https://aclanthology.org/D17-1004/)
- [Stoica, Platanios, and Poczos (2021), Re-TACRED](https://arxiv.org/abs/2104.08398)
- [Farkas et al. (2010), CoNLL-2010 shared task on uncertainty](https://aclanthology.org/W10-3001/)
- [Wadhwa, Amir, and Wallace (2023), Revisiting relation extraction in the era of LLMs](https://aclanthology.org/2023.acl-long.868/)
- [Jimenez Gutierrez et al. (2022), Biomedical relation extraction with large language models](https://aclanthology.org/2022.findings-emnlp.329/)

```{r}
#| include: false

stopifnot(
  identical(pipeline_info$version, "3.8.0"),
  exists("extracted_triples")
)
invisible(suppressMessages(spacy_finalize()))
```
