---
title: "Answering questions from retrieved passages"
subtitle: "Find candidate evidence first, then ask whether the answer is supported"
description: "Learn how question answering differs from search, and why a cited answer still needs support checks."
categories: [systems, question answering, retrieval]
---

Amara helps at the Riverton Skills Centre desk on days when staff are away from
the phone. A short handbook answers many routine questions, but callers do not
ask in handbook language. One person asks about a "kid" during night class;
another asks a question that assumes a Saturday forklift class exists.

**Question answering** returns a short answer, an abstention, or a correction of
a false premise. Retrieval is only the first step. The system has to choose
passages, read only those passages, cite what it used, and say when the passages
do not answer the question.

::: {.callout-note}
The Riverton Skills Centre handbook is an invented teaching fixture. It does
not describe a real school, program, deadline, stipend, or policy.
:::

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- retrieve handbook passages with tf-idf weights fitted on the handbook;
- show when a gold passage shares no content words with the question;
- compare an extractive sentence baseline with a generated short answer;
- parse a strict two-line model reply without rewriting it;
- keep action correctness, reference agreement, support screens, and human
  review separate; and
- explain why abstention, citations, and fluent wording are not enough.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the local reader and handbook

The handbook, questions, and answer variants were written before this lesson
retrieves or generates anything. The same author wrote all three, so they are a
teaching fixture rather than an independent benchmark.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(huggingfaceR)
  library(knitr)
  library(purrr)
  library(readr)
  library(reticulate)
  library(stringr)
  library(tibble)
  library(tidyr)
})

source("R/use-nlg.R")

handbook <- read_csv(
  "data/riverton/riverton-handbook.csv",
  na = character(),
  col_types = cols(
    passage_id = col_character(),
    topic = col_character(),
    text = col_character(),
    source = col_character(),
    author_note = col_character()
  )
)

questions <- read_csv(
  "data/riverton/riverton-handbook-questions.csv",
  na = c("", "NA"),
  col_types = cols(
    question_id = col_character(),
    question = col_character(),
    probe_type = col_character(),
    expected_action = col_character(),
    gold_passage_id = col_character(),
    acceptable_answers = col_character(),
    author_note = col_character()
  )
) |>
  mutate(
    acceptable_list = str_split(coalesce(acceptable_answers, ""), "\\|")
  )

metadata <- read_csv(
  "data/riverton/riverton-handbook-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_date(),
    rows = col_integer(),
    fingerprint = col_character()
  )
)

qa_model <- load_nlg_pipeline(
  "qwen_1_5b_instruct",
  "text-generation"
)

model_record <- qa_model$metadata |>
  select(model_id, revision, license, task_scope)

kable(
  model_record,
  col.names = c("Model", "Revision", "License", "Use in this course"),
  caption = "Pinned local model used as the passage reader",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c(
    "dplyr", "huggingfaceR", "knitr", "purrr", "readr",
    "reticulate", "stringr", "tibble", "tidyr"
  ) %in% loadedNamespaces()),
  identical(nrow(handbook), 15L),
  identical(nrow(questions), 6L),
  identical(
    questions$expected_action,
    c("answer", "answer", "answer", "answer", "abstain", "flag premise")
  ),
  identical(model_record$model_id, "Qwen/Qwen2.5-1.5B-Instruct"),
  identical(model_record$license, "Apache-2.0"),
  all(metadata$artifact %in% c(
    "riverton-handbook.csv",
    "riverton-handbook-questions.csv",
    "riverton-search-judgments.csv"
  ))
)
```

## Read the handbook once

Each passage keeps its passage ID. Later tables cite these IDs instead of row
positions.

```{r}
kable(
  handbook |>
    select(passage_id, topic, text),
  format = "html",
  escape = TRUE,
  col.names = c("Passage ID", "Topic", "Handbook text"),
  caption = "The complete fictional handbook used for retrieve-then-read question answering",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(handbook$passage_id, sprintf("H%02d", 1:15)),
  !anyDuplicated(handbook$passage_id),
  all(str_detect(handbook$text, "\\S")),
  all(handbook$passage_id %in% c(questions$gold_passage_id, handbook$passage_id))
)
```

## Declare the questions and reference answers

The expected action is defined before retrieval and generation. A gold passage
is the passage that answers the question or contradicts its false premise. The
unanswerable parking question has no gold passage.

```{r}
reference_table <- questions |>
  transmute(
    question_id,
    question,
    expected_action,
    gold_passage_id = coalesce(gold_passage_id, "not applicable"),
    acceptable_answers = coalesce(acceptable_answers, "not applicable")
  )

kable(
  reference_table,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Question",
    "Question text",
    "Expected action",
    "Gold evidence",
    "Acceptable short answers"
  ),
  caption = "Author-written reference actions and answer variants",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(reference_table$question_id, paste0("Q", 1:6)),
  identical(reference_table$expected_action, questions$expected_action),
  identical(reference_table$gold_passage_id[[5]], "not applicable"),
  all(
    questions$gold_passage_id[!is.na(questions$gold_passage_id)] %in%
      handbook$passage_id
  )
)
```

## Check overlap before retrieval

The retriever below uses simple English content tokens and removes the small
stop-word list printed in the code, including `at`, `while`, and other function
words. This choice matters. If stop words stayed in, Q3 would retrieve unrelated
passages because of `at`; with this rule, Q3 has no content-word match and
follows the `no lexical match` path.

```{r}
stop_words <- c(
  "a", "an", "and", "are", "as", "at", "be", "by", "can", "do", "does",
  "for", "from", "how", "i", "in", "is", "it", "my", "of", "on", "or",
  "the", "there", "this", "to", "when", "where", "while", "with", "your"
)

tokenize_text <- function(text) {
  str_extract_all(
    str_to_lower(text, locale = "en"),
    "[\\p{L}\\p{N}][\\p{L}\\p{N}-]*"
  )[[1]]
}

content_words <- function(text) {
  tokenize_text(text) |>
    setdiff(stop_words) |>
    unique() |>
    sort()
}

gold_overlap <- questions |>
  left_join(
    handbook |>
      select(gold_passage_id = passage_id, gold_text = text),
    by = "gold_passage_id"
  ) |>
  mutate(
    question_words = map(question, content_words),
    gold_words = map(gold_text, \(text) {
      if (is.na(text)) character() else content_words(text)
    }),
    shared_words = map2(question_words, gold_words, intersect),
    shared_content_words = map_chr(
      shared_words,
      \(words) if (length(words) == 0L) "none" else paste(words, collapse = ", ")
    )
  ) |>
  select(question_id, gold_passage_id, shared_content_words)

kable(
  gold_overlap,
  col.names = c("Question", "Gold evidence", "Shared content words"),
  caption = "Content-word overlap between each question and its gold passage",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(gold_overlap$question_id, questions$question_id),
  identical(gold_overlap$shared_content_words[[3]], "none"),
  identical(gold_overlap$gold_passage_id[[5]], NA_character_),
  "at" %in% stop_words,
  length(intersect(
    content_words(questions$question[[3]]),
    content_words(handbook$text[handbook$passage_id == "H13"])
  )) == 0L
)
```

## Retrieve candidate passages

Tf-idf gives more weight to terms that appear in fewer handbook passages. The
idf values are fitted once on the handbook and reused for each question. A
passage with score zero is not retrieved, and score ties are broken by passage
ID. The retrieval score is cosine similarity between the question and passage
tf-idf vectors, so it ranges from 0 to 1.

```{r}
passage_terms <- handbook |>
  transmute(
    passage_id,
    term = map(text, tokenize_text)
  ) |>
  unnest(term) |>
  filter(!term %in% stop_words)

document_frequency <- passage_terms |>
  distinct(passage_id, term) |>
  count(term, name = "document_frequency") |>
  mutate(
    idf = log(nrow(handbook) / document_frequency)
  )

passage_vectors <- passage_terms |>
  count(passage_id, term, name = "term_frequency") |>
  left_join(document_frequency, by = "term") |>
  mutate(weight = term_frequency * idf)

rank_question_passages <- function(question_id, question) {
  question_terms <- tibble(term = tokenize_text(question)) |>
    filter(!term %in% stop_words) |>
    count(term, name = "term_frequency") |>
    inner_join(document_frequency, by = "term") |>
    mutate(question_weight = term_frequency * idf)

  if (nrow(question_terms) == 0L) {
    return(handbook |>
      transmute(question_id = question_id, passage_id, score = 0))
  }

  question_norm <- sqrt(sum(question_terms$question_weight^2))
  passage_norms <- passage_vectors |>
    group_by(passage_id) |>
    summarise(passage_norm = sqrt(sum(weight^2)), .groups = "drop")

  scored <- passage_vectors |>
    inner_join(question_terms |> select(term, question_weight), by = "term") |>
    group_by(passage_id) |>
    summarise(dot_product = sum(weight * question_weight), .groups = "drop") |>
    left_join(passage_norms, by = "passage_id") |>
    mutate(score = dot_product / (passage_norm * question_norm)) |>
    select(passage_id, score)

  handbook |>
    select(passage_id) |>
    left_join(scored, by = "passage_id") |>
    mutate(
      score = coalesce(score, 0),
      question_id = question_id
    ) |>
    select(question_id, passage_id, score)
}

ranked_passages <- map2_dfr(
  questions$question_id,
  questions$question,
  rank_question_passages
) |>
  arrange(question_id, desc(score), passage_id) |>
  group_by(question_id) |>
  mutate(display_rank = if_else(score > 0, row_number(), NA_integer_)) |>
  ungroup()

retrieved <- ranked_passages |>
  filter(score > 0, display_rank <= 3L) |>
  transmute(question_id, passage_id, rank = display_rank, score)

gold_ranks <- ranked_passages |>
  right_join(
    questions |> select(question_id, gold_passage_id),
    by = c("question_id", "passage_id" = "gold_passage_id")
  ) |>
  transmute(
    question_id,
    gold_rank = display_rank,
    gold_score = score,
    gold_retrieval_label = case_when(
      is.na(passage_id) ~ "not applicable",
      coalesce(gold_score, 0) == 0 ~ "not retrieved (all scores 0)",
      !is.na(gold_rank) ~ paste0("rank ", gold_rank),
      TRUE ~ "not retrieved"
    )
  )

retrieval_display <- questions |>
  select(question_id, question, expected_action, gold_passage_id) |>
  left_join(
    retrieved |>
      group_by(question_id) |>
      summarise(
        retrieved_ids = paste(passage_id, collapse = ", "),
        retrieved_ranks = paste(
          paste0(passage_id, " (#", rank, ", ", sprintf("%.3f", score), ")"),
          collapse = "; "
        ),
        .groups = "drop"
      ),
    by = "question_id"
  ) |>
  left_join(gold_ranks, by = "question_id") |>
  mutate(
    retrieved_ids = if_else(is.na(retrieved_ids), "no lexical match", retrieved_ids),
    retrieved_ranks = if_else(is.na(retrieved_ranks), "no lexical match", retrieved_ranks),
    gold_passage_id = coalesce(gold_passage_id, "not applicable")
  )

kable(
  retrieval_display |>
    select(question_id, question, retrieved_ranks, gold_passage_id, gold_retrieval_label),
  format = "html",
  escape = TRUE,
  col.names = c(
    "Question",
    "Question text",
    "Retrieved candidates",
    "Gold evidence",
    "Gold evidence retrieval"
  ),
  caption = "Tf-idf retrieval results with gold evidence status",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(retrieved$score > 0),
  all(retrieved$rank <= 3L),
  identical(retrieval_display$retrieved_ids[[3]], "no lexical match"),
  identical(retrieval_display$gold_retrieval_label[[3]], "not retrieved (all scores 0)"),
  identical(retrieval_display$gold_score[[3]], 0),
  identical(retrieval_display$gold_retrieval_label[[5]], "not applicable"),
  all(
    retrieved |>
      group_by(question_id) |>
      summarise(ok = all(order(rank) == seq_along(rank)), .groups = "drop") |>
      pull(ok)
  )
)
```

Q3 shows the retrieval problem. The correct evidence is H13, but a lexical
retriever that shares no content word with the question gives the reader no
passage to read.

## Add a sentence baseline

An extractive baseline returns one sentence from the highest-ranked passage.
It is scored by a different unit from the generated answer: whether that
sentence contains one acceptable answer variant.

```{r}
normalize_answer <- function(text) {
  text |>
    str_to_lower(locale = "en") |>
    str_replace_all("[\\p{P}\\p{S}]", "") |>
    str_replace_all("\\b(a|an|the)\\b", " ") |>
    str_squish()
}

contains_variant <- function(text, variants) {
  variants <- variants[nzchar(variants)]
  if (length(variants) == 0L || is.na(text)) {
    return(NA)
  }
  normalized_text <- normalize_answer(text)
  any(map_lgl(
    variants,
    \(variant) str_detect(
      normalized_text,
      fixed(normalize_answer(variant))
    )
  ))
}

top_passage <- retrieved |>
  filter(rank == 1L) |>
  left_join(handbook |> select(passage_id, text), by = "passage_id") |>
  mutate(
    baseline_sentence = map_chr(
      str_split(text, "(?<=[.!?])\\s+"),
      \(sentences) sentences[[1]]
    )
  ) |>
  select(question_id, baseline_sentence)

extractive_baseline <- questions |>
  select(question_id, expected_action, acceptable_list) |>
  left_join(top_passage, by = "question_id") |>
  mutate(
    contains_acceptable_answer = map2_lgl(
      baseline_sentence,
      acceptable_list,
      contains_variant
    ),
    contains_acceptable_answer = if_else(
      expected_action == "answer",
      coalesce(contains_acceptable_answer, FALSE),
      NA
    ),
    baseline_sentence = if_else(
      expected_action == "answer" & is.na(baseline_sentence),
      "no sentence (no lexical match)",
      baseline_sentence
    )
  ) |>
  select(question_id, baseline_sentence, contains_acceptable_answer)

kable(
  extractive_baseline,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Question",
    "Sentence returned by the baseline",
    "Contains an acceptable short answer"
  ),
  caption = "Extractive sentence baseline scored in sentence units",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(extractive_baseline$question_id, questions$question_id),
  identical(extractive_baseline$baseline_sentence[[3]], "no sentence (no lexical match)"),
  identical(extractive_baseline$contains_acceptable_answer[[3]], FALSE),
  all(is.na(extractive_baseline$contains_acceptable_answer[5:6])),
  all(extractive_baseline$contains_acceptable_answer[1:2] %in% c(TRUE, FALSE))
)
```

The baseline is supported by construction because it copies a sentence. It can
still fail to answer the question, and it has no way to explain a false premise.

## Ask the reader to produce two lines

The generated reader sees the question and the retrieved passages only. If no
passage is retrieved, that fact is placed in the prompt. The required output has
exactly two lines: `Answer:` and `Sources:`.

```{r}
reader_system <- paste(
  "You are a strict formatter for handbook question answering.",
  "Your reply must begin with Answer: on the first line.",
  "The second and final line must begin with Sources:.",
  "Use only IDs that appear in PASSAGES.",
  "If no listed passage answers the question, write Answer: NOT IN HANDBOOK",
  "and Sources: none.",
  "If the question assumes a fact contradicted by a listed passage, write",
  "Answer: PREMISE NOT SUPPORTED: plus a short correction.",
  "Do not write a blank line, bullet, explanation, or extra source."
)

format_passages <- function(question_id) {
  rows <- retrieved |>
    filter(.data$question_id == .env$question_id) |>
    arrange(rank) |>
    left_join(handbook |> select(passage_id, text), by = "passage_id")

  if (nrow(rows) == 0L) {
    return("No passages were retrieved.")
  }

  rows |>
    transmute(line = paste0(passage_id, ": ", text)) |>
    pull(line) |>
    paste(collapse = "\n")
}

reader_inputs <- questions |>
  mutate(
    passage_block = map_chr(question_id, format_passages),
    user_prompt = paste0(
      "QUESTION ID: ", question_id,
      "\nQUESTION: ", question,
      "\nPASSAGES:\n", passage_block,
      "\n\nOUTPUT FORMAT:\nAnswer: <short answer, NOT IN HANDBOOK, or PREMISE NOT SUPPORTED: correction>",
      "\nSources: <listed passage IDs, or none>"
    ),
    model_prompt = map_chr(
      user_prompt,
      \(user) nlg_chat_prompt(qa_model$tokenizer, reader_system, user)
    ),
    input_tokens = map_int(
      model_prompt,
      \(prompt) nlg_token_count(qa_model$tokenizer, prompt)
    ),
    max_new_tokens = 48L
  )

nlg_assert_input_budget(
  reader_inputs$input_tokens,
  max_input_tokens = 768L,
  item_ids = reader_inputs$question_id
)

kable(
  reader_inputs |>
    select(question_id, input_tokens, max_new_tokens),
  col.names = c("Question", "Prompt tokens", "Maximum new tokens"),
  caption = "Input budgets checked before each reader call",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(reader_inputs$input_tokens > 0L),
  all(reader_inputs$input_tokens <= 768L),
  identical(reader_inputs$max_new_tokens, rep(48L, 6L))
)
```

Greedy decoding chooses the highest-scoring next token at each step. The helper
sets `do_sample = FALSE`. This lesson passes a repetition penalty of 1.05,
which overrides the checkpoint's generation-config value of 1.1.
Exact generated text can differ across machines; the page does not depend on
literal wording.

```{r}
reader_results <- reader_inputs |>
  mutate(
    generation = map2(
      model_prompt,
      max_new_tokens,
      \(prompt, budget) {
        nlg_generate(
          qa_model,
          prompt,
          max_new_tokens = budget,
          repetition_penalty = 1.05
        )
      }
    ),
    raw_output = map_chr(generation, "text"),
    output_tokens = map_int(generation, "output_tokens"),
    ended_by_eos = map_lgl(generation, "ended_by_eos"),
    hit_token_cap = map_lgl(generation, "hit_token_cap")
  )

kable(
  reader_results |>
    mutate(raw_output_visible = str_replace_all(raw_output, "\\r?\\n", " ⏎ ")) |>
    select(question_id, raw_output_visible, output_tokens, ended_by_eos, hit_token_cap),
  format = "html",
  escape = TRUE,
  col.names = c(
    "Question",
    "Raw reader output (`⏎` marks a line break)",
    "Output tokens",
    "EOS emitted",
    "Hit token cap"
  ),
  caption = "Raw local model output with visible line-break markers and stopping diagnostics",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(nzchar(reader_results$raw_output)),
  all(reader_results$output_tokens > 0L),
  all(reader_results$output_tokens <= 48L),
  all(reader_results$ended_by_eos | reader_results$hit_token_cap)
)
```

## Parse without rewriting

A strict parser accepts only the two requested lines. In this render every
output parsed; if a later output adds a third line or omits `Sources:`, the
lesson stops before publishing a malformed row.

```{r}
parse_reader_output <- function(text) {
  parsed <- str_match(
    str_trim(text),
    "^Answer: ([^\\r\\n]+)\\r?\\nSources: ([^\\r\\n]+)$"
  )

  if (is.na(parsed[1, 1])) {
    return(tibble(
      parse_ok = FALSE,
      parsed_answer = NA_character_,
      parsed_sources = list(character()),
      observed_action = "parse failure"
    ))
  }

  answer <- str_squish(parsed[1, 2])
  source_text <- str_squish(parsed[1, 3])
  sources <- if (str_to_lower(source_text) == "none") {
    character()
  } else {
    str_split(source_text, "\\s*,\\s*")[[1]]
  }
  action <- case_when(
    str_starts(answer, "NOT IN HANDBOOK") ~ "abstain",
    str_starts(answer, "PREMISE NOT SUPPORTED") ~ "flag premise",
    TRUE ~ "answer"
  )

  tibble(
    parse_ok = TRUE,
    parsed_answer = answer,
    parsed_sources = list(sources),
    observed_action = action
  )
}

parsed_results <- reader_results |>
  select(question_id, raw_output, output_tokens, ended_by_eos, hit_token_cap) |>
  mutate(parsed = map(raw_output, parse_reader_output)) |>
  unnest(parsed) |>
  left_join(questions, by = "question_id")

parsed_display <- parsed_results |>
  transmute(
    question_id,
    parsed_answer = coalesce(parsed_answer, "parse failure"),
    parsed_sources = map_chr(
      parsed_sources,
      \(sources) if (length(sources) == 0L) "none" else paste(sources, collapse = ", ")
    ),
    observed_action,
    expected_action
  )

kable(
  parsed_display,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Question",
    "Parsed answer",
    "Parsed sources",
    "Observed action",
    "Expected action"
  ),
  caption = "Strictly parsed reader outputs",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(parsed_results$question_id, questions$question_id),
  all(parsed_results$parse_ok),
  all(parsed_results$observed_action %in% c("answer", "abstain", "flag premise")),
  all(map_lgl(parsed_results$parsed_sources, \(sources) {
    all(sources %in% handbook$passage_id) || length(sources) == 0L
  }))
)
```

## Score answers, actions, and retrieval

**Exact match (EM)** is 1 only when the normalized prediction exactly equals one
acceptable answer. **Token F1** balances shared-token precision and recall after
normalization. SQuAD is a widely used question-answering benchmark; this lesson
copies its scoring steps: lowercase, delete punctuation and symbols, remove
English articles, and squeeze spaces. The only difference is that stringr's ICU
class removes Unicode punctuation and symbols, not only ASCII punctuation.

```{r}
answer_f1 <- function(prediction, reference) {
  prediction_tokens <- str_split(normalize_answer(prediction), "\\s+")[[1]]
  reference_tokens <- str_split(normalize_answer(reference), "\\s+")[[1]]
  prediction_tokens <- prediction_tokens[nzchar(prediction_tokens)]
  reference_tokens <- reference_tokens[nzchar(reference_tokens)]

  if (length(prediction_tokens) == 0L || length(reference_tokens) == 0L) {
    return(0)
  }

  common <- intersect(unique(prediction_tokens), unique(reference_tokens))
  overlap <- sum(pmin(
    tabulate(match(prediction_tokens, common), nbins = length(common)),
    tabulate(match(reference_tokens, common), nbins = length(common))
  ))

  if (overlap == 0L) {
    return(0)
  }

  precision <- overlap / length(prediction_tokens)
  recall <- overlap / length(reference_tokens)
  2 * precision * recall / (precision + recall)
}

best_reference_scores <- function(answer, variants) {
  variants <- variants[nzchar(variants)]
  if (length(variants) == 0L || is.na(answer)) {
    return(tibble(exact_match = NA_real_, f1 = NA_real_))
  }

  tibble(
    exact_match = as.numeric(map_lgl(
      variants,
      \(variant) identical(normalize_answer(answer), normalize_answer(variant))
    )),
    f1 = map_dbl(variants, \(variant) answer_f1(answer, variant))
  ) |>
    summarise(
      exact_match = max(exact_match),
      f1 = max(f1),
      .groups = "drop"
    )
}

retrieval_success <- questions |>
  select(question_id, expected_action, gold_passage_id) |>
  left_join(gold_ranks, by = "question_id") |>
  mutate(
    recall_at_3 = case_when(
      is.na(gold_passage_id) ~ NA,
      gold_rank <= 3L ~ TRUE,
      TRUE ~ FALSE
    )
  )

score_table <- parsed_results |>
  mutate(
    answer_for_reference_score = if_else(
      observed_action == "answer" & parse_ok,
      parsed_answer,
      ""
    ),
    reference_scores = map2(
      answer_for_reference_score,
      acceptable_list,
      best_reference_scores
    )
  ) |>
  unnest(reference_scores) |>
  left_join(retrieval_success, by = c("question_id", "expected_action", "gold_passage_id")) |>
  mutate(
    action_correct = observed_action == expected_action,
    reader_action = observed_action,
    reader_given_passages = case_when(
      expected_action == "answer" & recall_at_3 %in% FALSE & observed_action == "abstain" ~
        "reasonable abstention from retrieved passages",
      action_correct ~ "expected action",
      TRUE ~ "unexpected action"
    ),
    end_to_end_outcome = case_when(
      expected_action == "answer" & recall_at_3 %in% FALSE ~ "failed before reading",
      action_correct ~ "expected action",
      TRUE ~ "unexpected action"
    ),
    reference_agreement = case_when(
      expected_action != "answer" ~ "not applicable",
      is.na(exact_match) ~ "not applicable",
      TRUE ~ sprintf("EM %.0f; F1 %.2f", exact_match, f1)
    ),
    recall_at_3_label = case_when(
      is.na(recall_at_3) ~ "not applicable",
      recall_at_3 ~ "gold in top 3",
      TRUE ~ "gold not in top 3"
    )
  )

acceptable_variant_in_answer <- function(answer, variants) {
  variants <- variants[nzchar(variants)]
  if (length(variants) == 0L || is.na(answer) || !nzchar(answer)) {
    return(FALSE)
  }
  normalized_answer <- normalize_answer(answer)
  any(map_lgl(
    variants,
    \(variant) str_detect(
      normalized_answer,
      fixed(normalize_answer(variant))
    )
  ))
}

em_zero_reason_text <- function(rows, count, partial_overlap, variant_in_answer) {
  if (count == 0L) {
    return("no correctly handled answer row has exact match 0")
  }
  verb <- if (count == 1L) " has" else " have"
  if (all(partial_overlap) && all(variant_in_answer)) {
    return(paste0(
      rows,
      verb,
      " exact match 0 because ",
      if (count == 1L) "the answer contains" else "each answer contains",
      " an acceptable variant but includes extra words"
    ))
  }
  paste0(rows, verb, " exact match 0 for wording differences")
}

right_action_em_zero_rows <- score_table |>
  filter(expected_action == "answer", action_correct, exact_match == 0) |>
  mutate(
    acceptable_variant_in_answer = map2_lgl(
      parsed_answer,
      acceptable_list,
      acceptable_variant_in_answer
    ),
    partial_overlap = f1 > 0
  )

right_action_em_zero <- tibble(
  rows = paste(right_action_em_zero_rows$question_id, collapse = ", "),
  count = nrow(right_action_em_zero_rows),
  reason = em_zero_reason_text(
    rows,
    count,
    right_action_em_zero_rows$partial_overlap,
    right_action_em_zero_rows$acceptable_variant_in_answer
  )
)

kable(
  score_table |>
    select(
      question_id,
      reader_action,
      expected_action,
      action_correct,
      reader_given_passages,
      reference_agreement,
      recall_at_3_label,
      end_to_end_outcome
    ),
  col.names = c(
    "Question",
    "Reader action",
    "Expected action",
    "Action correct",
    "Reader given retrieved passages",
    "Reference agreement",
    "Retrieval recall at 3",
    "End-to-end outcome"
  ),
  caption = "Action, reference, retrieval, and end-to-end results kept separate",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(score_table), 6L),
  all(score_table$action_correct %in% c(TRUE, FALSE)),
  identical(score_table$recall_at_3_label[[5]], "not applicable"),
  identical(score_table$recall_at_3[[3]], FALSE),
  identical(
    score_table$reader_given_passages[[3]],
    "reasonable abstention from retrieved passages"
  ),
  all(score_table$reference_agreement[5:6] == "not applicable"),
  all(score_table$exact_match[score_table$expected_action != "answer"] |> is.na()),
  identical(score_table$reader_action[[3]], "abstain"),
  identical(score_table$answer_for_reference_score[[3]], ""),
  identical(score_table$exact_match[[3]], 0),
  identical(score_table$f1[[3]], 0),
  all(right_action_em_zero_rows$partial_overlap),
  all(right_action_em_zero_rows$acceptable_variant_in_answer),
  identical(right_action_em_zero$count, nrow(right_action_em_zero_rows)),
  identical(
    em_zero_reason_text("", 0L, logical(), logical()),
    "no correctly handled answer row has exact match 0"
  ),
  identical(
    em_zero_reason_text("Q2", 1L, TRUE, TRUE),
    paste0(
      "Q2 has exact match 0 because the answer contains an ",
      "acceptable variant but includes extra words"
    )
  ),
  identical(
    em_zero_reason_text("Q2, Q4", 2L, c(TRUE, TRUE), c(TRUE, TRUE)),
    paste0(
      "Q2, Q4 have exact match 0 because each answer contains an ",
      "acceptable variant but includes extra words"
    )
  )
)

qa_action_summary <- score_table |>
  summarise(
    answer_actions = sum(reader_action == "answer"),
    abstain_actions = sum(reader_action == "abstain"),
    premise_actions = sum(reader_action == "flag premise"),
    exact_action_matches = sum(action_correct),
    retrieval_failures = sum(end_to_end_outcome == "failed before reading"),
    .groups = "drop"
  )
```

In this render, the parsed reader actions include
`r qa_action_summary$answer_actions` answer actions,
`r qa_action_summary$abstain_actions` abstentions, and
`r qa_action_summary$premise_actions` premise flags. Exact action matched the
predeclared reference on `r qa_action_summary$exact_action_matches` of 6 rows,
and `r qa_action_summary$retrieval_failures` row failed before reading because
the gold passage was not retrieved. The generated short answer receives EM and
F1 only when the expected action is `answer`; parsed abstentions and parse
failures count as empty predictions, so they receive EM 0 and F1 0 on
answerable questions. The sentence baseline used a different unit, whether the
copied sentence contained an acceptable answer. These numbers should not be read
as a model comparison. In this render, `r right_action_em_zero$reason`.

## Screen citations and surface support

The support screen checks six mechanical facts: the row parsed, cited IDs
exist, cited IDs were retrieved, an answered row has a citation, an abstention
cites no passages, and answer numerals appear in cited passages after citation
tokens are removed. False-premise handling is scored in the action table using
the author's reference; no screen on this page detects a false premise from text
alone. A screen can miss number words and wrong attachments, such as using the
right number for the wrong item.

```{r}
strip_citation_tokens <- function(text) {
  str_remove_all(text, "\\bH\\d{2}\\b")
}

numbers_in <- function(text) {
  str_extract_all(
    strip_citation_tokens(coalesce(text, "")),
    "(?<![\\p{L}\\p{N}])\\d+"
  )[[1]]
}

screen_one <- function(answer, sources, parse_ok, observed_action, retrieved_ids) {
  cited_text <- handbook |>
    filter(passage_id %in% sources) |>
    pull(text) |>
    paste(collapse = " ")
  answer_numbers <- numbers_in(answer)
  cited_numbers <- numbers_in(cited_text)

  parsed <- isTRUE(parse_ok)
  cited_exist <- all(sources %in% handbook$passage_id)
  cited_retrieved <- all(sources %in% retrieved_ids)
  answer_has_citation <- if (observed_action == "answer") {
    length(sources) > 0L
  } else {
    TRUE
  }
  abstain_sources_ok <- if (observed_action == "abstain") {
    length(sources) == 0L
  } else {
    TRUE
  }
  numbers_supported <- if (length(answer_numbers) == 0L) {
    TRUE
  } else {
    all(answer_numbers %in% cited_numbers)
  }

  c(
    parsed = parsed,
    cited_exist = cited_exist,
    cited_retrieved = cited_retrieved,
    answer_has_citation = answer_has_citation,
    abstain_sources_ok = abstain_sources_ok,
    numbers_supported = numbers_supported
  )
}

screen_results <- parsed_results |>
  left_join(
    retrieved |>
      group_by(question_id) |>
      summarise(retrieved_ids = list(passage_id), .groups = "drop"),
    by = "question_id"
  ) |>
  mutate(
    retrieved_ids = map(retrieved_ids, \(ids) if (is.null(ids)) character() else ids),
    screen_values = pmap(
      list(parsed_answer, parsed_sources, parse_ok, observed_action, retrieved_ids),
      screen_one
    ),
    surface_support_screen = map_lgl(screen_values, all),
    failed_checks = map_chr(
      screen_values,
      \(values) {
        failed <- names(values)[!values]
        if (length(failed) == 0L) "none" else paste(failed, collapse = ", ")
      }
    ),
    screen_label = nlg_screen_label(surface_support_screen),
    human_support_review = "pending"
  )

screen_display <- screen_results |>
  transmute(
    question_id,
    parsed_answer,
    cited_passages = map_chr(
      parsed_sources,
      \(sources) if (length(sources) == 0L) "none" else paste(sources, collapse = ", ")
    ),
    cited_text = map_chr(
      parsed_sources,
      \(sources) {
        text <- handbook |>
          filter(passage_id %in% sources) |>
          transmute(line = paste0(passage_id, ": ", text)) |>
          pull(line)
        if (length(text) == 0L) "none" else paste(text, collapse = " ")
      }
    ),
    failed_checks,
    screen_label,
    human_support_review
  )

kable(
  screen_display,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Question",
    "Parsed answer",
    "Cited IDs",
    "Cited passage text",
    "Failed screen checks",
    "Surface support screen",
    "Human support review"
  ),
  caption = "Cited text and support-screen labels for every reader output",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(screen_results$screen_label %in% c(
    "not flagged by screen",
    "held for human review"
  )),
  identical(screen_results$human_support_review, rep("pending", 6L)),
  is.logical(screen_results$surface_support_screen),
  all(map_lgl(screen_results$screen_values, \(values) {
    identical(names(values), c(
      "parsed",
      "cited_exist",
      "cited_retrieved",
      "answer_has_citation",
      "abstain_sources_ok",
      "numbers_supported"
    ))
  })),
  identical(numbers_in("October 16th"), "16"),
  length(numbers_in("See H08 and H14.")) == 0L,
  !screen_one(
    "Applications close on October 16th.",
    c("H14"),
    TRUE,
    "answer",
    c("H14")
  )[["numbers_supported"]]
)

qa_screen_summary <- screen_results |>
  count(screen_label, name = "questions") |>
  complete(
    screen_label = c("not flagged by screen", "held for human review"),
    fill = list(questions = 0L)
  )
```

The plain-language human support test is: would someone who read only the cited
passage agree, "according to this passage, this answer is supported"? This page
does not make that human judgment. In this render, the surface screen labels
`r qa_screen_summary$questions[qa_screen_summary$screen_label == "held for human review"]`
reader rows as held for review and
`r qa_screen_summary$questions[qa_screen_summary$screen_label == "not flagged by screen"]`
as not flagged by the screen.

## Show screen controls

If every generated answer looks clean in one render, a reader still needs to see
the screen catch something. The next two rows are author-written control
strings, not model outputs.

```{r}
control_outputs <- tibble(
  control_id = c("control-unsupported-number", "control-unretrieved-citation"),
  question_id = c("Q1", "Q1"),
  parsed_answer = c(
    "The weekly training stipend is $200.",
    "The stipend is paid weekly."
  ),
  parsed_sources = list(c("H08"), c("H14")),
  parse_ok = TRUE,
  observed_action = "answer"
) |>
  left_join(
    retrieved |>
      group_by(question_id) |>
      summarise(retrieved_ids = list(passage_id), .groups = "drop"),
    by = "question_id"
  ) |>
  mutate(
    screen_values = pmap(
      list(parsed_answer, parsed_sources, parse_ok, observed_action, retrieved_ids),
      screen_one
    ),
    surface_support_screen = map_lgl(screen_values, all),
    failed_checks = map_chr(
      screen_values,
      \(values) {
        failed <- names(values)[!values]
        if (length(failed) == 0L) "none" else paste(failed, collapse = ", ")
      }
    ),
    screen_label = nlg_screen_label(surface_support_screen)
  )

kable(
  control_outputs |>
    transmute(
      control_id,
      parsed_answer,
      cited_passages = map_chr(parsed_sources, paste, collapse = ", "),
      failed_checks,
      screen_label
    ),
  format = "html",
  escape = TRUE,
  col.names = c("Control", "Control answer", "Cited IDs", "Failed screen checks", "Screen label"),
  caption = "Author-written control strings that the support screen holds for review",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(
    control_outputs$screen_label,
    c("held for human review", "held for human review")
  ),
  identical(control_outputs$failed_checks[[1]], "numbers_supported"),
  identical(control_outputs$failed_checks[[2]], "cited_retrieved"),
  !any(control_outputs$surface_support_screen)
)
```

These controls do not prove that the screen is complete. They show only that two
known failure shapes are caught.

## Count action choices

The fixture has three expected actions, so the table keeps all three:
`answer`, `abstain`, and `flag premise`. The simple-policy comparison uses the
same exact-action basis. A policy that always answers matches the four answer
rows; a policy that always abstains matches only Q5, because the false-premise
row is a separate expected action.

```{r}
action_table <- score_table |>
  count(expected_action, reader_action, name = "questions") |>
  complete(
    expected_action = c("answer", "abstain", "flag premise"),
    reader_action = c("answer", "abstain", "flag premise"),
    fill = list(questions = 0L)
  )

always_answer_exact <- sum(questions$expected_action == "answer")
always_abstain_exact <- sum(questions$expected_action == "abstain")
observed_reader_exact <- sum(score_table$action_correct)

policy_misses <- bind_rows(
  score_table |>
    filter(!action_correct) |>
    summarise(policy = "observed reader", misses = paste(question_id, collapse = ", ")),
  questions |>
    filter(expected_action != "answer") |>
    summarise(policy = "always answer", misses = paste(question_id, collapse = ", ")),
  questions |>
    filter(expected_action != "abstain") |>
    summarise(policy = "always abstain", misses = paste(question_id, collapse = ", "))
)

kable(
  action_table,
  col.names = c("Expected action", "Observed action", "Questions"),
  caption = "Expected versus observed action for all six questions",
  row.names = FALSE
)

kable(
  tibble(
    baseline = c("observed reader", "always answer", "always abstain"),
    exact_action_matches = c(
      observed_reader_exact,
      always_answer_exact,
      always_abstain_exact
    ),
    total_questions = nrow(questions)
  ) |>
    left_join(policy_misses, by = c("baseline" = "policy")),
  col.names = c("Policy", "Exact action matches", "Questions", "Rows missed"),
  caption = "Observed reader and simple policies scored on the same six questions",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  sum(action_table$questions) == 6L,
  identical(observed_reader_exact, 4L),
  identical(always_answer_exact, 4L),
  identical(always_abstain_exact, 1L),
  all(action_table$expected_action %in% c("answer", "abstain", "flag premise")),
  all(action_table$reader_action %in% c("answer", "abstain", "flag premise"))
)
```

The abstention row matters because an unanswerable question alone can flatter a
system that refuses everything. The premise row matters too: a false assumption
needs a correction, not just a generic refusal. The observed reader and the
always-answer policy both match `r observed_reader_exact` of 6 actions, but they
miss different rows: the reader misses `r policy_misses$misses[policy_misses$policy == "observed reader"]`,
while always-answer misses `r policy_misses$misses[policy_misses$policy == "always answer"]`.

## How to use this pattern

This is a retrieve-then-read demonstration over a small English (`en`)
handbook. The term retrieval-augmented generation, or RAG, usually refers to a
broader family of systems that combine a generator with an external memory; the
original RAG paper used a dense Wikipedia index and trained components together.
This lesson does neither.

Questions about money, eligibility, or deadlines need a person to check the
source before anyone acts. Typed questions are user text; this page stores none
after the render. Retrieved passages are evidence, not instructions. The next
lesson shows why dialogue systems should keep safety and routing rules in code
rather than trusting retrieved or typed text.

## What to remember

- Question answering is retrieval plus reading, not retrieval alone.
- A lexical retriever can miss an answerable question before the reader starts.
- EM and F1 score short answer strings; they do not measure support.
- A citation is a clue to inspect, not proof that the answer follows.
- Screens can hold obvious failures, but human support review stays pending.
- Abstention must be checked on answerable and unanswerable questions.

## Sources

- [Rajpurkar, Jia, and Liang on unanswerable questions for SQuAD 2.0](https://aclanthology.org/P18-2124/)
- [Rajpurkar et al. on SQuAD exact match and F1](https://aclanthology.org/D16-1264/)
- [Chen et al. on retrieval plus reading in DrQA](https://aclanthology.org/P17-1171/)
- [SQuAD explorer and evaluation script](https://rajpurkar.github.io/SQuAD-explorer/)
- [Qwen2.5-1.5B-Instruct model card](https://huggingface.co/Qwen/Qwen2.5-1.5B-Instruct)
- [Liu, Zhang, and Liang on verifiability in generative search](https://aclanthology.org/2023.findings-emnlp.467/)
- [Kim et al. on question answering with questionable assumptions](https://aclanthology.org/2023.acl-long.472/)
- [Kamalloo et al. on open-domain QA evaluation with lexical metrics](https://aclanthology.org/2023.acl-long.307/)
- [Rashkin et al. on attribution to identified sources](https://aclanthology.org/2023.cl-4.2/)
- [Lewis et al. on retrieval-augmented generation](https://proceedings.neurips.cc/paper/2020/hash/6b493230205f780e1bc26945df7481e5-Abstract.html)
