---
title: "Adding facts to a knowledge base"
subtitle: "Keep provenance, conflicts, NIL records, and review decisions visible"
description: "Learn how knowledge base population links mentions, adds typed facts, and records why some candidates stay out of the knowledge base."
categories: [systems, knowledge base population, entity linking]
---

Grace receives a week of short announcements about Riverton training programs.
Some repeat facts the team already has. One gives a new deadline. Another uses
the name `Riverton`, which could mean two different places in the reference
table.

**Knowledge base population** reads documents, links entity mentions, and adds
typed facts to a knowledge base with provenance. A knowledge base is not a truth
machine. It is a structured record of claims, sources, conflicts, and review
decisions.

::: {.callout-note}
The Riverton records in this lesson are fictional teaching examples about training programs, places, and organisations.
:::

::: {.callout-tip}
## What you will learn

By the end of this lesson, you will be able to:

- distinguish entity linking from fact insertion;
- declare single-valued and list-valued slots before merging;
- keep document dates, evidence text, and offsets for every fact;
- merge duplicate facts without losing provenance;
- keep conflicting dated values instead of overwriting them; and
- route negated, ambiguous, NIL, and type-failed candidates to review.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the Riverton reference list

The existing knowledge base starts with entity IDs and aliases. The lesson reads
the local CSV files with fixed column types and checks their recorded
fingerprints before using them.

```{r}
library(readr)
library(dplyr)
library(tibble)
library(tidyr)
library(purrr)
library(stringr)
library(digest)
library(knitr)

entities <- read_csv(
  "data/riverton/riverton-entities.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    canonical_name = col_character(),
    entity_type = col_character(),
    description = col_character()
  )
) |>
  select(entity_id, canonical_name, entity_type)

aliases <- read_csv(
  "data/riverton/riverton-aliases.csv",
  na = character(),
  col_types = cols(
    entity_id = col_character(),
    alias = col_character()
  )
) |>
  left_join(entities, by = "entity_id")

metadata <- read_csv(
  "data/riverton/riverton-reference-metadata.csv",
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    fingerprint = col_character()
  )
)

entity_hash <- digest(
  paste(read_lines("data/riverton/riverton-entities.csv"), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)
alias_hash <- digest(
  paste(read_lines("data/riverton/riverton-aliases.csv"), collapse = "\n"),
  algo = "sha256",
  serialize = FALSE
)

reference_summary <- tibble(
  artifact = c("riverton-entities.csv", "riverton-aliases.csv"),
  rows = c(nrow(entities), nrow(aliases)),
  sha256 = c(entity_hash, alias_hash)
)

knitr::kable(
  entities,
  col.names = c("Entity ID", "Canonical name", "Type"),
  caption = "Existing Riverton entity IDs used by the population step",
  row.names = FALSE
)

knitr::kable(
  reference_summary,
  col.names = c("Artifact", "Rows", "SHA-256"),
  caption = "Reference files checked before use",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(entities), 8L),
  identical(nrow(aliases), 17L),
  identical(
    entity_hash,
    metadata$fingerprint[metadata$artifact == "riverton-entities.csv"]
  ),
  identical(
    alias_hash,
    metadata$fingerprint[metadata$artifact == "riverton-aliases.csv"]
  ),
  sum(aliases$alias == "Riverton") == 2L,
  !any(str_detect(entities$canonical_name, "job board"))
)
```

The table prints names and types only. It does not repeat the description field
for the forklift credential because this lesson is about provenance, and that
description is not evidence for a job-board claim.

## Declare slot rules before merging

A **slot** is a typed place where the knowledge base stores a value. Some slots
can hold a list. Others are single-valued for a named time period, so a second
dated value becomes a conflict to review.

```{r}
slot_schema <- tribble(
  ~slot, ~subject_type, ~object_type, ~cardinality, ~merge_rule,
  "offers", "organisation", "credential", "list-valued",
  "One fact per subject-object pair; repeated evidence adds provenance.",
  "application_deadline", "organisation", "date literal", "single-valued",
  "Keep each dated value, flag conflict, and let review decide.",
  "located_in", "organisation", "place", "single-valued",
  "Keep source and date; do not infer a new location from an ambiguous name.",
  "serves", "organisation", "place", "list-valued",
  "One fact per served place.",
  "near", "place", "place", "list-valued",
  "One directional reference edge for this teaching graph."
)

knitr::kable(
  slot_schema,
  col.names = c("Slot", "Subject type", "Object type", "Cardinality", "Merge rule"),
  caption = "Slot cardinality and merge rules declared before population",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  "single-valued" %in% slot_schema$cardinality,
  "list-valued" %in% slot_schema$cardinality,
  all(c("offers", "application_deadline") %in% slot_schema$slot)
)
```

The slot schema tells the code what counts as a duplicate and what counts as a
conflict. Without that rule, a pipeline might replace a deadline without
reporting the value recorded on 4 September, or turn two offerings into an
error.

## Prepare dated announcements

The documents below are constructed announcements. They cover one new fact, one
duplicate, one single-valued conflict, one negated statement, an ambiguous
`Riverton` mention, one NIL organisation, and one visible type-check failure.

```{r}
extractor_version <- "riverton-kbp-rules-1.0"

documents <- tribble(
  ~document_id, ~document_date, ~text,
  "A001", as.Date("2026-09-01"), "Riverton Skills Centre offers the Data Support Certificate.",
  "A002", as.Date("2026-09-03"), "The Skills Centre offers DATA SUPPORT CERTIFICATE.",
  "A003", as.Date("2026-09-04"), "Riverton Skills Centre application deadline is October 15.",
  "A004", as.Date("2026-09-06"), "Riverton Skills Centre application deadline is November 1.",
  "A005", as.Date("2026-09-07"), "Marrow County Transit does not require the Forklift Operator Licence.",
  "A006", as.Date("2026-09-08"), "Riverton will host the Data Support Certificate.",
  "A007", as.Date("2026-09-09"), "Riverton Transit offers forklift certification.",
  "A008", as.Date("2026-09-10"), "Bellhaven offers the Data Support Certificate."
)

candidate_plan <- tribble(
  ~candidate_id, ~document_id, ~slot, ~subject_mention, ~object_mention, ~status, ~expected_action,
  "CAND-001", "A001", "offers", "Riverton Skills Centre", "Data Support Certificate", "asserted", "add fact",
  "CAND-002", "A002", "offers", "Skills Centre", "DATA SUPPORT CERTIFICATE", "asserted", "merge duplicate",
  "CAND-003", "A003", "application_deadline", "Riverton Skills Centre", "October 15", "asserted", "add fact",
  "CAND-004", "A004", "application_deadline", "Riverton Skills Centre", "November 1", "asserted", "flag conflict",
  "CAND-005", "A005", "requires", "Marrow County Transit", "Forklift Operator Licence", "negated", "reject negated",
  "CAND-006", "A006", "offers", "Riverton", "Data Support Certificate", "asserted", "hold ambiguous",
  "CAND-007", "A007", "offers", "Riverton Transit", "forklift certification", "asserted", "hold NIL",
  "CAND-008", "A008", "offers", "Bellhaven", "Data Support Certificate", "asserted", "reject type"
)

knitr::kable(
  documents,
  col.names = c("Document ID", "Document date", "Evidence text"),
  caption = "Constructed dated announcements for knowledge base population",
  row.names = FALSE
)

knitr::kable(
  candidate_plan,
  col.names = c(
    "Candidate ID", "Document ID", "Slot", "Subject mention", "Object mention",
    "Status", "Expected action"
  ),
  caption = "Same-author expected actions written before the pipeline runs",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(documents), 8L),
  identical(nrow(candidate_plan), 8L),
  all(candidate_plan$status %in% c("asserted", "negated")),
  any(candidate_plan$expected_action == "hold NIL"),
  any(candidate_plan$expected_action == "flag conflict")
)
```

These expected actions are not a benchmark. They are a checklist for a teaching
pipeline written by the same author as the examples.

## Link mentions and keep offsets

The linker below accepts an alias only when it has one candidate. If the mention
is absent from the alias table, the code creates a provisional NIL ID. If the
alias has several candidates, the candidate goes to review instead of choosing a
nearby name.

```{r}
normalize_text <- function(value) {
  str_to_lower(value, locale = "en")
}

alias_lookup <- aliases |>
  mutate(normalized_alias = normalize_text(alias)) |>
  distinct(normalized_alias, entity_id, canonical_name, entity_type)

link_mention <- function(mention) {
  rows <- alias_lookup |>
    filter(normalized_alias == normalize_text(mention))

  if (nrow(rows) == 0L) {
    return(tibble(
      linked_id = NA_character_,
      linked_type = NA_character_,
      link_status = "NIL"
    ))
  }

  if (n_distinct(rows$entity_id) > 1L) {
    return(tibble(
      linked_id = paste(sort(unique(rows$entity_id)), collapse = "; "),
      linked_type = paste(sort(unique(rows$entity_type)), collapse = "; "),
      link_status = "ambiguous"
    ))
  }

  tibble(
    linked_id = rows$entity_id[[1]],
    linked_type = rows$entity_type[[1]],
    link_status = "linked"
  )
}

locate_mention <- function(text, mention) {
  location <- str_locate(
    text,
    regex(escape_regex(mention), ignore_case = TRUE)
  )
  tibble(
    start = as.integer(location[1, 1]),
    end = as.integer(location[1, 2])
  )
}

escape_regex <- function(value) {
  str_replace_all(value, "([.\\\\|()\\[\\]{}^$*+?])", "\\\\\\1")
}

nil_mentions <- candidate_plan |>
  distinct(subject_mention) |>
  mutate(link = map(subject_mention, link_mention)) |>
  unnest(link) |>
  filter(link_status == "NIL") |>
  mutate(provisional_id = sprintf("NIL-%04d", row_number()))

candidates <- candidate_plan |>
  left_join(documents, by = "document_id") |>
  mutate(
    subject_link = map(subject_mention, link_mention),
    object_link = map(object_mention, link_mention),
    subject_offsets = map2(text, subject_mention, locate_mention),
    object_offsets = map2(text, object_mention, locate_mention)
  ) |>
  unnest(subject_link) |>
  rename(
    subject_id = linked_id,
    subject_type = linked_type,
    subject_link_status = link_status
  ) |>
  unnest(object_link) |>
  rename(
    object_id = linked_id,
    object_type = linked_type,
    object_link_status = link_status
  ) |>
  unnest(subject_offsets) |>
  rename(subject_start = start, subject_end = end) |>
  unnest(object_offsets) |>
  rename(object_start = start, object_end = end) |>
  left_join(
    nil_mentions |>
      select(subject_mention, provisional_id),
    by = "subject_mention"
  ) |>
  mutate(
    subject_id = if_else(subject_link_status == "NIL", provisional_id, subject_id),
    subject_type = if_else(subject_link_status == "NIL", "organisation", subject_type),
    object_value = case_when(
      slot == "application_deadline" ~ object_mention,
      TRUE ~ object_id
    )
  )

link_table <- candidates |>
  select(
    candidate_id,
    subject_mention,
    subject_id,
    subject_type,
    subject_link_status,
    object_mention,
    object_id,
    object_type,
    object_link_status
  )

knitr::kable(
  link_table,
  col.names = c(
    "Candidate ID", "Subject mention", "Subject ID", "Subject type",
    "Subject link status", "Object mention", "Object ID", "Object type",
    "Object link status"
  ),
  caption = "Mention linking outcomes before fact insertion",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(candidates$subject_link_status[candidates$candidate_id == "CAND-006"], "ambiguous"),
  identical(candidates$subject_link_status[candidates$candidate_id == "CAND-007"], "NIL"),
  identical(candidates$subject_id[candidates$candidate_id == "CAND-007"], "NIL-0001"),
  str_detect(candidates$subject_id[candidates$candidate_id == "CAND-006"], "LOC-0001"),
  str_detect(candidates$subject_id[candidates$candidate_id == "CAND-006"], "LOC-0002"),
  all(!is.na(candidates$subject_start)),
  all(!is.na(candidates$object_start))
)
```

The `Riverton` row keeps both candidate IDs in the review record. It is not
auto-linked to either place. The `Riverton Transit` row gets a provisional NIL
ID and is not merged into Marrow County Transit.

## Apply slot, status, and type decisions

The action table is the evaluation for this lesson. It shows whether the
pipeline made the expected decision for each candidate. It is not precision and
recall.

```{r}
schema_checks <- candidates |>
  left_join(slot_schema, by = "slot") |>
  mutate(
    subject_type_ok = subject_type.x == subject_type.y,
    object_type_ok = case_when(
      object_type.y == "date literal" ~ str_detect(object_mention, "^[A-Z][a-z]+ [0-9]+$"),
      TRUE ~ object_type.x == object_type.y
    ),
    type_check = subject_type_ok & object_type_ok,
    fact_key = paste(slot, subject_id, object_value, sep = "|")
  )

first_seen <- schema_checks |>
  filter(status == "asserted", subject_link_status == "linked", type_check) |>
  arrange(document_date, candidate_id) |>
  distinct(fact_key, .keep_all = TRUE) |>
  transmute(first_candidate_id = candidate_id, fact_key)

single_value_seen <- schema_checks |>
  filter(status == "asserted", subject_link_status == "linked", type_check) |>
  filter(cardinality == "single-valued") |>
  count(slot, subject_id, name = "single_value_count")

first_single_seen <- schema_checks |>
  filter(status == "asserted", subject_link_status == "linked", type_check) |>
  filter(cardinality == "single-valued") |>
  arrange(document_date, candidate_id) |>
  group_by(slot, subject_id) |>
  slice_head(n = 1) |>
  ungroup() |>
  transmute(slot, subject_id, first_single_candidate_id = candidate_id)

candidate_actions <- schema_checks |>
  left_join(first_seen, by = "fact_key") |>
  left_join(single_value_seen, by = c("slot", "subject_id")) |>
  left_join(first_single_seen, by = c("slot", "subject_id")) |>
  mutate(
    single_value_count = coalesce(single_value_count, 0L),
    actual_action = case_when(
      status == "negated" ~ "reject negated",
      subject_link_status == "ambiguous" ~ "hold ambiguous",
      subject_link_status == "NIL" ~ "hold NIL",
      !type_check ~ "reject type",
      candidate_id != first_candidate_id ~ "merge duplicate",
      cardinality == "single-valued" &
        single_value_count > 1L &
        candidate_id != first_single_candidate_id ~ "flag conflict",
      TRUE ~ "add fact"
    ),
    action_matches = actual_action == expected_action,
    action_reason = case_when(
      status == "negated" ~ "negated statements never become facts",
      subject_link_status == "ambiguous" ~ "alias maps to more than one ID",
      subject_link_status == "NIL" ~ "new name receives a provisional ID for review",
      !type_check ~ "slot type check failed",
      candidate_id != first_candidate_id ~ "same fact already exists; keep provenance",
      cardinality == "single-valued" &
        single_value_count > 1L &
        candidate_id != first_single_candidate_id ~ "single-valued slot has dated conflict",
      TRUE ~ "asserted, linked, and type-compatible"
    )
  )

knitr::kable(
  candidate_actions |>
    select(candidate_id, expected_action, actual_action, action_matches, action_reason),
  col.names = c("Candidate ID", "Expected action", "Pipeline action", "Match", "Reason"),
  caption = "Expected-versus-actual action table for candidate facts",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(candidate_actions$action_matches),
  identical(
    candidate_actions$actual_action[candidate_actions$candidate_id == "CAND-005"],
    "reject negated"
  ),
  identical(
    candidate_actions$actual_action[candidate_actions$candidate_id == "CAND-004"],
    "flag conflict"
  ),
  identical(
    candidate_actions$actual_action[candidate_actions$candidate_id == "CAND-008"],
    "reject type"
  )
)
```

The type failure is visible: `Bellhaven` is a place ID, not an organisation.
The negated requirement row is rejected rather than stored as a negative fact or
used to delete anything.

## Build facts and provenance

The final fact table keeps one fact for the duplicate offer and two dated
deadline values for the conflict. Provenance stays in a separate table so one
fact can point to several source records.

```{r}
accepted_candidates <- candidate_actions |>
  filter(actual_action %in% c("add fact", "merge duplicate", "flag conflict"))

fact_keys <- accepted_candidates |>
  distinct(slot, subject_id, object_value, fact_key) |>
  arrange(slot, subject_id, object_value) |>
  mutate(fact_id = sprintf("FACT-%03d", row_number()))

populated_facts <- accepted_candidates |>
  inner_join(fact_keys, by = c("slot", "subject_id", "object_value", "fact_key")) |>
  group_by(fact_id, slot, subject_id, object_value) |>
  summarise(
    value_count = n_distinct(object_value),
    provenance_records = n(),
    conflict_flag = any(actual_action == "flag conflict") |
      (first(slot) == "application_deadline" &
        n_distinct(object_value) > 0L &
        n_distinct(candidate_actions$object_value[
          candidate_actions$slot == first(slot) &
            candidate_actions$subject_id == first(subject_id) &
            candidate_actions$actual_action %in% c("add fact", "flag conflict")
        ]) > 1L),
    .groups = "drop"
  ) |>
  arrange(fact_id)

fact_provenance <- accepted_candidates |>
  inner_join(fact_keys, by = c("slot", "subject_id", "object_value", "fact_key")) |>
  transmute(
    fact_id,
    candidate_id,
    document_id,
    document_date,
    evidence_text = text,
    subject_start,
    subject_end,
    object_start,
    object_end,
    extractor_version
  ) |>
  arrange(fact_id, document_id)

review_queue <- candidate_actions |>
  filter(!actual_action %in% c("add fact", "merge duplicate", "flag conflict")) |>
  transmute(
    candidate_id,
    document_id,
    subject_mention,
    object_mention,
    provisional_or_candidate_id = subject_id,
    review_reason = action_reason
  )

knitr::kable(
  populated_facts,
  col.names = c(
    "Fact ID", "Slot", "Subject ID", "Object value",
    "Distinct values in fact", "Provenance records", "Conflict flagged"
  ),
  caption = "Populated facts after duplicate and conflict handling",
  row.names = FALSE
)

knitr::kable(
  fact_provenance,
  col.names = c(
    "Fact ID", "Candidate ID", "Document ID", "Document date",
    "Evidence text", "Subject start", "Subject end",
    "Object start", "Object end", "Extractor version"
  ),
  caption = "Provenance records for every populated fact",
  row.names = FALSE
)

knitr::kable(
  review_queue,
  col.names = c(
    "Candidate ID", "Document ID", "Subject mention", "Object mention",
    "Provisional or candidate ID", "Review reason"
  ),
  caption = "Candidates held out of the fact table for review",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(!is.na(fact_provenance$document_id)),
  all(!is.na(fact_provenance$document_date)),
  all(nzchar(fact_provenance$evidence_text)),
  all(!is.na(fact_provenance$subject_start)),
  all(!is.na(fact_provenance$object_start)),
  all(!is.na(fact_provenance$extractor_version)),
  !any(fact_provenance$candidate_id == "CAND-005"),
  populated_facts$provenance_records[populated_facts$slot == "offers"] == 2L,
  all(populated_facts$conflict_flag[populated_facts$slot == "application_deadline"]),
  any(review_queue$provisional_or_candidate_id == "NIL-0001")
)
```

One offer fact has two provenance records. The two deadline facts both remain,
and both are flagged because a later document might be an update, an error, or a
different session. The table records the conflict; it does not resolve it.

## Prepare the graph facts used in the visualization lesson

Lesson 81 redraws a small graph from the same IDs. The table below is a compact
edge view of the kept KB facts plus reference links that keep the two Riverton
places separate.

```{r}
graph_facts <- tribble(
  ~subject_id, ~relation, ~object_id, ~source,
  "ORG-0002", "offers", "CRD-0002", "populated fact",
  "ORG-0002", "application_deadline", "DATE-2026-10-15", "populated fact",
  "ORG-0002", "application_deadline", "DATE-2026-11-01", "populated fact",
  "ORG-0002", "located_in", "LOC-0001", "reference link",
  "ORG-0003", "serves", "LOC-0001", "reference link",
  "LOC-0003", "near", "LOC-0001", "reference link"
)

knitr::kable(
  graph_facts,
  col.names = c("Subject ID", "Relation", "Object ID", "Source"),
  caption = "Fact edges rebuilt in the graph visualization lesson",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  all(c("LOC-0001", "LOC-0002") %in% entities$entity_id),
  "LOC-0002" %in% setdiff(entities$entity_id, c(graph_facts$subject_id, graph_facts$object_id)),
  any(graph_facts$relation == "offers"),
  any(graph_facts$object_id == "DATE-2026-11-01"),
  !any(graph_facts$subject_id == "NIL-0001"),
  !any(graph_facts$relation == "shares_label_with")
)
```

These graph edges are still KB records, not proof about the real world. The
source column says whether an edge came from the constructed population step or
from the local reference setting used for the drawing. The two `Riverton` place
IDs remain in the entity table, but their shared label is a naming note for the
visualization lesson rather than a factual relation.

## What the knowledge base does not know

A missing edge is not evidence that a relation is false. This is the open-world
caveat: the KB records what it has seen and accepted under its current policy.
It may be incomplete, stale, or contradicted by a later document. Provenance
helps a reviewer inspect a claim; it does not certify the claim as true.

## What to remember

- Knowledge base population combines linking and slot filling.
- Slot cardinality must be declared before merging facts.
- Duplicate facts can share one fact ID with several provenance records.
- Single-valued conflicts keep both dated values until review.
- Negated statements do not become facts.
- `Riverton` stays ambiguous, and a NIL organisation receives a provisional ID.

Grace ends with fewer automatic facts than candidate statements. That is the
point: a useful KBP step records what it added, what it held, and why.

## Sources

- [Ji and Grishman (2011), Knowledge Base Population: Successful Approaches and Challenges](https://aclanthology.org/P11-1115/)
- [NIST TAC KBP 2014 Cold Start track](https://tac.nist.gov/2014/KBP/ColdStart/index.html)
- [Wikidata Help:Qualifiers](https://www.wikidata.org/wiki/Help:Qualifiers)
- [Wikidata Help:Ranking](https://www.wikidata.org/wiki/Help:Ranking)
- [Hogan et al. (2021), Knowledge Graphs](https://arxiv.org/abs/2003.02320)
- [Zhang et al. (2017), TACRED](https://aclanthology.org/D17-1004/)
