---
title: "Reviewing documents for discovery and alerts"
subtitle: "Estimate what search and ranked review leave unseen"
description: "Learn how keyword terms, ranked review, an elusion sample, and a small monitoring stream expose missed records."
categories: [systems, e-discovery, media monitoring]
---

A fictional request asks for records about the Calder Yard contract. It is not asking for a few good examples. It asks for the responsive records in a fixed collection, including the ones that do not use the words a searcher guesses first.

**E-discovery** is electronic discovery: finding records that must be reviewed or produced for a legal request. **Technology-assisted review** uses a model to rank unreviewed records so people can review likely responsive records first. **Media monitoring** watches new items as they arrive and sends alerts for items that matter.

Grossman and Cormack report that Blair and Maron's legal-search study found searchers believed they had found at least `r 75` percent of relevant documents, while average recall was about `r 20` percent for the studied requests. Lesson 9's active learning chooses uncertain examples to improve a model; ranked review here chooses high-scoring unreviewed records to find responsive material. A low elusion rate can still hide many misses: if a review found `r 70` of `r 100` responsive records in a `r format(10000, big.mark = ",")`-record collection, the unreviewed pile would have only `r sprintf("%.2f%%", 100 * 30 / 9400)` elusion but recall would be only `r sprintf("%.0f%%", 100 * 70 / 100)`.

::: {.callout-note}
The records in this lesson are constructed teaching examples. They do not describe a real legal matter, company, person, news source, or monitoring target.
:::

::: {.callout-tip}
## What you will learn

This lesson shows how to:

- predeclare keyword terms and read their precision and recall;
- train a keyword-seeded ranked review model without scoring its training rows;
- stop review by a fixed rule chosen before sampling;
- estimate missed responsive records with one finite-population sample;
- explain why the fictional answer key is stronger than real reviewers usually have; and
- process a short monitoring stream in arrival order with duplicate suppression.
:::

```{r}
#| include: false

options(warn = 2)
stopifnot(identical(getOption("warn"), 2L))
```

## Load the constructed review collection

The builder created a short review collection for one request. It also wrote a monitoring stream and metadata fingerprints. Each row carries a fiction notice. The reference labels were written by the page author, so they are labels for this constructed collection rather than ground truth.

```{r}
suppressPackageStartupMessages({
  library(digest)
  library(dplyr)
  library(ggplot2)
  library(glmnet)
  library(knitr)
  library(purrr)
  library(readr)
  library(stringdist)
  library(stringr)
  library(tibble)
  library(tidymodels)
  library(textrecipes)
  library(tidyr)
})

hash_lines <- function(path) {
  digest(
    paste(read_lines(path), collapse = "\n"),
    algo = "sha256",
    serialize = FALSE
  )
}

collection_path <- "data/riverton/riverton-review-collection.csv"
stream_path <- "data/riverton/riverton-monitoring-stream.csv"
metadata_path <- "data/riverton/riverton-review-collection-metadata.csv"

review_records <- read_csv(
  collection_path,
  na = c("", "NA"),
  col_types = cols(
    doc_id = col_character(),
    record_type = col_character(),
    doc_date = col_date(),
    source_type = col_character(),
    sender_role = col_character(),
    subject = col_character(),
    review_request = col_character(),
    reference_responsive = col_logical(),
    reference_reason = col_character(),
    construction_subtype = col_character(),
    near_duplicate_of = col_character(),
    text = col_character(),
    author_note = col_character(),
    construction_id = col_character(),
    file_position = col_integer()
  )
)

monitoring_stream <- read_csv(
  stream_path,
  na = c("", "NA"),
  col_types = cols(
    stream_id = col_character(),
    record_type = col_character(),
    arrival_order = col_integer(),
    doc_date = col_date(),
    source_type = col_character(),
    text = col_character(),
    reference_alert = col_logical(),
    reference_reason = col_character(),
    near_duplicate_of = col_character(),
    author_note = col_character()
  )
)

review_metadata <- read_csv(
  metadata_path,
  na = character(),
  col_types = cols(
    artifact = col_character(),
    description = col_character(),
    source = col_character(),
    license = col_character(),
    created_on = col_character(),
    purpose = col_character(),
    rows = col_integer(),
    class_counts = col_character(),
    fingerprint = col_character(),
    rng_kind = col_character(),
    seed = col_character(),
    effective_seeds = col_character(),
    review_request = col_character(),
    shortcut_sweep = col_character()
  )
)

actual_fingerprints <- tibble(
  artifact = c(
    "riverton-review-collection.csv",
    "riverton-monitoring-stream.csv"
  ),
  fingerprint = c(hash_lines(collection_path), hash_lines(stream_path))
)

collection_summary <- review_records |>
  count(reference_responsive, name = "records") |>
  mutate(label = if_else(reference_responsive, "responsive", "not responsive"))

knitr::kable(
  collection_summary |>
    select(label, records),
  col.names = c("Reference label", "Records"),
  caption = "Constructed review collection labels for one fictional request",
  row.names = FALSE
)

knitr::kable(
  tibble(review_request = review_metadata$review_request[[1]]),
  format = "html",
  escape = TRUE,
  col.names = "Review request",
  caption = "Fictional request used to define responsiveness",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(nrow(review_records), 240L),
  identical(sum(review_records$reference_responsive), 36L),
  identical(nrow(monitoring_stream), 24L),
  identical(
    actual_fingerprints$fingerprint,
    review_metadata$fingerprint[
      match(actual_fingerprints$artifact, review_metadata$artifact)
    ]
  ),
  all(nzchar(review_records$author_note)),
  all(nzchar(monitoring_stream$author_note)),
  identical(review_metadata$seed[1], "7401")
)
```

The constructed prevalence is `r sprintf("%.0f%%", 100 * mean(review_records$reference_responsive))`, much higher than many legal-review matters. It was chosen so a small teaching page has enough responsive examples to inspect. The builder also swept superficial features such as length, punctuation, dates, pronouns, source type, sentence frame, common non-topic words, file position, and ID pattern, and refused to write if a single feature separated the classes too well.

## Search predeclared keywords

The keyword terms below are fixed before scoring. They are plausible terms for the written request, but they are still guesses. Keyword recall is knowable here only because the collection is fictional and the reference labels are already in the file.

```{r}
keyword_terms <- c(
  "inspection",
  "inspect",
  "delete",
  "deleted",
  "delay",
  "delayed",
  "log",
  "Calder Yard"
)

keyword_pattern <- regex(
  paste0(
    "\\b(",
    paste(keyword_terms[keyword_terms != "Calder Yard"], collapse = "|"),
    ")\\b|\\bCalder Yard\\b"
  ),
  ignore_case = TRUE
)

review_scored <- review_records |>
  mutate(
    keyword_hit = str_detect(text, keyword_pattern),
    label = factor(
      if_else(reference_responsive, "responsive", "not responsive"),
      levels = c("responsive", "not responsive")
    )
  )

keyword_metrics <- review_scored |>
  summarise(
    hits = sum(keyword_hit),
    responsive_hits = sum(keyword_hit & reference_responsive),
    responsive_total = sum(reference_responsive),
    precision = responsive_hits / hits,
    recall = responsive_hits / responsive_total
  )

keyword_examples <- bind_rows(
  review_scored |>
    filter(keyword_hit, reference_responsive) |>
    arrange(doc_id) |>
    slice_head(n = 4),
  review_scored |>
    filter(keyword_hit, !reference_responsive) |>
    arrange(doc_id) |>
    slice_head(n = 4)
) |>
  transmute(
    doc_id,
    reference_label = if_else(reference_responsive, "responsive", "not responsive"),
    construction_subtype,
    text
  )

knitr::kable(
  keyword_metrics |>
    mutate(across(c(precision, recall), \(value) round(value, 3))),
  col.names = c("Keyword hits", "Responsive hits", "Responsive records", "Precision", "Recall"),
  caption = "Predeclared keyword terms on the constructed collection",
  row.names = FALSE
)

knitr::kable(
  keyword_examples,
  format = "html",
  escape = TRUE,
  col.names = c("Document", "Reference label", "Construction note", "Text"),
  caption = "Keyword hits include both responsive records and false hits",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(keyword_metrics$hits, 39L),
  identical(keyword_metrics$responsive_hits, 22L),
  identical(keyword_metrics$responsive_total, 36L),
  identical(round(keyword_metrics$precision, 3), 0.564),
  identical(round(keyword_metrics$recall, 3), 0.611),
  keyword_metrics$responsive_hits >= 8L,
  keyword_metrics$hits - keyword_metrics$responsive_hits >= 8L
)
```

Precision is the share of reviewed hits that are responsive. Recall is the share of all responsive records that were found. The terms match whole words, plus the phrase `Calder Yard`; for example, `log` does not match `chronology`. In this fixture, the terms find `r keyword_metrics$hits` records. Of those, `r keyword_metrics$responsive_hits` are responsive under the constructed labels. That precision, `r keyword_metrics$responsive_hits` divided by `r keyword_metrics$hits`, is visible after reviewing the hits. The recall, `r keyword_metrics$responsive_hits` divided by `r keyword_metrics$responsive_total`, is not normally visible; the fictional labels let the lesson show what the keyword search missed.

## Rank the rest from the reviewed seed

The reviewed seed is the keyword-hit set. It has at least eight records of each class, which avoids the `glmnet` warning for very small classes under `options(warn = 2)`. The model uses tf-idf features and penalized logistic regression. It trains only on reviewed records and scores only records not yet reviewed.

The stopping rule is also fixed before the model runs: review the keyword-hit seed, then review two ranked batches of ten. That rule has no statistical guarantee. It is a budget rule, and the elusion sample is what estimates the remaining miss.

```{r}
fit_review_model <- function(training_rows) {
  class_counts <- table(training_rows$label)
  stopifnot(min(class_counts) >= 8L)

  review_recipe <- recipe(label ~ text, data = training_rows) |>
    step_tokenize(text) |>
    step_tokenfilter(text, max_tokens = 80) |>
    step_tfidf(text)

  review_model <- logistic_reg(penalty = 0.01, mixture = 1) |>
    set_engine("glmnet") |>
    set_mode("classification")

  workflow() |>
    add_recipe(review_recipe) |>
    add_model(review_model) |>
    fit(data = training_rows)
}

base_columns <- names(review_scored)
reviewed_records <- review_scored |>
  filter(keyword_hit)
unreviewed_records <- review_scored |>
  filter(!keyword_hit)

review_batches <- list()
ranked_batches <- 2L
batch_size <- 10L
tie_rows <- integer(ranked_batches)

for (batch_id in seq_len(ranked_batches)) {
  fitted_review <- fit_review_model(reviewed_records)
  scored_pool <- unreviewed_records |>
    bind_cols(
      predict(
        fitted_review,
        new_data = unreviewed_records,
        type = "prob"
      )
    ) |>
    mutate(score = .pred_responsive) |>
    arrange(desc(score), doc_id)

  tie_rows[[batch_id]] <- sum(
    duplicated(scored_pool$score) |
      duplicated(scored_pool$score, fromLast = TRUE)
  )

  batch_rows <- scored_pool |>
    slice_head(n = batch_size) |>
    mutate(batch = batch_id)

  review_batches[[batch_id]] <- batch_rows
  reviewed_records <- bind_rows(
    reviewed_records,
    batch_rows |> select(all_of(base_columns))
  )
  unreviewed_records <- anti_join(
    unreviewed_records,
    batch_rows |> select(doc_id),
    by = join_by(doc_id)
  )
}

ranked_reviewed <- bind_rows(review_batches)

ranked_review_display <- ranked_reviewed |>
  transmute(
    batch,
    doc_id,
    score = round(score, 3),
    reference_label = if_else(reference_responsive, "responsive", "not responsive"),
    construction_subtype,
    text
  )

batch_summary <- ranked_reviewed |>
  count(batch, reference_responsive, name = "records") |>
  complete(
    batch,
    reference_responsive = c(FALSE, TRUE),
    fill = list(records = 0L)
  ) |>
  mutate(label = if_else(reference_responsive, "responsive", "not responsive")) |>
  select(batch, label, records)

review_progress <- bind_rows(
  tibble(
    stage = "keyword seed",
    reviewed_this_stage = nrow(review_scored |> filter(keyword_hit)),
    responsive_found = sum(review_scored$keyword_hit & review_scored$reference_responsive)
  ),
  ranked_reviewed |>
    group_by(batch) |>
    summarise(
      stage = paste0("ranked batch ", first(batch)),
      reviewed_this_stage = n(),
      responsive_found = sum(reference_responsive),
      .groups = "drop"
    ) |>
    select(stage, reviewed_this_stage, responsive_found)
) |>
  mutate(
    cumulative_reviewed = cumsum(reviewed_this_stage),
    cumulative_responsive_found = cumsum(responsive_found),
    fiction_only_recall = cumulative_responsive_found /
      sum(review_scored$reference_responsive)
  )

review_stop <- tibble(
  reviewed = nrow(reviewed_records),
  responsive_found = sum(reviewed_records$reference_responsive),
  unreviewed = nrow(unreviewed_records),
  fiction_known_missed = sum(unreviewed_records$reference_responsive)
)

missed_records <- unreviewed_records |>
  filter(reference_responsive) |>
  transmute(
    doc_id,
    construction_subtype,
    text
  )

knitr::kable(
  batch_summary,
  col.names = c("Ranked batch", "Reference label", "Records"),
  caption = "Ranked review batches after the keyword seed",
  row.names = FALSE
)

knitr::kable(
  review_progress |>
    mutate(fiction_only_recall = round(fiction_only_recall, 3)),
  col.names = c(
    "Stage",
    "Reviewed at stage",
    "Responsive found at stage",
    "Cumulative reviewed",
    "Cumulative responsive found",
    "Fiction-only recall"
  ),
  caption = "Review progress, with recall visible only because the collection is invented",
  row.names = FALSE
)

knitr::kable(
  review_stop,
  col.names = c("Reviewed", "Responsive found", "Unreviewed", "Fiction-known missed"),
  caption = "Predeclared stopping point before the elusion sample",
  row.names = FALSE
)

knitr::kable(
  ranked_review_display,
  format = "html",
  escape = TRUE,
  col.names = c(
    "Batch",
    "Document",
    "Review score",
    "Reference label",
    "Construction note",
    "Text"
  ),
  caption = "Ranked records reviewed after the keyword seed",
  row.names = FALSE
)

knitr::kable(
  missed_records,
  format = "html",
  escape = TRUE,
  col.names = c("Document", "Construction note", "Text"),
  caption = "Fiction-only responsive records still unreviewed at the stop",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(ranked_batches, 2L),
  identical(batch_size, 10L),
  identical(nrow(reviewed_records), 59L),
  identical(sum(reviewed_records$reference_responsive), 33L),
  identical(nrow(unreviewed_records), 181L),
  identical(sum(unreviewed_records$reference_responsive), 3L),
  identical(nrow(missed_records), 3L),
  all(table(review_scored$label[review_scored$keyword_hit]) >= 8L),
  !any(ranked_reviewed$doc_id %in% review_scored$doc_id[review_scored$keyword_hit])
)
```

The team can see how many responsive records it has found, because reviewers code those records as they review them. It cannot see recall in a real matter without more assessment. The recall column is labeled as fiction-only because it uses the hidden labels for records that were not reviewed. In the two scoring rounds, `r sum(tie_rows)` scored-pool rows were part of exact score ties; ties are ordered by document ID so the run is reproducible, not because lower IDs are more relevant.

```{r}
#| label: fig-review-gain
#| fig-cap: "Responsive records found as review proceeds through the keyword seed and two ranked batches."
#| fig-alt: "Line chart starting at zero. Responsive records found rises from 22 at 39 reviewed records to 23 at 49 and 33 at 59."

ggplot(
  review_progress,
  aes(x = cumulative_reviewed, y = cumulative_responsive_found)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(
    breaks = c(0, review_progress$cumulative_reviewed),
    limits = c(0, max(review_progress$cumulative_reviewed))
  ) +
  scale_y_continuous(
    breaks = c(0, review_progress$cumulative_responsive_found),
    limits = c(0, max(review_progress$cumulative_responsive_found))
  ) +
  labs(
    x = "Records reviewed",
    y = "Responsive records found"
  ) +
  theme_minimal()
```

```{r}
#| include: false

stopifnot(
  identical(nrow(review_progress), 3L),
  identical(review_progress$cumulative_reviewed, c(39L, 49L, 59L)),
  identical(review_progress$cumulative_responsive_found, c(22L, 23L, 33L))
)
```

The gain curve is the reviewer-visible curve: responsive found versus records reviewed. It does not claim that ranked review beats keyword search. The collection was written by the same person who wrote the lesson, so comparative claims from it would be weak evidence.

## Sample the unreviewed pile once

An **elusion sample** reviews a random sample from the documents that were not reviewed. The seed and sample size are declared before sampling. The sample is drawn once and is not fed back into the model or the stopping decision.

The interval below treats the unreviewed pile as a finite population sampled without replacement. It tests every possible number of missed responsive records, keeps the values that would not put the observed sample count in either 2.5 percent tail under `phyper()`, then converts missed-record bounds to recall bounds.

```{r}
elusion_seed <- 7401L
elusion_n <- 50L

set.seed(elusion_seed)
elusion_sample <- unreviewed_records |>
  slice_sample(n = elusion_n) |>
  arrange(doc_id)

observed_missed_in_sample <- sum(elusion_sample$reference_responsive)
unreviewed_total <- nrow(unreviewed_records)
found_total <- sum(reviewed_records$reference_responsive)

possible_missed <- 0:unreviewed_total
kept_missed <- possible_missed[
  vapply(
    possible_missed,
    \(missed) {
      lower_tail <- phyper(
        observed_missed_in_sample,
        missed,
        unreviewed_total - missed,
        elusion_n
      )
      upper_tail <- 1 - phyper(
        observed_missed_in_sample - 1L,
        missed,
        unreviewed_total - missed,
        elusion_n
      )
      lower_tail >= 0.025 && upper_tail >= 0.025
    },
    logical(1)
  )
]

missed_interval <- range(kept_missed)
recall_interval <- found_total / (found_total + rev(missed_interval))
point_missed <- unreviewed_total *
  observed_missed_in_sample / elusion_n
point_recall <- found_total / (found_total + point_missed)
known_recall <- found_total /
  (found_total + sum(unreviewed_records$reference_responsive))

elusion_result <- tibble(
  unreviewed = unreviewed_total,
  sample_size = elusion_n,
  responsive_in_sample = observed_missed_in_sample,
  point_recall = point_recall,
  recall_low = recall_interval[1],
  recall_high = recall_interval[2],
  fiction_known_recall = known_recall
)

binomial_interval <- stats::binom.test(
  observed_missed_in_sample,
  elusion_n
)$conf.int
binomial_missed <- unreviewed_total * binomial_interval
binomial_recall <- found_total / (found_total + rev(binomial_missed))
maximum_possible_recall <- found_total /
  (found_total + observed_missed_in_sample)

binomial_warning <- tibble(
  method = "binomial Clopper-Pearson on sampled elusion rate",
  printed_recall_low = binomial_recall[1],
  printed_recall_high = binomial_recall[2],
  maximum_possible_recall = maximum_possible_recall
)

knitr::kable(
  elusion_result |>
    mutate(across(where(is.numeric), \(value) round(value, 3))),
  col.names = c(
    "Unreviewed",
    "Sample size",
    "Responsive in sample",
    "Point recall",
    "Interval low",
    "Interval high",
    "Fiction-known recall"
  ),
  caption = "One finite-population elusion sample after the stop",
  row.names = FALSE
)

knitr::kable(
  binomial_warning |>
    mutate(across(where(is.numeric), \(value) round(value, 3))),
  col.names = c(
    "Method",
    "Printed recall low",
    "Printed recall high",
    "Maximum possible recall after the sample"
  ),
  caption = "A binomial interval treats the fixed unreviewed pile as if it were very large",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(elusion_seed, 7401L),
  identical(elusion_n, 50L),
  identical(observed_missed_in_sample, 0L),
  identical(unreviewed_total, 181L),
  identical(found_total, 33L),
  identical(missed_interval, c(0L, 11L)),
  round(recall_interval[1], 3) == 0.75,
  round(recall_interval[2], 3) == 1,
  round(known_recall, 3) == 0.917,
  binomial_warning$printed_recall_low < recall_interval[1]
)
```

The predeclared elusion seed is `7401`, reused by rule before drawing the sample. The sample found `r observed_missed_in_sample` responsive records in `r elusion_n`. `r if (observed_missed_in_sample == 0L) "The point estimate reads as no missed records in the unreviewed pile, but" else sprintf("The point estimate scales that to about %.1f missed records in the unreviewed pile, and", point_missed)` the exact finite-population interval still allows `r missed_interval[1]` to `r missed_interval[2]` missed records. That becomes recall from about `r round(recall_interval[1], 3)` to `r round(recall_interval[2], 3)`. The fiction reveals that the true missed count is `r sum(unreviewed_records$reference_responsive)`.

A binomial Clopper-Pearson interval treats the pile as if it were very large and samples with replacement. Here it gives a wider lower recall bound than the finite-population interval. Confidence intervals describe a repeated-sampling procedure; they are not a 95 percent chance statement about this one interval.

Repeatedly sampling, reviewing more records, and sampling again until the result looks acceptable would bias the report. The sample belongs after the stop, once.

## Monitor new items in arrival order

Media monitoring has a different shape. Items arrive over time, duplicate wire copies appear, and a name like Riverton can refer to more than one place. This short stream watches organizations and program names, not individual people. It stores item IDs, decisions, and duplicate links; real review sets may contain personal or privileged material, which this lesson leaves out.

This use of "monitoring" is not lesson 47's deployed-model monitoring, and it is not burst detection from lesson 56. It is an alerting rule over incoming text.

The watch phrases below were written after reading this constructed stream, so the alert count shows bookkeeping rather than transferable precision. The pattern watches `Calder Yard`, `contract records`, `inspection messages`, `route notes`, and `missed check`. A repeated alert is suppressed only when its character-trigram Jaccard distance to a prior alert is at most 0.35.

```{r}
normalize_stream_text <- function(text) {
  text |>
    str_to_lower(locale = "en") |>
    str_replace_all("[^a-z0-9]+", " ") |>
    str_squish()
}

stream_decisions <- monitoring_stream |>
  arrange(arrival_order) |>
  mutate(
    normalized_text = normalize_stream_text(text),
    topic_hit = str_detect(
      text,
      regex("Calder Yard|contract records|inspection messages|route notes|missed check", ignore_case = TRUE)
    ),
    ambiguous_riverton = str_detect(text, regex("\\bRiverton\\b", ignore_case = TRUE)) &
      !str_detect(text, regex("Calder|Skills Centre|Workforce Lab", ignore_case = TRUE)),
    raw_alert = topic_hit & !ambiguous_riverton
  )

previous_alerts <- character()
previous_alert_ids <- character()
duplicate_of <- rep(NA_character_, nrow(stream_decisions))
send_alert <- rep(FALSE, nrow(stream_decisions))

for (row in seq_len(nrow(stream_decisions))) {
  if (!stream_decisions$raw_alert[[row]]) {
    next
  }

  if (length(previous_alerts) == 0L) {
    send_alert[[row]] <- TRUE
    previous_alerts <- c(previous_alerts, stream_decisions$normalized_text[[row]])
    previous_alert_ids <- c(previous_alert_ids, stream_decisions$stream_id[[row]])
    next
  }

  distances <- stringdist(
    stream_decisions$normalized_text[[row]],
    previous_alerts,
    method = "jaccard",
    q = 3
  )
  if (min(distances) <= 0.35) {
    duplicate_of[[row]] <- previous_alert_ids[[which.min(distances)]]
  } else {
    send_alert[[row]] <- TRUE
    previous_alerts <- c(previous_alerts, stream_decisions$normalized_text[[row]])
    previous_alert_ids <- c(previous_alert_ids, stream_decisions$stream_id[[row]])
  }
}

stream_decisions <- stream_decisions |>
  mutate(
    matched_prior_alert = duplicate_of,
    alert_sent = send_alert,
    decision = case_when(
      alert_sent ~ "alert sent",
      raw_alert & !is.na(matched_prior_alert) ~ "suppressed duplicate",
      ambiguous_riverton ~ "needs review: ambiguous Riverton",
      TRUE ~ "no alert"
    )
  )

monitoring_results <- tibble(
  alerts_sent = sum(stream_decisions$alert_sent),
  relevant_alerts = sum(stream_decisions$alert_sent & stream_decisions$reference_alert),
  false_alerts = sum(stream_decisions$alert_sent & !stream_decisions$reference_alert),
  relevant_items_missed = sum(!stream_decisions$alert_sent & stream_decisions$reference_alert & is.na(stream_decisions$matched_prior_alert)),
  suppressed_duplicates = sum(!is.na(stream_decisions$matched_prior_alert)),
  repeated_alerts = sum(stream_decisions$alert_sent & !is.na(stream_decisions$near_duplicate_of)),
  alert_precision = relevant_alerts / alerts_sent
)

stream_display <- stream_decisions |>
  select(
    arrival_order,
    stream_id,
    near_duplicate_of,
    matched_prior_alert,
    text,
    decision,
    reference_alert
  ) |>
  mutate(
    near_duplicate_of = coalesce(near_duplicate_of, "not a copy"),
    matched_prior_alert = coalesce(matched_prior_alert, "not suppressed")
  )

knitr::kable(
  stream_display,
  format = "html",
  escape = TRUE,
  col.names = c("Arrival", "Item", "Fixture duplicate of", "Matched prior alert", "Text", "Rule decision", "Reference alert"),
  caption = "Monitoring items are processed in arrival order",
  row.names = FALSE
)

knitr::kable(
  monitoring_results |>
    mutate(alert_precision = round(alert_precision, 3)),
  col.names = c(
    "Alerts sent",
    "Relevant alerts",
    "False alerts",
    "Relevant items missed",
    "Suppressed duplicates",
    "Repeated alerts",
    "Alert precision"
  ),
  caption = "Alert precision uses alerts sent after duplicate suppression as the denominator",
  row.names = FALSE
)
```

```{r}
#| include: false

stopifnot(
  identical(stream_decisions$arrival_order, seq_len(nrow(stream_decisions))),
  any(stream_decisions$decision == "needs review: ambiguous Riverton"),
  monitoring_results$alerts_sent > 0L,
  monitoring_results$suppressed_duplicates >= 3L,
  monitoring_results$alert_precision ==
    monitoring_results$relevant_alerts / monitoring_results$alerts_sent,
  identical(round(monitoring_results$alert_precision, 3), 0.875)
)
```

Alert precision is `r monitoring_results$relevant_alerts` relevant alerts out of `r monitoring_results$alerts_sent` alerts sent. The denominator is alerts after duplicate suppression. Suppressed copies still count in the audit trail, but they do not consume a new alert slot. In a real stream, missed items are usually found by sampling non-alerts, the same idea as the elusion sample.

The ambiguity rule is crude. It sends `r stream_decisions$stream_id[str_detect(stream_decisions$text, "Tidewater seafood")]`, the Tidewater seafood item, to review even though the text itself points away from Calder Yard.

## What this does not show

This page does not give legal advice, prove a review is defensible, or estimate a real system's recall. It also leaves out privilege review, redaction, email threading, family coding, dense retrieval, and language-model relevance labeling. The labels are perfectly consistent because the author wrote them; real reviewers disagree.

The review model's scores are ranks for this workflow, not calibrated probabilities of responsiveness. The monitoring rule is a small constructed example and is not suitable for watching individual people.

## What to remember

- Keyword precision is visible after reviewing hits; keyword recall is not visible in real work without more assessment.
- Ranked review must train on reviewed records and score only unreviewed records.
- A gain curve shows responsive records found, not recall, unless a complete fictional answer key is being used.
- Draw an elusion sample once, after the stop.
- A finite-population interval matches a fixed pile sampled without replacement.
- Monitoring streams need arrival order, duplicate suppression, ambiguity handling, and a named precision denominator.

The review log goes forward with `r if (observed_missed_in_sample == 0L) "no sampled responsive records" else paste(observed_missed_in_sample, "sampled responsive records")` and a wide interval. The uncomfortable range is the lesson: stopping review does not make the unreviewed pile disappear.

## Sources

- [Blair and Maron on full-text retrieval effectiveness](https://doi.org/10.1145/3166.3197)
- [Cormack and Grossman on continuous active learning](https://doi.org/10.1145/2600428.2609601)
- [Lewis, Yang, and Frieder on recall estimation and elusion](https://doi.org/10.1145/3459637.3482415)
- [R hypergeometric distribution documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Hypergeometric.html)
- [R binomial test documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/binom.test.html)
- [TREC 2015 Total Recall Track overview](https://trec.nist.gov/pubs/trec24/papers/Overview-TR.pdf)
- [TREC 2009 Legal Track overview](https://trec.nist.gov/pubs/trec18/papers/LEGAL09.OVERVIEW.pdf)
- [Grossman and Cormack on technology-assisted review](https://jolt.richmond.edu/jolt-archive/v17i3/article11.pdf)
- [TDT pilot study final report](https://ciir.cs.umass.edu/pubfiles/ir-137.pdf)
- [Broder on resemblance and containment of documents](https://doi.org/10.1109/SEQUEN.1997.666900)
